### 2022-10-29 FDAgroup.Rmd
some new questions?

---
title: "Dynamic Prediction in Functional Concurrent Regression "
output: beamer_presentation
date: "2022-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(fcr)
library(face)
library(refund)
library(ggplot2)
library(tidyverse)

set.seed(1025)

df = content
df_train = subset(content, include == 1)
df_test = subset(content, include == 0)
```

## Dynamic prediction

- With observations up to $t_m$, predict outcomes (or probabilities of outcome) after that time point
- Prediction updates with new observations

## Functional Concurrent Regression (FCR)

- For a subject i, we observe a function over t
$$Y_i(t)=f_0(t)+b_i(t)+\epsilon_i(t)$$
- We usually observe $Y_i$ on a series of discrete $t_{ij}$

$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}$$
where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.

- Subject-specific random effect

$$b_i(t) = \Sigma_{k=1}^c u_{ik}B_k(t)$$
where $\boldsymbol{u_i}\sim N(0, \Gamma)$

## Dynamic prediction

- Use FPCA to estimate $f_0$, $\Gamma$ and $\sigma_{\epsilon}$

- For a new subject with observations up to $t_m$, estimate its score:

$$\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})$$
With the estimated score, we can predict its outcome in following time points

$$\hat{\boldsymbol{Y}} = \boldsymbol{B}^T\hat{\boldsymbol{u}} $$

## Simulated child growth data

- Predict length-for-age, observed with noise

```{r}
ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
  geom_line(show.legend = F)+
  labs(title = "Length-for-Age for the first 10 subjects",
       x = "time")
```

## FPCA on observed LAZ

```{r}
tnew = sort(unique(df_train$argvals))
df_laz <- df_train %>% select(argvals, subj, Y) %>%
  rename("y" = "Y")
fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15, argvals.new = tnew, calculate.scores = TRUE)

# mean function

par(mfrow = c(2, 2))
plot(tnew, fpca_laz$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "t", ylab = "PC1")
plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "t", ylab = "PC2")
plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "t", ylab = "PC3")

```

## Prediction of new partially observed sample

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.3] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.3, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.3]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

# figure 
bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.3", 
       x = "time", y = "")

```

## Prediction of new partially observed sample

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.5] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.5, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.5]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

# figure

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.5", 
       x = "time", y = "")

```

## Prediction of new partially observed sample

```{r}
df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
  rename("y" = "Y")

# observe up to t = 0.3
df_laz_test$y[df_laz_test$argvals > 0.8] <- NA
pred1 <- predict(fpca_laz, newdata = df_laz_test)

# prediction
Bnew <- fpca_laz$eigenfunctions[tnew > 0.8, ]
Yhat <- Bnew %*% t(pred1$rand_eff$scores)
# every column is a new function

Yhat <- data.frame(Yhat)
colnames(Yhat) <- pred1$rand_eff$subj

df_laz_pred <- Yhat %>% 
  mutate(argvals = tnew[tnew > 0.8]) %>% 
  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
  mutate(type = "pred")

# figure

bind_rows(df_test %>% 
            select(argvals, subj, Y)%>% 
            rename("y" = "Y") %>% 
            mutate(type = "true", subj = as.factor(subj)), 
          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
  filter(subj %in% 151:156) %>% 
  ggplot(aes(x = argvals, y = y, col = type))+
  geom_line()+
  facet_wrap(~subj)+
  labs(title = "Prediction with observation up to t=0.8", 
       x = "time", y = "")

```

## Next steps

- Establish interval prediction
- ? fPCA on pooled prediction, instead of original functions observed with noise
- Extension to exponential family data

### 2022-10-31 FDAgroup.Rmd
updated FDA group presentation slides

@@
 ---
-title: "Dynamic Prediction in Functional Concurrent Regression "
+title: "Dynamic Prediction with fPCA "
 output: beamer_presentation
-date: "2022-10-27"
+date: "2022-11-3"
 ---
@@
 
+- Goal: to predict future track based on observed track
+
 - For a subject i, we observe a function over t
@@
 
-$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}$$
+$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}, \hspace{0.5cm} j = 1...J_i$$
 where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.
@@
 
-- Use FPCA to estimate $f_0$, $\Gamma$ and $\sigma_{\epsilon}$
+- When there is no covariate in the model, this is essentially a fPCA problem.
+
+  - $\boldsymbol{B}$ is a matrix of eigenfunctions
+  - $\boldsymbol{u}$ is a matrix of PC scores/loadings
+
+- Use fPCA to estimate $f_0$, $\Gamma$ and $\sigma_{\epsilon}$
 
@@
 
-$$\hat{\boldsymbol{Y}} = \boldsymbol{B}^T\hat{\boldsymbol{u}} $$
+$$\hat{\boldsymbol{Y}} = \hat{\boldsymbol{f}}_0+ \boldsymbol{B}^T\hat{\boldsymbol{u}} $$
 
@@
 
-```{r}
-df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y")
+```{r, pred_func}
+# write a function for prediction up to t
+# df_test: datafram with columns argvals, subj, y, observation up to tm
 
-# observe up to t = 0.3
-df_laz_test$y[df_laz_test$argvals > 0.3] <- NA
-pred1 <- predict(fpca_laz, newdata = df_laz_test)
-
+DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){
+  Ntest = length(unique(df_test$subj))
   # prediction
-Bnew <- fpca_laz$eigenfunctions[tnew > 0.3, ]
-Yhat <- Bnew %*% t(pred1$rand_eff$scores)
+  pred1 <- predict(fpca_fit, newdata = df_test)
+  # Eigen functions
+  Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ]
+  Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores)
   # every column is a new function
-
   Yhat <- data.frame(Yhat)
   colnames(Yhat) <- pred1$rand_eff$subj
+  df_pred <- Yhat %>% 
+    mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>% 
+    pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>%
+    mutate(type = "pred")
 
-df_laz_pred <- Yhat %>% 
-  mutate(argvals = tnew[tnew > 0.3]) %>% 
-  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
-  mutate(type = "pred")
+  return(df_pred)
+}
+```
+
+```{r t0_3}
+df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
+  rename("y" = "Y") %>% 
+  filter(argvals <= 0.3))
 
@@
             mutate(type = "true", subj = as.factor(subj)), 
-          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
+          df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>% 
   filter(subj %in% 151:156) %>% 
@@
        x = "time", y = "")
-
 ```
@@
 ```{r}
-df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y")
-
-# observe up to t = 0.3
-df_laz_test$y[df_laz_test$argvals > 0.5] <- NA
-pred1 <- predict(fpca_laz, newdata = df_laz_test)
-
-# prediction
-Bnew <- fpca_laz$eigenfunctions[tnew > 0.5, ]
-Yhat <- Bnew %*% t(pred1$rand_eff$scores)
-# every column is a new function
-
-Yhat <- data.frame(Yhat)
-colnames(Yhat) <- pred1$rand_eff$subj
-
-df_laz_pred <- Yhat %>% 
-  mutate(argvals = tnew[tnew > 0.5]) %>% 
-  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
-  mutate(type = "pred")
+df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
+  rename("y" = "Y") %>% 
+  filter(argvals <= 0.5), tm = 0.5)
 
@@
             mutate(type = "true", subj = as.factor(subj)), 
-          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
+          df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>% 
   filter(subj %in% 151:156) %>% 
@@
 ```{r}
-df_laz_test <- df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y")
-
-# observe up to t = 0.3
-df_laz_test$y[df_laz_test$argvals > 0.8] <- NA
-pred1 <- predict(fpca_laz, newdata = df_laz_test)
-
-# prediction
-Bnew <- fpca_laz$eigenfunctions[tnew > 0.8, ]
-Yhat <- Bnew %*% t(pred1$rand_eff$scores)
-# every column is a new function
-
-Yhat <- data.frame(Yhat)
-colnames(Yhat) <- pred1$rand_eff$subj
-
-df_laz_pred <- Yhat %>% 
-  mutate(argvals = tnew[tnew > 0.8]) %>% 
-  pivot_longer(1:50, names_to = "subj", values_to = "y") %>%
-  mutate(type = "pred")
+df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
+  rename("y" = "Y") %>% 
+  filter(argvals <= 0.8), tm = 0.8)
 
 # figure 
-
 bind_rows(df_test %>% 
@@
             mutate(type = "true", subj = as.factor(subj)), 
-          df_laz_pred %>% mutate(subj = as.factor(subj))) %>% 
+          df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>% 
   filter(subj %in% 151:156) %>% 

### 2022-11-02 FDAgroup.Rmd
FDA presentation

@@
 library(ggplot2)
+theme_set(theme_minimal())
 library(tidyverse)
@@
 - Prediction updates with new observations
+- Mixed model prediction is difficult
+  - Not flexible enough for densely measured data
+  - Out-of-sample random effects cannot be estimated
+- Functional mixed effects model
 
@@
 
-- Goal: to predict future track based on observed track
+- Goal: to predict future track based on partially observed track
 
@@
 $$Y_i(t)=f_0(t)+b_i(t)+\epsilon_i(t)$$
-- We usually observe $Y_i$ on a series of discrete $t_{ij}$
-
-$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}, \hspace{0.5cm} j = 1...J_i$$
-where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.
 
@@
 
-## Dynamic prediction
+- We usually observe $Y_i$ on a series of discrete $t_{ij}$
+
+$$Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}, \hspace{0.5cm} j = 1...J_i$$
+where $\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})$.
+
+## Connection to fPCA
 
@@
 
-$$\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})$$
+$$\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}\boldsymbol{y}$$
 With the estimated score, we can predict its outcome in following time points
@@
 
-- Predict length-for-age, observed with noise
+- Predict length-for-age z-score (LAZ), observed with noise
+  - Calculated using the age- and sex-specific WHO standard references
 
@@
   geom_line(show.legend = F)+
-  labs(title = "Length-for-Age for the first 10 subjects",
-       x = "time")
+  labs(title = "Length-for-Age z-score tracks for the 10 subjects",
+       x = "time", y ="LAZ")
 ```
@@
   rename("y" = "Y")
-fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15, argvals.new = tnew, calculate.scores = TRUE)
+fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15,
+                        argvals.new = tnew, calculate.scores = TRUE)
 
@@
      xlab = "t", ylab = "f0(t)")
-plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "t", ylab = "PC1")
-plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "t", ylab = "PC2")
-plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "t", ylab = "PC3")
+plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
+plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
+plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")
 
@@
 
+## Extension to non-Gaussian data
+
+- fPCA on non-Gaussian data is very difficult and computationally intensive
+- fPCA on a latent Gaussian variable instead
+  - transformation using link function: $g(E(Y))=\boldsymbol{X}^T\boldsymbol{\beta}$
+- Get a smooth latent Gaussian function by pooling a series of GLMM
+  - Assume we have regularly observed functions $Y_{ij}$, $i = 1... N$ and $j = 1...J$
+  - Bin the functional domain into small intervals $t\in {1...T}$
+  - Fit a GLMM in each bin to get a latent variables
+
+    $$g(E(Y_{it})) = \beta_{0t}+b_{it}$$
+    Let $Z_{it} = g(E(Y_{it}))$, we can do Gaussian fPCA on this latent variable
+
+## Extension to non-Gaussian data
+
+$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$
+
+$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$
+
+## Extension to non-Gaussian data
+
+```{r gen_bi_outcome}
+times <- seq(0, 1, by = 0.001)
+gen_data <- function(id = 1){
+  scores <- rnorm(4)
+  pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times)
+  Z <- pcs %*% scores
+  probs <- exp(Z)/(1+exp(Z))
+  Y <- sapply(probs, rbinom, n = 1, size = 1)
+
+  return(data.frame(id = id, argvals = times, Y=Y, Z=Z))
+}
+
+df_train2 <- bind_rows(lapply(1:20, gen_data))
+
+ggplot(df_train2 %>% filter(id %in% 1:4))+
+  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
+            show.legend = F)+
+  geom_line(aes(x=argvals, y=Z, col = "Red"), 
+            show.legend = F)+
+  facet_wrap(~id)+
+  labs(title = "Simulated function for the 4 subjects",
+       x = "time", y ="")
+```
+
+## Extension to non-Gaussian data
+
+```{r, message=FALSE}
+# series of mixed models
+## bin time
+cutbrk <- seq(0, 1, by = 0.01)
+cutlabs <- (cutbrk+0.01/2)[1:100]
+df_train2$time_bin <- cut(df_train2$argvals, breaks = cutbrk, 
+                include.lowest = T,
+                labels = cutlabs)
+df_train2$time_bin <- as.numeric(as.character(df_train2$time_bin))
+
+# intercept only in each bin mixed models
+df_train2_lst <- split(df_train2, f = df_train2$time_bin)
+pred_latent <- function(df){
+  fit_glm <- lme4::glmer(Y ~ 1 + (1|id), 
+                         data = df, family = binomial)
+  Zhat <- predict(fit_glm, type = "link")
+  return(Zhat)
+}
+
+allZhat <- unlist(sapply(df_train2_lst, pred_latent))
+df_train_sm <- data.frame(subj = df_train2$id, 
+                          argvals = df_train2$time_bin, 
+                          y = allZhat) %>% 
+  distinct()
+
+ggplot(df_train_sm %>% filter(subj %in% 1:4))+
+  geom_line(aes(x=argvals, y=y), 
+            show.legend = F)+
+  facet_wrap(~subj)+
+  labs(title = "Estimated latent function tracks for the 4 subjects",
+       x = "time", y ="")
+
+```
+
+## Extension to non-Gaussian data
+
+```{r}
+# fPCA
+fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15,
+                        argvals.new = times, calculate.scores = TRUE)
+
+# plot
+par(mfrow = c(2, 2))
+plot(times, fpca_Z$mu.new, type = "l", 
+     xlab = "t", ylab = "f0(t)")
+plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
+plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
+plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")
+
+```
+
 ## Next steps
 
-- Establish interval prediction
-- ? fPCA on pooled prediction, instead of original functions observed with noise
-- Extension to exponential family data
+- Simulate a non-Gaussian function and implement methods above
+  - Improve estiation of latent functions very well
+  - Numeric problems
+- Transform latent function back to non-Gaussian function
+- Inclusion of covariates

### 2022-11-03 FDAgroup.Rmd
update rmd files for process tracking

@@
 
-```{r, message=FALSE}
-# series of mixed models
-## bin time
-cutbrk <- seq(0, 1, by = 0.01)
-cutlabs <- (cutbrk+0.01/2)[1:100]
-df_train2$time_bin <- cut(df_train2$argvals, breaks = cutbrk, 
-                include.lowest = T,
-                labels = cutlabs)
-df_train2$time_bin <- as.numeric(as.character(df_train2$time_bin))
-
-# intercept only in each bin mixed models
-df_train2_lst <- split(df_train2, f = df_train2$time_bin)
-pred_latent <- function(df){
-  fit_glm <- lme4::glmer(Y ~ 1 + (1|id), 
-                         data = df, family = binomial)
-  Zhat <- predict(fit_glm, type = "link")
-  return(Zhat)
-}
-
-allZhat <- unlist(sapply(df_train2_lst, pred_latent))
-df_train_sm <- data.frame(subj = df_train2$id, 
-                          argvals = df_train2$time_bin, 
-                          y = allZhat) %>% 
-  distinct()
-
-ggplot(df_train_sm %>% filter(subj %in% 1:4))+
-  geom_line(aes(x=argvals, y=y), 
-            show.legend = F)+
-  facet_wrap(~subj)+
-  labs(title = "Estimated latent function tracks for the 4 subjects",
-       x = "time", y ="")
-
-```
-
-## Extension to non-Gaussian data
-
 ```{r}

### 2022-11-03 f-GFPCA.Rmd
update rmd files for process tracking

---
title: "fGFPCA dynamic prediction"
output: html_document
date: "2022-11-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(here)
```

## Top simulation with a very simple function

```{r, message=FALSE}
source(here("Code/ToySimulation.R"))
```

- Below we generate 100 latent functions and their corresponding binary functions

$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$

$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$

```{r}
## four subjects
ggplot(df_train2 %>% filter(id %in% 15:18))+
  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
             show.legend = F)+
  geom_line(aes(x=argvals, y=Z, col = "Red"), 
            show.legend = F)+
  facet_wrap(~id)+
  labs(title = "Simulated function for the 4 subjects",
       x = "time", y ="")

## all subjects latent functions
ggplot(df_train2)+
  geom_line(aes(x=argvals, y=Z, col = as.factor(id)), 
            show.legend = F)+
  labs(title = "Simulated latent functions for all subjects")

```

- Bloew is the predicted latent functions from local GLMM models

```{r}

ggplot(df_pred_latent %>% filter(id %in% 15:18))+
  geom_line(aes(x = argvals, y=Zhat))+
  geom_line(aes(x=argvals, y = Z, col = "red"))+
  geom_point(aes(x=argvals, y = Y, col = "blue"), size = 0.01)+
  facet_wrap(~id)

```

### 2022-11-03 Code/ToySimulation.R

@@ -21,22 +21,7 @@ gen_data <- function(id = 1){
 }
 
 df_train2 <- bind_rows(lapply(1:100, gen_data))
-
-## four subjects
-ggplot(df_train2 %>% filter(id %in% 1:4))+
-  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
-             show.legend = F)+
-  geom_line(aes(x=argvals, y=Z, col = "Red"), 
-            show.legend = F)+
-  facet_wrap(~id)+
-  labs(title = "Simulated function for the 4 subjects",
-       x = "time", y ="")
-
-## all subjects
-
-ggplot(df_train2)+
-  geom_line(aes(x=argvals, y=Z, col = as.factor(id)), 
-            show.legend = F)
+table(df_train2$id)
 
 
 ##### Local GLMMs #####
@@ -52,41 +37,32 @@ df_train2$time_bin <- cut(df_train2$argvals, breaks = brks, include.lowest = T, 
 df_train2$time_bin <- as.numeric(as.character(df_train2$time_bin))
 df_train2$id <- as.factor(df_train2$id)
 
+table(df_train2$id, df_train2$time_bin)
+
+
 # fit local linear mixed models in each bin
-## df: function with binary outcome Y, id and time_bin
-pred_latent <- function(df = df_train2, t = 0.005){
-  this_df <- df[df$time_bin==t, ]
-  this_glm <- glmer(Y ~ 1 + (1|id), data = this_df, family = binomial)
-  coef(this_glm)
+## df: function with binary outcome Y, with observations in one time bein
+pred_latent <- function(df){
+  this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
   Zhat <- predict(this_glm, type = "link")
-  return(Zhat)
+  df$Zhat <- Zhat
+  return(df)
 }
 
-df_try <- df_train2 %>% select(id, Z, argvals, time_bin) %>% 
-  filter(time_bin == 0.025) %>% 
-  mutate(Zhat = pred_latent(t = 0.025))
 
-ggplot(df_try %>% filter(id %in% 1:8))+
-  geom_line(aes(x = argvals, y=Z, col = "red"))+
-  geom_line(aes(x = argvals, y=Zhat, cold = "blue"))+
-  facet_wrap(~id, )
-
-allZhat <- unlist(sapply(df_train2_lst, pred_latent))
[…]

### 2022-11-08 FDAgroup.Rmd
update token

@@
           df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>% 
-  filter(subj %in% 151:156) %>% 
+  filter(subj == 152) %>% 
   ggplot(aes(x = argvals, y = y, col = type))+

### 2022-11-09 f-GFPCA.Rmd
fix typos

@@
 
-## Top simulation with a very simple function
+## Toy simulation with a very simple function
 
@@
 
-- Bloew is the predicted latent functions from local GLMM models
+- Below is the predicted latent functions from local GLMM models
 
@@
 ```
+
+# Next steps:
+
+- Use fastGFPCA to generate more complicated functions
+- What happened to the latent gaussian function? Are they just noisy versions of true latent functions? Or it there something wrong with GLMMs? 
+- Establish prediction intervals of latent gaussian function
+- Write a function to calculate random effects (instead of using predict.fpca.sparse)

### 2022-11-11 f-GFPCA.Rmd
update simulation scheme

@@
 
-## Toy simulation with a very simple function
+## Toy simulation
+
+### Simulation set-up
 
@@
 
-- Below we generate 100 latent functions and their corresponding binary functions
+- For a random sample i:
 
-$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$
+$$\eta_i(t) =f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)$$
 
-$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$
+where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$
+
+$$Y_i(t) \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))})$$
+
+In this simulation, we set $f_0(t) = 0$. In the fPCA framework, $\Gamma$ is a diagonal matrix and diagonal elements correponds to eigen values.
 
 ```{r}
-## four subjects
-ggplot(df_train2 %>% filter(id %in% 15:18))+
-  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
-             show.legend = F)+
-  geom_line(aes(x=argvals, y=Z, col = "Red"), 
-            show.legend = F)+
+df %>% 
+  filter(id %in% 1:4) %>% 
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_i))+
+  geom_point(aes(x=sind, y=Y), size = 0.05)+
   facet_wrap(~id)+
-  labs(title = "Simulated function for the 4 subjects",
-       x = "time", y ="")
-
-## all subjects latent functions
-ggplot(df_train2)+
-  geom_line(aes(x=argvals, y=Z, col = as.factor(id)), 
-            show.legend = F)+
-  labs(title = "Simulated latent functions for all subjects")
-
+  labs(title = "Four simulated subjects")
 ```
 
-- Below is the predicted latent functions from local GLMM models
+### Local GLMMS
 
 ```{r}
+source(here("Code/GLMM-FPCA.R"))
+```
 
-ggplot(df_pred_latent %>% filter(id %in% 15:18))+
-  geom_line(aes(x = argvals, y=Zhat))+
-  geom_line(aes(x=argvals, y = Z, col = "red"))+
-  geom_point(aes(x=argvals, y = Y, col = "blue"), size = 0.01)+
-  facet_wrap(~id)
+```{r}
+par(mfrow=c(3, 2))
+## mean functions
+plot(mid, fpca_fit$mu, ylab = "mean")
 
+## eigenfunctions
+plot(mid, fpca_fit$efunctions[, 1], ylab='PC1')
+plot(mid, fpca_fit$efunctions[, 2], ylab='PC2')
+plot(mid, fpca_fit$efunctions[, 3], ylab='PC3')
+plot(mid, fpca_fit$efunctions[, 4], ylab='PC4')
 ```
+
+```{r}
+ggplot(df_pred_latent %>% filter(id %in% 1:4))+
+  geom_line(aes(x = sind_inx, y=eta_hat))+
+  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
+  geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
+  facet_wrap(~id)+
+  labs(title = "Predicted latent functions for four subjects")
+```
+
+### Prediction
+
+Assume we have a new observation with outcome $\boldsymbol{Y}$ up to $t_m$. If we have an Gaussian function, the scores of this subject can be estimated with empirical Bayes:
+
+$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}\boldsymbol{Y}$$
+Where $\hat{\boldsymbol{\Gamma}}$ is the diagonal matrix of eigenvalues, $\boldsymbol{\Phi}$ is the matrix of eigenfunctions. These should be constant across all subjects.
+
+However, for binary data with latent function, there are a couple of differences:
+
+1. We do not have a $\sigma_{\epsilon}^2$ in the model formula
+
+2. We do not have an Gaussian outcome $\boldsymbol{Y}$. The $\boldsymbol{Y}$ should be replaced with the values of latent function $\eta(t)$. However, we also could not observe the value of $\eta(t)$.
+
+Two options I am thinking of:
+
+1. If we'd like to make predictions on more than one sample, we may be able to fit GLMM on test samples and predict $\eta(t)$
+2. Do a different set of derivation with MLE+EB like [this paper](https://www.jstor.org/stable/pdf/2533715.pdf?refreqid=excelsior%3A135dd765c41e4b46eb39bb6ef7a71053&ab_segments=&origin=&acceptTC=1)
+
+Another thing I have been thinking of is how to extend the grid of eigenfunctions from binned-grid to full-grid.
+I am using interpolation now, but only because I am not sure how to do it with basis functions.
+It looks like we will need K more sparse PCA, but it will introduce a lot more scores.
 

### 2022-11-11 Code/GFPCA_simulate_data_Ying.R

@@ -1,0 +1,64 @@
+# clear workspace
+rm(list=ls())
+# set seed
+set.seed(1090)
+# load and, if necessary, install packages used in simulating data 
+# (some of these you'll need for model fitting)
+pckgs <- c("tidyverse","mgcv","refund","lme4")
+invisible(
+        sapply(pckgs, function(x) if(!require(x,character.only=TRUE,quietly=TRUE)) {
+                install.packages(x)
+                require(x, character.only=TRUE,quietly=TRUE)
+        })
+)
+
+### Simulate data ###
+
+## basic parameters
+# number of subjects
+N <- 500
+# number of observations per subject (observations per function)
+J <- 1000
+# functional domain, equally spaced grid on 0,1
+sind <- seq(0,1,len=J)
+
+## simulate outcomes g(E[Y_i(s)]) = \eta_i(s) = f_0(s) + f_1(s)X_i + b_i(s)
+# f_0(s) = 0
+# b_i(s) = \sum_{k=1}^4 \phi_k(s)\xi_{ik}
+#        \xi_{ik} \sim N(0, \lambda_k)
+
+## fixed effects f_0, f_1
+f_0 <- function(s) 0
+
+## random effects
+# eigenfunctions \phi_k(s) evaluated on the observed grid
+phi <- sqrt(2)*cbind(sin(2*pi*sind),cos(2*pi*sind),
+                     sin(4*pi*sind),cos(4*pi*sind))
+# eigenvalues \lambda
+K <- 4
+lambda <- 0.5^(0:(K-1))
+# subject-specific weights/coefficients \xi_ik
+xi <- matrix(rnorm(N*K),N,K);
+xi <- xi %*% diag(sqrt(lambda))
+b_i <- xi %*% t(phi); # of size N by J
+
+## linear predictor \eta_i(s)
+eta_i <- t(vapply(1:N, function(x){
+                        f_0(sind) + b_i[x,]
+                }, numeric(J)))
+## response Y_i(s) \sim Binom(p = expit(eta_i))
+Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
+              N, J, byrow=FALSE)
+
+## create data matrix in long format for estimation 
+# id = subject identifier (factor variable)
+# sind = numeric value corresponding to the observed functional domain
+# Y = functional response (binary)
+# sind_inx = numeric value associated with order of "sind"
+#            this is not necessary, but may be of convenience when you implement the method
[…]

### 2022-11-11 Code/GLMM-FPCA.R

@@ -1,0 +1,86 @@
+##### Local GLMMs #####
+
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+# bin observations
+df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
+df$bin <- as.numeric(as.character(df$bin))
+
+head(df, 20)
+
+# check overlap
+# df %>% 
+#   filter(id %in% 1:4) %>% 
+#   ggplot()+
+#   geom_line(aes(x=sind, y=eta_i, col=as.factor(bin)), show.legend = F)+
+#   geom_point(aes(x=sind, y=Y, col=as.factor(bin)), size = 0.05, show.legend = F)+
+#   facet_wrap(~id)
+
+# fit local linear mixed models in each bin
+## df: function with binary outcome Y, with observations in one time bein
+pred_latent <- function(df){
+  this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+  eta_hat <- predict(this_glm, type = "link")
+  df$eta_hat <- eta_hat
+  return(df)
+}
+
+## do that for all time bin
+df_bin_lst <- split(df, f = df$bin)
+df_pred_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+lapply(df_pred_latent, dim)
+df_pred_latent <- bind_rows(df_pred_latent)
+
+#### fPCA on predicted latent variables #####
+rm(df_bin_lst)
+
+# put predicted latent function to wide format
+df_pred_latent <- df_pred_latent %>% arrange(id, sind_inx)
+df_pred_unique <- df_pred_latent %>% select(id, bin, eta_hat) %>% distinct()
+
+mat_pred_unique <- matrix(df_pred_unique$eta_hat, nrow=N, ncol=n_bin, byrow = T)
+# dim(mat_pred_unique)
+
+# fPCA
+fpca_fit <- fpca.face(mat_pred_unique, pve = 0.95, argvals = unique(df_pred_unique$bin), knots=20)
+# fpca_fit <- face.sparse(df_pred_unique, argvals.new = 1:J, knots = 20, pve=0.95)
+
+
+#### Project eigen functions to a denser grid #####
[…]

### 2022-11-11 Code/ToySimulation.R

@@ -2,68 +2,67 @@
 set.seed(1025)
 ##### generate data #####
 
+rm(list=ls())
 library(tidyverse)
 library(ggplot2)
 library(fcr)
 library(refund)
 library(lme4)
 
-#####  generate data for one subject ####
-times <- seq(0, 1, by = 0.001)
-gen_data <- function(id = 1){
-  scores <- rnorm(4)
-  pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times)
-  Z <- pcs %*% scores
-  probs <- exp(Z)/(1+exp(Z))
-  Y <- sapply(probs, rbinom, n = 1, size = 1) 
-  
-  return(data.frame(id = id, argvals = times, Y=Y, Z=Z))
-}
 
-df_train2 <- bind_rows(lapply(1:100, gen_data))
-table(df_train2$id)
+#####  generate data f ####
 
+## basic parameters
+# number of subjects
+N <- 500
+# number of observations per subject (observations per function)
+J <- 1000
+# functional domain, equally spaced grid on 0,1
+sind <- seq(0,1,len=J)
 
-##### Local GLMMs #####
+## simulate outcomes g(E[Y_i(s)]) = \eta_i(s) = f_0(s) + f_1(s)X_i + b_i(s)
+# f_0(s) = 0
+# b_i(s) = \sum_{k=1}^4 \phi_k(s)\xi_{ik}
+#        \xi_{ik} \sim N(0, \lambda_k)
 
-# bins
-w <- 0.01 # bin width
-brks <- seq(0, 01, by = w) # cutoff points
-mid <- (brks+w/2)[1:100] # mid points
-nb <- length(mid) # number of bins
+## fixed effects f_0, f_1
+f_0 <- function(s) 0
 
-# bin observations
-df_train2$time_bin <- cut(df_train2$argvals, breaks = brks, include.lowest = T, labels = mid)
-df_train2$time_bin <- as.numeric(as.character(df_train2$time_bin))
-df_train2$id <- as.factor(df_train2$id)
+## random effects
+# eigenfunctions \phi_k(s) evaluated on the observed grid
+phi <- sqrt(2)*cbind(sin(2*pi*sind),cos(2*pi*sind),
+                     sin(4*pi*sind),cos(4*pi*sind))
+# eigenvalues \lambda
+K <- 4
+lambda <- 0.5^(0:(K-1))
[…]

### 2022-11-15 f-GFPCA.Rmd
modify in-sample formula

@@
 
+#### FPCA
+
 ```{r}
@@
 
-```{r}
+```{r, fig.height=12, fig.width=8}
 par(mfrow=c(3, 2))
@@
 ```
+#### Estimated latent function
 
@@
 
-$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}\boldsymbol{Y}$$
+$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}(\hat{\boldsymbol{\eta}}-\hat{\boldsymbol{f}}_0)$$
 Where $\hat{\boldsymbol{\Gamma}}$ is the diagonal matrix of eigenvalues, $\boldsymbol{\Phi}$ is the matrix of eigenfunctions. These should be constant across all subjects.

### 2022-11-21 f-GFPCA.Rmd
try in-sample prediction (point)

--- 2022-11-09 f-GFPCA.Rmd
@@
 ---
-title: "fGFPCA dynamic prediction"
+title: "fGFPCA dynamic prediction (In-sample)"
 output: html_document
@@
 library(here)
+library(tidyverse)
+library(ggplot2)
+library(fcr)
+library(refund)
+library(lme4)
+set.seed(1025)
 ```
 
-## Toy simulation with a very simple function
+## Toy simulation
+
+### Simulation set-up
 
@@
 
-- Below we generate 100 latent functions and their corresponding binary functions
+- For a random sample i:
 
-$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$
+$$\eta_i(t) =f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)$$
 
-$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$
+where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$. In the fPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix and diagonal elements corresponds to eigenvalues.
+
+$$Y_i(t) \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))})$$
+
+In this simulation, we set $f_0(t) = 0$, $\boldsymbol{\Gamma}$ is a diagonal matrix with diagonal elements {1, 0.5, 0.25, 0.125}.
 
 ```{r}
-## four subjects
-ggplot(df_train2 %>% filter(id %in% 15:18))+
-  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
-             show.legend = F)+
-  geom_line(aes(x=argvals, y=Z, col = "Red"), 
-            show.legend = F)+
+df %>% 
+  filter(id %in% 1:4) %>% 
+  mutate(probs = exp(eta_i)/(1+exp(eta_i))) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_i), col = "blue")+
+  geom_line(aes(x=sind, y=probs), col = "red")+
+  geom_point(aes(x=sind, y=Y), size = 0.05)+
   facet_wrap(~id)+
-  labs(title = "Simulated function for the 4 subjects",
-       x = "time", y ="")
+  labs(title = "Four simulated subjects")
+```
 
-## all subjects latent functions
-ggplot(df_train2)+
-  geom_line(aes(x=argvals, y=Z, col = as.factor(id)), 
-            show.legend = F)+
-  labs(title = "Simulated latent functions for all subjects")
+### Local GLMMS
+
+The entire time grid into 100 small intervals with equal length.
+The cut is made on the order index of measurement time, but not the actually time itself.
+Each bin its labeled by its midpoint (5, 15, 25, ..., 995).
+
+Within each interval labeled as $s$, we fit a local GLMM model:
+
+$$logit(E(Y_i^s)) = \beta_0^s+b_i^s$$
+Then we use this series of models to estimate the latent Gaussian function. Each subject will have a different value of latent function at each bin.
+
+```{r}
+source(here("Code/GLMM-FPCA.R"))
+```
+
+Figure below shows the estimated latent function for subjects 1-4:
+
+```{r}
+ggplot(df_pred_latent %>% filter(id %in% 1:4))+
+  geom_line(aes(x = sind_inx, y=eta_hat))+
+  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
+  # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
+  facet_wrap(~id)+
+  labs(title = "Predicted latent functions on original grid")
+```
+```{r}
+df_pred_latent %>% filter(id %in% 1:4 & sind_inx %in% mid) %>%
+  ggplot()+
+  geom_line(aes(x = sind_inx, y=eta_hat))+
+  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
+  # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
+  facet_wrap(~id)+
+  labs(title = "Predicted latent functions on binned grid")
 
@@
 
-- Below is the predicted latent functions from local GLMM models
+### FPCA
+
+Here we fit FPCA on the estimated latent function from the previous step:
 
 ```{r}
+## mean functions
+plot(mid, fpca_fit$mu, ylab = "mean")
+```
 
-ggplot(df_pred_latent %>% filter(id %in% 15:18))+
-  geom_line(aes(x = argvals, y=Zhat))+
-  geom_line(aes(x=argvals, y = Z, col = "red"))+
-  geom_point(aes(x=argvals, y = Y, col = "blue"), size = 0.01)+
-  facet_wrap(~id)
+```{r, fig.height=8, fig.width=8}
+
+## eigenfunctions
+par(mfrow=c(2, 2))
+plot(mid, fpca_fit$efunctions[, 1], ylab='PC1')
+plot(mid, fpca_fit$efunctions[, 2], ylab='PC2')
+plot(mid, fpca_fit$efunctions[, 3], ylab='PC3')
+plot(mid, fpca_fit$efunctions[, 4], ylab='PC4')
+```
+
+### In-sample prediction
+
+Now let's try to predict future outcomes based on observations $\boldsymbol{Y}$ up to $t_m$.
+
+For in-sample prediction, we have observed the full outcome track, thus have estimated the full latent Gaussian function $\eta(s)$ on binned grid. 
+For now, we will take the quantities up to $t_m$ and re-estimate the scores with empirical Bayes:
+
+$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}(\hat{\boldsymbol{\eta}}-\hat{\boldsymbol{f}}_0)$$
+Where:
+
+- $\hat{\boldsymbol{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
+- $\boldsymbol{\Phi}$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
+- $\hat{\sigma_{\epsilon}}^2$ is the residual variance from FPCA
+- $\hat{\boldsymbol{\eta}}$ is the estimated latent Gaussian function up to $t_m$
+- $\hat{\boldsymbol{f}}_0$ is the FPCA mean function up to $t_m$
+
+Below we try to do that with one subject:
+
+- Subject 1
+
+```{r}
+source(here("code/DynPred.R"))
+```
+
+```{r}
+# function to clean results
+# make predictions with partial track up to 20, 40, 60
+clean_pred <- function(subj = 1){
+  pred20 <- data.frame(bin = mid, 
+                       eta_i = in_samp_dyn_pred(fpca_fit, 20,
+                                                df_pred_unique, subj),
+                       type = "up to 195") %>% 
+    filter(bin>195)
+
+  pred40 <- data.frame(bin = mid, 
+                      eta_i = in_samp_dyn_pred(fpca_fit, 40,
+                                               df_pred_unique, subj),
+                       type = "up to 395") %>%
+    filter(bin>395)
+
+  pred60 <- data.frame(bin = mid, 
+                       eta_i = in_samp_dyn_pred(fpca_fit, 60,
+                                                df_pred_unique, subj),
+                        type = "up to 595") %>%
+    filter(bin>595)
+
+  df_pred <- df_pred_latent %>% filter(id==subj & sind_inx==bin) %>%
+    select(bin, eta_i) %>% mutate(type="true") %>%
+    add_row(pred20) %>%
+    add_row(pred40) %>% 
+    add_row(pred60)
+
+  return(df_pred)
+}
+
+```
+
+```{r}
+# Try on subjectw 1-4
+
+clean_pred(1) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+
+clean_pred(2) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+
+clean_pred(3) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+
+clean_pred(4) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+
+```
+- Subject 2
+
+```{r}
+# Try on subject 1
+pred20 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=2),
+                     type = "up to 195") %>% 
+  filter(bin>195)
+
+pred40 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=2),
+                     type = "up to 395") %>%
+  filter(bin>395)
+
+pred60 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=2),
+                     type = "up to 595") %>%
+  filter(bin>595)
+
+clean_pred(1)
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+```
+
+- Subject 3
+
+```{r}
+# Try on subject 1
+pred20 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=3),
+                     type = "up to 195") %>% 
+  filter(bin>195)
+
+pred40 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=3),
+                     type = "up to 395") %>%
+  filter(bin>395)
+
+pred60 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=3),
+                     type = "up to 595") %>%
+  filter(bin>595)
+
+df_pred_latent %>% filter(id==3 & sind_inx==bin) %>%
+  select(bin, eta_i) %>% mutate(type="true") %>%
+  add_row(pred20) %>%
+  add_row(pred40) %>% 
+  add_row(pred60) %>% 
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
+```
+- Subject 4
+
+```{r}
+# Try on subject 1
+pred20 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=4),
+                     type = "up to 195") %>% 
+  filter(bin>195)
+
+pred40 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=4),
+                     type = "up to 395") %>%
+  filter(bin>395)
+
+pred60 <- data.frame(bin = mid, 
+                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=4),
+                     type = "up to 595") %>%
+  filter(bin>595)
+
+df_pred_latent %>% filter(id==4 & sind_inx==bin) %>%
+  select(bin, eta_i) %>% mutate(type="true") %>%
+  add_row(pred20) %>%
+  add_row(pred40) %>% 
+  add_row(pred60) %>% 
+  ggplot()+
+  geom_line(aes(x=bin, y = eta_i, col=type))
 
@@
 
-- Use fastGFPCA to generate more complicated functions
-- What happened to the latent gaussian function? Are they just noisy versions of true latent functions? Or it there something wrong with GLMMs? 
-- Establish prediction intervals of latent gaussian function
-- Write a function to calculate random effects (instead of using predict.fpca.sparse)
+1. Interval prediction
+2. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
+3. Out-of-sample prediction: how to estimate scores?
+
+Thoughts on 3:
+
+1. If we'd like to make predictions on more than one sample, we may be able to fit GLMM on test samples and predict $\eta(t)$
+2. Do a different set of derivation with MLE+EB like [this paper](https://www.jstor.org/stable/pdf/2533715.pdf?refreqid=excelsior%3A135dd765c41e4b46eb39bb6ef7a71053&ab_segments=&origin=&acceptTC=1)

### 2022-11-21 Code/DynPred.R

@@ -1,0 +1,36 @@
+
+##### In-sample prediction on binned grid #####
+# fpca_fit: fpca model fit on estimated latent Gaussian function
+# tm: maximum observation time on binned grid, order index of bin
+# df_pred_unique: data frame with only id, bin, eta_hat
+# subj: subject ID to make prediction on
+# mid: label of binned grid
+
+# return: predictions on the full binned grid
+
+in_samp_dyn_pred <- function(fpca_fit, tm=20, df_pred_unique, subj=1){
+  # values for prediction
+  max_t = unique(df_pred_unique$bin)[tm] # maximum observation time on bin label scale
+  tao <- diag(fpca_fit$evalues) ## egienvalues, also var-cov matrix of scores
+  Phi <- fpca_fit$efunctions[1:tm, ] ## eigenfunctions on the entire binned grid
+  
+  # observed prediction
+  partial_df <- df_pred_unique %>% filter(bin <= max_t & id == subj) 
+  
+  # re-estimate score
+  score_hat <- tao %*% t(Phi) %*% solve(Phi%*%tao%*%t(Phi)+fpca_fit$sigma2*diag(tm)) %*% (partial_df$eta_hat-fpca_fit$mu[1:tm])
+  
+  # predicte latent function
+  eta_pred <- fpca_fit$mu+fpca_fit$efunctions%*%score_hat
+  
+  return(eta_pred)
+}
+
+
+# in_samp_dyn_pred(fpca_fit, 60,df_pred_unique, 7)
+
+
+
+
+
+

### 2022-11-21 Code/GLMM-FPCA.R

@@ -9,8 +9,6 @@ mid <- (brks+bin_w/2)[1:n_bin] # mid points
 df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
 df$bin <- as.numeric(as.character(df$bin))
 
-head(df, 20)
-
 # check overlap
 # df %>% 
 #   filter(id %in% 1:4) %>% 
@@ -31,7 +29,7 @@ pred_latent <- function(df){
 ## do that for all time bin
 df_bin_lst <- split(df, f = df$bin)
 df_pred_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
-lapply(df_pred_latent, dim)
+# lapply(df_pred_latent, dim)
 df_pred_latent <- bind_rows(df_pred_latent)
 
 #### fPCA on predicted latent variables #####
@@ -48,59 +46,8 @@ mat_pred_unique <- matrix(df_pred_unique$eta_hat, nrow=N, ncol=n_bin, byrow = T)
 fpca_fit <- fpca.face(mat_pred_unique, pve = 0.95, argvals = unique(df_pred_unique$bin), knots=20, var=T)
 # fpca_fit <- face.sparse(df_pred_unique, argvals.new = 1:J, knots = 20, pve=0.95)
 
-# Things for prediction
-tao <- diag(fpca_fit$evalues) ## egienvalues, also var-cov matrix of scores
-allB <- fpca_fit$efunctions ## eigen functions on the entire binned grid
-fpca_fit$mu ## mean function
-fpca_fit$sigma2 ## residual variance
 
 
-
-##### In-sample prediction on binned grid #####
-
-# start with one obesrvation
-df_id1 <- df_pred_latent %>% filter(id==1 & sind_inx <= 300) %>% 
-  select(id, bin, eta_hat) %>% distinct(.)
-
-obs_bin <- length(unique(df_id1$bin))
-
-B <- allB[1:obs_bin, ] # eigenfunctions up to observed time point
-
-# score
-score_hat <- tao %*% t(B) %*% solve(B%*%tao%*%t(B)+fpca_fit$sigma2*diag(obs_bin)) %*% (df_id1$eta_hat-fpca_fit$mu[1:obs_bin])
-
-fpca_fit$scores[1, ] ## scores from fPCA face
-
-# prediction
-df_pred_id1 <- df_pred_latent %>% filter(id==1 )
-
[…]

### 2022-11-21 Code/ToySimulation.R

@@ -1,13 +1,7 @@
 
-set.seed(1025)
 ##### generate data #####
 
-rm(list=ls())
-library(tidyverse)
-library(ggplot2)
-library(fcr)
-library(refund)
-library(lme4)
+# rm(list=ls())
 
 
 #####  generate data f ####

### 2022-11-21 f-GFPCA.Rmd
clean up figures

@@
 library(lme4)
+library(ggpubr)
 set.seed(1025)
@@
 ```{r}
-# Try on subjectw 1-4
-
+# Try on subjects 1-4
+ggarrange(
 clean_pred(1) %>%
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
+  geom_line(aes(x=bin, y = eta_i, col=type))+
+  labs(title = "Subject 1"),
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
+  geom_line(aes(x=bin, y = eta_i, col=type))+
+  labs(title = "Subject 2"),
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
+  geom_line(aes(x=bin, y = eta_i, col=type))+
+  labs(title = "Subject 3"),
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
-
-```
-- Subject 2
-
-```{r}
-# Try on subject 1
-pred20 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=2),
-                     type = "up to 195") %>% 
-  filter(bin>195)
-
-pred40 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=2),
-                     type = "up to 395") %>%
-  filter(bin>395)
-
-pred60 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=2),
-                     type = "up to 595") %>%
-  filter(bin>595)
-
-clean_pred(1)
-  ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
-```
-
-- Subject 3
-
-```{r}
-# Try on subject 1
-pred20 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=3),
-                     type = "up to 195") %>% 
-  filter(bin>195)
-
-pred40 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=3),
-                     type = "up to 395") %>%
-  filter(bin>395)
-
-pred60 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=3),
-                     type = "up to 595") %>%
-  filter(bin>595)
-
-df_pred_latent %>% filter(id==3 & sind_inx==bin) %>%
-  select(bin, eta_i) %>% mutate(type="true") %>%
-  add_row(pred20) %>%
-  add_row(pred40) %>% 
-  add_row(pred60) %>% 
-  ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
-```
-- Subject 4
-
-```{r}
-# Try on subject 1
-pred20 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj=4),
-                     type = "up to 195") %>% 
-  filter(bin>195)
-
-pred40 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj=4),
-                     type = "up to 395") %>%
-  filter(bin>395)
-
-pred60 <- data.frame(bin = mid, 
-                     eta_i = in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj=4),
-                     type = "up to 595") %>%
-  filter(bin>595)
-
-df_pred_latent %>% filter(id==4 & sind_inx==bin) %>%
-  select(bin, eta_i) %>% mutate(type="true") %>%
-  add_row(pred20) %>%
-  add_row(pred40) %>% 
-  add_row(pred60) %>% 
-  ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))
+  geom_line(aes(x=bin, y = eta_i, col=type))+
+  labs(title = "Subject 4"), nrow=2, ncol=2, common.legend = T)
 

### 2022-12-01 f-GFPCA.Rmd
edit report

@@
   ggplot()+
-  geom_line(aes(x=sind, y=eta_i), col = "blue")+
-  geom_line(aes(x=sind, y=probs), col = "red")+
-  geom_point(aes(x=sind, y=Y), size = 0.05)+
+  geom_line(aes(x=sind, y=eta_i, color = "True Gaussian function"))+
+  geom_line(aes(x=sind, y=probs, color = "Probability scale"))+
+  geom_point(aes(x=sind, y=Y, color = "Binary outcome"), size = 0.05)+
   facet_wrap(~id)+
-  labs(title = "Four simulated subjects")
+  labs(title = "Four simulated subjects", color = "Legend")+
+  scale_color_manual(values = c("True Gaussian function" = "blue", "Probability scale" = "red", "Binary outcome" = "black"))
 ```
@@
 ggplot(df_pred_latent %>% filter(id %in% 1:4))+
-  geom_line(aes(x = sind_inx, y=eta_hat))+
-  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
-  # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
+  geom_line(aes(x = sind_inx, y=eta_hat, color = "Estimated"))+
+  geom_line(aes(x=sind_inx, y = eta_i, color = "True"))+
   facet_wrap(~id)+
-  labs(title = "Predicted latent functions on original grid")
+  labs(title = "Predicted latent functions on original grid", color = "Lgend")+
+  scale_color_manual(values = c("Estimated" = "red", "True" = "Black"))
 ```
@@
   ggplot()+
-  geom_line(aes(x = sind_inx, y=eta_hat))+
-  geom_line(aes(x=sind_inx, y = eta_i, col = "red"))+
+  geom_line(aes(x = sind_inx, y=eta_hat, col = "Estimated"))+
+  geom_line(aes(x=sind_inx, y = eta_i, col = "True"))+
   # geom_point(aes(x=sind_inx, y = Y, col = "blue"), size = 0.01)+
   facet_wrap(~id)+
-  labs(title = "Predicted latent functions on binned grid")
-
+  labs(title = "Predicted latent functions on binned grid", color = "Lgend")+
+  scale_color_manual(values = c("Estimated" = "red", "True" = "Black"))
 ```
@@
 
-```{r}
+```{r, fig.height=4, fig.width=4}
 ## mean functions
@@
 
-Below we try to do that with one subject:
-
-- Subject 1
+Below we try to do that with a few subjects:
 

### 2022-12-01 f-GFPCA.Rmd
more edits to progress report

@@
 
-### Local GLMMS
+### Local GLMMs
 
@@
 
-Now let's try to predict future outcomes based on observations $\boldsymbol{Y}$ up to $t_m$.
+Now let's try to predict future outcomes based on observations $\boldsymbol{Y}_m$ up to $t_m$.
 
@@
 
-$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}) =  \boldsymbol{\hat{\Gamma}\Phi ^T}(\boldsymbol{\Phi\hat{\Gamma}\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}(\hat{\boldsymbol{\eta}}-\hat{\boldsymbol{f}}_0)$$
+$$\hat{\boldsymbol{\xi}} = E(\boldsymbol{\xi}|\boldsymbol{Y}_m) =  \boldsymbol{\hat{\Gamma}\Phi ^T_m}(\boldsymbol{\Phi_m\hat{\Gamma}\Phi^T_m}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}(\hat{\boldsymbol{\eta}}_m-\hat{\boldsymbol{f}}_0^m)$$
 Where:
@@
 - $\hat{\boldsymbol{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
-- $\boldsymbol{\Phi}$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
+- $\boldsymbol{\Phi}_m$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
 - $\hat{\sigma_{\epsilon}}^2$ is the residual variance from FPCA
-- $\hat{\boldsymbol{\eta}}$ is the estimated latent Gaussian function up to $t_m$
-- $\hat{\boldsymbol{f}}_0$ is the FPCA mean function up to $t_m$
+- $\hat{\boldsymbol{\eta}}_m$ is the estimated latent Gaussian function up to $t_m$
+- $\hat{\boldsymbol{f}}_0^m$ is the FPCA mean function up to $t_m$
+
+And then, we use the FPCA model for point prediction on the entire binned grid:
+
+$$\hat{\boldsymbol{\eta}} = \hat{\boldsymbol{f}}_0+\boldsymbol{\Phi}\hat{\boldsymbol{\xi}}$$
+
+Here, $\boldsymbol{\Phi}$ is the eigenfunctions on the entire binned grid. For $\hat{\boldsymbol{\eta}}$, even though it is calculated for the entire binned grid, we will take values after $t_m$.
 

### 2022-12-22 f-GFPCA.Rmd
start on out-of-sample pred

@@
 title: "fGFPCA dynamic prediction (In-sample)"
-output: html_document
+output: 
+  html_document:
+    toc: true
+    toc_float: true
+    toc_depth: 3
 date: "2022-11-03"
@@
 
-## Toy simulation
+# Toy simulation
 
-### Simulation set-up
+## Simulation set-up
 
@@
 
-### Local GLMMs
+## Local GLMMs
 
@@
 
-### FPCA
+## FPCA
 
@@
 
-### In-sample prediction
+## In-sample prediction
 
@@
 
+Let
+
+$$\boldsymbol{H} = \boldsymbol{\hat{\Gamma}\Phi ^T_m}(\boldsymbol{\Phi_m\hat{\Gamma}\Phi^T_m}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}$$
+
+The variance of $\hat{\boldsymbol{\eta}}$ can be estimated by:
+
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=\boldsymbol{\Phi}(\hat{\boldsymbol{\Gamma}}-\boldsymbol{H}\boldsymbol{\Phi_m}\hat{\boldsymbol{\Gamma}})\boldsymbol{\Phi}^T+\hat{\sigma_{\epsilon}}^2$$
+
+It should be a $S \times S$ matrix; diagonals are variance at a specific time and off-diagonals are covariance between different times on binned grid.
+
 Below we try to do that with a few subjects:
@@
 source(here("code/DynPred.R"))
+
+#in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, 1)
+
 ```
@@
 clean_pred <- function(subj = 1){
-  pred20 <- data.frame(bin = mid, 
-                       eta_i = in_samp_dyn_pred(fpca_fit, 20,
-                                                df_pred_unique, subj),
-                       type = "up to 195") %>% 
-    filter(bin>195)
 
-  pred40 <- data.frame(bin = mid, 
-                      eta_i = in_samp_dyn_pred(fpca_fit, 40,
-                                               df_pred_unique, subj),
-                       type = "up to 395") %>%
-    filter(bin>395)
+  pred20 <- in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj)
+  df_pred20 <- data.frame(bin = mid, 
+                          eta_i = pred20$eta_pred,
+                          eta_i_std=pred20$eta_std_var,
+                          type = "up to 195") 
+  df_pred20$eta_i[df_pred20$bin<=195] <- NA
+  df_pred20$eta_i_std[df_pred20$bin<=195] <- NA
 
-  pred60 <- data.frame(bin = mid, 
-                       eta_i = in_samp_dyn_pred(fpca_fit, 60,
-                                                df_pred_unique, subj),
-                        type = "up to 595") %>%
-    filter(bin>595)
+  pred40 <- in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj)
+  df_pred40 <- data.frame(bin = mid, 
+                          eta_i = pred40$eta_pred,
+                          eta_i_std=pred40$eta_std_var,
+                          type = "up to 395") 
+  df_pred40$eta_i[df_pred20$bin<=395] <- NA
+  df_pred40$eta_i_std[df_pred20$bin<=395] <- NA
 
-  df_pred <- df_pred_latent %>% filter(id==subj & sind_inx==bin) %>%
-    select(bin, eta_i) %>% mutate(type="true") %>%
-    add_row(pred20) %>%
-    add_row(pred40) %>% 
-    add_row(pred60)
+  pred60 <- in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj)
+  df_pred60 <- data.frame(bin = mid, 
+                          eta_i = pred60$eta_pred,
+                          eta_i_std=pred60$eta_std_var,
+                          type = "up to 595")
+  df_pred60$eta_i[df_pred20$bin<=595] <- NA
+  df_pred60$eta_i_std[df_pred20$bin<=595] <- NA
+
+  df_pred <- df_pred20 %>%
+    add_row(df_pred40) %>% 
+    add_row(df_pred60) %>%
+    mutate(eta_i_lower = eta_i-eta_i_std, 
+           eta_i_upper = eta_i+eta_i_std) %>%
+    left_join(df_pred_latent %>% 
+                 filter(id==subj & sind_inx==bin) %>%
+                 select(bin, eta_i) %>% 
+                 rename(true_eta = eta_i), by = "bin")
 
@@
 
-```{r}
+```{r, fig.width=12, fig.height=4}
 # Try on subjects 1-4
-ggarrange(
 clean_pred(1) %>%
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))+
-  labs(title = "Subject 1"),
+  geom_line(aes(x=bin, y = true_eta), col = "red")+
+  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
+  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 1", x = "bin", y="")+
+  facet_wrap(~type)
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))+
-  labs(title = "Subject 2"),
+  geom_line(aes(x=bin, y = true_eta), col = "red")+
+  geom_line(aes(x=bin, y = eta_i), na.rm=T)+
+  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 2", x = "bin", y="")+
+  facet_wrap(~type)
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))+
-  labs(title = "Subject 3"),
+  geom_line(aes(x=bin, y=true_eta), col = "red")+
+  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
+  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 3", x = "bin", y="")+
+  facet_wrap(~type)
 
@@
   ggplot()+
-  geom_line(aes(x=bin, y = eta_i, col=type))+
-  labs(title = "Subject 4"), nrow=2, ncol=2, common.legend = T)
+  geom_line(aes(x=bin, y=true_eta), col = "red")+
+  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
+  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 4", x = "bin", y="")+
+  facet_wrap(~type)
+```
 
-```
+## Out of sample prediction
+
+Now let's say if we have a new subject with observations up to $t_m$ and $t_m$ is in mth bin:
+
+1. Local GLMMs: we will be able to estimate random effect of this subject up to mth bin, thus estimate the latent function up to the mth bin ($\hat{\boldsymbol{\eta}}_m$).
+
+2. fPCA: We will have an incomplete function for this new observation. I think fpca.face can handle that.
+
+3. Use the save formulas as in-sample prediction
 

### 2022-12-22 Code/DynPred.R

@@ -18,19 +18,24 @@ in_samp_dyn_pred <- function(fpca_fit, tm=20, df_pred_unique, subj=1){
   partial_df <- df_pred_unique %>% filter(bin <= max_t & id == subj) 
   
   # re-estimate score
-  score_hat <- tao %*% t(Phi) %*% solve(Phi%*%tao%*%t(Phi)+fpca_fit$sigma2*diag(tm)) %*% (partial_df$eta_hat-fpca_fit$mu[1:tm])
+  H <-  tao %*% t(Phi) %*% solve(Phi%*%tao%*%t(Phi)+fpca_fit$sigma2*diag(tm))
+  score_hat <- H %*% (partial_df$eta_hat-fpca_fit$mu[1:tm])
   
   # predicte latent function
   eta_pred <- fpca_fit$mu+fpca_fit$efunctions%*%score_hat
   
-  return(eta_pred)
+  # predict std
+  eta_pred_var <- fpca_fit$efunctions %*% (tao - H %*% Phi %*% tao) %*% t(fpca_fit$efunctions) + fpca_fit$sigma2
+  
+  return(list(eta_pred = eta_pred, eta_pred_var = eta_pred_var,
+              eta_std_var = sqrt(diag(eta_pred_var))))
 }
 
 
-# in_samp_dyn_pred(fpca_fit, 60,df_pred_unique, 7)
-
-
-
-
-
-
+# try <- in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, 4)
+# 
+# try$eta_std_var
+# 
+# plot(1:length(try$eta_std_var), try$eta_std_var)
+# fpca_fit$diag.var
+# dim(fpca_fit$VarMats[[2]])

### 2022-12-26 f-GFPCA.Rmd
next steps

@@
 
-1. Interval prediction
-2. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
-3. Out-of-sample prediction: how to estimate scores?
+1. Write out the formula for the BLUPs in the local GLMMs. Since we have a random intercept only model this should not be too complicated. (this is for out of sample prediction only)
+2. Approximate these values numerically via numeric integration (like we did witht he AUC paper) and/or explicit laplace approximation 
+3. Evaluate performance of dynamic predictions (both using "in sample" data -- those individuals whose BLUPs we can extract directly from the local glmm fits and out of sample data)
+4. Write an R package (see Julia's fastGFPCA and Andrew's fcr pacakge as inspiration) which has functions for performing dynamic prediction using our framework
+5. Implement a brief simulation study showing both that the approach works
 
-Thoughts on 3:
+6. Deriving unbiased scores
+
+As in our first paper (attached), the scores we obtain from the approach you're using are biased.
+We can show that the scores obtained from the local models (i.e.
+Step 3 in the fastGFPCA approach) are a linear function of the true scores.
+1) Derive the result that the binned scores are linear in the true scores (I have already derived this result, but you should attempt to do this yourself.)
+- Attempt to find an analytic solution for the bias factor. This may be challenging, so don't spend too much time here, but it would be an amazing result if we could do so
+
+2) Assuming you cannot find an analytic solution for the bias factor, attempt to estimate this factor yourself.
+To do so implement the full fastGFPCA approach (i.e.
+Step 4, where we re-estimate the scores), then:
+- Take the estimated scores for each subject from step 4 and and regress them on the scores from step 3 using linear regression. Do this separately for each eigenfunction
+-  Identify and report the estimated coefficients and R^2 for each eigenfunction separately
+- Use these models to create a de-biasing model for the Step 3 scores
+Show these scores are closer (lower MSE) to the true scores than those obtained from Step 3, and that these result in lower MSE for the latent subject specific functoions
+- Incorporate this result into the dynamic prediction framework you developed per the instructions above
+
+7. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
+
+Thoughts on 1:
 

### 2023-01-02 f-GFPCA.Rmd
add GLMM intercept function

@@
     toc: true
-    toc_float: true
+    toc_float: 
+      collapsed: false
     toc_depth: 3
@@
 source(here("Code/GLMM-FPCA.R"))
+```
+
+Figure below is the "mean" latent function across subject, which essential is $\beta_0^s$ from the series of mixed models.
+
+```{r,fig.height=4, fig.width=4}
+ggplot(data.frame(x=mid, y=mean_latent))+
+  geom_line(aes(x=x, y=y))
 ```
@@
 
-Now let's say if we have a new subject with observations up to $t_m$ and $t_m$ is in mth bin:
+Now let's say if we have a new subject $\boldsymbol{Y}_{new}$ with observations up to $t_m$ and $t_m$ is in mth bin.
+Also, we do not wish to refit any model, including the local GLMMs.
+In the first step, we need to use the series of local GLMMs that's already been fitted to get the random intercepts for this new observations ($\boldsymbol{b}_{new}$).
+
+We first focus on one specific interval s.
 
@@
 3. Use the save formulas as in-sample prediction
+
+```{r}
+```
 

### 2023-01-02 Code/GLMM-FPCA.R

@@ -18,6 +18,7 @@ df$bin <- as.numeric(as.character(df$bin))
 #   facet_wrap(~id)
 
 # fit local linear mixed models in each bin
+# and extract subject level estimation
 ## df: function with binary outcome Y, with observations in one time bein
 pred_latent <- function(df){
   this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
@@ -26,11 +27,24 @@ pred_latent <- function(df){
   return(df)
 }
 
+# additionally, get the average estimation across subject 
+# by extacting the random intercepts from each bin
+mean_latent <- function(df){
+  this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+  beta_hat <- coef(summary(this_glm))[1]
+  return(beta_hat)
+}
+
 ## do that for all time bin
 df_bin_lst <- split(df, f = df$bin)
+
+###### subject specific
 df_pred_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
 # lapply(df_pred_latent, dim)
 df_pred_latent <- bind_rows(df_pred_latent)
+
+###### subject average
+mean_latent <- sapply(df_bin_lst, function(x){mean_latent(x)}) 
 
 #### fPCA on predicted latent variables #####
 rm(df_bin_lst)

### 2023-01-04 f-GFPCA.Rmd
try MLE of out-of-sample latent function

@@
 
+## Some clarification points
+
+1. When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval.
+
 ## Out of sample prediction
 
-Now let's say if we have a new subject $\boldsymbol{Y}_{new}$ with observations up to $t_m$ and $t_m$ is in mth bin.
-Also, we do not wish to refit any model, including the local GLMMs.
-In the first step, we need to use the series of local GLMMs that's already been fitted to get the random intercepts for this new observations ($\boldsymbol{b}_{new}$).
+Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:
 
-We first focus on one specific interval s.
+### Refit local GLMMs
 
@@
 
+### Do not refit GLMMs
+
+We can utilize the fitted series of local GLMMs, instead of refitting them.
+Since $\boldsymbol{Y}_{new}$ is a binary function, we can get its (incomplete) latent function value $\eta^s$ with Maximum Likelihood Estimation.
+Specifically, we need to estimate its (incomplete) random intercept function $b^s$.
+
+Let's focus on a specific bin $s$. If the new sample has $n_s$ observations in this bin, among which $m_s$ are success (value = 1). Then $\boldsymbol{Y}^s$ would follow a Binomial distribution:
+
+$$\boldsymbol{Y}^s \sim Binomial(n_s, \frac{exp(\eta^s)}{1+exp(\eta^s)})$$
+Where $\eta^s = \beta_0^s+b^s$, and $\beta_0^s$ is already estimated with training data.
+
+Now we write out the likelihood in interval s:
+
+$$L(b^s|\boldsymbol{Y}^s) = \binom{n^s}{m^s}p_s^{m^s}(1-p_s)^{n^s-m^s}$$
+where
+
+$$p_s = \frac{exp(\eta^s)}{1+exp(\eta^s)}, \hspace{0.5cm} 
+\eta^s = \beta_0^s+b^s$$
+
+Then the log likelihood is:
+
+$$\begin{aligned}
+l(b^s|\boldsymbol{Y}^s) &= log\binom{n^s}{m^s}+{m^s}logp_s+(n^s-m^s)log(1-p_s)\\
+&= log\binom{n^s}{m^s}+m^slog(\frac{p_s}{1-p_s})+n^slog(1-p_s)\\
+&= log\binom{n^s}{m^s}+m^s(\beta_0+b^s)-n^slog(1+exp(\beta_0^s+b^s))
+\end{aligned}$$
+
+Then we take the first derivative and set it to zero, and end up with MLE os $b^s$:
+
+$$\hat{b}^s=log(\frac{m^s}{n^s-m^s})-\beta_0^s$$
+Again, $\beta_0^s$ is the estimated mean latent function from the local GlMMs steps. In fact, this individual latent track can be simply estimated by success ratio in the corresponding interval:
+
+$$\eta^s = log\frac{m^s}{n^s-m^s}$$
+
 ```{r}
+# generate one subject from the smae generating mechanism
+xi_new <- matrix(rnorm(1*K),1,K);
+xi_new <- xi_new %*% diag(sqrt(lambda))
+# random intercept function
+b_new <- xi_new %*% t(phi)
+# latent function
+eta_new <- f_0(sind) + b_new
+## response binary function
+Y_new <- rbinom(length(eta_new), size=1, prob=plogis(eta_new))
+df_new_true <- data.frame(
+  id = N+1000,
+  sind = sind, 
+  Y = Y_new,
+  eta = t(eta_new),
+  sind_inx = 1:J)
+
+# cut observations after specific time point
+tm <- 0.5
+df_new_obs <- df_new_true %>%  select(-eta) %>% 
+  filter(sind<=0.5)
+```
+
+```{r}
+# bin observations
+df_new_obs$bin <- cut(df_new_obs$sind_inx, breaks = brks, 
+                      include.lowest = T,labels = mid)
+df_new_obs$bin <- as.numeric(as.character(df_new_obs$bin))
+
+# estimated MLE of latent function
+eta_hat_new <- df_new_obs %>% group_by(bin) %>% 
+  summarise_at("Y", function(x){log(sum(x==1)/sum(x==0))})
+
+# plot
+df_new_true %>% filter(sind_inx %in% mid) %>% 
+  select(sind_inx, eta) %>% 
+  rename(bin = sind_inx) %>% 
+  left_join(eta_hat_new) %>% 
+  ggplot()+
+  geom_line(aes(x=bin, y=eta, col = "True"))+
+  geom_line(aes(x=bin, y=Y, col = "Predicted"))
+
 ```

### 2023-01-06 f-GFPCA.Rmd
out-of-sample prediction using Empirical Bayes

--- 2023-01-02 f-GFPCA.Rmd
@@
 ```
+
+# Rationale
+
+We assume that a binary function $Y(t)$ is generated by a latent Gaussian function $\eta(t)$, as follows:
+
+$$Y_i(t) \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))})$$
+$$\eta_i(t) = \beta_0(t)+b_i(t)$$
+
+And this latent funciton can be approximated by fPCA:
+
+$$\eta_i(t) = f_0(t)+\sum_{k=1}^{K}\xi_k\phi_k(t)+\epsilon_i(t)$$
 
@@
 
+## Some clarification points
+
+1. When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval.
+
 ## Out of sample prediction
 
-Now let's say if we have a new subject $\boldsymbol{Y}_{new}$ with observations up to $t_m$ and $t_m$ is in mth bin.
-Also, we do not wish to refit any model, including the local GLMMs.
-In the first step, we need to use the series of local GLMMs that's already been fitted to get the random intercepts for this new observations ($\boldsymbol{b}_{new}$).
+Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:
 
-We first focus on one specific interval s.
+### Refit local GLMMs to estimate latent function
 
@@
 
-```{r}
-```
+### Do not estimate latent function
+
+Let say on the binned grid, at each interval s, new sample has $n_s$ observations, among which $m_s$ are successes.
+$m_s = \sum_{j=1}^{n_S} I(Y_{sj}=1)$, j is the index for observation in a specific interval.
+
+We use the expectation of latent function conditonal on fPCA score to estimate the true latent function, so that the error term can be removed from following calculation:
+
+$$\hat{\eta_s} = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}$$
+
+Here, $\hat{f}_{0}(s)$ is the mean function from fPCA and $\phi(s)$ are eigenfunctions at time s, which is a K $\times$ 1 vector.
+
+Then the data is generated from a binomial distribution:
+
+$$\begin{aligned}
+ \boldsymbol{Y}_s & \sim Binomial(p_s) \\
+ logit(p_s) & = \hat{\eta}_s = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}
+\end{aligned}$$
+
+Now let's find the empirical bayes estimator of score $E(\boldsymbol{\xi}|\boldsymbol{Y})$, which is not likely to have closed form solution.
+
+$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int_{\boldsymbol{\xi}} \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
+$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+
+Thus:
+
+$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int_{\boldsymbol{\xi}}\boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+
+We know that fPCA scores follow a multivariate normal distribution:
+
+$$p(\boldsymbol{\xi}) = (2\pi)^{-K/2}det(\boldsymbol{\Gamma})^{-1/2}exp(-\boldsymbol{\xi^T \Gamma^{-1} \xi}/2)$$
+
+And the binary outcome, conditional on score, follows a Binomial distribution:
+
+$$\begin{aligned}
+p(\boldsymbol{Y}|\boldsymbol{\xi}) &= \prod_{s=1}^{m}p(\boldsymbol{Y}_t|\boldsymbol{\xi})
+\end{aligned}$$
+<!-- We can utilize the fitted series of local GLMMs, instead of refitting them.
+Since $\boldsymbol{Y}_{new}$ is a binary function, we can get its (incomplete) latent function value $\eta^s$ with Maximum Likelihood Estimation.
+Specifically, we need to estimate its (incomplete) random intercept function $b^s$.
+-->
+
+<!-- Let's focus on a specific bin $s$.
+If the new sample has $n_s$ observations in this bin, among which $m_s$ are success (value = 1).
+Then $\boldsymbol{Y}^s$ would follow a Binomial distribution:  -->
+
+<!-- $$\boldsymbol{Y}^s \sim Binomial(n_s, \frac{exp(\eta^s)}{1+exp(\eta^s)})$$ -->
+<!-- Where $\eta^s = \beta_0^s+b^s$, and $\beta_0^s$ is already estimated with training data.  -->
+
+<!-- Now we write out the likelihood in interval s: -->
+
+<!-- $$L(b^s|\boldsymbol{Y}^s) = \binom{n^s}{m^s}p_s^{m^s}(1-p_s)^{n^s-m^s}$$ -->
+<!-- where  -->
+
+<!-- $$p_s = \frac{exp(\eta^s)}{1+exp(\eta^s)}, \hspace{0.5cm}  -->
+<!-- \eta^s = \beta_0^s+b^s$$ -->
+
+<!-- Then the log likelihood is: -->
+
+<!-- $$\begin{aligned} -->
+<!-- l(b^s|\boldsymbol{Y}^s) &= log\binom{n^s}{m^s}+{m^s}logp_s+(n^s-m^s)log(1-p_s)\\ -->
+<!-- &= log\binom{n^s}{m^s}+m^slog(\frac{p_s}{1-p_s})+n^slog(1-p_s)\\ -->
+<!-- &= log\binom{n^s}{m^s}+m^s(\beta_0+b^s)-n^slog(1+exp(\beta_0^s+b^s)) -->
+<!-- \end{aligned}$$ -->
+
+<!-- Then we take the first derivative and set it to zero, and end up with MLE os $b^s$: -->
+
+<!-- $$\hat{b}^s=log(\frac{m^s}{n^s-m^s})-\beta_0^s$$ -->
+<!-- Again, $\beta_0^s$ is the estimated mean latent function from the local GlMMs steps.
+In fact, this individual latent track can be simply estimated by success ratio in the corresponding interval: -->
+
+<!-- $$\eta^s = log\frac{m^s}{n^s-m^s}$$ -->
+
+<!-- ```{r} -->
+<!-- # generate one subject from the smae generating mechanism -->
+<!-- xi_new <- matrix(rnorm(1*K),1,K); -->
+<!-- xi_new <- xi_new %*% diag(sqrt(lambda)) -->
+<!-- # random intercept function -->
+<!-- b_new <- xi_new %*% t(phi) -->
+<!-- # latent function -->
+<!-- eta_new <- f_0(sind) + b_new -->
+<!-- ## response binary function -->
+<!-- Y_new <- rbinom(length(eta_new), size=1, prob=plogis(eta_new)) -->
+<!-- df_new_true <- data.frame( -->
+<!--   id = N+1000, -->
+<!--   sind = sind,  -->
+<!--   Y = Y_new, -->
+<!--   eta = t(eta_new), -->
+<!--   sind_inx = 1:J) -->
+
+<!-- # cut observations after specific time point -->
+<!-- tm <- 0.5 -->
+<!-- df_new_obs <- df_new_true %>%  select(-eta) %>%  -->
+<!--   filter(sind<=0.5) -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- # bin observations -->
+<!-- df_new_obs$bin <- cut(df_new_obs$sind_inx, breaks = brks,  -->
+<!--                       include.lowest = T,labels = mid) -->
+<!-- df_new_obs$bin <- as.numeric(as.character(df_new_obs$bin)) -->
+
+<!-- # estimated MLE of latent function -->
+<!-- eta_hat_new <- df_new_obs %>% group_by(bin) %>%  -->
+<!--   summarise_at("Y", function(x){log(sum(x==1)/sum(x==0))})  -->
+
+<!-- # plot -->
+<!-- df_new_true %>% filter(sind_inx %in% mid) %>%  -->
+<!--   select(sind_inx, eta) %>%  -->
+<!--   rename(bin = sind_inx) %>%  -->
+<!--   left_join(eta_hat_new) %>%  -->
+<!--   ggplot()+ -->
+<!--   geom_line(aes(x=bin, y=eta, col = "True"))+ -->
+<!--   geom_line(aes(x=bin, y=Y, col = "Predicted")) -->
+
+<!-- ``` -->
 

### 2023-01-09 f-GFPCA.Rmd
try out-of-sample mutlidimensional integral

@@
 
-Let say on the binned grid, at each interval s, new sample has $n_s$ observations, among which $m_s$ are successes.
-$m_s = \sum_{j=1}^{n_S} I(Y_{sj}=1)$, j is the index for observation in a specific interval.
+Let say on the binned grid, at each interval s, new sample has $n_s$ observations, among which $h_s$ are successes.
+$h_s = \sum_{j=1}^{n_S} I(Y_{sj}=1)$, j is the index for observation in a specific interval.
 
@@
 $$\begin{aligned}
-p(\boldsymbol{Y}|\boldsymbol{\xi}) &= \prod_{s=1}^{m}p(\boldsymbol{Y}_t|\boldsymbol{\xi})
+p(\boldsymbol{Y}|\boldsymbol{\xi}) &= \prod_{s=1}^{m}p(\boldsymbol{Y}_s|\boldsymbol{\xi}) \\
+& = \prod_{s=1}^{m} \binom{h_s}{n_s}p_s^{h_s}(1-p_s)^{n_s-h_s}\\
+logit(p_s) & = \hat{\eta}_s = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}
 \end{aligned}$$
+
+Please note that here, we are assuming $\boldsymbol{Y}_s$ is independent from each other given fPCA score.
+That is, all correlation across time within the same subject is introduced by underlying eigenfunctions.
+
+If we write this down for a single score:
+
+$$E(\xi_k|\boldsymbol{Y})=\frac{\int \xi_k \int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi_{(-k)}}d\xi_k}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+
+# Problems: 
+Numeric integration too slow (adaptIntegrate multidimensional)! Maybe try  MLE?
+
 <!-- We can utilize the fitted series of local GLMMs, instead of refitting them.
@@
 1. Write out the formula for the BLUPs in the local GLMMs. Since we have a random intercept only model this should not be too complicated. (this is for out of sample prediction only)
+
 2. Approximate these values numerically via numeric integration (like we did witht he AUC paper) and/or explicit laplace approximation 

### 2023-01-09 Code/out_of_samp_bayes.R

@@ -1,0 +1,49 @@
+
+# This script writes functions for 
+# calculating conditional expectation
+# of fPCA scores 
+# of an out-of-sample observation
+
+# first, only subject one, time up to 395
+rm(df_out)
+df_out1 <- df %>% filter(id == 1 & sind_inx <= 395) %>% select(-eta_i)
+max_bin <- which(mid==395)
+
+#### Denominator of E(xi|Y) ####
+# xi: scores. The value to integrate over
+# tao: estimated covariance matrix(eigenvalues) of score from fPCA
+# f0: mean function from fPCA (up to mth bin)
+# phi: basis function of fPCA (up to mth bin)
+# Y: newly observed data (up to mth bin)
+
+# just for mess around
+tao <- diag(fpca_fit$evalues)
+f0 <- fpca_fit$mu[1:max_bin]
+phi <- fpca_fit$efunctions[1:max_bin, ] 
+df_out <- df_out1
+xi <- fpca_fit$scores[1, ]
+  
+joint_pdf <- function(xi, tao, f0, phi, df_out){
+  
+  # numbers
+  ns <- table(df_out$bin) # number of observations
+  hs <- df_out %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+    select(Y) # number of success
+  nf <- ns-hs # number of failure
+  
+  # probability of conditional distribution of Y
+  eta = f0+phi%*%xi
+  pr = exp(eta)/(1+exp(eta))
+  # the part relevant to P(Y|xi)
+  prod((pr^hs)*(1-pr)^nf)*exp(-t(xi)%*%solve(tao)%*%xi/2)
+  # part relebant to marginal distribution of xi
+  
+  
+  
+  
+}
+
+condi_pdf_y <- function(Y)
+  
+  
+  

### 2023-01-11 f-GFPCA.Rmd
MLE scores and point prediction

@@
 ```{r}
-source(here("code/DynPred.R"))
-
-#in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, 1)
-
+source(here("code/InSampPred.R"))
 ```
@@
 # function to clean results
-# make predictions with partial track up to 20, 40, 60
-clean_pred <- function(subj = 1){
+# make predictions with partial track up to 195, 395, 595
 
-  pred20 <- in_samp_dyn_pred(fpca_fit, 20, df_pred_unique, subj)
-  df_pred20 <- data.frame(bin = mid, 
-                          eta_i = pred20$eta_pred,
-                          eta_i_std=pred20$eta_std_var,
-                          type = "up to 195") 
-  df_pred20$eta_i[df_pred20$bin<=195] <- NA
-  df_pred20$eta_i_std[df_pred20$bin<=195] <- NA
+in_pred_clean <- function(subjID){
+  # make predictions
+  pred_t195 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=195))
+  pred_t395 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=395))
+  pred_t595 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=595))
 
-  pred40 <- in_samp_dyn_pred(fpca_fit, 40, df_pred_unique, subj)
-  df_pred40 <- data.frame(bin = mid, 
-                          eta_i = pred40$eta_pred,
-                          eta_i_std=pred40$eta_std_var,
-                          type = "up to 395") 
-  df_pred40$eta_i[df_pred20$bin<=395] <- NA
-  df_pred40$eta_i_std[df_pred20$bin<=395] <- NA
+  df_pred_t195 <- data.frame(mid,
+                             eta=pred_t195$eta_pred,
+                             eta_std = pred_t195$eta_std_var) %>%
+    mutate(eta_lower = eta-eta_std, 
+             eta_upper = eta+eta_std) %>% 
+    filter(mid > 195)
 
-  pred60 <- in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, subj)
-  df_pred60 <- data.frame(bin = mid, 
-                          eta_i = pred60$eta_pred,
-                          eta_i_std=pred60$eta_std_var,
-                          type = "up to 595")
-  df_pred60$eta_i[df_pred20$bin<=595] <- NA
-  df_pred60$eta_i_std[df_pred20$bin<=595] <- NA
+  df_pred_t395 <- data.frame(mid,
+                             eta=pred_t395$eta_pred,
+                             eta_std = pred_t395$eta_std_var) %>%
+    mutate(eta_lower = eta-eta_std, 
+             eta_upper = eta+eta_std) %>% 
+    filter(mid > 395)
 
-  df_pred <- df_pred20 %>%
-    add_row(df_pred40) %>% 
-    add_row(df_pred60) %>%
-    mutate(eta_i_lower = eta_i-eta_i_std, 
-           eta_i_upper = eta_i+eta_i_std) %>%
-    left_join(df_pred_latent %>% 
-                 filter(id==subj & sind_inx==bin) %>%
-                 select(bin, eta_i) %>% 
-                 rename(true_eta = eta_i), by = "bin")
+  df_pred_t595 <- data.frame(mid,
+                             eta=pred_t595$eta_pred,
+                             eta_std = pred_t595$eta_std_var) %>%
+    mutate(eta_lower = eta-eta_std, 
+             eta_upper = eta+eta_std) %>% 
+  filter(mid > 595)
 
-  return(df_pred)
+  return(list(df_pred_t195, df_pred_t395, df_pred_t595))
 }
+
+# true latent function on the binned grid
+df_true_eta <- df_pred_latent %>% 
+  filter(sind_inx %in% mid) %>%
+  select(id, bin, eta_i) %>% rename("mid" = "bin", "true" = "eta_i")
 
@@
 # Try on subjects 1-4
-clean_pred(1) %>%
+bind_rows(in_pred_clean(1), .id = "type") %>%
+  mutate(type=factor(type, levels=1:3, 
+                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
+  left_join(df_true_eta %>% filter(id==1), by = "mid") %>% 
   ggplot()+
-  geom_line(aes(x=bin, y = true_eta), col = "red")+
-  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
-  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
-  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = true, col = "True"))+
+  geom_line(aes(x=mid, y = eta), na.rm = T)+
+  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
   labs(title = "Subject 1", x = "bin", y="")+
@@
 
-clean_pred(2) %>%
+bind_rows(in_pred_clean(2), .id = "type") %>%
+  mutate(type=factor(type, levels=1:3, 
+                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
+  left_join(df_true_eta %>% filter(id==2), by = "mid") %>% 
   ggplot()+
-  geom_line(aes(x=bin, y = true_eta), col = "red")+
-  geom_line(aes(x=bin, y = eta_i), na.rm=T)+
-  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
-  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = true, col = "True"))+
+  geom_line(aes(x=mid, y = eta), na.rm = T)+
+  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
   labs(title = "Subject 2", x = "bin", y="")+
-  facet_wrap(~type)
-
-clean_pred(3) %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=true_eta), col = "red")+
-  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
-  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
-  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
-  labs(title = "Subject 3", x = "bin", y="")+
-  facet_wrap(~type)
-
-clean_pred(4) %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=true_eta), col = "red")+
-  geom_line(aes(x=bin, y = eta_i), na.rm = T)+
-  geom_line(aes(x=bin, y=eta_i_lower), linetype = "dashed", na.rm=T)+
-  geom_line(aes(x=bin, y = eta_i_upper), linetype = "dashed", na.rm=T)+
-  labs(title = "Subject 4", x = "bin", y="")+
   facet_wrap(~type)
@@
 
+#### Empirical Bayes
+
 Now let's find the empirical bayes estimator of score $E(\boldsymbol{\xi}|\boldsymbol{Y})$, which is not likely to have closed form solution.
@@
 
-# Problems: 
-Numeric integration too slow (adaptIntegrate multidimensional)! Maybe try  MLE?
+In this case, I ran into the problem of numeric integration being too slow (adaptIntegrate multidimensional). The function value is also very small. Therefore, I turned to MLE instead.
 
-<!-- We can utilize the fitted series of local GLMMs, instead of refitting them.
-Since $\boldsymbol{Y}_{new}$ is a binary function, we can get its (incomplete) latent function value $\eta^s$ with Maximum Likelihood Estimation.
-Specifically, we need to estimate its (incomplete) random intercept function $b^s$.
--->
+#### MLE
 
-<!-- Let's focus on a specific bin $s$.
-If the new sample has $n_s$ observations in this bin, among which $m_s$ are success (value = 1).
-Then $\boldsymbol{Y}^s$ would follow a Binomial distribution:  -->
+We utilize the same conditional likelihood of scores:
 
-<!-- $$\boldsymbol{Y}^s \sim Binomial(n_s, \frac{exp(\eta^s)}{1+exp(\eta^s)})$$ -->
-<!-- Where $\eta^s = \beta_0^s+b^s$, and $\beta_0^s$ is already estimated with training data.  -->
+$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
 
-<!-- Now we write out the likelihood in interval s: -->
+But now we don't need to calculated the marginal distribution of $\boldsymbol{Y}$ anymore (denominator). So we'll get:
 
-<!-- $$L(b^s|\boldsymbol{Y}^s) = \binom{n^s}{m^s}p_s^{m^s}(1-p_s)^{n^s-m^s}$$ -->
-<!-- where  -->
+$$\begin{aligned}
+p(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi}) \\
 
-<!-- $$p_s = \frac{exp(\eta^s)}{1+exp(\eta^s)}, \hspace{0.5cm}  -->
-<!-- \eta^s = \beta_0^s+b^s$$ -->
+l(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto logp (\boldsymbol{Y}|\boldsymbol{\xi})+log p(\boldsymbol{\xi})
+\end{aligned}$$
 
-<!-- Then the log likelihood is: -->
+Plug in the conditional log-likelihood of $\boldsymbol{Y}$ and marginal log-likelihood of $\boldsymbol{\xi}$:
 
-<!-- $$\begin{aligned} -->
-<!-- l(b^s|\boldsymbol{Y}^s) &= log\binom{n^s}{m^s}+{m^s}logp_s+(n^s-m^s)log(1-p_s)\\ -->
-<!-- &= log\binom{n^s}{m^s}+m^slog(\frac{p_s}{1-p_s})+n^slog(1-p_s)\\ -->
-<!-- &= log\binom{n^s}{m^s}+m^s(\beta_0+b^s)-n^slog(1+exp(\beta_0^s+b^s)) -->
-<!-- \end{aligned}$$ -->
+$$\begin{aligned}
+l(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto \sum_{s=1}^mh_s\eta_s-\sum_{s=1}^mn_slog(1+exp(\eta_s))-\boldsymbol{\xi^T\Gamma^{-1}\xi}/2\\
+\frac{dl(\boldsymbol{\xi}|\boldsymbol{Y})}{d\boldsymbol{\xi}}& =\sum_{s=1}^mh_s\phi(s)-\sum_{s=1}^mn_s\frac{exp(\eta_s)}{1+exp(\eta_s)}\phi(s)-\boldsymbol{\xi^T\Gamma^{-1}} = 0
+\end{aligned}$$
 
-<!-- Then we take the first derivative and set it to zero, and end up with MLE os $b^s$: -->
+It is not easy to get a closed-form solution for the score equation above. We can try to get a numeric solution instead. Below are some visualization of the prediction performances
 
-<!-- $$\hat{b}^s=log(\frac{m^s}{n^s-m^s})-\beta_0^s$$ -->
-<!-- Again, $\beta_0^s$ is the estimated mean latent function from the local GlMMs steps.
-In fact, this individual latent track can be simply estimated by success ratio in the corresponding interval: -->
+#### In- and out-of-sample prediction of latent function
 
-<!-- $$\eta^s = log\frac{m^s}{n^s-m^s}$$ -->
+```{r}
+source(here("Code/OutSampMLE.R"))
+```
 
-<!-- ```{r} -->
-<!-- # generate one subject from the smae generating mechanism -->
-<!-- xi_new <- matrix(rnorm(1*K),1,K); -->
-<!-- xi_new <- xi_new %*% diag(sqrt(lambda)) -->
-<!-- # random intercept function -->
-<!-- b_new <- xi_new %*% t(phi) -->
-<!-- # latent function -->
-<!-- eta_new <- f_0(sind) + b_new -->
-<!-- ## response binary function -->
-<!-- Y_new <- rbinom(length(eta_new), size=1, prob=plogis(eta_new)) -->
-<!-- df_new_true <- data.frame( -->
-<!--   id = N+1000, -->
-<!--   sind = sind,  -->
-<!--   Y = Y_new, -->
-<!--   eta = t(eta_new), -->
-<!--   sind_inx = 1:J) -->
+```{r}
+# out-of-sample
+pred1_t195 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 195) %>%
+                                  select(-eta_i),
+                                fpca_fit = fpca_fit)
+pred1_t395 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 395) %>%
+                                  select(-eta_i),
+                                fpca_fit = fpca_fit)
 
-<!-- # cut observations after specific time point -->
-<!-- tm <- 0.5 -->
-<!-- df_new_obs <- df_new_true %>%  select(-eta) %>%  -->
-<!--   filter(sind<=0.5) -->
-<!-- ``` -->
+pred1_t595 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 595) %>%
+                                  select(-eta_i),
+                                fpca_fit = fpca_fit)
 
-<!-- ```{r} -->
-<!-- # bin observations -->
-<!-- df_new_obs$bin <- cut(df_new_obs$sind_inx, breaks = brks,  -->
-<!--                       include.lowest = T,labels = mid) -->
-<!-- df_new_obs$bin <- as.numeric(as.character(df_new_obs$bin)) -->
+df_out_pred1 <- data.frame(mid, pred1_t195,pred1_t395, pred1_t595)
+df_out_pred1$pred1_t195[1:which(mid==195)] <- NA
+df_out_pred1$pred1_t395[1:which(mid==395)] <- NA
+df_out_pred1$pred1_t595[1:which(mid==595)] <- NA
+df_out_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]
 
-<!-- # estimated MLE of latent function -->
-<!-- eta_hat_new <- df_new_obs %>% group_by(bin) %>%  -->
-<!--   summarise_at("Y", function(x){log(sum(x==1)/sum(x==0))})  -->
+```
 
-<!-- # plot -->
-<!-- df_new_true %>% filter(sind_inx %in% mid) %>%  -->
-<!--   select(sind_inx, eta) %>%  -->
-<!--   rename(bin = sind_inx) %>%  -->
-<!--   left_join(eta_hat_new) %>%  -->
-<!--   ggplot()+ -->
-<!--   geom_line(aes(x=bin, y=eta, col = "True"))+ -->
-<!--   geom_line(aes(x=bin, y=Y, col = "Predicted")) -->
+```{r}
+# in-sample  
+df_in_pred1 <- data.frame(mid,
+                          pred_t195 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=195))$eta_pred,
+                          pred_t395 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=395))$eta_pred,
+                          pred_t595 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=595))$eta_pred)
 
-<!-- ``` -->
+df_in_pred1$pred_t195[1:which(mid==195)] <- NA
+df_in_pred1$pred_t395[1:which(mid==395)] <- NA
+df_in_pred1$pred_t595[1:which(mid==595)] <- NA
+df_in_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]
+
+```
+
+```{r}
+# compare point prediction with in-sample
+
+## out-of-sample
+ggplot(df_out_pred1)+
+  geom_line(aes(x=mid, y=true, col="True"))+
+  geom_line(aes(x=mid, y=pred1_t195, col="Up to 195"), na.rm = T)+
+  geom_line(aes(x=mid, y=pred1_t395, col="Up to 395"), na.rm = T)+
+  geom_line(aes(x=mid, y=pred1_t595, col="Up to 595"), na.rm = T)+
+  labs(y = "Out-of-sample prediction", x = "Binned time")
+
+## in-sample
+ggplot(df_in_pred1)+
+  geom_line(aes(x=mid, y=true, col="True"))+
+  geom_line(aes(x=mid, y=pred_t195, col="Up to 195"), na.rm = T)+
+  geom_line(aes(x=mid, y=pred_t395, col="Up to 395"), na.rm = T)+
+  geom_line(aes(x=mid, y=pred_t595, col="Up to 595"), na.rm = T)+
+  labs(y = "In-sample prediction", x = "Binned time")
+
+```
+
+- Repeat this for more subjects
+- Interval prediction for MLE
+- Bias of score for both in-sample and out-of-sample estimation
 

### 2023-01-11 Code/InSampPred.R

@@ -1,34 +1,32 @@
 
 ##### In-sample prediction on binned grid #####
 # fpca_fit: fpca model fit on estimated latent Gaussian function
-# tm: maximum observation time on binned grid, order index of bin
-# df_pred_unique: data frame with only id, bin, eta_hat
-# subj: subject ID to make prediction on
-# mid: label of binned grid
+# df_part: data frame including partial observations, with at least columns "bin", "eta_hat"
+
 
 # return: predictions on the full binned grid
 
-in_samp_dyn_pred <- function(fpca_fit, tm=20, df_pred_unique, subj=1){
+in_samp_dyn_pred <- function(fpca_fit, df_part){
   # values for prediction
-  max_t = unique(df_pred_unique$bin)[tm] # maximum observation time on bin label scale
+  tm <- length(unique(df_part$bin)) # maximum observation time on bin label scale
   tao <- diag(fpca_fit$evalues) ## egienvalues, also var-cov matrix of scores
   Phi <- fpca_fit$efunctions[1:tm, ] ## eigenfunctions on the entire binned grid
   
-  # observed prediction
-  partial_df <- df_pred_unique %>% filter(bin <= max_t & id == subj) 
   
   # re-estimate score
   H <-  tao %*% t(Phi) %*% solve(Phi%*%tao%*%t(Phi)+fpca_fit$sigma2*diag(tm))
-  score_hat <- H %*% (partial_df$eta_hat-fpca_fit$mu[1:tm])
+  score_hat <- H %*% (df_part$eta_hat-fpca_fit$mu[1:tm])
   
-  # predicte latent function
+  # predict latent function
   eta_pred <- fpca_fit$mu+fpca_fit$efunctions%*%score_hat
   
   # predict std
   eta_pred_var <- fpca_fit$efunctions %*% (tao - H %*% Phi %*% tao) %*% t(fpca_fit$efunctions) + fpca_fit$sigma2
   
-  return(list(eta_pred = eta_pred, eta_pred_var = eta_pred_var,
-              eta_std_var = sqrt(diag(eta_pred_var))))
+  return(list(eta_pred = eta_pred, 
+              eta_pred_var = eta_pred_var,
+              eta_std_var = sqrt(diag(eta_pred_var)),
+              score_hat = score_hat))
 }
 
 

### 2023-01-11 Code/OutSampMLE.R

@@ -1,0 +1,57 @@
+
+library(rootSolve)
+
+# This script writes functions for 
+# calculating MLE solution of fPCA scores 
+# based on conditonal distribution of score (conditional on observations)
+# of an out-of-sample observation
+# just for mess around
+
+
+#### MLE of out-of-sample score ####
+
+# First derivative of Log-likelihood
+# df_new: new observations, including at least columns Y and bin
+# fpca_fit: fpca object
+
+
+llh_div <- function(xi, df_new, fpca_fit){
+  
+  # observations
+  ns <- as.vector(table(df_new$bin)) # number of observations
+  hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+    select(Y) %>% unlist()# number of success
+  nf <- ns-hs # number of failure
+  max_bin <- length(unique(df_new$bin)) # assume new skipped bins 
+  
+  # fpca objects
+  phi_m <- fpca_fit$efunctions[1:max_bin, ]
+  f0_m <- fpca_fit$mu[1:max_bin]
+  eta_s <- f0_m + phi_m %*% xi
+  tao <- diag(fpca_fit$evalues)
+  
+  # derivative
+  val1 <- hs-ns*(exp(eta_s)/(1+exp(eta_s)))
+  div <- colSums(apply(phi_m, 2, function(x)x*val1))-xi%*%solve(tao)
+  return(div)
+}
+
+# test
+# llh_div(rep(0, 4), df_new=df_new, fpca_fit = fpca_fit)
+
+#### Prediction with MLE scores #####
+
+out_samp_dyn_pred <- function(df_new, fpca_fit){
+  score_out <- multiroot(f=llh_div, start = rep(0, 4), df_new=df_new, fpca_fit=fpca_fit)$root
+  eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions%*%score_out
+  
+  return(eta_pred_out)
+  
+}
+
+
+# test
+# out_samp_dyn_pred(df_new, fpca_fit)
+
+
+

### 2023-01-13 f-GFPCA.Rmd
bias of score

@@
       collapsed: false
-    toc_depth: 3
+    toc_depth: 4
 date: "2022-11-03"
@@
 library(ggplot2)
+theme_set(theme_minimal())
 library(fcr)
@@
 library(ggpubr)
+library(gridExtra)
 set.seed(1025)
@@
 Then we use this series of models to estimate the latent Gaussian function. Each subject will have a different value of latent function at each bin.
+
+<span style="color: red;">When fitting local GLMMs, it seems that the intervals are essentially independent.
+We do not consider the effect of neighborhood intervals on a specific interval.
+This means that correlation between repeated measures from the same subject is only introduced by the latent function.</span>
 
@@
 
-in_pred_clean <- function(subjID){
+in_pred_clean <- function(subjID=1, mid=mid){
   # make predictions
@@
   pred_t595 <- in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==subjID & bin<=595))
-
+  # put predicted function and standard error in to a data frame
+  # calculate confidence interval
   df_pred_t195 <- data.frame(mid,
@@
     mutate(eta_lower = eta-eta_std, 
-             eta_upper = eta+eta_std) %>% 
-    filter(mid > 195)
+             eta_upper = eta+eta_std)
 
@@
     mutate(eta_lower = eta-eta_std, 
-             eta_upper = eta+eta_std) %>% 
-    filter(mid > 395)
+             eta_upper = eta+eta_std)
 
@@
     mutate(eta_lower = eta-eta_std, 
-             eta_upper = eta+eta_std) %>% 
-  filter(mid > 595)
+             eta_upper = eta+eta_std) 
+  # remove observed part
+  df_pred_t195[df_pred_t195$mid<=195, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
+  df_pred_t395[df_pred_t195$mid<=395, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
+  df_pred_t595[df_pred_t195$mid<=595, c("eta", "eta_std", "eta_lower", "eta_upper")] <- NA
+  # put into one data frame
+  df_pred <- bind_rows(df_pred_t195, df_pred_t395, df_pred_t595, .id="type") %>%
+    mutate(type = factor(type, levels=1:3, 
+                         labels = c("Up to 195", "Up to 395", "Up to 595")))
 
-  return(list(df_pred_t195, df_pred_t395, df_pred_t595))
+  return(df_pred)
 }
@@
 
+```{r}
+# Try on subjects 1-4
+in_pred_subj1 <- in_pred_clean(1, mid) %>% 
+  left_join(df_true_eta %>% filter(id==1), by = "mid") 
+in_pred_subj2 <- in_pred_clean(2, mid) %>% 
+  left_join(df_true_eta %>% filter(id==2), by = "mid")
+in_pred_subj3 <- in_pred_clean(3, mid) %>% 
+  left_join(df_true_eta %>% filter(id==3), by = "mid") 
+in_pred_subj4 <- in_pred_clean(4, mid) %>% 
+  left_join(df_true_eta %>% filter(id==4), by = "mid") 
+```
+
 ```{r, fig.width=12, fig.height=4}
-# Try on subjects 1-4
-bind_rows(in_pred_clean(1), .id = "type") %>%
-  mutate(type=factor(type, levels=1:3, 
-                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
-  left_join(df_true_eta %>% filter(id==1), by = "mid") %>% 
+in_pred_subj1 %>%
   ggplot()+
-  geom_line(aes(x=mid, y = true, col = "True"))+
+  geom_line(aes(x=mid, y = true), col = "red")+
   geom_line(aes(x=mid, y = eta), na.rm = T)+
@@
 
-bind_rows(in_pred_clean(2), .id = "type") %>%
-  mutate(type=factor(type, levels=1:3, 
-                     labels = c("Up to 195", "Up to 395", "Up to 595"))) %>%
-  left_join(df_true_eta %>% filter(id==2), by = "mid") %>% 
+in_pred_subj2 %>% 
   ggplot()+
-  geom_line(aes(x=mid, y = true, col = "True"))+
+  geom_line(aes(x=mid, y = true), col = "red")+
   geom_line(aes(x=mid, y = eta), na.rm = T)+
@@
   facet_wrap(~type)
+
+in_pred_subj3 %>% 
+  ggplot()+
+  geom_line(aes(x=mid, y = true), col = "red")+
+  geom_line(aes(x=mid, y = eta), na.rm = T)+
+  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 3", x = "bin", y="")+
+  facet_wrap(~type)
+
+in_pred_subj4 %>% 
+  ggplot()+
+  geom_line(aes(x=mid, y = true), col = "red")+
+  geom_line(aes(x=mid, y = eta), na.rm = T)+
+  geom_line(aes(x=mid, y=eta_lower), linetype = "dashed", na.rm=T)+
+  geom_line(aes(x=mid, y = eta_upper), linetype = "dashed", na.rm=T)+
+  labs(title = "Subject 4", x = "bin", y="")+
+  facet_wrap(~type)
 ```
-
-## Some clarification points
-
-1. When fitting local GLMMs, it seems that the intervals are essentially independent. We do not consider the effect of neighborhood intervals on a specific interval.
 
@@
 
-#### In- and out-of-sample prediction of latent function
+### In-/Out-of-sample prediction of latent function
 
@@
 ```{r}
-# out-of-sample
-pred1_t195 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 195) %>%
+# clean out-of-sample prediction results
+out_pred_clean <- function(subjID, mid=mid){
+  # Prediction
+  pred_t195 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=195) %>%
                                   select(-eta_i),
                                 fpca_fit = fpca_fit)
-pred1_t395 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 395) %>%
+  pred_t395 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=395) %>%
+                                  select(-eta_i),
+                                fpca_fit = fpca_fit)
+  pred_t595 <- out_samp_dyn_pred(df_new = df %>% filter(id==subjID & sind_inx<=595) %>%
                                   select(-eta_i),
@@
 
-pred1_t595 <- out_samp_dyn_pred(df_new = df %>% filter(id == 1 & sind_inx <= 595) %>%
-                                  select(-eta_i),
-                                fpca_fit = fpca_fit)
+  # clean
+  df_pred <- data.frame(mid, 
+                        pred_t195=pred_t195$eta_pred,
+                        pred_t395=pred_t395$eta_pred, 
+                        pred_t595=pred_t595$eta_pred)
+  df_pred[df_pred$mid<=195, "pred_t195"] <- NA
+  df_pred[df_pred$mid<=395, "pred_t395"] <- NA
+  df_pred[df_pred$mid<=595, "pred_t595"] <- NA
 
-df_out_pred1 <- data.frame(mid, pred1_t195,pred1_t395, pred1_t595)
-df_out_pred1$pred1_t195[1:which(mid==195)] <- NA
-df_out_pred1$pred1_t395[1:which(mid==395)] <- NA
-df_out_pred1$pred1_t595[1:which(mid==595)] <- NA
-df_out_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]
+  df_pred <- df_pred %>% pivot_longer(starts_with("pred_t"), 
+                                      names_to = "type", values_to = "eta") %>%
+    mutate(type = factor(type, levels=c("pred_t195", "pred_t395", "pred_t595"),
+                         labels=c("Up to 195", "Up to 395", "Up to 595")))
+
+  return(df_pred)
+}
 
@@
 ```{r}
-# in-sample  
-df_in_pred1 <- data.frame(mid,
-                          pred_t195 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=195))$eta_pred,
-                          pred_t395 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=395))$eta_pred,
-                          pred_t595 = in_samp_dyn_pred(fpca_fit, df_pred_unique %>% filter(id==1 & bin<=595))$eta_pred)
+out_pred_subj1 <- out_pred_clean(1, mid) %>% 
+  left_join(df_true_eta %>% filter(id==1), by = "mid")
+out_pred_subj2 <- out_pred_clean(2, mid) %>% 
+  left_join(df_true_eta %>% filter(id==2), by = "mid")
+out_pred_subj3 <- out_pred_clean(3, mid) %>% 
+  left_join(df_true_eta %>% filter(id==3), by = "mid")
+out_pred_subj4 <- out_pred_clean(4, mid) %>% 
+  left_join(df_true_eta %>% filter(id==4), by = "mid")
+```
 
-df_in_pred1$pred_t195[1:which(mid==195)] <- NA
-df_in_pred1$pred_t395[1:which(mid==395)] <- NA
-df_in_pred1$pred_t595[1:which(mid==595)] <- NA
-df_in_pred1$true <- df[df$id==1 & df$sind_inx%in%mid, "eta_i"]
+```{r, fig.width=8, fig.height=4}
+# compare point prediction with in-sample
+# subject 1
+grid.arrange(
+  out_pred_subj1 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 1 out-of-sample"),
+
+  in_pred_subj1 %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 1 in-sample"),
+  nrow=1)
+
+# subject 2
+grid.arrange(
+  out_pred_subj2 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 2 out-of-sample"),
+
+  in_pred_subj2 %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 2 in-sample"),
+  nrow=1)
+
+# subject 3
+grid.arrange(
+  out_pred_subj3 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 3 out-of-sample"),
+
+  in_pred_subj3 %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 3 in-sample"),
+  nrow=1)
+
+# subject 4
+grid.arrange(
+  out_pred_subj4 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 4 out-of-sample"),
+
+  in_pred_subj4 %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col=type, group=type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 4 in-sample"),
+  nrow=1)
 
@@
 
+### In-/Out-of-sample PC scores
+
+<span style="color: red;">Doesn't look biased? </span>
+<span style="color: red;">Perhaps because the function value range is so small? </span>
+
 ```{r}
-# compare point prediction with in-sample
+# true scores:xi
 
-## out-of-sample
-ggplot(df_out_pred1)+
-  geom_line(aes(x=mid, y=true, col="True"))+
-  geom_line(aes(x=mid, y=pred1_t195, col="Up to 195"), na.rm = T)+
-  geom_line(aes(x=mid, y=pred1_t395, col="Up to 395"), na.rm = T)+
-  geom_line(aes(x=mid, y=pred1_t595, col="Up to 595"), na.rm = T)+
-  labs(y = "Out-of-sample prediction", x = "Binned time")
+# scores
+in_score <- matrix(NA, nrow=N, ncol=4)
+out_score <- matrix(NA, nrow=N, ncol=4)
+for(i in 1:N){
+  in_score[i, ] <- in_samp_dyn_pred(fpca_fit=fpca_fit, 
+                   df_part=df_pred_unique %>% filter(id==i & bin<=595))$score
+  out_score[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
+                                  select(-eta_i),
+                                  fpca_fit=fpca_fit)$score_out
+}
 
-## in-sample
-ggplot(df_in_pred1)+
-  geom_line(aes(x=mid, y=true, col="True"))+
-  geom_line(aes(x=mid, y=pred_t195, col="Up to 195"), na.rm = T)+
-  geom_line(aes(x=mid, y=pred_t395, col="Up to 395"), na.rm = T)+
-  geom_line(aes(x=mid, y=pred_t595, col="Up to 595"), na.rm = T)+
-  labs(y = "In-sample prediction", x = "Binned time")
+# bias
+in_bias <- in_score-xi
+out_bias <- out_score-xi
+```
+
+```{r, fig.height=4, fig.width=8}
+grid.arrange(
+  out_bias %>% data.frame() %>%
+    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
+    ggplot()+
+    geom_boxplot(aes(x=PC, y=Bias))+
+    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
+    labs(title = "Out-of-sample"),
+
+  in_bias %>% data.frame() %>%
+    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
+    ggplot()+
+    geom_boxplot(aes(x=PC, y=Bias))+
+    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
+    labs(title = "In-sample"), 
+nrow=1)
 
@@
 
-- Repeat this for more subjects
-- Interval prediction for MLE
-- Bias of score for both in-sample and out-of-sample estimation
-
 # Next steps:
 
-1. Write out the formula for the BLUPs in the local GLMMs. Since we have a random intercept only model this should not be too complicated. (this is for out of sample prediction only)
-
-2. Approximate these values numerically via numeric integration (like we did witht he AUC paper) and/or explicit laplace approximation 
-3. Evaluate performance of dynamic predictions (both using "in sample" data -- those individuals whose BLUPs we can extract directly from the local glmm fits and out of sample data)
-4. Write an R package (see Julia's fastGFPCA and Andrew's fcr pacakge as inspiration) which has functions for performing dynamic prediction using our framework
-5. Implement a brief simulation study showing both that the approach works
-
-6. Deriving unbiased scores
-
-As in our first paper (attached), the scores we obtain from the approach you're using are biased.
-We can show that the scores obtained from the local models (i.e.
-Step 3 in the fastGFPCA approach) are a linear function of the true scores.
-1) Derive the result that the binned scores are linear in the true scores (I have already derived this result, but you should attempt to do this yourself.)
-- Attempt to find an analytic solution for the bias factor. This may be challenging, so don't spend too much time here, but it would be an amazing result if we could do so
-
-2) Assuming you cannot find an analytic solution for the bias factor, attempt to estimate this factor yourself.
-To do so implement the full fastGFPCA approach (i.e.
-Step 4, where we re-estimate the scores), then:
-- Take the estimated scores for each subject from step 4 and and regress them on the scores from step 3 using linear regression. Do this separately for each eigenfunction
--  Identify and report the estimated coefficients and R^2 for each eigenfunction separately
-- Use these models to create a de-biasing model for the Step 3 scores
-Show these scores are closer (lower MSE) to the true scores than those obtained from Step 3, and that these result in lower MSE for the latent subject specific functoions
-- Incorporate this result into the dynamic prediction framework you developed per the instructions above
-
-7. FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?
-
-Thoughts on 1:
-
-1. If we'd like to make predictions on more than one sample, we may be able to fit GLMM on test samples and predict $\eta(t)$
-2. Do a different set of derivation with MLE+EB like [this paper](https://www.jstor.org/stable/pdf/2533715.pdf?refreqid=excelsior%3A135dd765c41e4b46eb39bb6ef7a71053&ab_segments=&origin=&acceptTC=1)
+- Interval prediction for MLE
+- FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?

### 2023-01-13 Code/InSampPred.R

@@ -29,6 +29,7 @@ in_samp_dyn_pred <- function(fpca_fit, df_part){
               score_hat = score_hat))
 }
 
+# return: predicted latent function, its variance/std, PC scores
 
 # try <- in_samp_dyn_pred(fpca_fit, 60, df_pred_unique, 4)
 # 

### 2023-01-13 Code/OutSampMLE.R

@@ -45,7 +45,8 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
   score_out <- multiroot(f=llh_div, start = rep(0, 4), df_new=df_new, fpca_fit=fpca_fit)$root
   eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions%*%score_out
   
-  return(eta_pred_out)
+  return(list(eta_pred = eta_pred_out,
+              score_out = score_out))
   
 }
 

### 2023-01-23 f-GFPCA.Rmd
try ssh
prediction interval

@@
 ---
-title: "fGFPCA dynamic prediction (In-sample)"
+title: "fGFPCA dynamic prediction"
 output: 
@@
 
+### Prediction interval
+
+With MLE of scores $\hat{\boldsymbol{\xi}}$
+
+$$\boldsymbol{\hat{\eta}}=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}$$
+$$\begin{aligned}
+\hat{\boldsymbol{\eta}}-\boldsymbol{\eta}&=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}-\boldsymbol{\eta}\\
+&= \boldsymbol{\hat{f_0}-f_0}+\boldsymbol{\Phi(\hat{\xi}-\xi)}-\boldsymbol{\epsilon}
+\end{aligned}$$
+
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi\Gamma \Phi^T}+\hat{\sigma_{\epsilon}}^2$$
+Note:
+
+- $\boldsymbol{\Gamma}$ is estimated from FPCA. $\hat{\boldsymbol{\xi}}$ is also derived conditional on this covariance matrix.
+- <span style="color: red;"> ? Covariance of $\boldsymbol{\hat{f_0}-\eta}$ and $\hat{\boldsymbol{\xi}}$ should be 0 by assumption? </span>
+- <span style="color: red;"> $Var(\boldsymbol{\hat{f_0}-f_0})$ should not change from in-sample prediction formula since it is only related to the FPCA part? </span>
+
+$$Var(\boldsymbol{\hat{f_0}-f_0})=\boldsymbol{\Phi(H\Phi_m\Gamma)\Phi^T}$$
+
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=\boldsymbol{\Phi(H\Phi_m+I_K)\Gamma\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I}_m$$
+
 # Next steps:

### 2023-01-23 Code/OutSampMLE.R

@@ -47,11 +47,24 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
   
   return(list(eta_pred = eta_pred_out,
               score_out = score_out))
-  
 }
 
 
-# test
+#### Estimate standard error of MLE ####
+
+# # Jacabian matrix of first dirivative of llh
+# try <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+#                     select(-eta_i),
+#                   fpca_fit = fpca_fit)$score_out
+# 
+# I_mat <- -gradient(f = llh_div, x = try,
+#          df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+#            select(-eta_i),
+#          fpca_fit=fpca_fit)
+# 
+# var <- diag(solve(I_mat))
+# 1.96*sqrt(var)
+# # test
 # out_samp_dyn_pred(df_new, fpca_fit)
 
 

### 2023-01-25 f-GFPCA.Rmd
try laplace demon package

@@
 
+#### Explicit laplace approximation
+
+- Idea: the posterior distribution approaches a Gaussian distribution as sample size grows. 
+- <span style="color: red;"> Looks like it is also maximizing posterior log-likelihood </span>
+- <span style="color: red;"> How to use Laplace Approximation to approximate integrals? </span>
+
+https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
+
 ### In-/Out-of-sample PC scores

### 2023-01-25 Code/OutSampBayes.R

@@ -1,0 +1,121 @@
+library(cubature)
+# This script writes functions for 
+# calculating conditional expectation
+# of fPCA scores 
+# of an out-of-sample observation
+
+# first, only subject one, time up to 395
+
+#### Numeric integration: denominator of E(xi|Y) ####
+# xi: scores. The value to integrate over
+# tao: estimated covariance matrix(eigenvalues) of score from fPCA
+# f0: mean function from fPCA (up to mth bin)
+# phi: basis function of fPCA (up to mth bin)
+# Y: newly observed data (up to mth bin)
+
+# just for mess around
+tao <- diag(fpca_fit$evalues)
+f0 <- fpca_fit$mu[1:max_bin]
+phi <- ç
+df_out <- df %>% filter(id == 1 & sind_inx <= 395) %>% select(-eta_i)
+max_bin <- which(mid==395)
+xi <- fpca_fit$scores[1, ]
+xi <- matrix(xi, nrow = 4, ncol = 1)
+k <- 4
+  
+joint_pdf <- function(xi, tao, f0, phi, df_out){
+  
+  # numbers
+  ns <- table(df_out$bin) # number of observations
+  hs <- df_out %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+    select(Y) # number of success
+  nf <- ns-hs # number of failure
+  
+  # probability of conditional distribution of Y
+  eta = f0+phi%*%xi
+  pr = exp(eta)/(1+exp(eta))
+  # the part relevant to P(Y|xi)
+  p1 <- prod(dbinom(x=unlist(hs), size = ns, prob=pr))
+  p2 <- exp(-t(xi)%*%solve(tao)%*%xi/2)
+  
+  return(p1*p2)
+  
+}
+
+#test 
+joint_pdf(xi, tao, f0, phi, df_out)
+
+adaptIntegrate(joint_pdf, lower = rep(-500, K), upper = rep(500, K), 
+          tao=tao, f0=f0, phi=phi, df_out=df_out)
+  
+# takes a really long time to run!
+
+#### Laplace approximation ####
+
+library(LaplacesDemon)
+library(mvtnorm)
+
+# experiment with one subject
+
+# data
+df_new <- df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i)
+ns <- as.vector(table(df_new$bin)) # number of observations
[…]

### 2023-01-25 Scripts/LaplacesDemon.R

@@ -1,0 +1,3 @@
+library(LaplacesDemon)
+
+data("demonsnacks")

### 2023-01-26 f-GFPCA.Rmd
more notes
try but failed to get prediction interval

--- 2023-01-23 f-GFPCA.Rmd
@@
 
-```{r}
+```{r, results='hide'}
 source(here("Code/OutSampMLE.R"))
+source(here("Code/OutSampBayes.R"))
 ```
 
-```{r}
-# clean out-of-sample prediction results
+```{r, MLE_pred}
+# clean MLE out-of-sample prediction results
 out_pred_clean <- function(subjID, mid=mid){
@@
 
-```{r, fig.width=8, fig.height=4}
+```{r, laplace_ped}
+# clean MLE out-of-sample prediction results
+out_pred_clean_laplace <- function(subjID=1, mid=mid){
+  # Prediction
+  pred_t195 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=195) %>% select(-eta_i))
+  pred_t395 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=395) %>% select(-eta_i))
+  pred_t595 <- out_pred_laplace(fpca_fit = fpca_fit, df_new = df %>% filter(id==subjID & sind_inx<=595) %>% select(-eta_i))
+
+  # clean
+  df_pred <- data.frame(mid, 
+                        pred_t195=pred_t195$eta_pred,
+                        pred_t395=pred_t395$eta_pred, 
+                        pred_t595=pred_t595$eta_pred)
+  df_pred[df_pred$mid<=195, "pred_t195"] <- NA
+  df_pred[df_pred$mid<=395, "pred_t395"] <- NA
+  df_pred[df_pred$mid<=595, "pred_t595"] <- NA
+
+  df_pred <- df_pred %>% pivot_longer(starts_with("pred_t"), 
+                                      names_to = "type", values_to = "eta") %>%
+    mutate(type = factor(type, levels=c("pred_t195", "pred_t395", "pred_t595"),
+                         labels=c("Up to 195", "Up to 395", "Up to 595")))
+
+  return(df_pred)
+}
+
+```
+
+```{r, results='hide'}
+out_pred_lap_subj1 <- out_pred_clean_laplace(1, mid) %>% 
+  left_join(df_true_eta %>% filter(id==1), by = "mid")
+out_pred_lap_subj2 <- out_pred_clean_laplace(2, mid) %>% 
+  left_join(df_true_eta %>% filter(id==2), by = "mid")
+out_pred_lap_subj3 <- out_pred_clean_laplace(3, mid) %>% 
+  left_join(df_true_eta %>% filter(id==3), by = "mid")
+out_pred_lap_subj4 <- out_pred_clean_laplace(4, mid) %>% 
+  left_join(df_true_eta %>% filter(id==4), by = "mid")
+```
+
+```{r, fig.width=12, fig.height=4}
 # compare point prediction with in-sample
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 1 out-of-sample"),
+    labs(x="Binned time", title = "Subject 1 out-of-sample MLE"),
+
+  out_pred_lap_subj1 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample"),
+    labs(x="Binned time", title = "Subject 2 out-of-sample MLE"),
+
+  out_pred_lap_subj2 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 3 out-of-sample"),
+    labs(x="Binned time", title = "Subject 3 out-of-sample MLE"),
+
+  out_pred_lap_subj3 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 3 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 4 out-of-sample"),
+    labs(x="Binned time", title = "Subject 4 out-of-sample MLE"),
+
+  out_pred_lap_subj4 %>% 
+    group_by("type") %>%
+    ggplot()+
+    geom_line(aes(x=mid, y=true), col = "black")+
+    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+              show.legend = FALSE)+
+    labs(x="Binned time", title = "Subject 4 out-of-sample Laplace"),
 
@@
 
+#### Laplace approximation
+
+This method essential finds $\boldsymbol{\xi}$ that maximizes $p(\boldsymbol{Y}|\boldsymbol{\xi})$.
+
+- <span style="color: red;"> Is it equivalent to posterior mean $E(\boldsymbol{\xi}|\boldsymbol{Y})$ under certain conditionss? (according to decision theory) </span>
+- <span style="color: red;"> How to use Laplace Approximation to approximate integrals? </span>
+
+https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
+
 ### In-/Out-of-sample PC scores
@@
 
-```{r}
+```{r, results='hide'}
 # true scores:xi
@@
 in_score <- matrix(NA, nrow=N, ncol=4)
-out_score <- matrix(NA, nrow=N, ncol=4)
+out_score_MLE <- matrix(NA, nrow=N, ncol=4)
+out_score_laplace <- matrix(NA, nrow=N, ncol=4)
+
 for(i in 1:N){
@@
                    df_part=df_pred_unique %>% filter(id==i & bin<=595))$score
-  out_score[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
+  out_score_MLE[i,] <- out_samp_dyn_pred(df %>% filter(id==i & sind_inx<=595) %>%
                                   select(-eta_i),
                                   fpca_fit=fpca_fit)$score_out
+  out_score_laplace[i,] <- out_pred_laplace(fpca_fit=fpca_fit, 
+                                            df_new = df %>% filter(id==i & sind_inx<=595) %>% select(-eta_i))$score_out
 }
@@
 in_bias <- in_score-xi
-out_bias <- out_score-xi
+out_bias_MLE <- out_score_MLE-xi
+out_bias_laplace <- out_score_laplace-xi
 ```
@@
 grid.arrange(
-  out_bias %>% data.frame() %>%
+  out_bias_MLE %>% data.frame() %>%
     pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
@@
     stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
-    labs(title = "Out-of-sample"),
+    labs(title = "Out-of-sample MLE"),
+
+  out_bias_laplace %>% data.frame() %>%
+    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
+    ggplot()+
+    geom_boxplot(aes(x=PC, y=Bias))+
+    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
+    labs(title = "Out-of-sample laplace"),
 
@@
 $$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi\Gamma \Phi^T}+\hat{\sigma_{\epsilon}}^2$$
+
+What happens to the variance of mean function? In practice it is estimated by smoothed population mean at each observation time.
+
 Note:

### 2023-01-26 Code/OutSampBayes.R

@@ -14,14 +14,14 @@ library(cubature)
 # Y: newly observed data (up to mth bin)
 
 # just for mess around
-tao <- diag(fpca_fit$evalues)
-f0 <- fpca_fit$mu[1:max_bin]
-phi <- ç
-df_out <- df %>% filter(id == 1 & sind_inx <= 395) %>% select(-eta_i)
-max_bin <- which(mid==395)
-xi <- fpca_fit$scores[1, ]
-xi <- matrix(xi, nrow = 4, ncol = 1)
-k <- 4
+# tao <- diag(fpca_fit$evalues)
+# f0 <- fpca_fit$mu[1:max_bin]
+# phi <- ç
+# df_out <- df %>% filter(id == 1 & sind_inx <= 395) %>% select(-eta_i)
+# max_bin <- which(mid==395)
+# xi <- fpca_fit$scores[1, ]
+# xi <- matrix(xi, nrow = 4, ncol = 1)
+# k <- 4
   
 joint_pdf <- function(xi, tao, f0, phi, df_out){
   
@@ -43,10 +43,10 @@ joint_pdf <- function(xi, tao, f0, phi, df_out){
 }
 
 #test 
-joint_pdf(xi, tao, f0, phi, df_out)
-
-adaptIntegrate(joint_pdf, lower = rep(-500, K), upper = rep(500, K), 
-          tao=tao, f0=f0, phi=phi, df_out=df_out)
+# joint_pdf(xi, tao, f0, phi, df_out)
+# 
+# adaptIntegrate(joint_pdf, lower = rep(-500, K), upper = rep(500, K), 
+#           tao=tao, f0=f0, phi=phi, df_out=df_out)
   
 # takes a really long time to run!
 
@@ -55,42 +55,7 @@ adaptIntegrate(joint_pdf, lower = rep(-500, K), upper = rep(500, K),
 library(LaplacesDemon)
 library(mvtnorm)
 
-# experiment with one subject
-
-# data
-df_new <- df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i)
-ns <- as.vector(table(df_new$bin)) # number of observations
-hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
-  select(Y) %>% unlist()# number of success
-nf <- ns-hs # number of failure
-max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
-
[…]

### 2023-01-26 f-GFPCA.Rmd
interval prediction

@@
 
-$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi\Gamma \Phi^T}+\hat{\sigma_{\epsilon}}^2$$
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi Var(\hat{\xi}-\xi) \Phi^T}+Var(\epsilon)$$
 
-What happens to the variance of mean function? In practice it is estimated by smoothed population mean at each observation time.
+- The three components should be mutally independent? 
+- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}\boldsymbol{\Phi\Gamma \Phi}+\sigma_{\epsilon}^2\boldsymbol{I}_m$.
+- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation
+- $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
+- Very different approach from in-sample estimation
 
-Note:
+Additional questions:
 
-- $\boldsymbol{\Gamma}$ is estimated from FPCA. $\hat{\boldsymbol{\xi}}$ is also derived conditional on this covariance matrix.
-- <span style="color: red;"> ? Covariance of $\boldsymbol{\hat{f_0}-\eta}$ and $\hat{\boldsymbol{\xi}}$ should be 0 by assumption? </span>
-- <span style="color: red;"> $Var(\boldsymbol{\hat{f_0}-f_0})$ should not change from in-sample prediction formula since it is only related to the FPCA part? </span>
+Variance of $\boldsymbol{\xi}$ and random error is in fact given from the FPCA procedure. But we need to variance of mean function for the actual prediction interval.
 
-$$Var(\boldsymbol{\hat{f_0}-f_0})=\boldsymbol{\Phi(H\Phi_m\Gamma)\Phi^T}$$
-
-$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=\boldsymbol{\Phi(H\Phi_m+I_K)\Gamma\Phi^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I}_m$$
+What happens to the variance of mean function?
+In practice it is estimated by smoothed population mean at each observation time.
+It looks like in existing derivations, outcomes are usually assumed to be centered.
 

### 2023-01-26 Scripts/MessAround.R

@@ -1,0 +1,4 @@
+head(fpca_fit$mu)
+
+df_pred_latent %>% select(id, bin, eta_hat) %>% distinct(.) %>%
+  group_by(bin) %>% summarise_at("eta_hat", mean)

### 2023-01-30 Code/OutSampBayes.R
hand code laplace approximation

@@ -50,7 +50,7 @@ joint_pdf <- function(xi, tao, f0, phi, df_out){
   
 # takes a really long time to run!
 
-#### Laplace approximation ####
+#### Laplace approximation with Bayes package ####
 
 library(LaplacesDemon)
 library(mvtnorm)
@@ -127,3 +127,5 @@ out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=
 
 # out_pred_laplace(fpca_fit,  df %>% filter(id==2 & sind_inx<=395) %>% select(-eta_i))
 
+##### Hand-coded Laplace Approximation #####
+s

### 2023-01-30 Code/OutSampLaplace.R

@@ -1,0 +1,48 @@
+
+# This script trys to approximate the integrals involved in posterior mean of scores
+# using laplace approximation
+
+library(mvtnorm)
+library(rootSolve)
+##### Marginal distribution of outcome #####
+
+# first, approximate the denominator
+# We need: 
+# joint log-likelihood of outcome and score
+# MLE of score
+# second derivative of joint log-likelihood wrt scores
+
+# subject 1 with observations up to t=195
+
+df_new <- df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i)
+
+ns <- as.vector(table(df_new$bin)) # number of observations
+hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
+nf <- ns-hs # number of failure
+max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
+
+# additional information needed from FPCA 
+tao <- diag(fpca_fit$evalues)
+f0 <- fpca_fit$mu[1:max_bin]
+Phi <- fpca_fit$efunctions[1:max_bin, ]
+K <- ncol(Phi)
+
+# function for joint likelihood
+
+joint_llh <- function(xi){
+  eta <- f0+ Phi%*%xi
+  p <- exp(eta)/(1+exp(eta))
+  pos_y_llh <- sum(dbinom(x=hs, size = ns, prob=p, log=T)) # posterior log-likelihood of outcome
+  pri_xi_llh <- dmvnorm(xi, mean = rep(0, K), sigma=tao, log = T) # prior log-likelihood of scores
+  return(pos_y_llh+pri_xi_llh)
+}
+
+# MLE
+maximum <- optim(par=rep(0, K), fn = joint_llh, control = list(fnscale=-1))
+mle_score <-maximum$par
+mle_max <- maximum$value
+
+# 2nd derivative
+joint_llh_div2 <- function(xi = mle_score){
+  eta <- f0+ Phi%*%xi
+}

### 2023-02-01 f-GFPCA.Rmd
hand code laplace approximation
Laplace and MLE are actually the same thing

@@
 
+#### Laplace approximation
+
+Posterior mean from Laplace method is the same as posterior mode. In this sense, empirical bayes method and MLE are in fact equivalent.
+
+https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
+https://bookdown.org/rdpeng/advstatcomp/laplace-approximation.html
+
 ### In-/Out-of-sample prediction of latent function
 
-```{r, results='hide'}
+```{r, results='hide', message=FALSE}
 source(here("Code/OutSampMLE.R"))
@@
 
-```{r, results='hide'}
+```{r, results='hide', message=FALSE}
 out_pred_lap_subj1 <- out_pred_clean_laplace(1, mid) %>% 
@@
 
-#### Laplace approximation
-
-This method essential finds $\boldsymbol{\xi}$ that maximizes $p(\boldsymbol{Y}|\boldsymbol{\xi})$.
-
-- <span style="color: red;"> Is it equivalent to posterior mean $E(\boldsymbol{\xi}|\boldsymbol{Y})$ under certain conditionss? (according to decision theory) </span>
-- <span style="color: red;"> How to use Laplace Approximation to approximate integrals? </span>
-
-https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
-
 ### In-/Out-of-sample PC scores
@@
 
-```{r, results='hide'}
+```{r, results='hide', message=FALSE}
 # true scores:xi
@@
 
-- The three components should be mutally independent? 
-- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}\boldsymbol{\Phi\Gamma \Phi}+\sigma_{\epsilon}^2\boldsymbol{I}_m$.
-- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation
+- The three components should be mutually independent? 
+- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
+- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?ß
 - $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA

### 2023-02-01 Code/OutSampLaplace.R

@@ -75,5 +75,9 @@ maximum2 <- optim(par=mle_score, fn = tfunc, control = list(fnscale=-1))
 mle_score <-maximum$par
 mle_max <- maximum$value
 
+##### joint likelihood of scores and outcome #####
+
+
+
 
     

### 2023-02-07 f-GFPCA.Rmd
update progress report

@@
 
-And this latent funciton can be approximated by fPCA:
+And this latent funciton can be approximated based on Karhunen–Loève expansion:
 
@@
 
-where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$. In the fPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix and diagonal elements corresponds to eigenvalues.
+where $\boldsymbol{\xi}_{i} \sim N(0, \boldsymbol{\Gamma})$.
 
@@
 
+<span style="color: red;"> Question 1: in FPCA framework, $\boldsymbol{\Gamma}$ is a diagonal matrix of eigenvalues.
+However in my inplemetation, the eigenvalues are much larger than true $\boldsymbol{\Gamma}$.
+I am not sure why this happened.
+Does it has to do with the GLMM step, or decreased density when binning observations?
+</span>
+
 ## Local GLMMs
 
-The entire time grid into 100 small intervals with equal length.
+The entire time grid is binned into 100 small intervals with equal length.
 The cut is made on the order index of measurement time, but not the actually time itself.
@@
 
-<span style="color: red;">When fitting local GLMMs, it seems that the intervals are essentially independent.
+<span style="color: red;">Question 2: When fitting local GLMMs, it seems that the intervals are essentially independent.
 We do not consider the effect of neighborhood intervals on a specific interval.
-This means that correlation between repeated measures from the same subject is only introduced by the latent function.</span>
+This means that correlation between repeated measures from the same subject is only introduced by the latent function?
+</span>
 
@@
 
+<span style="color: red;">Follow-up on question 1: It looks like the PC function ranges are different from true PC functions.
+Maybe some of its variation is absorbed by the score, causing $\boldsymbol{\Gamma}$ to be much greater than its true value?</span>
+
 ## In-sample prediction
@@
 
-- $\hat{\boldsymbol{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
+- $\boldsymbol{\hat{\Gamma}}$ is the diagonal covariance matrix of eigenvalues
 - $\boldsymbol{\Phi}_m$ is the matrix of eigenfunctions up to $t_m$. These should be constant across all subjects.
@@
 
-$$\hat{\boldsymbol{\eta}} = \hat{\boldsymbol{f}}_0+\boldsymbol{\Phi}\hat{\boldsymbol{\xi}}$$
+$$\boldsymbol{\hat{\eta}} = \hat{\boldsymbol{f}}_0+\boldsymbol{\Phi}\boldsymbol{\hat{\xi}}$$
 
@@
 
-Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin. I can think of a few ways to  predict the following track, as below:
+Now let's say if we have a new subject $\boldsymbol{Y}$ with observations up to $t_m$ and $t_m$ is in mth bin.
+It seems there are more than one ways to  predict the following track.
+The first thing coming to mind was in fact to 1) refit local GLMMs to estimate a incomplete latent function; 2) refit local FPCA with this additional incomplete latent function track; and 3) repeat score estimation from in-sample procedure.
 
-### Refit local GLMMs to estimate latent function
+However, if we wish to make prediction without re-estimating any model (neither local GLMMs nor FPCA), we need to use the fitted FPCA model (specifically, $\boldsymbol{\hat{f}}_0$, $\boldsymbol{\Phi}$, $\boldsymbol{\hat{\Gamma}}$, $\hat{\sigma}_{\epsilon}$) to estimate the scores for this new observations (out-of-sample scores).
+I can think of two ways of doing this:
 
-1. Local GLMMs: we will be able to estimate random effect of this subject up to mth bin, thus estimate the latent function up to the mth bin ($\hat{\boldsymbol{\eta}}_m$).
+1. Empirical Bayes: similar to in-sample procedure, estimate the posterior mean 
+2. MLE, as often used in longitudinal models.
 
-2. fPCA: We will have an incomplete function for this new observation. I think fpca.face can handle that.
+Below I experimented with both procedures.
 
-3. Use the save formulas as in-sample prediction
-
-### Do not estimate latent function
+#### Empirical Bayes
 
@@
 
-We use the expectation of latent function conditonal on fPCA score to estimate the true latent function, so that the error term can be removed from following calculation:
-
-$$\hat{\eta_s} = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}$$
-
-Here, $\hat{f}_{0}(s)$ is the mean function from fPCA and $\phi(s)$ are eigenfunctions at time s, which is a K $\times$ 1 vector.
-
-Then the data is generated from a binomial distribution:
-
-$$\begin{aligned}
- \boldsymbol{Y}_s & \sim Binomial(p_s) \\
- logit(p_s) & = \hat{\eta}_s = \hat{f}_{0}(s)+\phi(s)\boldsymbol{\xi}
-\end{aligned}$$
-
-#### Empirical Bayes
-
 Now let's find the empirical bayes estimator of score $E(\boldsymbol{\xi}|\boldsymbol{Y})$, which is not likely to have closed form solution.
 
-$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int_{\boldsymbol{\xi}} \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
-$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\int \boldsymbol{\xi} p(\boldsymbol{\xi}|\boldsymbol{Y})d\boldsymbol{\xi}$$
+$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
 
@@
 
-$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int_{\boldsymbol{\xi}}\boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+$$E(\boldsymbol{\xi}|\boldsymbol{Y})=\frac{\int \boldsymbol{\xi}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
 
-We know that fPCA scores follow a multivariate normal distribution:
+We know that FPCA scores follow a multivariate normal distribution:
 
@@
 
-Please note that here, we are assuming $\boldsymbol{Y}_s$ is independent from each other given fPCA score.
+<span style="color:red;">Follow up on question 2: Please note that here, we are assuming $\boldsymbol{Y}_s$ is independent from each other given fPCA score.
 That is, all correlation across time within the same subject is introduced by underlying eigenfunctions.
+</span>
 
-If we write this down for a single score:
+<!-- If we write this down for a single score:  -->
 
-$$E(\xi_k|\boldsymbol{Y})=\frac{\int \xi_k \int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi_{(-k)}}d\xi_k}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+<!-- $$E(\xi_k|\boldsymbol{Y})=\frac{\int \xi_k \int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi_{(-k)}}d\xi_k}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$ -->
 
-In this case, I ran into the problem of numeric integration being too slow (adaptIntegrate multidimensional). The function value is also very small. Therefore, I turned to MLE instead.
+It is very difficult to get a closed-form solution of this formula.
+Therefore, I tried to approximate the integrals with numeric integration (*adaptIntegrate* function from *cubature* package).
+In this case, I ran into the problem of computation time.
+Numeric integration was very, very slow, taking roughly an hour for just the denominator of one subject.
+The function value is also very small.
 
@@
 
-We utilize the same conditional likelihood of scores:
+We maximize the same conditional likelihood of scores (conditional on observations):
 
-$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int_{\boldsymbol{\xi}}p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
+$$p(\boldsymbol{\xi}|\boldsymbol{Y}) = \frac{p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})}{\int p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi})d\boldsymbol{\xi}}$$
 
-But now we don't need to calculated the marginal distribution of $\boldsymbol{Y}$ anymore (denominator). So we'll get:
+But now we don't need to calculated the marginal distribution of $\boldsymbol{Y}$ anymore (denominator), since it would be constant wrt $\boldsymbol{\xi}$:
 
@@
 p(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto p(\boldsymbol{Y}|\boldsymbol{\xi})p(\boldsymbol{\xi}) \\
-
 l(\boldsymbol{\xi}|\boldsymbol{Y}) & \propto logp (\boldsymbol{Y}|\boldsymbol{\xi})+log p(\boldsymbol{\xi})
@@
 
-It is not easy to get a closed-form solution for the score equation above. We can try to get a numeric solution instead. Below are some visualization of the prediction performances
+It is not easy to get a closed-form solution for the score equation above. We can try to get a numeric solution instead (*multiroot* function from *rootSolve* package).
 
@@
 
-Posterior mean from Laplace method is the same as posterior mode. In this sense, empirical bayes method and MLE are in fact equivalent.
+It seems that $E(\boldsymbol{\xi}|\boldsymbol{Y})$ and $p(\boldsymbol{\xi}|\boldsymbol{Y})$ can be connected through Laplace approximation.
+Specifically, posterior mean estimated from Laplace method is the same as posterior mode.
+In this sense, the two methods above are essentially equivalent.
 
-https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
-https://bookdown.org/rdpeng/advstatcomp/laplace-approximation.html
+Below I have experimented with *LaplaceDemon::LaplaceApproximation*: https://search.r-project.org/CRAN/refmans/LaplacesDemon/html/LaplaceApproximation.html
 
-### In-/Out-of-sample prediction of latent function
+<!-- https://bookdown.org/rdpeng/advstatcomp/laplace-approximation.html -->
+
+### In-/Out-of-sample prediction results of latent function
 
@@
 
-<span style="color: red;">Doesn't look biased? </span>
-<span style="color: red;">Perhaps because the function value range is so small? </span>
+Here I explored the estimated scores, comparing them with the true scores from simulation setup.
+The figures actually didn't look biased to me.
+Perhaps it is an artifact of small functional range.
+Need to think about that.
 
@@
 
-With MLE of scores $\hat{\boldsymbol{\xi}}$
+<span style="color:red;"> Questions 3: a lot of problems on this part </span>
+
+With an estimate of scores $\hat{\boldsymbol{\xi}}$
 
@@
 - Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
-- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?ß
+- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?
 - $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
-- Very different approach from in-sample estimation
-
-Additional questions:
-
-Variance of $\boldsymbol{\xi}$ and random error is in fact given from the FPCA procedure. But we need to variance of mean function for the actual prediction interval.
-
-What happens to the variance of mean function?
-In practice it is estimated by smoothed population mean at each observation time.
-It looks like in existing derivations, outcomes are usually assumed to be centered.
+- May need to circle back to in-sample estimation and dig into where that formula came from
 
@@
 - Interval prediction for MLE
+- Bias of PC scores
 - FPCA eigenfunctions: how to extend to the original grid? Interpolation or projection?

### 2023-02-15 f-GFPCA.Rmd
more updates interval prediction
model evaluation metrics

@@
 source(here("Code/OutSampMLE.R"))
-source(here("Code/OutSampBayes.R"))
+# source(here("Code/OutSampBayes.R"))
 ```
+
+#### Figure
 
@@
 
-```{r, laplace_ped}
+```{r, laplace_ped, eval=FALSE}
 # clean MLE out-of-sample prediction results
@@
 
-```{r, results='hide', message=FALSE}
+```{r, results='hide', message=FALSE, eval=FALSE}
 out_pred_lap_subj1 <- out_pred_clean_laplace(1, mid) %>% 
@@
 
-```{r, fig.width=12, fig.height=4}
+```{r, fig.width=8, fig.height=4}
 # compare point prediction with in-sample
@@
     labs(x="Binned time", title = "Subject 1 out-of-sample MLE"),
-
-  out_pred_lap_subj1 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),
 
@@
 
-  out_pred_lap_subj2 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
-
   in_pred_subj2 %>%
@@
     labs(x="Binned time", title = "Subject 3 out-of-sample MLE"),
-
-  out_pred_lap_subj3 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 3 out-of-sample Laplace"),
 
@@
 
-  out_pred_lap_subj4 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 4 out-of-sample Laplace"),
-
   in_pred_subj4 %>%
@@
 
+```
+
+#### Predictive performance measures
+
+- Modeled after dynamic prediction paper (Leroux 2017), focusing on prediction of individual tracks
+- Integrated error: for latent process
+
+$$\frac{1}{N}\sum_{i=1}^N\sum_s(\hat{\eta}_{is}-\eta_{is})^2$$
+where s covers only unobserved tracks
+
+- Integrated AUC: for binary outcome
+- Also, compare with a existing method (e.g, AMM, FRI)
+- Perhaps, also computation time?
+
+<span style="color:red;">The problem with measures above is that predictions are made on a different grid than observations.
+I can think of two potential ways to address this issue: 1) only evaluate at bin midpoints; 2) use interpolation to extend latend function from binned grid to the original grid (in GFPCA paper this is done through projection.
+) </span>
+
+```{r}
+df_true_eta %>% filter(id==1)
+eta_hat_mat <- matrix(NA, ncol = )
+
+out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+                                  select(-eta_i),
+                                fpca_fit = fpca_fit)$eta_pred
 ```
@@
 
+Some new updates:
+
+Actually, after going over the full derivation in Leroux2017 paper, I think the preidction interval completely ignored the randomness from fixed effects.
+In other worlds, prediction intervals are made conditioning on fixed effect coefficients (in our case $\boldsymbol{f_0}$).
+
 # Next steps:

### 2023-02-20 f-GFPCA.Rmd
clean scripts

--- 2023-02-07 f-GFPCA.Rmd
@@
 
-```{r, laplace_ped}
+```{r, laplace_pred}
 # clean MLE out-of-sample prediction results
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 3 out-of-sample Laplace"),
+    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 4 out-of-sample Laplace"),
+    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
 
-### In-/Out-of-sample PC scores
+## Predictive performance measures
+
+#### Measures
+
+- Focusing on prediction of individual tracks
+- Integrated error: for latent process
+
+$$\frac{1}{N}\sum_{i=1}^N\sum_s(\hat{\eta}_{is}-\eta_{is})^2$$
+where s covers only unobserved tracks
+
+- Integrated AUC: for binary outcome
+- Perhaps, also computation time?
+
+<span style="color:red;">The problem with measures above is that predictions are made on a different grid than observations.
+I can think of two potential ways to address this issue: 1) only evaluate at bin midpoints; 2) use interpolation to extend latend function from binned grid to the original grid (in GFPCA paper this is done through projection.
+) </span>
+
+#### Reference methods
+
+Also, we need to compare with a existing method for dynamic prediction of binary outcomes. Ideals, there should be a FDA component to this methods.
+
+1. registr::bfpca:
+
+This method is not a method for prediction.
+If we would like to use it for binary prediction, the two methods will only differ in the first half (FPCA).
+But this comparision has already been made in the GFPCA paper.
+
+This method also has convergence issue.
+
+2. Functional mixed models (Intercept-only)
+
+Can these methods handle binary function? They are also very, very slow to fit on the full dataset.
+
+3. GLMM
+Also had convergence issues.
+
+## In-/Out-of-sample PC scores
 

### 2023-02-20 Code/OutSampBayes.R

@@ -4,24 +4,13 @@ library(cubature)
 # of fPCA scores 
 # of an out-of-sample observation
 
-# first, only subject one, time up to 395
-
 #### Numeric integration: denominator of E(xi|Y) ####
 # xi: scores. The value to integrate over
 # tao: estimated covariance matrix(eigenvalues) of score from fPCA
 # f0: mean function from fPCA (up to mth bin)
 # phi: basis function of fPCA (up to mth bin)
-# Y: newly observed data (up to mth bin)
+# df_out: newly observed data (up to mth bin)
 
-# just for mess around
-# tao <- diag(fpca_fit$evalues)
-# f0 <- fpca_fit$mu[1:max_bin]
-# phi <- ç
-# df_out <- df %>% filter(id == 1 & sind_inx <= 395) %>% select(-eta_i)
-# max_bin <- which(mid==395)
-# xi <- fpca_fit$scores[1, ]
-# xi <- matrix(xi, nrow = 4, ncol = 1)
-# k <- 4
   
 joint_pdf <- function(xi, tao, f0, phi, df_out){
   
@@ -127,5 +116,4 @@ out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=
 
 # out_pred_laplace(fpca_fit,  df %>% filter(id==2 & sind_inx<=395) %>% select(-eta_i))
 
-##### Hand-coded Laplace Approximation #####
-s
+

### 2023-02-20 Code/PredEval.R

@@ -41,11 +41,28 @@ summary(auc_perf)
 
 
 
+##### Reference methods #####
 
+# binary FPCA for dynamic prediction
+# non-convergence
+library(registr)
 
+bfpca_fit <- bfpca(Y= df %>% select(id, Y, sind_inx) %>% rename(index=sind_inx, value = Y), 
+      npc_varExplained = 0.95, t_min=1, t_max = J)
 
+# A standard mxied model
+# time as the only fixed effect
+# random intercept for ID and random slope for time
+# slow, also with numeric issues
+glmm_fit <- glmer(Y ~ sind_inx + (id | sind_inx),
+                  data = df %>% select(id, Y, sind_inx), family = binomial)
 
-
+# functional mixed models
+# generalized function-on-scalar regression
+# marginal constrants? 
+fit_gam <- gam(Y ~ s(sind_inx, k=20, bs = "cr") + ti(id, sind_inx, bs=c("re","cr"), mc=c(TRUE,FALSE), k=c(5,5)), 
+              family = "binomial",
+              data =  df %>% select(id, Y, sind_inx))
 
 
 

### 2023-02-21 f-GFPCA.Rmd
make figure for proposal report:

@@
 
-  out_pred_lap_subj1 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),
+  # out_pred_lap_subj1 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),
 
@@
 
-  out_pred_lap_subj2 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
+  # out_pred_lap_subj2 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
 
-  out_pred_lap_subj3 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
+  # out_pred_lap_subj3 %>% 
+    # group_by("type") %>%
+    # ggplot()+
+    # geom_line(aes(x=mid, y=true), col = "black")+
+    # geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+    #           show.legend = FALSE)+
+    # labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
 
-  out_pred_lap_subj4 %>% 
-    group_by("type") %>%
-    ggplot()+
-    geom_line(aes(x=mid, y=true), col = "black")+
-    geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
-              show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
+  # out_pred_lap_subj4 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
+  #
 

### 2023-02-21 Scripts/Figures.R

@@ -1,0 +1,135 @@
+# compare point prediction with in-sample
+
+in_pred_subj1 %>% select(mid, type, eta, id, true)
+
+p1 <- bind_rows(out_pred_subj1 %>% mutate(sample="Out-of-sample prediction"),
+          in_pred_subj1 %>% select(mid, type, eta, id, true) %>%
+            mutate(sample = "In-sample estimation")) %>%
+  ggplot()+
+  geom_line(aes(x=mid, y=true), col = "black")+
+  geom_line(aes(x=mid, y=eta, col= type, group = type), 
+            linetype="dashed", na.rm = T)+
+  facet_wrap(~sample)+
+  labs(x="", y = "", title = "Subject 1", col = "")
+
+p2 <- bind_rows(out_pred_subj2 %>% mutate(sample="Out-of-sample prediction"),
+                in_pred_subj2 %>% select(mid, type, eta, id, true) %>%
+                  mutate(sample = "In-sample estimation")) %>%
+  ggplot()+
+  geom_line(aes(x=mid, y=true), col = "black")+
+  geom_line(aes(x=mid, y=eta, col= type, group = type), 
+            linetype="dashed", na.rm = T)+
+  facet_wrap(~sample)+
+  labs(x="", y = "", title = "Subject 2", col = "")
+
+p3 <- bind_rows(out_pred_subj3 %>% mutate(sample="Out-of-sample prediction"),
+                in_pred_subj3 %>% select(mid, type, eta, id, true) %>%
+                  mutate(sample = "In-sample estimation")) %>%
+  ggplot()+
+  geom_line(aes(x=mid, y=true), col = "black")+
+  geom_line(aes(x=mid, y=eta, col= type, group = type), 
+            linetype="dashed", na.rm = T)+
+  facet_wrap(~sample)+
+  labs(x="", y = "", title = "Subject 3", col = "")
+
+p4 <- bind_rows(out_pred_subj4 %>% mutate(sample="Out-of-sample prediction"),
+                in_pred_subj4 %>% select(mid, type, eta, id, true) %>%
+                  mutate(sample = "In-sample estimation")) %>%
+  ggplot()+
+  geom_line(aes(x=mid, y=true), col = "black")+
+  geom_line(aes(x=mid, y=eta, col= type, group = type), 
+            linetype="dashed", na.rm = T)+
+  facet_wrap(~sample)+
[…]

### 2023-02-24 f-GFPCA.Rmd
prediction performance evaluation

--- 2023-02-15 f-GFPCA.Rmd
@@
 library(gridExtra)
+library(kableExtra)
 set.seed(1025)
@@
 
-#### Figure
-
 ```{r, MLE_pred}
@@
 
-```{r, laplace_ped, eval=FALSE}
+```{r, laplace_pred, eval=FALSE}
 # clean MLE out-of-sample prediction results
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 1 out-of-sample MLE"),
+    labs(x="Binned time", title = "Subject 1 out-of-sample prediction"),
+
+  # out_pred_lap_subj1 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 1 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 1 in-sample"),
+    labs(x="Binned time", title = "Subject 1 in-sample estimation"),
   nrow=1)
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 out-of-sample MLE"),
+    labs(x="Binned time", title = "Subject 2 out-of-sample prediction"),
+
+  # out_pred_lap_subj2 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 2 in-sample"),
+    labs(x="Binned time", title = "Subject 2 in-sample estimation"),
   nrow=1)
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 3 out-of-sample MLE"),
+    labs(x="Binned time", title = "Subject 3 out-of-sample prediction"),
+
+  # out_pred_lap_subj3 %>% 
+    # group_by("type") %>%
+    # ggplot()+
+    # geom_line(aes(x=mid, y=true), col = "black")+
+    # geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+    #           show.legend = FALSE)+
+    # labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 3 in-sample"),
+    labs(x="Binned time", title = "Subject 3 in-sample estimation"),
   nrow=1)
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 4 out-of-sample MLE"),
+    labs(x="Binned time", title = "Subject 4 out-of-sample prediction"),
+
+  # out_pred_lap_subj4 %>% 
+  #   group_by("type") %>%
+  #   ggplot()+
+  #   geom_line(aes(x=mid, y=true), col = "black")+
+  #   geom_line(aes(x=mid, y=eta, col= type, group = type), linetype="dashed", na.rm = T,
+  #             show.legend = FALSE)+
+  #   labs(x="Binned time", title = "Subject 2 out-of-sample Laplace"),
+  #
 
@@
               show.legend = FALSE)+
-    labs(x="Binned time", title = "Subject 4 in-sample"),
+    labs(x="Binned time", title = "Subject 4 in-sample estimation"),
   nrow=1)
@@
 
-#### Predictive performance measures
+## Predictive performance measures
 
-- Modeled after dynamic prediction paper (Leroux 2017), focusing on prediction of individual tracks
-- Integrated error: for latent process
+This section evaluates the performance of out-of-sample prediction of individual tracks.
+Out reference method is Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive), with a fixed effect of time and subject-level random intercept and slope:
 
-$$\frac{1}{N}\sum_{i=1}^N\sum_s(\hat{\eta}_{is}-\eta_{is})^2$$
-where s covers only unobserved tracks
+$$g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$ 
+We use two metrics for evaluation:
 
-- Integrated AUC: for binary outcome
-- Also, compare with a existing method (e.g, AMM, FRI)
-- Perhaps, also computation time?
+1. Integrated squared error.
 
-<span style="color:red;">The problem with measures above is that predictions are made on a different grid than observations.
-I can think of two potential ways to address this issue: 1) only evaluate at bin midpoints; 2) use interpolation to extend latend function from binned grid to the original grid (in GFPCA paper this is done through projection.
-) </span>
+- Evaluation of the latent Gaussian function
+- For a single subject, only evaluated at unobserved bin midpoints
+
+$$\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_{is}-\eta_{is})^2$$
+
+2. Average AUC
+
+- Evaluation of the binary outcome function
+- AUC for a single subject is calculated over the unobserved time points on the original un-binned grid
+- Predicted probability between bin midpoints are estimated with interpolation
+- Please note that the grid ends at the maximum bin midpoint, instead of the maximum observed time, due to how interpolation works
+
+3. Computation time
+
+This is a bit more complicated. The full compuation time consists of two parts: 
+- Time spent on model fitting
+- Time spent on prediction **given** the fitted FPCA or adaptive GLMM models.
+
+```{r, results='hide', message=FALSE}
+source(here("Code/PredEval.R"))
+```
 
 ```{r}
-df_true_eta %>% filter(id==1)
-eta_hat_mat <- matrix(NA, ncol = )
-
-out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
-                                  select(-eta_i),
-                                fpca_fit = fpca_fit)$eta_pred
+# table displaying predictive performance
+tb_pred_eval <- tibble(
+  c("195", "395", "595"),
+  ise_fgfpca, ise_adglmm, auc_fgfpca, auc_adglmm, comp_t_fgfpca, comp_t_adglmm)
+colnames(tb_pred_eval) <- 
+  c("Maximum observed time", rep(c("fGFPCA", "GLMMadaptive"), 3))
 ```
 
-### In-/Out-of-sample PC scores
+```{r, results='asis'}
+tb_pred_eval %>% kable(digits = 3, align = "lcccccc") %>%
+  kable_styling(full_width = F) %>%
+  add_header_above(c(" "=1, "ISE"=2, "AUC "=2, "Prediction time"=2))
+```
+
+## In-/Out-of-sample PC scores
 
@@
 out_score_MLE <- matrix(NA, nrow=N, ncol=4)
-out_score_laplace <- matrix(NA, nrow=N, ncol=4)
+# out_score_laplace <- matrix(NA, nrow=N, ncol=4)
 
@@
                                   fpca_fit=fpca_fit)$score_out
-  out_score_laplace[i,] <- out_pred_laplace(fpca_fit=fpca_fit, 
-                                            df_new = df %>% filter(id==i & sind_inx<=595) %>% select(-eta_i))$score_out
+  # out_score_laplace[i,] <- out_pred_laplace(fpca_fit=fpca_fit, 
+  #                                           df_new = df %>% filter(id==i & sind_inx<=595) %>% select(-eta_i))$score_out
 }
@@
 out_bias_MLE <- out_score_MLE-xi
-out_bias_laplace <- out_score_laplace-xi
+# out_bias_laplace <- out_score_laplace-xi
 ```
 
-```{r, fig.height=4, fig.width=8}
+```{r, fig.height=4, fig.width=6}
 grid.arrange(
@@
 
-  out_bias_laplace %>% data.frame() %>%
-    pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
-    ggplot()+
-    geom_boxplot(aes(x=PC, y=Bias))+
-    stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
-    labs(title = "Out-of-sample laplace"),
+  # out_bias_laplace %>% data.frame() %>%
+  #   pivot_longer(1:4, names_to = "PC", values_to = "Bias") %>%
+  #   ggplot()+
+  #   geom_boxplot(aes(x=PC, y=Bias))+
+  #   stat_summary(aes(x=PC, y=Bias), fun = mean, geom = "point", col = "red")+
+  #   labs(title = "Out-of-sample laplace"),
 

### 2023-02-24 Code/PredEval.R

@@ -1,136 +1,101 @@
 # This script codes the computaiton of evaluation of predictive performance
 
 library(ROCR)
-
-##### Interpolated ISE and AUC #####
-
-# true eta and true y
-eta_true_mat <- df %>%
-  select(id, sind_inx, eta_i) %>%
-  pivot_wider(names_from = id, values_from = eta_i) %>%
-  column_to_rownames("sind_inx")
-
-# View(eta_true_mat)
-
-y_mat <- df %>%
-  select(id, sind_inx, Y) %>%
-  pivot_wider(names_from = id, values_from = Y) %>%
-  column_to_rownames("sind_inx")
-
-# View(y_mat)
-
-##### Observations up to 195 #####
-pred_t <- 196:max(mid)
-eta_hat_mat <- matrix(NA, nrow=length(pred_t), ncol = N) # prediction container
-dim(eta_hat_mat)
-for(i in 1:N){
-  pred <- out_samp_dyn_pred(df_new = df %>% 
-                              filter(id==i & sind_inx<=195) %>%
-                              select(-eta_i),
-                            fpca_fit = fpca_fit)$eta_pred
-  interp_pred <- approx(x=mid, y=pred, xout = pred_t)$y
-  eta_hat_mat[, i] <- interp_pred
-}
-
-# ISE
-ise_t195 <- mean(apply((eta_hat_mat-eta_true_mat[pred_t, ])^2, 2, sum))
-# AUC
-prob_mat <- exp(eta_hat_mat)/(1+exp(eta_hat_mat))
-auc_perf <- performance(prediction(prob_mat, y_mat[pred_t, ]), "auc")
-mean(unlist(auc_perf@y.values))
-
-
-
-##### Reference methods #####
-
-# binary FPCA for dynamic prediction
-# non-convergence
-library(registr)
-
-bfpca_fit <- bfpca(Y= df %>% select(id, Y, sind_inx) %>% rename(index=sind_inx, value = Y), 
-      npc_varExplained = 0.95, t_min=1, t_max = J)
-
-# A standard mxied model
-# time as the only fixed effect
-# random intercept for ID and random slope for time
-# slow, also with numeric issues
-glmm_fit <- glmer(Y ~ sind_inx + (id | sind_inx),
-                  data = df %>% select(id, Y, sind_inx), family = binomial)
-
-
-# Adaptive GLMM
-
 library(GLMMadaptive)
 
[…]

### 2023-02-24 f-GFPCA.Rmd
some formatting

@@
 
-This is a bit more complicated. The full compuation time consists of two parts: 
+This is a bit more complicated. The full computation time consists of two parts:
+
 - Time spent on model fitting
-- Time spent on prediction **given** the fitted FPCA or adaptive GLMM models.
+- Time spent on prediction **given** the fitted FPCA or adaptive GLMM models
+
+Table below include time spent only for prediction.
 

### 2023-02-28 f-GFPCA.Rmd
prediction interval

@@
   left_join(df_true_eta %>% filter(id==4), by = "mid")
+
+# in_pred_subj1 %>% filter(mid >595)
+# in_pred_subj2 %>% filter(mid >595)
 ```
@@
 
+## Prediction interval
+
+With an estimate of scores $\hat{\boldsymbol{\xi}}$
+
+$$\boldsymbol{\hat{\eta}}=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}$$
+$$\begin{aligned}
+\hat{\boldsymbol{\eta}}-\boldsymbol{\eta}&=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}-\boldsymbol{\eta}\\
+&= \boldsymbol{\hat{f_0}-f_0}+\boldsymbol{\Phi(\hat{\xi}-\xi)}-\boldsymbol{\epsilon}
+\end{aligned}$$
+
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi Var(\hat{\xi}-\xi) \Phi^T}+Var(\epsilon)$$
+
+- The three components should be mutually independent? 
+- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
+- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?
+- $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
+- May need to circle back to in-sample estimation and dig into where that formula came from
+
+Below are some potential methods:
+
+1. Estimate the variance with Empirical Bayes estimators of Gaussian process
+
+- Essentially the same formula as in-sample prediction interval
+
+$$\boldsymbol{H} = \boldsymbol{\hat{\Gamma}\Phi ^T_m}(\boldsymbol{\Phi_m\hat{\Gamma}\Phi^T_m}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}$$
+
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=\boldsymbol{\Phi}(\hat{\boldsymbol{\Gamma}}-\boldsymbol{H}\boldsymbol{\Phi_m}\hat{\boldsymbol{\Gamma}})\boldsymbol{\Phi}^T+\hat{\sigma_{\epsilon}}^2$$
+
+- This estimate does not change across subject? 
+- Also does not change from in- to out-of-sample? 
+- Ignore randomness from FPCA estimates (that is, $\boldsymbol{f}_0$, $\boldsymbol{\Gamma}$ and $\sigma_{\epsilon}$).
+
+2. Fisher information matrix
+
+- Based on posterior likelihood
+- Standard deviation from this method ended up being very big
+- Also does not change from in-sample to out-of-sample, nor across subjects
+
+3. Bootstrap
+
+```{r, warning=FALSE}
+source(here("Code/OutSampPredInterv.R"))
+
+# EB
+pred_std_eb <- data.frame(mid,
+                          t195 = pred_std_eb(fpca_fit, 195),
+                          t395 = pred_std_eb(fpca_fit, 395),
+                          t595 = pred_std_eb(fpca_fit, 595)) %>%
+  pivot_longer(2:4, values_to = "std_eb", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("t195", "t395", "t595"),
+                       labels=c("Up to 195", "Up to 395", "Up to 595")))
+
+```
+
+```{r}
+pred_std_mle <- data.frame(mid)
+
+for(t in c(195, 395, 595)){
+  mle_xi <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=t) %>%
+                           select(-eta_i),fpca_fit = fpca_fit)$score_out
+  I_mat <- -gradient(f = llh_div, x = mle_xi, 
+                     df_new = df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i),
+                   fpca_fit=fpca_fit)
+  std <- sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions))+fpca_fit$sigma2)
+
+  pred_std_mle <- data.frame(pred_std_mle, std)
+}
+pred_std_mle <- pred_std_mle %>%  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("std", "std.1", "std.2"),
+                       labels=c("Up to 195", "Up to 395", "Up to 595")))
+```
+
+```{r}
+out_pred_subj1 %>%
+  left_join(pred_std_eb, by = c("mid", "type")) %>% 
+  mutate(eta_upper_eb = eta+std_eb,
+         eta_lower_eb = eta-std_eb) %>%
+  left_join(pred_std_mle, by = c("mid", "type")) %>% 
+  mutate(eta_upper_mle = eta+std_mle,
+         eta_lower_mle = eta-std_mle) %>%
+  ggplot()+
+    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col="Estimated"))+
+    geom_line(aes(x=mid, y=eta_upper_eb, col="EB"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_eb, col="EB"), linetype="dashed")+
+  geom_line(aes(x=mid, y=eta_upper_mle, col="MLE"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_mle, col="MLE"), linetype="dashed")+
+    facet_wrap(~type)
+```
+
 ## In-/Out-of-sample PC scores
@@
 
-### Prediction interval
-
-<span style="color:red;"> Questions 3: a lot of problems on this part </span>
-
-With an estimate of scores $\hat{\boldsymbol{\xi}}$
-
-$$\boldsymbol{\hat{\eta}}=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}$$
-$$\begin{aligned}
-\hat{\boldsymbol{\eta}}-\boldsymbol{\eta}&=\boldsymbol{\hat{f_0}}+\boldsymbol{\Phi\hat{\xi}}-\boldsymbol{\eta}\\
-&= \boldsymbol{\hat{f_0}-f_0}+\boldsymbol{\Phi(\hat{\xi}-\xi)}-\boldsymbol{\epsilon}
-\end{aligned}$$
-
-$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=Var(\boldsymbol{\hat{f_0}-f_0})+\boldsymbol{\Phi Var(\hat{\xi}-\xi) \Phi^T}+Var(\epsilon)$$
-
-- The three components should be mutually independent? 
-- Through some derivation, I think $Var(\boldsymbol{f_0})=\frac{Var(\boldsymbol{Y})}{N} = \frac{1}{N}[\boldsymbol{\Phi\Gamma \Phi^T}+\sigma_{\epsilon}^2\boldsymbol{I}_m]$.
-- Variance of $\hat{\boldsymbol{\xi}}$ will depend on the method used for its estimation. But isn't the estimated done conditioning on $Var(\boldsymbol{\xi}) = \boldsymbol{\Gamma}$?
-- $Var(\epsilon) =\hat{\sigma}_{\epsilon}^2$ from FPCA
-- May need to circle back to in-sample estimation and dig into where that formula came from
-
-Some new updates:
-
-Actually, after going over the full derivation in Leroux2017 paper, I think the preidction interval completely ignored the randomness from fixed effects.
-In other worlds, prediction intervals are made conditioning on fixed effect coefficients (in our case $\boldsymbol{f_0}$).
-
 # Next steps:

### 2023-02-28 Code/OutSampMLE.R

@@ -52,20 +52,48 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
 
 #### Estimate standard error of MLE ####
 
-# # Jacabian matrix of first dirivative of llh
-# try <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
-#                     select(-eta_i),
-#                   fpca_fit = fpca_fit)$score_out
-# 
-# I_mat <- -gradient(f = llh_div, x = try,
-#          df_new = df %>% filter(id==1 & sind_inx<=195) %>%
-#            select(-eta_i),
-#          fpca_fit=fpca_fit)
+# Jacabian matrix of first dirivative of llh
+try <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+                    select(-eta_i),
+                  fpca_fit = fpca_fit)$score_out
+
+I_mat <- -gradient(f = llh_div, x = try,
+         df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+           select(-eta_i),
+         fpca_fit=fpca_fit)
+
+var <- diag(solve(I_mat))
+sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions)))
+
+# I_mat <- function(xi, df_new, fpca_fit){
+#   
+#   # observations
+#   ns <- as.vector(table(df_new$bin)) # number of observations
+#   hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+#     select(Y) %>% unlist()# number of success
+#   max_bin <- length(unique(df_new$bin)) # assume new skipped bins 
+#   
+#   # fpca objects
+#   phi_m <- fpca_fit$efunctions[1:max_bin, ]
+#   f0_m <- fpca_fit$mu[1:max_bin]
+#   eta_s <- f0_m + phi_m %*% xi
+#   tao <- diag(fpca_fit$evalues)
+#   
+#   # derivative
+#   val1 <- ns*(exp(eta_s)/(1+exp(eta_s))^2)
+#   div2 <- lapply(1:max_bin,  function(x)val1[x]*phi_m[x, ] %*% t(phi_m[x, ]))
+#   div2 <- array(unlist(div2), dim = c(K, K, max_bin))
[…]

### 2023-03-01 f-GFPCA.Rmd
have trouble knitting computation time

@@
 
-Table below include time spent only for prediction.
-
 ```{r, results='hide', message=FALSE}
@@
   c("195", "395", "595"),
-  ise_fgfpca, ise_adglmm, auc_fgfpca, auc_adglmm, comp_t_fgfpca, comp_t_adglmm)
+  ise_fgfpca, ise_adglmm, auc_fgfpca, auc_adglmm)
 colnames(tb_pred_eval) <- 
   c("Maximum observed time", rep(c("fGFPCA", "GLMMadaptive"), 3))
+
+# print(tb_pred_eval)
+# print(comp_t_fgfpca)
+# print(comp_t_adglmm)
 ```
@@
 ```{r, results='asis'}
-tb_pred_eval %>% kable(digits = 3, align = "lcccccc") %>%
+tb_pred_eval  %>% 
+  kable(format = "html", digits = 3, align = "lcccccc", 
+        table.attr = "style = \"color: black;\"") %>%
   kable_styling(full_width = F) %>%
-  add_header_above(c(" "=1, "ISE"=2, "AUC "=2, "Prediction time"=2))
+  add_header_above(c(" "=1, "ISE"=2, "AUC "=2))
 ```
@@
 - Based on posterior likelihood
-- Standard deviation from this method ended up being very big
-- Also does not change from in-sample to out-of-sample, nor across subjects
+- Also ignore randomness from FPCA estimates
 
-3. Bootstrap
+$$Var(\hat{\boldsymbol{\eta}}-\boldsymbol{\eta})=\boldsymbol{\Phi}I(\hat{\boldsymbol{\xi}})\boldsymbol{\Phi}^T+\hat{\sigma_{\epsilon}}^2\boldsymbol{I}_m $$
+- Standard deviation would change across subjects.
 
-```{r, warning=FALSE}
+```{r}
 source(here("Code/OutSampPredInterv.R"))
@@
 ```{r}
-pred_std_mle <- data.frame(mid)
+df_std_mle_subj1 <- data.frame(mid, 
+                               t195 = pred_std_mle(df %>% 
+                                                     filter(id==1 & sind_inx<=195) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t395 =  pred_std_mle(df %>% 
+                                                     filter(id==1 & sind_inx<=395) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t595 =  pred_std_mle(df %>% 
+                                                     filter(id==1 & sind_inx<=595) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit)) %>%
+  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("t195", "t395", "t595"),
+                       labels=c("Up to 195", "Up to 395", "Up to 595")))
 
-for(t in c(195, 395, 595)){
-  mle_xi <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=t) %>%
-                           select(-eta_i),fpca_fit = fpca_fit)$score_out
-  I_mat <- -gradient(f = llh_div, x = mle_xi, 
-                     df_new = df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i),
-                   fpca_fit=fpca_fit)
-  std <- sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions))+fpca_fit$sigma2)
+df_std_mle_subj2 <- data.frame(mid, 
+                               t195 = pred_std_mle(df %>% 
+                                                     filter(id==2 & sind_inx<=195) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t395 =  pred_std_mle(df %>% 
+                                                     filter(id==2 & sind_inx<=395) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t595 =  pred_std_mle(df %>% 
+                                                     filter(id==2 & sind_inx<=595) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit))%>%
+  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("t195", "t395", "t595"),
+                       labels=c("Up to 195", "Up to 395", "Up to 595")))
 
-  pred_std_mle <- data.frame(pred_std_mle, std)
-}
-pred_std_mle <- pred_std_mle %>%  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
-  mutate(type = factor(type, levels = c("std", "std.1", "std.2"),
+df_std_mle_subj3 <- data.frame(mid, 
+                               t195 = pred_std_mle(df %>% 
+                                                     filter(id==3 & sind_inx<=195) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t395 =  pred_std_mle(df %>% 
+                                                     filter(id==3 & sind_inx<=395) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t595 =  pred_std_mle(df %>% 
+                                                     filter(id==3 & sind_inx<=595) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit))%>%
+  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("t195", "t395", "t595"),
                        labels=c("Up to 195", "Up to 395", "Up to 595")))
+
+df_std_mle_subj4 <- data.frame(mid, 
+                               t195 = pred_std_mle(df %>% 
+                                                     filter(id==4 & sind_inx<=195) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t395 =  pred_std_mle(df %>% 
+                                                     filter(id==4 & sind_inx<=395) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit),
+                               t595 =  pred_std_mle(df %>% 
+                                                     filter(id==4 & sind_inx<=595) %>%
+                                                     select(-eta_i),
+                                                   fpca_fit))%>%
+  pivot_longer(2:4, values_to = "std_mle", names_to = "type") %>%
+  mutate(type = factor(type, levels = c("t195", "t395", "t595"),
+                       labels=c("Up to 195", "Up to 395", "Up to 595")))
+
 ```
 
-```{r}
+```{r, warning=FALSE,fig.height=3, fig.width=10}
 out_pred_subj1 %>%
@@
          eta_lower_eb = eta-std_eb) %>%
-  left_join(pred_std_mle, by = c("mid", "type")) %>% 
+  left_join(df_std_mle_subj1, by = c("mid", "type")) %>% 
   mutate(eta_upper_mle = eta+std_mle,
@@
   ggplot()+
-    geom_line(aes(x=mid, y=true), col="black")+
+    geom_line(aes(x=mid, y=eta, col="Estimated"))+
+    geom_line(aes(x=mid, y=eta_upper_eb, col="EB"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_eb, col="EB"), linetype="dashed")+
+  geom_line(aes(x=mid, y=eta_upper_mle, col="MLE"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_mle, col="MLE"), linetype="dashed")+
+    facet_wrap(~type)
+
+out_pred_subj2 %>%
+  left_join(pred_std_eb, by = c("mid", "type")) %>% 
+  mutate(eta_upper_eb = eta+std_eb,
+         eta_lower_eb = eta-std_eb) %>%
+  left_join(df_std_mle_subj2, by = c("mid", "type")) %>% 
+  mutate(eta_upper_mle = eta+std_mle,
+         eta_lower_mle = eta-std_mle) %>%
+  ggplot()+
+    geom_line(aes(x=mid, y=eta, col="Estimated"))+
+    geom_line(aes(x=mid, y=eta_upper_eb, col="EB"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_eb, col="EB"), linetype="dashed")+
+  geom_line(aes(x=mid, y=eta_upper_mle, col="MLE"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_mle, col="MLE"), linetype="dashed")+
+    facet_wrap(~type)
+
+out_pred_subj3 %>%
+  left_join(pred_std_eb, by = c("mid", "type")) %>% 
+  mutate(eta_upper_eb = eta+std_eb,
+         eta_lower_eb = eta-std_eb) %>%
+  left_join(df_std_mle_subj3, by = c("mid", "type")) %>% 
+  mutate(eta_upper_mle = eta+std_mle,
+         eta_lower_mle = eta-std_mle) %>%
+  ggplot()+
+    geom_line(aes(x=mid, y=eta, col="Estimated"))+
+    geom_line(aes(x=mid, y=eta_upper_eb, col="EB"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_eb, col="EB"), linetype="dashed")+
+  geom_line(aes(x=mid, y=eta_upper_mle, col="MLE"), linetype="dashed")+
+    geom_line(aes(x=mid, y=eta_lower_mle, col="MLE"), linetype="dashed")+
+    facet_wrap(~type)
+
+out_pred_subj4 %>%
+  left_join(pred_std_eb, by = c("mid", "type")) %>% 
+  mutate(eta_upper_eb = eta+std_eb,
+         eta_lower_eb = eta-std_eb) %>%
+  left_join(df_std_mle_subj4, by = c("mid", "type")) %>% 
+  mutate(eta_upper_mle = eta+std_mle,
+         eta_lower_mle = eta-std_mle) %>%
+  ggplot()+
     geom_line(aes(x=mid, y=eta, col="Estimated"))+

### 2023-03-01 Code/OutSampPredInterv.R

@@ -25,3 +25,14 @@ pred_std_eb <- function(fpca_fit, tm){
 #### MLE ####
 
 
+pred_std_mle <- function(df, fpca_fit){
+  mle_xi <- out_samp_dyn_pred(df_new = df ,fpca_fit = fpca_fit)$score_out
+  I_mat <- -gradient(f = llh_div, x = mle_xi, df_new = df, fpca_fit=fpca_fit)
+  std <- sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions))+fpca_fit$sigma2)
+  
+  return(std)
+}
+
+# pred_std_mle(df %>% filter(id==1 & sind_inx<=395) %>% select(-eta_i), fpca_fit)
+
+

### 2023-03-07 Write-ups/Manuscript.Rmd
start manuscript

---
title: "Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis"
author:
  - Ying Jin
  - Andrew Leroux
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin = 1in
bibliography: refs.bib
nocite: '@*'
link-citations: yes
linkcolor: blue
csl: ASA.cls

---
Abstract: 
=================================================================================================================================
  Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction. Traditional methods are often limited in flexibility and computationally intensive, especially with non-Gaussian data. To address these issues, we propose a novel method for dynamic prediction based on Generalized Functional Principal Component Analysis (FPCA). Assume the observed outcome follows an exponential family distribution parameterized by a latent Gaussian function, the proposed method consists of the following steps: 1) Bin the data across functional domain into small, equal-length intervals; 2) Fit local generalized mixed models at every bin to estimate individual latent functions; 3) Fit FPCA model to smooth latent functions and 4) Obtain estimates of subject-specific PC scores using partial observations and recover the unobserved part on the binned grid. Our simulation study showed the proposed method achieved significantly better out-of-sample predictive performance compared to existing methods with much shorter computation time, thus has the potential to be widely applicable to large datasets.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Introduction
=================================================================================================================================

-	Overview of dynamic prediction methods

Prediction of repeated measures has been a problem of interest in the biomedical field. Typically, such predictions are made based on the correlation between repeated measures from the same subject, and/or covariates that can be either fixed or time-varying. Traditionally, repeated measures have been modeled using marginal models (generalized estimating equations) or conditional models (mixed effect models) [@liang1986; @Laird1982; @lindstrom1990; @davidian2003; @GLMMadaptive]. These methods, while allowing for correlation between repeated measures, are limited in terms of flexibility of correlation structure and the ability to handle out-of-sample prediction. Therefore, one may turn to functional mixed effect models when measures are dense across the domain. Such methods accommodates more flexible correlation structure by modeling subject-specific random effects as a function, but often cause nontrivial computational burden. A feasible approach to address this issue is non-parametric smoothing [@Scheipl2014], such as spline basis functions or eigenfunctions from functional principal component analysis (fPCA). The introduction of basis functions also makes out-of-sample prediction more straightforward. Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing. In this project, we will focus on the prediction of non-Gaussian outcomes (e.g. binary or count) from a random-intercept only model, with no covariates considered. In other words, we aim to propose a new fast, scalable method for dynamic prediction of discrete function tracks based only on past observations using functional mixed effect model with fPCA smoothing.

-	Dynamic prediction with functional methods

Research on dynamic prediction of functional outcomes has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors. However, little extension was made on prediction of non-Gaussian functions, such as binary and count outcomes.

-	fPCA and GFPCA [@leroux2022]

Unlike FPCA on Gaussian data, fewer papers have focused on its extension to non-Gaussian data, such as series of binary or count outcomes. Existing methods also tend to be computationally intensive. For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes. @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*. @gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage, joint estimation strategy. @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data. In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit GFPCA on binary data using EM algorithm, accompanied by the an open source R package *registr*.

Method
=================================================================================================================================
-	Need better notation system

Result
=================================================================================================================================
-	Repeat simulation: is it necessary in this case?
-	Different set-up:
a.	Different eigenfunctions: with or without periodicity
b.	Outcome: binary or count
c.	Sample size
d.	Grid density
e.	Bin width, overlap or not
-	Real data application

Discussion
=================================================================================================================================
-	Grid 
-	Score bias: cannot demonstrate without repeat simulation

References
=================================================================================================================================
<div id="refs"></div>

### 2023-03-07 Write-ups/Manuscript.tex
start manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 07, 2023}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract:}\label{abstract}}

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, referred to as
dynamic prediction. Traditional methods are often limited in flexibility
and computationally intensive, especially with non-Gaussian data. To
address these issues, we propose a novel method for dynamic prediction
based on Generalized Functional Principal Component Analysis (FPCA).
Assume the observed outcome follows an exponential family distribution
parameterized by a latent Gaussian function, the proposed method
consists of the following steps: 1) Bin the data across functional
domain into small, equal-length intervals; 2) Fit local generalized
mixed models at every bin to estimate individual latent functions; 3)
Fit FPCA model to smooth latent functions and 4) Obtain estimates of
subject-specific PC scores using partial observations and recover the
unobserved part on the binned grid. Our simulation study showed the
proposed method achieved significantly better out-of-sample predictive
performance compared to existing methods with much shorter computation
time, thus has the potential to be widely applicable to large datasets.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{itemize}
\tightlist
\item
  Overview of dynamic prediction methods
\end{itemize}

Prediction of repeated measures has been a problem of interest in the
biomedical field. Typically, such predictions are made based on the
correlation between repeated measures from the same subject, and/or
covariates that can be either fixed or time-varying. Traditionally,
repeated measures have been modeled using marginal models (generalized
estimating equations) or conditional models (mixed effect models)
(\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003};
\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). These methods,
while allowing for correlation between repeated measures, are limited in
terms of flexibility of correlation structure and the ability to handle
out-of-sample prediction. Therefore, one may turn to functional mixed
effect models when measures are dense across the domain. Such methods
accommodates more flexible correlation structure by modeling
subject-specific random effects as a function, but often cause
nontrivial computational burden. A feasible approach to address this
issue is non-parametric smoothing
(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}), such as
spline basis functions or eigenfunctions from functional principal
component analysis (fPCA). The introduction of basis functions also
makes out-of-sample prediction more straightforward. Instead of
estimating subject-specific random effects of new observations, we can
simply estimate coefficients/loadings on the basis function used for
smoothing. In this project, we will focus on the prediction of
non-Gaussian outcomes (e.g.~binary or count) from a random-intercept
only model, with no covariates considered. In other words, we aim to
propose a new fast, scalable method for dynamic prediction of discrete
function tracks based only on past observations using functional mixed
effect model with fPCA smoothing.

\begin{itemize}
\tightlist
\item
  Dynamic prediction with functional methods
\end{itemize}

Research on dynamic prediction of functional outcomes has been focusing
on continuous/Gaussian outcomes, modelling subject-specific random
effects with FPCA (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors. However, little extension was
made on prediction of non-Gaussian functions, such as binary and count
outcomes.

\begin{itemize}
\tightlist
\item
  fPCA and GFPCA (\protect\hyperlink{ref-leroux2022}{Leroux et al.
  n.d.})
\end{itemize}

Unlike FPCA on Gaussian data, fewer papers have focused on its extension
to non-Gaussian data, such as series of binary or count outcomes.
Existing methods also tend to be computationally intensive. For example,
Chen et al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches
to fit marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage, joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

\hypertarget{method}{%
\section{Method}\label{method}}

\begin{itemize}
\tightlist
\item
  Need better notation system
\end{itemize}

\hypertarget{result}{%
\section{Result}\label{result}}

\begin{itemize}
\tightlist
\item
  Repeat simulation: is it necessary in this case?
\item
  Different set-up:
\end{itemize}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\tightlist
\item
  Different eigenfunctions: with or without periodicity
\item
  Outcome: binary or count
\item
  Sample size
\item
  Grid density
\item
  Bin width, overlap or not
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Real data application
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\begin{itemize}
\tightlist
\item
  Grid
\item
  Score bias: cannot demonstrate without repeat simulation
\end{itemize}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-chen2013}{}}%
Chen, H., Wang, Y., Paik, M. cho, and Choi, H. A. (2013), {``A marginal
approach to reduced-rank penalized spline smoothing with application to
multilevel functional data,''} \emph{J Am Stat Assoc.}, 108, 1216--1229.
\url{https://doi.org/10.1080/01621459.2013.826134}.

\leavevmode\vadjust pre{\hypertarget{ref-chiou2012}{}}%
Chiou, J.-M. (2012), {``Dynamical functional prediction and
classification, with application to traffic flow prediction,''}
\emph{The Annals of Applied Statistics}, Institute of Mathematical
Statistics, 6, 1588--1614. \url{https://doi.org/10.1214/12-AOAS595}.

\leavevmode\vadjust pre{\hypertarget{ref-delaigo2016}{}}%
Delaigle, A., and Hall, P. (2016), {``Approximating fragmented
functional data by segments of markov chains,''} \emph{Biometrika}, 103,
779--799. \url{https://doi.org/10.1093/biomet/asw040}.

\leavevmode\vadjust pre{\hypertarget{ref-gertheiss2016}{}}%
Gertheiss, J., Goldsmith, J., and Staicu, A. (2016), {``A note on
modeling sparse exponential-family functional response curves,''}
\emph{Comput Stat Data Anal}, 105, 46--52.
\url{https://doi.org/10.1016/j.csda.2016.07.010}.

\leavevmode\vadjust pre{\hypertarget{ref-goldberg2014}{}}%
Goldberg, Y., Ritov, Y., and Mandelbaum, A. (2014), {``Predicting the
continuation of a function with applications to call center data,''}
\emph{Journal of Statistical Planning and Inference}, 147, 53--65.
https://doi.org/\url{https://doi.org/10.1016/j.jspi.2013.11.006}.

\leavevmode\vadjust pre{\hypertarget{ref-goldsmith2015}{}}%
Goldsmith, J., Zipunnikov, V., and Schrack, J. (2015), {``Generalized
multilevel function-on-scalar regression and principal component
analysis,''} \emph{Biometrics}, 71, 344--53.
\url{https://doi.org/10.1111/biom.12278}.

\leavevmode\vadjust pre{\hypertarget{ref-hall2018}{}}%
Hall, P., Müller, H.-G., and Yao, F. (2008), {``Modelling sparse
generalized longitudinal observations with latent gaussian processes,''}
\emph{Journal of the Royal Statistical Society: Series B (Statistical
Methodology)}, 70, 703--723.
https://doi.org/\url{https://doi.org/10.1111/j.1467-9868.2008.00656.x}.

\leavevmode\vadjust pre{\hypertarget{ref-kraus2015}{}}%
Kraus, D. (2015),
{``\href{http://www.jstor.org/stable/24775309}{Components and completion
of partially observed functional data},''} \emph{Journal of the Royal
Statistical Society. Series B (Statistical Methodology)}, {[}Royal
Statistical Society, Wiley{]}, 77, 777--801.

\leavevmode\vadjust pre{\hypertarget{ref-Laird1982}{}}%
Laird, N. M., and Ware, J. H. (1982),
{``\href{http://www.jstor.org/stable/2529876}{Random-effects models for
longitudinal data},''} \emph{Biometrics}, {[}Wiley, International
Biometric Society{]}, 38, 963--974.

\leavevmode\vadjust pre{\hypertarget{ref-leroux2022}{}}%
Leroux, A., Crainiceanu, C. M., and Wrobel, J. (n.d.). {``Fast
generalized functional principal component analysis.''}

\leavevmode\vadjust pre{\hypertarget{ref-leroux2016}{}}%
Leroux, A., Xiao, L., Crainiceanu, C., and Checkley, W. (2018),
{``Dynamic prediction in functional concurrent regression with an
application to child growth,''} \emph{Statistics in medicine}, 37,
1376--1388.

\leavevmode\vadjust pre{\hypertarget{ref-liang1986}{}}%
LIANG, K.-Y., and ZEGER, S. L. (1986), {``Longitudinal data analysis
using generalized linear models,''} \emph{Biometrika}, 73, 13--22.
\url{https://doi.org/10.1093/biomet/73.1.13}.

\leavevmode\vadjust pre{\hypertarget{ref-linde2019}{}}%
Linde, van der (2009), {``A bayesian latent variable approach to
functional principal components analysi with binary and count data,''}
\emph{A StA Adv Stat Anal}, 307--333.
\url{https://doi.org/10.1007/s10182-009-0113-6}.

\leavevmode\vadjust pre{\hypertarget{ref-lindstrom1990}{}}%
Lindstrom, M. J., and Bates, D. M. (1990),
{``\href{http://www.jstor.org/stable/2532087}{Nonlinear mixed effects
models for repeated measures data},''} \emph{Biometrics}, {[}Wiley,
International Biometric Society{]}, 46, 673--687.

\leavevmode\vadjust pre{\hypertarget{ref-davidian2003}{}}%
{``\href{http://www.jstor.org/stable/1400665}{Nonlinear models for
repeated measurement data: An overview and update}''} (2003),
{[}International Biometric Society, Springer{]}, 8, 387--419.

\leavevmode\vadjust pre{\hypertarget{ref-GLMMadaptive}{}}%
Rizopoulos, D. (2022),
\emph{\href{https://CRAN.R-project.org/package=GLMMadaptive}{GLMMadaptive:
Generalized linear mixed models using adaptive gaussian quadrature}}.

\leavevmode\vadjust pre{\hypertarget{ref-Scheipl2014}{}}%
Scheipl, F., Staicu, A.-M., and Greven, S. (2014), {``Functional
additive mixed models,''} \emph{J Comput Graph Stat}, 24, 447--501.
\url{https://doi.org/10.1080/10618600.2014.901914}.

\leavevmode\vadjust pre{\hypertarget{ref-shang2017}{}}%
Shang, H. L. (2017), {``Functional time series forecasting with dynamic
updating: An application to intraday particulate matter
concentration,''} \emph{Econometrics and Statistics}, 1, 184--200.
https://doi.org/\url{https://doi.org/10.1016/j.ecosta.2016.08.004}.

\leavevmode\vadjust pre{\hypertarget{ref-suresh2017}{}}%
Suresh, K., Taylor, J. M. G., Spratt, D. E., Daignault, S., and
Tsodikov, A. (2017), {``Comparison of joint modeling and landmarking for
dynamic prediction under an illness-death model,''} \emph{Biom J}, 59,
1277--1300. \url{https://doi.org/10.1002/bimj.201600235}.

\leavevmode\vadjust pre{\hypertarget{ref-wrobel2019}{}}%
Wrobel, J., Zipunnikov, V., Schrack, J., and Goldsmith, J. (2019),
{``Registration for exponential family functional data,''}
\emph{Biometrics}, 75, 48--57.
https://doi.org/\url{https://doi.org/10.1111/biom.12963}.

\end{CSLReferences}

\end{document}

### 2023-03-07 f-GFPCA.Rmd
start manuscript

@@
 
-```{r, fig.width=8, fig.height=4}
+```{r, fig.width=6, fig.height=3}
 # compare point prediction with in-sample

### 2023-03-07 Write-ups/Manuscript.Rmd
manuscript outline

@@
 =================================================================================================================================
+
 -	Need better notation system
+- Change to general exponential family
+- Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate
 
@@
 =================================================================================================================================
--	Repeat simulation: is it necessary in this case?
+-	Repeat simulation: repeat 10-100 times
+
 -	Different set-up:
-a.	Different eigenfunctions: with or without periodicity
+a.	Different eigenfunctions: with or without periodicity (start with the current one)
 b.	Outcome: binary or count
-c.	Sample size
-d.	Grid density
-e.	Bin width, overlap or not
+c.	Sample size: start with N=500
+d.	Grid density: start with J=1000
+e.	Bin width, overlap or not: start with non-overlap, bin width = 10 (100 bins)
+
 -	Real data application

### 2023-03-07 Write-ups/Manuscript.tex
manuscript outline

@@
   Need better notation system
+\item
+  Change to general exponential family
+\item
+  Add section of de-bias score: conditional on subject. Need repeat
+  simulation to demonstrate
 \end{itemize}
@@
 \begin{itemize}
-\tightlist
 \item
-  Repeat simulation: is it necessary in this case?
+  Repeat simulation: repeat 10-100 times
 \item
@@
 \item
-  Different eigenfunctions: with or without periodicity
+  Different eigenfunctions: with or without periodicity (start with the
+  current one)
 \item
@@
 \item
-  Sample size
+  Sample size: start with N=500
 \item
-  Grid density
+  Grid density: start with J=1000
 \item
-  Bin width, overlap or not
+  Bin width, overlap or not: start with non-overlap, bin width = 10 (100
+  bins)
 \end{enumerate}

### 2023-03-14 Write-ups/Manuscript.Rmd
simulation part 1

@@
 csl: ASA.cls
+header-includes:
+  - \usepackage{setspace}\doublespacing
 
 ---
-Abstract: 
-=================================================================================================================================
-  Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction.
+<!-- Abstract:  -->
+<!-- ================================================================================================================================= -->
+<!--   Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction.
 Traditional methods are often limited in flexibility and computationally intensive, especially with non-Gaussian data.
@@
 Our simulation study showed the proposed method achieved significantly better out-of-sample predictive performance compared to existing methods with much shorter computation time, thus has the potential to be widely applicable to large datasets.
+-->
 
@@
 
--	Overview of dynamic prediction methods
+<!-- -	Overview of dynamic prediction methods -->
 
-Prediction of repeated measures has been a problem of interest in the biomedical field.
-Typically, such predictions are made based on the correlation between repeated measures from the same subject, and/or covariates that can be either fixed or time-varying.
-Traditionally, repeated measures have been modeled using marginal models (generalized estimating equations) or conditional models (mixed effect models) [@liang1986; @Laird1982; @lindstrom1990; @davidian2003; @GLMMadaptive].
-These methods, while allowing for correlation between repeated measures, are limited in terms of flexibility of correlation structure and the ability to handle out-of-sample prediction.
-Therefore, one may turn to functional mixed effect models when measures are dense across the domain.
-Such methods accommodates more flexible correlation structure by modeling subject-specific random effects as a function, but often cause nontrivial computational burden.
-A feasible approach to address this issue is non-parametric smoothing [@Scheipl2014], such as spline basis functions or eigenfunctions from functional principal component analysis (fPCA).
+Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction.
+Traditionally, this type of data has been modeled using marginal models (generalized estimating equations) or conditional models (mixed effect models) [@liang1986; @Laird1982; @lindstrom1990; @davidian2003], and predictions are made based on the correlation between repeated measures from the same subject, and/or covariates that can be either fixed or time-varying.
+However, these methods are limited in terms of flexibility of correlation structure and the ability to handle out-of-sample prediction.
+When sample size is large or the density of repeated measures is high, they also tend to cause severe computational burden [@GLMMadaptive].
+
+To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain.
+Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (fPCA).
 The introduction of basis functions also makes out-of-sample prediction more straightforward.
 Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing.
-In this project, we will focus on the prediction of non-Gaussian outcomes (e.g.
-binary or count) from a random-intercept only model, with no covariates considered.
-In other words, we aim to propose a new fast, scalable method for dynamic prediction of discrete function tracks based only on past observations using functional mixed effect model with fPCA smoothing.
-
--	Dynamic prediction with functional methods
-
-Research on dynamic prediction of functional outcomes has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
+Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
 @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains.
@@
 
--	fPCA and GFPCA [@leroux2022]
-
-Unlike FPCA on Gaussian data, fewer papers have focused on its extension to non-Gaussian data, such as series of binary or count outcomes.
-Existing methods also tend to be computationally intensive.
+Unfortunately, fewer papers have focused on its extension to non-Gaussian data, such as series of binary or count outcomes.
+Existing methods also tend to be very computationally intensive.
 For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes.
@@
 
+In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with fPCA smoothing.
+Section 2 presents the procedure of the proposed method.
+In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study.
+In Section 4, we apply this method to a real-world dataset.
+Section 5 presents a discussion of davantages and limitation of the proposed method.
+
 Method
@@
 
--	Need better notation system
-- Change to general exponential family
-- Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate
+The observed data for a single subject $i$ is ($t$, $Y_i(t)$), where $t$ consists of dense, discrete points along the functional domain, and $Y_i(t)$ is the non-Gaussian outcome observed at $t$.
+We assume that the outcome $Y_i(t)$ can be characterized by a latent continuous function $\eta_i(t)$.
+That is, at a specific t, $Y_i(t)$ follows a exponential family distribution such that:
 
-Result
+$$
+g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)
+$$
+
+where g is a appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function follows a zero-mean Gaussian process.
+
+While $b_i(t)$ is not observed, it can be approximated under the FPCA framework $b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)$.
+Here $\phi_{ik}(t)$ is a set of orthogonal eigenfunctions that explains the most variation in $b_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction.
+Additionally, $\xi_{ik}$ are mutually independent obver both subject ($i$) and eigenfunctions ($k$).
+That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues: $\int \phi_k^2(t)dt=\lambda_k$.
+$\epsilon_i(t)$ here is a residual function that accounts for the variation not explained by the first K eigenfunctions from FPCA.
+We assume it follows a zero-mean Gaussian process.
+At a specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+
+Based on the problem set up above, we propose the following algorithm for PFCA on the unobserved latent process $\eta_i(t)$:
+
+1. Bin the observed outcomes in to small, non-overlapping, equal length intervals. We hereafter index the bins by their midpoints $s$. 
+2.
+Fit a local Generalized Mixed Model at every bin.
+Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$.
+From this series of models we can get estiamtes of population mean $\hat{\beta}_0(s)$ and subject-level random effect $\hat{b}_i(s)$, thus estimate of the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$.
+3.
+Fit FPCA on the estimated latent functions $\hat{\eta}_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, population mean $\hat{\beta}_0(s)$ and residual variance $\hat{\sigma}^2$.
+4.
+With components extracted above, calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$ of new samples based on their partially observed data.
+Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$
+
+Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects.
+Since the bins are set up to be small in length, the binned grid would still be dense enough.
+However, in situations where we need predictions on the original, un-binned grid instead, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.
+
+Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework.
+In step 4 we calculated the MLE of $\hat{\xi}_{ik}$.
+Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$.
+Therefore, the variation of prediction interval is:
+
+$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
+Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.
+
+<!-- -	Need better notation system -->
+<!-- - Change to general exponential family -->
+<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->
+
+Simulation
 =================================================================================================================================
--	Repeat simulation: repeat 10-100 times
 
--	Different set-up:
-a.	Different eigenfunctions: with or without periodicity (start with the current one)
-b.	Outcome: binary or count
-c.	Sample size: start with N=500
-d.	Grid density: start with J=1000
-e.	Bin width, overlap or not: start with non-overlap, bin width = 10 (100 bins)
+In this section, we illustrate the predictive performance and computational efficiency of the proposed method through a simulation study.
+We simulated 50 datasets, each with 500 subjects.
+For every subject, we generate 1000 binary outcomes $Y_i(t) \in (0, 1)$ across functional domain $t \in [0, 1]$, where the distribution of outcome is characterized by a continuous latent function.
+The data generation mechanism can be expressed as follows:
 
--	Real data application
+$$\begin{aligned}
+Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
+\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
+\end{aligned}$$
+
+In this simulation, we set $f_0(t) = 0$.
+$\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \gamma_k)$.
+Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$.
+In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1]$ and are the same for all subjects.
+
+We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC).
+ISE assess the prediction accucay of latent continuous function.
+It is evaluated on the binned grid at midpoints of each unobserved bin.
+If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.
+
+The second metric, AUC, focuses on evaluation of prediction of the binary outcome.
+Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation.
+Just like ISE, the AUC report is also the average value across the whole sample.
+
+As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive).
+
+<!-- -	Repeat simulation: repeat 10-100 times -->
+
+<!-- -	Different set-up: -->
+<!-- a.	Different eigenfunctions: with or without periodicity (start with the current one) -->
+<!-- b.	Outcome: binary or count -->
+<!-- c.	Sample size: start with N=500 -->
+<!-- d.	Grid density: start with J=1000 -->
+<!-- e.	Bin width, overlap or not: start with non-overlap, bin width = 10 (100 bins) -->
+
+Data application
+=================================================================================================================================
 

### 2023-03-14 Write-ups/Manuscript.tex
simulation part 1

--- 2023-03-07 Write-ups/Manuscript.tex
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{setspace}\doublespacing
 \ifLuaTeX
@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 07, 2023}
+\date{March 13, 2023}
 
@@
 
-\hypertarget{abstract}{%
-\section{Abstract:}\label{abstract}}
-
-Biomedical investigators are often interested in predicting future
-observations of subjects based on their historical data, referred to as
-dynamic prediction. Traditional methods are often limited in flexibility
-and computationally intensive, especially with non-Gaussian data. To
-address these issues, we propose a novel method for dynamic prediction
-based on Generalized Functional Principal Component Analysis (FPCA).
-Assume the observed outcome follows an exponential family distribution
-parameterized by a latent Gaussian function, the proposed method
-consists of the following steps: 1) Bin the data across functional
-domain into small, equal-length intervals; 2) Fit local generalized
-mixed models at every bin to estimate individual latent functions; 3)
-Fit FPCA model to smooth latent functions and 4) Obtain estimates of
-subject-specific PC scores using partial observations and recover the
-unobserved part on the binned grid. Our simulation study showed the
-proposed method achieved significantly better out-of-sample predictive
-performance compared to existing methods with much shorter computation
-time, thus has the potential to be widely applicable to large datasets.
-
 \hypertarget{introduction}{%
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Overview of dynamic prediction methods
-\end{itemize}
-
-Prediction of repeated measures has been a problem of interest in the
-biomedical field. Typically, such predictions are made based on the
-correlation between repeated measures from the same subject, and/or
-covariates that can be either fixed or time-varying. Traditionally,
-repeated measures have been modeled using marginal models (generalized
-estimating equations) or conditional models (mixed effect models)
+Biomedical investigators are often interested in predicting future
+observations of subjects based on their historical data, typically
+referred to as dynamic prediction. Traditionally, this type of data has
+been modeled using marginal models (generalized estimating equations) or
+conditional models (mixed effect models)
 (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
@@
 \protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
-measurement data''} 2003};
-\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). These methods,
-while allowing for correlation between repeated measures, are limited in
-terms of flexibility of correlation structure and the ability to handle
-out-of-sample prediction. Therefore, one may turn to functional mixed
-effect models when measures are dense across the domain. Such methods
-accommodates more flexible correlation structure by modeling
-subject-specific random effects as a function, but often cause
-nontrivial computational burden. A feasible approach to address this
-issue is non-parametric smoothing
-(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}), such as
-spline basis functions or eigenfunctions from functional principal
-component analysis (fPCA). The introduction of basis functions also
-makes out-of-sample prediction more straightforward. Instead of
-estimating subject-specific random effects of new observations, we can
-simply estimate coefficients/loadings on the basis function used for
-smoothing. In this project, we will focus on the prediction of
-non-Gaussian outcomes (e.g.~binary or count) from a random-intercept
-only model, with no covariates considered. In other words, we aim to
-propose a new fast, scalable method for dynamic prediction of discrete
-function tracks based only on past observations using functional mixed
-effect model with fPCA smoothing.
+measurement data''} 2003}), and predictions are made based on the
+correlation between repeated measures from the same subject, and/or
+covariates that can be either fixed or time-varying. However, these
+methods are limited in terms of flexibility of correlation structure and
+the ability to handle out-of-sample prediction. When sample size is
+large or the density of repeated measures is high, they also tend to
+cause severe computational burden
+(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}).
 
-\begin{itemize}
-\tightlist
-\item
-  Dynamic prediction with functional methods
-\end{itemize}
-
-Research on dynamic prediction of functional outcomes has been focusing
-on continuous/Gaussian outcomes, modelling subject-specific random
-effects with FPCA (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
+To address these problems, one may turn to functional mixed effect
+models instead when measures are dense across the domain. Such methods
+accommodate more flexible correlation structure by modeling
+subject-specific random effects as a function, and non-parametric
+smoothing (\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) can
+be incorporated to speed up the computation, such as spline basis
+functions or eigenfunctions from functional principal component analysis
+(fPCA). The introduction of basis functions also makes out-of-sample
+prediction more straightforward. Instead of estimating subject-specific
+random effects of new observations, we can simply estimate
+coefficients/loadings on the basis function used for smoothing. Existing
+research on dynamic prediction with functional data analysis methods has
+been focusing on continuous/Gaussian outcomes, modelling
+subject-specific random effects with FPCA
+(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
 \protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
@@
 
-\begin{itemize}
-\tightlist
-\item
-  fPCA and GFPCA (\protect\hyperlink{ref-leroux2022}{Leroux et al.
-  n.d.})
-\end{itemize}
-
-Unlike FPCA on Gaussian data, fewer papers have focused on its extension
-to non-Gaussian data, such as series of binary or count outcomes.
-Existing methods also tend to be computationally intensive. For example,
+Unfortunately, fewer papers have focused on its extension to
+non-Gaussian data, such as series of binary or count outcomes. Existing
+methods also tend to be very computationally intensive. For example,
 Chen et al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches
@@
 
+In this paper, we aim to develop a fast, scalable method for dynamic
+prediction of discrete function tracks based on functional mixed effect
+model with fPCA smoothing. Section 2 presents the procedure of the
+proposed method. In Section 3, we illustrate the performance and
+efficiency of our proposed method in a simulation study. In Section 4,
+we apply this method to a real-world dataset. Section 5 presents a
+discussion of davantages and limitation of the proposed method.
+
 \hypertarget{method}{%
@@
 
-\begin{itemize}
+The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
+where \(t\) consists of dense, discrete points along the functional
+domain, and \(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We
+assume that the outcome \(Y_i(t)\) can be characterized by a latent
+continuous function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\)
+follows a exponential family distribution such that:
+
+\[
+g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)
+\]
+
+where g is a appropriate link function, \(\beta_0(t)\) is the population
+mean of latent function, and \(b_i(t)\) is a subjects-level random
+effect function follows a zero-mean Gaussian process.
+
+While \(b_i(t)\) is not observed, it can be approximated under the FPCA
+framework \(b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)\).
+Here \(\phi_{ik}(t)\) is a set of orthogonal eigenfunctions that
+explains the most variation in \(b_i(t)\), and \(\xi_{ik}\) are
+subject-specific PC scores (or loadings) on each eigenfunction.
+Additionally, \(\xi_{ik}\) are mutually independent obver both subject
+(\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
+normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
+eigenvalues: \(\int \phi_k^2(t)dt=\lambda_k\). \(\epsilon_i(t)\) here is
+a residual function that accounts for the variation not explained by the
+first K eigenfunctions from FPCA. We assume it follows a zero-mean
+Gaussian process. At a specific point t,
+\(\epsilon_i(t) \sim N(0, \sigma^2)\).
+
+Based on the problem set up above, we propose the following algorithm
+for PFCA on the unobserved latent process \(\eta_i(t)\):
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  Need better notation system
-\end{itemize}
-
-\hypertarget{result}{%
-\section{Result}\label{result}}
-
-\begin{itemize}
-\tightlist
+  Bin the observed outcomes in to small, non-overlapping, equal length
+  intervals. We hereafter index the bins by their midpoints \(s\).
 \item
-  Repeat simulation: is it necessary in this case?
+  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
+  \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
+  bin \(s\). From this series of models we can get estiamtes of
+  population mean \(\hat{\beta}_0(s)\) and subject-level random effect
+  \(\hat{b}_i(s)\), thus estimate of the individual latent functions at
+  every bin: \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
 \item
-  Different set-up:
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\alph{enumi}.}
-\tightlist
+  Fit FPCA on the estimated latent functions \(\hat{\eta}_i(s)\), and
+  obtain estimates of basis functions
+  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
+  \(\hat{\lambda}_1...\hat{\lambda}_k\), population mean
+  \(\hat{\beta}_0(s)\) and residual variance \(\hat{\sigma}^2\).
 \item
-  Different eigenfunctions: with or without periodicity
-\item
-  Outcome: binary or count
-\item
-  Sample size
-\item
-  Grid density
-\item
-  Bin width, overlap or not
+  With components extracted above, calculate the maximum likelihood
+  estimate (MLE) of the subject-specific PC scores \(\hat{\xi}_{ik}\) of
+  new samples based on their partially observed data. Then the value of
+  latent functions at unobserved points can be estimated as
+  \(\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)
 \end{enumerate}
 
-\begin{itemize}
-\tightlist
-\item
-  Real data application
-\end{itemize}
+Following the algorithm above, predictions of individual latent
+functions are made on the binned grid based on partially observed
+non-Gaussian functions tracks of new subjects. Since the bins are set up
+to be small in length, the binned grid would still be dense enough.
+However, in situations where we need predictions on the original,
+un-binned grid instead, linear interpolation turns out to be a fast,
+convenient way with good performance for prediction at points between
+bins.
+
+Precision of prediction, usually quantified by the variance of
+prediction error \(Var(\hat{\eta_i}(s)-\eta_i(s))\), is also
+straightforward under this framework. In step 4 we calculated the MLE of
+\(\hat{\xi}_{ik}\). Based on likelihood theory, its variance can be
+estimated with observed information \({I}(\hat{\xi}_{ik})\), which is
+essentially the second derivative of likelihood at \(\hat{\xi}_{ik}\).
+Therefore, the variation of prediction interval is:
+
+\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
+Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
+\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).
+
+\hypertarget{simulation}{%
+\section{Simulation}\label{simulation}}
+
+In this section, we illustrate the predictive performance and
+computational efficiency of the proposed method through a simulation
+study. We simulated 50 datasets, each with 500 subjects. For every
+subject, we generate 1000 binary outcomes \(Y_i(t) \in (0, 1)\) across
+functional domain \(t \in [0, 1]\), where the distribution of outcome is
+characterized by a continuous latent function. The data generation
+mechanism can be expressed as follows:
+
+\[\begin{aligned}
+Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
+\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
+\end{aligned}\]
+
+In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
+independent normal random variables \(\xi_{ik}\sim N(0, \gamma_k)\).
+Here we set the values of \(\gamma_k\) to be \(0.5^{k-1}\),
+\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
+generate data on a regular grid, which means observations points are
+equally distributed across \([0, 1]\) and are the same for all subjects.
+
+We use two metrics to evaluate the out-of-sample predictive performance:
+integrated squared error (ISE) and
+Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
+accucay of latent continuous function. It is evaluated on the binned
+grid at midpoints of each unobserved bin. If the entire functional
+domian has S bins, but we have observations up to the mth bin, then ISE
+is defined as
+\(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
+
+The second metric, AUC, focuses on evaluation of prediction of the
+binary outcome. Since the binary outcomes are generated on the original,
+un-binned grid, we evaluated AUC on this grid as well and estimated
+values of latent functions between bins with linear interpolation. Just
+like ISE, the AUC report is also the average value across the whole
+sample.
+
+As a reference method, we compare our method to Generalized Linear Mixed
+Models using Adaptive Gaussian Quadrature (GLMMadaptive).
+
+\hypertarget{data-application}{%
+\section{Data application}\label{data-application}}
 

### 2023-03-14 Code/GLMM-FPCA.R

@@ -1,66 +1,23 @@
 ##### Local GLMMs #####
 
-bin_w <- 10 # bin width
-n_bin <- J/bin_w # number of bins
-brks <- seq(0, J, by = bin_w) # cutoff points
-mid <- (brks+bin_w/2)[1:n_bin] # mid points
+# get predicted latent function
+# df must have: outcome Y, subject id
 
-# bin observations
-df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
-df$bin <- as.numeric(as.character(df$bin))
-
-# check overlap
-# df %>% 
-#   filter(id %in% 1:4) %>% 
-#   ggplot()+
-#   geom_line(aes(x=sind, y=eta_i, col=as.factor(bin)), show.legend = F)+
-#   geom_point(aes(x=sind, y=Y, col=as.factor(bin)), size = 0.05, show.legend = F)+
-#   facet_wrap(~id)
-
-# fit local linear mixed models in each bin
-# and extract subject level estimation
-## df: function with binary outcome Y, with observations in one time bein
 pred_latent <- function(df){
-  this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
-  eta_hat <- predict(this_glm, type = "link")
-  df$eta_hat <- eta_hat
-  return(df)
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+    eta_hat <- predict(this_glm, type = "link")
+    df$eta_hat <- eta_hat
+    return(df)
 }
-
-# additionally, get the average estimation across subject 
+  
+# Additionally, get the average estimation across subject 
 # by extacting the random intercepts from each bin
 mean_latent <- function(df){
-  this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
-  beta_hat <- coef(summary(this_glm))[1]
-  return(beta_hat)
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+    beta_hat <- coef(summary(this_glm))[1]
+    return(beta_hat)
 }
-
-## do that for all time bin
-df_bin_lst <- split(df, f = df$bin)
-
-###### subject specific
-df_pred_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
-# lapply(df_pred_latent, dim)
[…]

### 2023-03-14 Code/OutSampMLE.R

@@ -45,6 +45,7 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
   score_out <- multiroot(f=llh_div, start = rep(0, 4), df_new=df_new, fpca_fit=fpca_fit)$root
   eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions%*%score_out
   
+  
   return(list(eta_pred = eta_pred_out,
               score_out = score_out))
 }
@@ -52,18 +53,18 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
 
 #### Estimate standard error of MLE ####
 
-# Jacabian matrix of first dirivative of llh
-try <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
-                    select(-eta_i),
-                  fpca_fit = fpca_fit)$score_out
-
-I_mat <- -gradient(f = llh_div, x = try,
-         df_new = df %>% filter(id==1 & sind_inx<=195) %>%
-           select(-eta_i),
-         fpca_fit=fpca_fit)
-
-var <- diag(solve(I_mat))
-sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions)))
+# # Jacabian matrix of first dirivative of llh
+# try <- out_samp_dyn_pred(df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+#                     select(-eta_i),
+#                   fpca_fit = fpca_fit)$score_out
+# 
+# I_mat <- -gradient(f = llh_div, x = try,
+#          df_new = df %>% filter(id==1 & sind_inx<=195) %>%
+#            select(-eta_i),
+#          fpca_fit=fpca_fit)
+# 
+# var <- diag(solve(I_mat))
+# sqrt(diag(fpca_fit$efunctions %*% solve(I_mat) %*% t(fpca_fit$efunctions)))
 
 # I_mat <- function(xi, df_new, fpca_fit){
 #   

### 2023-03-14 Code/WNARSim.R

@@ -1,0 +1,124 @@
+# This script implements simulation for WNAR manuscript
+
+set.seed(1025)
+
+library(here)
+library(tidyverse)
+# library(ggplot2)
+# theme_set(theme_minimal())
+# library(fcr)
+library(refund)
+library(lme4)
+# library(ggpubr)
+# library(gridExtra)
+# library(kableExtra)
+
+
+
+
+#### Generate data ####
+source(here("Code/ToySimulation.R"))
+
+
+sim_data <- list()
+
+M <- 500
+
+# generate 500 subject with four eigenfunctions, observation over 1000 time points
+for(m in 1:M){
+  sim_data[[m]] <- gen_data(N = 500, J = 1000, sind = seq(0,1,len=1000) , lambda = 0.5^(0:(4-1)))
+}
+
+save(sim_data, file = here("Data/sim_data.RData"))
+
+load(here("Data/sim_data.RData"))
+
+#### fGFPCA ####
+source(here("Code/GLMM-FPCA.R"))
+source(here("Code/OutSampMLE.R"))
+
+J <- 1000
+N <- 500
+K <- 4
+
+
+# result container
+pred_lst <- list()
+score_lst <- list()
+
+
+
+t1 <- Sys.time()
+for(m in 1:50){
+  
+  df <- sim_data[[m]]
+  
+  # bin observations
+  bin_w <- 10 # bin width
+  n_bin <- J/bin_w # number of bins
+  brks <- seq(0, J, by = bin_w) # cutoff points
+  mid <- (brks+bin_w/2)[1:n_bin] # mid points
+  df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
+  df$bin <- as.numeric(as.character(df$bin))
+  
+  df_bin_lst <- split(df, f = df$bin)
+  
+  # fit local GLMM and estimate latent function
+  df_pred_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+  df_pred_latent <- bind_rows(df_pred_latent)
+
+  # put predicted latent function to wide format
+  df_pred_latent <- df_pred_latent %>% arrange(id, sind_inx)
+  df_pred_unique <- df_pred_latent %>% select(id, bin, eta_hat) %>% distinct()
[…]

### 2023-03-14 Write-ups/Manuscript.Rmd
evaluation metrix

@@
 knitr::opts_chunk$set(echo = FALSE)
+library(tidyverse)
+library(here)
+library(ROCR)
+
+set.seed(314)
 ```
@@
 As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive).
+This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes.
+Just like many mixed models, this method is very limited in terms of flexibility.
+For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$.
+While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.
+
+## Result
+
+```{r}
+load(here("Data/sim_data.RData"))
+load(here("Data/SimOutput_fGFPCA.RData"))
+
+# other quantities
+mid <- unique(exp_df$bin)
+N <- 500
+M <- length(pred_lst)
+J <- 1000
+
+# code prediction value at observed points as NA
+for(m in 1:length(pred_lst)){
+  pred_lst[[m]]$pred_t200[pred_lst[[m]]$bin <= 200] <- NA
+  pred_lst[[m]]$pred_t400[pred_lst[[m]]$bin <= 400] <- NA
+  pred_lst[[m]]$pred_t600[pred_lst[[m]]$bin <= 600] <- NA
+  pred_lst[[m]]$pred_t800[pred_lst[[m]]$bin <= 800] <- NA
+}
+
+```
+
+```{r}
+# true data
+df_true <- sim_data[[1]] %>% select(id, sind_inx, eta_i)
+
+# prediction results used for visualization
+exp_df <- pred_lst[[1]]
+
+# randomly draw four samples for visualization
+rand_id <- sample(exp_df$id, 4)
+
+exp_df %>%
+  left_join(df_true %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_i))+
+  geom_line(aes(x=bin, y = pred_t200, col = "200"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t400, col = "400"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t600, col = "600"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t800, col = "800"), linetype = "dashed", na.rm = T)+
+  facet_wrap(~id, scales = "free_y")
+
+```
+
+```{r}
+runtime
+# calculate ISE on binned grid
+ise_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
+colnames(ise_mat) <- c("t200", "t400", "t600", "t800")
+
+for(m in 1:M){
+  # true value at midpoints
+  true_eta <- sim_data[[m]] %>% filter(sind_inx %in% mid) %>% select(id, sind_inx, eta_i) %>% rename(bin=sind_inx)
+  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
+  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))
+
+  ise_m <- c(
+    t200 = sum((pred_eta$pred_t200-pred_eta$eta_i)^2, na.rm = T)/N,
+    t400 = sum((pred_eta$pred_t400-pred_eta$eta_i)^2, na.rm = T)/N,
+    t600 = sum((pred_eta$pred_t600-pred_eta$eta_i)^2, na.rm = T)/N,
+    t800 = sum((pred_eta$pred_t800-pred_eta$eta_i)^2, na.rm = T)/N
+  )
+
+  ise_mat[m, ] <- ise_m
+}
+```
+```{r}
+# calculate AUC on unbinned grid
+auc_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
+colnames(auc_mat) <- c("t200", "t400", "t600", "t800")
+
+for(m in 1:M){
+  # true value as matrix
+  Y_mat <- sim_data[[m]]  %>% select(id, sind_inx, Y) %>% rename(bin=sind_inx) %>% 
+    pivot_wider(names_from = id, values_from = Y) %>%
+    column_to_rownames("bin")
+
+  # calculate AUC
+  auc_mt <- rep(NA, 4) 
+  for(t in c(200, 400, 600, 800)){
+
+    # predicted value
+    pred_name <- paste("pred_t", t, sep = "")
+    pred_mat <- pred_lst[[m]] %>% select(id, bin, all_of(pred_name)) %>% 
+      pivot_wider(names_from = id, values_from = pred_name) %>%
+      column_to_rownames("bin")
+    pred_full <- sapply(1:N,
+                           function(x){approx(x=mid, y=pred_mat[ , x], xout = 1:J)$y})
+    pred_full <- exp(pred200_full)/(1+exp(pred200_full))
+    row_id <- complete.cases(pred_full)
+    auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
+    auc_mt[t] <- mean(unlist(auc_perf@y.values))
+  }
+
+  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
+  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))
+}
+```
 

### 2023-03-14 Write-ups/Manuscript.tex
evaluation metrix

@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 13, 2023}
+\date{March 14, 2023}
 
@@
 As a reference method, we compare our method to Generalized Linear Mixed
-Models using Adaptive Gaussian Quadrature (GLMMadaptive).
+Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
+the fastest existing method developed for dynamic prediction of repeated
+generalized outcomes. Just like many mixed models, this method is very
+limited in terms of flexibility. For example, the model used for
+prediction of our simulated datset would simply be an linear model with
+one covariate indicating observation time:
+\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
+of this mixed model can be increased using spline functions, the
+dimension of spline basis is also restricted by computational ability,
+and is unfeasible to implement under the scale of our simulated data or
+the complexity of our proposed method.
+
+\hypertarget{result}{%
+\subsection{Result}\label{result}}
+
+\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
+
+\begin{verbatim}
+## [1] 5.033985 5.134139 5.195642 5.152902 5.082466
+\end{verbatim}
 

### 2023-03-14 Code/OutSampMLE.R

@@ -22,7 +22,7 @@ llh_div <- function(xi, df_new, fpca_fit){
   hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
     select(Y) %>% unlist()# number of success
   nf <- ns-hs # number of failure
-  max_bin <- length(unique(df_new$bin)) # assume new skipped bins 
+  max_bin <- length(unique(df_new$bin)) # assume no skipped bins 
   
   # fpca objects
   phi_m <- fpca_fit$efunctions[1:max_bin, ]
@@ -50,6 +50,8 @@ out_samp_dyn_pred <- function(df_new, fpca_fit){
               score_out = score_out))
 }
 
+# test
+# multiroot(f=llh_div, start = rep(0, 4), df_new=df_i %>% filter(sind_inx<=800), fpca_fit=fpca_mod)
 
 #### Estimate standard error of MLE ####
 

### 2023-03-14 Code/ToySimulation.R

@@ -1,58 +1,81 @@
 
-##### generate data #####
+# This sciprt generateds 500 data sets
 
-# rm(list=ls())
+set.seed(314)
+N <- 500 # sample size
+J <- 1000 # number of observation points
+
+sind = seq(0,1,len=J) # observations points
 
 
-#####  generate data f ####
+f_0 <- function(s) 0 # mean functions
+f_0(5)
 
-## basic parameters
-# number of subjects
 
-## simulate outcomes g(E[Y_i(s)]) = \eta_i(s) = f_0(s) + f_1(s)X_i + b_i(s)
-# f_0(s) = 0
-# b_i(s) = \sum_{k=1}^4 \phi_k(s)\xi_{ik}
-#        \xi_{ik} \sim N(0, \lambda_k)
+#### eigenfunctions #####
+K <- 4 # number of eigenfunctions
+phi <- sqrt(2)*cbind(sin(2*pi*sind),cos(2*pi*sind),
+                     sin(4*pi*sind),cos(4*pi*sind))
 
-## fixed effects f_0, f_1
+par(mfrow=c(2, 2))
+plot(sind, phi[,1])
+plot(sind, phi[,2])
+plot(sind, phi[,3])
+plot(sind, phi[,4])
 
-gen_data <- function(N = 500, J = 1000, sind = seq(0,1,len=J) , lambda = 0.5^(0:(K-1))){
+
+#### generated data ####
+
+sim_data <- list()
+M <- 500 # number of datasets to generate
+lambda = 0.5^(0:(K-1)) # eigenvalues
+
+for(m in 1:M){
   
-  # mean
-  f_0 <- function(s) 0
-
-  ## random effects
-  # eigenfunctions \phi_k(s) evaluated on the observed grid
-  phi <- sqrt(2)*cbind(sin(2*pi*sind),cos(2*pi*sind),
-                       sin(4*pi*sind),cos(4*pi*sind))
-  
-  # subject-specific weights/coefficients \xi_ik
-  K <- 4
+  # generate score
   xi <- matrix(rnorm(N*K),N,K)
   xi <- xi %*% diag(sqrt(lambda))
+  
+  # subject-specific random effect
   b_i <- xi %*% t(phi); # of size N by J
-
-  ## linear predictor \eta_i(s), also the latent function
+  
+  # latent gaussian function
   eta_i <- t(vapply(1:N, function(x){
     f_0(sind) + b_i[x,]
[…]

### 2023-03-14 Code/WNARSim.R

@@ -1,6 +1,6 @@
 # This script implements simulation for WNAR manuscript
 
-set.seed(1025)
+set.seed(314)
 
 library(here)
 library(tidyverse)
@@ -9,6 +9,7 @@ library(tidyverse)
 # library(fcr)
 library(refund)
 library(lme4)
+library(ROCR)
 # library(ggpubr)
 # library(gridExtra)
 # library(kableExtra)
@@ -16,103 +17,130 @@ library(lme4)
 
 
 
-#### Generate data ####
-source(here("Code/ToySimulation.R"))
-
-
-sim_data <- list()
-
-M <- 500
-
-# generate 500 subject with four eigenfunctions, observation over 1000 time points
-for(m in 1:M){
-  sim_data[[m]] <- gen_data(N = 500, J = 1000, sind = seq(0,1,len=1000) , lambda = 0.5^(0:(4-1)))
-}
-
-save(sim_data, file = here("Data/sim_data.RData"))
+#### load simulated data ####
 
 load(here("Data/sim_data.RData"))
 
+
+#### simulation set up ####
+
+J <- 1000 # number of observations points
+N <- 500 # sample size
+K <- 4 # number of eigenfunctions
+M <- 500 # number of simulation
+
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
 #### fGFPCA ####
-source(here("Code/GLMM-FPCA.R"))
+source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
 source(here("Code/OutSampMLE.R"))
-
-J <- 1000
-N <- 500
-K <- 4
-
 
 # result container
 pred_lst <- list()
 score_lst <- list()
+runtime <- rep(NA, 5)
 
 
 
-t1 <- Sys.time()
-for(m in 1:50){
+pb = txtProgressBar(min = 0, max = 5, initial = 0, style = 3) 
+
+
+for(m in 1:5){
   
   df <- sim_data[[m]]
   
+  t1 <- Sys.time()
   # bin observations
-  bin_w <- 10 # bin width
-  n_bin <- J/bin_w # number of bins
-  brks <- seq(0, J, by = bin_w) # cutoff points
-  mid <- (brks+bin_w/2)[1:n_bin] # mid points
[…]

### 2023-03-14 Write-ups/Manuscript.Rmd
calculated AUC and ISE of fGFPCA

@@
 # other quantities
-mid <- unique(exp_df$bin)
+mid <- unique(pred_lst[[1]]$bin)
 N <- 500
@@
   # calculate AUC
-  auc_mt <- rep(NA, 4) 
-  for(t in c(200, 400, 600, 800)){
+  t_vec <- c(200, 400, 600, 800)
+  for(i in seq_along(t_vec)){
 
+    t <- t_vec[i]
     # predicted value
@@
                            function(x){approx(x=mid, y=pred_mat[ , x], xout = 1:J)$y})
-    pred_full <- exp(pred200_full)/(1+exp(pred200_full))
+    pred_full <- exp(pred_full)/(1+exp(pred_full))
     row_id <- complete.cases(pred_full)
     auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
-    auc_mt[t] <- mean(unlist(auc_perf@y.values))
+    auc_mat[m, i] <- mean(unlist(auc_perf@y.values))
   }
 
-  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
-  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))
 }
+
+auc_mat
 ```

### 2023-03-14 Write-ups/Manuscript.tex
calculated AUC and ISE of fGFPCA

@@
 
+\begin{verbatim}
+## Warning: Using an external vector in selections was deprecated in tidyselect 1.1.0.
+## i Please use `all_of()` or `any_of()` instead.
+##   # Was:
+##   data %>% select(pred_name)
+## 
+##   # Now:
+##   data %>% select(all_of(pred_name))
+## 
+## See <https://tidyselect.r-lib.org/reference/faq-external-vector.html>.
+\end{verbatim}
+
+\begin{verbatim}
+##           t200      t400      t600      t800
+## [1,] 0.6831917 0.7072828 0.6962669 0.6216755
+## [2,] 0.6881236 0.7011135 0.6886562 0.6203330
+## [3,] 0.6903756 0.7055576 0.6949729 0.6219840
+## [4,] 0.6869610 0.6983328 0.6946955 0.6178570
+## [5,] 0.6857926 0.7086377 0.6983496 0.6223322
+\end{verbatim}
+
 \hypertarget{data-application}{%

### 2023-03-15 Write-ups/Manuscript.Rmd
solve AUC error

@@
 ```{r}
-runtime
+mean(runtime)
 # calculate ISE on binned grid
@@
 }
+
 ```
+
 ```{r}
@@
 colnames(auc_mat) <- c("t200", "t400", "t600", "t800")
+
+skip_count <- 0
 
@@
     pred_mat <- pred_lst[[m]] %>% select(id, bin, all_of(pred_name)) %>% 
-      pivot_wider(names_from = id, values_from = pred_name) %>%
+      pivot_wider(names_from = id, values_from = all_of(pred_name)) %>%
       column_to_rownames("bin")
@@
     row_id <- complete.cases(pred_full)
-    auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
-    auc_mat[m, i] <- mean(unlist(auc_perf@y.values))
+
+    # calculate AUC, skip subjects with no variation in outcome Y
+    auc_lst <- lapply(1:N, function(x){
+      auc_try <- try(performance(prediction(pred_full[row_id, x], Y_mat[row_id, x]), "auc"), 
+                     silent = T)
+      auc_pref <- ifelse(class(auc_try)=="try-error",NA, auc_try@y.values)
+      return(auc_pref)               
+    })
+
+    auc_mat[m, i] <- mean(unlist(auc_lst), na.rm = T)
+    skip_count <- skip_count+sum(is.na(auc_lst))
   }
@@
 
-auc_mat
 ```

### 2023-03-15 Code/WNARSim.R

@@ -27,7 +27,7 @@ load(here("Data/sim_data.RData"))
 J <- 1000 # number of observations points
 N <- 500 # sample size
 K <- 4 # number of eigenfunctions
-M <- 500 # number of simulation
+M <- 50 # number of simulation
 
 # bin data
 bin_w <- 10 # bin width
@@ -42,14 +42,14 @@ source(here("Code/OutSampMLE.R"))
 # result container
 pred_lst <- list()
 score_lst <- list()
-runtime <- rep(NA, 5)
+runtime <- rep(NA, M)
 
 
 
-pb = txtProgressBar(min = 0, max = 5, initial = 0, style = 3) 
+pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 
 
-for(m in 1:5){
+for(m in 1:M){
   
   df <- sim_data[[m]]
   

### 2023-03-16 Write-ups/Manuscript.Rmd
simulation of GLMM adaptvie

@@
 
+colMeans(ise_mat)
+```
+
+```{r, eval=FALSE}
+# calculate ISE on binned grid
+ise_ref_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
+colnames(ise_ref_mat) <- c("t200", "t400", "t600", "t800")
+
+for(m in 1:M){
+  # true value at midpoints
+  pred_eta <- pred_lst_ref[[m]]
+
+  ise_m <- c(
+    t200 = sum((pred_eta$pred_t200-pred_eta$eta_i)^2, na.rm = T)/N,
+    t400 = sum((pred_eta$pred_t400-pred_eta$eta_i)^2, na.rm = T)/N,
+    t600 = sum((pred_eta$pred_t600-pred_eta$eta_i)^2, na.rm = T)/N,
+    t800 = sum((pred_eta$pred_t800-pred_eta$eta_i)^2, na.rm = T)/N
+  )
+
+  ise_ref_mat[m, ] <- ise_m
+}
+
+colMeans(ise_ref_mat)
 ```
@@
 
+colMeans(auc_mat)
+```
+
+```{r, eval=FALSE}
+# calculate AUC on unbinned grid
+auc_mat_ref <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
+colnames(auc_mat_ref) <- c("t200", "t400", "t600", "t800")
+
+skip_count <- 0
+
+for(m in 1:M){
+  # true value as matrix
+  Y_mat <- sim_data[[m]]  %>% select(id, sind_inx, Y) %>% 
+    pivot_wider(names_from = id, values_from = Y) %>%
+    column_to_rownames("sind_inx")
+
+  # calculate AUC
+  t_vec <- c(200, 400, 600, 800)
+  for(i in seq_along(t_vec)){
+
+    t <- t_vec[i]
+    # predicted value
+    pred_name <- paste("pred_t", t, sep = "")
+    pred_mat <- pred_lst_ref[[m]] %>% select(id, sind, all_of(pred_name)) %>% 
+      pivot_wider(names_from = id, values_from = all_of(pred_name)) %>%
+      column_to_rownames("sind")
+    pred_mat <- exp(pred_mat)/(1+exp(pred_mat))
+    row_id <- complete.cases(pred_mat)
+
+    # calculate AUC, skip subjects with no variation in outcome Y
+    auc_lst <- lapply(1:N, function(x){
+      auc_try <- try(performance(prediction(pred_mat[row_id, x], Y_mat[row_id, x]), "auc"), 
+                     silent = T)
+      auc_pref <- ifelse(class(auc_try)=="try-error",NA, auc_try@y.values)
+      return(auc_pref)               
+    })
+
+    auc_mat[m, i] <- mean(unlist(auc_lst), na.rm = T)
+    skip_count <- skip_count+sum(is.na(auc_lst))
+  }
+
+}
+
 ```

### 2023-03-16 Code/WNARSim.R

@@ -10,6 +10,7 @@ library(tidyverse)
 library(refund)
 library(lme4)
 library(ROCR)
+library(GLMMadaptive)
 # library(ggpubr)
 # library(gridExtra)
 # library(kableExtra)
@@ -106,30 +107,52 @@ save(pred_lst, score_lst, runtime, file = here("Data/SimOutput_fGFPCA.RData"))
 #### GLMMadaptive ####
 
 pred_lst_ref <- list()
-runtime_ref <- rep(NA, 5)
+runtime_ref <- rep(NA, M)
 
 
 
-pb = txtProgressBar(min = 0, max = 5, initial = 0, style = 3) 
+pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 
 for(m in 1:5){
   df <- sim_data[[m]]
   
+  t1 <- Sys.time()
   # GLMM adaptive model
   adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df, family = binomial())
   
   # container
-  
-  
+  df_est_latent <- df
+  df_est_latent[, 'pred_t200'] <- df_est_latent[, 'pred_t400'] <- df_est_latent[, 'pred_t600'] <- df_est_latent[, 'pred_t800'] <- NA 
   
   # time up to t
-  pred_t200 <- predict(adglmm, newdata = df %>% filter(sind_inx <= t200),
+  pred_t200 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 200),
                         newdata2 =  df %>% filter(sind_inx > 200), type = "subject_specific", type_pred = "link",
                           se.fit = TRUE, return_newdata = TRUE)
-  preds_adglmm$newdata2 %>% select(id, sind_inx, pred) %>% 
-    filter(sind_inx <= max(mid)) %>% 
-    pivot_wider(names_from = id, values_from = pred) %>% 
-    column_to_rownames("sind_inx") %>% as.matrix()
+  df_est_latent$pred_t200[df_est_latent$sind_inx>200] <- pred_t200$newdata2$pred
+  
+  pred_t400 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 400),
+                       newdata2 =  df %>% filter(sind_inx > 400), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
+  df_est_latent$pred_t400[df_est_latent$sind_inx>400] <- pred_t400$newdata2$pred
+  
[…]

### 2023-03-17 Write-ups/Manuscript.Rmd
simulation section of WNAR paper

--- 2023-03-14 Write-ups/Manuscript.Rmd
@@
 library(tidyverse)
+theme_set(theme_minimal())
 library(here)
 library(ROCR)
+library(kableExtra)
+library(knitr)
+options(knitr.kable.NA = "")
 
 set.seed(314)
+
+load(here("Data/sim_data.RData"))
+load(here("Data/SimOutput_fGFPCA.RData"))
+load(here("Data/SimOutput_GLMMadaptive.RData"))
 ```
@@
 If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.
-
 The second metric, AUC, focuses on evaluation of prediction of the binary outcome.
 Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation.
-Just like ISE, the AUC report is also the average value across the whole sample.
 
@@
 
-## Result
+The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation.
+The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time.
+As Table 1 reveals, the fGFPCA outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accomodate simple model structures.
+In addition, fGFPCA also takes less computation time.
+For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.
+
+Because of the flexibility of our proposed model framework, the accuracy of prediction at specific time points would improve with more observed data.
+This is revealed in Table 1 as ISE decreases and AUC increases with maximum observed time (left-to-right), also in Figure 1 as predicted curves get closer with longer observed track.
+However the same tendency is not observed with GLMMadaptive models, as a result of restricted model flexibility.
+A linear model that fits well to a specific part of latent function can fit very badly to the following parts, especially when the underlying latent function has cyclic patterns.
+Therefore, more observations do not necessarily make the model more predictive.
 
 ```{r}
-load(here("Data/sim_data.RData"))
-load(here("Data/SimOutput_fGFPCA.RData"))
 
@@
 
+# for fGFPCA
 # code prediction value at observed points as NA
@@
 
+# for GLMM adaptive
+# add in observation time index
+for(m in 1:length(pred_lst_ref)){
+  pred_lst_ref[[m]]$sind_inx <- rep(1:J, N)
+}
 ```
 
-```{r}
+```{r Figure, fig.cap="Predicted track of four randomly selected subjects.
+The grey dash line indicates the true latent continuous function.
+The dashed lines indicated predicted latent function tracks, and color indicates different observations time."}
+# figure of fGFPCA prediction
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
 # true data
-df_true <- sim_data[[1]] %>% select(id, sind_inx, eta_i)
+df_true_sim1 <- sim_data[[1]] %>% select(id, sind_inx, eta_i)
 
 # prediction results used for visualization
-exp_df <- pred_lst[[1]]
+exp_df_sim1 <- pred_lst[[1]]
 
 # randomly draw four samples for visualization
-rand_id <- sample(exp_df$id, 4)
+rand_id <- sample(exp_df_sim1$id, 4)
 
-exp_df %>%
-  left_join(df_true %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
+exp_df_sim1 %>%
+  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
   filter(id %in% rand_id) %>%
   ggplot()+
-  geom_line(aes(x=bin, y=eta_i))+
-  geom_line(aes(x=bin, y = pred_t200, col = "200"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t400, col = "400"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t600, col = "600"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t800, col = "800"), linetype = "dashed", na.rm = T)+
-  facet_wrap(~id, scales = "free_y")
-
+  geom_line(aes(x=bin, y=eta_i, col = "True"))+
+  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
+  facet_wrap(~id, scales = "free_y")+
+  labs(y = "Latent function", x = "Bin", colour= "Observed time")+
+  scale_color_manual(values=cbPalette)+
+  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
 ```
 
-```{r}
-runtime
+```{r ISE_fGFPCA}
 # calculate ISE on binned grid
-ise_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
-colnames(ise_mat) <- c("t200", "t400", "t600", "t800")
+ise_mat <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation
 
 for(m in 1:M){
+
   # true value at midpoints
@@
 
-  ise_m <- c(
-    t200 = sum((pred_eta$pred_t200-pred_eta$eta_i)^2, na.rm = T)/N,
-    t400 = sum((pred_eta$pred_t400-pred_eta$eta_i)^2, na.rm = T)/N,
-    t600 = sum((pred_eta$pred_t600-pred_eta$eta_i)^2, na.rm = T)/N,
-    t800 = sum((pred_eta$pred_t800-pred_eta$eta_i)^2, na.rm = T)/N
-  )
+  # split unobserved time into intervals of length 200
+  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
+  pred_eta$bin_cat <- cut(pred_eta$bin, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
 
-  ise_mat[m, ] <- ise_m
+  # SE at each observed time
+  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
+                                  ise_t400 = (pred_t400-eta_i)^2,
+                                  ise_t600 = (pred_t600-eta_i)^2,
+                                  ise_t800 = (pred_t800-eta_i)^2)
+
+  # subject-average SE
+  ise_mat[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
+    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()
 }
+
+ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
+colnames(ave_ise_mat) <- c("200", "400", "600", "800")
+rownames(ave_ise_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+
 ```
-```{r}
-# calculate AUC on unbinned grid
-auc_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
-colnames(auc_mat) <- c("t200", "t400", "t600", "t800")
+
+```{r, ISE_GLMMadaptive}
+# calculate ISE on binned grid
+ise_mat_ref <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation
 
 for(m in 1:M){
-  # true value as matrix
-  Y_mat <- sim_data[[m]]  %>% select(id, sind_inx, Y) %>% rename(bin=sind_inx) %>% 
-    pivot_wider(names_from = id, values_from = Y) %>%
-    column_to_rownames("bin")
 
-  # calculate AUC
-  t_vec <- c(200, 400, 600, 800)
-  for(i in seq_along(t_vec)){
+  # true value at midpoints
+  pred_eta <- pred_lst_ref[[m]] %>% filter(sind_inx %in% mid)
 
-    t <- t_vec[i]
-    # predicted value
-    pred_name <- paste("pred_t", t, sep = "")
-    pred_mat <- pred_lst[[m]] %>% select(id, bin, all_of(pred_name)) %>% 
-      pivot_wider(names_from = id, values_from = pred_name) %>%
-      column_to_rownames("bin")
-    pred_full <- sapply(1:N,
-                           function(x){approx(x=mid, y=pred_mat[ , x], xout = 1:J)$y})
-    pred_full <- exp(pred_full)/(1+exp(pred_full))
-    row_id <- complete.cases(pred_full)
-    auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
-    auc_mat[m, i] <- mean(unlist(auc_perf@y.values))
-  }
+  # split unobserved time into intervals of length 200
+  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
+  pred_eta$bin_cat <- cut(pred_eta$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
+
+  # SE at each observed time
+  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
+                                  ise_t400 = (pred_t400-eta_i)^2,
+                                  ise_t600 = (pred_t600-eta_i)^2,
+                                  ise_t800 = (pred_t800-eta_i)^2)
+
+  # subject-average SE
+  ise_mat_ref[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
+    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()
 
@@
 
-auc_mat
+ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
+colnames(ave_ise_mat_ref) <- c("200", "400", "600", "800")
+rownames(ave_ise_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+
+# cbind(ave_ise_mat, ave_ise_mat_ref)
+```
+
+```{r AUC_fGFPCA, cache=TRUE}
+# maximum observations time
+t_vec <- c(200, 400, 600, 800)
+
+# calculate AUC on unbinned grid
+auc_mat <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time
+
+for(m in 1:M){
+
+  Y_df <- sim_data[[m]]
+  eta_pred <- pred_lst[[m]]
+
+  # use interpolation to extend prediction to the unbinned grid
+  for(i in 1:N){
+   Y_df$pred_200[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t200[eta_pred$id==i], xout = 1:J)$y
+   Y_df$pred_400[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t400[eta_pred$id==i], xout = 1:J)$y
+   Y_df$pred_600[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t600[eta_pred$id==i], xout = 1:J)$y
+   Y_df$pred_800[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t800[eta_pred$id==i], xout = 1:J)$y
+  }
+
+  # split unobserved time into intervals of length 200
+  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
+  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
+  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
+  Y_df_lst <- Y_df_lst[-1]
+
+  for(j in seq_along(t_vec)){
+
+    t <- t_vec[j]
+    pred_name <- paste("pred_", t, sep = "")
+
+    # calculate AUC
+    auc_lst <- lapply(Y_df_lst[j:4], 
+                       function(x){
+                         row_id <- !is.na(x[, pred_name])
+                         this_eta <- x[row_id, pred_name]
+                         this_Y <- x[row_id, "Y"]
+                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
+                         return(auc_pref@y.values)
+                })
+
+    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
+    auc_mat[ , j, m] <- auc_vec
+
+}}
+
+ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
+colnames(ave_auc_mat) <- c("200", "400", "600", "800")
+rownames(ave_auc_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+
+```
+
+```{r AUC_GLMMadaptive}
+# calculate AUC on unbinned grid
+auc_mat_ref <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time
+
+for(m in 1:M){
+
+  Y_df <- sim_data[[m]]
+  eta_pred <- pred_lst_ref[[m]]
+
+  Y_df <- Y_df %>% left_join(eta_pred %>% select(id, sind_inx, starts_with("pred_")), by = c("id", "sind_inx"))
+
+  # split unobserved time into intervals of length 200
+  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
+  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
+  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
+  Y_df_lst <- Y_df_lst[-1]
+
+  for(j in seq_along(t_vec)){
+
+    t <- t_vec[j]
+    pred_name <- paste("pred_t", t, sep = "")
+
+    # calculate AUC
+    auc_lst <- lapply(Y_df_lst[j:4], 
+                       function(x){
+                         row_id <- !is.na(x[, pred_name])
+                         this_eta <- x[row_id, pred_name]
+                         this_Y <- x[row_id, "Y"]
+                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
+                         return(auc_pref@y.values)
+                })
+
+    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
+    auc_mat_ref[ , j, m] <- auc_vec
+
+  }
+}
+
+ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
+colnames(ave_auc_mat_ref) <- c("200", "400", "600", "800")
+rownames(ave_auc_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+
+```
+
+```{r format table}
+ise_tb <- cbind(ave_ise_mat, ave_ise_mat_ref)
+auc_tb <- cbind(ave_auc_mat, ave_auc_mat_ref)
+
+rbind(ise_tb, auc_tb) %>%
+  kable(booktabs = T, digit=2,
+        caption = "Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.",
+        align = "lcccccccc") %>%
+  kable_styling() %>%
+  add_header_above(c(" " = 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" " = 1, "Maximum observed time" = 8)) %>% 
+  group_rows(index = c("Prediction time window" = 8)) %>%
+  group_rows(index = c("ISE" = 4, "AUC" = 4)) 
 ```

### 2023-03-17 Write-ups/Manuscript.tex
solve AUC error
simulation of GLMM adaptvie
simulation section of WNAR paper

--- 2023-03-14 Write-ups/Manuscript.tex
@@
 \usepackage{setspace}\doublespacing
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \ifLuaTeX
@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 14, 2023}
+\date{March 17, 2023}
 
@@
 \(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
-
 The second metric, AUC, focuses on evaluation of prediction of the
@@
 un-binned grid, we evaluated AUC on this grid as well and estimated
-values of latent functions between bins with linear interpolation. Just
-like ISE, the AUC report is also the average value across the whole
-sample.
+values of latent functions between bins with linear interpolation.
 
@@
 
-\hypertarget{result}{%
-\subsection{Result}\label{result}}
+The average ISE and AUC across all simulation is presented in Table 1,
+and Figure 1 presents the predicted function curves of four
+randomly-drawn subjects from the first simulation. The prediction is
+conditioning on different length of observed track (specifically, with
+observations up to t = 200, 400, 600, 800 respectively), and evaluation
+is made on equal-length time windows on the unobserved tracks following
+the maximum observation time. As Table 1 reveals, the fGFPCA
+outperformed GLMMadaptive under every scenario, which was expected since
+GLMMadaptive can only accomodate simple model structures. In addition,
+fGFPCA also takes less computation time. For one simulated dataset,
+fGFPCA spent 4.94 minutes on model fitting and out-of-sample prediction,
+while GLMMadaptive took 5.11 minutes.
 
-\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}
+Because of the flexibility of our proposed model framework, the accuracy
+of prediction at specific time points would improve with more observed
+data. This is revealed in Table 1 as ISE decreases and AUC increases
+with maximum observed time (left-to-right), also in Figure 1 as
+predicted curves get closer with longer observed track. However the same
+tendency is not observed with GLMMadaptive models, as a result of
+restricted model flexibility. A linear model that fits well to a
+specific part of latent function can fit very badly to the following
+parts, especially when the underlying latent function has cyclic
+patterns. Therefore, more observations do not necessarily make the model
+more predictive.
 
-\begin{verbatim}
-## [1] 5.033985 5.134139 5.195642 5.152902 5.082466
-\end{verbatim}
+\begin{figure}
+\centering
+\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
+\caption{Predicted track of four randomly selected subjects. The grey
+dash line indicates the true latent continuous function. The dashed
+lines indicated predicted latent function tracks, and color indicates
+different observations time.}
+\end{figure}
+
+\begin{table}
+
+\caption{\label{tab:format table}Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.}
+\centering
+\begin{tabular}[t]{llccccccc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observed time} \\
+\cmidrule(l{3pt}r{3pt}){2-9}
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
+  & 200 & 400 & 600 & 800 & 200 & 400 & 600 & 800\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{9}{l}{\textbf{Prediction time window}}\\
+\addlinespace[0.3em]
+\multicolumn{9}{l}{\textbf{ISE}}\\
+\hspace{1em}\hspace{1em}(200, 400] & 15.29 &  &  &  & 38.20 &  &  & \\
+\hspace{1em}\hspace{1em}(400, 600] & 18.68 & 8.53 &  &  & 28.59 & 26.46 &  & \\
+\hspace{1em}\hspace{1em}(600, 800] & 22.53 & 5.51 & 1.65 &  & 31.32 & 27.97 & 27.40 & \\
+\hspace{1em}\hspace{1em}(800, 1000] & 11.34 & 8.72 & 1.90 & 1.30 & 55.19 & 46.89 & 58.57 & 59.25\\
+\addlinespace[0.3em]
+\multicolumn{9}{l}{\textbf{AUC}}\\
+\hspace{1em}\hspace{1em}(200, 400] & 0.74 &  &  &  & 0.59 &  &  & \\
+\hspace{1em}\hspace{1em}(400, 600] & 0.66 & 0.73 &  &  & 0.52 & 0.59 &  & \\
+\hspace{1em}\hspace{1em}(600, 800] & 0.71 & 0.79 & 0.80 &  & 0.67 & 0.70 & 0.69 & \\
+\hspace{1em}\hspace{1em}(800, 1000] & 0.74 & 0.75 & 0.78 & 0.78 & 0.52 & 0.56 & 0.53 & 0.57\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2023-03-30 Write-ups/Manuscript.Rmd
data application use daily weigh-in data

@@
   - \usepackage{setspace}\doublespacing
-
+  - \usepackage[utf8]{inputenc}
+  - \usepackage[T1]{fontenc}
 ---
@@
 library(knitr)
+library(ggpubr)
 options(knitr.kable.NA = "")
@@
 
+# simulation
 load(here("Data/sim_data.RData"))
@@
 load(here("Data/SimOutput_GLMMadaptive.RData"))
+
+# data application
+df_appl <- read.csv(here("Data/Daily_Weigh_ins.csv"))
+load(here("Data/ApplOutput_fGFPCA.RData"))
+load(here("Data/ApplOutput_GLMMadaptive.RData"))
+
 ```
@@
 =================================================================================================================================
-
-<!-- -	Overview of dynamic prediction methods -->
 
@@
 
-To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain.
-Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (fPCA).
+To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures.
+Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (FPCA).
 The introduction of basis functions also makes out-of-sample prediction more straightforward.
@@
 While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors.
-However, little extension was made on prediction of non-Gaussian functions, such as binary and count outcomes.
 
@@
 
-In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with fPCA smoothing.
+In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with FPCA smoothing.
 Section 2 presents the procedure of the proposed method.
@@
 In Section 4, we apply this method to a real-world dataset.
-Section 5 presents a discussion of davantages and limitation of the proposed method.
+Section 5 presents a discussion of advantages and limitation of the proposed method.
 
@@
 
-where g is a appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function follows a zero-mean Gaussian process.
+where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process.
 
-While $b_i(t)$ is not observed, it can be approximated under the FPCA framework $b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)$.
-Here $\phi_{ik}(t)$ is a set of orthogonal eigenfunctions that explains the most variation in $b_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction.
-Additionally, $\xi_{ik}$ are mutually independent obver both subject ($i$) and eigenfunctions ($k$).
-That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues: $\int \phi_k^2(t)dt=\lambda_k$.
-$\epsilon_i(t)$ here is a residual function that accounts for the variation not explained by the first K eigenfunctions from FPCA.
-We assume it follows a zero-mean Gaussian process.
-At a specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$.
+We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation, following a zero-mean Gaussian process.
+At each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
 
-Based on the problem set up above, we propose the following algorithm for PFCA on the unobserved latent process $\eta_i(t)$:
+In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$.
+In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction.
+Additionally, $\xi_{ik}$ are mutually independent over both subject ($i$) and eigenfunctions ($k$).
+That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues.
+
+**Fast Generalized FPCA** Based on the problem set up above, we propose the following algorithm for fast, generalizable implementation of FPCA (fGFPCA) on the unobserved latent process $\eta_i(t)$:
 
@@
 Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$.
-From this series of models we can get estiamtes of population mean $\hat{\beta}_0(s)$ and subject-level random effect $\hat{b}_i(s)$, thus estimate of the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$.
+From this series of models we can estimate the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$.
 3.
-Fit FPCA on the estimated latent functions $\hat{\eta}_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, population mean $\hat{\beta}_0(s)$ and residual variance $\hat{\sigma}^2$.
-4.
-With components extracted above, calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$ of new samples based on their partially observed data.
-Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$
+Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
+
+**Dynamic prediction** Now assume we have a new observations with partially observed outcome $Y_{new}(t)$, $t\leq t_m$.
+The maximum observed point is $t_m$, which is a point along the functional domain and belongs to bin $s_m$ on the binned grid from step 1 above.
+With components extracted from the FPCA model in step 3, we will be able to make predictions on the future, unobserved bins.
+Specifically, with the partially observed binary function $Y_{new}(t)$, we will be able to calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$.
+Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$
 
 Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects.
-Since the bins are set up to be small in length, the binned grid would still be dense enough.
-However, in situations where we need predictions on the original, un-binned grid instead, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.
+Since the bins are set up to be small in length, the binned grid would still be dense.
+However, in situations where predictions on the original, un-binned grid is needed, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.
 
-Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework.
+<!-- Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework.
 In step 4 we calculated the MLE of $\hat{\xi}_{ik}$.
 Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$.
-Therefore, the variation of prediction interval is:
+Therefore, the variation of prediction interval is: -->
 
-$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
-Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.
+<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
+<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->
 
-<!-- -	Need better notation system -->
-<!-- - Change to general exponential family -->
 <!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->
@@
 We simulated 50 datasets, each with 500 subjects.
-For every subject, we generate 1000 binary outcomes $Y_i(t) \in (0, 1)$ across functional domain $t \in [0, 1]$, where the distribution of outcome is characterized by a continuous latent function.
+For every subject, we generate 1000 binary outcomes $Y_i(t) \in \{0, 1\}$ across functional domain $t \in [0, 1000]$, where the distribution of outcome is characterized by a continuous latent function.
 The data generation mechanism can be expressed as follows:
@@
 Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
-\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
+\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
 \end{aligned}$$
@@
 Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$.
-In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1]$ and are the same for all subjects.
+In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1000]$ and are the same for all subjects.
 
 We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC).
-ISE assess the prediction accucay of latent continuous function.
+ISE assess the prediction accuracy of latent continuous function.
 It is evaluated on the binned grid at midpoints of each unobserved bin.
-If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.
+If the entire functional domain has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.
 The second metric, AUC, focuses on evaluation of prediction of the binary outcome.
-Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation.
+Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid and estimated values of latent functions between bins with linear interpolation.
 
@@
 The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time.
-As Table 1 reveals, the fGFPCA outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accomodate simple model structures.
-In addition, fGFPCA also takes less computation time.
+As Table 1 reveals, our proposed method (fGFPCA) outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accommodate simple model structures.
+In addition, fGFPCA also took less computation time.
 For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.
@@
 ```{r}
-
 # other quantities
@@
 
-```{r Figure, fig.cap="Predicted track of four randomly selected subjects.
-The grey dash line indicates the true latent continuous function.
-The dashed lines indicated predicted latent function tracks, and color indicates different observations time."}
+```{r Figure, fig.cap="Predicted track of four randomly selected subjects from one simulated dataset.
+The grey solid line indicates the true latent continuous function.
+The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
 # figure of fGFPCA prediction
@@
 
+exp_df_sim1_ref <- pred_lst_ref[[1]]
+
 # randomly draw four samples for visualization
@@
 
-exp_df_sim1 %>%
+# fGFPCA
+fig1a <- exp_df_sim1 %>%
   left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
@@
   facet_wrap(~id, scales = "free_y")+
-  labs(y = "Latent function", x = "Bin", colour= "Observed time")+
+  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "fGFPCA")+
   scale_color_manual(values=cbPalette)+
   scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
+
+# GLMMadaptive
+fig1b <- exp_df_sim1_ref %>%
+  filter(sind_inx %in% mid) %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
+  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
+  facet_wrap(~id, scales = "free_y")+
+  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "GLMMadaptive")+
+  scale_color_manual(values=cbPalette)+
+  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
+
+ggarrange(fig1a, fig1b, nrow = 1, common.legend = T)
+
 ```
@@
 
+```{r}
+# data clean
+mid2 <- unique(df_est_latent$bin)
+N2 <- max(df_est_latent$id)
+J2 <- 400
+
+# for fGFPCA
+# code prediction value at observed points as NA
+df_est_latent$pred_t100[df_est_latent$bin <= 100] <- NA
+df_est_latent$pred_t200[df_est_latent$bin <= 200] <- NA
+df_est_latent$pred_t300[df_est_latent$bin <= 300] <- NA
+
+# for GLMM adaptive
+# add in observation time index
+for(m in 1:length(pred_lst_ref)){
+  pred_lst_ref[[m]]$sind_inx <- rep(1:J, N)
+}
+```
+
+```{r AUC_fGFPCA_Appl, cache=TRUE}
+# maximum observations time
+t_vec2 <- c(100, 200, 300)
+
+# calculate AUC on unbinned grid
+auc_mat2 <- matrix(NA, nrow = 3, ncol = 3) # row index iteration, col index max observation time
+
+# result container for interpolation
+Y_df2 <- df_appl %>% select(participant_id, starts_with("day")) %>% 
+  pivot_longer(2:401, values_to = "Y", names_to = "sind_inx") %>%
+  mutate(sind_inx = gsub("day_", "", sind_inx)) %>%
+  mutate(sind_inx = as.numeric(sind_inx)) %>%
+  rename("id" = "participant_id")
+Y_df2$pred_100 <- Y_df2$pred_200 <- Y_df2$pred_300 <- NA
+
+  # use interpolation to extend prediction to the unbinned grid
+for(i in 1:N2){
+   Y_df2$pred_100[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
+                                         y=df_est_latent$pred_t100[df_est_latent$id==i], 
+                                         xout = 1:J2)$y
+   Y_df2$pred_200[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
+                                         y=df_est_latent$pred_t200[df_est_latent$id==i], 
+                                         xout = 1:J2)$y
+   Y_df2$pred_300[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
+                                         y=df_est_latent$pred_t300[df_est_latent$id==i], 
+                                         xout = 1:J2)$y
+  }
+
+# split unobserved time into intervals of length 100
+Y_df2$bin_cat <- cut(Y_df2$sind_inx, breaks=c(0, 100, 200, 300, 400), include.lowest = T)
+Y_df_lst2 <- split(Y_df2, f = Y_df2$bin_cat)
+Y_df_lst2 <- Y_df_lst2[-1]
+
+for(j in seq_along(t_vec2)){
+
+    t <- t_vec2[j]
+    pred_name <- paste("pred_", t, sep = "")
+
+    # calculate AUC
+    auc_lst <- lapply(Y_df_lst2[j:3], 
+                       function(x){
+                         row_id <- !is.na(x[, pred_name])
+                         this_eta <- x[row_id, pred_name]
+                         this_Y <- x[row_id, "Y"]
+                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
+                         return(auc_pref@y.values)
+                })
+
+    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
+    auc_mat2[ , j] <- auc_vec
+
+}
+
+colnames(auc_mat2) <- c("100", "200", "300")
+rownames(auc_mat2) <- c("(100, 200]", "(200, 300]", "(300, 400]")
+
+```
+
+```{r AUC_GLMMadaptive_Appl}
+# calculate AUC on unbinned grid
+auc_mat_ref2 <- matrix(NA, nrow = 3, ncol = 3)  # row index iteration, col index max observation time
+
+# split unobserved time into intervals of length 100
+df_est_latent_ref$bin_cat <- cut(df_est_latent_ref$sind_inx, 
+                                 breaks=c(0, 100, 200, 300, 400), include.lowest = T)
+Y_df_lst2_ref <- split(df_est_latent_ref, f = df_est_latent_ref$bin_cat)
+Y_df_lst2_ref <- Y_df_lst2_ref[-1]
+
+for(j in seq_along(t_vec2)){
+
+    t <- t_vec2[j]
+    pred_name <- paste("pred_t", t, sep = "")
+
+    # calculate AUC
+    auc_lst <- lapply(Y_df_lst2_ref[j:3], 
+                       function(x){
+                         row_id <- !is.na(x[, pred_name])
+                         this_eta <- x[row_id, pred_name]
+                         this_Y <- x[row_id, "Y"]
+                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
+                         return(auc_pref@y.values)
+                })
+
+    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
+    auc_mat_ref2[ , j] <- auc_vec
+
+  }
+
+colnames(auc_mat_ref2) <- c("100", "200", "300")
+rownames(auc_mat_ref2) <- c("(100, 200]", "(200, 300]", "(300, 400]")
+
+```
+
+```{r, Figure_appl, fig.cap="Predicted track of four randomly selected subjects from the daily weigh-in dataset.
+The grey solid line indicates the true latent continuous function.
+The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
+rand_id2 <- sample(N2, 4)
+fig2a <- Y_df2 %>%
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(pred_100, pred_200, pred_300), function(x)exp(x)/(1+exp(x))) %>%
+  ggplot()+
+  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
+  geom_line(aes(x=sind_inx, y = pred_100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
+  facet_wrap(~id, scales = "free_y")+
+  labs(y = "Predicted latent function on probability scale", x = "Bin", 
+       colour= "Observed time", title = "fGFPCA")+
+  scale_color_manual(values=cbPalette)+
+  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))
+
+fig2b <- df_est_latent_ref %>%
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(pred_t100, pred_t200, pred_t300), function(x)exp(x)/(1+exp(x))) %>%
+  ggplot()+
+  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
+  geom_line(aes(x=sind_inx, y = pred_t100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
+  facet_wrap(~id, scales = "free_y")+
+  labs(y = "Predicted latent function on probability scale", x = "Bin", 
+       colour= "Observed time", title = "GLMMadaptive")+
+  scale_color_manual(values=cbPalette)+
+  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))
+
+ggarrange(fig2a, fig2b, nrow = 1, common.legend = T)
+```
+
+```{r}
+cbind(auc_mat2, auc_mat_ref2) %>%
+  kable(booktabs = T, digit=2,
+        caption = "AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset",
+        align = "lcccccccc") %>%
+  kable_styling() %>%
+  add_header_above(c(" " = 1, "fGFPCA" = 3, "GLMMadaptive" = 3)) %>%
+  add_header_above(c(" " = 1, "Maximum observed time" = 6)) %>% 
+  group_rows(index = c("Prediction time window" = 3))
+
+```
 Discussion

### 2023-03-30 Write-ups/Manuscript.tex
data application use daily weigh-in data

@@
 \usepackage{setspace}\doublespacing
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{booktabs}
@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 17, 2023}
+\date{March 30, 2023}
 
@@
 To address these problems, one may turn to functional mixed effect
-models instead when measures are dense across the domain. Such methods
-accommodate more flexible correlation structure by modeling
+models instead when measures are dense across the domain of measures.
+Such methods accommodate more flexible correlation structure by modeling
 subject-specific random effects as a function, and non-parametric
@@
 functions or eigenfunctions from functional principal component analysis
-(fPCA). The introduction of basis functions also makes out-of-sample
+(FPCA). The introduction of basis functions also makes out-of-sample
 prediction more straightforward. Instead of estimating subject-specific
@@
 Functional Concurrent Regression (FCR) framework which can incorporate
-the effect of subject-specific predictors. However, little extension was
-made on prediction of non-Gaussian functions, such as binary and count
-outcomes.
+the effect of subject-specific predictors.
 
@@
 prediction of discrete function tracks based on functional mixed effect
-model with fPCA smoothing. Section 2 presents the procedure of the
+model with FPCA smoothing. Section 2 presents the procedure of the
 proposed method. In Section 3, we illustrate the performance and
@@
 we apply this method to a real-world dataset. Section 5 presents a
-discussion of davantages and limitation of the proposed method.
+discussion of advantages and limitation of the proposed method.
 
@@
 
-where g is a appropriate link function, \(\beta_0(t)\) is the population
-mean of latent function, and \(b_i(t)\) is a subjects-level random
-effect function follows a zero-mean Gaussian process.
+where g is an appropriate link function, \(\beta_0(t)\) is the
+population mean of latent function, and \(b_i(t)\) is a subjects-level
+random effect function following a zero-mean Gaussian process.
 
-While \(b_i(t)\) is not observed, it can be approximated under the FPCA
-framework \(b_i(t)=\sum_{k=1}^K\xi_{ik}\phi_{ik}(t)+\epsilon_i(t)\).
-Here \(\phi_{ik}(t)\) is a set of orthogonal eigenfunctions that
-explains the most variation in \(b_i(t)\), and \(\xi_{ik}\) are
+According to the Karhunen-Loève theorem, the unobserved random effect
+function \(b_i(t)\) can be represented as the linear combination of a
+infinite set of orthogonal functions:
+\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
+function by a finite set of \(\phi_{k}(t)\), such that
+\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
+\(\epsilon_i(t)\) indicates random noise introduced by this
+approximation, following a zero-mean Gaussian process. At each specific
+point t, \(\epsilon_i(t) \sim N(0, \sigma^2)\).
+
+In practice, this approximation could be implemented through an FPCA
+procedure on \(\eta_i(t)\). In this case
+\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
+explain the most variation in \(\eta_i(t)\), and \(\xi_{ik}\) are
 subject-specific PC scores (or loadings) on each eigenfunction.
-Additionally, \(\xi_{ik}\) are mutually independent obver both subject
+Additionally, \(\xi_{ik}\) are mutually independent over both subject
 (\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
 normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
-eigenvalues: \(\int \phi_k^2(t)dt=\lambda_k\). \(\epsilon_i(t)\) here is
-a residual function that accounts for the variation not explained by the
-first K eigenfunctions from FPCA. We assume it follows a zero-mean
-Gaussian process. At a specific point t,
-\(\epsilon_i(t) \sim N(0, \sigma^2)\).
+eigenvalues.
 
-Based on the problem set up above, we propose the following algorithm
-for PFCA on the unobserved latent process \(\eta_i(t)\):
+\textbf{Fast Generalized FPCA} Based on the problem set up above, we
+propose the following algorithm for fast, generalizable implementation
+of FPCA (fGFPCA) on the unobserved latent process \(\eta_i(t)\):
 
@@
   \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
-  bin \(s\). From this series of models we can get estiamtes of
-  population mean \(\hat{\beta}_0(s)\) and subject-level random effect
-  \(\hat{b}_i(s)\), thus estimate of the individual latent functions at
-  every bin: \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
+  bin \(s\). From this series of models we can estimate the individual
+  latent functions at every bin:
+  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
 \item
-  Fit FPCA on the estimated latent functions \(\hat{\eta}_i(s)\), and
-  obtain estimates of basis functions
+  Fit FPCA on the estimated latent functions:
+  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
+  and obtain estimates of basis functions
   \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
-  \(\hat{\lambda}_1...\hat{\lambda}_k\), population mean
-  \(\hat{\beta}_0(s)\) and residual variance \(\hat{\sigma}^2\).
-\item
-  With components extracted above, calculate the maximum likelihood
-  estimate (MLE) of the subject-specific PC scores \(\hat{\xi}_{ik}\) of
-  new samples based on their partially observed data. Then the value of
-  latent functions at unobserved points can be estimated as
-  \(\hat{\eta_i}(s)=\hat{\beta}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)
+  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
+  and residual variance \(\hat{\sigma}^2\).
 \end{enumerate}
+
+\textbf{Dynamic prediction} Now assume we have a new observations with
+partially observed outcome \(Y_{new}(t)\), \(t\leq t_m\). The maximum
+observed point is \(t_m\), which is a point along the functional domain
+and belongs to bin \(s_m\) on the binned grid from step 1 above. With
+components extracted from the FPCA model in step 3, we will be able to
+make predictions on the future, unobserved bins. Specifically, with the
+partially observed binary function \(Y_{new}(t)\), we will be able to
+calculate the maximum likelihood estimate (MLE) of the subject-specific
+PC scores \(\hat{\xi}_{ik}\). Then the value of latent functions at
+unobserved points can be estimated as
+\(\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)
 
@@
 non-Gaussian functions tracks of new subjects. Since the bins are set up
-to be small in length, the binned grid would still be dense enough.
-However, in situations where we need predictions on the original,
-un-binned grid instead, linear interpolation turns out to be a fast,
-convenient way with good performance for prediction at points between
-bins.
-
-Precision of prediction, usually quantified by the variance of
-prediction error \(Var(\hat{\eta_i}(s)-\eta_i(s))\), is also
-straightforward under this framework. In step 4 we calculated the MLE of
-\(\hat{\xi}_{ik}\). Based on likelihood theory, its variance can be
-estimated with observed information \({I}(\hat{\xi}_{ik})\), which is
-essentially the second derivative of likelihood at \(\hat{\xi}_{ik}\).
-Therefore, the variation of prediction interval is:
-
-\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
-Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
-\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).
+to be small in length, the binned grid would still be dense. However, in
+situations where predictions on the original, un-binned grid is needed,
+linear interpolation turns out to be a fast, convenient way with good
+performance for prediction at points between bins.
 
@@
 study. We simulated 50 datasets, each with 500 subjects. For every
-subject, we generate 1000 binary outcomes \(Y_i(t) \in (0, 1)\) across
-functional domain \(t \in [0, 1]\), where the distribution of outcome is
-characterized by a continuous latent function. The data generation
+subject, we generate 1000 binary outcomes \(Y_i(t) \in \{0, 1\}\) across
+functional domain \(t \in [0, 1000]\), where the distribution of outcome
+is characterized by a continuous latent function. The data generation
 mechanism can be expressed as follows:
@@
 Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
-\eta_i(t) &= f_0(t)+ \xi_{i1}sin(2\pi t)+\xi_{i2}cos(2\pi t)+\xi_{i3}sin(4\pi t)+\xi_{i4}cos(4\pi t)
+\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
 \end{aligned}\]
@@
 generate data on a regular grid, which means observations points are
-equally distributed across \([0, 1]\) and are the same for all subjects.
+equally distributed across \([0, 1000]\) and are the same for all
+subjects.
 
@@
 Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
-accucay of latent continuous function. It is evaluated on the binned
+accuracy of latent continuous function. It is evaluated on the binned
 grid at midpoints of each unobserved bin. If the entire functional
-domian has S bins, but we have observations up to the mth bin, then ISE
+domain has S bins, but we have observations up to the mth bin, then ISE
 is defined as
-\(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
+\(\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
 The second metric, AUC, focuses on evaluation of prediction of the
 binary outcome. Since the binary outcomes are generated on the original,
-un-binned grid, we evaluated AUC on this grid as well and estimated
-values of latent functions between bins with linear interpolation.
+un-binned grid, we evaluated AUC on this grid and estimated values of
+latent functions between bins with linear interpolation.
 
@@
 is made on equal-length time windows on the unobserved tracks following
-the maximum observation time. As Table 1 reveals, the fGFPCA
-outperformed GLMMadaptive under every scenario, which was expected since
-GLMMadaptive can only accomodate simple model structures. In addition,
-fGFPCA also takes less computation time. For one simulated dataset,
-fGFPCA spent 4.94 minutes on model fitting and out-of-sample prediction,
-while GLMMadaptive took 5.11 minutes.
+the maximum observation time. As Table 1 reveals, our proposed method
+(fGFPCA) outperformed GLMMadaptive under every scenario, which was
+expected since GLMMadaptive can only accommodate simple model
+structures. In addition, fGFPCA also took less computation time. For one
+simulated dataset, fGFPCA spent 4.94 minutes on model fitting and
+out-of-sample prediction, while GLMMadaptive took 5.11 minutes.
 
@@
 \includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
-\caption{Predicted track of four randomly selected subjects. The grey
-dash line indicates the true latent continuous function. The dashed
-lines indicated predicted latent function tracks, and color indicates
-different observations time.}
+\caption{Predicted track of four randomly selected subjects from one
+simulated dataset. The grey solid line indicates the true latent
+continuous function. The dashed lines indicated predicted latent
+function tracks, and color indicates different observation time.}
 \end{figure}
@@
 \section{Data application}\label{data-application}}
+
+\begin{figure}
+\centering
+\includegraphics{Manuscript_files/figure-latex/Figure_appl-1.pdf}
+\caption{Predicted track of four randomly selected subjects from the
+daily weigh-in dataset. The grey solid line indicates the true latent
+continuous function. The dashed lines indicated predicted latent
+function tracks, and color indicates different observation time.}
+\end{figure}
+
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-3}AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset}
+\centering
+\begin{tabular}[t]{llccccc}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Maximum observed time} \\
+\cmidrule(l{3pt}r{3pt}){2-7}
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{fGFPCA} & \multicolumn{3}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
+  & 100 & 200 & 300 & 100 & 200 & 300\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{7}{l}{\textbf{Prediction time window}}\\
+\hspace{1em}(100, 200] & 0.75 &  &  & 0.76 &  & \\
+\hspace{1em}(200, 300] & 0.77 & 0.82 &  & 0.76 & 0.80 & \\
+\hspace{1em}(300, 400] & 0.76 & 0.81 & 0.84 & 0.74 & 0.78 & 0.82\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2023-03-30 Code/DataAppl.R

@@ -1,0 +1,125 @@
+# This script runs the data application section
+
+set.seed(330)
+
+library(here)
+library(here)
+library(tidyverse)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+
+# code
+source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
+source(here("Code/OutSampMLE.R"))
+
+# data
+df <- read.csv("Data/Daily_Weigh_ins.csv")
+
+#### Data clean ####
+
+J <- ncol(df %>% select(starts_with("day")))  # number of observations points
+N <- nrow(df) # sample size
+rand_id <- sample(N, 4) # sample used to visualization
+
+
+# move from wide to long
+df_long <- df %>% select(participant_id, starts_with("day")) %>% 
+  pivot_longer(2:401, values_to = "Y", names_to = "sind_inx") %>%
+  mutate(sind_inx = gsub("day_", "", sind_inx)) %>%
+  mutate(sind_inx = as.numeric(sind_inx)) %>%
+  rename("id" = "participant_id")
+
+df_long %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind_inx, y=Y))+
+  facet_wrap(~id)
+
+
+##### fGFPCA #####
+
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+df_long$bin <- cut(df_long$sind_inx, breaks = brks, include.lowest = T, labels = mid)
+df_long$bin <- as.numeric(as.character(df_long$bin))
+
+df_bin_lst <- split(df_long, f = df_long$bin)
+
+lapply(df_bin_lst, function(x)table(x$Y))
+
+# fit local GLMM and estimate latent function
+# unidentifiability issues !!!
+df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+View(df_est_latent[[1]])
+df_est_latent <- bind_rows(df_est_latent) %>% select(id, bin, eta_hat) %>% distinct() # on the binned grid, one unique value each bin
+df_est_latent %>%
+  left_join(df_long, by = c("id", "bin")) %>%
+  filter(id %in% rand_id) %>%
[…]

### 2023-03-30 Code/OutSampMLE.R

@@ -41,8 +41,8 @@ llh_div <- function(xi, df_new, fpca_fit){
 
 #### Prediction with MLE scores #####
 
-out_samp_dyn_pred <- function(df_new, fpca_fit){
-  score_out <- multiroot(f=llh_div, start = rep(0, 4), df_new=df_new, fpca_fit=fpca_fit)$root
+out_samp_dyn_pred <- function(df_new, fpca_fit, K){
+  score_out <- multiroot(f=llh_div, start = rep(0, K), df_new=df_new, fpca_fit=fpca_fit)$root
   eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions%*%score_out
   
   

### 2023-03-31 Write-ups/Manuscript.Rmd
data application

@@
 Therefore, more observations do not necessarily make the model more predictive.
+Since the latent Gaussian function tracks are not known in this case, prediction performance on evaluated with AUC along.
+
+Table 2 compares the predictive performance of both models.
+When the observed time is short, the two method performend similarly when predicting recent days, however fGFPCA did better in predicting further days along the trial.
+As more observations coming in, fGFPCA sees increasingly higher AUC compared to GLMMadaptive model.
+This could be a result of the difference of model flexibility of the two methods.
+While the linear relationship in underlying latent functions are not likely to change immediately following the observed track, the whole track may need to be much more flexible, which is a better fit for fGFPCA model.
+This possible explanation is illustrated in Figure 2 using predicted latent function tracks of four randomly drawn subjects.
 
@@
 
+We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@].
+The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing.
+This record was then scaled into a binary indicator of participant adherence to the daily-weigh-in procedure, where 1 indicates participants followed the requirement and 0 otherwise.
+In this section, we will use this series of daily indicators as the binary functional outcome, and use the proposed method to dynamically predict participant adherence on future days given historical track.
+
+The dataset includes 55 participants aged 22-56 at baseline.
+Adherence to daily weigh-in is availabe four 400 consecutive days for all participants.
+In other words, for each individual, 400 binary measuresments were taken on a regular grid along the functional domain (number of days into the study).
+We will fit both fGFPCA and GLMMadaptive models and compared their performance on this data.
+For fGFPCA, every 10 observations are binned together, resulting in 40 bins for final model fitting.
+Linear interpolation was used for extension back to the unbinned grid.
+
 ```{r}

### 2023-03-31 Write-ups/Manuscript.tex
data application

@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 30, 2023}
+\date{March 31, 2023}
 
@@
 
+We will illustrate the performance of our proposed method using data
+from a randomized trial on weight loss at University of Colorado
+Anschutz Medical Campus {[}@{]}. The study originally aimed to study the
+effect of dietary strategies on weight loss, and participants were given
+a smart scale for daily at-home weighing. This record was then scaled
+into a binary indicator of participant adherence to the daily-weigh-in
+procedure, where 1 indicates participants followed the requirement and 0
+otherwise. In this section, we will use this series of daily indicators
+as the binary functional outcome, and use the proposed method to
+dynamically predict participant adherence on future days given
+historical track.
+
+The dataset includes 55 participants aged 22-56 at baseline. Adherence
+to daily weigh-in is availabe four 400 consecutive days for all
+participants. In other words, for each individual, 400 binary
+measuresments were taken on a regular grid along the functional domain
+(number of days into the study). We will fit both fGFPCA and
+GLMMadaptive models and compared their performance on this data. For
+fGFPCA, every 10 observations are binned together, resulting in 40 bins
+for final model fitting. Linear interpolation was used for extension
+back to the unbinned grid.
+
 \begin{figure}

### 2023-03-31 Code/DataAppl.R

@@ -54,9 +54,8 @@ df_bin_lst <- split(df_long, f = df_long$bin)
 lapply(df_bin_lst, function(x)table(x$Y))
 
 # fit local GLMM and estimate latent function
-# unidentifiability issues !!!
+# near-unidentifiability issues !!!
 df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
-View(df_est_latent[[1]])
 df_est_latent <- bind_rows(df_est_latent) %>% select(id, bin, eta_hat) %>% distinct() # on the binned grid, one unique value each bin
 df_est_latent %>%
   left_join(df_long, by = c("id", "bin")) %>%

### 2023-04-01 Write-ups/Manuscript.Rmd
discussion

@@
 However, in situations where predictions on the original, un-binned grid is needed, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.
-
-<!-- Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework.
-In step 4 we calculated the MLE of $\hat{\xi}_{ik}$.
-Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$.
-Therefore, the variation of prediction interval is: -->
-
-<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
-<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->
-
-<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->
 
@@
 
-```{r AUC_fGFPCA, cache=TRUE}
+```{r AUC_fGFPCA}
 # maximum observations time
@@
 
-We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@].
+We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@bothwell2022; @ostendorf2022].
 The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing.
@@
 
-```{r AUC_fGFPCA_Appl, cache=TRUE}
+```{r AUC_fGFPCA_Appl}
 # maximum observations time
@@
 =================================================================================================================================
--	Grid 
--	Score bias: cannot demonstrate without repeat simulation
+
+The simulation study and case study of daily weigh-in data above have revealed the proposed method can achieve better predictive performance with much less computational burden.
+Compared to existing method based on linear mixed model, fGFPCA can accomodate flexible correlation struture between repeated measure, as well as high density of measurement.
+This is particularly useful when underlying outcome shows cyclic pattern, such as the simulation setting in Section 3.
+While the binary outcomes are generated from a cyclic Gaussian process, fGFPCA has much lower ISE and higher AUC than GLMM.
+Its prediction accuracy also improves as more outcomes are observed, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data.
+
+While the proposed method achieved good accuracy of point prediction, it is more complicated to develop a straightforward, interpretable metric to evaluate the precision of prediction.
+The challenge arises from the fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both introduce uncertainty to the final prediction.
+Since they are implemented consecutively, it is not clear how the uncertainty from both procedures should be integrated into one single interval estimator.
+In practice, one could consider developing interval predictions conditional on the local GLMMs, quantifying uncertainty of individual score estimates from FPCA model alone.
+Since the score $\hat{\xi}_{ik}$ is essentially a maximum likelihood estimate, it is natural to estimate its variance based on likelihood theory with observed information ${I}(\hat{\xi}_{ik})$.
+Therefore, the variation of prediction interval is:
+
+$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
+Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.
+
+The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quanitfy such effects.
+Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid.
+Moreover, if the bins are very narrow with only a few observations, unidentifiability issues can happen when fitting local mixed models.
+While this may not necessarily affect point predictions, it can cause numeric failure to estimate standard error and prediction intervals.
+In practice, we expect this to happen when the measurements are sparse or sample size is small.
+Under such scenarios, one may choose traditional mixed model framework instead to get valid estimates at the expense of computational time.
+
+It should also be noted that the binning procedure in Step 1 changes the density of repeated measures.
+The final prediction of latent continuous function tracks are made on the binned grid, integrating information contained in several observations in the same bin.
+In Section 3 and 4, we have transformed this sparser grid to the original, unbinned grid using linear interpolation of the predicted latent function.
+However, this is by no means the only method to be considered.
+Such transformation can also be made by projection to a higher-dimensional space, similar as general additive model with a large number of basis functions.
+We suspect that this method would work better than linear interpolation when the original grid is sparse or irregular, but take longer time for computation.
+
+<!-- -	Score bias: cannot demonstrate without repeat simulation -->
+
+<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->
 

### 2023-04-01 Write-ups/Manuscript.tex
discussion

@@
 \author{Ying Jin \and Andrew Leroux}
-\date{March 31, 2023}
+\date{April 01, 2023}
 
@@
 patterns. Therefore, more observations do not necessarily make the model
-more predictive.
+more predictive. Since the latent Gaussian function tracks are not known
+in this case, prediction performance on evaluated with AUC along.
+
+Table 2 compares the predictive performance of both models. When the
+observed time is short, the two method performend similarly when
+predicting recent days, however fGFPCA did better in predicting further
+days along the trial. As more observations coming in, fGFPCA sees
+increasingly higher AUC compared to GLMMadaptive model. This could be a
+result of the difference of model flexibility of the two methods. While
+the linear relationship in underlying latent functions are not likely to
+change immediately following the observed track, the whole track may
+need to be much more flexible, which is a better fit for fGFPCA model.
+This possible explanation is illustrated in Figure 2 using predicted
+latent function tracks of four randomly drawn subjects.
 
@@
 from a randomized trial on weight loss at University of Colorado
-Anschutz Medical Campus {[}@{]}. The study originally aimed to study the
-effect of dietary strategies on weight loss, and participants were given
-a smart scale for daily at-home weighing. This record was then scaled
-into a binary indicator of participant adherence to the daily-weigh-in
-procedure, where 1 indicates participants followed the requirement and 0
-otherwise. In this section, we will use this series of daily indicators
-as the binary functional outcome, and use the proposed method to
-dynamically predict participant adherence on future days given
-historical track.
+Anschutz Medical Campus (\protect\hyperlink{ref-bothwell2022}{Bothwell
+et al. 2022}; \protect\hyperlink{ref-ostendorf2022}{Ostendorf et al.
+2022}). The study originally aimed to study the effect of dietary
+strategies on weight loss, and participants were given a smart scale for
+daily at-home weighing. This record was then scaled into a binary
+indicator of participant adherence to the daily-weigh-in procedure,
+where 1 indicates participants followed the requirement and 0 otherwise.
+In this section, we will use this series of daily indicators as the
+binary functional outcome, and use the proposed method to dynamically
+predict participant adherence on future days given historical track.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Grid
-\item
-  Score bias: cannot demonstrate without repeat simulation
-\end{itemize}
+The simulation study and case study of daily weigh-in data above have
+revealed the proposed method can achieve better predictive performance
+with much less computational burden. Compared to existing method based
+on linear mixed model, fGFPCA can accomodate flexible correlation
+struture between repeated measure, as well as high density of
+measurement. This is particularly useful when underlying outcome shows
+cyclic pattern, such as the simulation setting in Section 3. While the
+binary outcomes are generated from a cyclic Gaussian process, fGFPCA has
+much lower ISE and higher AUC than GLMM. Its prediction accuracy also
+improves as more outcomes are observed, which is a desirable property
+for dynamic prediction where the datasets would be periodically updated
+with newly collected data.
+
+While the proposed method achieved good accuracy of point prediction, it
+is more complicated to develop a straightforward, interpretable metric
+to evaluate the precision of prediction. The challenge arises from the
+fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both
+introduce uncertainty to the final prediction. Since they are
+implemented consecutively, it is not clear how the uncertainty from both
+procedures should be integrated into one single interval estimator. In
+practice, one could consider developing interval predictions conditional
+on the local GLMMs, quantifying uncertainty of individual score
+estimates from FPCA model alone. Since the score \(\hat{\xi}_{ik}\) is
+essentially a maximum likelihood estimate, it is natural to estimate its
+variance based on likelihood theory with observed information
+\({I}(\hat{\xi}_{ik})\). Therefore, the variation of prediction interval
+is:
+
+\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
+Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
+\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).
+
+The hyperparameters introduced by the binning procedure in Step 1 can
+also affect the final predictive performance, thus further
+investigations are needed to quanitfy such effects. Predictive accuracy
+and computational speed can change with bin width, number of
+observations in each bin, whether the bins overlap with each other, also
+whether the bins are equally spaced along the functional grid. Moreover,
+if the bins are very narrow with only a few observations,
+unidentifiability issues can happen when fitting local mixed models.
+While this may not necessarily affect point predictions, it can cause
+numeric failure to estimate standard error and prediction intervals. In
+practice, we expect this to happen when the measurements are sparse or
+sample size is small. Under such scenarios, one may choose traditional
+mixed model framework instead to get valid estimates at the expense of
+computational time.
+
+It should also be noted that the binning procedure in Step 1 changes the
+density of repeated measures. The final prediction of latent continuous
+function tracks are made on the binned grid, integrating information
+contained in several observations in the same bin. In Section 3 and 4,
+we have transformed this sparser grid to the original, unbinned grid
+using linear interpolation of the predicted latent function. However,
+this is by no means the only method to be considered. Such
+transformation can also be made by projection to a higher-dimensional
+space, similar as general additive model with a large number of basis
+functions. We suspect that this method would work better than linear
+interpolation when the original grid is sparse or irregular, but take
+longer time for computation.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-bothwell2022}{}}%
+Bothwell, S., Kaizer, A., Catenacci, V., and Wrobel, J. (2022),
+{``Pattern-based clustering of daily weigh-in trajectories using dynamic
+time warping,''} \emph{Biometrics}.
+\url{https://doi.org/10.1111/biom.13773}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-chen2013}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ostendorf2022}{}}%
+Ostendorf, D. M., Caldwell, A. E., Zaman, A., Pan, Z., Bing, K.,
+Wayland, L. T., Creasy, S. A., Bessesen, D. H., MacLean, P., Melanson,
+E. L., and Catenacci, V. A. (2022), {``Comparison of weight loss induced
+by daily caloric restriction versus intermittent fasting (DRIFT) in
+individuals with obesity: Study protocol for a 52-week randomized
+clinical trial,''} \emph{Trials}, 23, 718.
+\url{https://doi.org/10.1186/s13063-022-06523-2.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-GLMMadaptive}{}}%

### 2023-04-03 Write-ups/Manuscript.Rmd
spell checks

@@
 Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction.
-Traditionally, this type of data has been modeled using marginal models (generalized estimating equations) or conditional models (mixed effect models) [@liang1986; @Laird1982; @lindstrom1990; @davidian2003], and predictions are made based on the correlation between repeated measures from the same subject, and/or covariates that can be either fixed or time-varying.
-However, these methods are limited in terms of flexibility of correlation structure and the ability to handle out-of-sample prediction.
+Traditionally, this type of data has been modeled either marginally, such as generalized estimating equations, or constitutionally on specific subjects, such as mixed effect models [@liang1986; @Laird1982; @lindstrom1990; @davidian2003].
+For subject-specific models, predictions are made based on the correlation between repeated measures from the same subject and covariates that can be either fixed or varying.
+However, most of these methods are based on linear model structure, which is very limited in terms of flexibility of correlation.
 When sample size is large or the density of repeated measures is high, they also tend to cause severe computational burden [@GLMMadaptive].
+In addition, with generalized, non-Gaussian outcomes, these methods lack the ability to handle out-of-sample prediction because there is usually no close-form solution for subject-specific random effects
 
 To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures.
-Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (FPCA).
+Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function.
+Non-parametric smoothing [@Scheipl2014] are often incorporated to speed up the computation, where random effects are approximated by spline basis functions or eigenfunctions from functional principal component analysis (FPCA).
 The introduction of basis functions also makes out-of-sample prediction more straightforward.
 Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing.
-Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
+Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, usually modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
 @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains.
@@
 
-Unfortunately, fewer papers have focused on its extension to non-Gaussian data, such as series of binary or count outcomes.
+Unfortunately, fewer papers have focused on its extension to generalized, non-Gaussian outcomes, such as series of binary or count outcomes.
 Existing methods also tend to be very computationally intensive.
@@
 @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*.
-@gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage, joint estimation strategy.
+@gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage joint estimation strategy.
 @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data.
@@
 
-In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with FPCA smoothing.
-Section 2 presents the procedure of the proposed method.
+In this paper, we aim to develop a fast, scalable method for dynamic prediction of generalized functional outcome based on functional mixed effect model with FPCA smoothing.
+Specifically, "Generalized" indicates that the outcome follows non-Gaussian exponential family distribution, and the term "functional" indicates that repeated measure are collected densely across the study domain (e.g., time).
+Section 2 explains the implementation procedure of the proposed method.
 In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study.
-In Section 4, we apply this method to a real-world dataset.
+In Section 4, we apply this method to a real world dataset from a weight loss study.
 Section 5 presents a discussion of advantages and limitation of the proposed method.
@@
 
-The observed data for a single subject $i$ is ($t$, $Y_i(t)$), where $t$ consists of dense, discrete points along the functional domain, and $Y_i(t)$ is the non-Gaussian outcome observed at $t$.
+The observed data for a single subject $i$ is ($t$, $Y_i(t)$), where $t$ consists of dense, discrete points along the domain of data collection.
+In practice, $t$ is usually a time index along the duration of study.
+We hereafter refer to the range of $t$ as the "functional domain".
+For simplicity, we further assume that outcomes are collected on a regular grid over the functional domain, meaning that $t$ are evenly distributed and does not change across subjects.
+
+$Y_i(t)$ is the non-Gaussian outcome observed at $t$.
 We assume that the outcome $Y_i(t)$ can be characterized by a latent continuous function $\eta_i(t)$.
@@
 According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$.
-We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation, following a zero-mean Gaussian process.
-At each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation.
+This random noise is assumed to follow a zero-mean Gaussian process, such that at each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
 
 In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$.
-In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction.
-Additionally, $\xi_{ik}$ are mutually independent over both subject ($i$) and eigenfunctions ($k$).
-That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues.
+In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$.
+These eigenfunctions, also referred to as Principal Component (PC) functions, are orthogonal are can be orthonormal with proper scaling.
+$\xi_{ik}$ are subject-specific scores (or loadings) on each PC function.
+According to properties of FPCA method, these scores are mutually independent over both subject ($i$) and eigenfunctions ($k$).
+That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$, where $\lambda_k$ is the kth eigenvalues.
 
@@
 
-1. Bin the observed outcomes in to small, non-overlapping, equal length intervals. We hereafter index the bins by their midpoints $s$. 
+1.
+Bin the observed outcomes in to small, non-overlapping, equal length intervals.
+We hereafter index the bins by their midpoints $s$.
+The number of observations from the same subject in bin $s$ is $n_s$, and $t_{sj}$ refers to the jth observation point in bin $s$.
+
 2.
 Fit a local Generalized Mixed Model at every bin.
-Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$.
+Specifically, at bin $s$, we fit $g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)$,  $j\in\{1, ..., n_s\}$.
 From this series of models we can estimate the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$.
+
 3.
@@
 
-Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects.
+Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed tracks on the un-binned grid.
 Since the bins are set up to be small in length, the binned grid would still be dense.
-However, in situations where predictions on the original, un-binned grid is needed, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins.
+However, in situations where predictions on the original, un-binned grid is needed, linear interpolation of estimated latent function tracks turns out to be a fast, convenient way with good performance for prediction at points between bins.
 
@@
 We simulated 50 datasets, each with 500 subjects.
-For every subject, we generate 1000 binary outcomes $Y_i(t) \in \{0, 1\}$ across functional domain $t \in [0, 1000]$, where the distribution of outcome is characterized by a continuous latent function.
+For every subject, we generate 1000 binary outcomes $Y_i(t) \in \{0, 1\}$ across functional domain $t \in (0, 1000]$, where the distribution of outcome is characterized by a continuous latent function.
 The data generation mechanism can be expressed as follows:
@@
 In this simulation, we set $f_0(t) = 0$.
-$\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \gamma_k)$.
-Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$.
-In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1000]$ and are the same for all subjects.
+$\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \lambda_k)$.
+Here we set the values of $\lambda_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$.
+In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $(0, 1000]$ and are the same for all subjects.
 
 We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC).
-ISE assess the prediction accuracy of latent continuous function.
+ISE assesses the prediction accuracy of latent continuous functions.
 It is evaluated on the binned grid at midpoints of each unobserved bin.
-If the entire functional domain has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.
+With observations up to bin $s_m$, ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s>s_m} (\hat{\eta}_i(s)-\eta_i(s))^2$.
 The second metric, AUC, focuses on evaluation of prediction of the binary outcome.
-Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid and estimated values of latent functions between bins with linear interpolation.
+Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid by estimating values of latent functions between bins with linear interpolation.
 
@@
 
-The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation.
-The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time.
-As Table 1 reveals, our proposed method (fGFPCA) outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accommodate simple model structures.
+The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted latent function of four randomly selected subjects from the first simulated dataset.
+The prediction is made conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time.
+As Table 1 reveals, the proposed method (fGFPCA) outperforms GLMMadaptive under every scenario, which was expected since the linear model structure of GLMM is too simple for the cyclic data generation mechanism.
 In addition, fGFPCA also took less computation time.
-For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.
+For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction overall, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.
 
-Because of the flexibility of our proposed model framework, the accuracy of prediction at specific time points would improve with more observed data.
-This is revealed in Table 1 as ISE decreases and AUC increases with maximum observed time (left-to-right), also in Figure 1 as predicted curves get closer with longer observed track.
+Because of the flexibility of our proposed model framework, the accuracy of prediction at specific time points would improve as more observed data is collected.
+This is revealed in Table 1 as ISE decreases and AUC increases with maximum observed time (left-to-right), also in Figure 1 as predicted curves get closer to the true latent function with longer observed track.
 However the same tendency is not observed with GLMMadaptive models, as a result of restricted model flexibility.
-A linear model that fits well to a specific part of latent function can fit very badly to the following parts, especially when the underlying latent function has cyclic patterns.
+A linear model that fits well to a specific part can fit very badly to other parts along the underlying latent function tracks, especially when the shapes of functions are wiggly or cyclic.
 Therefore, more observations do not necessarily make the model more predictive.
-Since the latent Gaussian function tracks are not known in this case, prediction performance on evaluated with AUC along.
-
-Table 2 compares the predictive performance of both models.
-When the observed time is short, the two method performend similarly when predicting recent days, however fGFPCA did better in predicting further days along the trial.
-As more observations coming in, fGFPCA sees increasingly higher AUC compared to GLMMadaptive model.
-This could be a result of the difference of model flexibility of the two methods.
-While the linear relationship in underlying latent functions are not likely to change immediately following the observed track, the whole track may need to be much more flexible, which is a better fit for fGFPCA model.
-This possible explanation is illustrated in Figure 2 using predicted latent function tracks of four randomly drawn subjects.
 
@@
 
-We will illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@bothwell2022; @ostendorf2022].
+In this section, we illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@bothwell2022; @ostendorf2022].
 The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing.
@@
 The dataset includes 55 participants aged 22-56 at baseline.
-Adherence to daily weigh-in is availabe four 400 consecutive days for all participants.
-In other words, for each individual, 400 binary measuresments were taken on a regular grid along the functional domain (number of days into the study).
+Adherence to daily weigh-in is available along 400 consecutive days for all participants.
+In other words, for each individual, 400 binary measurements were taken on a regular grid along the functional domain (number of days into the study).
 We will fit both fGFPCA and GLMMadaptive models and compared their performance on this data.
 For fGFPCA, every 10 observations are binned together, resulting in 40 bins for final model fitting.
-Linear interpolation was used for extension back to the unbinned grid.
+Linear interpolation was used for extension from the binned grid back to the un-binned grid.
+Since the latent continuous function tracks are not known in this case, prediction performance on evaluated with AUC alone.
+
+Table 2 compares the predictive performance of both models.
+Similar to the simulatio study, prediction is made conditioning on observations up to the 100th, 200th and 300th day respectively, and evaluation is made every 100 days following the maximum observation time.
+As is revealed, when the observed time is short (up to 100th day), the two method performed similarly when predicting recent days, however fGFPCA did better in predicting further days along the trial (prediction time window of $(300, 400]$).
+As more observations coming in, fGFPCA sees increasingly higher AUC compared to GLMMadaptive model.
+This could be a result of the difference in model flexibility of the two methods.
+While the linear relationship in underlying latent functions are not likely to change immediately following the observed track, the whole track could be much more complex.
+Therefore, while fGFPCA performs similarly to GLMMadaptive right after the end of observation, it does a much better job capturing the underlying pattern along the whole track.
+Figure 2 further illustrate such difference using predicted latent function tracks of four randomly drawn subjects.
 
@@
 ```
+
 Discussion
@@
 The simulation study and case study of daily weigh-in data above have revealed the proposed method can achieve better predictive performance with much less computational burden.
-Compared to existing method based on linear mixed model, fGFPCA can accomodate flexible correlation struture between repeated measure, as well as high density of measurement.
-This is particularly useful when underlying outcome shows cyclic pattern, such as the simulation setting in Section 3.
-While the binary outcomes are generated from a cyclic Gaussian process, fGFPCA has much lower ISE and higher AUC than GLMM.
-Its prediction accuracy also improves as more outcomes are observed, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data.
+Compared to existing methods based on linear mixed model, fGFPCA can accommodate flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with wiggly or cyclic pattern.
+As in the simulation setting in Section 3, the binary outcomes are generated from a cyclic process.
+fGFPCA has much lower ISE and higher AUC than GLMM in this case, and its prediction accuracy also improves as more outcomes are collected, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data.
+Moreover, fGFPCA also significantly reduce time spent on model fitting and predict, which is beneficial when repeated measures are dense or sample size is large.
+In fact, as far as we are aware of, the proposed method is the only feasible method that can handle the scale of the simulated datasets with compatible flexibility.
 
@@
 
-The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quanitfy such effects.
+The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quantify such effects.
 Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid.
@@
 It should also be noted that the binning procedure in Step 1 changes the density of repeated measures.
-The final prediction of latent continuous function tracks are made on the binned grid, integrating information contained in several observations in the same bin.
-In Section 3 and 4, we have transformed this sparser grid to the original, unbinned grid using linear interpolation of the predicted latent function.
+We essentially assume that latent functions are smooth in small intervals, and treat adjacent observations as if they are generated from the same time point.
+The final prediction of latent continuous function tracks are made on the binned grid, integrating information from several adjacent observations in the same bin.
+In Section 3 and 4, this sparser grid was transformed to the original, un-binned grid using linear interpolation of the predicted latent function.
 However, this is by no means the only method to be considered.

### 2023-04-03 Write-ups/Manuscript.tex
spell checks

@@
 \author{Ying Jin \and Andrew Leroux}
-\date{April 01, 2023}
+\date{April 03, 2023}
 
@@
 referred to as dynamic prediction. Traditionally, this type of data has
-been modeled using marginal models (generalized estimating equations) or
-conditional models (mixed effect models)
-(\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
+been modeled either marginally, such as generalized estimating
+equations, or constitutionally on specific subjects, such as mixed
+effect models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
 \protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
@@
 \protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
-measurement data''} 2003}), and predictions are made based on the
-correlation between repeated measures from the same subject, and/or
-covariates that can be either fixed or time-varying. However, these
-methods are limited in terms of flexibility of correlation structure and
-the ability to handle out-of-sample prediction. When sample size is
+measurement data''} 2003}). For subject-specific models, predictions are
+made based on the correlation between repeated measures from the same
+subject and covariates that can be either fixed or varying. However,
+most of these methods are based on linear model structure, which is very
+limited in terms of flexibility of correlation. When sample size is
 large or the density of repeated measures is high, they also tend to
 cause severe computational burden
-(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}).
+(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). In addition,
+with generalized, non-Gaussian outcomes, these methods lack the ability
+to handle out-of-sample prediction because there is usually no
+close-form solution for subject-specific random effects
 
@@
 Such methods accommodate more flexible correlation structure by modeling
-subject-specific random effects as a function, and non-parametric
-smoothing (\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) can
-be incorporated to speed up the computation, such as spline basis
-functions or eigenfunctions from functional principal component analysis
-(FPCA). The introduction of basis functions also makes out-of-sample
-prediction more straightforward. Instead of estimating subject-specific
-random effects of new observations, we can simply estimate
-coefficients/loadings on the basis function used for smoothing. Existing
-research on dynamic prediction with functional data analysis methods has
-been focusing on continuous/Gaussian outcomes, modelling
-subject-specific random effects with FPCA
+subject-specific random effects as a function. Non-parametric smoothing
+(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) are often
+incorporated to speed up the computation, where random effects are
+approximated by spline basis functions or eigenfunctions from functional
+principal component analysis (FPCA). The introduction of basis functions
+also makes out-of-sample prediction more straightforward. Instead of
+estimating subject-specific random effects of new observations, we can
+simply estimate coefficients/loadings on the basis function used for
+smoothing. Existing research on dynamic prediction with functional data
+analysis methods has been focusing on continuous/Gaussian outcomes,
+usually modelling subject-specific random effects with FPCA
 (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
@@
 Unfortunately, fewer papers have focused on its extension to
-non-Gaussian data, such as series of binary or count outcomes. Existing
-methods also tend to be very computationally intensive. For example,
-Chen et al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches
-to fit marginal functional models that is compatible to multi-level,
+generalized, non-Gaussian outcomes, such as series of binary or count
+outcomes. Existing methods also tend to be very computationally
+intensive. For example, Chen et al.
+(\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
+marginal functional models that is compatible to multi-level,
 generalized outcomes. Goldsmith et al.
@@
 by directly applying FPCA methods to generalized functions, and proposed
-to address this problem using a two-stage, joint estimation strategy.
+to address this problem using a two-stage joint estimation strategy.
 Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
@@
 In this paper, we aim to develop a fast, scalable method for dynamic
-prediction of discrete function tracks based on functional mixed effect
-model with FPCA smoothing. Section 2 presents the procedure of the
-proposed method. In Section 3, we illustrate the performance and
-efficiency of our proposed method in a simulation study. In Section 4,
-we apply this method to a real-world dataset. Section 5 presents a
-discussion of advantages and limitation of the proposed method.
+prediction of generalized functional outcome based on functional mixed
+effect model with FPCA smoothing. Specifically, ``Generalized''
+indicates that the outcome follows non-Gaussian exponential family
+distribution, and the term ``functional'' indicates that repeated
+measure are collected densely across the study domain (e.g., time).
+Section 2 explains the implementation procedure of the proposed method.
+In Section 3, we illustrate the performance and efficiency of our
+proposed method in a simulation study. In Section 4, we apply this
+method to a real world dataset from a weight loss study. Section 5
+presents a discussion of advantages and limitation of the proposed
+method.
 
@@
 The observed data for a single subject \(i\) is (\(t\), \(Y_i(t)\)),
-where \(t\) consists of dense, discrete points along the functional
-domain, and \(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We
-assume that the outcome \(Y_i(t)\) can be characterized by a latent
-continuous function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\)
-follows a exponential family distribution such that:
+where \(t\) consists of dense, discrete points along the domain of data
+collection. In practice, \(t\) is usually a time index along the
+duration of study. We hereafter refer to the range of \(t\) as the
+``functional domain''. For simplicity, we further assume that outcomes
+are collected on a regular grid over the functional domain, meaning that
+\(t\) are evenly distributed and does not change across subjects.
+
+\(Y_i(t)\) is the non-Gaussian outcome observed at \(t\). We assume that
+the outcome \(Y_i(t)\) can be characterized by a latent continuous
+function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows a
+exponential family distribution such that:
 
@@
 \(\epsilon_i(t)\) indicates random noise introduced by this
-approximation, following a zero-mean Gaussian process. At each specific
-point t, \(\epsilon_i(t) \sim N(0, \sigma^2)\).
+approximation. This random noise is assumed to follow a zero-mean
+Gaussian process, such that at each specific point t,
+\(\epsilon_i(t) \sim N(0, \sigma^2)\).
 
@@
 \(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
-explain the most variation in \(\eta_i(t)\), and \(\xi_{ik}\) are
-subject-specific PC scores (or loadings) on each eigenfunction.
-Additionally, \(\xi_{ik}\) are mutually independent over both subject
-(\(i\)) and eigenfunctions (\(k\)). That is, each \(\xi_{ik}\) follows
-normal distribution \(N(0, \lambda_k)\) where \(\lambda_k\) is the kth
-eigenvalues.
+explain the most variation in \(\eta_i(t)\). These eigenfunctions, also
+referred to as Principal Component (PC) functions, are orthogonal are
+can be orthonormal with proper scaling. \(\xi_{ik}\) are
+subject-specific scores (or loadings) on each PC function. According to
+properties of FPCA method, these scores are mutually independent over
+both subject (\(i\)) and eigenfunctions (\(k\)). That is, each
+\(\xi_{ik}\) follows normal distribution \(N(0, \lambda_k)\), where
+\(\lambda_k\) is the kth eigenvalues.
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
   Bin the observed outcomes in to small, non-overlapping, equal length
-  intervals. We hereafter index the bins by their midpoints \(s\).
+  intervals. We hereafter index the bins by their midpoints \(s\). The
+  number of observations from the same subject in bin \(s\) is \(n_s\),
+  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
 \item
   Fit a local Generalized Mixed Model at every bin. Specifically, at bin
-  \(s\), we fit \(g[E(Y_i(t))] = \beta_0(s)+b_i(s)\) for all \(t\) in
-  bin \(s\). From this series of models we can estimate the individual
-  latent functions at every bin:
+  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
+  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
+  the individual latent functions at every bin:
   \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
@@
 Following the algorithm above, predictions of individual latent
-functions are made on the binned grid based on partially observed
-non-Gaussian functions tracks of new subjects. Since the bins are set up
-to be small in length, the binned grid would still be dense. However, in
-situations where predictions on the original, un-binned grid is needed,
-linear interpolation turns out to be a fast, convenient way with good
-performance for prediction at points between bins.
+functions are made on the binned grid based on partially observed tracks
+on the un-binned grid. Since the bins are set up to be small in length,
+the binned grid would still be dense. However, in situations where
+predictions on the original, un-binned grid is needed, linear
+interpolation of estimated latent function tracks turns out to be a
+fast, convenient way with good performance for prediction at points
+between bins.
 
@@
 subject, we generate 1000 binary outcomes \(Y_i(t) \in \{0, 1\}\) across
-functional domain \(t \in [0, 1000]\), where the distribution of outcome
+functional domain \(t \in (0, 1000]\), where the distribution of outcome
 is characterized by a continuous latent function. The data generation
@@
 In this simulation, we set \(f_0(t) = 0\). \(\xi_{ik}\) are mutually
-independent normal random variables \(\xi_{ik}\sim N(0, \gamma_k)\).
-Here we set the values of \(\gamma_k\) to be \(0.5^{k-1}\),
+independent normal random variables \(\xi_{ik}\sim N(0, \lambda_k)\).
+Here we set the values of \(\lambda_k\) to be \(0.5^{k-1}\),
 \(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
 generate data on a regular grid, which means observations points are
-equally distributed across \([0, 1000]\) and are the same for all
+equally distributed across \((0, 1000]\) and are the same for all
 subjects.
@@
 integrated squared error (ISE) and
-Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
-accuracy of latent continuous function. It is evaluated on the binned
-grid at midpoints of each unobserved bin. If the entire functional
-domain has S bins, but we have observations up to the mth bin, then ISE
-is defined as
-\(\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
+Area-Under-the-Receiver-Operator-Curve (AUC). ISE assesses the
+prediction accuracy of latent continuous functions. It is evaluated on
+the binned grid at midpoints of each unobserved bin. With observations
+up to bin \(s_m\), ISE is defined as
+\(\frac{1}{N}\sum_{i=1}^N\sum_{s>s_m} (\hat{\eta}_i(s)-\eta_i(s))^2\).
 The second metric, AUC, focuses on evaluation of prediction of the
 binary outcome. Since the binary outcomes are generated on the original,
-un-binned grid, we evaluated AUC on this grid and estimated values of
+un-binned grid, we evaluated AUC on this grid by estimating values of
 latent functions between bins with linear interpolation.
@@
 The average ISE and AUC across all simulation is presented in Table 1,
-and Figure 1 presents the predicted function curves of four
-randomly-drawn subjects from the first simulation. The prediction is
-conditioning on different length of observed track (specifically, with
-observations up to t = 200, 400, 600, 800 respectively), and evaluation
-is made on equal-length time windows on the unobserved tracks following
-the maximum observation time. As Table 1 reveals, our proposed method
-(fGFPCA) outperformed GLMMadaptive under every scenario, which was
-expected since GLMMadaptive can only accommodate simple model
-structures. In addition, fGFPCA also took less computation time. For one
-simulated dataset, fGFPCA spent 4.94 minutes on model fitting and
-out-of-sample prediction, while GLMMadaptive took 5.11 minutes.
+and Figure 1 presents the predicted latent function of four randomly
+selected subjects from the first simulated dataset. The prediction is
+made conditioning on different length of observed track (specifically,
+with observations up to t = 200, 400, 600, 800 respectively), and
+evaluation is made on equal-length time windows on the unobserved tracks
+following the maximum observation time. As Table 1 reveals, the proposed
+method (fGFPCA) outperforms GLMMadaptive under every scenario, which was
+expected since the linear model structure of GLMM is too simple for the
+cyclic data generation mechanism. In addition, fGFPCA also took less
+computation time. For one simulated dataset, fGFPCA spent 4.94 minutes
+on model fitting and out-of-sample prediction overall, while
+GLMMadaptive took 5.11 minutes.
 
 Because of the flexibility of our proposed model framework, the accuracy
-of prediction at specific time points would improve with more observed
-data. This is revealed in Table 1 as ISE decreases and AUC increases
-with maximum observed time (left-to-right), also in Figure 1 as
-predicted curves get closer with longer observed track. However the same
-tendency is not observed with GLMMadaptive models, as a result of
-restricted model flexibility. A linear model that fits well to a
-specific part of latent function can fit very badly to the following
-parts, especially when the underlying latent function has cyclic
-patterns. Therefore, more observations do not necessarily make the model
-more predictive. Since the latent Gaussian function tracks are not known
-in this case, prediction performance on evaluated with AUC along.
-
-Table 2 compares the predictive performance of both models. When the
-observed time is short, the two method performend similarly when
-predicting recent days, however fGFPCA did better in predicting further
-days along the trial. As more observations coming in, fGFPCA sees
-increasingly higher AUC compared to GLMMadaptive model. This could be a
-result of the difference of model flexibility of the two methods. While
-the linear relationship in underlying latent functions are not likely to
-change immediately following the observed track, the whole track may
-need to be much more flexible, which is a better fit for fGFPCA model.
-This possible explanation is illustrated in Figure 2 using predicted
-latent function tracks of four randomly drawn subjects.
+of prediction at specific time points would improve as more observed
+data is collected. This is revealed in Table 1 as ISE decreases and AUC
+increases with maximum observed time (left-to-right), also in Figure 1
+as predicted curves get closer to the true latent function with longer
+observed track. However the same tendency is not observed with
+GLMMadaptive models, as a result of restricted model flexibility. A
+linear model that fits well to a specific part can fit very badly to
+other parts along the underlying latent function tracks, especially when
+the shapes of functions are wiggly or cyclic. Therefore, more
+observations do not necessarily make the model more predictive.
 
@@
 
-We will illustrate the performance of our proposed method using data
-from a randomized trial on weight loss at University of Colorado
-Anschutz Medical Campus (\protect\hyperlink{ref-bothwell2022}{Bothwell
-et al. 2022}; \protect\hyperlink{ref-ostendorf2022}{Ostendorf et al.
-2022}). The study originally aimed to study the effect of dietary
-strategies on weight loss, and participants were given a smart scale for
-daily at-home weighing. This record was then scaled into a binary
-indicator of participant adherence to the daily-weigh-in procedure,
-where 1 indicates participants followed the requirement and 0 otherwise.
-In this section, we will use this series of daily indicators as the
-binary functional outcome, and use the proposed method to dynamically
-predict participant adherence on future days given historical track.
+In this section, we illustrate the performance of our proposed method
+using data from a randomized trial on weight loss at University of
+Colorado Anschutz Medical Campus
+(\protect\hyperlink{ref-bothwell2022}{Bothwell et al. 2022};
+\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The study
+originally aimed to study the effect of dietary strategies on weight
+loss, and participants were given a smart scale for daily at-home
+weighing. This record was then scaled into a binary indicator of
+participant adherence to the daily-weigh-in procedure, where 1 indicates
+participants followed the requirement and 0 otherwise. In this section,
+we will use this series of daily indicators as the binary functional
+outcome, and use the proposed method to dynamically predict participant
+adherence on future days given historical track.
 
 The dataset includes 55 participants aged 22-56 at baseline. Adherence
-to daily weigh-in is availabe four 400 consecutive days for all
+to daily weigh-in is available along 400 consecutive days for all
 participants. In other words, for each individual, 400 binary
-measuresments were taken on a regular grid along the functional domain
+measurements were taken on a regular grid along the functional domain
 (number of days into the study). We will fit both fGFPCA and
@@
 for final model fitting. Linear interpolation was used for extension
-back to the unbinned grid.
+from the binned grid back to the un-binned grid. Since the latent
+continuous function tracks are not known in this case, prediction
+performance on evaluated with AUC alone.
+
+Table 2 compares the predictive performance of both models. Similar to
+the simulatio study, prediction is made conditioning on observations up
+to the 100th, 200th and 300th day respectively, and evaluation is made
+every 100 days following the maximum observation time. As is revealed,
+when the observed time is short (up to 100th day), the two method
+performed similarly when predicting recent days, however fGFPCA did
+better in predicting further days along the trial (prediction time
+window of \((300, 400]\)). As more observations coming in, fGFPCA sees
+increasingly higher AUC compared to GLMMadaptive model. This could be a
+result of the difference in model flexibility of the two methods. While
+the linear relationship in underlying latent functions are not likely to
+change immediately following the observed track, the whole track could
+be much more complex. Therefore, while fGFPCA performs similarly to
+GLMMadaptive right after the end of observation, it does a much better
+job capturing the underlying pattern along the whole track. Figure 2
+further illustrate such difference using predicted latent function
+tracks of four randomly drawn subjects.
 
@@
 revealed the proposed method can achieve better predictive performance
-with much less computational burden. Compared to existing method based
-on linear mixed model, fGFPCA can accomodate flexible correlation
-struture between repeated measure, as well as high density of
-measurement. This is particularly useful when underlying outcome shows
-cyclic pattern, such as the simulation setting in Section 3. While the
-binary outcomes are generated from a cyclic Gaussian process, fGFPCA has
-much lower ISE and higher AUC than GLMM. Its prediction accuracy also
-improves as more outcomes are observed, which is a desirable property
+with much less computational burden. Compared to existing methods based
+on linear mixed model, fGFPCA can accommodate flexible correlation
+structure between repeated measure, which is particularly useful when
+outcome is generated from latent functions with wiggly or cyclic
+pattern. As in the simulation setting in Section 3, the binary outcomes
+are generated from a cyclic process. fGFPCA has much lower ISE and
+higher AUC than GLMM in this case, and its prediction accuracy also
+improves as more outcomes are collected, which is a desirable property
 for dynamic prediction where the datasets would be periodically updated
-with newly collected data.
+with newly collected data. Moreover, fGFPCA also significantly reduce
+time spent on model fitting and predict, which is beneficial when
+repeated measures are dense or sample size is large. In fact, as far as
+we are aware of, the proposed method is the only feasible method that
+can handle the scale of the simulated datasets with compatible
+flexibility.
 
@@
 also affect the final predictive performance, thus further
-investigations are needed to quanitfy such effects. Predictive accuracy
+investigations are needed to quantify such effects. Predictive accuracy
 and computational speed can change with bin width, number of
@@
 It should also be noted that the binning procedure in Step 1 changes the
-density of repeated measures. The final prediction of latent continuous
-function tracks are made on the binned grid, integrating information
-contained in several observations in the same bin. In Section 3 and 4,
-we have transformed this sparser grid to the original, unbinned grid
-using linear interpolation of the predicted latent function. However,
-this is by no means the only method to be considered. Such
-transformation can also be made by projection to a higher-dimensional
-space, similar as general additive model with a large number of basis
-functions. We suspect that this method would work better than linear
-interpolation when the original grid is sparse or irregular, but take
-longer time for computation.
+density of repeated measures. We essentially assume that latent
+functions are smooth in small intervals, and treat adjacent observations
+as if they are generated from the same time point. The final prediction
+of latent continuous function tracks are made on the binned grid,
+integrating information from several adjacent observations in the same
+bin. In Section 3 and 4, this sparser grid was transformed to the
+original, un-binned grid using linear interpolation of the predicted
+latent function. However, this is by no means the only method to be
+considered. Such transformation can also be made by projection to a
+higher-dimensional space, similar as general additive model with a large
+number of basis functions. We suspect that this method would work better
+than linear interpolation when the original grid is sparse or irregular,
+but take longer time for computation.
 

### 2023-04-04 Write-ups/Manuscript.Rmd
edit method and add new data

@@
 According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$.
-We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation.
-This random noise is assumed to follow a zero-mean Gaussian process, such that at each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)$.
 
@@
 3.
-Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
+Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, where $\epsilon_i(t)$ indicates random noise introduced by the binning process in previous steps.
+This random noise is assumed to follow a zero-mean Gaussian process, such that at each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+The FPCA output includes estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
 

### 2023-04-04 Write-ups/Manuscript.tex
edit method and add new data

@@
 \author{Ying Jin \and Andrew Leroux}
-\date{April 03, 2023}
+\date{April 04, 2023}
 
@@
 function by a finite set of \(\phi_{k}(t)\), such that
-\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
-\(\epsilon_i(t)\) indicates random noise introduced by this
-approximation. This random noise is assumed to follow a zero-mean
-Gaussian process, such that at each specific point t,
-\(\epsilon_i(t) \sim N(0, \sigma^2)\).
+\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)\).
 
@@
   \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
-  and obtain estimates of basis functions
+  where \(\epsilon_i(t)\) indicates random noise introduced by the
+  binning process in previous steps. This random noise is assumed to
+  follow a zero-mean Gaussian process, such that at each specific point
+  t, \(\epsilon_i(t) \sim N(0, \sigma^2)\). The FPCA output includes
+  estimates of basis functions
   \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues

### 2023-04-04 DataApplNHANES.R

@@ -1,0 +1,25 @@
+# data application of NHANES
+
+set.seed(404)
+
+library(here)
+library(tidyverse)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+
+df <- read_rds(here("Data/nhanes_bi.rds"))
+
+N <- length(unique(df$SEQN)) # sample size 8763
+J <- max(df$sind) # 1440 measures for each subject
+
+# visulization
+rand_id <- sample(unique(df$SEQN), size = 4)
+
+df %>% 
+  filter(SEQN %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Z))+
+  facet_wrap(~SEQN)
+    

### 2023-04-06 FDAgroup.Rmd
edit introduction

@@
 
-df = content
-df_train = subset(content, include == 1)
-df_test = subset(content, include == 0)
+# df = content
+# df_train = subset(content, include == 1)
+# df_test = subset(content, include == 0)
 ```
@@
 
-## Simulated child growth data
+<!-- ## Simulated child growth data -->
 
-- Predict length-for-age z-score (LAZ), observed with noise
-  - Calculated using the age- and sex-specific WHO standard references
+<!-- - Predict length-for-age z-score (LAZ), observed with noise -->
+<!--   - Calculated using the age- and sex-specific WHO standard references -->
 
-```{r}
-ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
-  geom_line(show.legend = F)+
-  labs(title = "Length-for-Age z-score tracks for the 10 subjects",
-       x = "time", y ="LAZ")
-```
+<!-- ```{r} -->
+<!-- ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+ -->
+<!--   geom_line(show.legend = F)+ -->
+<!--   labs(title = "Length-for-Age z-score tracks for the 10 subjects", -->
+<!--        x = "time", y ="LAZ") -->
+<!-- ```  -->
 
-## FPCA on observed LAZ
+<!-- ## FPCA on observed LAZ -->
 
-```{r}
-tnew = sort(unique(df_train$argvals))
-df_laz <- df_train %>% select(argvals, subj, Y) %>%
-  rename("y" = "Y")
-fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15,
-                        argvals.new = tnew, calculate.scores = TRUE)
+<!-- ```{r} -->
+<!-- tnew = sort(unique(df_train$argvals)) -->
+<!-- df_laz <- df_train %>% select(argvals, subj, Y) %>% -->
+<!--   rename("y" = "Y") -->
+<!-- fpca_laz <- face.sparse(df_laz, newdata = df_laz, knots = 15, -->
+<!--                         argvals.new = tnew, calculate.scores = TRUE) -->
 
-# mean function
+<!-- # mean function -->
 
-par(mfrow = c(2, 2))
-plot(tnew, fpca_laz$mu.new, type = "l", 
-     xlab = "t", ylab = "f0(t)")
-plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
-plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
-plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")
+<!-- par(mfrow = c(2, 2)) -->
+<!-- plot(tnew, fpca_laz$mu.new, type = "l",  -->
+<!--      xlab = "t", ylab = "f0(t)") -->
+<!-- plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1") -->
+<!-- plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2") -->
+<!-- plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3") -->
 
-```
+<!-- ``` -->
 
-## Prediction of new partially observed sample
+<!-- ## Prediction of new partially observed sample -->
 
-```{r, pred_func}
-# write a function for prediction up to t
-# df_test: datafram with columns argvals, subj, y, observation up to tm
+<!-- ```{r, pred_func} -->
+<!-- # write a function for prediction up to t -->
+<!-- # df_test: datafram with columns argvals, subj, y, observation up to tm -->
 
-DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){
-  Ntest = length(unique(df_test$subj))
-  # prediction
-  pred1 <- predict(fpca_fit, newdata = df_test)
-  # Eigen functions
-  Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ]
-  Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores)
-  # every column is a new function
-  Yhat <- data.frame(Yhat)
-  colnames(Yhat) <- pred1$rand_eff$subj
-  df_pred <- Yhat %>% 
-    mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>% 
-    pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>%
-    mutate(type = "pred")
+<!-- DynPred <- function(df_test = df_test1, tm = 0.3, fpca_fit = fpca_laz){ -->
+<!--   Ntest = length(unique(df_test$subj)) -->
+<!--   # prediction -->
+<!--   pred1 <- predict(fpca_fit, newdata = df_test) -->
+<!--   # Eigen functions -->
+<!--   Bnew <- fpca_fit$eigenfunctions[fpca_fit$argvals.new > tm, ] -->
+<!--   Yhat <- fpca_fit$mu.new[fpca_fit$argvals.new > tm] + Bnew%*%t(pred1$rand_eff$scores) -->
+<!--   # every column is a new function -->
+<!--   Yhat <- data.frame(Yhat) -->
+<!--   colnames(Yhat) <- pred1$rand_eff$subj -->
+<!--   df_pred <- Yhat %>%  -->
+<!--     mutate(argvals = fpca_fit$argvals.new[fpca_fit$argvals.new > tm]) %>%  -->
+<!--     pivot_longer(all_of(1:Ntest), names_to = "subj", values_to = "y") %>% -->
+<!--     mutate(type = "pred") -->
 
-  return(df_pred)
-}
-```
+<!--   return(df_pred) -->
+<!-- } -->
+<!-- ``` -->
 
-```{r t0_3}
-df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y") %>% 
-  filter(argvals <= 0.3))
+<!-- ```{r t0_3} -->
+<!-- df_test_t0.3 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
+<!--   rename("y" = "Y") %>%  -->
+<!--   filter(argvals <= 0.3)) -->
 
-# figure 
-bind_rows(df_test %>% 
-            select(argvals, subj, Y)%>% 
-            rename("y" = "Y") %>% 
-            mutate(type = "true", subj = as.factor(subj)), 
-          df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>% 
-  filter(subj == 152) %>% 
-  ggplot(aes(x = argvals, y = y, col = type))+
-  geom_line()+
-  facet_wrap(~subj)+
-  labs(title = "Prediction with observation up to t=0.3", 
-       x = "time", y = "")
-```
+<!-- # figure  -->
+<!-- bind_rows(df_test %>%  -->
+<!--             select(argvals, subj, Y)%>%  -->
+<!--             rename("y" = "Y") %>%  -->
+<!--             mutate(type = "true", subj = as.factor(subj)),  -->
+<!--           df_test_t0.3 %>% mutate(subj = as.factor(subj))) %>%  -->
+<!--   filter(subj == 152) %>%  -->
+<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
+<!--   geom_line()+ -->
+<!--   facet_wrap(~subj)+ -->
+<!--   labs(title = "Prediction with observation up to t=0.3",  -->
+<!--        x = "time", y = "") -->
+<!-- ``` -->
 
-## Prediction of new partially observed sample
+<!-- ## Prediction of new partially observed sample -->
 
-```{r}
-df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y") %>% 
-  filter(argvals <= 0.5), tm = 0.5)
+<!-- ```{r} -->
+<!-- df_test_t0.5 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
+<!--   rename("y" = "Y") %>%  -->
+<!--   filter(argvals <= 0.5), tm = 0.5) -->
 
-# figure
+<!-- # figure  -->
 
-bind_rows(df_test %>% 
-            select(argvals, subj, Y)%>% 
-            rename("y" = "Y") %>% 
-            mutate(type = "true", subj = as.factor(subj)), 
-          df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>% 
-  filter(subj %in% 151:156) %>% 
-  ggplot(aes(x = argvals, y = y, col = type))+
-  geom_line()+
-  facet_wrap(~subj)+
-  labs(title = "Prediction with observation up to t=0.5", 
-       x = "time", y = "")
+<!-- bind_rows(df_test %>%  -->
+<!--             select(argvals, subj, Y)%>%  -->
+<!--             rename("y" = "Y") %>%  -->
+<!--             mutate(type = "true", subj = as.factor(subj)),  -->
+<!--           df_test_t0.5 %>% mutate(subj = as.factor(subj))) %>%  -->
+<!--   filter(subj %in% 151:156) %>%  -->
+<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
+<!--   geom_line()+ -->
+<!--   facet_wrap(~subj)+ -->
+<!--   labs(title = "Prediction with observation up to t=0.5",  -->
+<!--        x = "time", y = "") -->
 
-```
+<!-- ``` -->
 
-## Prediction of new partially observed sample
+<!-- ## Prediction of new partially observed sample -->
 
-```{r}
-df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>%
-  rename("y" = "Y") %>% 
-  filter(argvals <= 0.8), tm = 0.8)
+<!-- ```{r} -->
+<!-- df_test_t0.8 <- DynPred(df_test %>%  select(argvals, subj, Y) %>% -->
+<!--   rename("y" = "Y") %>%  -->
+<!--   filter(argvals <= 0.8), tm = 0.8) -->
 
-# figure 
-bind_rows(df_test %>% 
-            select(argvals, subj, Y)%>% 
-            rename("y" = "Y") %>% 
-            mutate(type = "true", subj = as.factor(subj)), 
-          df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>% 
-  filter(subj %in% 151:156) %>% 
-  ggplot(aes(x = argvals, y = y, col = type))+
-  geom_line()+
-  facet_wrap(~subj)+
-  labs(title = "Prediction with observation up to t=0.8", 
-       x = "time", y = "")
+<!-- # figure  -->
+<!-- bind_rows(df_test %>%  -->
+<!--             select(argvals, subj, Y)%>%  -->
+<!--             rename("y" = "Y") %>%  -->
+<!--             mutate(type = "true", subj = as.factor(subj)),  -->
+<!--           df_test_t0.8 %>% mutate(subj = as.factor(subj))) %>%  -->
+<!--   filter(subj %in% 151:156) %>%  -->
+<!--   ggplot(aes(x = argvals, y = y, col = type))+ -->
+<!--   geom_line()+ -->
+<!--   facet_wrap(~subj)+ -->
+<!--   labs(title = "Prediction with observation up to t=0.8",  -->
+<!--        x = "time", y = "") -->
 
-```
+<!-- ``` -->
 
@@
 
-## Extension to non-Gaussian data
+<!-- ## Extension to non-Gaussian data -->
 
-```{r gen_bi_outcome}
-times <- seq(0, 1, by = 0.001)
-gen_data <- function(id = 1){
-  scores <- rnorm(4)
-  pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times)
-  Z <- pcs %*% scores
-  probs <- exp(Z)/(1+exp(Z))
-  Y <- sapply(probs, rbinom, n = 1, size = 1)
+<!-- ```{r gen_bi_outcome} -->
+<!-- times <- seq(0, 1, by = 0.001) -->
+<!-- gen_data <- function(id = 1){ -->
+<!--   scores <- rnorm(4) -->
+<!--   pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times) -->
+<!--   Z <- pcs %*% scores -->
+<!--   probs <- exp(Z)/(1+exp(Z)) -->
+<!--   Y <- sapply(probs, rbinom, n = 1, size = 1)  -->
 
-  return(data.frame(id = id, argvals = times, Y=Y, Z=Z))
-}
+<!--   return(data.frame(id = id, argvals = times, Y=Y, Z=Z)) -->
+<!-- } -->
 
-df_train2 <- bind_rows(lapply(1:20, gen_data))
+<!-- df_train2 <- bind_rows(lapply(1:20, gen_data)) -->
 
-ggplot(df_train2 %>% filter(id %in% 1:4))+
-  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
-            show.legend = F)+
-  geom_line(aes(x=argvals, y=Z, col = "Red"), 
-            show.legend = F)+
-  facet_wrap(~id)+
-  labs(title = "Simulated function for the 4 subjects",
-       x = "time", y ="")
-```
+<!-- ggplot(df_train2 %>% filter(id %in% 1:4))+ -->
+<!--   geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1, -->
+<!--             show.legend = F)+ -->
+<!--   geom_line(aes(x=argvals, y=Z, col = "Red"),  -->
+<!--             show.legend = F)+ -->
+<!--   facet_wrap(~id)+ -->
+<!--   labs(title = "Simulated function for the 4 subjects", -->
+<!--        x = "time", y ="") -->
+<!-- ``` -->
 
-## Extension to non-Gaussian data
+<!-- ## Extension to non-Gaussian data -->
 
-```{r}
-# fPCA
-fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15,
-                        argvals.new = times, calculate.scores = TRUE)
+<!-- ```{r} -->
+<!-- # fPCA -->
+<!-- fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15, -->
+<!--                         argvals.new = times, calculate.scores = TRUE) -->
 
-# plot
-par(mfrow = c(2, 2))
-plot(times, fpca_Z$mu.new, type = "l", 
-     xlab = "t", ylab = "f0(t)")
-plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
-plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
-plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")
+<!-- # plot -->
+<!-- par(mfrow = c(2, 2)) -->
+<!-- plot(times, fpca_Z$mu.new, type = "l",  -->
+<!--      xlab = "t", ylab = "f0(t)") -->
+<!-- plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1") -->
+<!-- plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2") -->
+<!-- plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3") -->
 
-```
+<!-- ``` -->
 

### 2023-04-06 FDAgroup.tex
edit introduction

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ignorenonframetext,
]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{part title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{part title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\newif\ifbibliography
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction with fPCA},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction with fPCA}
\author{}
\date{\vspace{-2.5em}2022-11-3}

\begin{document}
\frame{\titlepage}

\begin{frame}{Dynamic prediction}
\protect\hypertarget{dynamic-prediction}{}
\begin{itemize}
\tightlist
\item
  With observations up to \(t_m\), predict outcomes (or probabilities of
  outcome) after that time point
\item
  Prediction updates with new observations
\item
  Mixed model prediction is difficult

  \begin{itemize}
  \tightlist
  \item
    Not flexible enough for densely measured data
  \item
    Out-of-sample random effects cannot be estimated
  \end{itemize}
\item
  Functional mixed effects model
\end{itemize}
\end{frame}

\begin{frame}{Functional Concurrent Regression (FCR)}
\protect\hypertarget{functional-concurrent-regression-fcr}{}
\begin{itemize}
\item
  Goal: to predict future track based on partially observed track
\item
  For a subject i, we observe a function over t
  \[Y_i(t)=f_0(t)+b_i(t)+\epsilon_i(t)\]
\item
  Subject-specific random effect
\end{itemize}

\[b_i(t) = \Sigma_{k=1}^c u_{ik}B_k(t)\] where
\(\boldsymbol{u_i}\sim N(0, \Gamma)\)

\begin{itemize}
\tightlist
\item
  We usually observe \(Y_i\) on a series of discrete \(t_{ij}\)
\end{itemize}

\[Y_{ij} = f_0(t_{ij})+b_i(t_{ij})+\epsilon_{ij}, \hspace{0.5cm} j = 1...J_i\]
where \(\epsilon_{ij} \sim N(0, \sigma_{\epsilon^2})\).
\end{frame}

\begin{frame}{Connection to fPCA}
\protect\hypertarget{connection-to-fpca}{}
\begin{itemize}
\item
  When there is no covariate in the model, this is essentially a fPCA
  problem.

  \begin{itemize}
  \tightlist
  \item
    \(\boldsymbol{B}\) is a matrix of eigenfunctions
  \item
    \(\boldsymbol{u}\) is a matrix of PC scores/loadings
  \end{itemize}
\item
  Use fPCA to estimate \(f_0\), \(\Gamma\) and \(\sigma_{\epsilon}\)
\item
  For a new subject with observations up to \(t_m\), estimate its score:
\end{itemize}

\[\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}\boldsymbol{y}\]
With the estimated score, we can predict its outcome in following time
points

\[\hat{\boldsymbol{Y}} = \hat{\boldsymbol{f}}_0+ \boldsymbol{B}^T\hat{\boldsymbol{u}} \]
\end{frame}

\begin{frame}{Extension to non-Gaussian data}
\protect\hypertarget{extension-to-non-gaussian-data}{}
\begin{itemize}
\tightlist
\item
  fPCA on non-Gaussian data is very difficult and computationally
  intensive
\item
  fPCA on a latent Gaussian variable instead

  \begin{itemize}
  \tightlist
  \item
    transformation using link function:
    \(g(E(Y))=\boldsymbol{X}^T\boldsymbol{\beta}\)
  \end{itemize}
\item
  Get a smooth latent Gaussian function by pooling a series of GLMM

  \begin{itemize}
  \item
    Assume we have regularly observed functions \(Y_{ij}\),
    \(i = 1... N\) and \(j = 1...J\)
  \item
    Bin the functional domain into small intervals \(t\in {1...T}\)
  \item
    Fit a GLMM in each bin to get a latent variables

    \[g(E(Y_{it})) = \beta_{0t}+b_{it}\] Let \(Z_{it} = g(E(Y_{it}))\),
    we can do Gaussian fPCA on this latent variable
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Extension to non-Gaussian data}
\protect\hypertarget{extension-to-non-gaussian-data-1}{}
\[Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t\]

\[Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})\]
\end{frame}

\begin{frame}{Next steps}
\protect\hypertarget{next-steps}{}
\begin{itemize}
\tightlist
\item
  Simulate a non-Gaussian function and implement methods above

  \begin{itemize}
  \tightlist
  \item
    Improve estiation of latent functions very well
  \item
    Numeric problems
  \end{itemize}
\item
  Transform latent function back to non-Gaussian function
\item
  Inclusion of covariates
\end{itemize}
\end{frame}

\end{document}

### 2023-04-06 Write-ups/Manuscript_edit.Rmd
edit introduction

--- 2023-04-03 Write-ups/Manuscript.Rmd
@@
 Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction.
-Traditionally, this type of data has been modeled either marginally, such as generalized estimating equations, or constitutionally on specific subjects, such as mixed effect models [@liang1986; @Laird1982; @lindstrom1990; @davidian2003].
+Traditionally, this type of data has been modeled either marginally, such as generalized estimating equations, or conditionally on specific subjects, such as mixed effect models [@liang1986; @Laird1982; @lindstrom1990; @davidian2003].
 For subject-specific models, predictions are made based on the correlation between repeated measures from the same subject and covariates that can be either fixed or varying.
-However, most of these methods are based on linear model structure, which is very limited in terms of flexibility of correlation.
-When sample size is large or the density of repeated measures is high, they also tend to cause severe computational burden [@GLMMadaptive].
-In addition, with generalized, non-Gaussian outcomes, these methods lack the ability to handle out-of-sample prediction because there is usually no close-form solution for subject-specific random effects
+Software and methods for dynamic prediction have been developed for Gaussian [leroux2016; @shang2017; @goldberg2014], non-Gaussian data [@chiou2012; @GLMMadaptive], and time-to-event outcomes [@suresh2017; @jmbook].
+However, with the notable exception of @leroux2016, existing methods are either limited in their flexibility (both fixed and random effect structure, e.g.
+random slope and intercept), computationally unfeasible for large datasets, or both.
+These issues are particularly acute for data which are measured densely, as is often the case with functional data, or when data are non-Gaussian.
 
-To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures.
-Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function.
-Non-parametric smoothing [@Scheipl2014] are often incorporated to speed up the computation, where random effects are approximated by spline basis functions or eigenfunctions from functional principal component analysis (FPCA).
-The introduction of basis functions also makes out-of-sample prediction more straightforward.
-Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing.
-Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, usually modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
-@kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains.
-While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors.
-
-Unfortunately, fewer papers have focused on its extension to generalized, non-Gaussian outcomes, such as series of binary or count outcomes.
-Existing methods also tend to be very computationally intensive.
+While methods with accompanying software for modeling complex functional data with non-linear fixed and random effects exist for both Gaussian and non-Gaussian data, software is largely limited to the refund package in the statistical software R [@Scheipl2014; @wood2014; @refundpkg].
+However, these methods and software implementations do not currently allow for dynamic prediction.
+In addition to the methods implemented in the refund package, other work has been done on the fitting of functional regression models to non-Gaussian functional data.
 For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes.
@@
 
-In this paper, we aim to develop a fast, scalable method for dynamic prediction of generalized functional outcome based on functional mixed effect model with FPCA smoothing.
-Specifically, "Generalized" indicates that the outcome follows non-Gaussian exponential family distribution, and the term "functional" indicates that repeated measure are collected densely across the study domain (e.g., time).
+Existing research on dynamic prediction with functional data has focused primarily on continuous/Gaussian outcomes, usually modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017].
+@kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains.
+While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors.
+Although methodological development exists, flexible software for implementing these methods is still relatively sparse with the exception of the fcr [@fcrpkg] package accompanying [@leroux2016].
+
+In this paper, we aim to develop a fast, scalable method for dynamic prediction of generalized functional outcomes using the principals of local generalized linear mixed effects models combined with functional principal components analysis of the estimated latent processes.
+The proposed method is designed to handle dense, regularly measured functional data and is extremely computationally efficient relative to the model complexity.
 Section 2 explains the implementation procedure of the proposed method.
@@
 
-where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process.
+where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process with covariance operator $\Sigma_b$.
+By the Karhunen-Loève theorem, $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$.
+In practice, $b_i(t)$ is approximated using a finite $K$ number of basis functions.
 
-According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$.
-We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation.
-This random noise is assumed to follow a zero-mean Gaussian process, such that at each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.
+Estimation of $b_i(t)$ involves estimating both $\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}$ and $\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}$.
+For Gaussian data this may be done by directly estimating the covariance operator $\Sigma_b$ using either method of moments (if the data are relatively small) or using faster methods such as the FACE method [reference - FACE].
+However, for non-Gaussian data, direct estimation is not possible.
+Instead, one may choose to estimate $b_i(t)$ first using a flexible spline basis  (e.g.
+B-splines)  with dimension $K^b$ such that $K^B > K$, assuming an unstructured correlation between subject-specific spline coefficients, then orthogonalizing the resulting estimates via eigendecomposition or singular value decomposition.
+This general approach forms the basis for estimation for several GFPCA methods [references - goldsmith, julia's EM algorithm].
+However, this approach results in rapidly escalating computational burden as both the number of subjects, $N$, and the density of the observed functional predictor increases.
+Note that if the covariance operator $\Sigma_b$ is relatively complex, then $K$ may be large and not well approximated using simple basis functions, $\phi_k$ (e.g.
+random intercept and slope).
 
-In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$.
-In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$.
-These eigenfunctions, also referred to as Principal Component (PC) functions, are orthogonal are can be orthonormal with proper scaling.
-$\xi_{ik}$ are subject-specific scores (or loadings) on each PC function.
-According to properties of FPCA method, these scores are mutually independent over both subject ($i$) and eigenfunctions ($k$).
-That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$, where $\lambda_k$ is the kth eigenvalues.
+Beyond the computational burden associated with fitting the model of interest, for dynamic prediction one must then estimate the scores for a new subject $u$ not used in model fitting.
+Typically this takes the form of the conditional expectation of the scores $E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]$ where $\bf{\Theta}$ represents the set of parameters estimated by the model (i.e.
+parameters associated with $\beta_0(t), \Sigma_b$) and $\bf{y}_u$ are the non-Gaussian data for the new subject.
+Estimation of the random effects for a new subject is non-trivial and not implemented for even very simple models in most random effects software packages such as lme4 [reference] and nlme [reference].
+Our proposed method bypasses this problem by using local generalized linear mixed models to estimate the covariance operator $\Sigma_b$ directly on the latent scale, then using the closed form solution for the best linear unbiased predictor of Gaussian random effects.
+Estimation is made fast for even very large data through the use of the FACE methodology for FPCA on the latent estimates obtained from local models.
+Because the local models are very simple GLMMs, estimation is also fast for the local models.
 
@@
 Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
+
+**Ying, include the formulas for your conditional expectation here to obtain random effects estimates **
 
@@
 This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes.
-Just like many mixed models, this method is very limited in terms of flexibility.
+Just like many mixed models, this method is very limited in terms of flexibility if one wishes to obtain estimates in a reasonable time frame.
 For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$.
@@
 
-The simulation study and case study of daily weigh-in data above have revealed the proposed method can achieve better predictive performance with much less computational burden.
-Compared to existing methods based on linear mixed model, fGFPCA can accommodate flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with wiggly or cyclic pattern.
-As in the simulation setting in Section 3, the binary outcomes are generated from a cyclic process.
-fGFPCA has much lower ISE and higher AUC than GLMM in this case, and its prediction accuracy also improves as more outcomes are collected, which is a desirable property for dynamic prediction where the datasets would be periodically updated with newly collected data.
-Moreover, fGFPCA also significantly reduce time spent on model fitting and predict, which is beneficial when repeated measures are dense or sample size is large.
+The simulation study and data application above have illustrated the feasibility and utility of the proposed method, even for large datasets and complex latent processes/random effect structures.
+can achieve better predictive performance with much less computational burden.
+Compared to competing methods, fGFPCA can accommodate extremely flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with highly non-linear patterns.
+Due to the comparative flexibility of the fGFPCA approach, fGFPCA has much lower ISE and higher AUC than GLMM.
+As expected, the dynamic prediction accuracy of fGFPCA improves as more data are availble for predictions.
+Moreover, fGFPCA also significantly reduce time spent on both model fitting and prediction, allowing for scaling of models and predictions which are infeasible with existing approaches.
 In fact, as far as we are aware of, the proposed method is the only feasible method that can handle the scale of the simulated datasets with compatible flexibility.
+
+**Andrew: I would delete this paragraph, honestly**
 
@@
 
-The hyperparameters introduced by the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quantify such effects.
+A few open questions remain with the fGFPCA approach.
+The choices involved in the binning procedure in Step 1 can also affect the final predictive performance, thus further investigations are needed to quantify such effects.
 Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid.
-Moreover, if the bins are very narrow with only a few observations, unidentifiability issues can happen when fitting local mixed models.
-While this may not necessarily affect point predictions, it can cause numeric failure to estimate standard error and prediction intervals.
+Moreover, if the bins are very narrow with only a few observations, local models may be non-identifiabile, resulting in bins with no point estimates.
+While this may not overly impact point final predictions, it is currently unclear as to when this issue warrants concern.
 In practice, we expect this to happen when the measurements are sparse or sample size is small.
-Under such scenarios, one may choose traditional mixed model framework instead to get valid estimates at the expense of computational time.
-
-It should also be noted that the binning procedure in Step 1 changes the density of repeated measures.
-We essentially assume that latent functions are smooth in small intervals, and treat adjacent observations as if they are generated from the same time point.
-The final prediction of latent continuous function tracks are made on the binned grid, integrating information from several adjacent observations in the same bin.
-In Section 3 and 4, this sparser grid was transformed to the original, un-binned grid using linear interpolation of the predicted latent function.
-However, this is by no means the only method to be considered.
-Such transformation can also be made by projection to a higher-dimensional space, similar as general additive model with a large number of basis functions.
-We suspect that this method would work better than linear interpolation when the original grid is sparse or irregular, but take longer time for computation.
+However, the methodology here is designed for large, dense functional data and thus not currently appropriate in such scenarios.
+In this case, existing methods may be sufficiently flexible and computationally feasible.
 

### 2023-04-06 Write-ups/Manuscript_edit.tex
edit introduction

--- 2023-04-03 Write-ups/Manuscript.tex
@@
 \author{Ying Jin \and Andrew Leroux}
-\date{April 03, 2023}
+\date{April 06, 2023}
 
@@
 been modeled either marginally, such as generalized estimating
-equations, or constitutionally on specific subjects, such as mixed
-effect models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
+equations, or conditionally on specific subjects, such as mixed effect
+models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
 \protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
@@
 made based on the correlation between repeated measures from the same
-subject and covariates that can be either fixed or varying. However,
-most of these methods are based on linear model structure, which is very
-limited in terms of flexibility of correlation. When sample size is
-large or the density of repeated measures is high, they also tend to
-cause severe computational burden
-(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). In addition,
-with generalized, non-Gaussian outcomes, these methods lack the ability
-to handle out-of-sample prediction because there is usually no
-close-form solution for subject-specific random effects
+subject and covariates that can be either fixed or varying. Software and
+methods for dynamic prediction have been developed for Gaussian
+{[}leroux2016; Shang (\protect\hyperlink{ref-shang2017}{2017}); Goldberg
+et al. (\protect\hyperlink{ref-goldberg2014}{2014}){]}, non-Gaussian
+data (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
+\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}), and
+time-to-event outcomes (\protect\hyperlink{ref-jmbook}{Rizopoulos 2012};
+\protect\hyperlink{ref-suresh2017}{Suresh et al. 2017}). However, with
+the notable exception of Leroux et al.
+(\protect\hyperlink{ref-leroux2016}{2018a}), existing methods are either
+limited in their flexibility (both fixed and random effect structure,
+e.g.~random slope and intercept), computationally unfeasible for large
+datasets, or both. These issues are particularly acute for data which
+are measured densely, as is often the case with functional data, or when
+data are non-Gaussian.
 
-To address these problems, one may turn to functional mixed effect
-models instead when measures are dense across the domain of measures.
-Such methods accommodate more flexible correlation structure by modeling
-subject-specific random effects as a function. Non-parametric smoothing
-(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) are often
-incorporated to speed up the computation, where random effects are
-approximated by spline basis functions or eigenfunctions from functional
-principal component analysis (FPCA). The introduction of basis functions
-also makes out-of-sample prediction more straightforward. Instead of
-estimating subject-specific random effects of new observations, we can
-simply estimate coefficients/loadings on the basis function used for
-smoothing. Existing research on dynamic prediction with functional data
-analysis methods has been focusing on continuous/Gaussian outcomes,
-usually modelling subject-specific random effects with FPCA
-(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
-\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
-\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
-(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
-predict missing observations in partially observed function tracks, and
-Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
-similar goals using Markov Chains. While methods mentioned above used
-only partial observations for prediction with an intercept-only model,
-Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
-Functional Concurrent Regression (FCR) framework which can incorporate
-the effect of subject-specific predictors.
-
-Unfortunately, fewer papers have focused on its extension to
-generalized, non-Gaussian outcomes, such as series of binary or count
-outcomes. Existing methods also tend to be very computationally
-intensive. For example, Chen et al.
-(\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
+While methods with accompanying software for modeling complex functional
+data with non-linear fixed and random effects exist for both Gaussian
+and non-Gaussian data, software is largely limited to the refund package
+in the statistical software R
+(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
+\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014};
+\protect\hyperlink{ref-refundpkg}{\textbf{refundpkg?}}). However, these
+methods and software implementations do not currently allow for dynamic
+prediction. In addition to the methods implemented in the refund
+package, other work has been done on the fitting of functional
+regression models to non-Gaussian functional data. For example, Chen et
+al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
 marginal functional models that is compatible to multi-level,
@@
 
+Existing research on dynamic prediction with functional data has focused
+primarily on continuous/Gaussian outcomes, usually modelling
+subject-specific random effects with FPCA
+(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
+\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
+\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
+(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
+predict missing observations in partially observed function tracks, and
+Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
+similar goals using Markov Chains. While methods mentioned above used
+only partial observations for prediction with an intercept-only model,
+Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018a}) proposed
+Functional Concurrent Regression (FCR) framework which can incorporate
+the effect of subject-specific predictors. Although methodological
+development exists, flexible software for implementing these methods is
+still relatively sparse with the exception of the fcr
+(\protect\hyperlink{ref-fcrpkg}{\textbf{fcrpkg?}}) package accompanying
+(\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).
+
 In this paper, we aim to develop a fast, scalable method for dynamic
-prediction of generalized functional outcome based on functional mixed
-effect model with FPCA smoothing. Specifically, ``Generalized''
-indicates that the outcome follows non-Gaussian exponential family
-distribution, and the term ``functional'' indicates that repeated
-measure are collected densely across the study domain (e.g., time).
-Section 2 explains the implementation procedure of the proposed method.
-In Section 3, we illustrate the performance and efficiency of our
-proposed method in a simulation study. In Section 4, we apply this
-method to a real world dataset from a weight loss study. Section 5
-presents a discussion of advantages and limitation of the proposed
-method.
+prediction of generalized functional outcomes using the principals of
+local generalized linear mixed effects models combined with functional
+principal components analysis of the estimated latent processes. The
+proposed method is designed to handle dense, regularly measured
+functional data and is extremely computationally efficient relative to
+the model complexity. Section 2 explains the implementation procedure of
+the proposed method. In Section 3, we illustrate the performance and
+efficiency of our proposed method in a simulation study. In Section 4,
+we apply this method to a real world dataset from a weight loss study.
+Section 5 presents a discussion of advantages and limitation of the
+proposed method.
 
@@
 population mean of latent function, and \(b_i(t)\) is a subjects-level
-random effect function following a zero-mean Gaussian process.
+random effect function following a zero-mean Gaussian process with
+covariance operator \(\Sigma_b\). By the Karhunen-Loève theorem,
+\(b_i(t)\) can be represented as the linear combination of a infinite
+set of orthogonal functions:
+\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). In practice,
+\(b_i(t)\) is approximated using a finite \(K\) number of basis
+functions.
 
-According to the Karhunen-Loève theorem, the unobserved random effect
-function \(b_i(t)\) can be represented as the linear combination of a
-infinite set of orthogonal functions:
-\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
-function by a finite set of \(\phi_{k}(t)\), such that
-\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)\), where
-\(\epsilon_i(t)\) indicates random noise introduced by this
-approximation. This random noise is assumed to follow a zero-mean
-Gaussian process, such that at each specific point t,
-\(\epsilon_i(t) \sim N(0, \sigma^2)\).
+Estimation of \(b_i(t)\) involves estimating both
+\(\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}\) and
+\(\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}\). For Gaussian data
+this may be done by directly estimating the covariance operator
+\(\Sigma_b\) using either method of moments (if the data are relatively
+small) or using faster methods such as the FACE method {[}reference -
+FACE{]}. However, for non-Gaussian data, direct estimation is not
+possible. Instead, one may choose to estimate \(b_i(t)\) first using a
+flexible spline basis (e.g.~B-splines) with dimension \(K^b\) such that
+\(K^B > K\), assuming an unstructured correlation between
+subject-specific spline coefficients, then orthogonalizing the resulting
+estimates via eigendecomposition or singular value decomposition. This
+general approach forms the basis for estimation for several GFPCA
+methods {[}references - goldsmith, julia's EM algorithm{]}. However,
+this approach results in rapidly escalating computational burden as both
+the number of subjects, \(N\), and the density of the observed
+functional predictor increases. Note that if the covariance operator
+\(\Sigma_b\) is relatively complex, then \(K\) may be large and not well
+approximated using simple basis functions, \(\phi_k\) (e.g.~random
+intercept and slope).
 
-In practice, this approximation could be implemented through an FPCA
-procedure on \(\eta_i(t)\). In this case
-\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
-explain the most variation in \(\eta_i(t)\). These eigenfunctions, also
-referred to as Principal Component (PC) functions, are orthogonal are
-can be orthonormal with proper scaling. \(\xi_{ik}\) are
-subject-specific scores (or loadings) on each PC function. According to
-properties of FPCA method, these scores are mutually independent over
-both subject (\(i\)) and eigenfunctions (\(k\)). That is, each
-\(\xi_{ik}\) follows normal distribution \(N(0, \lambda_k)\), where
-\(\lambda_k\) is the kth eigenvalues.
+Beyond the computational burden associated with fitting the model of
+interest, for dynamic prediction one must then estimate the scores for a
+new subject \(u\) not used in model fitting. Typically this takes the
+form of the conditional expectation of the scores
+\(E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]\) where \(\bf{\Theta}\) represents
+the set of parameters estimated by the model (i.e.~parameters associated
+with \(\beta_0(t), \Sigma_b\)) and \(\bf{y}_u\) are the non-Gaussian
+data for the new subject. Estimation of the random effects for a new
+subject is non-trivial and not implemented for even very simple models
+in most random effects software packages such as lme4 {[}reference{]}
+and nlme {[}reference{]}. Our proposed method bypasses this problem by
+using local generalized linear mixed models to estimate the covariance
+operator \(\Sigma_b\) directly on the latent scale, then using the
+closed form solution for the best linear unbiased predictor of Gaussian
+random effects. Estimation is made fast for even very large data through
+the use of the FACE methodology for FPCA on the latent estimates
+obtained from local models. Because the local models are very simple
+GLMMs, estimation is also fast for the local models.
 
@@
 \end{enumerate}
+
+\textbf{Ying, include the formulas for your conditional expectation here
+to obtain random effects estimates }
 
@@
 generalized outcomes. Just like many mixed models, this method is very
-limited in terms of flexibility. For example, the model used for
-prediction of our simulated datset would simply be an linear model with
-one covariate indicating observation time:
+limited in terms of flexibility if one wishes to obtain estimates in a
+reasonable time frame. For example, the model used for prediction of our
+simulated datset would simply be an linear model with one covariate
+indicating observation time:
 \(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
@@
 \centering
-\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
+\includegraphics{Manuscript_edit_files/figure-latex/Figure-1.pdf}
 \caption{Predicted track of four randomly selected subjects from one
@@
 \centering
-\includegraphics{Manuscript_files/figure-latex/Figure_appl-1.pdf}
+\includegraphics{Manuscript_edit_files/figure-latex/Figure_appl-1.pdf}
 \caption{Predicted track of four randomly selected subjects from the
@@
 
-The simulation study and case study of daily weigh-in data above have
-revealed the proposed method can achieve better predictive performance
-with much less computational burden. Compared to existing methods based
-on linear mixed model, fGFPCA can accommodate flexible correlation
-structure between repeated measure, which is particularly useful when
-outcome is generated from latent functions with wiggly or cyclic
-pattern. As in the simulation setting in Section 3, the binary outcomes
-are generated from a cyclic process. fGFPCA has much lower ISE and
-higher AUC than GLMM in this case, and its prediction accuracy also
-improves as more outcomes are collected, which is a desirable property
-for dynamic prediction where the datasets would be periodically updated
-with newly collected data. Moreover, fGFPCA also significantly reduce
-time spent on model fitting and predict, which is beneficial when
-repeated measures are dense or sample size is large. In fact, as far as
-we are aware of, the proposed method is the only feasible method that
-can handle the scale of the simulated datasets with compatible
-flexibility.
+The simulation study and data application above have illustrated the
+feasibility and utility of the proposed method, even for large datasets
+and complex latent processes/random effect structures. can achieve
+better predictive performance with much less computational burden.
+Compared to competing methods, fGFPCA can accommodate extremely flexible
+correlation structure between repeated measure, which is particularly
+useful when outcome is generated from latent functions with highly
+non-linear patterns. Due to the comparative flexibility of the fGFPCA
+approach, fGFPCA has much lower ISE and higher AUC than GLMM. As
+expected, the dynamic prediction accuracy of fGFPCA improves as more
+data are availble for predictions. Moreover, fGFPCA also significantly
+reduce time spent on both model fitting and prediction, allowing for
+scaling of models and predictions which are infeasible with existing
+approaches. In fact, as far as we are aware of, the proposed method is
+the only feasible method that can handle the scale of the simulated
+datasets with compatible flexibility.
+
+\textbf{Andrew: I would delete this paragraph, honestly}
 
@@
 
-The hyperparameters introduced by the binning procedure in Step 1 can
-also affect the final predictive performance, thus further
-investigations are needed to quantify such effects. Predictive accuracy
-and computational speed can change with bin width, number of
-observations in each bin, whether the bins overlap with each other, also
-whether the bins are equally spaced along the functional grid. Moreover,
-if the bins are very narrow with only a few observations,
-unidentifiability issues can happen when fitting local mixed models.
-While this may not necessarily affect point predictions, it can cause
-numeric failure to estimate standard error and prediction intervals. In
-practice, we expect this to happen when the measurements are sparse or
-sample size is small. Under such scenarios, one may choose traditional
-mixed model framework instead to get valid estimates at the expense of
-computational time.
-
-It should also be noted that the binning procedure in Step 1 changes the
-density of repeated measures. We essentially assume that latent
-functions are smooth in small intervals, and treat adjacent observations
-as if they are generated from the same time point. The final prediction
-of latent continuous function tracks are made on the binned grid,
-integrating information from several adjacent observations in the same
-bin. In Section 3 and 4, this sparser grid was transformed to the
-original, un-binned grid using linear interpolation of the predicted
-latent function. However, this is by no means the only method to be
-considered. Such transformation can also be made by projection to a
-higher-dimensional space, similar as general additive model with a large
-number of basis functions. We suspect that this method would work better
-than linear interpolation when the original grid is sparse or irregular,
-but take longer time for computation.
+A few open questions remain with the fGFPCA approach. The choices
+involved in the binning procedure in Step 1 can also affect the final
+predictive performance, thus further investigations are needed to
+quantify such effects. Predictive accuracy and computational speed can
+change with bin width, number of observations in each bin, whether the
+bins overlap with each other, also whether the bins are equally spaced
+along the functional grid. Moreover, if the bins are very narrow with
+only a few observations, local models may be non-identifiabile,
+resulting in bins with no point estimates. While this may not overly
+impact point final predictions, it is currently unclear as to when this
+issue warrants concern. In practice, we expect this to happen when the
+measurements are sparse or sample size is small. However, the
+methodology here is designed for large, dense functional data and thus
+not currently appropriate in such scenarios. In this case, existing
+methods may be sufficiently flexible and computationally feasible.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-refundpck}{}}%
+Goldsmith, J., Scheipl, F., Huang, L., Wrobel, J., Di, C., Gellar, J.,
+Harezlak, J., McLean, M. W., Swihart, B., Xiao, L., Crainiceanu, C., and
+Reiss, P. T. (2022),
+\emph{\href{https://CRAN.R-project.org/package=refund}{Refund:
+Regression with functional data}}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-goldsmith2015}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-leroux2016}{}}%
-Leroux, A., Xiao, L., Crainiceanu, C., and Checkley, W. (2018),
+Leroux, A., Xiao, L., Crainiceanu, C., and Checkley, W. (2018a),
 {``Dynamic prediction in functional concurrent regression with an
@@
 1376--1388.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fcrpck}{}}%
+Leroux, A., Xiao, L., Crainiceanu, C., and Checkly, W. (2018b),
+\emph{\href{https://CRAN.R-project.org/package=fcr}{Fcr: Functional
+concurrent regression for sparse data}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-jmbook}{}}%
+Rizopoulos, D. (2012), \emph{Joint models for longitudinal and
+time-to-event data: With applications in r}, Chapman; Hall/CRC.
+https://doi.org/\url{https://doi.org/10.1201/b12208}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-GLMMadaptive}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wood2014}{}}%
+Wood, S., and Scheipl, F. (2014), {``gamm4: Generalized additive mixed
+models using mgcv and lme4,''} \emph{R package version 0.2-3}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wrobel2019}{}}%

### 2023-04-07 Write-ups/Manuscript_edit.Rmd
WNAR paper compitition submission

@@
 title: "Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis"
-author:
-  - Ying Jin
-  - Andrew Leroux
-date: "`r format(Sys.time(), '%B %d, %Y')`"
 output: 
@@
 For subject-specific models, predictions are made based on the correlation between repeated measures from the same subject and covariates that can be either fixed or varying.
-Software and methods for dynamic prediction have been developed for Gaussian [leroux2016; @shang2017; @goldberg2014], non-Gaussian data [@chiou2012; @GLMMadaptive], and time-to-event outcomes [@suresh2017; @jmbook].
+Software and methods for dynamic prediction have been developed for Gaussian [@leroux2016; @shang2017; @goldberg2014], non-Gaussian data [@chiou2012; @GLMMadaptive], and time-to-event outcomes [@suresh2017; @jmbook].
 However, with the notable exception of @leroux2016, existing methods are either limited in their flexibility (both fixed and random effect structure, e.g.
@@
 
-While methods with accompanying software for modeling complex functional data with non-linear fixed and random effects exist for both Gaussian and non-Gaussian data, software is largely limited to the refund package in the statistical software R [@Scheipl2014; @wood2014; @refundpkg].
+While methods with accompanying software for modeling complex functional data with non-linear fixed and random effects exist for both Gaussian and non-Gaussian data, software is largely limited to the *refund* package in the statistical software R [@Scheipl2014; @wood2014; @refundpkg].
 However, these methods and software implementations do not currently allow for dynamic prediction.
@@
 @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*.
-@gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage joint estimation strategy.
+@gertheiss2016 identified bias introduced by directly applying functional principal analysis (FPCA) methods to generalized functions, and proposed to address this problem using a two-stage joint estimation strategy.
 @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data.
-In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit GFPCA on binary data using EM algorithm, accompanied by the an open source R package *registr*.
+In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit generalized FPCA (GFPCA) on binary data using EM algorithm, accompanied by the an open source R package *registr*.
 
@@
 In practice, $t$ is usually a time index along the duration of study.
-We hereafter refer to the range of $t$ as the "functional domain".
+We hereafter refer to the range of $t$ as the "functional domain" denoted by $T$.
 For simplicity, we further assume that outcomes are collected on a regular grid over the functional domain, meaning that $t$ are evenly distributed and does not change across subjects.
@@
 We assume that the outcome $Y_i(t)$ can be characterized by a latent continuous function $\eta_i(t)$.
-That is, at a specific t, $Y_i(t)$ follows a exponential family distribution such that:
+That is, at a specific t, $Y_i(t)$ follows an exponential family distribution such that:
 
@@
 
-Estimation of $b_i(t)$ involves estimating both $\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}$ and $\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}$.
-For Gaussian data this may be done by directly estimating the covariance operator $\Sigma_b$ using either method of moments (if the data are relatively small) or using faster methods such as the FACE method [reference - FACE].
+Estimation of $b_i(t)$ involves estimating both $\{\phi_k(t): t \in T, , 1 \leq k \leq K\}$ and $\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}$.
+For Gaussian data this may be done by directly estimating the covariance operator $\Sigma_b$ using either method of moments (if the data are relatively small) or faster methods such as fast covariance estimation (FACE) [@face].
 However, for non-Gaussian data, direct estimation is not possible.
 Instead, one may choose to estimate $b_i(t)$ first using a flexible spline basis  (e.g.
-B-splines)  with dimension $K^b$ such that $K^B > K$, assuming an unstructured correlation between subject-specific spline coefficients, then orthogonalizing the resulting estimates via eigendecomposition or singular value decomposition.
-This general approach forms the basis for estimation for several GFPCA methods [references - goldsmith, julia's EM algorithm].
+B-splines)  with dimension $K^B$ such that $K^B > K$, assuming an unstructured correlation between subject-specific spline coefficients, then orthogonalize the resulting estimates via eigendecomposition or singular value decomposition.
+This general approach forms the basis for estimation for several GFPCA methods [@goldsmith2015; @wrobel2019].
 However, this approach results in rapidly escalating computational burden as both the number of subjects, $N$, and the density of the observed functional predictor increases.
@@
 Beyond the computational burden associated with fitting the model of interest, for dynamic prediction one must then estimate the scores for a new subject $u$ not used in model fitting.
-Typically this takes the form of the conditional expectation of the scores $E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]$ where $\bf{\Theta}$ represents the set of parameters estimated by the model (i.e.
-parameters associated with $\beta_0(t), \Sigma_b$) and $\bf{y}_u$ are the non-Gaussian data for the new subject.
-Estimation of the random effects for a new subject is non-trivial and not implemented for even very simple models in most random effects software packages such as lme4 [reference] and nlme [reference].
-Our proposed method bypasses this problem by using local generalized linear mixed models to estimate the covariance operator $\Sigma_b$ directly on the latent scale, then using the closed form solution for the best linear unbiased predictor of Gaussian random effects.
+Typically for non-Gaussian data, this takes the form of the maximum likelihood estimator (MLE) of the scores $\hat{\xi}_{uk} = \underset{\xi_{uk}}{\mathrm{argmax}} L(\xi_{uk}|\boldsymbol{\Theta}, \boldsymbol{Y}_{u})$, where $\bf{\Theta}$ represents the set of parameters estimated by the model (i.e.
+parameters associated with $\beta_0(t), \Sigma_b$) and $\boldsymbol{Y}_u$ are the non-Gaussian data for the new subject.
+Estimation of the random effects for a new subject is non-trivial and not implemented for even very simple models in most random effects software packages such as *lme4* [@lme4] and *nlme* [@nlme].
+Our proposed method bypasses this problem by using local generalized linear mixed models (GLMMs) to estimate the covariance operator $\Sigma_b$ directly on the latent scale, then using the MLE of Gaussian random effects.
 Estimation is made fast for even very large data through the use of the FACE methodology for FPCA on the latent estimates obtained from local models.
@@
 3.
-Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
+Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain $\hat{\boldsymbol{\Theta}}$ which includes the estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$.
 
-**Ying, include the formulas for your conditional expectation here to obtain random effects estimates **
+<!-- **Ying, include the formulas for your conditional expectation here to obtain random effects estimates ** -->
 
-**Dynamic prediction** Now assume we have a new observations with partially observed outcome $Y_{new}(t)$, $t\leq t_m$.
+**Dynamic prediction** Now assume we have a new observations $u$ with partially observed outcome $\boldsymbol{Y}_u = \{Y_u(t), t\leq t_m\}$.
 The maximum observed point is $t_m$, which is a point along the functional domain and belongs to bin $s_m$ on the binned grid from step 1 above.
 With components extracted from the FPCA model in step 3, we will be able to make predictions on the future, unobserved bins.
-Specifically, with the partially observed binary function $Y_{new}(t)$, we will be able to calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$.
-Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$
+Specifically, with the partially observed binary function $Y_u(t)$, we will be able to calculate the MLE of the subject-specific PC scores $\hat{\boldsymbol{\xi}}_u= \{\hat{\xi}_{u1},...,\hat{\xi}_{uK}\}$ that maximizes its posterior likelihood:
+
+$$\hat{\boldsymbol{\xi}}_u = \underset{\boldsymbol{\xi}_u}{\mathrm{argmin}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
+\hat{\boldsymbol{\Theta}})= \frac{p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})}{\int p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u,\hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})d\boldsymbol{\xi}_u}$$
+
+Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_u}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{uk}{\phi}_k(s)$.
+In fact, since the denominator in the formula above is constant with respect to $\boldsymbol{\xi}_u$, we only need to maximum the numerator.
+Since $\boldsymbol{Y}_u|\boldsymbol{\xi}_u$ follows an exponential family distribution and $\boldsymbol{\xi}_u$ follows multivariate Guassian distribution, it could be numerically maximized efficiently.
 
@@
 $\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \lambda_k)$.
-Here we set the values of $\lambda_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$.
+Here we set the values of $\lambda_k$ to be $0.5^{k-1}$, $k \in \{1,..., 4\}$.
 In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $(0, 1000]$ and are the same for all subjects.
@@
 
-As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive).
+As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive)[@GLMMadaptive].
 This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes.
 Just like many mixed models, this method is very limited in terms of flexibility if one wishes to obtain estimates in a reasonable time frame.
-For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$.
+For example, the model used for prediction of our simulated datasets would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$.
 While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.
@@
 
-```{r Figure, fig.cap="Predicted track of four randomly selected subjects from one simulated dataset.
+```{r Figure, fig.cap="Predicted tracks of four randomly selected subjects from one simulated dataset.
 The grey solid line indicates the true latent continuous function.
-The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
+The dashed lines indicate predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
 # figure of fGFPCA prediction
@@
   kable(booktabs = T, digit=2,
-        caption = "Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.",
+        caption = "Predictive performance of fGFPCA and GLMMadaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.",
         align = "lcccccccc") %>%
@@
 In this section, we illustrate the performance of our proposed method using data from a randomized trial on weight loss at University of Colorado Anschutz Medical Campus [@bothwell2022; @ostendorf2022].
-The study originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing.
+The trial originally aimed to study the effect of dietary strategies on weight loss, and participants were given a smart scale for daily at-home weighing.
 This record was then scaled into a binary indicator of participant adherence to the daily-weigh-in procedure, where 1 indicates participants followed the requirement and 0 otherwise.
@@
 In other words, for each individual, 400 binary measurements were taken on a regular grid along the functional domain (number of days into the study).
-We will fit both fGFPCA and GLMMadaptive models and compared their performance on this data.
+We will fit both fGFPCA and GLMMadaptive models and compare their performance on this data.
 For fGFPCA, every 10 observations are binned together, resulting in 40 bins for final model fitting.
 Linear interpolation was used for extension from the binned grid back to the un-binned grid.
-Since the latent continuous function tracks are not known in this case, prediction performance on evaluated with AUC alone.
+Since the latent continuous function tracks are not known in this case, prediction performance is evaluated with AUC alone.
 
 Table 2 compares the predictive performance of both models.
-Similar to the simulatio study, prediction is made conditioning on observations up to the 100th, 200th and 300th day respectively, and evaluation is made every 100 days following the maximum observation time.
-As is revealed, when the observed time is short (up to 100th day), the two method performed similarly when predicting recent days, however fGFPCA did better in predicting further days along the trial (prediction time window of $(300, 400]$).
+Similar to the simulation study, prediction is made conditioning on observations up to the 100th, 200th and 300th day respectively, and evaluation is made every 100 days following the maximum observation time.
+As is revealed, when the observed time is short (up to 100th day), the two method performed similarly when predicting recent days, however fGFPCA did better in predicting further days along the trial.
 As more observations coming in, fGFPCA sees increasingly higher AUC compared to GLMMadaptive model.
-This could be a result of the difference in model flexibility of the two methods.
+This is a result of the difference in model flexibility of the two methods.
 While the linear relationship in underlying latent functions are not likely to change immediately following the observed track, the whole track could be much more complex.
@@
 
-```{r, Figure_appl, fig.cap="Predicted track of four randomly selected subjects from the daily weigh-in dataset.
-The grey solid line indicates the true latent continuous function.
-The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
+```{r, Figure_appl, fig.cap="Predicted tracks of four randomly selected subjects from the daily weigh-in dataset.
+The grey points are observed binary outcomes.
+The dashed lines indicate predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
 rand_id2 <- sample(N2, 4)
@@
   kable(booktabs = T, digit=2,
-        caption = "AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset",
+        caption = "AUC of fGFPCA and GLMMadaptive on the daily weigh-in dataset",
         align = "lcccccccc") %>%
@@
 
-The simulation study and data application above have illustrated the feasibility and utility of the proposed method, even for large datasets and complex latent processes/random effect structures.
-can achieve better predictive performance with much less computational burden.
+The simulation study and data application above have illustrated the feasibility and utility of the proposed method.
+Even for large datasets and complex latent processes/random effect structures, fGFPCA can achieve better predictive performance with much less computational burden.
 Compared to competing methods, fGFPCA can accommodate extremely flexible correlation structure between repeated measure, which is particularly useful when outcome is generated from latent functions with highly non-linear patterns.
 Due to the comparative flexibility of the fGFPCA approach, fGFPCA has much lower ISE and higher AUC than GLMM.
-As expected, the dynamic prediction accuracy of fGFPCA improves as more data are availble for predictions.
+As expected, the dynamic prediction accuracy of fGFPCA improves as more data are available for predictions.
 Moreover, fGFPCA also significantly reduce time spent on both model fitting and prediction, allowing for scaling of models and predictions which are infeasible with existing approaches.
@@
 
-**Andrew: I would delete this paragraph, honestly**
+<!-- **Andrew: I would delete this paragraph, honestly** -->
 
-While the proposed method achieved good accuracy of point prediction, it is more complicated to develop a straightforward, interpretable metric to evaluate the precision of prediction.
+<!-- While the proposed method achieved good accuracy of point prediction, it is more complicated to develop a straightforward, interpretable metric to evaluate the precision of prediction.
 The challenge arises from the fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both introduce uncertainty to the final prediction.
@@
 Since the score $\hat{\xi}_{ik}$ is essentially a maximum likelihood estimate, it is natural to estimate its variance based on likelihood theory with observed information ${I}(\hat{\xi}_{ik})$.
-Therefore, the variation of prediction interval is:
+Therefore, the variation of prediction interval is:  -->
 
-$$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$
-Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$.
+<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
+<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->
 
@@
 Predictive accuracy and computational speed can change with bin width, number of observations in each bin, whether the bins overlap with each other, also whether the bins are equally spaced along the functional grid.
-Moreover, if the bins are very narrow with only a few observations, local models may be non-identifiabile, resulting in bins with no point estimates.
-While this may not overly impact point final predictions, it is currently unclear as to when this issue warrants concern.
+Moreover, if the bins are very narrow with only a few observations, local models may be non-identifiable, resulting in bins with no point estimates.
+While this may not overly impact final point predictions, it is currently unclear as to when this issue warrants concern.
 In practice, we expect this to happen when the measurements are sparse or sample size is small.
-However, the methodology here is designed for large, dense functional data and thus not currently appropriate in such scenarios.
-In this case, existing methods may be sufficiently flexible and computationally feasible.
+However, the methodology here is designed for large, dense functional data and thus not currently appropriate in such scenarios, in which case existing methods may be sufficiently flexible and computationally feasible.
 

### 2023-04-07 Write-ups/Manuscript_edit.tex
WNAR paper compitition submission

@@
   pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
-  pdfauthor={Ying Jin; Andrew Leroux},
   colorlinks=true,
@@
 Functional Principal Analysis}
-\author{Ying Jin \and Andrew Leroux}
-\date{April 06, 2023}
+\author{}
+\date{\vspace{-2.5em}}
 
@@
 equations, or conditionally on specific subjects, such as mixed effect
-models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
-\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
-\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
-\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
-measurement data''} 2003}). For subject-specific models, predictions are
-made based on the correlation between repeated measures from the same
-subject and covariates that can be either fixed or varying. Software and
-methods for dynamic prediction have been developed for Gaussian
-{[}leroux2016; Shang (\protect\hyperlink{ref-shang2017}{2017}); Goldberg
-et al. (\protect\hyperlink{ref-goldberg2014}{2014}){]}, non-Gaussian
-data (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
+models (\protect\hyperlink{ref-davidian2003}{Davidian and Giltinan
+2003}; \protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
+\protect\hyperlink{ref-liang1986}{Liang and Zeger 1986};
+\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990}). For
+subject-specific models, predictions are made based on the correlation
+between repeated measures from the same subject and covariates that can
+be either fixed or varying. Software and methods for dynamic prediction
+have been developed for Gaussian
+(\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
+\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a};
+\protect\hyperlink{ref-shang2017}{Shang 2017}), non-Gaussian data
+(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
 \protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}), and
@@
 data with non-linear fixed and random effects exist for both Gaussian
-and non-Gaussian data, software is largely limited to the refund package
-in the statistical software R
-(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
-\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014};
-\protect\hyperlink{ref-refundpkg}{\textbf{refundpkg?}}). However, these
+and non-Gaussian data, software is largely limited to the \emph{refund}
+package in the statistical software R
+(\protect\hyperlink{ref-refundpkg}{Goldsmith et al. 2022};
+\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
+\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014}). However, these
 methods and software implementations do not currently allow for dynamic
@@
 (\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
-by directly applying FPCA methods to generalized functions, and proposed
-to address this problem using a two-stage joint estimation strategy.
-Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
+by directly applying functional principal analysis (FPCA) methods to
+generalized functions, and proposed to address this problem using a
+two-stage joint estimation strategy. Linde
+(\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
 variational algorithm for FPCA of binary and count data. In terms of
 implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
-proposed a fast, efficient way to fit GFPCA on binary data using EM
-algorithm, accompanied by the an open source R package \emph{registr}.
+proposed a fast, efficient way to fit generalized FPCA (GFPCA) on binary
+data using EM algorithm, accompanied by the an open source R package
+\emph{registr}.
 
@@
 still relatively sparse with the exception of the fcr
-(\protect\hyperlink{ref-fcrpkg}{\textbf{fcrpkg?}}) package accompanying
-(\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).
+(\protect\hyperlink{ref-fcrpkg}{Leroux et al. 2018b}) package
+accompanying (\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).
 
@@
 duration of study. We hereafter refer to the range of \(t\) as the
-``functional domain''. For simplicity, we further assume that outcomes
-are collected on a regular grid over the functional domain, meaning that
-\(t\) are evenly distributed and does not change across subjects.
+``functional domain'' denoted by \(T\). For simplicity, we further
+assume that outcomes are collected on a regular grid over the functional
+domain, meaning that \(t\) are evenly distributed and does not change
+across subjects.
 
@@
 the outcome \(Y_i(t)\) can be characterized by a latent continuous
-function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows a
+function \(\eta_i(t)\). That is, at a specific t, \(Y_i(t)\) follows an
 exponential family distribution such that:
@@
 Estimation of \(b_i(t)\) involves estimating both
-\(\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}\) and
+\(\{\phi_k(t): t \in T, , 1 \leq k \leq K\}\) and
 \(\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}\). For Gaussian data
@@
 \(\Sigma_b\) using either method of moments (if the data are relatively
-small) or using faster methods such as the FACE method {[}reference -
-FACE{]}. However, for non-Gaussian data, direct estimation is not
-possible. Instead, one may choose to estimate \(b_i(t)\) first using a
-flexible spline basis (e.g.~B-splines) with dimension \(K^b\) such that
-\(K^B > K\), assuming an unstructured correlation between
-subject-specific spline coefficients, then orthogonalizing the resulting
-estimates via eigendecomposition or singular value decomposition. This
-general approach forms the basis for estimation for several GFPCA
-methods {[}references - goldsmith, julia's EM algorithm{]}. However,
-this approach results in rapidly escalating computational burden as both
-the number of subjects, \(N\), and the density of the observed
-functional predictor increases. Note that if the covariance operator
-\(\Sigma_b\) is relatively complex, then \(K\) may be large and not well
-approximated using simple basis functions, \(\phi_k\) (e.g.~random
-intercept and slope).
+small) or faster methods such as fast covariance estimation (FACE)
+(\protect\hyperlink{ref-face}{Xiao et al. 2016}). However, for
+non-Gaussian data, direct estimation is not possible. Instead, one may
+choose to estimate \(b_i(t)\) first using a flexible spline basis
+(e.g.~B-splines) with dimension \(K^B\) such that \(K^B > K\), assuming
+an unstructured correlation between subject-specific spline
+coefficients, then orthogonalize the resulting estimates via
+eigendecomposition or singular value decomposition. This general
+approach forms the basis for estimation for several GFPCA methods
+(\protect\hyperlink{ref-goldsmith2015}{Goldsmith et al. 2015};
+\protect\hyperlink{ref-wrobel2019}{Wrobel et al. 2019}). However, this
+approach results in rapidly escalating computational burden as both the
+number of subjects, \(N\), and the density of the observed functional
+predictor increases. Note that if the covariance operator \(\Sigma_b\)
+is relatively complex, then \(K\) may be large and not well approximated
+using simple basis functions, \(\phi_k\) (e.g.~random intercept and
+slope).
 
@@
 interest, for dynamic prediction one must then estimate the scores for a
-new subject \(u\) not used in model fitting. Typically this takes the
-form of the conditional expectation of the scores
-\(E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]\) where \(\bf{\Theta}\) represents
-the set of parameters estimated by the model (i.e.~parameters associated
-with \(\beta_0(t), \Sigma_b\)) and \(\bf{y}_u\) are the non-Gaussian
-data for the new subject. Estimation of the random effects for a new
-subject is non-trivial and not implemented for even very simple models
-in most random effects software packages such as lme4 {[}reference{]}
-and nlme {[}reference{]}. Our proposed method bypasses this problem by
-using local generalized linear mixed models to estimate the covariance
-operator \(\Sigma_b\) directly on the latent scale, then using the
-closed form solution for the best linear unbiased predictor of Gaussian
-random effects. Estimation is made fast for even very large data through
-the use of the FACE methodology for FPCA on the latent estimates
-obtained from local models. Because the local models are very simple
-GLMMs, estimation is also fast for the local models.
+new subject \(u\) not used in model fitting. Typically for non-Gaussian
+data, this takes the form of the maximum likelihood estimator (MLE) of
+the scores
+\(\hat{\xi}_{uk} = \underset{\xi_{uk}}{\mathrm{argmax}} L(\xi_{uk}|\boldsymbol{\Theta}, \boldsymbol{Y}_{u})\),
+where \(\bf{\Theta}\) represents the set of parameters estimated by the
+model (i.e.~parameters associated with \(\beta_0(t), \Sigma_b\)) and
+\(\boldsymbol{Y}_u\) are the non-Gaussian data for the new subject.
+Estimation of the random effects for a new subject is non-trivial and
+not implemented for even very simple models in most random effects
+software packages such as \emph{lme4}
+(\protect\hyperlink{ref-lme4}{Bates et al. 2015}) and \emph{nlme}
+(\protect\hyperlink{ref-nlme}{Pinheiro et al. 2022}). Our proposed
+method bypasses this problem by using local generalized linear mixed
+models (GLMMs) to estimate the covariance operator \(\Sigma_b\) directly
+on the latent scale, then using the MLE of Gaussian random effects.
+Estimation is made fast for even very large data through the use of the
+FACE methodology for FPCA on the latent estimates obtained from local
+models. Because the local models are very simple GLMMs, estimation is
+also fast for the local models.
 
@@
   \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
-  and obtain estimates of basis functions
+  and obtain \(\hat{\boldsymbol{\Theta}}\) which includes the estimates
+  of basis functions
   \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
@@
 
-\textbf{Ying, include the formulas for your conditional expectation here
-to obtain random effects estimates }
+\textbf{Dynamic prediction} Now assume we have a new observations \(u\)
+with partially observed outcome
+\(\boldsymbol{Y}_u = \{Y_u(t), t\leq t_m\}\). The maximum observed point
+is \(t_m\), which is a point along the functional domain and belongs to
+bin \(s_m\) on the binned grid from step 1 above. With components
+extracted from the FPCA model in step 3, we will be able to make
+predictions on the future, unobserved bins. Specifically, with the
+partially observed binary function \(Y_u(t)\), we will be able to
+calculate the MLE of the subject-specific PC scores
+\(\hat{\boldsymbol{\xi}}_u= \{\hat{\xi}_{u1},...,\hat{\xi}_{uK}\}\) that
+maximizes its posterior likelihood:
 
-\textbf{Dynamic prediction} Now assume we have a new observations with
-partially observed outcome \(Y_{new}(t)\), \(t\leq t_m\). The maximum
-observed point is \(t_m\), which is a point along the functional domain
-and belongs to bin \(s_m\) on the binned grid from step 1 above. With
-components extracted from the FPCA model in step 3, we will be able to
-make predictions on the future, unobserved bins. Specifically, with the
-partially observed binary function \(Y_{new}(t)\), we will be able to
-calculate the maximum likelihood estimate (MLE) of the subject-specific
-PC scores \(\hat{\xi}_{ik}\). Then the value of latent functions at
-unobserved points can be estimated as
-\(\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)\)
+\[\hat{\boldsymbol{\xi}}_u = \underset{\boldsymbol{\xi}_u}{\mathrm{argmin}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
+\hat{\boldsymbol{\Theta}})= \frac{p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})}{\int p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u,\hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})d\boldsymbol{\xi}_u}\]
+
+Then the value of latent functions at unobserved points can be estimated
+as
+\(\hat{\eta_u}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{uk}{\phi}_k(s)\).
+In fact, since the denominator in the formula above is constant with
+respect to \(\boldsymbol{\xi}_u\), we only need to maximum the
+numerator. Since \(\boldsymbol{Y}_u|\boldsymbol{\xi}_u\) follows an
+exponential family distribution and \(\boldsymbol{\xi}_u\) follows
+multivariate Guassian distribution, it could be numerically maximized
+efficiently.
 
@@
 Here we set the values of \(\lambda_k\) to be \(0.5^{k-1}\),
-\(k \in (1,..., 4)\). In addition, for simplicity of presentation, we
+\(k \in \{1,..., 4\}\). In addition, for simplicity of presentation, we
 generate data on a regular grid, which means observations points are
@@
 As a reference method, we compare our method to Generalized Linear Mixed
-Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
-the fastest existing method developed for dynamic prediction of repeated
-generalized outcomes. Just like many mixed models, this method is very
-limited in terms of flexibility if one wishes to obtain estimates in a
-reasonable time frame. For example, the model used for prediction of our
-simulated datset would simply be an linear model with one covariate
-indicating observation time:
+Models using Adaptive Gaussian Quadrature
+(GLMMadaptive)(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}).
+This is one of the fastest existing method developed for dynamic
+prediction of repeated generalized outcomes. Just like many mixed
+models, this method is very limited in terms of flexibility if one
+wishes to obtain estimates in a reasonable time frame. For example, the
+model used for prediction of our simulated datasets would simply be an
+linear model with one covariate indicating observation time:
 \(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
@@
 \includegraphics{Manuscript_edit_files/figure-latex/Figure-1.pdf}
-\caption{Predicted track of four randomly selected subjects from one
+\caption{Predicted tracks of four randomly selected subjects from one
 simulated dataset. The grey solid line indicates the true latent
-continuous function. The dashed lines indicated predicted latent
-function tracks, and color indicates different observation time.}
+continuous function. The dashed lines indicate predicted latent function
+tracks, and color indicates different observation time.}
 \end{figure}
@@
 
-\caption{\label{tab:format table}Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.}
+\caption{\label{tab:format table}Predictive performance of fGFPCA and GLMMadaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.}
 \centering
@@
 (\protect\hyperlink{ref-bothwell2022}{Bothwell et al. 2022};
-\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The study
+\protect\hyperlink{ref-ostendorf2022}{Ostendorf et al. 2022}). The trial
 originally aimed to study the effect of dietary strategies on weight
@@
 (number of days into the study). We will fit both fGFPCA and
-GLMMadaptive models and compared their performance on this data. For
+GLMMadaptive models and compare their performance on this data. For
 fGFPCA, every 10 observations are binned together, resulting in 40 bins
@@
 continuous function tracks are not known in this case, prediction
-performance on evaluated with AUC alone.
+performance is evaluated with AUC alone.
 
 Table 2 compares the predictive performance of both models. Similar to
-the simulatio study, prediction is made conditioning on observations up
+the simulation study, prediction is made conditioning on observations up
 to the 100th, 200th and 300th day respectively, and evaluation is made
@@
 performed similarly when predicting recent days, however fGFPCA did
-better in predicting further days along the trial (prediction time
-window of \((300, 400]\)). As more observations coming in, fGFPCA sees
-increasingly higher AUC compared to GLMMadaptive model. This could be a
-result of the difference in model flexibility of the two methods. While
-the linear relationship in underlying latent functions are not likely to
-change immediately following the observed track, the whole track could
-be much more complex. Therefore, while fGFPCA performs similarly to
-GLMMadaptive right after the end of observation, it does a much better
-job capturing the underlying pattern along the whole track. Figure 2
-further illustrate such difference using predicted latent function
-tracks of four randomly drawn subjects.
+better in predicting further days along the trial. As more observations
+coming in, fGFPCA sees increasingly higher AUC compared to GLMMadaptive
+model. This is a result of the difference in model flexibility of the
+two methods. While the linear relationship in underlying latent
+functions are not likely to change immediately following the observed
+track, the whole track could be much more complex. Therefore, while
+fGFPCA performs similarly to GLMMadaptive right after the end of
+observation, it does a much better job capturing the underlying pattern
+along the whole track. Figure 2 further illustrate such difference using
+predicted latent function tracks of four randomly drawn subjects.
 
@@
 \includegraphics{Manuscript_edit_files/figure-latex/Figure_appl-1.pdf}
-\caption{Predicted track of four randomly selected subjects from the
-daily weigh-in dataset. The grey solid line indicates the true latent
-continuous function. The dashed lines indicated predicted latent
-function tracks, and color indicates different observation time.}
+\caption{Predicted tracks of four randomly selected subjects from the
+daily weigh-in dataset. The grey points are observed binary outcomes.
+The dashed lines indicate predicted latent function tracks, and color
+indicates different observation time.}
 \end{figure}
@@
 
-\caption{\label{tab:unnamed-chunk-3}AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset}
+\caption{\label{tab:unnamed-chunk-3}AUC of fGFPCA and GLMMadaptive on the daily weigh-in dataset}
 \centering
@@
 The simulation study and data application above have illustrated the
-feasibility and utility of the proposed method, even for large datasets
-and complex latent processes/random effect structures. can achieve
-better predictive performance with much less computational burden.
-Compared to competing methods, fGFPCA can accommodate extremely flexible
-correlation structure between repeated measure, which is particularly
-useful when outcome is generated from latent functions with highly
-non-linear patterns. Due to the comparative flexibility of the fGFPCA
-approach, fGFPCA has much lower ISE and higher AUC than GLMM. As
+feasibility and utility of the proposed method. Even for large datasets
+and complex latent processes/random effect structures, fGFPCA can
+achieve better predictive performance with much less computational
+burden. Compared to competing methods, fGFPCA can accommodate extremely
+flexible correlation structure between repeated measure, which is
+particularly useful when outcome is generated from latent functions with
+highly non-linear patterns. Due to the comparative flexibility of the
+fGFPCA approach, fGFPCA has much lower ISE and higher AUC than GLMM. As
 expected, the dynamic prediction accuracy of fGFPCA improves as more
-data are availble for predictions. Moreover, fGFPCA also significantly
+data are available for predictions. Moreover, fGFPCA also significantly
 reduce time spent on both model fitting and prediction, allowing for
@@
 datasets with compatible flexibility.
-
-\textbf{Andrew: I would delete this paragraph, honestly}
-
-While the proposed method achieved good accuracy of point prediction, it
-is more complicated to develop a straightforward, interpretable metric
-to evaluate the precision of prediction. The challenge arises from the
-fact that fitting local GLMMs (Step 2) and FPCA (Step 3) can both
-introduce uncertainty to the final prediction. Since they are
-implemented consecutively, it is not clear how the uncertainty from both
-procedures should be integrated into one single interval estimator. In
-practice, one could consider developing interval predictions conditional
-on the local GLMMs, quantifying uncertainty of individual score
-estimates from FPCA model alone. Since the score \(\hat{\xi}_{ik}\) is
-essentially a maximum likelihood estimate, it is natural to estimate its
-variance based on likelihood theory with observed information
-\({I}(\hat{\xi}_{ik})\). Therefore, the variation of prediction interval
-is:
-
-\[Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2\]
-Where \(\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))\) and
-\(\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})\).
 
@@
 along the functional grid. Moreover, if the bins are very narrow with
-only a few observations, local models may be non-identifiabile,
-resulting in bins with no point estimates. While this may not overly
-impact point final predictions, it is currently unclear as to when this
-issue warrants concern. In practice, we expect this to happen when the
+only a few observations, local models may be non-identifiable, resulting
+in bins with no point estimates. While this may not overly impact final
+point predictions, it is currently unclear as to when this issue
+warrants concern. In practice, we expect this to happen when the
 measurements are sparse or sample size is small. However, the
 methodology here is designed for large, dense functional data and thus
-not currently appropriate in such scenarios. In this case, existing
+not currently appropriate in such scenarios, in which case existing
 methods may be sufficiently flexible and computationally feasible.
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-lme4}{}}%
+Bates, D., Mächler, M., Bolker, B., and Walker, S. (2015), {``Fitting
+linear mixed-effects models using {lme4},''} \emph{Journal of
+Statistical Software}, 67, 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-bothwell2022}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-davidian2003}{}}%
+Davidian, M., and Giltinan, D. M. (2003),
+{``\href{http://www.jstor.org/stable/1400665}{Nonlinear models for
+repeated measurement data: An overview and update},''} {[}International
+Biometric Society, Springer{]}, 8, 387--419.
+
 \leavevmode\vadjust pre{\hypertarget{ref-delaigo2016}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-refundpck}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-refundpkg}{}}%
 Goldsmith, J., Scheipl, F., Huang, L., Wrobel, J., Di, C., Gellar, J.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fcrpck}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-fcrpkg}{}}%
 Leroux, A., Xiao, L., Crainiceanu, C., and Checkly, W. (2018b),
@@
 \leavevmode\vadjust pre{\hypertarget{ref-liang1986}{}}%
-LIANG, K.-Y., and ZEGER, S. L. (1986), {``Longitudinal data analysis
+Liang, K.-Y., and Zeger, S. L. (1986), {``Longitudinal data analysis
 using generalized linear models,''} \emph{Biometrika}, 73, 13--22.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-davidian2003}{}}%
-{``\href{http://www.jstor.org/stable/1400665}{Nonlinear models for
-repeated measurement data: An overview and update}''} (2003),
-{[}International Biometric Society, Springer{]}, 8, 387--419.
-
 \leavevmode\vadjust pre{\hypertarget{ref-ostendorf2022}{}}%
@@
 \url{https://doi.org/10.1186/s13063-022-06523-2.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlme}{}}%
+Pinheiro, J., Bates, D., and R Core Team (2022),
+\emph{\href{https://CRAN.R-project.org/package=nlme}{Nlme: Linear and
+nonlinear mixed effects models}}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-face}{}}%
+Xiao, L., Zipunnikov, V., Ruppert, D., and Crainiceanu, C. (2016),
+{``Fast covariance estimation for high-dimensional functional data,''}
+\emph{Statistics and Computing}, 26, 409--421.
+https://doi.org/\url{https://doi.org/10.1007/s11222-014-9485-x}.
+
 \end{CSLReferences}

### 2023-04-14 Code/OutSampMLE.R
delete files

@@ -17,7 +17,7 @@ library(rootSolve)
 
 llh_div <- function(xi, df_new, fpca_fit){
   
-  # observations
+  # observations 
   ns <- as.vector(table(df_new$bin)) # number of observations
   hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
     select(Y) %>% unlist()# number of success

### 2023-05-17 Docs/FDAgroup.Rmd
calculated AUC for NHANES data

--- 2022-11-08 FDAgroup.Rmd
@@
   labs(title = "Prediction with observation up to t=0.8",
-       x = "time", y = "")
+       x = "time", y = "Length for age z-score")
+
+```
+
+```{r}
+bind_rows(df_test %>%
+            select(argvals, subj, Y)%>%
+            rename("y" = "Y") %>%
+            mutate(type = "true", subj = as.factor(subj)),
+          df_test_t0.3 %>% mutate(subj = as.factor(subj),
+                                  type = "Up to t=0.3"),
+          df_test_t0.5 %>% mutate(subj = as.factor(subj),
+                                  type = "Up to t=0.5"),
+          df_test_t0.8 %>% mutate(subj = as.factor(subj),
+                                  type = "Up to t=0.8")) %>%
+  filter(subj == 152) %>%
+  ggplot(aes(x = argvals, y = y, col = type))+
+  geom_line()+
+  labs(title = "Dynamic prediction of child growth data",
+       x = "Time", y = "Length for age z-score")
+
+ggsave(here("Images/dynpred_exp.png"), width = 8, height = 6, bg = "white")
 

### 2023-05-17 Docs/Manuscript_edit.Rmd
calculated AUC for NHANES data

--- 2023-04-07 Write-ups/Manuscript_edit.Rmd
@@
 
-$$\hat{\boldsymbol{\xi}}_u = \underset{\boldsymbol{\xi}_u}{\mathrm{argmin}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
+$$\hat{\boldsymbol{\xi}}_u = \underset{\boldsymbol{\xi}_u}{\mathrm{argmax}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
 \hat{\boldsymbol{\Theta}})= \frac{p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})}{\int p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u,\hat{\boldsymbol{\Theta}})p(\boldsymbol{\xi}_u|\hat{\boldsymbol{\Theta}})d\boldsymbol{\xi}_u}$$
@@
   geom_line(aes(x=bin, y=eta_i, col = "True"))+
-  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t200, col = "Up to 0.2"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t400, col = "Up to 0.4"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t600, col = "Up to 0.6"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=bin, y = pred_t800, col= "Up to 0.8"), linetype = "dashed", na.rm = T)+
   facet_wrap(~id, scales = "free_y")+
-  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "fGFPCA")+
+  labs(y = "Latent function", x = "Time", colour= "Observed time", title = "fGFPCA")+
   scale_color_manual(values=cbPalette)+
-  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
+  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000), labels = c(0, 0.2, 0.4, 0.6, 0.8, 1.0))
 
@@
   geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
-  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
-  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 0.2"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 0.4"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 0.6"), linetype = "dashed", na.rm = T)+
+  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 0.8"), linetype = "dashed", na.rm = T)+
   facet_wrap(~id, scales = "free_y")+
-  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "GLMMadaptive")+
+  labs(y = "Latent function", x = "Time", colour= "Observed time", title = "GLMMadaptive")+
   scale_color_manual(values=cbPalette)+
-  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
+  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000),
+                     labels = c(0, 0.2, 0.4, 0.6, 0.8, 1.0))
 
@@
 ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
-colnames(ave_ise_mat) <- c("200", "400", "600", "800")
-rownames(ave_ise_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+colnames(ave_ise_mat) <- c("0.2", "0.4", "0.6", "0.8")
+rownames(ave_ise_mat) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")
 
@@
 ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
-colnames(ave_ise_mat_ref) <- c("200", "400", "600", "800")
-rownames(ave_ise_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+colnames(ave_ise_mat_ref) <- c("0.2", "0.4", "0.6", "0.8")
+rownames(ave_ise_mat_ref) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")
 
@@
 ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
-colnames(ave_auc_mat) <- c("200", "400", "600", "800")
-rownames(ave_auc_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+colnames(ave_auc_mat) <- c("0.2", "0.4", "0.6", "0.8")
+rownames(ave_auc_mat) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")
 
@@
 ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
-colnames(ave_auc_mat_ref) <- c("200", "400", "600", "800")
-rownames(ave_auc_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")
+colnames(ave_auc_mat_ref) <- c("0.2", "0.4", "0.6", "0.8")
+rownames(ave_auc_mat_ref) <- c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]")
 
@@
   facet_wrap(~id, scales = "free_y")+
-  labs(y = "Predicted latent function on probability scale", x = "Bin", 
+  labs(y = "Predicted latent function on probability scale", x = "Time", 
        colour= "Observed time", title = "fGFPCA")+
@@
   facet_wrap(~id, scales = "free_y")+
-  labs(y = "Predicted latent function on probability scale", x = "Bin", 
+  labs(y = "Predicted latent function on probability scale", x = "Time", 
        colour= "Observed time", title = "GLMMadaptive")+

### 2023-05-17 Docs/ProgressReport.Rmd
calculated AUC for NHANES data

---
title: "Progress Report"
author: "Ying Jin"
date: "2023-05-16"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(516)

library(here)
library(tidyverse)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
theme_set(theme_minimal())

df <- read_rds(here("Data/nhanes_bi.rds"))
load(here("Data/ApplOutput_fGFPCA.RData"))

```

# NHANES data application

## Failed prediction

- Laplace approximation failed for some subjects
- Only when the observed track is short (up to 360)
- These subjects have something in common

```{r}
# compare subjects that failed at Laplace Approximation with others
rand_id <- sample(unique(df$SEQN), size = 4)
com_id <- c(rand_id, sample(skip_id, 4))

df_est_latent %>% filter(id %in% com_id) %>%
  left_join(df %>% select(SEQN, Z, sind) %>%
              rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>%
  mutate(type = ifelse(id %in% rand_id, "Prediction", "Failed prediction")) %>%
  ggplot(aes(x=bin, y=eta_hat, col = type))+
  geom_line()+
  geom_vline(xintercept = 360)+
  facet_wrap(~type+id, nrow = 2, ncol = 4)
```

## Subjects with complete prediction

```{r fig_prob, warning=FALSE}
df_est_latent %>%
  filter(id %in% rand_id) %>%
  mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080),
         pred_t720 = ifelse(bin<=720, NA, pred_t720),
         pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>%
  mutate_at(vars(pred_t360, pred_t720, pred_t1080), function(x)exp(x)/(1+exp(x))) %>%
  left_join(df %>% select(SEQN, Z, sind) %>%
              rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>%
  ggplot()+
    geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.5)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r fig_eta, warning=FALSE}
df_est_latent %>%
  filter(id %in% rand_id) %>%
  mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080),
         pred_t720 = ifelse(bin<=720, NA, pred_t720),
         pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>%
  ggplot()+
    geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+
    geom_line(aes(x=bin, y = eta_hat, col = "eta_hat"), linetype = "solid" )+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function")
```

## AUC on subjects with complete prediction

- Use interpolation to fill in gaps between bins

```{r}
# subjects with complete prediction
N <- length(unique(df$SEQN))-length(skip_id)
eval_id <- setdiff(unique(df$SEQN), skip_id)
J <- max(df$sind)

# containter: row index subject, column index time
eta_mat_t360 <- eta_mat_t720 <- eta_mat_t1080 <- matrix(NA, nrow = N, ncol = J)

# interpolation
for(i in seq_along(eval_id)){
   eta_mat_t360[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
                               y=df_est_latent$pred_t360[df_est_latent$id==eval_id[i]], 
                               xout = 1:J)$y
   eta_mat_t720[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
                               y=df_est_latent$pred_t720[df_est_latent$id==eval_id[i]], 
                               xout = 1:J)$y
   eta_mat_t1080[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
                               y=df_est_latent$pred_t1080[df_est_latent$id==eval_id[i]], 
                               xout = 1:J)$y
  }

```

```{r}
# true outcome
Y_mat <- df$Z[df$SEQN %in% eval_id]
Y_mat <- matrix(Y_mat, nrow = N, ncol = J, byrow = T) 
```

```{r AUC_fGFPCA}
# maximum observations time
t_vec <- c(360, 720, 1080)

# AUC container
auc_mat <- matrix(NA, 3, 3)
colnames(auc_mat) <- c("360", "720", "1080")
rownames(auc_mat) <- c("(360, 720]", "(720, 1080]", "(1080, 1440]")

# Calculate AUC

auc_mat[1, 1] <- performance(prediction(as.vector(eta_mat_t360[, 361:720]), 
                                   as.vector(Y_mat[, 361:720])),
                        "auc")@y.values[[1]]
auc_mat[2, 1] <- performance(prediction(as.vector(eta_mat_t360[, 721:1080]), 
                                   as.vector(Y_mat[, 721:1080])),
                        "auc")@y.values[[1]]
auc_mat[3, 1] <- performance(prediction(as.vector(eta_mat_t360[, 1081:1435]), 
                                   as.vector(Y_mat[, 1081:1435])),
                        "auc")@y.values[[1]]

auc_mat[2, 2] <- performance(prediction(as.vector(eta_mat_t720[, 721:1080]), 
                                   as.vector(Y_mat[, 721:1080])),
                        "auc")@y.values[[1]]
auc_mat[3, 2] <- performance(prediction(as.vector(eta_mat_t720[, 1081:1435]), 
                                   as.vector(Y_mat[, 1081:1435])),
                        "auc")@y.values[[1]]

auc_mat[3, 3] <- performance(prediction(as.vector(eta_mat_t1080[, 1081:1435]), 
                                   as.vector(Y_mat[, 1081:1435])),
                        "auc")@y.values[[1]]
auc_mat
```

## Reference method

- GLMMadaptive?
- fcr on lowess smoothed track?

### 2023-05-17 Docs/manuscript.Rmd
calculated AUC for NHANES data

---
title: "Dynamic Prediction with Fast Generalized Functional Principal Component analysis"
author: "Ying Jin"
date: "2023-03-07"
output: 
  pdf_document:
    citation_package: natbib
    keep_text: true
    fig_caption: true
    latex_engine: pdflatex
    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-ms.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2023-05-17 Code/DataApplNHANES.R

@@ -1,6 +1,6 @@
 # data application of NHANES
 
-set.seed(404)
+set.seed(516)
 
 library(here)
 library(tidyverse)
@@ -12,7 +12,8 @@ theme_set(theme_minimal())
 
 # code
 source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
-source(here("Code/OutSampMLE.R"))
+# source(here("Code/OutSampMLE.R"))
+source(here("Code/OutsampBayes.R"))
 
 # data 
 df <- read_rds(here("Data/nhanes_bi.rds"))
@@ -30,7 +31,10 @@ df %>%
   ggplot()+
   geom_point(aes(x=sind, y=Z), size = 0.5)+
   facet_wrap(~SEQN)+
-  labs(x="Time", y = "Activity")
+  labs(x="Time", y = "Activity")+
+  theme(plot.margin = margin(0, 1, 0, 0))
+ggsave(here("Images/NHANES/data.png"), width=7, height=5, bg="white")
+
 
 
 ##### fGFPCA #####
@@ -44,6 +48,7 @@ mid <- (brks+bin_w/2)[1:n_bin] # mid points
 
 df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
 df$bin <- as.numeric(as.character(df$bin))
+head(df, 20)
 
 df <- df %>% rename(id = SEQN, Y=Z)
 
@@ -52,64 +57,165 @@ df_bin_lst <- split(df, f = df$bin)
 lapply(df_bin_lst, function(x)table(x$Y))
 
 # fit local GLMM and estimate latent function
-# near-unidentifiability issues !!!
+# near-unidentifiability issues 
 df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
-df_est_latent <- bind_rows(df_est_latent) %>% select(id, bin, eta_hat) %>% distinct() # on the binned grid, one unique value each bin
+lapply(df_est_latent, function(x)range(x$eta_hat))
+
+# on the binned grid, one unique value each bin
+df_est_latent <- bind_rows(df_est_latent) %>% select(id, bin, eta_hat) %>% distinct() 
+
 df_est_latent %>%
-  left_join(df, by = c("id", "bin")) %>%
+  left_join(df %>% filter(sind==bin), by = c("id", "bin")) %>%
   filter(id %in% rand_id) %>%
   ggplot()+
[…]

### 2023-05-17 Code/OutSampBayes.R

@@ -65,7 +65,7 @@ Model <- function(parm, Data){
 
 
 # data: 
-out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=195) %>% select(-eta_i)){
+out_pred_laplace <- function(fpca_fit, df_new){
   
   # put data into correct format
   ns <- as.vector(table(df_new$bin)) # number of observations
@@ -83,7 +83,11 @@ out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=
   X <- fpca_fit$efunctions[1:max_bin, ] # "covariates", in this case PC functions
   J <- ncol(fpca_fit$efunctions) # number of parameters/scores
   mon.names <- "LP"
-  parm.names <- as.parm.names(list(xi1=0, xi2=0, xi3=0, xi4=0))
+  
+  # parameter names
+  name_lst <- as.list(rep(0, J))
+  names(name_lst) <- paste("xi", 1:J, sep = "")
+  parm.names <- as.parm.names(name_lst)
   pos.xi <- grep("xi", parm.names)
   PGF <- function(Data) {
     xi <- rnorm(Data$J)
@@ -95,7 +99,7 @@ out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=
   
   
   # fit laplace approximation
-  Fit <- LaplaceApproximation(Model, parm = rep(0, 4), Data=MyData)
+  Fit <- LaplaceApproximation(Model, parm = rep(0, J), Data=MyData)
   score <- Fit$Summary1[, "Mode"]
   
   # prediction
@@ -114,6 +118,6 @@ out_pred_laplace <- function(fpca_fit, df_new = df %>% filter(id==1 & sind_inx<=
 
 # test
 
-# out_pred_laplace(fpca_fit,  df %>% filter(id==2 & sind_inx<=395) %>% select(-eta_i))
-
+# score_out <- out_pred_laplace(fpca_fit,  df %>% filter(id==rand_id[1] & sind<=720))$score_out
+# eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions%*%score_out
 

### 2023-05-17 Code/OutSampMLE.R

@@ -7,6 +7,33 @@ library(rootSolve)
 # of an out-of-sample observation
 # just for mess around
 
+#### Likelihood function ####
+joint_pdf <- function(xi, df_new, fpca_fit){
+  
+  # observations 
+  ns <- as.vector(table(df_new$bin)) # number of observations
+  hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+    select(Y) %>% unlist()# number of success
+  nf <- ns-hs # number of failure
+  max_bin <- length(unique(df_new$bin)) # assume no skipped bins 
+  
+  # numbers
+  ns <- table(df_out$bin) # number of observations
+  hs <- df_out %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+    select(Y) # number of success
+  nf <- ns-hs # number of failure
+  
+  # probability of conditional distribution of Y
+  eta = f0+phi%*%xi
+  pr = exp(eta)/(1+exp(eta))
+  # the part relevant to P(Y|xi)
+  p1 <- prod(dbinom(x=unlist(hs), size = ns, prob=pr))
+  p2 <- exp(-t(xi)%*%solve(tao)%*%xi/2)
+  
+  return(p1*p2)
+  
+}
+
 
 #### MLE of out-of-sample score ####
 

### 2023-05-17 Scripts/Figures.R

@@ -1,135 +1,175 @@
-# compare point prediction with in-sample
 
-in_pred_subj1 %>% select(mid, type, eta, id, true)
 
-p1 <- bind_rows(out_pred_subj1 %>% mutate(sample="Out-of-sample prediction"),
-          in_pred_subj1 %>% select(mid, type, eta, id, true) %>%
-            mutate(sample = "In-sample estimation")) %>%
+
+#### Example figures for presentation ####
+
+library(here)
+library(tidyverse)
+library(mgcv)
+theme_set(theme_minimal())
+
+# binary function
+
+df <- readRDS(here("Data/nhanes_bi.rds"))
+length(unique(df$SEQN))
+length(unique(df$sind))
+
+## all subjects
+
+df1 <- df %>% filter(SEQN == 62161)
+df2 <- df %>% filter(SEQN == 62209)
+
+df1$eta_hat <- predict(loess(Z~sind, df1))
+df2$eta_hat <- predict(loess(Z~sind, df2))
+
+df1$eta_hat[df1$eta_hat<0] <- 0
+df2$eta_hat[df2$eta_hat<0] <- 0
+
+bind_rows(df1, df2) %>%
+  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
   ggplot()+
-  geom_line(aes(x=mid, y=true), col = "black")+
-  geom_line(aes(x=mid, y=eta, col= type, group = type), 
-            linetype="dashed", na.rm = T)+
-  facet_wrap(~sample)+
-  labs(x="", y = "", title = "Subject 1", col = "")
+  geom_point(aes(x=sind, y=Z), size = 0.5)+
+  geom_line(aes(x=sind, y=eta_hat), col = "blue")+
+  facet_wrap(~gender, ncol = 2, nrow = 1)+
+  labs(title = "Binary activity indicator from NHANES data", y = "", x = "Minute")
+ggsave(filename = "bifunc.png", path = "Images", width = 12, height = 6, bg = "white")
 
-p2 <- bind_rows(out_pred_subj2 %>% mutate(sample="Out-of-sample prediction"),
-                in_pred_subj2 %>% select(mid, type, eta, id, true) %>%
-                  mutate(sample = "In-sample estimation")) %>%
+
+# try glm
+library(lme4)
+
+fit_glm <- glmer(Z ~ sind + (1 | sind), data = df, family = binomial)
+
+# continous function
+
+list.files("Data")
[…]

### 2023-09-11 Docs/ProgressReport.Rmd
explore unidentifiability issue

@@
   html_document:
+    self_contained: no
+    number_sections: true
+    code_folding: hide
     toc: true
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 
@@
 
-## Failed prediction
+## Data overview
 
-- Laplace approximation failed for some subjects
-- Only when the observed track is short (up to 360)
-- These subjects have something in common
+- 8763 subjects, 1440 measures each
+- no missingness
 
 ```{r}
-# compare subjects that failed at Laplace Approximation with others
-rand_id <- sample(unique(df$SEQN), size = 4)
-com_id <- c(rand_id, sample(skip_id, 4))
-
-df_est_latent %>% filter(id %in% com_id) %>%
-  left_join(df %>% select(SEQN, Z, sind) %>%
-              rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>%
-  mutate(type = ifelse(id %in% rand_id, "Prediction", "Failed prediction")) %>%
-  ggplot(aes(x=bin, y=eta_hat, col = type))+
-  geom_line()+
-  geom_vline(xintercept = 360)+
-  facet_wrap(~type+id, nrow = 2, ncol = 4)
+N <- length(unique(df$SEQN)) # sample size 8763
+J <- max(df$sind) # 1440 measures per subject
 ```
 
-## Subjects with complete prediction
+```{r}
+rand_id <- sample(unique(df$SEQN), size = 4) # "toy" sample
 
-```{r fig_prob, warning=FALSE}
-df_est_latent %>%
+df %>% 
+  filter(SEQN %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Z), size = 0.5)+
+  facet_wrap(~SEQN)+
+  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
+```
+
+## fGFPCA Estimation
+
+```{r}
+# code
+source(here("Code/GLMM-FPCA.R")) 
+# use pred_latent function to estimate latent function 
+source(here("Code/OutSampMLE.R"))
+# source(here("Code/OutsampBayes.R"))
+
+```
+
+### Binning
+
+- Bin every 10 measures
+
+```{r, message=FALSE}
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
+df$bin <- as.numeric(as.character(df$bin))
+# head(df)
+
+df %>% 
+  filter(SEQN %in% rand_id) %>%
+  group_by(SEQN, bin) %>%
+  summarise(num = sum(Z)) %>%
+  ggplot()+
+  geom_point(aes(x=bin, y=num), size = 0.5)+
+  facet_wrap(~SEQN)+
+  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
+```
+
+### Local GLMMs
+
+- This is where near-unidentifiability issues first appear
+- Though point estimation not necessarily affected
+
+```{r, cache=TRUE, warning=FALSE}
+# split by each bin
+df <- df %>% rename(id = SEQN, Y=Z)
+df_bin_lst <- split(df, f = df$bin)
+
+# fit local GLMM and estimate latent function
+# near-unidentifiability issues 
+df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+df_est_latent <- bind_rows(df_est_latent)
+
+# keep midpoint values
+df_est_latent <- df_est_latent %>%
+  select(-sind, -Y) %>% distinct(.)
+```
+
+```{r, message=FALSE}
+df %>% 
   filter(id %in% rand_id) %>%
-  mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080),
-         pred_t720 = ifelse(bin<=720, NA, pred_t720),
-         pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>%
-  mutate_at(vars(pred_t360, pred_t720, pred_t1080), function(x)exp(x)/(1+exp(x))) %>%
-  left_join(df %>% select(SEQN, Z, sind) %>%
-              rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>%
+  left_join(df_est_latent, by = c("id", "bin")) %>%
+  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
   ggplot()+
-    geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+
-    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.5)+
-    facet_wrap(~id)+
+  geom_line(aes(x=sind, y=eta_hat, group = id))+
+  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+             alpha = 0.5, linewidth = 0.5)+
   labs(x = "Time", y = "Estimated latent function (probablity scale)")
@@
 
-```{r fig_eta, warning=FALSE}
-df_est_latent %>%
+```{r, message=FALSE}
+# overview at estimated latent function
+df %>% 
   filter(id %in% rand_id) %>%
-  mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080),
-         pred_t720 = ifelse(bin<=720, NA, pred_t720),
-         pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  left_join(df_est_latent, by = c("id", "bin")) %>%
+  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
   ggplot()+
-    geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+
-    geom_line(aes(x=bin, y = eta_hat, col = "eta_hat"), linetype = "solid" )+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function")
+  geom_line(aes(x=bin, y=eta_hat, group = id, linetype = "dashed"))+
+  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+             linewidth = 0.5, alpha = 0.6)+
+  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
 ```
 
-## AUC on subjects with complete prediction
+- Take a closer look at the unidentifiable bins?
 
-- Use interpolation to fill in gaps between bins
+I just realized glmer actually used Laplace approximation as well. Could this also be due to numeric stability? In some bins (465, 545, 555, 1185, 1195, 1205), the approximation failed to converge.
+
+Below is the warning messages and one model summary.
+
+For this bin, the problem is with the fixed effect standard error. It is clearly near-zero.
+
+https://www.learn-mlms.com/07-module-7.html
+
+https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
+
+```{r, eval=FALSE}
+try_fit_lst<-list()
+for(i in seq_along(df_bin_lst)){
+  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]],
+                             family = binomial)},
+                       warning=function(w){NA})
+  try_fit_lst[[i]] <- this_glm
+
+}
+
+# bins where glmer returned errors
+# names(df_bin_lst)[is.na(try_fit_lst)]
+```
 
 ```{r}
-# subjects with complete prediction
-N <- length(unique(df$SEQN))-length(skip_id)
-eval_id <- setdiff(unique(df$SEQN), skip_id)
-J <- max(df$sind)
+issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']],
+                             family = binomial)
+summary(issue_fit)
 
-# containter: row index subject, column index time
-eta_mat_t360 <- eta_mat_t720 <- eta_mat_t1080 <- matrix(NA, nrow = N, ncol = J)
+# table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)
+```
 
-# interpolation
-for(i in seq_along(eval_id)){
-   eta_mat_t360[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
-                               y=df_est_latent$pred_t360[df_est_latent$id==eval_id[i]], 
-                               xout = 1:J)$y
-   eta_mat_t720[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
-                               y=df_est_latent$pred_t720[df_est_latent$id==eval_id[i]], 
-                               xout = 1:J)$y
-   eta_mat_t1080[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]],
-                               y=df_est_latent$pred_t1080[df_est_latent$id==eval_id[i]], 
-                               xout = 1:J)$y
+```{r eval=FALSE}
+colSums(table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)==0)
+colSums(table(df_bin_lst[['465']]$id, df_bin_lst[['465']]$Y)==0)
+```
+
+### FPCA
+
+```{r}
+mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
+# row index subject, column binned time
+fpca_mod <- fpca.face(mat_est_unique, pve = 0.95, argvals = mid, var=T)
+
+K <- ncol(fpca_mod$efunctions) # 18 eigenfunctions
+```
+
+```{r}
+# population mean
+plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+```
+
+```{r}
+# eigenfunctions
+par(mfrow=c(2,2))
+plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
+plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+```
+
+```{r}
+# covariance
+heatmap(fpca_mod$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
+        Rowv = NA, Colv = NA, main = "Covariance matrix")
+```
+
+## Dynamic prediction
+
+### Maximum likelihood
+
+- Boundary hitting issues with numeric maximization
+- Warning message: Steady solution not reached
+- Large variation in both predicted values and scores. The numeric solution from maximum likelihood is obviously not stable. It is very likely to fail to converge.
+- Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot.
+
+```{r}
+# Out-of-sample prediction
+# maximum observations time: 360, 720, 1080
+df_pred <- df_est_latent %>% filter(id %in% rand_id)
+df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
+score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
+# dim indexes subject, eigenfunction and max obs time respectively
+```
+
+```{r, warning=FALSE}
+# prediction for a single subject
+for(i in rand_id){
+  df_i <- df %>% filter(id==i) %>% mutate(bin=as.numeric(bin))
+  # prediction
+  pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
+                                 fpca_fit = fpca_mod, K = K)
+  pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
+                                 fpca_fit = fpca_mod, K = K)
+  pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
+                                  fpca_fit = fpca_mod, K = K)
+  # prediction in container
+  df_pred[df_pred$id==i, 'pred_t360'] <- pred_t360$eta_pred
+  df_pred[df_pred$id==i, 'pred_t720'] <- pred_t720$eta_pred
+  df_pred[df_pred$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
+  # score in container
+  # score_out_mat[i, ,1] <- pred_t360$score_out
+  # score_out_mat[i, ,2] <- pred_t720$score_out
+  # score_out_mat[i, ,3] <- pred_t1440$score_out
 }
@@
 ```{r}
-# true outcome
-Y_mat <- df$Z[df$SEQN %in% eval_id]
-Y_mat <- matrix(Y_mat, nrow = N, ncol = J, byrow = T) 
+df_pred$pred_t360[df_pred$bin<=360] <- NA
+df_pred$pred_t720[df_pred$bin<=720] <- NA
+df_pred$pred_t1080[df_pred$bin<=1080] <- NA
+# df_pred %>% filter(id == rand_id[1])
 ```
 
-```{r AUC_fGFPCA}
-# maximum observations time
-t_vec <- c(360, 720, 1080)
-
-# AUC container
-auc_mat <- matrix(NA, 3, 3)
-colnames(auc_mat) <- c("360", "720", "1080")
-rownames(auc_mat) <- c("(360, 720]", "(720, 1080]", "(1080, 1440]")
-
-# Calculate AUC
-
-auc_mat[1, 1] <- performance(prediction(as.vector(eta_mat_t360[, 361:720]), 
-                                   as.vector(Y_mat[, 361:720])),
-                        "auc")@y.values[[1]]
-auc_mat[2, 1] <- performance(prediction(as.vector(eta_mat_t360[, 721:1080]), 
-                                   as.vector(Y_mat[, 721:1080])),
-                        "auc")@y.values[[1]]
-auc_mat[3, 1] <- performance(prediction(as.vector(eta_mat_t360[, 1081:1435]), 
-                                   as.vector(Y_mat[, 1081:1435])),
-                        "auc")@y.values[[1]]
-
-auc_mat[2, 2] <- performance(prediction(as.vector(eta_mat_t720[, 721:1080]), 
-                                   as.vector(Y_mat[, 721:1080])),
-                        "auc")@y.values[[1]]
-auc_mat[3, 2] <- performance(prediction(as.vector(eta_mat_t720[, 1081:1435]), 
-                                   as.vector(Y_mat[, 1081:1435])),
-                        "auc")@y.values[[1]]
-
-auc_mat[3, 3] <- performance(prediction(as.vector(eta_mat_t1080[, 1081:1435]), 
-                                   as.vector(Y_mat[, 1081:1435])),
-                        "auc")@y.values[[1]]
-auc_mat
+```{r, warning=FALSE}
+df_pred %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
+  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
+  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
+  facet_wrap(~id)
 ```
 
-## Reference method
+### Bayes method like Laplace approximation
 
-- GLMMadaptive?
-- fcr on lowess smoothed track?
+```{r, warning=FALSE, message=FALSE, results='hide'}
+df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
+df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
+score_out_mat2 <- array(NA, dim = c(length(rand_id), K, 3))
+
+# unique id
+# prediction for a single subject
+source(here("Code/OutsampBayes.R"))
+
+for(i in rand_id){
+  df_i <- df %>% filter(id==i)
+
+  # prediction 
+  ## up to 360
+  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=360))
+  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t1$score_out
+  df_pred2[df_pred2$id == i, 'pred_t360'] <- eta_pred_t1
+  ## up to 720
+  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=720))
+  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t2$score_out
+  df_pred2[df_pred2$id == i, 'pred_t720'] <- eta_pred_t2
+  ## up to 1080
+  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1080))
+  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t3$score_out
+  df_pred2[df_pred2$id == i, 'pred_t1080'] <- eta_pred_t3
+
+}
+```
+
+```{r}
+df_pred2$pred_t360[df_pred2$bin<=360] <- NA
+df_pred2$pred_t720[df_pred2$bin<=720] <- NA
+df_pred2$pred_t1080[df_pred2$bin<=1080] <- NA
+# df_pred %>% filter(id == rand_id[1])
+```
+
+```{r, warning=FALSE}
+df_pred2 %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
+  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
+  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
+  facet_wrap(~id)
+```
+
+<!-- ## Failed prediction  -->
+
+<!-- - Laplace approximation failed for some subjects -->
+<!-- - Only when the observed track is short (up to 360) -->
+<!-- - These subjects have something in common -->
+
+<!-- ```{r} -->
+<!-- # compare subjects that failed at Laplace Approximation with others -->
+<!-- rand_id <- sample(unique(df$SEQN), size = 4) -->
+<!-- com_id <- c(rand_id, sample(skip_id, 4)) -->
+
+<!-- df_est_latent %>% filter(id %in% com_id) %>% -->
+<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
+<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
+<!--   mutate(type = ifelse(id %in% rand_id, "Prediction", "Failed prediction")) %>% -->
+<!--   ggplot(aes(x=bin, y=eta_hat, col = type))+ -->
+<!--   geom_line()+ -->
+<!--   geom_vline(xintercept = 360)+ -->
+<!--   facet_wrap(~type+id, nrow = 2, ncol = 4)+ -->
+<!--   labs(y = "Estimated latend function", x = "Time", col = "") -->
+<!-- ``` -->
+
+<!-- ## Subjects with complete prediction -->
+
+<!-- ```{r fig_prob, warning=FALSE} -->
+<!-- df_est_latent %>% -->
+<!--   filter(id %in% rand_id) %>% -->
+<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
+<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
+<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
+<!--   mutate_at(vars(pred_t360, pred_t720, pred_t1080), function(x)exp(x)/(1+exp(x))) %>% -->
+<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
+<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
+<!--   ggplot()+ -->
+<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
+<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
+<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
+<!--     geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.5)+ -->
+<!--     facet_wrap(~id)+ -->
+<!--     labs(x = "Time", y = "Estimated latent function (probablity scale)", -->
+<!--          col = "Maximum observation time") -->
+<!-- ``` -->
+
+<!-- ```{r fig_eta, warning=FALSE} -->
+<!-- df_est_latent %>% -->
+<!--   filter(id %in% rand_id) %>% -->
+<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
+<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
+<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
+<!--   ggplot()+ -->
+<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
+<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
+<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
+<!--     geom_line(aes(x=bin, y = eta_hat, col = "eta_hat"), linetype = "solid" )+ -->
+<!--     facet_wrap(~id)+ -->
+<!--     labs(x = "Time", y = "Estimated latent function") -->
+<!-- ``` -->
+
+<!-- ## AUC on subjects with complete prediction -->
+
+<!-- - Use interpolation to fill in gaps between bins -->
+
+<!-- ```{r} -->
+<!-- # subjects with complete prediction -->
+<!-- N <- length(unique(df$SEQN))-length(skip_id) -->
+<!-- eval_id <- setdiff(unique(df$SEQN), skip_id) -->
+<!-- J <- max(df$sind) -->
+
+<!-- # containter: row index subject, column index time -->
+<!-- eta_mat_t360 <- eta_mat_t720 <- eta_mat_t1080 <- matrix(NA, nrow = N, ncol = J) -->
+
+<!-- # interpolation -->
+<!-- for(i in seq_along(eval_id)){ -->
+<!--    eta_mat_t360[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
+<!--                                y=df_est_latent$pred_t360[df_est_latent$id==eval_id[i]],  -->
+<!--                                xout = 1:J)$y -->
+<!--    eta_mat_t720[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
+<!--                                y=df_est_latent$pred_t720[df_est_latent$id==eval_id[i]],  -->
+<!--                                xout = 1:J)$y -->
+<!--    eta_mat_t1080[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
+<!--                                y=df_est_latent$pred_t1080[df_est_latent$id==eval_id[i]],  -->
+<!--                                xout = 1:J)$y -->
+<!--   } -->
+
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- # true outcome -->
+<!-- Y_mat <- df$Z[df$SEQN %in% eval_id] -->
+<!-- Y_mat <- matrix(Y_mat, nrow = N, ncol = J, byrow = T)  -->
+<!-- ``` -->
+
+<!-- ```{r AUC_fGFPCA} -->
+<!-- # maximum observations time -->
+<!-- t_vec <- c(360, 720, 1080) -->
+
+<!-- # AUC container -->
+<!-- auc_mat <- matrix(NA, 3, 3) -->
+<!-- colnames(auc_mat) <- c("360", "720", "1080") -->
+<!-- rownames(auc_mat) <- c("(360, 720]", "(720, 1080]", "(1080, 1440]") -->
+
+<!-- # Calculate AUC -->
+
+<!-- auc_mat[1, 1] <- performance(prediction(as.vector(eta_mat_t360[, 361:720]),  -->
+<!--                                    as.vector(Y_mat[, 361:720])), -->
+<!--                         "auc")@y.values[[1]] -->
+<!-- auc_mat[2, 1] <- performance(prediction(as.vector(eta_mat_t360[, 721:1080]),  -->
+<!--                                    as.vector(Y_mat[, 721:1080])), -->
+<!--                         "auc")@y.values[[1]] -->
+<!-- auc_mat[3, 1] <- performance(prediction(as.vector(eta_mat_t360[, 1081:1435]),  -->
+<!--                                    as.vector(Y_mat[, 1081:1435])), -->
+<!--                         "auc")@y.values[[1]] -->
+
+<!-- auc_mat[2, 2] <- performance(prediction(as.vector(eta_mat_t720[, 721:1080]),  -->
+<!--                                    as.vector(Y_mat[, 721:1080])), -->
+<!--                         "auc")@y.values[[1]] -->
+<!-- auc_mat[3, 2] <- performance(prediction(as.vector(eta_mat_t720[, 1081:1435]),  -->
+<!--                                    as.vector(Y_mat[, 1081:1435])), -->
+<!--                         "auc")@y.values[[1]] -->
+
+<!-- auc_mat[3, 3] <- performance(prediction(as.vector(eta_mat_t1080[, 1081:1435]),  -->
+<!--                                    as.vector(Y_mat[, 1081:1435])), -->
+<!--                         "auc")@y.values[[1]] -->
+<!-- auc_mat -->
+<!-- ``` -->
+
+<!-- ## Reference method -->
+
+<!-- - GLMMadaptive? -->
+<!-- - fcr on lowess smoothed track? -->

### 2023-09-11 Code/Sim_Subset.R

@@ -1,0 +1,86 @@
+set.seed(314)
+
+library(here)
+library(tidyverse)
+# library(ggplot2)
+# theme_set(theme_minimal())
+# library(fcr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+# library(ggpubr)
+# library(gridExtra)
+# library(kableExtra)
+
+
+
+
+#### load simulated data ####
+
+load(here("Data/sim_data.RData"))
+
+
+#### simulation set up ####
+
+J <- 1000 # number of observations points
+N <- 500 # sample size
+M <- 50 
+
+
+#### GLMM adaptive ####
+
+# cubic spline basis functions
+df_basis <- smoothCon(s(sind, bs = "cr", k = 20), data=sim_data[[1]])[[1]]$X
+colnames(df_basis) <- paste("B", 1:20, sep="")
+this_df <- data.frame(sim_data[[1]] %>% select(id, Y), df_basis)
+fix_f <- paste("Y ~", 
+                paste("B", 1:20, sep = "", collapse = " + "))
+fix_f <- as.formula(fix_f)
+rand_f <- paste
+
+adglmm <- mixed_model(this_f, random = ~ 1 | id, data =  this_df, family = binomial())
+
+
+for(m in 1:M){
+  df <- sim_data[[m]]
+  
+  t1 <- Sys.time()
+  # GLMM adaptive model
+  adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df, family = binomial())
+  
+  # container
+  df_est_latent <- df
+  df_est_latent[, 'pred_t200'] <- df_est_latent[, 'pred_t400'] <- df_est_latent[, 'pred_t600'] <- df_est_latent[, 'pred_t800'] <- NA 
+  
+  # time up to t
+  pred_t200 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 200),
+                       newdata2 =  df %>% filter(sind_inx > 200), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
+  df_est_latent$pred_t200[df_est_latent$sind_inx>200] <- pred_t200$newdata2$pred
+  
+  pred_t400 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 400),
+                       newdata2 =  df %>% filter(sind_inx > 400), type = "subject_specific", type_pred = "link",
[…]

### 2023-09-12 Docs/ProgressReport.Rmd
explore local GLMM unidentifiability issue

@@
   ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, group = id, linetype = "dashed"))+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
   geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
@@
 
-- Take a closer look at the unidentifiable bins?
-
-I just realized glmer actually used Laplace approximation as well. Could this also be due to numeric stability? In some bins (465, 545, 555, 1185, 1195, 1205), the approximation failed to converge.
-
-Below is the warning messages and one model summary.
-
-For this bin, the problem is with the fixed effect standard error. It is clearly near-zero.
-
-https://www.learn-mlms.com/07-module-7.html
-
-https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
+#### Take a closer look at the unidentifiable bins
 
@@
 # bins where glmer returned errors
-# names(df_bin_lst)[is.na(try_fit_lst)]
+names(df_bin_lst)[is.na(try_fit_lst)]
 ```
 
-```{r}
-issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']],
-                             family = binomial)
+The fitting problem occurs in the following bins: 465, 545, 555, 1185, 1195, 1205.
+All of them had two warnings: 1) Non-convergence; 2) near-unidenetifiability.
+Summary results show that standar error estimates of fixed effects (intercept) are very close two zero.
+
+Below is an example of bin 465:
+
+```{r class.source='fold-show'}
+issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial)
+
 summary(issue_fit)
-
-# table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)
 ```
 
-```{r eval=FALSE}
-colSums(table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)==0)
-colSums(table(df_bin_lst[['465']]$id, df_bin_lst[['465']]$Y)==0)
+However, this issue seems to disappear after setting nAGO=2.
+This argument controls the number of nodes in adaptive Gauss-Hermite quadrature approximation.
+This also seems to be faster and more accurate!
+
+Reference: https://arxiv.org/pdf/2202.07864.pdf
+
+```{r class.source='fold-show'}
+issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial, nAGQ=2)
+
+summary(issue_fit)
 ```
+
+<!-- https://www.learn-mlms.com/07-module-7.html -->
+
+<!-- https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html -->
+
+So I decided to re-do latent function estimation while adjusting the nAGQ parameter. Below are some findings:
+
+- nAGQ=0 for all bins: no warnings appeared. This is supposed to be fast but not accurate
+- nAGQ=1, 2: warnings appeared in some bins.
+
+```{r, eval=FALSE}
+# re-fit local GLMM and check numeric issues
+try_fit_lst<-list()
+for(i in seq_along(df_bin_lst)){
+  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=0)},
+                       warning=function(w){NA})
+  try_fit_lst[[i]] <- this_glm
+
+}
+
+# bins where glmer returned errors
+names(df_bin_lst)[is.na(try_fit_lst)]
+lapply(head(try_fit_lst), summary)
+```
+
+So perhaps we could try either setting nAGQ = 0 or a variable nAGQ:
+
+Below is the result for nAGQ = 0:
+
+```{r}
+# re-define pre_latent funciton with nAGQ = 0
+pred_latent_0AGQ <- function(df){
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ = 0)
+    eta_hat <- predict(this_glm, type = "link")
+    df$eta_hat <- eta_hat
+    return(df)
+}
+
+# re-fit local GLMMs
+df_est_latent_0AGQ <- lapply(df_bin_lst, function(x){pred_latent_0AGQ(x)}) 
+# head(df_est_latent_0AGQ[[1]])
+df_est_latent_0AGQ <- bind_rows(df_est_latent_0AGQ)
+
+# keep midpoint values
+df_est_latent_0AGQ <- df_est_latent_0AGQ %>%
+  select(-sind, -Y) %>% distinct(.)
+```
+
+```{r, message=FALSE}
+df %>% 
+  filter(id %in% rand_id) %>%
+  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
+  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_hat, group = id))+
+  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+             # alpha = 0.5, linewidth = 0.5)+
+  labs(x = "Time", y = "Estimated latent function (probablity scale)")
+```
+
+```{r, message=FALSE}
+# overview at estimated latent function
+df %>% 
+  filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
+  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+  #            linewidth = 0.5, alpha = 0.6)+
+  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
+```
+
+Obviously the lower bound of estimated latent function got a lot higher. Nothing else seems to change much.
+
+For FPCA models below, I'll use the result from nAGQ=0 since it is free of fitting issues.
 
@@
 ```{r}
-mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
+mat_est_unique <- matrix(df_est_latent_0AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
 # row index subject, column binned time
@@
 
+Compared the nAGQ=1:
+
+- The lower bound of mean function went up
+- Shapes of PC functions didn't change a lot
+- Covariance matrix looks a lot more different! Near-diagonal correlation got higher and off-diagonal got lower.
+
 ## Dynamic prediction
@@
 - Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot. 
+- After setting nAGQ=0, the problem got better for some subjects, but not the others.
 
@@
 # maximum observations time: 360, 720, 1080
-df_pred <- df_est_latent %>% filter(id %in% rand_id)
+df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
 df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
@@
 
+# df_pred
+
 ```
@@
   geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
-  facet_wrap(~id)
+  facet_wrap(~id,scales = "free")
 ```
@@
 ```{r, warning=FALSE, message=FALSE, results='hide'}
-df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
+df_pred2 <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
 df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA

### 2023-09-12 Docs/ProgressReport.Rmd
more exploration on local GLMM unidentifiability

@@
 library(tidyverse)
+library(ggpubr)
 library(refund)
@@
 
-However, this issue seems to disappear after setting nAGO=2.
+However, this issue seems to disappear after changing nAGQ.
 This argument controls the number of nodes in adaptive Gauss-Hermite quadrature approximation.
-This also seems to be faster and more accurate!
+If nAGQ=0, optimitization is done through penalized iteratively reweighted least square (PIRLS); if nAGQ=1, Laplace approximation and if nAGQ>1, adaptive Gauss-Hermite Quadrature.
 
-Reference: https://arxiv.org/pdf/2202.07864.pdf
+Reference: https://arxiv.org/pdf/2202.07864.pdf (In fact based on this reference the recommended number of node is 0)
+
+Another reference: https://stats.stackexchange.com/questions/544937/when-is-it-appropriate-to-set-nagq-0-in-glmer.
+According to this one, when setting nAGQ=0, fixed effect parameters are not optimized.
+
+For example set nAGQ=2 for bin 465:
 
@@
 
-- nAGQ=0 for all bins: no warnings appeared. This is supposed to be fast but not accurate
-- nAGQ=1, 2: warnings appeared in some bins.
+- nAGQ=0 for all bins: no warnings appeared. This is supposed to be fast but not accurate. 
+- nAGQ=1, 2, 3: warnings appeared in some bins. 
+- nAGQ=5: no warnings!
 
@@
 for(i in seq_along(df_bin_lst)){
-  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=0)},
+  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
                        warning=function(w){NA})
@@
 names(df_bin_lst)[is.na(try_fit_lst)]
-lapply(head(try_fit_lst), summary)
+glmer(Y ~ 1 + (1|id), data = df_bin_lst[['565']], family = binomial, nAGQ=3)
 ```
@@
 
-Below is the result for nAGQ = 0:
+Below compares the result for nAGQ = 0 and nAGQ = 5:
 
-```{r}
+```{r, cache=TRUE}
 # re-define pre_latent funciton with nAGQ = 0
@@
 # re-fit local GLMMs
+t1=Sys.time()
 df_est_latent_0AGQ <- lapply(df_bin_lst, function(x){pred_latent_0AGQ(x)}) 
+t2=Sys.time()
 # head(df_est_latent_0AGQ[[1]])
 df_est_latent_0AGQ <- bind_rows(df_est_latent_0AGQ)
+t_0AGQ <- t2-t1
 
@@
 
+```{r, cache=TRUE}
+# re-define pre_latent funciton with nAGQ = 0
+pred_latent_5AGQ <- function(df){
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ = 5)
+    eta_hat <- predict(this_glm, type = "link")
+    df$eta_hat <- eta_hat
+    return(df)
+}
+
+# re-fit local GLMMs
+t1=Sys.time()
+df_est_latent_5AGQ <- lapply(df_bin_lst, function(x){pred_latent_5AGQ(x)}) 
+t2=Sys.time()
+# head(df_est_latent_0AGQ[[1]])
+df_est_latent_5AGQ <- bind_rows(df_est_latent_5AGQ)
+t_5AGQ <- t2-t1
+
+# keep midpoint values
+df_est_latent_5AGQ <- df_est_latent_5AGQ %>%
+  select(-sind, -Y) %>% distinct(.)
+```
+
+- Time spent: `r t_0AGQ` for nAGQ=0, `r t_5AGQ` for nAGQ=5
+
 ```{r, message=FALSE}
+ggarrange(
+# nAQG=0
 df %>% 
@@
              # alpha = 0.5, linewidth = 0.5)+
-  labs(x = "Time", y = "Estimated latent function (probablity scale)")
+  labs(x = "Time", y = "Estimated latent function (probablity scale)", title = "nAGQ=0"),
+#nAGQ=5
+df %>% 
+  filter(id %in% rand_id) %>%
+  left_join(df_est_latent_5AGQ, by = c("id", "bin")) %>%
+  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_hat, group = id))+
+  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+             # alpha = 0.5, linewidth = 0.5)+
+  labs(x = "Time", y = "Estimated latent function (probablity scale)", title = "nAGQ=5"), nrow = 1)
 ```
@@
 # overview at estimated latent function
+ggarrange(
+#nAGQ=0
 df %>% 
@@
   #            linewidth = 0.5, alpha = 0.6)+
-  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
+  labs(x = "Bin", y = "Estimated latent function/number of active minutes", title = "nAGQ=0")+
+  ylim(-25, 10),
+
+#nAGQ=5
+df %>% 
+  filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  left_join(df_est_latent_5AGQ, by = c("id", "bin")) %>%
+  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
+  #            linewidth = 0.5, alpha = 0.6)+
+  labs(x = "Bin", y = "Estimated latent function/number of active minutes", title = "nAGQ=5")+
+  ylim(-25, 10),
+
+nrow = 1
+)
 ```
 
-Obviously the lower bound of estimated latent function got a lot higher. Nothing else seems to change much.
-
-For FPCA models below, I'll use the result from nAGQ=0 since it is free of fitting issues.
+Obviously the lower bound of estimated latent function got a lot lower using AGQ than PIRLS. Nothing else seems to change much.
 
@@
 
+I would also like to compare the difference between AGQ (nAGQ=5) and PIRLS (nAGQ=0)
+
+- AGQ has 16 eigenfunctions while PIRLS has 18
+
 ```{r}
-mat_est_unique <- matrix(df_est_latent_0AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
+# nAGQ=0
+mat_est_unique_0AGQ <- matrix(df_est_latent_0AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
 # row index subject, column binned time
-fpca_mod <- fpca.face(mat_est_unique, pve = 0.95, argvals = mid, var=T)
+fpca_mod_0AGQ <- fpca.face(mat_est_unique_0AGQ, pve = 0.95, argvals = mid, var=T)
 
-K <- ncol(fpca_mod$efunctions) # 18 eigenfunctions
+K_0AGQ <- ncol(fpca_mod_0AGQ$efunctions) # 18 eigenfunctions
 ```
@@
 ```{r}
-# population mean
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+# nAGQ=5
+mat_est_unique_5AGQ <- matrix(df_est_latent_5AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
+# row index subject, column binned time
+fpca_mod_5AGQ <- fpca.face(mat_est_unique_5AGQ, pve = 0.95, argvals = mid, var=T)
+
+K_5AGQ <- ncol(fpca_mod_5AGQ$efunctions) # 18 eigenfunctions
 ```
 
+- Population mean function
+
 ```{r}
+# population mean
+par(mfrow=c(1, 2))
+plot(mid, fpca_mod_0AGQ$mu, type = "l", xlab = "bin", ylab = "mean", main = "nAGQ=0", ylim = c(-25, 0))
+plot(mid, fpca_mod_5AGQ$mu, type = "l", xlab = "bin", ylab = "mean", main = "nAGQ=5", ylim = c(-25, 0))
+```
+- PC functions
+```{r, fig.height=10, fig.width=6}
 # eigenfunctions
-par(mfrow=c(2,2))
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+par(mfrow=c(4,2))
+
+plot(mid, fpca_mod_0AGQ$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "nAGQ=0")
+plot(mid, fpca_mod_5AGQ$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "nAGQ=5")
+
+plot(mid, fpca_mod_0AGQ$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid, fpca_mod_5AGQ$efunctions[, 2], type="l", xlab="bin", ylab="PC3")
+
+plot(mid, fpca_mod_0AGQ$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid, fpca_mod_5AGQ$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+
+plot(mid, fpca_mod_0AGQ$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+plot(mid, fpca_mod_5AGQ$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
 ```
 
+- Individual covariance matrix
 ```{r}
 # covariance
-heatmap(fpca_mod$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
-        Rowv = NA, Colv = NA, main = "Covariance matrix")
+heatmap(fpca_mod_0AGQ$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
+        Rowv = NA, Colv = NA, main = "nAGQ=0")
+heatmap(fpca_mod_5AGQ$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
+        Rowv = NA, Colv = NA, main = "nAGQ=5")
 ```
@@
 # maximum observations time: 360, 720, 1080
+# nAGQ=0
 df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
 df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
-score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
+score_out_mat <- array(NA, dim = c(length(rand_id), K_0AGQ, 3))
 # dim indexes subject, eigenfunction and max obs time respectively
+
+# nAGQ=5
+df_pred2 <- df_est_latent_5AGQ %>% filter(id %in% rand_id)
+df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
+score_out_mat2 <- array(NA, dim = c(length(rand_id), K_5AGQ, 3))
 ```
@@
 # prediction for a single subject
+# nAGQ=0
 for(i in rand_id){
@@
   pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
-                                 fpca_fit = fpca_mod, K = K)
+                                 fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
   pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
-                                 fpca_fit = fpca_mod, K = K)
+                                 fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
   pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
-                                  fpca_fit = fpca_mod, K = K)
+                                  fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
   # prediction in container
@@
   df_pred[df_pred$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
+  # score in container
+  # score_out_mat[i, ,1] <- pred_t360$score_out
+  # score_out_mat[i, ,2] <- pred_t720$score_out
+  # score_out_mat[i, ,3] <- pred_t1440$score_out
+}
+
+# df_pred
+
+```
+
+```{r, warning=FALSE}
+# prediction for a single subject
+# nAGQ=5
+for(i in rand_id){
+  df_i <- df %>% filter(id==i) %>% mutate(bin=as.numeric(bin))
+  # prediction
+  pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
+                                 fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
+  pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
+                                 fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
+  pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
+                                  fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
+  # prediction in container
+  df_pred2[df_pred2$id==i, 'pred_t360'] <- pred_t360$eta_pred
+  df_pred2[df_pred2$id==i, 'pred_t720'] <- pred_t720$eta_pred
+  df_pred2[df_pred2$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
   # score in container
@@
 # df_pred %>% filter(id == rand_id[1])
+
+df_pred2$pred_t360[df_pred2$bin<=360] <- NA
+df_pred2$pred_t720[df_pred2$bin<=720] <- NA
+df_pred2$pred_t1080[df_pred2$bin<=1080] <- NA
 ```
@@
 ```{r, warning=FALSE}
+ggarrange(
 df_pred %>%
@@
   geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
-  facet_wrap(~id,scales = "free")
+  facet_wrap(~id,scales = "free")+
+  labs(title = "nAGQ=0"),
+
+df_pred2 %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
+  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
+  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
+  facet_wrap(~id,scales = "free")+
+  labs(title = "nAGQ=5"),
+
+nrow = 1
+)
 ```
@@
 
-```{r, warning=FALSE, message=FALSE, results='hide'}
-df_pred2 <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
+```{r, warning=FALSE, message=FALSE}
+#PIRLS
+df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
+df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
+score_out_mat <- array(NA, dim = c(length(rand_id), K_0AGQ, 3))
+
+# AGQ
+df_pred2 <- df_est_latent_5AGQ %>% filter(id %in% rand_id)
 df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
-score_out_mat2 <- array(NA, dim = c(length(rand_id), K, 3))
+score_out_mat <- array(NA, dim = c(length(rand_id), K_5AGQ, 3))
 
+```
+
+```{r, results='hide'}
 # unique id
@@
 
+# PRILS
 for(i in rand_id){
@@
   ## up to 360
-  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=360))
-  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t1$score_out
+  pred_t1 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=360))
+  eta_pred_t1 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t1$score_out
+  df_pred[df_pred$id == i, 'pred_t360'] <- eta_pred_t1
+  ## up to 720
+  pred_t2 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=720))
+  eta_pred_t2 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t2$score_out
+  df_pred[df_pred$id == i, 'pred_t720'] <- eta_pred_t2
+  ## up to 1080
+  pred_t3 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=1080))
+  eta_pred_t3 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t3$score_out
+  df_pred[df_pred$id == i, 'pred_t1080'] <- eta_pred_t3
+
+}
+
+df_pred$pred_t360[df_pred$bin<=360] <- NA
+df_pred$pred_t720[df_pred$bin<=720] <- NA
+df_pred$pred_t1080[df_pred$bin<=1080] <- NA
+```
+
+```{r, results='hide'}
+# AGQ
+for(i in rand_id){
+  df_i <- df %>% filter(id==i)
+
+  # prediction 
+  ## up to 360
+  pred_t1 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=360))
+  eta_pred_t1 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t1$score_out
   df_pred2[df_pred2$id == i, 'pred_t360'] <- eta_pred_t1
   ## up to 720
-  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=720))
-  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t2$score_out
+  pred_t2 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=720))
+  eta_pred_t2 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t2$score_out
   df_pred2[df_pred2$id == i, 'pred_t720'] <- eta_pred_t2
   ## up to 1080
-  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1080))
-  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t3$score_out
+  pred_t3 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=1080))
+  eta_pred_t3 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t3$score_out
   df_pred2[df_pred2$id == i, 'pred_t1080'] <- eta_pred_t3
@@
 }
-```
 
-```{r}
 df_pred2$pred_t360[df_pred2$bin<=360] <- NA
@@
 ```{r, warning=FALSE}
+ggarrange(
+df_pred %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
+  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
+  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
+  facet_wrap(~id,scales = "free")+
+  labs(title = "nAGQ=0"),
+
 df_pred2 %>%
@@
   geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
-  facet_wrap(~id,scales = "free")
+  facet_wrap(~id,scales = "free")+
+  labs(title = "nAGQ=5"), nrow = 1
+)
 ```

### 2023-09-18 Docs/ProgressReport.Rmd
explore glmmTMB

--- 2023-09-11 Docs/ProgressReport.Rmd
@@
 author: "Ying Jin"
-date: "2023-05-16"
+date: "2023-09-15"
 output: 
@@
     toc_float: true
+    font: 12pt
 ---
@@
 
-- This is where near-unidentifiability issues first appear
-- Though point estimation not necessarily affected
+Following up on last report, when using glmer, the results from PRILS (nAGQ=0) seems more in line with experiences.
+As validation, let's fit the marginal model and look at the estimated population-average latent function, and compare it with the FPCA mean function.
+Specifically, at bin s the model would be $g(E(Y_i(s)))=f_0(s)$, with no subject-specific random effect.
 
-```{r, cache=TRUE, warning=FALSE}
+Another validation I came up with is to look at the local intercepts.
+That is the $\beta_0(s)$ in local mixed models $g(E(Y_i(s)))=\beta_0(s)+b_i(x)$.
+This should correspond to the FPCA mean functions, as the population average level.
+
+Since the individual random effects are centered around zero, I believe the marginal $f_0(s)$ should look very much like the conditional $\beta_0(x)$ if the model local fit well.
+The result obviously is in favor of PRILS.n *But what is the reason for AGQ to overshoot downward?*
+
+```{r}
 # split by each bin
@@
 df_bin_lst <- split(df, f = df$bin)
+```
 
+```{r cache=TRUE}
+# local marginal GLM
+local_marg <- function(df){
+    this_glm <- glm(Y ~ 1, data = df, family = binomial)
+    eta_hat <- predict(this_glm, type = "link")
+    df$eta_hat <- eta_hat
+    return(df)
+}
+
+df_est_local_marg <- lapply(df_bin_lst, function(x){local_marg(x)})
+
+# local conditional GLMM
+## PIRLS vs AGQ
+df_est_cond_mean_pirls <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 0)}) 
+df_est_cond_mean_agq <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 5)}) 
+```
+
+```{r, fig.width=12, fig.height=4}
+ggarrange(
+df_est_local_marg %>% bind_rows() %>%
+  select(bin, eta_hat) %>% distinct(.) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLM (marginal)", y = "latent mean")+
+  ylim(-25, 1),
+
+data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLMM (conditional) using PRILS", y = "latent mean")+
+  ylim(-25, 1),
+
+data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_agq)) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLMM (conditional) using AGQ", y = "latent mean")+
+  ylim(-25, 1),
+
+nrow = 1, align = "h"
+)
+```
+
+Let's for now trust the results from PIRLS more than AGQ, if we are using glmer.
+
+```{r cache=TRUE}
 # fit local GLMM and estimate latent function
-# near-unidentifiability issues 
-df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+# to avoid near-unidentifiability issues, using PIRLS maximization with nAGQ = 0
+
+# fit local models
+df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
 df_est_latent <- bind_rows(df_est_latent)
@@
   facet_wrap(~id, scales = "free")+
-  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
-             alpha = 0.5, linewidth = 0.5)+
   labs(x = "Time", y = "Estimated latent function (probablity scale)")
@@
   ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, group = id, linetype = "dashed"))+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
   geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
   facet_wrap(~id, scales = "free")+
-  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
-             linewidth = 0.5, alpha = 0.6)+
   labs(x = "Bin", y = "Estimated latent function/number of active minutes")
@@
 
-- Take a closer look at the unidentifiable bins?
+#### An alternative: glmmTMB
 
-I just realized glmer actually used Laplace approximation as well. Could this also be due to numeric stability? In some bins (465, 545, 555, 1185, 1195, 1205), the approximation failed to converge.
+The default optimizer of glmmTMB is nlminb (quasi-Newton BFGS), whereas glmer uses bobyaqa (nAGQ=0) or Nelder_Mead (nAGQ>0)?
 
-Below is the warning messages and one model summary.
+Let's first try to re-do the who local GLMM step using glmmTMB. Looks like no numeric issue occurred. It takes slightly longer time than PIRLS in glmer
 
-For this bin, the problem is with the fixed effect standard error. It is clearly near-zero.
-
-https://www.learn-mlms.com/07-module-7.html
-
-https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
-
-```{r, eval=FALSE}
+```{r}
+library(glmmTMB)
+# re-fit local GLMM and check numeric issues
 try_fit_lst<-list()
+t1=Sys.time()
 for(i in seq_along(df_bin_lst)){
-  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]],
-                             family = binomial)},
-                       warning=function(w){NA})
+  # this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
+  #                      warning=function(w){NA})
+  this_glm <- glmmTMB(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial)
   try_fit_lst[[i]] <- this_glm
@@
 }
-
-# bins where glmer returned errors
-# names(df_bin_lst)[is.na(try_fit_lst)]
+df_est_latent_tmb <- lapply(try_fit_lst, function(x){predict(x, type = "link")})
+t2=Sys.time()
+t_tmb <- t2-t1
+# population mean
 ```
 
-```{r}
-issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']],
-                             family = binomial)
-summary(issue_fit)
+- Time: `r t_tmb` minutes
+- Mean/Intercept function: It looks like glmmTMB estimates lie right between PIRLS and AGQ! It takes slightly shorter time than PIRLS and mean function shoots more downward, not as bad as AGQ though.
 
-# table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)
+```{r, fig.width=12, fig.height=4}
+mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
+
+ggarrange(
+df_est_local_marg %>% bind_rows() %>%
+  select(bin, eta_hat) %>% distinct(.) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLM (marginal)", y = "latent mean")+
+  ylim(-15, 1),
+
+data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLMM (conditional) using PRILS (glmer)", y = "latent mean")+
+  ylim(-15, 1),
+
+data.frame(bin=mid, eta_hat = mean_tmb) %>%
+  ggplot(aes(x=bin, y=eta_hat))+
+  geom_line()+
+  labs(x="local GLMM (conditional) using glmmTMB", y = "latent mean")+
+  ylim(-15, 1),
+
+nrow = 1
+)
 ```
 
-```{r eval=FALSE}
-colSums(table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)==0)
-colSums(table(df_bin_lst[['465']]$id, df_bin_lst[['465']]$Y)==0)
+- Individual latent functions
+
+```{r}
+for(i in seq_along(df_est_latent_tmb)){
+  df_est_latent_tmb[[i]] <- data.frame(df_bin_lst[[i]], eta_hat = df_est_latent_tmb[[i]])
+}
+
+df_est_latent_tmb <- bind_rows(df_est_latent_tmb)
+
+# keep midpoint values
+df_est_latent_tmb <- df_est_latent_tmb %>%
+  select(-sind, -Y) %>% distinct(.)
+
+```
+
+```{r, message=FALSE, fig.height=8, fig.width=16}
+df %>% 
+  filter(id %in% rand_id) %>%
+  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
+  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
+            by = c("id", "bin")) %>%
+  mutate_at(vars(eta_hat, eta_hat_tmb), function(x){exp(x)/(1+exp(x))}) %>% 
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_hat, col = "glmer", alpha = 0.5))+
+  geom_line(aes(x=sind, y=eta_hat_tmb, col = "glmmTMB", alpha = 0.5))+
+  geom_point(aes(x=sind, y = Y), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  labs(x = "Time", y = "Estimated latent function (probablity scale)")
+```
+
+```{r, message=FALSE, fig.height=8, fig.width=16}
+# overview at estimated latent function
+df %>% 
+  filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
+  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
+            by = c("id", "bin")) %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, col = "glmer", alpha = 0.5))+
+  geom_line(aes(x=bin, y=eta_hat_tmb, col = "glmmTMB", alpha = 0.5))+
+  geom_point(aes(x=bin, y = num), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  labs(x = "Time", y = "Estimated latent function")
+```
+
+So it looks like glmmTMB still overshoot downwards.
+Between the two methods, the mid-day part (active part for most people) seems to be very similar.
+Difference lies mostly at both ends, with consecutive inactive minutes.
+
+```{r, eval=FALSE}
+# numeric values of individual latent function
+df %>% 
+  filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
+  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
+            by = c("id", "bin")) %>%
+  head(20)
 ```
@@
 
+Follow up from late time, let's use PIRLS estimator and fix the number of PC functions instead of using all of them. We start with four PC functions.
+
+I will use estimated latent functions from both glmer (nAGQ=0) and glmmTMB, and compare their results.
+
 ```{r}
+# glmer nAGQ=0
 mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
-# row index subject, column binned time
-fpca_mod <- fpca.face(mat_est_unique, pve = 0.95, argvals = mid, var=T)
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
 
-K <- ncol(fpca_mod$efunctions) # 18 eigenfunctions
+# lmmTMB
+mat_est_unique2 <- matrix(df_est_latent_tmb$eta_hat, 
+                          nrow=N, ncol=n_bin, byrow = F) 
+fpca_mod2 <- fpca.face(mat_est_unique2, argvals = mid, var=T)
+
+# ncol(fpca_mod$efunctions) # 27 eigenfunctions
+K <- 4
 ```
+
+- Population mean function
+
+Hmm...FPCA seems to shoot the population mean downward in both methods, more so for glmmTMB.
+It looks like more rounds of model would make the population mean function more variable.
+Does that make sense?
 
@@
 # population mean
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+par(mfrow=c(1, 2))
+plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmer",
+     ylim = c(-12, 1))
+plot(mid, fpca_mod2$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmmTMB", ylim = c(-12, 1))
+```
+- PC functions
+
+```{r, fig.height=16, fig.width=8}
+# eigenfunctions
+par(mfrow=c(4,2))
+
+plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmer", ylim = c(-0.21, 0.13))
+plot(mid, fpca_mod2$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmmTMB", ylim = c(-0.21, 0.13))
+
+plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2", ylim = c(-0.2, 0.14))
+plot(mid, fpca_mod2$efunctions[, 2], type="l", xlab="bin", ylab="PC2",ylim = c(-0.2, 0.14))
+
+plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3", ylim = c(-0.2, 0.2))
+plot(mid, fpca_mod2$efunctions[, 3], type="l", xlab="bin", ylab="PC3",ylim = c(-0.2, 0.2))
+
+plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4", ylim = c(-0.14, 0.23))
+plot(mid, fpca_mod2$efunctions[, 4], type="l", xlab="bin", ylab="PC4",
+     ylim = c(-0.14, 0.23))
 ```
 
-```{r}
-# eigenfunctions
-par(mfrow=c(2,2))
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
-```
+- Individual covariance matrix
+
+Wow! This covariacne matrix changed a lot from when using all of the PCs! It still looks cyclic, but now off-diagnoal region seems much more stable. Fluctuations mostly centered around diagnoal line.
 
@@
 heatmap(fpca_mod$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
-        Rowv = NA, Colv = NA, main = "Covariance matrix")
+        Rowv = NA, Colv = NA, main = "glmer")
+
+heatmap(fpca_mod2$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
+        Rowv = NA, Colv = NA, main = "glmmTMB")
+
 ```
@@
 
-### Maximum likelihood
+Still from last time, we're gonna focus on Bayes methods now!
 
-- Boundary hitting issues with numeric maximization
-- Warning message: Steady solution not reached
-- Large variation in both predicted values and scores. The numeric solution from maximum likelihood is obviously not stable. It is very likely to fail to converge.
-- Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot.
+We are also going to extend the observed track, so that there are some variation in the observed data (instead of all zeros). I am going to start with 9am (540), 1pm (780),  5pm (1020)
 
-```{r}
-# Out-of-sample prediction
-# maximum observations time: 360, 720, 1080
+Extending the track seems to solve the problem with failed Laplace Approximation. Not sure. Should try on the full sample.
+
+```{r, warning=FALSE, message=FALSE}
 df_pred <- df_est_latent %>% filter(id %in% rand_id)
-df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
+df_pred[, 'pred_t540'] <- df_pred[, 'pred_t780'] <- df_pred[, 'pred_t1020'] <- NA
 score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
-# dim indexes subject, eigenfunction and max obs time respectively
 ```
 
-```{r, warning=FALSE}
-# prediction for a single subject
-for(i in rand_id){
-  df_i <- df %>% filter(id==i) %>% mutate(bin=as.numeric(bin))
-  # prediction
-  pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
-                                 fpca_fit = fpca_mod, K = K)
-  pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
-                                 fpca_fit = fpca_mod, K = K)
-  pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
-                                  fpca_fit = fpca_mod, K = K)
-  # prediction in container
-  df_pred[df_pred$id==i, 'pred_t360'] <- pred_t360$eta_pred
-  df_pred[df_pred$id==i, 'pred_t720'] <- pred_t720$eta_pred
-  df_pred[df_pred$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
-  # score in container
-  # score_out_mat[i, ,1] <- pred_t360$score_out
-  # score_out_mat[i, ,2] <- pred_t720$score_out
-  # score_out_mat[i, ,3] <- pred_t1440$score_out
-}
-
-```
-
-```{r}
-df_pred$pred_t360[df_pred$bin<=360] <- NA
-df_pred$pred_t720[df_pred$bin<=720] <- NA
-df_pred$pred_t1080[df_pred$bin<=1080] <- NA
-# df_pred %>% filter(id == rand_id[1])
-```
-
-```{r, warning=FALSE}
-df_pred %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, group = id))+
-  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
-  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
-  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
-  facet_wrap(~id)
-```
-
-### Bayes method like Laplace approximation
-
-```{r, warning=FALSE, message=FALSE, results='hide'}
-df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
-df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
-score_out_mat2 <- array(NA, dim = c(length(rand_id), K, 3))
-
+```{r, results='hide'}
 # unique id
@@
 source(here("Code/OutsampBayes.R"))
+
+# PRILS
+for(i in rand_id){
+  df_i <- df %>% filter(id==i)
+
+  # prediction 
+  ## up to 360
+  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=540), kpc=K)
+  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t1$score_out
+  df_pred[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
+  ## up to 720
+  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=780), kpc=K)
+  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t2$score_out
+  df_pred[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
+  ## up to 1080
+  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1020), kpc = K)
+  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t3$score_out
+  df_pred[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3
+
+}
+
+df_pred$pred_t540[df_pred$bin<=540] <- NA
+df_pred$pred_t780[df_pred$bin<=780] <- NA
+df_pred$pred_t1020[df_pred$bin<=1020] <- NA
+```
+
+```{r, results='hide'}
+# glmmTMB
+df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
+df_pred2[, 'pred_t540'] <- df_pred2[, 'pred_t780'] <- df_pred2[, 'pred_t1020'] <- NA
 
@@
   ## up to 360
-  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=360))
-  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t1$score_out
-  df_pred2[df_pred2$id == i, 'pred_t360'] <- eta_pred_t1
+  pred_t1 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=540), kpc=K)
+  eta_pred_t1 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t1$score_out
+  df_pred2[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
   ## up to 720
-  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=720))
-  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t2$score_out
-  df_pred2[df_pred2$id == i, 'pred_t720'] <- eta_pred_t2
+  pred_t2 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=780), kpc=K)
+  eta_pred_t2 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t2$score_out
+  df_pred2[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
   ## up to 1080
-  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1080))
-  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions%*%pred_t3$score_out
-  df_pred2[df_pred2$id == i, 'pred_t1080'] <- eta_pred_t3
+  pred_t3 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=1020), kpc = K)
+  eta_pred_t3 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t3$score_out
+  df_pred2[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3
 
 }
+
+df_pred2$pred_t540[df_pred$bin<=540] <- NA
+df_pred2$pred_t780[df_pred$bin<=780] <- NA
+df_pred2$pred_t1020[df_pred$bin<=1020] <- NA
 ```
 
-```{r}
-df_pred2$pred_t360[df_pred2$bin<=360] <- NA
-df_pred2$pred_t720[df_pred2$bin<=720] <- NA
-df_pred2$pred_t1080[df_pred2$bin<=1080] <- NA
-# df_pred %>% filter(id == rand_id[1])
-```
+```{r, warning=FALSE, fig.width=16, fig.height=8}
+ggarrange(
+df_pred %>%
+  ggplot()+
+  geom_line(aes(x=bin, y=eta_hat, group = id))+
+  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
+  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
+  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
+  facet_wrap(~id,scales = "free")+labs(title = "glmer"),
 
-```{r, warning=FALSE}
 df_pred2 %>%
@@
   geom_line(aes(x=bin, y=eta_hat, group = id))+
-  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
-  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
-  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
-  facet_wrap(~id)
+  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
+  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
+  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
+  facet_wrap(~id,scales = "free")+labs(title = "glmmTMB"),
+nrow = 1, common.legend = T)
 ```
-
-<!-- ## Failed prediction  -->
-
-<!-- - Laplace approximation failed for some subjects -->
-<!-- - Only when the observed track is short (up to 360) -->
-<!-- - These subjects have something in common -->
-
-<!-- ```{r} -->
-<!-- # compare subjects that failed at Laplace Approximation with others -->
-<!-- rand_id <- sample(unique(df$SEQN), size = 4) -->
-<!-- com_id <- c(rand_id, sample(skip_id, 4)) -->
-
-<!-- df_est_latent %>% filter(id %in% com_id) %>% -->
-<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
-<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
-<!--   mutate(type = ifelse(id %in% rand_id, "Prediction", "Failed prediction")) %>% -->
-<!--   ggplot(aes(x=bin, y=eta_hat, col = type))+ -->
-<!--   geom_line()+ -->
-<!--   geom_vline(xintercept = 360)+ -->
-<!--   facet_wrap(~type+id, nrow = 2, ncol = 4)+ -->
-<!--   labs(y = "Estimated latend function", x = "Time", col = "") -->
-<!-- ``` -->
-
-<!-- ## Subjects with complete prediction -->
-
-<!-- ```{r fig_prob, warning=FALSE} -->
-<!-- df_est_latent %>% -->
-<!--   filter(id %in% rand_id) %>% -->
-<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
-<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
-<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
-<!--   mutate_at(vars(pred_t360, pred_t720, pred_t1080), function(x)exp(x)/(1+exp(x))) %>% -->
-<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
-<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
-<!--   ggplot()+ -->
-<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
-<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
-<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
-<!--     geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.5)+ -->
-<!--     facet_wrap(~id)+ -->
-<!--     labs(x = "Time", y = "Estimated latent function (probablity scale)", -->
-<!--          col = "Maximum observation time") -->
-<!-- ``` -->
-
-<!-- ```{r fig_eta, warning=FALSE} -->
-<!-- df_est_latent %>% -->
-<!--   filter(id %in% rand_id) %>% -->
-<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
-<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
-<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
-<!--   ggplot()+ -->
-<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
-<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
-<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
-<!--     geom_line(aes(x=bin, y = eta_hat, col = "eta_hat"), linetype = "solid" )+ -->
-<!--     facet_wrap(~id)+ -->
-<!--     labs(x = "Time", y = "Estimated latent function") -->
-<!-- ``` -->
-
-<!-- ## AUC on subjects with complete prediction -->
-
-<!-- - Use interpolation to fill in gaps between bins -->
-
-<!-- ```{r} -->
-<!-- # subjects with complete prediction -->
-<!-- N <- length(unique(df$SEQN))-length(skip_id) -->
-<!-- eval_id <- setdiff(unique(df$SEQN), skip_id) -->
-<!-- J <- max(df$sind) -->
-
-<!-- # containter: row index subject, column index time -->
-<!-- eta_mat_t360 <- eta_mat_t720 <- eta_mat_t1080 <- matrix(NA, nrow = N, ncol = J) -->
-
-<!-- # interpolation -->
-<!-- for(i in seq_along(eval_id)){ -->
-<!--    eta_mat_t360[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
-<!--                                y=df_est_latent$pred_t360[df_est_latent$id==eval_id[i]],  -->
-<!--                                xout = 1:J)$y -->
-<!--    eta_mat_t720[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
-<!--                                y=df_est_latent$pred_t720[df_est_latent$id==eval_id[i]],  -->
-<!--                                xout = 1:J)$y -->
-<!--    eta_mat_t1080[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
-<!--                                y=df_est_latent$pred_t1080[df_est_latent$id==eval_id[i]],  -->
-<!--                                xout = 1:J)$y -->
-<!--   } -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- # true outcome -->
-<!-- Y_mat <- df$Z[df$SEQN %in% eval_id] -->
-<!-- Y_mat <- matrix(Y_mat, nrow = N, ncol = J, byrow = T)  -->
-<!-- ``` -->
-
-<!-- ```{r AUC_fGFPCA} -->
-<!-- # maximum observations time -->
-<!-- t_vec <- c(360, 720, 1080) -->
-
-<!-- # AUC container -->
-<!-- auc_mat <- matrix(NA, 3, 3) -->
-<!-- colnames(auc_mat) <- c("360", "720", "1080") -->
-<!-- rownames(auc_mat) <- c("(360, 720]", "(720, 1080]", "(1080, 1440]") -->
-
-<!-- # Calculate AUC -->
-
-<!-- auc_mat[1, 1] <- performance(prediction(as.vector(eta_mat_t360[, 361:720]),  -->
-<!--                                    as.vector(Y_mat[, 361:720])), -->
-<!--                         "auc")@y.values[[1]] -->
-<!-- auc_mat[2, 1] <- performance(prediction(as.vector(eta_mat_t360[, 721:1080]),  -->
-<!--                                    as.vector(Y_mat[, 721:1080])), -->
-<!--                         "auc")@y.values[[1]] -->
-<!-- auc_mat[3, 1] <- performance(prediction(as.vector(eta_mat_t360[, 1081:1435]),  -->
-<!--                                    as.vector(Y_mat[, 1081:1435])), -->
-<!--                         "auc")@y.values[[1]] -->
-
-<!-- auc_mat[2, 2] <- performance(prediction(as.vector(eta_mat_t720[, 721:1080]),  -->
-<!--                                    as.vector(Y_mat[, 721:1080])), -->
-<!--                         "auc")@y.values[[1]] -->
-<!-- auc_mat[3, 2] <- performance(prediction(as.vector(eta_mat_t720[, 1081:1435]),  -->
-<!--                                    as.vector(Y_mat[, 1081:1435])), -->
-<!--                         "auc")@y.values[[1]] -->
-
-<!-- auc_mat[3, 3] <- performance(prediction(as.vector(eta_mat_t1080[, 1081:1435]),  -->
-<!--                                    as.vector(Y_mat[, 1081:1435])), -->
-<!--                         "auc")@y.values[[1]] -->
-<!-- auc_mat -->
-<!-- ``` -->
-
-<!-- ## Reference method -->
-
-<!-- - GLMMadaptive? -->
-<!-- - fcr on lowess smoothed track? -->

### 2023-09-18 Code/GLMM-FPCA.R

@@ -3,8 +3,8 @@
 # get predicted latent function
 # df must have: outcome Y, subject id
 
-pred_latent <- function(df){
-    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+pred_latent <- function(df, n_node){
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ=n_node)
     eta_hat <- predict(this_glm, type = "link")
     df$eta_hat <- eta_hat
     return(df)
@@ -12,8 +12,8 @@ pred_latent <- function(df){
   
 # Additionally, get the average estimation across subject 
 # by extacting the random intercepts from each bin
-mean_latent <- function(df){
-    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial)
+mean_latent <- function(df, n_node){
+    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ=n_node)
     beta_hat <- coef(summary(this_glm))[1]
     return(beta_hat)
 }

### 2023-09-18 Code/OutSampBayes.R

@@ -65,7 +65,7 @@ Model <- function(parm, Data){
 
 
 # data: 
-out_pred_laplace <- function(fpca_fit, df_new){
+out_pred_laplace <- function(fpca_fit, df_new, kpc){
   
   # put data into correct format
   ns <- as.vector(table(df_new$bin)) # number of observations
@@ -75,13 +75,13 @@ out_pred_laplace <- function(fpca_fit, df_new){
   df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)  
   
   # into a list
-  tao <- diag(fpca_fit$evalues)
+  tao <- diag(fpca_fit$evalues[1:kpc])
   f0 <- fpca_fit$mu[1:max_bin]
   N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
   n <- df_new2$ns # number of experiements at each bin
   y <- df_new2$hs # outcome, number of 1
-  X <- fpca_fit$efunctions[1:max_bin, ] # "covariates", in this case PC functions
-  J <- ncol(fpca_fit$efunctions) # number of parameters/scores
+  X <- fpca_fit$efunctions[1:max_bin, 1:kpc] # "covariates", in this case PC functions
+  J <- kpc # number of parameters/scores
   mon.names <- "LP"
   
   # parameter names
@@ -103,7 +103,7 @@ out_pred_laplace <- function(fpca_fit, df_new){
   score <- Fit$Summary1[, "Mode"]
   
   # prediction
-  eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions%*%score
+  eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions[, 1:kpc]%*%score
   
   return(list(eta_pred = eta_pred_out,
                score_out = score))

### 2023-09-18 Code/OutSampMLE.R

@@ -8,31 +8,31 @@ library(rootSolve)
 # just for mess around
 
 #### Likelihood function ####
-joint_pdf <- function(xi, df_new, fpca_fit){
-  
-  # observations 
-  ns <- as.vector(table(df_new$bin)) # number of observations
-  hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
-    select(Y) %>% unlist()# number of success
-  nf <- ns-hs # number of failure
-  max_bin <- length(unique(df_new$bin)) # assume no skipped bins 
-  
-  # numbers
-  ns <- table(df_out$bin) # number of observations
-  hs <- df_out %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
-    select(Y) # number of success
-  nf <- ns-hs # number of failure
-  
-  # probability of conditional distribution of Y
-  eta = f0+phi%*%xi
-  pr = exp(eta)/(1+exp(eta))
-  # the part relevant to P(Y|xi)
-  p1 <- prod(dbinom(x=unlist(hs), size = ns, prob=pr))
-  p2 <- exp(-t(xi)%*%solve(tao)%*%xi/2)
-  
-  return(p1*p2)
-  
-}
+# joint_pdf <- function(xi, df_new, fpca_fit){
+#   
+#   # observations 
+#   ns <- as.vector(table(df_new$bin)) # number of observations
+#   hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+#     select(Y) %>% unlist()# number of success
+#   nf <- ns-hs # number of failure
+#   max_bin <- length(unique(df_new$bin)) # assume no skipped bins 
+#   
+#   # numbers
+#   ns <- table(df_out$bin) # number of observations
+#   hs <- df_out %>% group_by(bin) %>% summarize_at("Y", sum) %>% 
+#     select(Y) # number of success
+#   nf <- ns-hs # number of failure
+#   
+#   # probability of conditional distribution of Y
+#   eta = f0+phi%*%xi
+#   pr = exp(eta)/(1+exp(eta))
+#   # the part relevant to P(Y|xi)
+#   p1 <- prod(dbinom(x=unlist(hs), size = ns, prob=pr))
+#   p2 <- exp(-t(xi)%*%solve(tao)%*%xi/2)
+#   
+#   return(p1*p2)
+#   
+# }
 
 
 #### MLE of out-of-sample score ####

### 2023-09-19 Docs/ProgressReport.Rmd
data application

@@
 library(GLMMadaptive)
+library(kableExtra)
 theme_set(theme_minimal())
@@
 df <- read_rds(here("Data/nhanes_bi.rds"))
-load(here("Data/ApplOutput_fGFPCA.RData"))
+# load(here("Data/ApplOutput_fGFPCA.RData"))
 
@@
 
-# NHANES data application
+# NHANES data
 
@@
 
-```{r}
+```{r, cache=TRUE}
 library(glmmTMB)
@@
 # population mean
+mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
 ```
@@
 ```{r, fig.width=12, fig.height=4}
-mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
-
 ggarrange(
@@
 ```
+
+# NHANES data application output
+
+## fGFPCA
+
+- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
+- 3.3 minutes spent on model fitting, 20 minutes spent on prediction (about 0.01 per subject)
+- Three subjects had failed approximation, all happened with the shortest observed track
+
+```{r}
+load(here("Data/ApplOutput_fGFPCA.RData"))
+```
+
+- Example of out-of-sample prediction
+
+```{r message=FALSE, warning=FALSE}
+rand_id2 <- sample(unique(df_test$id), 4)
+df %>% 
+  filter(id %in% rand_id2) %>%
+  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020)) %>%
+  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
+  ggplot()+
+    geom_line(aes(x=bin, y = pred_t540, col = "9am"), linetype = "dashed")+
+    geom_line(aes(x=bin, y = pred_t780, col = "1pm"),linetype = "dashed")+
+    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"),linetype = "dashed")+
+    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "Estimated latent function (probablity scale)")
+```
+
+- Subjects with numeric issue
+
+```{r}
+df %>% filter(id %in% skip_id) %>% 
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  facet_wrap(~id)+
+  geom_vline(xintercept = c(540))
+```
+
+- Calculated AUC
+
+Since we are assuming value of latent function is constant within a bin, why can't we just do the same when calculating AUC without interpolation?
+
+Remove the three subjects with failed approximation
+
+```{r}
+lst_auc <- df %>% 
+  filter(id %in% unique(df_test$id) & !id %in% skip_id ) %>%
+  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin")) %>% 
+  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
+                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
+                      include.lowest = T))
+lst_auc <- split(lst_auc, f=lst_auc$window)
+
+# up to 9am
+auc_t540 <- lst_auc[2:4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
+auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 1pm
+auc_t780 <- lst_auc[3:4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t780, x$Y), measure = "auc"))
+auc_t780 <- lapply(auc_t780, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 5pm
+auc_t1020 <- lst_auc[4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t1020, x$Y), measure = "auc"))
+auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
+```
+
+## GLMMadaptvie
+
+- when fitting model with **random intercept + slope** on the same full training set: Error: vector memory exhausted (limit reached?)
+- when fitting model with **only random intercept** on the same full training set: Error regarding a large coefficient value.
+One fix is to re-scale covariates.
+So I re-scaled minute index by dividing them by J, so the range is within (0, 1], and tried again.
+This fitting procedure took 20.72 minutes to finish
+- Out-of-sample prediction took 12.18 minutes. About 0.007 minute per subject.
+
+```{r}
+load(here("Data/ApplOutput_GLMMadaptive.RData"))
+```
+
+```{r}
+# up to 540
+auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
+  mutate(sind=sind*J) %>%
+  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
+                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
+                      include.lowest = T))
+auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
+auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
+  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
+auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 780
+auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
+  mutate(sind=sind*J) %>%
+  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
+                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
+                      include.lowest = T))
+auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
+auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
+  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
+auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 1020
+auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
+  mutate(sind=sind*J) %>%
+  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
+                      labels = c("5pm-9pm","9pm-12pm"), 
+                      include.lowest = T))
+auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
+auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
+  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
+auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
+
+```
+
+## Compare fGFPCA and GLMMadaptive
+
+```{r}
+options(knitr.kable.NA = '')
+tb_auc <- data.frame(auc_t540, 
+           c(NA, auc_t780), 
+           c(NA, NA, auc_t1020), 
+           auc_t540_adglmm,
+           c(NA, auc_t780_adglmm), 
+           c(NA, NA, auc_t1020_adglmm))
+
+colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
+
+tb_auc %>%
+  kable(digit = 2) %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 3, "20.72" = 3), bold = F) %>%
+  add_header_above(c("Time spent on prediction" = 1, "20.0" = 3, "12.2" = 3), bold = F) %>% 
+  add_header_above(c(" "=1, "fGPFCA"=3, "GLMMadaptive"=3))
+```
+
+- fGFPCA almost always have higher AUC. It at least is as good as GLMMadaptive.
+- Computationally, fGFPCA is about 10 minute faster (33%).
+
+## Discussion
+
+- What would be the reference method to compare our performance to?
+Original GLMMadaptvie perhaps is feasible on teh entire dataset.
+But what should we use if we wanna do sub-sample comparison?
+Should we even do that for this data application (or simulation alone)?
+- There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.

### 2023-09-19 Code/DataApplNHANES.R

@@ -21,8 +21,6 @@ df <- read_rds(here("Data/nhanes_bi.rds"))
 N <- length(unique(df$SEQN)) # sample size 8763
 J <- max(df$sind) # 1440 measures for each subject
 
-
-
 #### visulization ####
 rand_id <- sample(unique(df$SEQN), size = 4)
 
@@ -33,11 +31,11 @@ df %>%
   facet_wrap(~SEQN)+
   labs(x="Time", y = "Activity")+
   theme(plot.margin = margin(0, 1, 0, 0))
-ggsave(here("Images/NHANES/data.png"), width=7, height=5, bg="white")
+# ggsave(here("Images/NHANES/data.png"), width=7, height=5, bg="white")
 
 
 
-##### fGFPCA #####
+##### Bin data #####
 
 # bin data
 bin_w <- 10 # bin width
@@ -45,177 +43,215 @@ n_bin <- J/bin_w # number of bins
 brks <- seq(0, J, by = bin_w) # cutoff points
 mid <- (brks+bin_w/2)[1:n_bin] # mid points
 
-
+# bin labels
 df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
 df$bin <- as.numeric(as.character(df$bin))
-head(df, 20)
 
 df <- df %>% rename(id = SEQN, Y=Z)
 
-df_bin_lst <- split(df, f = df$bin)
 
-lapply(df_bin_lst, function(x)table(x$Y))
+##### Data split #####
 
-# fit local GLMM and estimate latent function
+id_vec <- unique(df$id)
+train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
+test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
+
+#### fGFPCA estimation ####
+# fit model on the training set
+df_train <- df %>% filter(id %in% train_id)
+train_bin_lst <- split(df_train, f = df_train$bin)
+# length(train_bin_lst)
+# lapply(train_bin_lst, nrow)
+# train_bin_lst[[1]] %>% View()
+
+# local GLMM and estimate latent function
 # near-unidentifiability issues 
-df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
+t1=Sys.time()
+df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+t2= Sys.time()
[…]

### 2023-09-20 Docs/ProgressReport.Rmd
investigate effect of linear interpolation

@@
 - Three subjects had failed approximation, all happened with the shortest observed track
+- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints.
+Here I wanna compare if this choice would make any different.
+Also, there are a log of other methods to fill in the gaps, such as projection.
+- For linear interpolation, I set the edges to be the closest data extreme.
 
@@
 
-- Example of out-of-sample prediction
+```{r}
+# linear interpolation
+# remove subjects with numeric problem
+df_test_full <- df %>% 
+  filter(id %in% unique(df_test$id)) %>%
+  filter(!id %in% skip_id) %>%
+  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
+df_test_full$pred_t540_interp <- df_test_full$pred_t780_interp <- df_test_full$pred_t1020_interp <- NA
+
+# interpolation for each subject
+for(i in unique(df_test_full$id)){
+  # interpolation t540
+  interp_i_t540 <- approx(x = df_test[df_test$id==i, "bin"], 
+                     y = df_test[df_test$id==i, "pred_t540"],
+                     xout = 541:J, rule = 2)
+  df_test_full$pred_t540_interp[df_test_full$id==i & df_test_full$sind>540] <- interp_i_t540$y
+
+  # interpolation t780
+  interp_i_t780 <- approx(x = df_test[df_test$id==i, "bin"], 
+                     y = df_test[df_test$id==i, "pred_t780"],
+                     xout = 781:J, rule = 2)
+  df_test_full$pred_t780_interp[df_test_full$id==i & df_test_full$sind>780] <- interp_i_t780$y
+
+  # interpolation t1020
+  interp_i_t1020 <- approx(x = df_test[df_test$id==i, "bin"], 
+                     y = df_test[df_test$id==i, "pred_t1020"],
+                     xout = 1021:J, rule = 2)
+  df_test_full$pred_t1020_interp[df_test_full$id==i & df_test_full$sind>1020] <- interp_i_t1020$y
+}
+
+```
+
+Let's look at four subjects as an example:
 
@@
 rand_id2 <- sample(unique(df_test$id), 4)
-df %>% 
+# without interpolation, assume constant latent function value in each bin
+df_test_full%>% 
   filter(id %in% rand_id2) %>%
-  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020)) %>%
   mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
   ggplot()+
-    geom_line(aes(x=bin, y = pred_t540, col = "9am"), linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t780, col = "1pm"),linetype = "dashed")+
-    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"),linetype = "dashed")+
+    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
+    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
+    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
     geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
     facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)")
+    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
+         title = "Constant latent function within bin")
 ```
 
+```{r, warning=FALSE}
+# with interpolation
+df_test_full%>% 
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(pred_t540_interp, pred_t780_interp, pred_t1020_interp), function(x)exp(x)/(1+exp(x))) %>% 
+  ggplot()+
+    geom_line(aes(x=bin, y = pred_t540_interp, col = "9am"))+
+    geom_line(aes(x=bin, y = pred_t780_interp, col = "1pm"))+
+    geom_line(aes(x=bin, y = pred_t1020_interp, col = "5pm"))+
+    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
+         title = "Interpolated latent function")
+
+```
 - Subjects with numeric issue
 
-```{r}
+```{r, fig.height=3, fig.width=9}
 df %>% filter(id %in% skip_id) %>% 
@@
 
-Since we are assuming value of latent function is constant within a bin, why can't we just do the same when calculating AUC without interpolation?
-
-Remove the three subjects with failed approximation
-
 ```{r}
-lst_auc <- df %>% 
-  filter(id %in% unique(df_test$id) & !id %in% skip_id ) %>%
-  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin")) %>% 
+# break by prediction window
+lst_auc <- df_test_full %>% 
   mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
@@
 lst_auc <- split(lst_auc, f=lst_auc$window)
+```
 
+```{r}
+# no interpolation
 # up to 9am
@@
 auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
+```
+
+```{r}
+# with interpolation
+# up to 9am
+auc_t540_interp <- lst_auc[2:4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t540_interp, x$Y), measure = "auc"))
+auc_t540_interp <- lapply(auc_t540_interp, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 1pm
+auc_t780_interp <- lst_auc[3:4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t780_interp, x$Y), measure = "auc"))
+auc_t780_interp <- lapply(auc_t780_interp, function(x){x@y.values[[1]]}) %>% unlist()
+
+# up to 5pm
+auc_t1020_interp <- lst_auc[4] %>% 
+  lapply(function(x)performance(prediction(x$pred_t1020_interp, x$Y), measure = "auc"))
+auc_t1020_interp <- lapply(auc_t1020_interp, function(x){x@y.values[[1]]}) %>% unlist()
 ```
@@
            c(NA, NA, auc_t1020), 
+           auc_t540_interp, 
+           c(NA, auc_t780_interp), 
+           c(NA, NA, auc_t1020_interp), 
            auc_t540_adglmm,
@@
 
-colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
+colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 3)
 
 tb_auc %>%
-  kable(digit = 2) %>%
+  kable(digit = 4) %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 3, "20.72" = 3), bold = F) %>%
-  add_header_above(c("Time spent on prediction" = 1, "20.0" = 3, "12.2" = 3), bold = F) %>% 
-  add_header_above(c(" "=1, "fGPFCA"=3, "GLMMadaptive"=3))
+  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 6, "20.72" = 3), bold = F) %>%
+  add_header_above(c("Time spent on prediction" = 1, "20.0" = 6, "12.2" = 3), bold = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA"=3, "fGFPCA+interpolation" = 3, "GLMMadaptive"=3))
 ```
@@
 - There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
+- Is interplation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension?

### 2023-09-20 Docs/ProgressReport.Rmd
adjust test sample of GLMMadaptive

@@
 
-- when fitting model with **random intercept + slope** on the same full training set: Error: vector memory exhausted (limit reached?)
+- when fitting model with **random intercept + slope** on the same full training set (even after scaling covariates): Error: vector memory exhausted (limit reached?). 
 - when fitting model with **only random intercept** on the same full training set: Error regarding a large coefficient value.
@@
 - Out-of-sample prediction took 12.18 minutes. About 0.007 minute per subject.
+- I'd also exclude the subjects with failed Laplace approximation here. 
+- Should I calculate AUC subject-wise?
 
@@
 auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
+  filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%
@@
 auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
+  filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%
@@
 auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
+  filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%

### 2023-09-20 Code/DataApplNHANES.R

@@ -221,8 +221,10 @@ save(df_test, skip_id, file = here("Data/ApplOutput_fGFPCA.RData"))
 # fit GLMMadaptvie model on the training set
 head(df_train)
 t1 <- Sys.time()
-adglmm_mod <- mixed_model(Y ~ sind, random = ~ 1 | id, data = df_train %>% mutate(sind=sind/J),
-                      family = binomial())
+# adglmm_mod <- mixed_model(Y ~ sind, random = ~ 1 | id, data = df_train %>% mutate(sind=sind/J),
+#                       family = binomial())
+adglmm_mod <- mixed_model(Y ~ sind, random = ~ sind | id, data = df_train %>% mutate(sind=sind/J),
+                          family = binomial())
 t2 <- Sys.time()
 t_est_adglmm <- t2-t1 # model fitting took 20.72 mins
 summary(adglmm_mod)

### 2023-09-21 Docs/ProgressReport.Rmd
try to calculate subject-wise AUC
GLMMadaptive results

@@
 
-```{r, fig.height=3, fig.width=9}
+```{r, fig.height=6, fig.width=9}
 df %>% filter(id %in% skip_id) %>% 
@@
 - I'd also exclude the subjects with failed Laplace approximation here.
-- Should I calculate AUC subject-wise?
 
@@
 load(here("Data/ApplOutput_GLMMadaptive.RData"))
+```
+
+```{r}
+df_test_full2 <- df_test_full %>%
+  filter(id %in% rand_id2)%>%
+  select(id, sind, Y)
+
+df_test_full2$pred_t540 <- df_test_full2$pred_t780 <- df_test_full2$pred_t1020 <- NA 
+for(i in rand_id2){
+  df_test_full2[df_test_full2$id==i & df_test_full2$sind>540, "pred_t540"]<- adglmm_pred_t540$newdata2 %>% filter(id ==i) %>% select(pred)
+
+  df_test_full2[df_test_full2$id==i & df_test_full2$sind>780, "pred_t780"]<- adglmm_pred_t780$newdata2 %>% filter(id ==i) %>% select(pred)
+
+  df_test_full2[df_test_full2$id==i & df_test_full2$sind>1020, "pred_t1020"]<- adglmm_pred_t1020$newdata2 %>% filter(id ==i) %>% select(pred)
+}
+
+# colSums(is.na(df_test_full2))
+```
+
+```{r, warning=FALSE}
+df_test_full2 %>% 
+  mutate_at(vars(pred_t540, pred_t780, pred_t1020), 
+            function(x)exp(x)/(1+exp(x))) %>%
+  ggplot()+
+    geom_line(aes(x=sind, y = pred_t540, col = "9am"))+
+    geom_line(aes(x=sind, y = pred_t780, col = "1pm"))+
+    geom_line(aes(x=sind, y = pred_t1020, col = "5pm"))+
+    geom_point(aes(x=sind, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "Estimated latent function (probablity scale)")
 ```
@@
 - Is interplation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
+- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the predictio window, making AUC for this single subjects impossible to calculate...

### 2023-09-21 Code/DataApplNHANES.R
add GLMM adaptive model summary

@@ -230,30 +230,40 @@ t_est_adglmm <- t2-t1 # model fitting took 20.72 mins
 summary(adglmm_mod)
 
 # use the fitted model for prediction
-df_test <- df %>% filter(id %in% test_id)
-df_test$sind <- df_test$sind/J
-head(df_test)
+df_test2 <- df %>% filter(id %in% test_id)
+df_test2$sind <- df_test2$sind/J
+head(df_test2)
 
 ## up to 540
 t1 <- Sys.time()
 adglmm_pred_t540 <- predict(adglmm_mod, 
-        newdata = df_test %>% filter(sind <= 540/J),
-        newdata2 =  df_test %>% filter(sind > 540/J), 
+        newdata = df_test2 %>% filter(sind <= 540/J),
+        newdata2 =  df_test2 %>% filter(sind > 540/J), 
         type = "subject_specific", type_pred = "link",
         se.fit = TRUE, return_newdata = TRUE)
 
 adglmm_pred_t780 <- predict(adglmm_mod, 
-                            newdata = df_test %>% filter(sind <= 780/J),
-                            newdata2 =  df_test %>% filter(sind > 780/J), 
+                            newdata = df_test2 %>% filter(sind <= 780/J),
+                            newdata2 =  df_test2 %>% filter(sind > 780/J), 
                             type = "subject_specific", type_pred = "link",
                             se.fit = TRUE, return_newdata = TRUE)
 
 adglmm_pred_t1020 <- predict(adglmm_mod, 
-                            newdata = df_test %>% filter(sind <= 1020/J),
-                            newdata2 =  df_test %>% filter(sind > 1020/J), 
+                            newdata = df_test2 %>% filter(sind <= 1020/J),
+                            newdata2 =  df_test2 %>% filter(sind > 1020/J), 
                             type = "subject_specific", type_pred = "link",
                             se.fit = TRUE, return_newdata = TRUE)
 t2 <- Sys.time()
 t_pred_adglmm <- t2-t1 # 12.18 minutes spent on prediction
 
-save(adglmm_pred_t540, adglmm_pred_t780, adglmm_pred_t1020, file = here("Data/ApplOutput_GLMMadaptive.RData"))
+#### sanity checks ####
+
+id1 <- unique(df_test$id)
+id2 <- unique(adglmm_pred_t540$newdata2$id)
+setdiff(id1, id2)
[…]

### 2023-09-22 Docs/ProgressReport.Rmd
add GLMM adaptive model summary
re-run data application part and record time

@@
 - 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
-- 3.3 minutes spent on model fitting, 20 minutes spent on prediction (about 0.01 per subject)
-- Three subjects had failed approximation, all happened with the shortest observed track
-- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints.
-Here I wanna compare if this choice would make any different.
-Also, there are a log of other methods to fill in the gaps, such as projection.
-- For linear interpolation, I set the edges to be the closest data extreme.
+- Used PIRLS for local GLMMs step
 
@@
 # interpolation for each subject
+t1 <- Sys.time()
 for(i in unique(df_test_full$id)){
@@
 }
+t2 <- Sys.time()
+t_interp <- t2-t1
 
@@
 
-Let's look at four subjects as an example:
+- `r round(t_local_glmm, 2)` minutes spent on fitting local GLMMs, `r round(t_fpca, 2)` seconds on FPCA model, `r round(t_pred, 2)` minutes on out-of-sample prediction, `r round(t_interp, 2)` minutes on interpolation to the original grid
+
+- Let's look at four subjects as an example:
 
@@
 
+### A closer look at interpolation
+
+- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints.
+Here I wanna compare if this choice would make any different.
+Also, there are a lot of other methods to fill in the gaps, such as projection.
+- For linear interpolation, I set the edges to be the closest data extreme.
+- The interpolation itself spent `r round(t_interp, 2)` minutes.
+
 ```{r, warning=FALSE}
@@
 ```
-- Subjects with numeric issue
 
-```{r, fig.height=6, fig.width=9}
+### Numeric problems
+
+- Three subjects had failed approximation, all happened with the shortest observed track
+- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width).
+
+```{r, fig.height=3, fig.width=9}
 df %>% filter(id %in% skip_id) %>% 
@@
 
-- Calculated AUC
+<!-- Calculate AUC -->
 
@@
 This fitting procedure took 20.72 minutes to finish
-- Out-of-sample prediction took 12.18 minutes. About 0.007 minute per subject.
-- I'd also exclude the subjects with failed Laplace approximation here.
 
@@
 load(here("Data/ApplOutput_GLMMadaptive.RData"))
+```
+
+- `r round(t_est_adglmm, 2)` minutes on model fitting and `r round(t_pred_adglmm, 2)` minutes on out-of-sample prediction.
+
+- Below is the model fit on the trainign set
+
+```{r}
+summary(adglmm_mod)
 ```
@@
 
+- Below is an example of predicted probablity on four random subjects
+
 ```{r, warning=FALSE}
@@
 ```
+
+<!-- Calculate AUC  -->
 
@@
 
+- To use the same sample for comparison, I'd exclude the subjects with failed Laplace approximation from the test set for both methods
+
 ```{r}
@@
 colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 3)
+```
 
+```{r}
+# time
+t_fit <- c(t_local_glmm+t_fpca, t_local_glmm+t_fpca, t_est_adglmm)
+units(t_fit) <- "mins"
+
+t_test <- c(t_pred, t_pred+t_interp, t_pred_adglmm)
+# units(t_pred)
+
+t_total <- t_fit+t_test
+
+idx1 <- idx2 <- idx3 <- c(1, 3, 3, 3)
+names(idx1) <- c("Total time", as.character(round(t_total, 2)))
+names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
+names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))
+
+t_fit <- as.character(round(t_fit, 2))
+t_test <- as.character(round(t_test, 2))
+t_total <- as.character(round(t_total, 2))
+
+```
+
+```{r}
 tb_auc %>%
-  kable(digit = 4) %>%
+  kable(digit = 4, table.attr = "style = \"color: black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 6, "20.72" = 3), bold = F) %>%
-  add_header_above(c("Time spent on prediction" = 1, "20.0" = 6, "12.2" = 3), bold = F) %>% 
+  add_header_above(header=idx1) %>%
+  add_header_above(header=idx2) %>%
+  add_header_above(header=idx3) %>%
   add_header_above(c(" "=1, "fGFPCA"=3, "fGFPCA+interpolation" = 3, "GLMMadaptive"=3))
@@
 
-- fGFPCA almost always have higher AUC. It at least is as good as GLMMadaptive.
-- Computationally, fGFPCA is about 10 minute faster (33%).
+- fGFPCA does not always have higher AUC, but definitely most of the time. 
+- In fact there are two scenarios where GLMMadaptive overperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm.
+Both of them have a 4-hour interval apart from the maximum observation time and and prediction window.
+Could it have anything to do the cyclic trend in the data?
+Is 4h the length of a cycle?
+I tried to look at it but really couldn't see anything...there are too many points lumped together!
+
+```{r}
+df %>% filter(id %in% rand_id2) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  geom_vline(xintercept = c(540, 780, 1020, 1260))+
+  scale_x_continuous(breaks = c(540, 780, 1020, 1260),
+                     labels = c("9am", "1pm", "5pm", "9pm"))+
+  facet_wrap(~id)
+```
 

### 2023-09-22 Code/DataApplNHANES.R

@@ -65,11 +65,11 @@ train_bin_lst <- split(df_train, f = df_train$bin)
 # train_bin_lst[[1]] %>% View()
 
 # local GLMM and estimate latent function
-# near-unidentifiability issues 
+# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
 t1=Sys.time()
 df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
 t2= Sys.time()
-t_local_glmm <- t2-t1 # 3.3 minutes for model estimation
+t_local_glmm <- t2-t1 # 3.25 minutes for model estimation
 # no numeric warnings appeared
 
 lapply(df_est_latent, function(x)range(x$eta_hat))
@@ -81,9 +81,8 @@ df_est_latent <- bind_rows(df_est_latent) %>%
   select(-sind, -Y) %>% distinct(.) 
 
 # visualization of the estimated latent function
-rand_id <- sample(train_id, size = 4)
 df %>% 
-  filter(id %in% rand_id) %>%
+  filter(id %in% sample(train_id, size = 4)) %>%
   left_join(df_est_latent, by = c("id", "bin")) %>%
   mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
   ggplot()+
@@ -97,7 +96,11 @@ mat_est_unique <- matrix(df_est_latent$eta_hat,
                          nrow=length(train_id), 
                          ncol=n_bin, byrow = F) # row index subject, column binned time
 dim(mat_est_unique)
+
+t1 <- Sys.time()
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
+t2 <- Sys.time()
+t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
 
 dim(fpca_mod$efunctions) # 27 eigenfunctions total
 fpca_mod$evalues
@@ -156,7 +159,7 @@ for(i in seq_along(test_id)){
 }
 t2=Sys.time()
 
-t_pred = t2-t1 # total 20min for out-of-sample prediction
+t_pred = t2-t1 # total 19.79 min for out-of-sample prediction
 t_pred/length(test_id) # about 0.01 mins per subject 
 
 
@@ -184,10 +187,9 @@ df_test$pred_t540[df_test$bin<=540] <- NA
 df_test$pred_t780[df_test$bin<=780] <- NA
[…]

### 2023-09-25 Docs/ProgressReport.Rmd
some format

@@
 # fit local models
+t1 <- Sys.time()
 df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+t2 <- Sys.time()
+t_pirls <- t2-t1
 df_est_latent <- bind_rows(df_est_latent)
@@
 
-Let's first try to re-do the who local GLMM step using glmmTMB. Looks like no numeric issue occurred. It takes slightly longer time than PIRLS in glmer
+Let's first try to re-do the who local GLMM step using glmmTMB. Looks like no numeric issue occurred. It takes slightly shorter time (about 1 minute) than PIRLS in glmer
 
@@
 
-- Time: `r t_tmb` minutes
-- Mean/Intercept function: It looks like glmmTMB estimates lie right between PIRLS and AGQ! It takes slightly shorter time than PIRLS and mean function shoots more downward, not as bad as AGQ though.
+- Time: `r round(t_tmb, 2)` minutes
+- Mean/Intercept function: It looks like glmmTMB estimates lie right between PIRLS and AGQ!
+It takes slightly shorter time than PIRLS (`r round(t_pirls, 2)` minutes) and mean function shoots more downward, not as bad as AGQ though.
 
@@
   ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, col = "glmer", alpha = 0.5))+
-  geom_line(aes(x=sind, y=eta_hat_tmb, col = "glmmTMB", alpha = 0.5))+
+  geom_line(aes(x=sind, y=eta_hat, col = "glmer"))+
+  geom_line(aes(x=sind, y=eta_hat_tmb, col = "glmmTMB"))+
   geom_point(aes(x=sind, y = Y), size = 0.5)+
@@
   ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, col = "glmer", alpha = 0.5))+
-  geom_line(aes(x=bin, y=eta_hat_tmb, col = "glmmTMB", alpha = 0.5))+
+  geom_line(aes(x=bin, y=eta_hat, col = "glmer"))+
+  geom_line(aes(x=bin, y=eta_hat_tmb, col = "glmmTMB"))+
   geom_point(aes(x=bin, y = num), size = 0.5)+
@@
 
-So it looks like glmmTMB still overshoot downwards.
+So it looks like glmmTMB still overshoots downwards.
 Between the two methods, the mid-day part (active part for most people) seems to be very similar.
@@
 
-Hmm...FPCA seems to shoot the population mean downward in both methods, more so for glmmTMB.
-It looks like more rounds of model would make the population mean function more variable.
-Does that make sense?
+Hmm...Compared to conditional intercept, FPCA seems to shoot the population mean downward in PIRLS but upward for glmmTMB. Does that make sense?
 
@@
 - fGFPCA does not always have higher AUC, but definitely most of the time. 
-- In fact there are two scenarios where GLMMadaptive overperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm.
+- In fact there are two scenarios where GLMMadaptive outperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm.
 Both of them have a 4-hour interval apart from the maximum observation time and and prediction window.
@@
 - What would be the reference method to compare our performance to?
-Original GLMMadaptvie perhaps is feasible on teh entire dataset.
+Original GLMMadaptvie perhaps is feasible on the entire dataset.
 But what should we use if we wanna do sub-sample comparison?
 Should we even do that for this data application (or simulation alone)?
-- There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
-- Is interplation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
-- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the predictio window, making AUC for this single subjects impossible to calculate...
+- There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
+- Is interpolation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
+- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

### 2023-09-26 Docs/ProgressReport.Rmd
report on laplace approximation procedure

@@
 library(kableExtra)
+library(mvtnorm)
 theme_set(theme_minimal())
@@
 ```
-
-## fGFPCA Estimation
+## fGFPCA
 
@@
 # source(here("Code/OutsampBayes.R"))
-
 ```
 
-### Binning
+### Bin Data
 
-- Bin every 10 measures
+- Bin every 10 consecutive observations
 
@@
   labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
+
+df <- df %>% rename(id = SEQN, Y=Z)
+```
+### Data split
+
+```{r}
+id_vec <- unique(df$id)
+train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
+test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
 ```
 
-### Local GLMMs
+- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
 
-Following up on last report, when using glmer, the results from PRILS (nAGQ=0) seems more in line with experiences.
-As validation, let's fit the marginal model and look at the estimated population-average latent function, and compare it with the FPCA mean function.
-Specifically, at bin s the model would be $g(E(Y_i(s)))=f_0(s)$, with no subject-specific random effect.
+### Local GLMM
 
-Another validation I came up with is to look at the local intercepts.
-That is the $\beta_0(s)$ in local mixed models $g(E(Y_i(s)))=\beta_0(s)+b_i(x)$.
-This should correspond to the FPCA mean functions, as the population average level.
+- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
+- Fit on the training set
 
-Since the individual random effects are centered around zero, I believe the marginal $f_0(s)$ should look very much like the conditional $\beta_0(x)$ if the model local fit well.
-The result obviously is in favor of PRILS.n *But what is the reason for AGQ to overshoot downward?*
+```{r, cache=TRUE}
+# fit model on the training set
+df_train <- df %>% filter(id %in% train_id)
+train_bin_lst <- split(df_train, f = df_train$bin)
+
+# local GLMM and estimate latent function
+# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
+t1=Sys.time()
+df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+t2= Sys.time()
+t_local_glmm <- t2-t1 
+```
 
 ```{r}
-# split by each bin
-df <- df %>% rename(id = SEQN, Y=Z)
-df_bin_lst <- split(df, f = df$bin)
-```
-
-```{r cache=TRUE}
-# local marginal GLM
-local_marg <- function(df){
-    this_glm <- glm(Y ~ 1, data = df, family = binomial)
-    eta_hat <- predict(this_glm, type = "link")
-    df$eta_hat <- eta_hat
-    return(df)
-}
-
-df_est_local_marg <- lapply(df_bin_lst, function(x){local_marg(x)})
-
-# local conditional GLMM
-## PIRLS vs AGQ
-df_est_cond_mean_pirls <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 0)}) 
-df_est_cond_mean_agq <- lapply(df_bin_lst, function(x){mean_latent(x, n_node = 5)}) 
-```
-
-```{r, fig.width=12, fig.height=4}
-ggarrange(
-df_est_local_marg %>% bind_rows() %>%
-  select(bin, eta_hat) %>% distinct(.) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLM (marginal)", y = "latent mean")+
-  ylim(-25, 1),
-
-data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLMM (conditional) using PRILS", y = "latent mean")+
-  ylim(-25, 1),
-
-data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_agq)) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLMM (conditional) using AGQ", y = "latent mean")+
-  ylim(-25, 1),
-
-nrow = 1, align = "h"
-)
-```
-
-Let's for now trust the results from PIRLS more than AGQ, if we are using glmer.
-
-```{r cache=TRUE}
-# fit local GLMM and estimate latent function
-# to avoid near-unidentifiability issues, using PIRLS maximization with nAGQ = 0
-
-# fit local models
-t1 <- Sys.time()
-df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
-t2 <- Sys.time()
-t_pirls <- t2-t1
-df_est_latent <- bind_rows(df_est_latent)
-
-# keep midpoint values
-df_est_latent <- df_est_latent %>%
+df_est_latent <- bind_rows(df_est_latent) %>% 
   select(-sind, -Y) %>% distinct(.) 
-```
-
-```{r, message=FALSE}
+# example estimated latent function
 df %>% 
-  filter(id %in% rand_id) %>%
+  filter(id %in% sample(train_id, size = 4)) %>%
   left_join(df_est_latent, by = c("id", "bin")) %>%
@@
 ```{r, message=FALSE}
-# overview at estimated latent function
+
 df %>% 
-  filter(id %in% rand_id) %>%
+  filter(id %in% sample(train_id, size = 4)) %>%
   group_by(id, bin) %>%
@@
   left_join(df_est_latent, by = c("id", "bin")) %>%
-  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
   ggplot()+
+  geom_point(aes(x=bin, y=num), size = 0.5)+
   geom_line(aes(x=bin, y=eta_hat, group = id))+
-  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
-  facet_wrap(~id, scales = "free")+
-  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
-```
-
-#### An alternative: glmmTMB
-
-The default optimizer of glmmTMB is nlminb (quasi-Newton BFGS), whereas glmer uses bobyaqa (nAGQ=0) or Nelder_Mead (nAGQ>0)?
-
-Let's first try to re-do the who local GLMM step using glmmTMB. Looks like no numeric issue occurred. It takes slightly shorter time (about 1 minute) than PIRLS in glmer
-
-```{r, cache=TRUE}
-library(glmmTMB)
-# re-fit local GLMM and check numeric issues
-try_fit_lst<-list()
-t1=Sys.time()
-for(i in seq_along(df_bin_lst)){
-  # this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
-  #                      warning=function(w){NA})
-  this_glm <- glmmTMB(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial)
-  try_fit_lst[[i]] <- this_glm
-
-}
-df_est_latent_tmb <- lapply(try_fit_lst, function(x){predict(x, type = "link")})
-t2=Sys.time()
-t_tmb <- t2-t1
-# population mean
-mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
-```
-
-- Time: `r round(t_tmb, 2)` minutes
-- Mean/Intercept function: It looks like glmmTMB estimates lie right between PIRLS and AGQ!
-It takes slightly shorter time than PIRLS (`r round(t_pirls, 2)` minutes) and mean function shoots more downward, not as bad as AGQ though.
-
-```{r, fig.width=12, fig.height=4}
-ggarrange(
-df_est_local_marg %>% bind_rows() %>%
-  select(bin, eta_hat) %>% distinct(.) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLM (marginal)", y = "latent mean")+
-  ylim(-15, 1),
-
-data.frame(bin=mid, eta_hat = unlist(df_est_cond_mean_pirls)) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLMM (conditional) using PRILS (glmer)", y = "latent mean")+
-  ylim(-15, 1),
-
-data.frame(bin=mid, eta_hat = mean_tmb) %>%
-  ggplot(aes(x=bin, y=eta_hat))+
-  geom_line()+
-  labs(x="local GLMM (conditional) using glmmTMB", y = "latent mean")+
-  ylim(-15, 1),
-
-nrow = 1
-)
-```
-
-- Individual latent functions
-
-```{r}
-for(i in seq_along(df_est_latent_tmb)){
-  df_est_latent_tmb[[i]] <- data.frame(df_bin_lst[[i]], eta_hat = df_est_latent_tmb[[i]])
-}
-
-df_est_latent_tmb <- bind_rows(df_est_latent_tmb)
-
-# keep midpoint values
-df_est_latent_tmb <- df_est_latent_tmb %>%
-  select(-sind, -Y) %>% distinct(.)
-
-```
-
-```{r, message=FALSE, fig.height=8, fig.width=16}
-df %>% 
-  filter(id %in% rand_id) %>%
-  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
-  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
-            by = c("id", "bin")) %>%
-  mutate_at(vars(eta_hat, eta_hat_tmb), function(x){exp(x)/(1+exp(x))}) %>% 
-  ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, col = "glmer"))+
-  geom_line(aes(x=sind, y=eta_hat_tmb, col = "glmmTMB"))+
-  geom_point(aes(x=sind, y = Y), size = 0.5)+
-  facet_wrap(~id, scales = "free")+
-  labs(x = "Time", y = "Estimated latent function (probablity scale)")
-```
-
-```{r, message=FALSE, fig.height=8, fig.width=16}
-# overview at estimated latent function
-df %>% 
-  filter(id %in% rand_id) %>%
-  group_by(id, bin) %>%
-  summarise(num = sum(Y)) %>%
-  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
-  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
-            by = c("id", "bin")) %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, col = "glmer"))+
-  geom_line(aes(x=bin, y=eta_hat_tmb, col = "glmmTMB"))+
-  geom_point(aes(x=bin, y = num), size = 0.5)+
-  facet_wrap(~id, scales = "free")+
-  labs(x = "Time", y = "Estimated latent function")
-```
-
-So it looks like glmmTMB still overshoots downwards.
-Between the two methods, the mid-day part (active part for most people) seems to be very similar.
-Difference lies mostly at both ends, with consecutive inactive minutes.
-
-```{r, eval=FALSE}
-# numeric values of individual latent function
-df %>% 
-  filter(id %in% rand_id) %>%
-  group_by(id, bin) %>%
-  summarise(num = sum(Y)) %>%
-  left_join(df_est_latent %>% select(id, bin, eta_hat), by = c("id", "bin")) %>%
-  left_join(df_est_latent_tmb %>% select(id, bin, eta_hat) %>% rename(eta_hat_tmb=eta_hat), 
-            by = c("id", "bin")) %>%
-  head(20)
+  facet_wrap(~id)+
+  labs(x="Time", y = "Estimated latent function (original scale")
 ```
@@
 
-Follow up from late time, let's use PIRLS estimator and fix the number of PC functions instead of using all of them. We start with four PC functions.
+```{r}
+mat_est_unique <- matrix(df_est_latent$eta_hat,
+                         nrow=length(train_id), 
+                         ncol=n_bin, byrow = F) # row index subject, column binned time
+#dim(mat_est_unique)
 
-I will use estimated latent functions from both glmer (nAGQ=0) and glmmTMB, and compare their results.
+t1 <- Sys.time()
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
+t2 <- Sys.time()
+t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
 
-```{r}
-# glmer nAGQ=0
-mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
-
-# lmmTMB
-mat_est_unique2 <- matrix(df_est_latent_tmb$eta_hat, 
-                          nrow=N, ncol=n_bin, byrow = F) 
-fpca_mod2 <- fpca.face(mat_est_unique2, argvals = mid, var=T)
-
-# ncol(fpca_mod$efunctions) # 27 eigenfunctions
+# dim(fpca_mod$efunctions) # 27 eigenfunctions total
+# fpca_mod$evalues
 K <- 4
@@
 
-- Population mean function
+```{r}
+plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+```
 
-Hmm...Compared to conditional intercept, FPCA seems to shoot the population mean downward in PIRLS but upward for glmmTMB. Does that make sense?
+```{r, fig.height=8, fig.width=8}
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
+plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+```
 
 ```{r}
-# population mean
-par(mfrow=c(1, 2))
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmer",
-     ylim = c(-12, 1))
-plot(mid, fpca_mod2$mu, type = "l", xlab = "bin", ylab = "mean", main = "glmmTMB", ylim = c(-12, 1))
-```
-- PC functions
-
-```{r, fig.height=16, fig.width=8}
-# eigenfunctions
-par(mfrow=c(4,2))
-
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmer", ylim = c(-0.21, 0.13))
-plot(mid, fpca_mod2$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "glmmTMB", ylim = c(-0.21, 0.13))
-
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2", ylim = c(-0.2, 0.14))
-plot(mid, fpca_mod2$efunctions[, 2], type="l", xlab="bin", ylab="PC2",ylim = c(-0.2, 0.14))
-
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3", ylim = c(-0.2, 0.2))
-plot(mid, fpca_mod2$efunctions[, 3], type="l", xlab="bin", ylab="PC3",ylim = c(-0.2, 0.2))
-
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4", ylim = c(-0.14, 0.23))
-plot(mid, fpca_mod2$efunctions[, 4], type="l", xlab="bin", ylab="PC4",
-     ylim = c(-0.14, 0.23))
+# plot correlation matrix
+sum(fpca_mod$VarMats[[1]]!=fpca_mod$VarMats[[2]]) 
+heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
 ```
 
-- Individual covariance matrix
+### Debias?
 
-Wow! This covariacne matrix changed a lot from when using all of the PCs! It still looks cyclic, but now off-diagnoal region seems much more stable. Fluctuations mostly centered around diagnoal line.
+### Out-of-sample prediction
+
+follow up on the last report, I'll add some details here.
+
+Let's set:
+
+- $s$: bin index
+- $t_{m_s}$: midpoint time of bin s
+- $\mathcal{T}_s$: all observation points in bin s
+- $Y_i^s)$ all observations in bin s from subject i
+
+Let's assume we have a new subject $u$ with maximum observation time $T_u$. Then the log-likelihood of this new subject would be:
+
+$$
+l_u=\sum_{t_{m_s}<T_u}log(h(Y_u^s))+\hat{\eta}_u(t_{m_s})T(Y_u^s)-log(A[\hat{\eta}_u(t_{m_s})])
+$$
+
+where $\hat{\eta}_u(t_{m_s}) = \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})$.
+
+From FPCA model (or the debias step to be added), we obtain $\hat{\mu}_0$, $\hat{\phi}_k$, as well as the variance estimates of $\xi_k$: $\hat{\lambda}_k$, and variance of residual process $\hat{\sigma}^2$.
+
+With all these estimates, we wanna use Laplace approximation to find the $\xi_{uk}$ that maximizes $l_u$.
+
+I am using a package called **LaplacesDemon**, following this example https://cran.r-project.org/web/packages/LaplacesDemon/vignettes/LaplacesDemonTutorial.pdf
+
+```{r, class.source = "fold-show"}
+library(LaplacesDemon)
+```
+
+Now I am going to use one subject as an example to go over the approximation procedure:
+
+First we write out the model:
+
+- Prior distribution: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$
+- Posterior distribution:
+
+For a single bin
+$$
+\begin{aligned}
+l(\mathbf{Y_u}|\mathbf{\xi}_u) &= \sum_{t_{m_s}<T_u}l(Y_u^s|\mathbf{\xi}_u) \\
+
+Y_u^s|\mathbf{\xi}_u & \sim Binomial(n_s, p_s) \\
+
+g^{-1}(p_s) = \hat{\eta}(t_{m_s}) &= \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})
+\end{aligned}
+$$
+And $n_s$ is the number of observations in bin s.
+
+Bayes theroem:
+
+$$
+l(\mathbf{\xi}_u|\mathbf{Y}_u) \propto l(\mathbf{Y}_u|\mathbf{\xi}_u)+l(\mathbf{\xi}_u)
+$$
+
+```{r, class.source = "fold-show"}
+## model
+Model <- function(parm, Data){
+  xi <- parm[Data$pos.xi]
+
+  # log-prior
+  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)
+
+  # log-posterior likelihood
+  eta <- Data$f0+Data$X %*% xi
+  p <- exp(eta)/(1+exp(eta))
+  LL <- sum(dbinom(x=Data$y, size = Data$n, prob=p, log = TRUE)) # log likelihood of Y|xi
+  LP <- LL+sum(xi.prior) # unnormalized joint log likelihood of (Y, xi)
+
+  # output
+  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LP,
+                   yhat=Data$y, parm=parm)
+  return(Modelout)
+}
+```
+
+- Then we feed in data for one observation
+
+```{r, class.source = "fold-show"}
+rand_id <- sample(test_id, 1)
+df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am
+
+# put data into correct format
+ns <- as.vector(table(df_new$bin)) # number of observations
+hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
+nf <- ns-hs # number of failure
+max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
+df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)
+
+# extract estimates from FPCA model to be used in prior distribution
+tao <- diag(fpca_mod$evalues[1:K]) # variance of xi
+
+# extract estimates from FPCA model to be used in posterior distribution
+f0 <- fpca_mod$mu[1:max_bin]
+N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
+n <- df_new2$ns # number of experiements at each bin
+y <- df_new2$hs # outcome, number of 1
+X <- fpca_mod$efunctions[1:max_bin, 1:K] # eigen/PC functions
+J <- K # number of parameters/scores
+mon.names <- "LP"
+
+# parameter names
+name_lst <- as.list(rep(0, J))
+names(name_lst) <- paste("xi", 1:J, sep = "")
+parm.names <- as.parm.names(name_lst) # names of parameters to estimate
+pos.xi <- grep("xi", parm.names)
+
+# data
+MyData <- list(J=J, X=X, mon.names=mon.names,
+                 parm.names=parm.names, pos.xi=pos.xi, y=y, n=n, tao=tao, f0=f0)
+```
+
+```{r, class.source = "fold-show"}
+# fit laplace approximation
+# initial values generated by GIV function in the same package
+Fit <- LaplaceApproximation(Model, Data=MyData)
+```
+
+- Summarizing output
+
+```{r, class.source = "fold-show", fig.height=8, fig.width=8}
+# summarizing output
+print(Fit)
+plot(Fit, MyData)
+```
 
 ```{r}
-# covariance
-heatmap(fpca_mod$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
-        Rowv = NA, Colv = NA, main = "glmer")
+# prediction
+eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%Fit$Summary1[, "Mode"]
 
-heatmap(fpca_mod2$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
-        Rowv = NA, Colv = NA, main = "glmmTMB")
+data.frame(bin = mid, eta_pred = eta_pred_out) %>% 
+  right_join(df %>% filter(id==rand_id) %>%  select(Y, sind, bin), by = "bin") %>%
+  ggplot()+
+  geom_point(aes(x=sind, y = Y))+
+  geom_line(aes(x=sind, y = exp(eta_pred)/1+exp(eta_pred)))+
+  labs(title = rand_id)
 
@@
 
-## Dynamic prediction
+## Full data application output
 
-Still from last time, we're gonna focus on Bayes methods now!
+### fGFPCA
 
-We are also going to extend the observed track, so that there are some variation in the observed data (instead of all zeros). I am going to start with 9am (540), 1pm (780),  5pm (1020)
-
-Extending the track seems to solve the problem with failed Laplace Approximation. Not sure. Should try on the full sample.
-
-```{r, warning=FALSE, message=FALSE}
-df_pred <- df_est_latent %>% filter(id %in% rand_id)
-df_pred[, 'pred_t540'] <- df_pred[, 'pred_t780'] <- df_pred[, 'pred_t1020'] <- NA
-score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
-```
-
-```{r, results='hide'}
-# unique id
-# prediction for a single subject
-source(here("Code/OutsampBayes.R"))
-
-# PRILS
-for(i in rand_id){
-  df_i <- df %>% filter(id==i)
-
-  # prediction 
-  ## up to 360
-  pred_t1 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=540), kpc=K)
-  eta_pred_t1 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t1$score_out
-  df_pred[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
-  ## up to 720
-  pred_t2 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=780), kpc=K)
-  eta_pred_t2 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t2$score_out
-  df_pred[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
-  ## up to 1080
-  pred_t3 <- out_pred_laplace(fpca_mod, df_i %>% filter(sind<=1020), kpc = K)
-  eta_pred_t3 <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%pred_t3$score_out
-  df_pred[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3
-
-}
-
-df_pred$pred_t540[df_pred$bin<=540] <- NA
-df_pred$pred_t780[df_pred$bin<=780] <- NA
-df_pred$pred_t1020[df_pred$bin<=1020] <- NA
-```
-
-```{r, results='hide'}
-# glmmTMB
-df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
-df_pred2[, 'pred_t540'] <- df_pred2[, 'pred_t780'] <- df_pred2[, 'pred_t1020'] <- NA
-
-for(i in rand_id){
-  df_i <- df %>% filter(id==i)
-
-  # prediction 
-  ## up to 360
-  pred_t1 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=540), kpc=K)
-  eta_pred_t1 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t1$score_out
-  df_pred2[df_pred$id == i, 'pred_t540'] <- eta_pred_t1
-  ## up to 720
-  pred_t2 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=780), kpc=K)
-  eta_pred_t2 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t2$score_out
-  df_pred2[df_pred$id == i, 'pred_t780'] <- eta_pred_t2
-  ## up to 1080
-  pred_t3 <- out_pred_laplace(fpca_mod2, df_i %>% filter(sind<=1020), kpc = K)
-  eta_pred_t3 <- fpca_mod2$mu+fpca_mod2$efunctions[, 1:K]%*%pred_t3$score_out
-  df_pred2[df_pred$id == i, 'pred_t1020'] <- eta_pred_t3
-
-}
-
-df_pred2$pred_t540[df_pred$bin<=540] <- NA
-df_pred2$pred_t780[df_pred$bin<=780] <- NA
-df_pred2$pred_t1020[df_pred$bin<=1020] <- NA
-```
-
-```{r, warning=FALSE, fig.width=16, fig.height=8}
-ggarrange(
-df_pred %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, group = id))+
-  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
-  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
-  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
-  facet_wrap(~id,scales = "free")+labs(title = "glmer"),
-
-df_pred2 %>%
-  ggplot()+
-  geom_line(aes(x=bin, y=eta_hat, group = id))+
-  geom_line(aes(x=bin, y = pred_t540, col = "540", group = id))+
-  geom_line(aes(x=bin, y = pred_t780, col = "780", group = id))+
-  geom_line(aes(x=bin, y = pred_t1020, col = "1020", group = id))+
-  facet_wrap(~id,scales = "free")+labs(title = "glmmTMB"),
-nrow = 1, common.legend = T)
-```
-
-# NHANES data application output
-
-## fGFPCA
-
-- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
-- Used PIRLS for local GLMMs step
+- Still need to project eigenfunctions to the full grid
 
 ```{r}
+# results generated from DataApplNHANES.R
 load(here("Data/ApplOutput_fGFPCA.RData"))
@@
 
-```{r}
-# linear interpolation
-# remove subjects with numeric problem
+```{r message=FALSE, warning=FALSE}
 df_test_full <- df %>% 
-  filter(id %in% unique(df_test$id)) %>%
+  filter(id %in% test_id) %>%
   filter(!id %in% skip_id) %>%
   left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
-df_test_full$pred_t540_interp <- df_test_full$pred_t780_interp <- df_test_full$pred_t1020_interp <- NA
 
-# interpolation for each subject
-t1 <- Sys.time()
-for(i in unique(df_test_full$id)){
-  # interpolation t540
-  interp_i_t540 <- approx(x = df_test[df_test$id==i, "bin"], 
-                     y = df_test[df_test$id==i, "pred_t540"],
-                     xout = 541:J, rule = 2)
-  df_test_full$pred_t540_interp[df_test_full$id==i & df_test_full$sind>540] <- interp_i_t540$y
-
-  # interpolation t780
-  interp_i_t780 <- approx(x = df_test[df_test$id==i, "bin"], 
-                     y = df_test[df_test$id==i, "pred_t780"],
-                     xout = 781:J, rule = 2)
-  df_test_full$pred_t780_interp[df_test_full$id==i & df_test_full$sind>780] <- interp_i_t780$y
-
-  # interpolation t1020
-  interp_i_t1020 <- approx(x = df_test[df_test$id==i, "bin"], 
-                     y = df_test[df_test$id==i, "pred_t1020"],
-                     xout = 1021:J, rule = 2)
-  df_test_full$pred_t1020_interp[df_test_full$id==i & df_test_full$sind>1020] <- interp_i_t1020$y
-}
-t2 <- Sys.time()
-t_interp <- t2-t1
-
-```
-
-- `r round(t_local_glmm, 2)` minutes spent on fitting local GLMMs, `r round(t_fpca, 2)` seconds on FPCA model, `r round(t_pred, 2)` minutes on out-of-sample prediction, `r round(t_interp, 2)` minutes on interpolation to the original grid
-
-- Let's look at four subjects as an example:
-
-```{r message=FALSE, warning=FALSE}
 rand_id2 <- sample(unique(df_test$id), 4)
@@
 
-### A closer look at interpolation
-
-- We can either assume that the latent function value is constant within the same bin, or use linear interpolation to fill in the gaps between bin midpoints.
-Here I wanna compare if this choice would make any different.
-Also, there are a lot of other methods to fill in the gaps, such as projection.
-- For linear interpolation, I set the edges to be the closest data extreme.
-- The interpolation itself spent `r round(t_interp, 2)` minutes.
-
-```{r, warning=FALSE}
-# with interpolation
-df_test_full%>% 
-  filter(id %in% rand_id2) %>%
-  mutate_at(vars(pred_t540_interp, pred_t780_interp, pred_t1020_interp), function(x)exp(x)/(1+exp(x))) %>% 
-  ggplot()+
-    geom_line(aes(x=bin, y = pred_t540_interp, col = "9am"))+
-    geom_line(aes(x=bin, y = pred_t780_interp, col = "1pm"))+
-    geom_line(aes(x=bin, y = pred_t1020_interp, col = "5pm"))+
-    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)", 
-         title = "Interpolated latent function")
-
-```
-
 ### Numeric problems
@@
 - Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width). 
+- Per last meeting, adding some details on laplace approximation
 
@@
 
-```{r}
-# with interpolation
-# up to 9am
-auc_t540_interp <- lst_auc[2:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t540_interp, x$Y), measure = "auc"))
-auc_t540_interp <- lapply(auc_t540_interp, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 1pm
-auc_t780_interp <- lst_auc[3:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t780_interp, x$Y), measure = "auc"))
-auc_t780_interp <- lapply(auc_t780_interp, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 5pm
-auc_t1020_interp <- lst_auc[4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t1020_interp, x$Y), measure = "auc"))
-auc_t1020_interp <- lapply(auc_t1020_interp, function(x){x@y.values[[1]]}) %>% unlist()
-```
-
 ## GLMMadaptvie
@@
 
-- Below is the model fit on the trainign set
+- Below is the model fit on the training set
 
@@
 # up to 540
+J <- 1440
 auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
@@
            c(NA, NA, auc_t1020),
-           auc_t540_interp, 
-           c(NA, auc_t780_interp), 
-           c(NA, NA, auc_t1020_interp), 
+
            auc_t540_adglmm,
@@
 
-colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 3)
+colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
 ```
@@
 # time
-t_fit <- c(t_local_glmm+t_fpca, t_local_glmm+t_fpca, t_est_adglmm)
+t_fit <- c(t_local_glmm+t_fpca, t_est_adglmm)
 units(t_fit) <- "mins"
 
-t_test <- c(t_pred, t_pred+t_interp, t_pred_adglmm)
+t_test <- c(t_pred, t_pred_adglmm)
 # units(t_pred)
@@
 
-idx1 <- idx2 <- idx3 <- c(1, 3, 3, 3)
+idx1 <- idx2 <- idx3 <- c(1, 3, 3)
 names(idx1) <- c("Total time", as.character(round(t_total, 2)))
@@
   add_header_above(header=idx3) %>%
-  add_header_above(c(" "=1, "fGFPCA"=3, "fGFPCA+interpolation" = 3, "GLMMadaptive"=3))
+  add_header_above(c(" "=1, "fGFPCA"=3, "GLMMadaptive"=3))
 ```

### 2023-09-26 Docs/ProgressReport.Rmd
format

@@
 ```
-## fGFPCA
+## fGFPCA model fitting
 
@@
 
-### Out-of-sample prediction
+## Out-of-sample prediction
 
@@
 - $t_{m_s}$: midpoint time of bin s
-- $\mathcal{T}_s$: all observation points in bin s
-- $Y_i^s)$ all observations in bin s from subject i
+- $\mathscr{T}_s$: all observation points in bin s
+- $Y_i^s$ all observations in bin s from subject i
 
@@
   geom_point(aes(x=sind, y = Y))+
-  geom_line(aes(x=sind, y = exp(eta_pred)/1+exp(eta_pred)))+
+  geom_line(aes(x=sind, y = exp(eta_pred)/(1+exp(eta_pred))))+
   labs(title = rand_id)
@@
 
-## Full data application output
+# Full data application output
 
-### fGFPCA
+## fGFPCA
 

### 2023-09-26 Docs/ProgressReport.Rmd
more formatting

@@
 ```
+
 ## fGFPCA model fitting
@@
 df %>% 
-  filter(id %in% sample(train_id, size = 4)) %>%
+  filter(id %in% rand_id) %>%
   left_join(df_est_latent, by = c("id", "bin")) %>%
@@
 # plot correlation matrix
-sum(fpca_mod$VarMats[[1]]!=fpca_mod$VarMats[[2]]) 
 heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
@@
 - $\mathscr{T}_s$: all observation points in bin s
-- $Y_i^s$ all observations in bin s from subject i
+- $Y_i^s$ all observations in bin s from subject i. $Y_i^S := \{Y_i(t_j), t_j \in \mathscr{T}_s\}$
 
@@
 
-From FPCA model (or the debias step to be added), we obtain $\hat{\mu}_0$, $\hat{\phi}_k$, as well as the variance estimates of $\xi_k$: $\hat{\lambda}_k$, and variance of residual process $\hat{\sigma}^2$.
+From FPCA model (or the debias step to be added), we obtain $\hat{\mu}_0$, $\hat{\phi}_k$, as well as the variance estimates of $\xi_k$: $\hat{\lambda}_k$, and variance of residual process $\hat{\sigma^2}$.
 
@@
 
-For a single bin
-$$
-\begin{aligned}
+$$\begin{aligned}
 l(\mathbf{Y_u}|\mathbf{\xi}_u) &= \sum_{t_{m_s}<T_u}l(Y_u^s|\mathbf{\xi}_u) \\
@@
 g^{-1}(p_s) = \hat{\eta}(t_{m_s}) &= \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})
-\end{aligned}
-$$
+\end{aligned}$$
+
 And $n_s$ is the number of observations in bin s.
@@
 ```{r, class.source = "fold-show"}
-rand_id <- sample(test_id, 1)
+rand_id <- test_id[15]
 df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am

### 2023-09-27 Docs/ProgressReport.Rmd
try splines interpolation of mean and eigenfunctions

@@
 library(mvtnorm)
+library(mgcv)
+library(splines)
 theme_set(theme_minimal())
@@
 
-### Debias?
+### Debias (step 4)
+
+According to the fGFPCA paper, in this step we do two things:
+
+1. Interpolate the eigenfunctions with B-spline basis back to the original grid
+2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation
+- we do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
+
+First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. The model failed because of vector memory exhaustion. Shall I: 
+- Decrease the density of grid (increase bin width)
+- Increase memory (re-set to 100GB). But still took a lot of time (more than one hour). The model actually did not finish fitting.
+
+The bias was caused by the misspecification of latent process (assume constant effect over smooth function).
+And what needs re-evaluation are eigenvalues.
+Basis functions and mean functions are not biased.
+
+```{r, eval=FALSE}
+df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
+colnames(df_phi) <- c("bin", paste0("phi", 1:4))
+
+df_train <- df_train %>% left_join(df_phi, by = "bin")
+df_train$id <- as.factor(df_train$id)
+
+# usethis::edit_r_environ()
+t1 <- Sys.time()
+debias_glmm <- bam(Y ~ s(bin, bs="cr", k=10)+
+                     s(id, by=phi1, bs="re")+
+                     s(id, by=phi2, bs="re")+
+                     s(id, by=phi3, bs="re")+
+                     s(id, by=phi4, bs="re"), 
+                   family = binomial, data=df_train, method = "fREML")
+t2 <- Sys.time()
+t_debias <- t2=t1
+
+```
+
+Then I would like to interpolate eigenfunctions using B-spline basis functions:
+
+- Would you interpolate mean and eigenfunctions separately (in my case, five interpolations)? 
+- I did not find the reeval_efunctions function
+
+```{r, fig.height=5, fig.width=5}
+# mean
+interp_mean <- interpSpline(mid, fpca_mod$mu)
+full_mean <- predict(interp_mean, x=1:J)
+plot(mid, fpca_mod$mu, ylab = "mean", pch = 20)
+lines(full_mean$x, full_mean$y, col = "blue")
+
+# PCs
+interp_phi1 <- interpSpline(mid, fpca_mod$efunctions[, 1])
+full_phi1 <- predict(interp_phi1, x=1:J)
+plot(mid, fpca_mod$efunctions[, 1], ylab = "pc1", pch=20)
+lines(full_phi1$x, full_phi1$y, col = "blue")
+
+interp_phi2 <- interpSpline(mid, fpca_mod$efunctions[, 2])
+full_phi2 <- predict(interp_phi2, x=1:J)
+plot(mid, fpca_mod$efunctions[, 2], ylab = "pc2", pch=20)
+lines(full_phi2$x, full_phi2$y, col = "blue")
+
+interp_phi3 <- interpSpline(mid, fpca_mod$efunctions[, 3])
+full_phi3 <- predict(interp_phi3, x=1:J)
+plot(mid, fpca_mod$efunctions[, 3], ylab = "pc3", pch=20)
+lines(full_phi3$x, full_phi3$y, col = "blue")
+
+interp_phi4 <- interpSpline(mid, fpca_mod$efunctions[, 4])
+full_phi4 <- predict(interp_phi4, x=1:J)
+plot(mid, fpca_mod$efunctions[, 4], ylab = "pc4", pch=20)
+lines(full_phi4$x, full_phi4$y, col = "blue")
+```
 

### 2023-10-02 Docs/ProgressReport.Rmd
try the debias step

@@
 2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation
-- we do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
+
+- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
+- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
+- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
+- How are the eigenvalues biased? 
+- It looks like the FPCA step now estimates mean and eigenfunctions, while each eigenfunction serves as a random slope. Can we fit such model with an offset of mean/intercept function?
 
 First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. The model failed because of vector memory exhaustion. Shall I: 
-- Decrease the density of grid (increase bin width)
+- Decrease the density of grid (increase bin width)?
+- Decrease sample size? 
 - Increase memory (re-set to 100GB). But still took a lot of time (more than one hour). The model actually did not finish fitting.
 
-The bias was caused by the misspecification of latent process (assume constant effect over smooth function).
-And what needs re-evaluation are eigenvalues.
-Basis functions and mean functions are not biased.
-
-```{r, eval=FALSE}
+```{r, eval=FALSE, class.source = "fold-show"}
 df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
@@
 
-```{r, fig.height=5, fig.width=5}
+```{r, fig.height=5, fig.width=5, class.source = "fold-show"}
 # mean
@@
 ```
+
+Perhaps I should try with the simulation data.
 
@@
 
-$$\begin{aligned}
+\begin{aligned}
 l(\mathbf{Y_u}|\mathbf{\xi}_u) &= \sum_{t_{m_s}<T_u}l(Y_u^s|\mathbf{\xi}_u) \\
@@
 g^{-1}(p_s) = \hat{\eta}(t_{m_s}) &= \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})
-\end{aligned}$$
+\end{aligned}
 

### 2023-10-02 Code/Simulation.R

@@ -1,16 +1,17 @@
 # This script implements simulation for WNAR manuscript
 
-set.seed(314)
+set.seed(516)
 
 library(here)
 library(tidyverse)
-# library(ggplot2)
-# theme_set(theme_minimal())
+theme_set(theme_minimal())
 # library(fcr)
 library(refund)
 library(lme4)
 library(ROCR)
 library(GLMMadaptive)
+library(mgcv)
+library(splines)
 # library(ggpubr)
 # library(gridExtra)
 # library(kableExtra)
@@ -22,13 +23,27 @@ library(GLMMadaptive)
 
 load(here("Data/sim_data.RData"))
 
+# data structure
+M <- length(sim_data) # number of simulation
+N <- length(unique(sim_data[[1]]$id))
+J <- length(unique(sim_data[[1]]$sind))
 
-#### simulation set up ####
+# data preview
+rand_id <- sample(as.numeric(unique(sim_data[[1]]$id)), size=4)
 
-J <- 1000 # number of observations points
-N <- 500 # sample size
-K <- 4 # number of eigenfunctions
-M <- 50 # number of simulation
+sim_data[[1]] %>% 
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=exp(eta_i)/(1+exp(eta_i))))+
+  facet_wrap(~id)
+
+
+##### fGFPCA Model estimation #####
+
+source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
+# source(here("Code/OutSampMLE.R"))
+
 
 # bin data
 bin_w <- 10 # bin width
@@ -36,16 +51,13 @@ n_bin <- J/bin_w # number of bins
 brks <- seq(0, J, by = bin_w) # cutoff points
 mid <- (brks+bin_w/2)[1:n_bin] # mid points
 
-#### fGFPCA ####
-source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
-source(here("Code/OutSampMLE.R"))
 
 # result container
 pred_lst <- list()
 score_lst <- list()
 runtime <- rep(NA, M)
 
-
+m <- 1
 
 pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 
@@ -62,12 +74,40 @@ for(m in 1:M){
   df_bin_lst <- split(df, f = df$bin)
   
   # fit local GLMM and estimate latent function
[…]

### 2023-10-02 Code/SimulationOneIter.R

@@ -1,0 +1,188 @@
+# This script implements simulation 
+# but only for one iterarion
+# as an explorative example
+
+set.seed(516)
+
+library(here)
+library(tidyverse)
+theme_set(theme_minimal())
+# library(fcr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+library(splines)
+# library(ggpubr)
+# library(gridExtra)
+# library(kableExtra)
+
+
+
+
+#### load simulated data ####
+
+load(here("Data/sim_data.RData"))
+
+# data structure
+M <- length(sim_data) # number of simulation
+N <- length(unique(sim_data[[1]]$id))
+J <- length(unique(sim_data[[1]]$sind))
+
+# data preview
+rand_id <- sample(as.numeric(unique(sim_data[[1]]$id)), size=4)
+
+sim_data[[1]] %>% 
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=exp(eta_i)/(1+exp(eta_i))))+
+  facet_wrap(~id)
+
+
+##### fGFPCA Model estimation #####
+
+df <- sim_data[[1]]
+source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
+# source(here("Code/OutSampMLE.R"
+
+# Step 1: bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
+table(df$bin)
+df$bin <- as.numeric(as.character(df$bin))
+
+## preview of binned data
+df %>% filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
+  ggplot(aes(x=bin, y=num))+
+  geom_point(size = 0.5)+
+  facet_wrap(~id)
+
+# Step 2: Local GLMMs
+df_bin_lst <- split(df, f = df$bin)
+lapply(df_bin_lst, dim)
+
+t1 <- Sys.time()
+df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+t2 <- Sys.time()
[…]

### 2023-10-02 Docs/ProgressReport.Rmd
fit debias model in the fast way

@@
 
-df <- read_rds(here("Data/nhanes_bi.rds"))
+# df <- read_rds(here("Data/nhanes_bi.rds"))
 # load(here("Data/ApplOutput_fGFPCA.RData"))
 
-```
-
-# NHANES data
-
-## Data overview
-
-- 8763 subjects, 1440 measures each
-- no missingness
-
-```{r}
-N <- length(unique(df$SEQN)) # sample size 8763
-J <- max(df$sind) # 1440 measures per subject
-```
-
-```{r}
-rand_id <- sample(unique(df$SEQN), size = 4) # "toy" sample
-
-df %>% 
-  filter(SEQN %in% rand_id) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Z), size = 0.5)+
-  facet_wrap(~SEQN)+
-  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
-```
-
-## fGFPCA model fitting
-
-```{r}
 # code
@@
 # source(here("Code/OutsampBayes.R"))
+
 ```
 
-### Bin Data
+# Simulation
+
+I am now on the way to add the debias step to the whole algorithm. Just so that I have a set of "true parameters" to comparision, I am gonna use the simulation dataset for this part.
+
+## Generation mechanism
+
+- I am gonna generate only one dataset as an example
+- 500 individuals each with 1000 measures
+
+\[\begin{aligned}
+Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
+\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
+\end{aligned}\]
+
+where:
+
+- $t$ is equal-spaced on $[0, 1]$
+- $f_0(t)=0$
+- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively.
+
+```{r gen_data, class.source='show'}
+N <- 500 # sample size
+J <- 1000 # number of observation points
+
+t = seq(0,1,len=J) # observations points
+
+# mean function
+f_0 <- function(s) 0
+
+#eigenfunctions 
+K <- 4 # number of eigenfunctions
+phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
+                     sin(4*pi*t),cos(4*pi*t))
+
+# eigenvalues
+lambda = 0.5^(0:(K-1))
+
+# generated data 
+## score
+xi <- matrix(rnorm(N*K),N,K)
+xi <- xi %*% diag(sqrt(lambda))
+
+# subject-specific random effect
+b_i <- xi %*% t(phi); # of size N by J
+
+# latent gaussian function
+eta_i <- t(vapply(1:N, function(x){
+    f_0(t) + b_i[x,]
+  }, numeric(J)))
+
+# outcome binary function
+Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
+                N, J, byrow=FALSE)
+
+# format into dataframe
+# id = subject identifier (factor variable)
+# sind = numeric value corresponding to the observed functional domain
+# Y = functional response (binary)
+# sind_inx = numeric value associated with order of "sind"
+#            this is not necessary, but may be of convenience when you implement the method
+df <- data.frame(id = factor(rep(1:N, each=J)),
+                 t = rep(t, N), 
+                 Y = as.vector(t(Y_i)),
+                 eta_i = as.vector(t(eta_i)),
+                 sind = rep(1:J, N))
+# visualization
+rand_id <- sample(N, size = 4)
+df %>% filter(id %in% rand_id) %>% 
+  ggplot()+
+  geom_point(aes(x=sind, y=Y))+
+  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
+  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
+  facet_wrap(~id)
+```
+
+## fGFPCA model fitting
+
+### Bin ata
 
@@
 df$bin <- as.numeric(as.character(df$bin))
-# head(df)
+# unique(df$bin)
 
 df %>% 
-  filter(SEQN %in% rand_id) %>%
-  group_by(SEQN, bin) %>%
-  summarise(num = sum(Z)) %>%
+  filter(id %in% rand_id) %>%
+  group_by(id, bin) %>%
+  summarise(num = sum(Y)) %>%
   ggplot()+
   geom_point(aes(x=bin, y=num), size = 0.5)+
-  facet_wrap(~SEQN)+
+  facet_wrap(~id)+
   labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
+```
 
-df <- df %>% rename(id = SEQN, Y=Z)
-```
-### Data split
+## Data split
+
+- 400 subjects for training, 100 for out-of-sample prediction
 
 ```{r}
-id_vec <- unique(df$id)
-train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
-test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
+train_id <- sample(unique(df$id), size = N*0.8)
+test_id <- setdiff(unique(df$id), train_id)
+
+train_df <- df %>% filter(id %in% train_id)
+test_df <- df %>% filter(id %in% test_id)
 ```
-
-- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
 
@@
 # fit model on the training set
-df_train <- df %>% filter(id %in% train_id)
-train_bin_lst <- split(df_train, f = df_train$bin)
+train_bin_lst <- split(train_df, f = train_df$bin)
 
@@
 ```{r}
-df_est_latent <- bind_rows(df_est_latent) %>% 
-  select(-sind, -Y) %>% distinct(.) 
+df_est_latent <- bind_rows(df_est_latent) 
+# head(df_est_latent)
+
 # example estimated latent function
-df %>% 
+df_est_latent %>% 
   filter(id %in% rand_id) %>%
-  left_join(df_est_latent, by = c("id", "bin")) %>%
   mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  mutate(eta_i = exp(eta_i)/(1+exp(eta_i))) %>%
   ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, group = id))+
+  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
+  geom_line(aes(x=sind, y=eta_i, group = id, col = "true"))+
   geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
@@
 
-```{r, message=FALSE}
-
-df %>% 
-  filter(id %in% sample(train_id, size = 4)) %>%
-  group_by(id, bin) %>%
-  summarise(num = sum(Y)) %>%
-  left_join(df_est_latent, by = c("id", "bin")) %>%
-  ggplot()+
-  geom_point(aes(x=bin, y=num), size = 0.5)+
-  geom_line(aes(x=bin, y=eta_hat, group = id))+
-  facet_wrap(~id)+
-  labs(x="Time", y = "Estimated latent function (original scale")
-```
-
 ### FPCA
@@
 ```{r}
-mat_est_unique <- matrix(df_est_latent$eta_hat,
+uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+
+mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                          nrow=length(train_id), 
-                         ncol=n_bin, byrow = F) # row index subject, column binned time
-#dim(mat_est_unique)
+                         ncol=n_bin, byrow = F) 
+# row index subject, column binned time
+# dim(mat_est_unique)
 
@@
 t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
-
-# dim(fpca_mod$efunctions) # 27 eigenfunctions total
-# fpca_mod$evalues
-K <- 4
 ```
@@
 
-1. Interpolate the eigenfunctions with B-spline basis back to the original grid
+1. Project the eigenfunctions with B-spline basis back to the original grid
 2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation
@@
 - We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
-- How are the eigenvalues biased? 
-- It looks like the FPCA step now estimates mean and eigenfunctions, while each eigenfunction serves as a random slope. Can we fit such model with an offset of mean/intercept function?
 
-First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. The model failed because of vector memory exhaustion. Shall I: 
-- Decrease the density of grid (increase bin width)?
-- Decrease sample size? 
-- Increase memory (re-set to 100GB). But still took a lot of time (more than one hour). The model actually did not finish fitting.
+#### Debias with GLMM model
 
-```{r, eval=FALSE, class.source = "fold-show"}
+First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions.
+I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
+
+```{r}
 df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
@@
 
-df_train <- df_train %>% left_join(df_phi, by = "bin")
-df_train$id <- as.factor(df_train$id)
+# train_df$bin <- as.numeric(as.character(train_df$bin))
+train_df <- train_df %>% left_join(df_phi, by = "bin")
+train_df$id <- as.factor(train_df$id)
+```
 
+- Here, what if I replace "bin" with "t"?
+
+```{r, class.source = "fold-show"}
 # usethis::edit_r_environ()
 t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(bin, bs="cr", k=10)+
+debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                      s(id, by=phi1, bs="re")+
@@
                      s(id, by=phi4, bs="re"), 
-                   family = binomial, data=df_train, method = "fREML")
+                   family = binomial, data=train_df %>% filter(bin==sind), 
+                   method = "fREML",
+                   discrete = TRUE)
 t2 <- Sys.time()
-t_debias <- t2=t1
-
+t_debias <- t2-t1
 ```
 
-Then I would like to interpolate eigenfunctions using B-spline basis functions:
+- And now I try to extract the variance of scores (debiased eigenvalues). 
+- Use the conclusion that connects smoothing parameters to variance
+
+$$\hat{p}_k=1/\hat{\lambda}_k^2$$
+
+```{r}
+fpca_mod$evalues
+new_lambda <- 1/debias_glmm$sp[2:5]
+new_lambda/lambda
+fpca_mod$evalues[1:4]/new_lambda
+```
+
+- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.
+
+#### Projection
+
+Then I would like to project eigenfunctions using B-spline basis functions:
 
@@
 Perhaps I should try with the simulation data.
+
+# NHANES data
+
+## Data overview
+
+- 8763 subjects, 1440 measures each
+- no missingness
+
+```{r}
+N <- length(unique(df$SEQN)) # sample size 8763
+J <- max(df$sind) # 1440 measures per subject
+```
+
+```{r}
+rand_id <- sample(unique(df$SEQN), size = 4) # "toy" sample
+
+df %>% 
+  filter(SEQN %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Z), size = 0.5)+
+  facet_wrap(~SEQN)+
+  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
+```
 

### 2023-10-03 Docs/ProgressReportAppl.Rmd
explore projection of eigenfunctions

--- 2023-09-27 Docs/ProgressReport.Rmd
@@
 ---
-title: "Progress Report"
+title: "Progress Report on NHANES Data"
 author: "Ying Jin"
-date: "2023-09-15"
-output: 
-  html_document:
-    self_contained: no
-    number_sections: true
-    code_folding: hide
-    toc: true
-    toc_depth: 3
-    toc_float: true
-    font: 12pt
+date: "2023-10-03"
+output: html_document
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
-
-set.seed(516)
-
-library(here)
-library(tidyverse)
-library(ggpubr)
-library(refund)
-library(lme4)
-library(ROCR)
-library(GLMMadaptive)
-library(kableExtra)
-library(mvtnorm)
-library(mgcv)
-library(splines)
-theme_set(theme_minimal())
-
-df <- read_rds(here("Data/nhanes_bi.rds"))
-# load(here("Data/ApplOutput_fGFPCA.RData"))
-
 ```
@@
   labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
-```
-
-## fGFPCA model fitting
-
-```{r}
-# code
-source(here("Code/GLMM-FPCA.R")) 
-# use pred_latent function to estimate latent function 
-source(here("Code/OutSampMLE.R"))
-# source(here("Code/OutsampBayes.R"))
-```
-
-### Bin Data
-
-- Bin every 10 consecutive observations
-
-```{r, message=FALSE}
-# bin data
-bin_w <- 10 # bin width
-n_bin <- J/bin_w # number of bins
-brks <- seq(0, J, by = bin_w) # cutoff points
-mid <- (brks+bin_w/2)[1:n_bin] # mid points
-
-df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
-df$bin <- as.numeric(as.character(df$bin))
-# head(df)
-
-df %>% 
-  filter(SEQN %in% rand_id) %>%
-  group_by(SEQN, bin) %>%
-  summarise(num = sum(Z)) %>%
-  ggplot()+
-  geom_point(aes(x=bin, y=num), size = 0.5)+
-  facet_wrap(~SEQN)+
-  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
-
-df <- df %>% rename(id = SEQN, Y=Z)
-```
-### Data split
-
-```{r}
-id_vec <- unique(df$id)
-train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
-test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
-```
-
-- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
-
-### Local GLMM
-
-- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
-- Fit on the training set
-
-```{r, cache=TRUE}
-# fit model on the training set
-df_train <- df %>% filter(id %in% train_id)
-train_bin_lst <- split(df_train, f = df_train$bin)
-
-# local GLMM and estimate latent function
-# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
-t1=Sys.time()
-df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
-t2= Sys.time()
-t_local_glmm <- t2-t1 
-```
-
-```{r}
-df_est_latent <- bind_rows(df_est_latent) %>% 
-  select(-sind, -Y) %>% distinct(.) 
-# example estimated latent function
-df %>% 
-  filter(id %in% rand_id) %>%
-  left_join(df_est_latent, by = c("id", "bin")) %>%
-  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, group = id))+
-  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
-  facet_wrap(~id, scales = "free")+
-  labs(x = "Time", y = "Estimated latent function (probablity scale)")
-```
-
-```{r, message=FALSE}
-
-df %>% 
-  filter(id %in% sample(train_id, size = 4)) %>%
-  group_by(id, bin) %>%
-  summarise(num = sum(Y)) %>%
-  left_join(df_est_latent, by = c("id", "bin")) %>%
-  ggplot()+
-  geom_point(aes(x=bin, y=num), size = 0.5)+
-  geom_line(aes(x=bin, y=eta_hat, group = id))+
-  facet_wrap(~id)+
-  labs(x="Time", y = "Estimated latent function (original scale")
-```
-
-### FPCA
-
-```{r}
-mat_est_unique <- matrix(df_est_latent$eta_hat,
-                         nrow=length(train_id), 
-                         ncol=n_bin, byrow = F) # row index subject, column binned time
-#dim(mat_est_unique)
-
-t1 <- Sys.time()
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
-t2 <- Sys.time()
-t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
-
-# dim(fpca_mod$efunctions) # 27 eigenfunctions total
-# fpca_mod$evalues
-K <- 4
-```
-
-```{r}
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
-```
-
-```{r, fig.height=8, fig.width=8}
-par(mfrow=c(2,2))
-
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
-```
-
-```{r}
-# plot correlation matrix
-heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
-```
-
-### Debias (step 4)
-
-According to the fGFPCA paper, in this step we do two things:
-
-1. Interpolate the eigenfunctions with B-spline basis back to the original grid
-2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation
-- we do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
-
-First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. The model failed because of vector memory exhaustion. Shall I: 
-- Decrease the density of grid (increase bin width)
-- Increase memory (re-set to 100GB). But still took a lot of time (more than one hour). The model actually did not finish fitting.
-
-The bias was caused by the misspecification of latent process (assume constant effect over smooth function).
-And what needs re-evaluation are eigenvalues.
-Basis functions and mean functions are not biased.
-
-```{r, eval=FALSE}
-df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
-colnames(df_phi) <- c("bin", paste0("phi", 1:4))
-
-df_train <- df_train %>% left_join(df_phi, by = "bin")
-df_train$id <- as.factor(df_train$id)
-
-# usethis::edit_r_environ()
-t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(bin, bs="cr", k=10)+
-                     s(id, by=phi1, bs="re")+
-                     s(id, by=phi2, bs="re")+
-                     s(id, by=phi3, bs="re")+
-                     s(id, by=phi4, bs="re"), 
-                   family = binomial, data=df_train, method = "fREML")
-t2 <- Sys.time()
-t_debias <- t2=t1
-
-```
-
-Then I would like to interpolate eigenfunctions using B-spline basis functions:
-
-- Would you interpolate mean and eigenfunctions separately (in my case, five interpolations)? 
-- I did not find the reeval_efunctions function
-
-```{r, fig.height=5, fig.width=5}
-# mean
-interp_mean <- interpSpline(mid, fpca_mod$mu)
-full_mean <- predict(interp_mean, x=1:J)
-plot(mid, fpca_mod$mu, ylab = "mean", pch = 20)
-lines(full_mean$x, full_mean$y, col = "blue")
-
-# PCs
-interp_phi1 <- interpSpline(mid, fpca_mod$efunctions[, 1])
-full_phi1 <- predict(interp_phi1, x=1:J)
-plot(mid, fpca_mod$efunctions[, 1], ylab = "pc1", pch=20)
-lines(full_phi1$x, full_phi1$y, col = "blue")
-
-interp_phi2 <- interpSpline(mid, fpca_mod$efunctions[, 2])
-full_phi2 <- predict(interp_phi2, x=1:J)
-plot(mid, fpca_mod$efunctions[, 2], ylab = "pc2", pch=20)
-lines(full_phi2$x, full_phi2$y, col = "blue")
-
-interp_phi3 <- interpSpline(mid, fpca_mod$efunctions[, 3])
-full_phi3 <- predict(interp_phi3, x=1:J)
-plot(mid, fpca_mod$efunctions[, 3], ylab = "pc3", pch=20)
-lines(full_phi3$x, full_phi3$y, col = "blue")
-
-interp_phi4 <- interpSpline(mid, fpca_mod$efunctions[, 4])
-full_phi4 <- predict(interp_phi4, x=1:J)
-plot(mid, fpca_mod$efunctions[, 4], ylab = "pc4", pch=20)
-lines(full_phi4$x, full_phi4$y, col = "blue")
 ```

### 2023-10-03 Docs/ProgressReportSim.Rmd
explore projection of eigenfunctions

--- 2023-10-02 Docs/ProgressReport.Rmd
@@
 ---
-title: "Progress Report"
+title: "Progress Report on Simulation"
 author: "Ying Jin"
@@
 
-# Simulation
+I am now on the way to add the debias step to the whole algorithm. Just so that I have a set of "true parameters" for comparison, I am gonna use the simulation dataset for this part.
 
-I am now on the way to add the debias step to the whole algorithm. Just so that I have a set of "true parameters" to comparision, I am gonna use the simulation dataset for this part.
-
-## Generation mechanism
+# Generation mechanism
 
@@
   ggplot()+
-  geom_point(aes(x=sind, y=Y))+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
   geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
@@
 
-## fGFPCA model fitting
+# fGFPCA model fitting
 
-### Bin ata
+## Bin data
 
@@
 
-### Local GLMM
+## Local GLMM
 
@@
 # example estimated latent function
+rand_id <- sample(train_id, 4)
+
 df_est_latent %>% 
@@
 
-### Debias (step 4)
+## Debias (step 4)
 
@@
 
+### Porjection with B-spline basis
+
+- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly
+- Not sure how to decide the knots. I am using the bin boundaries for now (0, 10, 20, ...,1000), which is 101 knots.
+
+```{r, class.source='fold-show'}
+# order of b splines
+p <- 3
+
+# number of knots
+# not sure how to determine
+# maybe for now use the bin endpoints? 
+# knot_pos <- seq(0, J, by = 50)
+knot_pos <- brks
+
+# evaluate B-splines on binned grid
+B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
+                            outer.ok = TRUE)$design
+# evaluate B-splines on original grid
+Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
+                  outer.ok = TRUE)$design
+
+# project binned eigenfunctions onto the original grid
+efunctions_new <- matrix(NA, J, K)
+for(k in 1:K){
+    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+    efunctions_new[,k] <- Bnew %*% coef(lm_mod)
+}
+```
+
+```{r, fig.height=8, fig.width=8}
+# visualization
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
+lines(1:J, efunctions_new[, 1], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
+lines(1:J, efunctions_new[, 2], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
+lines(1:J, efunctions_new[, 3], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
+lines(1:J, efunctions_new[, 4], col="blue")
+
+```
+
+- The projected eigenfunctions seem to be more wiggly than we would like?
+- What is the reason for choosing projection over linear/spline interpolation?
+
+#### Number of knots
+
+- Try fewer knots?
+
+```{r}
+knot_pos <- seq(0, J, by = 100)
+
+# evaluate B-splines on binned grid
+B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
+                            outer.ok = TRUE)$design
+# evaluate B-splines on original grid
+Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
+                  outer.ok = TRUE)$design
+
+# project binned eigenfunctions onto the original grid
+efunctions_new_11knots <- matrix(NA, J, K)
+for(k in 1:K){
+    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+    efunctions_new_11knots[,k] <- Bnew %*% coef(lm_mod)
+}
+```
+
+```{r, fig.height=8, fig.width=8}
+# visualization
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_11knots[, 1], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_11knots[, 2], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_11knots[, 3], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_11knots[, 4], col="blue")
+
+```
+
+#### Order of spline
+
+- Try lower order? (order = 3)
+
+```{r}
+knot_pos <- brks
+
+# evaluate B-splines on binned grid
+B <- spline.des(knots = knot_pos, x = mid, ord = 3,
+                            outer.ok = TRUE)$design
+# evaluate B-splines on original grid
+Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = 3,
+                  outer.ok = TRUE)$design
+
+# project binned eigenfunctions onto the original grid
+efunctions_new_2p <- matrix(NA, J, K)
+for(k in 1:K){
+    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+    efunctions_new_2p[,k] <- Bnew %*% coef(lm_mod)
+}
+```
+
+```{r, fig.height=8, fig.width=8}
+# visualization
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_2p[, 1], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_2p[, 2], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_2p[, 3], col="blue")
+
+plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
+lines(1:J, efunctions_new_2p[, 4], col="blue")
+
+```
+
+### Debias with GLMM model
+
 - The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
@@
 - We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
-
-#### Debias with GLMM model
-
-First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions.
-I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
+- I will try to fit the model with projected eigenfunctions as above (with 101 knots). I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
 
 ```{r}
-df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
-colnames(df_phi) <- c("bin", paste0("phi", 1:4))
+df_phi <- data.frame(sind = 1:J, efunctions_new)
+colnames(df_phi) <- c("sind", paste0("phi", 1:4))
 
 # train_df$bin <- as.numeric(as.character(train_df$bin))
-train_df <- train_df %>% left_join(df_phi, by = "bin")
+train_df <- train_df %>% left_join(df_phi, by = "sind")
 train_df$id <- as.factor(train_df$id)
@@
 
-- Here, what if I replace "bin" with "t"?
+- Here, what if I replace "bin" with "t"? Should I regress of Y on t or sind?
 
@@
                      s(id, by=phi4, bs="re"), 
-                   family = binomial, data=train_df %>% filter(bin==sind), 
+                   family = binomial, data=train_df, 
                    method = "fREML",
@@
 
-```{r}
-fpca_mod$evalues
+```{r, class.source='fold-show'}
+# eigenvalues from FPCA
+fpca_mod$evalues[1:4]
+# transformed smoothing parameters
 new_lambda <- 1/debias_glmm$sp[2:5]
-new_lambda/lambda
-fpca_mod$evalues[1:4]/new_lambda
+new_lambda
+# true eigenvalues
+lambda
 ```
@@
 - Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.
-
-#### Projection
-
-Then I would like to project eigenfunctions using B-spline basis functions:
-
-- Would you interpolate mean and eigenfunctions separately (in my case, five interpolations)? 
-- I did not find the reeval_efunctions function
-
-```{r, fig.height=5, fig.width=5, class.source = "fold-show"}
-# mean
-interp_mean <- interpSpline(mid, fpca_mod$mu)
-full_mean <- predict(interp_mean, x=1:J)
-plot(mid, fpca_mod$mu, ylab = "mean", pch = 20)
-lines(full_mean$x, full_mean$y, col = "blue")
-
-# PCs
-interp_phi1 <- interpSpline(mid, fpca_mod$efunctions[, 1])
-full_phi1 <- predict(interp_phi1, x=1:J)
-plot(mid, fpca_mod$efunctions[, 1], ylab = "pc1", pch=20)
-lines(full_phi1$x, full_phi1$y, col = "blue")
-
-interp_phi2 <- interpSpline(mid, fpca_mod$efunctions[, 2])
-full_phi2 <- predict(interp_phi2, x=1:J)
-plot(mid, fpca_mod$efunctions[, 2], ylab = "pc2", pch=20)
-lines(full_phi2$x, full_phi2$y, col = "blue")
-
-interp_phi3 <- interpSpline(mid, fpca_mod$efunctions[, 3])
-full_phi3 <- predict(interp_phi3, x=1:J)
-plot(mid, fpca_mod$efunctions[, 3], ylab = "pc3", pch=20)
-lines(full_phi3$x, full_phi3$y, col = "blue")
-
-interp_phi4 <- interpSpline(mid, fpca_mod$efunctions[, 4])
-full_phi4 <- predict(interp_phi4, x=1:J)
-plot(mid, fpca_mod$efunctions[, 4], ylab = "pc4", pch=20)
-lines(full_phi4$x, full_phi4$y, col = "blue")
-```
-
-Perhaps I should try with the simulation data.
-
-# NHANES data
-
-## Data overview
-
-- 8763 subjects, 1440 measures each
-- no missingness
-
-```{r}
-N <- length(unique(df$SEQN)) # sample size 8763
-J <- max(df$sind) # 1440 measures per subject
-```
-
-```{r}
-rand_id <- sample(unique(df$SEQN), size = 4) # "toy" sample
-
-df %>% 
-  filter(SEQN %in% rand_id) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Z), size = 0.5)+
-  facet_wrap(~SEQN)+
-  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
-```
-
-## Out-of-sample prediction
-
-follow up on the last report, I'll add some details here.
-
-Let's set:
-
-- $s$: bin index
-- $t_{m_s}$: midpoint time of bin s
-- $\mathscr{T}_s$: all observation points in bin s
-- $Y_i^s$ all observations in bin s from subject i. $Y_i^S := \{Y_i(t_j), t_j \in \mathscr{T}_s\}$
-
-Let's assume we have a new subject $u$ with maximum observation time $T_u$. Then the log-likelihood of this new subject would be:
-
-$$
-l_u=\sum_{t_{m_s}<T_u}log(h(Y_u^s))+\hat{\eta}_u(t_{m_s})T(Y_u^s)-log(A[\hat{\eta}_u(t_{m_s})])
-$$
-
-where $\hat{\eta}_u(t_{m_s}) = \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})$.
-
-From FPCA model (or the debias step to be added), we obtain $\hat{\mu}_0$, $\hat{\phi}_k$, as well as the variance estimates of $\xi_k$: $\hat{\lambda}_k$, and variance of residual process $\hat{\sigma^2}$.
-
-With all these estimates, we wanna use Laplace approximation to find the $\xi_{uk}$ that maximizes $l_u$.
-
-I am using a package called **LaplacesDemon**, following this example https://cran.r-project.org/web/packages/LaplacesDemon/vignettes/LaplacesDemonTutorial.pdf
-
-```{r, class.source = "fold-show"}
-library(LaplacesDemon)
-```
-
-Now I am going to use one subject as an example to go over the approximation procedure:
-
-First we write out the model:
-
-- Prior distribution: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$
-- Posterior distribution:
-
-\begin{aligned}
-l(\mathbf{Y_u}|\mathbf{\xi}_u) &= \sum_{t_{m_s}<T_u}l(Y_u^s|\mathbf{\xi}_u) \\
-
-Y_u^s|\mathbf{\xi}_u & \sim Binomial(n_s, p_s) \\
-
-g^{-1}(p_s) = \hat{\eta}(t_{m_s}) &= \hat{\mu}_0(t_{m_s})+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_{m_s})
-\end{aligned}
-
-And $n_s$ is the number of observations in bin s.
-
-Bayes theroem:
-
-$$
-l(\mathbf{\xi}_u|\mathbf{Y}_u) \propto l(\mathbf{Y}_u|\mathbf{\xi}_u)+l(\mathbf{\xi}_u)
-$$
-
-```{r, class.source = "fold-show"}
-## model
-Model <- function(parm, Data){
-  xi <- parm[Data$pos.xi]
-
-  # log-prior
-  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)
-
-  # log-posterior likelihood
-  eta <- Data$f0+Data$X %*% xi
-  p <- exp(eta)/(1+exp(eta))
-  LL <- sum(dbinom(x=Data$y, size = Data$n, prob=p, log = TRUE)) # log likelihood of Y|xi
-  LP <- LL+sum(xi.prior) # unnormalized joint log likelihood of (Y, xi)
-
-  # output
-  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LP,
-                   yhat=Data$y, parm=parm)
-  return(Modelout)
-}
-```
-
-- Then we feed in data for one observation
-
-```{r, class.source = "fold-show"}
-rand_id <- test_id[15]
-df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am
-
-# put data into correct format
-ns <- as.vector(table(df_new$bin)) # number of observations
-hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
-nf <- ns-hs # number of failure
-max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
-df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)
-
-# extract estimates from FPCA model to be used in prior distribution
-tao <- diag(fpca_mod$evalues[1:K]) # variance of xi
-
-# extract estimates from FPCA model to be used in posterior distribution
-f0 <- fpca_mod$mu[1:max_bin]
-N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
-n <- df_new2$ns # number of experiements at each bin
-y <- df_new2$hs # outcome, number of 1
-X <- fpca_mod$efunctions[1:max_bin, 1:K] # eigen/PC functions
-J <- K # number of parameters/scores
-mon.names <- "LP"
-
-# parameter names
-name_lst <- as.list(rep(0, J))
-names(name_lst) <- paste("xi", 1:J, sep = "")
-parm.names <- as.parm.names(name_lst) # names of parameters to estimate
-pos.xi <- grep("xi", parm.names)
-
-# data
-MyData <- list(J=J, X=X, mon.names=mon.names,
-                 parm.names=parm.names, pos.xi=pos.xi, y=y, n=n, tao=tao, f0=f0)
-```
-
-```{r, class.source = "fold-show"}
-# fit laplace approximation
-# initial values generated by GIV function in the same package
-Fit <- LaplaceApproximation(Model, Data=MyData)
-```
-
-- Summarizing output
-
-```{r, class.source = "fold-show", fig.height=8, fig.width=8}
-# summarizing output
-print(Fit)
-plot(Fit, MyData)
-```
-
-```{r}
-# prediction
-eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%Fit$Summary1[, "Mode"]
-
-data.frame(bin = mid, eta_pred = eta_pred_out) %>% 
-  right_join(df %>% filter(id==rand_id) %>%  select(Y, sind, bin), by = "bin") %>%
-  ggplot()+
-  geom_point(aes(x=sind, y = Y))+
-  geom_line(aes(x=sind, y = exp(eta_pred)/(1+exp(eta_pred))))+
-  labs(title = rand_id)
-
-```
-
-# Full data application output
-
-## fGFPCA
-
-- Still need to project eigenfunctions to the full grid
-
-```{r}
-# results generated from DataApplNHANES.R
-load(here("Data/ApplOutput_fGFPCA.RData"))
-```
-
-```{r message=FALSE, warning=FALSE}
-df_test_full <- df %>% 
-  filter(id %in% test_id) %>%
-  filter(!id %in% skip_id) %>%
-  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
-
-rand_id2 <- sample(unique(df_test$id), 4)
-# without interpolation, assume constant latent function value in each bin
-df_test_full%>%
-  filter(id %in% rand_id2) %>%
-  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>%
-  ggplot()+
-    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
-    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
-    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
-    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)",
-         title = "Constant latent function within bin")
-```
-
-### Numeric problems
-
-- Three subjects had failed approximation, all happened with the shortest observed track
-- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width). 
-- Per last meeting, adding some details on laplace approximation
-
-```{r, fig.height=3, fig.width=9}
-df %>% filter(id %in% skip_id) %>% 
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  facet_wrap(~id)+
-  geom_vline(xintercept = c(540))
-```
-
-<!-- Calculate AUC -->
-
-```{r}
-# break by prediction window
-lst_auc <- df_test_full %>% 
-  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
-                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-lst_auc <- split(lst_auc, f=lst_auc$window)
-```
-
-```{r}
-# no interpolation
-# up to 9am
-auc_t540 <- lst_auc[2:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
-auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 1pm
-auc_t780 <- lst_auc[3:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t780, x$Y), measure = "auc"))
-auc_t780 <- lapply(auc_t780, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 5pm
-auc_t1020 <- lst_auc[4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t1020, x$Y), measure = "auc"))
-auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
-```
-
-## GLMMadaptvie
-
-- when fitting model with **random intercept + slope** on the same full training set (even after scaling covariates): Error: vector memory exhausted (limit reached?).
-- when fitting model with **only random intercept** on the same full training set: Error regarding a large coefficient value.
-One fix is to re-scale covariates.
-So I re-scaled minute index by dividing them by J, so the range is within (0, 1], and tried again.
-This fitting procedure took 20.72 minutes to finish
-
-```{r}
-load(here("Data/ApplOutput_GLMMadaptive.RData"))
-```
-
-- `r round(t_est_adglmm, 2)` minutes on model fitting and `r round(t_pred_adglmm, 2)` minutes on out-of-sample prediction.
-
-- Below is the model fit on the training set
-
-```{r}
-summary(adglmm_mod)
-```
-
-```{r}
-df_test_full2 <- df_test_full %>%
-  filter(id %in% rand_id2)%>%
-  select(id, sind, Y)
-
-df_test_full2$pred_t540 <- df_test_full2$pred_t780 <- df_test_full2$pred_t1020 <- NA 
-for(i in rand_id2){
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>540, "pred_t540"]<- adglmm_pred_t540$newdata2 %>% filter(id ==i) %>% select(pred)
-
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>780, "pred_t780"]<- adglmm_pred_t780$newdata2 %>% filter(id ==i) %>% select(pred)
-
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>1020, "pred_t1020"]<- adglmm_pred_t1020$newdata2 %>% filter(id ==i) %>% select(pred)
-}
-
-# colSums(is.na(df_test_full2))
-```
-
-- Below is an example of predicted probablity on four random subjects
-
-```{r, warning=FALSE}
-df_test_full2 %>% 
-  mutate_at(vars(pred_t540, pred_t780, pred_t1020), 
-            function(x)exp(x)/(1+exp(x))) %>%
-  ggplot()+
-    geom_line(aes(x=sind, y = pred_t540, col = "9am"))+
-    geom_line(aes(x=sind, y = pred_t780, col = "1pm"))+
-    geom_line(aes(x=sind, y = pred_t1020, col = "5pm"))+
-    geom_point(aes(x=sind, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)")
-```
-
-<!-- Calculate AUC  -->
-
-```{r}
-# up to 540
-J <- 1440
-auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
-  filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
-                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
-auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 780
-auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
-  filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
-                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
-auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 1020
-auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
-  filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
-                      labels = c("5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
-auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
-
-```
-
-## Compare fGFPCA and GLMMadaptive
-
-- To use the same sample for comparison, I'd exclude the subjects with failed Laplace approximation from the test set for both methods
-
-```{r}
-options(knitr.kable.NA = '')
-tb_auc <- data.frame(auc_t540, 
-           c(NA, auc_t780), 
-           c(NA, NA, auc_t1020),
-
-           auc_t540_adglmm,
-           c(NA, auc_t780_adglmm), 
-           c(NA, NA, auc_t1020_adglmm))
-
-colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
-```
-
-```{r}
-# time
-t_fit <- c(t_local_glmm+t_fpca, t_est_adglmm)
-units(t_fit) <- "mins"
-
-t_test <- c(t_pred, t_pred_adglmm)
-# units(t_pred)
-
-t_total <- t_fit+t_test
-
-idx1 <- idx2 <- idx3 <- c(1, 3, 3)
-names(idx1) <- c("Total time", as.character(round(t_total, 2)))
-names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
-names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))
-
-t_fit <- as.character(round(t_fit, 2))
-t_test <- as.character(round(t_test, 2))
-t_total <- as.character(round(t_total, 2))
-
-```
-
-```{r}
-tb_auc %>%
-  kable(digit = 4, table.attr = "style = \"color: black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(header=idx1) %>%
-  add_header_above(header=idx2) %>%
-  add_header_above(header=idx3) %>%
-  add_header_above(c(" "=1, "fGFPCA"=3, "GLMMadaptive"=3))
-```
-
-- fGFPCA does not always have higher AUC, but definitely most of the time. 
-- In fact there are two scenarios where GLMMadaptive outperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm.
-Both of them have a 4-hour interval apart from the maximum observation time and and prediction window.
-Could it have anything to do the cyclic trend in the data?
-Is 4h the length of a cycle?
-I tried to look at it but really couldn't see anything...there are too many points lumped together!
-
-```{r}
-df %>% filter(id %in% rand_id2) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  geom_vline(xintercept = c(540, 780, 1020, 1260))+
-  scale_x_continuous(breaks = c(540, 780, 1020, 1260),
-                     labels = c("9am", "1pm", "5pm", "9pm"))+
-  facet_wrap(~id)
-```
-
-## Discussion
-
-- What would be the reference method to compare our performance to?
-Original GLMMadaptvie perhaps is feasible on the entire dataset.
-But what should we use if we wanna do sub-sample comparison?
-Should we even do that for this data application (or simulation alone)?
-- There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
-- Is interpolation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
-- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

### 2023-10-03 Code/reeval_efunctions.R

@@ -1,0 +1,41 @@
+#' Get new spline basis for eigenfunctions
+#'
+#' Internal function for generating spline basis from fpca.face results to evaluate eigenfunctions
+#' on expanded grid.
+#'
+#' @param knots Number of knots for spline basis
+#' @param argvals_bin Index on for binned data.
+#' @param efunctions Eigenfunctions from an fpca object.
+#' @param argvals Full index on which to generate new basis.
+#' @param npc Number of principal components.
+#'
+#' @importFrom stats lm
+#' @importFrom splines spline.des
+#'
+reeval_efunctions <- function(knots, argvals_bin, argvals, efunctions, npc){
+
+  p <- 3 # order of b splines
+
+  ## get knot location in the same way that fpca.face() does internally
+  ## (see refund:::pspline.setting())
+  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+  knots_values <- knots_values * (max(argvals_bin) - min(argvals_bin)) + min(argvals_bin)
+
+  ## evaluate B splines on original grid
+  B <- splines::spline.des(knots = knots_values, x = argvals_bin, ord = p + 1,
+                            outer.ok = TRUE)$design
+
+  Bnew <- splines::spline.des(knots = knots_values, x = argvals, ord = p + 1,
+                              outer.ok = TRUE)$design
+
+  ## project first eigenfunction onto the B spline basis used in model fitting
+  efunctions_new <- matrix(NA, length(argvals), npc)
+  for(k in 1:npc){
+    lm_mod <- lm(efunctions[,k] ~ B-1)
+    efunctions_new[,k] <- Bnew %*% coef(lm_mod)
+  }
+
+  efunctions_new
+
+}
+

### 2023-10-03 Docs/ProgressReportAppl.Rmd
fix laplace failures

@@
 date: "2023-10-03"
-output: html_document
+output: 
+  html_document:
+    self_contained: no
+    number_sections: true
+    code_folding: hide
+    toc: true
+    toc_depth: 3
+    toc_float: true
+    font: 12pt
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+set.seed(516)
+
+library(here)
+library(tidyverse)
+library(ggpubr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(kableExtra)
+library(mvtnorm)
+library(mgcv)
+library(splines)
+theme_set(theme_minimal())
 ```
 
-# NHANES data
+```{r}
+# load data
+df <- read_rds(here("Data/nhanes_bi.rds"))
 
-## Data overview
+# code
+source(here("Code/GLMM-FPCA.R")) 
+# use pred_latent function to estimate latent function 
+source(here("Code/OutSampMLE.R"))
+# source(here("Code/OutsampBayes.R"))
+```
+
+# Data overview
 
@@
 
-## Out-of-sample prediction
+# Fit fGFPCA model
+
+## Bin data
+
+- Bin every 10 observations
+
+```{r, message=FALSE}
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
+df$bin <- as.numeric(as.character(df$bin))
+# unique(df$bin)
+
+df %>% 
+  filter(SEQN %in% rand_id) %>%
+  group_by(SEQN, bin) %>%
+  summarise(num = sum(Z)) %>%
+  ggplot()+
+  geom_point(aes(x=bin, y=num), size = 0.5)+
+  facet_wrap(~SEQN)+
+  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
+```
+## Data split
+
+- 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction
+
+```{r}
+# head(df)
+df <- df %>% rename(id=SEQN, Y=Z)
+
+train_id <- sample(unique(df$id), size = N*0.8)
+test_id <- setdiff(unique(df$id), train_id)
+
+train_df <- df %>% filter(id %in% train_id)
+test_df <- df %>% filter(id %in% test_id)
+```
+
+## Local GLMM
+
+- Used glmer with PIRLS (nAGQ=0) for local GLMMs step
+- Fit on the training set
+
+```{r, cache=TRUE}
+# fit model on the training set
+train_bin_lst <- split(train_df, f = train_df$bin)
+
+# local GLMM and estimate latent function
+# use PIRLS (nAGQ=0) to avoid near-unidentifiability issues 
+t1=Sys.time()
+df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+t2= Sys.time()
+t_local_glmm <- t2-t1 
+```
+
+```{r}
+df_est_latent <- bind_rows(df_est_latent) 
+# head(df_est_latent)
+
+# example estimated latent function
+rand_id <- sample(train_id, 4)
+
+df_est_latent %>% 
+  filter(id %in% rand_id) %>%
+  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
+  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
+  facet_wrap(~id, scales = "free")+
+  labs(x = "Time", y = "Estimated latent function (probablity scale)")
+```
+
+## FPCA
+
+```{r}
+uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+# head(uni_eta_hat)
+
+mat_est_unique <- matrix(uni_eta_hat$eta_hat,
+                         nrow=length(train_id), 
+                         ncol=n_bin, byrow = F) 
+# row index subject, column binned time
+# dim(mat_est_unique)
+
+t1 <- Sys.time()
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
+t2 <- Sys.time()
+t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
+
+# dim(fpca_mod$efunctions) # 27 eigenfunctions
+K <- 4
+```
+
+```{r}
+plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+```
+
+```{r, fig.height=8, fig.width=8}
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
+plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+```
+
+```{r}
+# plot correlation matrix
+heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
+```
+
+## Debias (step 4)
+
+According to the fGFPCA paper, in this step we do two things:
+
+1. Project the eigenfunctions with B-spline basis back to the original grid
+2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation
+
+Becuase I would like to focus this report on subjects with failed Laplace approximation, I will not do these two things for now.
+
+# Out-of-sample prediction
 
@@
 
-Now I am going to use one subject as an example to go over the approximation procedure:
-
 First we write out the model:
@@
 
+## One subject example
+
+Now I am going to use one subject as an example to go over the approximation procedure:
 ```{r, class.source = "fold-show"}
@@
 
-# Full data application output
+## Full data application output
 
-## fGFPCA
-
-- Still need to project eigenfunctions to the full grid
+### fGFPCA
 
@@
   filter(id %in% test_id) %>%
-  filter(!id %in% skip_id) %>%
+  # filter(!id %in% skip_id) %>%
   left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
@@
          title = "Constant latent function within bin")
-```
-
-### Numeric problems
-
-- Three subjects had failed approximation, all happened with the shortest observed track
-- Does it have to do with still the length of observed track? Or the arbitrary choices in the binning procedure (bin width). 
-- Per last meeting, adding some details on laplace approximation
-
-```{r, fig.height=3, fig.width=9}
-df %>% filter(id %in% skip_id) %>% 
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  facet_wrap(~id)+
-  geom_vline(xintercept = c(540))
 ```
@@
 auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
-  filter(!id %in% skip_id) %>%
+  # filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%
@@
 auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
-  filter(!id %in% skip_id) %>%
+  # filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%
@@
 auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
-  filter(!id %in% skip_id) %>%
+  # filter(!id %in% skip_id) %>%
   mutate(sind=sind*J) %>%
@@
 
-## Compare fGFPCA and GLMMadaptive
+### AUC
 
@@
 
+## Failed Laplace approximation
+
+- This problem seems to be solved after changing the optimizer? (Method = "BFGS")
+- The new optimizer also took less time
+- The old one was SPG (large scale optimization). The new one is BFGS (quasi-newton, the same in glmmTMB)
+
 ## Discussion
@@
 Should we even do that for this data application (or simulation alone)?
-- There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects. 
-- Is interpolation really necessary? It does not improve performance very much but consumes time. Also, should I try more sophisticated methods for grid extension? 
-- Should I calculate subject-wise AUC and average them? But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...
+<!-- - There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.  -->
+<!-- - Is interpolation really necessary?
+It does not improve perfor- Should I calculate subject-wise AUC and average them?
+But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

### 2023-10-04 Docs/ProgressReportAppl.Rmd
calculated ISE for one iterarion

@@
 - The new optimizer also took less time
-- The old one was SPG (large scale optimization). The new one is BFGS (quasi-newton, the same in glmmTMB)
+- The old one was SPG (Spectral projected gradient), which is a non-monotone line search. 
+- The new one is BFGS (quasi-newton, the same in glmmTMB)
 

### 2023-10-04 Docs/ProgressReportSim.Rmd
calculated ISE for one iterarion

@@
 ---
-title: "Progress Report on Simulation"
+title: "Progress Report on One Simulated Dataset"
 author: "Ying Jin"
@@
 library(kableExtra)
+library(knitr)
 library(mvtnorm)
@@
 source(here("Code/GLMM-FPCA.R")) 
-# use pred_latent function to estimate latent function 
-source(here("Code/OutSampMLE.R"))
+# source(here("Code/OutSampMLE.R"))
+source(here("Code/OutsampBayes.R"))
 # source(here("Code/OutsampBayes.R"))
@@
 - $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively. 
+- In fact, since there is a constant factor $\sqrt{2}$ before each eigenfunction, perhaps the real eigenvalues are $2, 1, 0,5, 0.25$.
 
@@
 
-- 400 subjects for training, 100 for out-of-sample prediction
+- 400 (80%) subjects for training, 100 (20%) for out-of-sample prediction
 
@@
 train_id <- sample(unique(df$id), size = N*0.8)
+# length(train_id)
 test_id <- setdiff(unique(df$id), train_id)
@@
 
+- Estimated mean
+
 ```{r}
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "Mean")
 ```
+
+- Estimated eigenfunctions
 
@@
 
+- Correlation matrix
+
 ```{r}
@@
 heatmap(cov2cor(fpca_mod$VarMats[[1]]), Rowv = NA, Colv = NA, main = "Correlation")
+```
+
+- Estimated eigenvalues
+
+```{r, class.source='fold-show'}
+# estimated eigenvalues from FPCA (biased)
+fpca_mod$evalues[1:4]
 ```
@@
 # knot_pos <- seq(0, J, by = 50)
-knot_pos <- brks
 
 # evaluate B-splines on binned grid
-B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
+B <- spline.des(knots = brks, x = mid, ord = p + 1,
                             outer.ok = TRUE)$design
 # evaluate B-splines on original grid
-Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
+Bnew <- spline.des(knots = brks, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design
@@
 
-- The projected eigenfunctions seem to be more wiggly than we would like?
+- The projected eigenfunctions seem to be more wiggly than we would like.
 - What is the reason for choosing projection over linear/spline interpolation?
@@
 
-- Try fewer knots?
+- Try fewer knots (11 knots)? 
+- Looks like the shape does not fit to the binned grid very well.
 
 ```{r}
+# 11 knots
 knot_pos <- seq(0, J, by = 100)
@@
 - Try lower order? (order = 3)
+- While the middle part fits alright, the end really goes crazy off.
 
 ```{r}
-knot_pos <- brks
-
 # evaluate B-splines on binned grid
-B <- spline.des(knots = knot_pos, x = mid, ord = 3,
+B <- spline.des(knots = brks, x = mid, ord = 3,
                             outer.ok = TRUE)$design
 # evaluate B-splines on original grid
-Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = 3,
+Bnew <- spline.des(knots = brks, x = 1:J, ord = 3,
                   outer.ok = TRUE)$design
@@
 
-### Debias with GLMM model
+### Debias with eigenvalues
 
@@
 - What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
-- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
-- I will try to fit the model with projected eigenfunctions as above (with 101 knots). I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
+- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction.
+And out-of-sample scores are not estimated by FPCA but Laplace approximation.
+- Since I don't yet know what to do with the projection, I will here use the un-projected basis functions and do things on the binned grid.
+I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
 
 ```{r}
-df_phi <- data.frame(sind = 1:J, efunctions_new)
-colnames(df_phi) <- c("sind", paste0("phi", 1:4))
+# dim(fpca_mod$efunctions)
+df_phi <- data.frame(bin = mid, fpca_mod$efunctions[, 1:K])
+colnames(df_phi) <- c("bin", paste0("phi", 1:4))
 
 # train_df$bin <- as.numeric(as.character(train_df$bin))
-train_df <- train_df %>% left_join(df_phi, by = "sind")
+train_df <- train_df %>% left_join(df_phi, by = "bin")
 train_df$id <- as.factor(train_df$id)
@@
 
-- Here, what if I replace "bin" with "t"? Should I regress of Y on t or sind?
+- Also, I am not exactly sure if I should regress on t (original time between 0 and 1) or the index of bin midpoints.
+I will do the bin midpoints for now because I think that's in the fGFPCA manuscript?
 
@@
 t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
+debias_glmm <- bam(Y ~ s(bin, bs="cr", k=10)+
                      s(id, by=phi1, bs="re")+
@@
 
-- And now I try to extract the variance of scores (debiased eigenvalues). 
-- Use the conclusion that connects smoothing parameters to variance
+And now I try to extract the variance of scores (debiased eigenvalues), using the conclusion that connects smoothing parameters to variance
 
@@
 ```{r, class.source='fold-show'}
-# eigenvalues from FPCA
-fpca_mod$evalues[1:4]
 # transformed smoothing parameters
@@
 new_lambda
-# true eigenvalues
-lambda
+# new_lambda/lambda
+# new_lambda/fpca_mod$evalues[1:4]
 ```
 
-- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.
+Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.
+
+How can we know this is the actual true eigenvalues?
+
+# Out-of-sample prediction with Laplace approximation
+
+I will now estimate individual score of test sample using Laplace Approximation, and calculated the predicted track.
+- Mximum observations time: 200, 400, 600. 800
+- Prediction window: 200-400, 400-600, 600-800, 800-1000
+
+## With eigenvalues from FPCA
+
+```{r, results='hide'}
+# container
+pred_list <- list()
+# score <- array(NA, dim=c(length(test_id), K, 4))
+# dimensions: subject, eigenfunction, maximum observation time
+
+# prediction for a single subject
+t1 <- Sys.time()
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+  # prediction 
+  pred1 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = fpca_mod$evalues[1:K], 
+                            phi_mat = fpca_mod$efunctions[, 1:K], 
+                            df_new = df_i %>% filter(sind<=200), kpc = K)
+  pred2 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = fpca_mod$evalues[1:K], 
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=400), kpc = K)
+  pred3 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = fpca_mod$evalues[1:K], 
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=600), kpc = K)
+  pred4 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = fpca_mod$evalues[1:K], 
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=800), kpc = K)
+
+  pred_i <- data.frame(id = test_id[i], bin=mid, 
+                       pred1=pred1$eta_pred, pred2=pred2$eta_pred,
+                       pred3=pred3$eta_pred, pred4=pred4$eta_pred)
+
+  pred_i$pred1[pred_i$bin<=200] <- NA
+  pred_i$pred2[pred_i$bin<=400] <- NA
+  pred_i$pred3[pred_i$bin<=600] <- NA
+  pred_i$pred4[pred_i$bin<=800] <- NA
+
+  pred_list[[i]] <- pred_i
+}
+t2 <- Sys.time()
+t_pred <- t2-t1
+```
+
+```{r}
+# overview of predicted track
+rand_test_id <- sample(test_id, 4)
+
+test_df %>%
+  left_join(bind_rows(pred_list), by = c("id", "bin")) %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred1, pred2, pred3, pred4), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  geom_line(aes(x=sind, y=eta_i, col = "True"))+
+  geom_line(aes(x=sind, y=pred1, col = "200"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred2, col = "400"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred3, col = "600"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred4, col = "800"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)
+
+```
+
+## With debiased eigenvalues
+
+```{r, results='hide'}
+# container
+pred_list2 <- list()
+# score <- array(NA, dim=c(length(test_id), K, 4))
+# dimensions: subject, eigenfunction, maximum observation time
+
+# prediction for a single subject
+t1 <- Sys.time()
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+  # prediction 
+  pred1 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = new_lambda, 
+                            phi_mat = fpca_mod$efunctions[, 1:K], 
+                            df_new = df_i %>% filter(sind<=200), kpc = K)
+  pred2 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = new_lambda, 
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=400), kpc = K)
+  pred3 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = new_lambda,
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=600), kpc = K)
+  pred4 <- out_pred_laplace(mu = fpca_mod$mu, 
+                            evalues = new_lambda,
+                            phi_mat = fpca_mod$efunctions[, 1:K],
+                            df_i %>% filter(sind<=800), kpc = K)
+
+  pred_i <- data.frame(id = test_id[i], bin=mid, 
+                       pred1=pred1$eta_pred, pred2=pred2$eta_pred,
+                       pred3=pred3$eta_pred, pred4=pred4$eta_pred)
+
+  pred_i$pred1[pred_i$bin<=200] <- NA
+  pred_i$pred2[pred_i$bin<=400] <- NA
+  pred_i$pred3[pred_i$bin<=600] <- NA
+  pred_i$pred4[pred_i$bin<=800] <- NA
+
+  pred_list2[[i]] <- pred_i
+}
+t2 <- Sys.time()
+t_pred <- t2-t1
+```
+
+```{r}
+# overview of predicted track
+test_df %>%
+  left_join(bind_rows(pred_list2), by = c("id", "bin")) %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred1, pred2, pred3, pred4), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  geom_line(aes(x=sind, y=eta_i, col = "True"))+
+  geom_line(aes(x=sind, y=pred1, col = "200"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred2, col = "400"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred3, col = "600"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred4, col = "800"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)
+```
+
+```{r}
+t_1sim <- (t_local_glmm+t_fpca+t_debias+t_pred)
+```
+
+- Time spent on one iteration: `r round(t_1sim, 2)` seconds
+- Did bias improved performance? The lines above look like so. I also wanna see the squared errors and AUC.
+
+## ISE
+
+```{r}
+# calcualte ISE
+err1 <- test_df %>%
+  left_join(bind_rows(pred_list), by = c("id", "bin")) %>%
+  filter(bin==sind) %>%
+  mutate(err1 = (pred1-eta_i)^2,
+         err2 = (pred2-eta_i)^2,
+         err3 = (pred3-eta_i)^2,
+         err4 = (pred4-eta_i)^2) %>%
+  select(id, bin, starts_with("err"))
+
+err1$window = cut(err1$bin, breaks = seq(0, 1000, by = 200), include.lowest = T)
+err1 <- split(err1, f = err1$window)
+
+tb1 <- lapply(err1, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+err2 <- test_df %>%
+  left_join(bind_rows(pred_list2), by = c("id", "bin")) %>%
+  filter(bin==sind) %>%
+  mutate(err1 = (pred1-eta_i)^2,
+         err2 = (pred2-eta_i)^2,
+         err3 = (pred3-eta_i)^2,
+         err4 = (pred4-eta_i)^2) %>%
+  select(id, bin, starts_with("err"))
+
+err2$window = cut(err2$bin, breaks = seq(0, 1000, by = 200), include.lowest = T)
+err2 <- split(err2, f = err2$window)
+
+tb2 <- lapply(err2, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id="Window")
+
+```
+
+```{r}
+# calcualte ISE
+options(knitr.kable.NA = "")
+
+ise <- full_join(tb1, tb2, by = "Window")
+colnames(ise) <-c("Window",  rep(seq(200, 800, by =200), 2))
+ise %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "FPCA eigenvalues" = 4, "Debiased eigenvalues" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+```
+
+- Looks like the debiased eigenvalues helped on most cases except the very last two, when trying to predict (800, 1000].
+
+## AUC
+
+```{r}
+# calcualte AUC
+auc1 <- test_df %>%
+  left_join(bind_rows(pred_list), by = c("id", "bin")) %>%
+  filter(bin==sind) %>%
+  mutate(window = cut(bin, breaks = seq(0, 1000, by = 200), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
+
+auc1 <- split(auc1, f = auc1$window)
+
+tb_auc1 <- matrix(NA, 5, 4)
+
+for(i in 2:4=5){
+
+  this_lst <- auc1[i:5]
+
+  for(j in seq_along(this_lst)){
+     tb_auc[ ,i]
+
+  }
+
+  tb_auc1[(i+1):5, i]
+}
+
+```
+
+```{r}
+# calcualte ISE
+err2 <- test_df %>%
+  left_join(bind_rows(pred_list2), by = c("id", "bin")) %>%
+  filter(bin==sind) %>%
+  mutate(err1 = (pred1-eta_i)^2,
+         err2 = (pred2-eta_i)^2,
+         err3 = (pred3-eta_i)^2,
+         err4 = (pred4-eta_i)^2) %>%
+  select(id, bin, starts_with("err"))
+
+err2$window = cut(err2$bin, breaks = seq(0, 1000, by = 200), include.lowest = T)
+err2 <- split(err2, f = err2$window)
+
+tb2 <- lapply(err2, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id="Window")

### 2023-10-04 Code/OutSampBayes.R

@@ -65,7 +65,7 @@ Model <- function(parm, Data){
 
 
 # data: 
-out_pred_laplace <- function(fpca_fit, df_new, kpc){
+out_pred_laplace <- function(mu, evalues, phi_mat, df_new, kpc){
   
   # put data into correct format
   ns <- as.vector(table(df_new$bin)) # number of observations
@@ -75,12 +75,12 @@ out_pred_laplace <- function(fpca_fit, df_new, kpc){
   df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)  
   
   # into a list
-  tao <- diag(fpca_fit$evalues[1:kpc])
-  f0 <- fpca_fit$mu[1:max_bin]
+  tao <- diag(evalues)
+  f0 <- mu[1:max_bin]
   N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
   n <- df_new2$ns # number of experiements at each bin
   y <- df_new2$hs # outcome, number of 1
-  X <- fpca_fit$efunctions[1:max_bin, 1:kpc] # "covariates", in this case PC functions
+  X <- phi_mat[1:max_bin, ] # "covariates", in this case PC functions
   J <- kpc # number of parameters/scores
   mon.names <- "LP"
   
@@ -103,7 +103,7 @@ out_pred_laplace <- function(fpca_fit, df_new, kpc){
   score <- Fit$Summary1[, "Mode"]
   
   # prediction
-  eta_pred_out <- fpca_fit$mu+fpca_fit$efunctions[, 1:kpc]%*%score
+  eta_pred_out <- mu+phi_mat%*%score
   
   return(list(eta_pred = eta_pred_out,
                score_out = score))

### 2023-10-06 Docs/ProgressReportSim.Rmd
calculated ISE and AUC

@@
 # calcualte AUC
-auc1 <- test_df %>%
+# FPCA eigenvalues
+auc_df1 <- test_df %>%
   left_join(bind_rows(pred_list), by = c("id", "bin")) %>%
@@
 
-auc1 <- split(auc1, f = auc1$window)
+obs_track <- seq(200, 800, by = 200)
+pred_window <- unique(auc_df1$window)
+auc_mat <- matrix(NA, length(pred_window), length(obs_track))
 
-tb_auc1 <- matrix(NA, 5, 4)
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df1$Y[auc_df1$window==pred_window[i]]
+    this_pred <- auc_df1[auc_df1$window==pred_window[i],
+                         paste0("pred", j)]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat[i ,j] <- this_perf@y.values[[1]]
+  }
+}
+```
 
-for(i in 2:4=5){
+```{r}
+# calcualte AUC
+# FPCA eigenvalues
+auc_df2 <- test_df %>%
+  left_join(bind_rows(pred_list2), by = c("id", "bin")) %>%
+  filter(bin==sind) %>%
+  mutate(window = cut(bin, breaks = seq(0, 1000, by = 200), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
 
-  this_lst <- auc1[i:5]
+auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))
 
-  for(j in seq_along(this_lst)){
-     tb_auc[ ,i]
-
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
+    this_pred <- auc_df2[auc_df2$window==pred_window[i],
+                         paste0("pred", j)]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat2[i ,j] <- this_perf@y.values[[1]]
   }
-
-  tb_auc1[(i+1):5, i]
 }
-
 ```
@@
 # calcualte ISE
-err2 <- test_df %>%
-  left_join(bind_rows(pred_list2), by = c("id", "bin")) %>%
-  filter(bin==sind) %>%
-  mutate(err1 = (pred1-eta_i)^2,
-         err2 = (pred2-eta_i)^2,
-         err3 = (pred3-eta_i)^2,
-         err4 = (pred4-eta_i)^2) %>%
-  select(id, bin, starts_with("err"))
+options(knitr.kable.NA = "")
 
-err2$window = cut(err2$bin, breaks = seq(0, 1000, by = 200), include.lowest = T)
-err2 <- split(err2, f = err2$window)
-
-tb2 <- lapply(err2, function(x){
-  x %>%  group_by(id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>%
-    summarise_at(vars(err1, err2, err3, err4), mean)
-}) %>% bind_rows(.id="Window")
+auc <- cbind(auc_mat, auc_mat2)
+rownames(auc) <- unique(auc_df1$window)
+colnames(auc) <-c(rep(seq(200, 800, by =200), 2))
+auc %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "FPCA eigenvalues" = 4, "Debiased eigenvalues" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+```

### 2023-10-06 Code/Sim_GLMMadaptive.R

@@ -1,0 +1,75 @@
+
+#### GLMMadaptive ####
+
+pred_lst_ref <- list()
+runtime_ref <- rep(NA, M)
+
+
+
+pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
+
+for(m in 1:M){
+  df <- sim_data[[m]]
+  
+  t1 <- Sys.time()
+  # GLMM adaptive model
+  adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df, family = binomial())
+  
+  # container
+  df_est_latent <- df
+  df_est_latent[, 'pred_t200'] <- df_est_latent[, 'pred_t400'] <- df_est_latent[, 'pred_t600'] <- df_est_latent[, 'pred_t800'] <- NA 
+  
+  # time up to t
+  pred_t200 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 200),
+                       newdata2 =  df %>% filter(sind_inx > 200), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
+  df_est_latent$pred_t200[df_est_latent$sind_inx>200] <- pred_t200$newdata2$pred
+  
+  pred_t400 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 400),
+                       newdata2 =  df %>% filter(sind_inx > 400), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
+  df_est_latent$pred_t400[df_est_latent$sind_inx>400] <- pred_t400$newdata2$pred
+  
+  pred_t600 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 600),
+                       newdata2 =  df %>% filter(sind_inx > 600), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
+  df_est_latent$pred_t600[df_est_latent$sind_inx>600] <- pred_t600$newdata2$pred
+  
+  pred_t800 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 800),
+                       newdata2 =  df %>% filter(sind_inx > 800), type = "subject_specific", type_pred = "link",
+                       se.fit = TRUE, return_newdata = TRUE)
[…]

### 2023-10-06 Code/Sim_fGFPCA.R

@@ -1,0 +1,200 @@
+# This script implements the full simulation 
+# for dynamic prediction of fGFPCA
+
+#### set up ####
+
+set.seed(516)
+
+library(here)
+library(tidyverse)
+theme_set(theme_minimal())
+
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+library(splines)
+
+
+
+#### load simulated data ####
+
+load(here("Data/sim_data.RData"))
+
+# data structure
+M <- length(sim_data) # number of simulation
+N <- length(unique(sim_data[[1]]$id)) # sample size 
+J <- length(unique(sim_data[[1]]$sind)) # number of measurements for each subject
+
+# data preview
+rand_id <- sample(as.numeric(unique(sim_data[[1]]$id)), size=4)
+
+sim_data[[1]] %>% 
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=exp(eta_i)/(1+exp(eta_i))))+
+  facet_wrap(~id)
+
+
+##### fGFPCA Model estimation #####
+
+source(here("Code/GLMM-FPCA.R")) 
+source(here("Code/OutSampBayes.R"))
+# use pred_latent function to estimate latent function 
+
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+
+# result container
+pred_list_all <- list()
+# score <- array()
+# score_lst <- list()
+fit_time <- pred_time <- rep(NA, M)
+
+K <- 4 # number of eigenfunctions to use
+
+pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
+
+# The whole process
+for(m in 1:M){
+  
+  # for every simulated dataset
+  df <- sim_data[[m]]
+  
+  ## Step 1: bin observations
+  df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
+  df$bin <- as.numeric(as.character(df$bin))
+  
+  # split into training and test set
+  # 80% subjects for training, 20% for testing
+  train_id <- sample(unique(df$id), size = N*0.8)
+  test_id <- setdiff(unique(df$id), train_id)
[…]

### 2023-10-06 Code/Simulation.R

@@ -1,220 +1,0 @@
-# This script implements simulation for WNAR manuscript
-
-set.seed(516)
-
-library(here)
-library(tidyverse)
-theme_set(theme_minimal())
-# library(fcr)
-library(refund)
-library(lme4)
-library(ROCR)
-library(GLMMadaptive)
-library(mgcv)
-library(splines)
-# library(ggpubr)
-# library(gridExtra)
-# library(kableExtra)
-
-
-
-
-#### load simulated data ####
-
-load(here("Data/sim_data.RData"))
-
-# data structure
-M <- length(sim_data) # number of simulation
-N <- length(unique(sim_data[[1]]$id))
-J <- length(unique(sim_data[[1]]$sind))
-
-# data preview
-rand_id <- sample(as.numeric(unique(sim_data[[1]]$id)), size=4)
-
-sim_data[[1]] %>% 
-  filter(id %in% rand_id) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  geom_line(aes(x=sind, y=exp(eta_i)/(1+exp(eta_i))))+
-  facet_wrap(~id)
-
-
-##### fGFPCA Model estimation #####
-
-source(here("Code/GLMM-FPCA.R")) # use pred_latent function to estimate latent function 
-# source(here("Code/OutSampMLE.R"))
-
-
-# bin data
-bin_w <- 10 # bin width
-n_bin <- J/bin_w # number of bins
-brks <- seq(0, J, by = bin_w) # cutoff points
-mid <- (brks+bin_w/2)[1:n_bin] # mid points
-
-
-# result container
-pred_lst <- list()
-score_lst <- list()
-runtime <- rep(NA, M)
-
-m <- 1
-
-pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
-
-
-for(m in 1:M){
-  
-  df <- sim_data[[m]]
-  
-  t1 <- Sys.time()
-  # bin observations
-  df$bin <- cut(df$sind_inx, breaks = brks, include.lowest = T, labels = mid)
-  df$bin <- as.numeric(as.character(df$bin))
-  
-  df_bin_lst <- split(df, f = df$bin)
-  
-  # fit local GLMM and estimate latent function
-  df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
[…]

### 2023-10-10 Docs/ProgressReportSim.Rmd
re-knit report
fix projection steps with FPCA parameters

--- 2023-10-03 Docs/ProgressReportSim.Rmd
@@
 - $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively. 
+- In fact, since there is a constant factor $\sqrt{2}$ before each eigenfunction so that they are orthogonal
 
@@
 lambda = 0.5^(0:(K-1)) 
+```
 
-# generated data 
+```{r}
+# generated training data 
 ## score
@@
 #            this is not necessary, but may be of convenience when you implement the method
-df <- data.frame(id = factor(rep(1:N, each=J)),
+train_df <- data.frame(id = factor(rep(1:N, each=J)),
                  t = rep(t, N), 
@@
                  sind = rep(1:J, N))
+
 # visualization
 rand_id <- sample(N, size = 4)
-df %>% filter(id %in% rand_id) %>% 
+train_df %>% filter(id %in% rand_id) %>% 
   ggplot()+
@@
 # fGFPCA model fitting
+
+- Training sample: N=500
 
@@
 
-df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
-df$bin <- as.numeric(as.character(df$bin))
+train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
+train_df$bin <- as.numeric(as.character(train_df$bin))
 # unique(df$bin)
 
-df %>% 
+train_df %>% 
   filter(id %in% rand_id) %>%
@@
 
-## Data split
-
-- 400 subjects for training, 100 for out-of-sample prediction
-
-```{r}
-train_id <- sample(unique(df$id), size = N*0.8)
-test_id <- setdiff(unique(df$id), train_id)
-
-train_df <- df %>% filter(id %in% train_id)
-test_df <- df %>% filter(id %in% test_id)
-```
-
 ## Local GLMM
@@
 - Used glmer with PIRLS (nAGQ=0) for local GLMMs step
-- Fit on the training set
+- Fit on the training set (N=500)
+- Use original time index ([0, 1]) as independent variable
 
@@
 train_bin_lst <- split(train_df, f = train_df$bin)
+# length(train_bin_lst)
+# lapply(train_bin_lst, dim)
 
@@
 t2= Sys.time()
-t_local_glmm <- t2-t1 
+t_local_glmm <- t2-t1 # less than 5 seconds
 ```
@@
 # example estimated latent function
+train_id <- unique(train_df$id)
 rand_id <- sample(train_id, 4)
@@
   ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
-  geom_line(aes(x=sind, y=eta_i, group = id, col = "true"))+
-  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
+  geom_line(aes(x=t, y=eta_hat, group = id, col = "estimated"))+
+  geom_line(aes(x=t, y=eta_i, group = id, col = "true"))+
+  geom_point(aes(x=t, y = Y, group = id), size = 0.5)+
   facet_wrap(~id, scales = "free")+
@@
 uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+mid_t <- df_est_latent$t[df_est_latent$bin==df_est_latent$sind]
+mid_t <- unique(mid_t) # the time points correspoinding to bin midpoints
 
@@
 t1 <- Sys.time()
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T)
 t2 <- Sys.time()
-t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
+t_fpca <- t2-t1 # less than one second
 ```
 
+- Estimated mean
+
 ```{r}
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
+plot(mid_t, fpca_mod$mu, type = "l", xlab = "bin", ylab = "Mean")
 ```
+
+- Estimated eigenfunctions
 
@@
 
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+plot(mid_t, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
+plot(mid_t, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid_t, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid_t, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
 ```
+
+- Correlation matrix
 
@@
 
+<!-- https://bookdown.org/ssjackson300/Machine-Learning-Lecture-Notes/splines.html -->
+
 - Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly
-- Not sure how to decide the knots. I am using the bin boundaries for now (0, 10, 20, ...,1000), which is 101 knots.
+- I would like to adjust use time points on the [0, 1] scale as evaluation points
+- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, p = 3)
+- Do the same kind of projection to mean functions as well
 
@@
 
-# number of knots
-# not sure how to determine
-# maybe for now use the bin endpoints? 
-# knot_pos <- seq(0, J, by = 50)
-knot_pos <- brks
+# knots
+knots <- 35
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
 
 # evaluate B-splines on binned grid
-B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
+B <- spline.des(knots = knots_values, x = mid_t, ord = p + 1,
                             outer.ok = TRUE)$design
 # evaluate B-splines on original grid
-Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
+Bnew <- spline.des(knots = knots_values, x = unique(train_df$t), ord = p + 1,
                   outer.ok = TRUE)$design
@@
 }
+
+# project mean functions 
+lm_mod <- lm(fpca_mod$mu ~ B-1)
+mu_new <- Bnew %*% coef(lm_mod)
 ```
@@
 
-plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
-lines(1:J, efunctions_new[, 1], col="blue")
+plot(mid_t, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
+lines(t, efunctions_new[, 1], col="blue")
 
-plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
-lines(1:J, efunctions_new[, 2], col="blue")
+plot(mid_t, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
+lines(t, efunctions_new[, 2], col="blue")
 
-plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
-lines(1:J, efunctions_new[, 3], col="blue")
+plot(mid_t, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
+lines(t, efunctions_new[, 3], col="blue")
 
-plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
-lines(1:J, efunctions_new[, 4], col="blue")
-
+plot(mid_t, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
+lines(t, efunctions_new[, 4], col="blue")
 ```
 
-- The projected eigenfunctions seem to be more wiggly than we would like?
-- What is the reason for choosing projection over linear/spline interpolation?
-
-#### Number of knots
-
-- Try fewer knots?
-
 ```{r}
-knot_pos <- seq(0, J, by = 100)
-
-# evaluate B-splines on binned grid
-B <- spline.des(knots = knot_pos, x = mid, ord = p + 1,
-                            outer.ok = TRUE)$design
-# evaluate B-splines on original grid
-Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = p + 1,
-                  outer.ok = TRUE)$design
-
-# project binned eigenfunctions onto the original grid
-efunctions_new_11knots <- matrix(NA, J, K)
-for(k in 1:K){
-    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
-    efunctions_new_11knots[,k] <- Bnew %*% coef(lm_mod)
-}
+plot(mid_t, fpca_mod$mu, xlab="bin", ylab="Mean", pch=20, cex = 0.5)
+lines(t, mu_new, col="blue")
 ```
 
-```{r, fig.height=8, fig.width=8}
-# visualization
-par(mfrow=c(2,2))
+- The projection seems to work very well!
 
-plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_11knots[, 1], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_11knots[, 2], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_11knots[, 3], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_11knots[, 4], col="blue")
-
-```
-
-#### Order of spline
-
-- Try lower order? (order = 3)
-
-```{r}
-knot_pos <- brks
-
-# evaluate B-splines on binned grid
-B <- spline.des(knots = knot_pos, x = mid, ord = 3,
-                            outer.ok = TRUE)$design
-# evaluate B-splines on original grid
-Bnew <- spline.des(knots = knot_pos, x = 1:J, ord = 3,
-                  outer.ok = TRUE)$design
-
-# project binned eigenfunctions onto the original grid
-efunctions_new_2p <- matrix(NA, J, K)
-for(k in 1:K){
-    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
-    efunctions_new_2p[,k] <- Bnew %*% coef(lm_mod)
-}
-```
-
-```{r, fig.height=8, fig.width=8}
-# visualization
-par(mfrow=c(2,2))
-
-plot(mid, fpca_mod$efunctions[, 1], xlab="bin", ylab="PC1", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_2p[, 1], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 2], xlab="bin", ylab="PC2", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_2p[, 2], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 3], xlab="bin", ylab="PC3", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_2p[, 3], col="blue")
-
-plot(mid, fpca_mod$efunctions[, 4], xlab="bin", ylab="PC4", pch=20, cex = 0.5)
-lines(1:J, efunctions_new_2p[, 4], col="blue")
-
-```
-
-### Debias with GLMM model
+### Debias with eigenvalues
 
@@
 - What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
-- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
-- I will try to fit the model with projected eigenfunctions as above (with 101 knots). I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
+- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction.
+And out-of-sample scores are not estimated by FPCA but Laplace approximation.
+- I will use the projected eigenfunctions from above. 
+- I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation
 
 ```{r}
-df_phi <- data.frame(sind = 1:J, efunctions_new)
-colnames(df_phi) <- c("sind", paste0("phi", 1:4))
+# dim(fpca_mod$efunctions)
+df_phi <- data.frame(t = t, efunctions_new)
+colnames(df_phi) <- c("t", paste0("phi", 1:4))
 
 # train_df$bin <- as.numeric(as.character(train_df$bin))
-train_df <- train_df %>% left_join(df_phi, by = "sind")
+train_df <- train_df %>% left_join(df_phi, by = "t")
 train_df$id <- as.factor(train_df$id)
 ```
-
-- Here, what if I replace "bin" with "t"? Should I regress of Y on t or sind?
 
@@
 t2 <- Sys.time()
-t_debias <- t2-t1
+t_debias <- t2-t1 # less than 45 seconds
 ```
 
-- And now I try to extract the variance of scores (debiased eigenvalues). 
-- Use the conclusion that connects smoothing parameters to variance
+And now I try to extract the variance of scores (debiased eigenvalues), using the conclusion that connects smoothing parameters to variance
 
@@
 new_lambda
-# true eigenvalues
-lambda
+# new_lambda/lambda
+# new_lambda/fpca_mod$evalues[1:4]
 ```
 
-- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.
+This eigenvalues, as well as the eigenfunctions, are actually not on the same scale of true values.
+By looking into the fpca.face source code, eigenfunctions are scaled by the square root of the number of total observations points.
+Since our model was fitted on the binnd grid, the actual eigenvaleus and eigenfunctions would be:
+
+\begin{aligned}
+\phi_k(t) & = \sqrt{n_{bin}}\hat{\phi}_k(t)  \\
+\lambda_k & = \hat{\lambda}_k/n_{bin}
+
+\end{aligned}
+
+Let's rescale these quatities as below
+
+- Re-scaled eigenvalues
+```{r}
+new_lambda/n_bin
+```
+
+- Re-scaled eigenfunctions
+
+```{r, fig.height=8, fig.width=8}
+par(mfrow=c(2,2))
+
+plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
+plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
+plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
+plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+
+```
+
+- Both much closer to their true values
+
+# Out-of-sample prediction with Laplace approximation
+
+I will now estimate individual score of test sample using Laplace Approximation, and calculated the predicted track.
+
+- Generate additional 200 subjects for out-of-sample performance evaluation
+- Maximum observations time: 0.2, 0.4, 0.6, 0.8
+- Prediction window: 0.2-0.4, 0.4-0.6, 0.6-0.8, 0.8-1.0
+
+## Generate test data
+
+```{r}
+# generated testing data 
+N_test <- 200
+b_test <- matrix(rnorm(N_test*K), N_test, K) %*% diag(sqrt(lambda)) %*% t(phi) # of size N by J
+# dim(b_test)
+
+# latent gaussian function
+eta_test <- t(vapply(1:N_test, function(x){
+    f_0(x) + b_i[x,]
+  }, numeric(J)))
+
+# outcome binary function
+Y_test <- matrix(rbinom(N_test*J, size=1, prob=plogis(eta_test)), 
+                N_test, J, byrow=FALSE)
+
+# format into dataframe
+test_df <- data.frame(id = factor(rep(1:N_test, each=J)),
+                 t = rep(t, N_test), 
+                 Y = as.vector(t(Y_test)),
+                 eta_i = as.vector(t(eta_test)),
+                 sind = rep(1:J, N_test))
+# bin
+test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
+test_df$bin <- as.numeric(as.character(test_df$bin))
+test_df$id <- as.factor(test_df$id)
+test_id <- unique(test_df$id)
+```
+
+## Laplace approximation
+
+```{r}
+# the model
+## now, with expanded mean and eigenfuncitons, a Bernoulli distribution at each time point
+Model <- function(parm, Data){
+  xi <- parm[Data$pos.xi]
+
+  # log-prior
+  xi.prior <- dmvnorm(xi, mean = rep(0, Data$K), sigma=Data$tao, log = TRUE)
+
+  # log-posterior likelihood
+  eta <- Data$f0+Data$X %*% xi
+  p <- exp(eta)/(1+exp(eta))
+  LL <- sum(dbern(x=Data$y, prob=p, log = TRUE)) # log likelihood of Y|xi
+  LP <- LL+sum(xi.prior) # joint log likelihood of (Y, xi)
+
+  # output
+  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LL,
+                   yhat=Data$y, parm=parm)
+  return(Modelout)
+}
+```
+
+```{r}
+# parameter names
+name_lst <- as.list(rep(0, K))
+names(name_lst) <- paste("xi", 1:K, sep = "")
+parm.names <- as.parm.names(name_lst)
+pos.xi <- grep("xi", parm.names)
+PGF <- function(Data) {
+  xi <- rnorm(Data$K)
+  return(xi)
+}
+mon.names <- "LP"
+```
+
+```{r, results='hide'}
+# approximation process
+# container
+pred_list <- list()
+# score <- array(NA, dim=c(length(test_id), K, 4))
+# dimensions: subject, eigenfunction, maximum observation time
+
+# prediction for a single subject
+t1 <- Sys.time()
+# per subject
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+
+  # per max obs time
+  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
+    df_it <- df_i %>% filter(t <= tmax)
+    max_rid <- nrow(df_it)
+
+    # into a list
+    MyData <- list(K=K, PGF=PGF, 
+                   X=efunctions_new[1: max_rid, ], 
+                   mon.names=mon.names,
+                   parm.names=parm.names, 
+                   pos.xi=pos.xi, 
+                   y=df_it$Y, 
+                   tao=diag(new_lambda), f0=mu_new[1:max_rid, ])
+
+    # fit laplace approximation
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
+    score <- Fit$Summary1[, "Mode"]
+
+    # prediction
+    eta_pred_out <- mu_new+efunctions_new%*%score
+    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
+  }
+
+  # df_i$pred0.2[df_i$t<=0.2] <- NA
+  # df_i$pred0.4[df_i$t<=0.4] <- NA
+  # df_i$pred0.6[df_i$t<=0.6] <- NA
+  # df_i$pred0.8[df_i$t<=0.8] <- NA
+
+  pred_list[[i]] <- df_i
+}
+t2 <- Sys.time()
+t_pred <- t2-t1 # About 3.5 minutes
+```
+
+```{r}
+df_pred <- bind_rows(pred_list)
+
+df_pred$pred0.2[df_pred$t<=0.2] <- NA
+df_pred$pred0.4[df_pred$t<=0.4] <- NA
+df_pred$pred0.6[df_pred$t<=0.6] <- NA
+df_pred$pred0.8[df_pred$t<=0.8] <- NA
+
+rand_test_id <- sample(test_id, 4)
+
+p1<-df_pred %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "Unscaled, debiased eigenvalues")
+
+```
+
+### Compared to biased eigenvalues
+
+```{r, results='hide'}
+## With eigenvalues from FPCA
+# container
+pred_list2 <- list()
+
+# per subject
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+
+  # per max obs time
+  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
+    df_it <- df_i %>% filter(t <= tmax)
+    max_rid <- nrow(df_it)
+
+    # into a list
+    MyData <- list(K=K, PGF=PGF, 
+                   X=efunctions_new[1: max_rid, ], 
+                   mon.names=mon.names,
+                   parm.names=parm.names, 
+                   pos.xi=pos.xi, 
+                   y=df_it$Y, 
+                   tao=diag(fpca_mod$evalues[1:K]), f0=mu_new[1:max_rid, ])
+
+    # fit laplace approximation
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
+    score <- Fit$Summary1[, "Mode"]
+
+    # prediction
+    eta_pred_out <- mu_new+efunctions_new%*%score
+    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
+  }
+
+  # df_i$pred0.2[df_i$t<=0.2] <- NA
+  # df_i$pred0.4[df_i$t<=0.4] <- NA
+  # df_i$pred0.6[df_i$t<=0.6] <- NA
+  # df_i$pred0.8[df_i$t<=0.8] <- NA
+
+  pred_list2[[i]] <- df_i
+}
+```
+
+```{r}
+df_pred2 <- bind_rows(pred_list2)
+
+df_pred2$pred0.2[df_pred2$t<=0.2] <- NA
+df_pred2$pred0.4[df_pred2$t<=0.4] <- NA
+df_pred2$pred0.6[df_pred2$t<=0.6] <- NA
+df_pred2$pred0.8[df_pred2$t<=0.8] <- NA
+# overview of predicted track
+p2<-df_pred2 %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "Unscaled, biased eigenvalues")
+
+```
+
+### Scaled eigenfunctions and eigenvalues
+
+- No need to scale mean functions
+
+```{r, results='hide'}
+## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
+# container
+pred_list3 <- list()
+efunctions_scaled <- efunctions_new*sqrt(n_bin)
+evalues_scaled <- new_lambda/n_bin
+
+# per subject
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+
+  # per max obs time
+  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
+    df_it <- df_i %>% filter(t <= tmax)
+    max_rid <- nrow(df_it)
+
+    # into a list
+    MyData <- list(K=K, PGF=PGF, 
+                   X=efunctions_scaled[1: max_rid, ], 
+                   mon.names=mon.names,
+                   parm.names=parm.names, 
+                   pos.xi=pos.xi, 
+                   y=df_it$Y, 
+                   tao=diag(evalues_scaled), f0=mu_new[1:max_rid, ])
+
+    # fit laplace approximation
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
+    score <- Fit$Summary1[, "Mode"]
+
+    # prediction
+    eta_pred_out <- mu_new+efunctions_scaled%*%score
+    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
+  }
+
+  # df_i$pred0.2[df_i$t<=0.2] <- NA
+  # df_i$pred0.4[df_i$t<=0.4] <- NA
+  # df_i$pred0.6[df_i$t<=0.6] <- NA
+  # df_i$pred0.8[df_i$t<=0.8] <- NA
+
+  pred_list3[[i]] <- df_i
+}
+```
+
+```{r}
+df_pred3 <- bind_rows(pred_list3)
+
+df_pred3$pred0.2[df_pred3$t<=0.2] <- NA
+df_pred3$pred0.4[df_pred3$t<=0.4] <- NA
+df_pred3$pred0.6[df_pred3$t<=0.6] <- NA
+df_pred3$pred0.8[df_pred3$t<=0.8] <- NA
+# overview of predicted track
+p3 <- df_pred3 %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "Scaled, unbiased eigenvalues")
+
+```
+
+```{r}
+t_1sim <- (t_local_glmm+t_fpca+t_debias+t_pred)
+```
+
+- Time spent on one iteration: `r round(t_1sim, 2)` seconds
+- Did debiased eigenvalues improve performance? The lines above look like so. I also wanna see the squared errors and AUC.
+- Scaling eigenfunctions and values will change a prediction a little bit
+
+```{r, fig.height=5, fig.width=15}
+ggarrange(p1, p2, p3, nrow = 1, common.legend = T)
+```
+
+## ISE
+
+```{r}
+# calcualte ISE
+err1 <- df_pred %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err1$window = cut(err1$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err1 <- split(err1, f = err1$window)
+
+tb1 <- lapply(err1, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+err2 <- df_pred2 %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err2 <- split(err2, f = err2$window)
+
+tb2 <- lapply(err2, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+err3 <- df_pred3 %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err3$window = cut(err3$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err3 <- split(err3, f = err3$window)
+
+tb3 <- lapply(err3, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+options(knitr.kable.NA = "")
+
+ise <- full_join(tb1, tb2, by = "Window") %>% full_join(tb3, by = "Window")
+colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 3))
+ise %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 12))
+```
+
+- Looks like the debiased eigenvalues helped.
+- Scaling, on the other hand, does seem to help in general. 
+- But shouldn't scaling improve performance? 
+- Also, it seems like the longer the observed track, the closer the three cases are.
+
+## AUC
+
+```{r}
+# calcualte AUC
+auc_df1 <- df_pred %>%
+  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
+
+obs_track <- seq(0.2, 0.8, by = 0.2)
+pred_window <- unique(auc_df1$window)
+auc_mat <- matrix(NA, length(pred_window), length(obs_track))
+
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df1$Y[auc_df1$window==pred_window[i]]
+    this_pred <- auc_df1[auc_df1$window==pred_window[i],
+                         paste0("pred", obs_track[j])]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat[i ,j] <- this_perf@y.values[[1]]
+  }
+}
+
+```
+
+```{r}
+# calcualte AUC
+auc_df2 <- df_pred2 %>%
+  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
+
+auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))
+
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
+    this_pred <- auc_df2[auc_df2$window==pred_window[i],
+                         paste0("pred", obs_track[j])]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat2[i ,j] <- this_perf@y.values[[1]]
+  }
+}
+
+```
+
+```{r}
+# calcualte AUC
+auc_df3 <- df_pred3 %>%
+  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
+
+auc_mat3 <- matrix(NA, length(pred_window), length(obs_track))
+
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df3$Y[auc_df3$window==pred_window[i]]
+    this_pred <- auc_df3[auc_df3$window==pred_window[i],
+                         paste0("pred", obs_track[j])]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat3[i ,j] <- this_perf@y.values[[1]]
+  }
+}
+
+```
+
+```{r}
+# calcualte ISE
+options(knitr.kable.NA = "")
+
+auc <- cbind(auc_mat, auc_mat2, auc_mat3)
+rownames(auc) <- unique(auc_df1$window)
+colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 3))
+auc %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 12))
+```

### 2023-10-12 Docs/ProgressReportSim.Rmd
update report

@@
 
-- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly
-- I would like to adjust use time points on the [0, 1] scale as evaluation points
-- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, p = 3)
+- Use fastGFPCA::reeval_efunctions. Cannot install the package so used the source code directly.
+- I would like to use time points on the [0, 1] scale as evaluation points when creating basis functions.
+- Also, take number of knots and order from the FPCA model in step 3 (knots = 35, cubic spline basis). 
 - Do the same kind of projection to mean functions as well
@@
 
-### Compared to biased eigenvalues
+<!-- ### Compared to biased eigenvalues -->
 
@@
 
-### Scaled eigenfunctions and eigenvalues
-
-- No need to scale mean functions
+<!-- ### Scaled eigenfunctions and eigenvalues -->
 
@@
 
-- Time spent on one iteration: `r round(t_1sim, 2)` seconds
-- Did debiased eigenvalues improve performance? The lines above look like so. I also wanna see the squared errors and AUC.
-- Scaling eigenfunctions and values will change a prediction a little bit
+I would like to see if the following operations can improve predictive performance:
+
+1. Debiasing eigenvalues from FPCA
+2. Scale the eigenfunctions and debiased eigenvalues (from mgcv::bam) so that they are close to the true values for data generation
 
@@
 ```
+
+- Time spent on one iteration: `r round(t_1sim, 2)` seconds
+- Debiasing eigenvalues helped with prediction. Scaling eigenfunctiosn and eigenvalues, on the other hand, ended up being more complicated.
 
@@
 - Looks like the debiased eigenvalues helped.
-- Scaling, on the other hand, does seem to help in general. 
-- But shouldn't scaling improve performance? 
-- Also, it seems like the longer the observed track, the closer the three cases are.
+- Scaling, on the other hand, doesn't seem to help in general. But shouldn't scaling improve performance? 
+- Also, it seems like the longer the observed track, the less different the performance is.
 

### 2023-10-18 Docs/ProgressReportSim.Rmd
explore GLMMadaptive + spline basis functions

@@
 ```
+
+# Reference method
+
+## GLMMadaptive
+
+- Here we can fit a model with random intercept and slope for time
+
+$$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
+- But this model is obviously too simple for the data generation scheme
+
+```{r, class.source='fold-show'}
+# model fit
+t1 <- Sys.time()
+fit_adglmm <- mixed_model(Y ~ t, random = ~ t | id, data =  train_df, family = binomial())
+t2 <- Sys.time()
+# time 
+t2-t1
+# model
+summary(fit_adglmm)
+```
+
+## Flexible GLMM
+
+### BAM
+
+- The bam model below couldn't do out-of-sample prediction
+- It also take a long time to fit
+
+```{r,class.source='show', eval=FALSE}
+# try to fit a flexible GLMM
+t1 <- Sys.time()
+fit_glmm_sp <- bam(Y ~ s(t, bs="cr", k=10)+ti(id, t, bs=c("re", "cr"), k = c(5, 5), mc = c(TRUE, FALSE)), 
+                   family = binomial, data=train_df, 
+                   method = "fREML",
+                   discrete = TRUE)
+t2 <- Sys.time()
+```
+
+- I should construct basis functions and fit model with GLMMadaptvie::mixed_model instead
+
+### GLMMadaptive + basis functions
+
+- Fit on all 500 subjects take too long. I reduced the sample size to 10 
+- In fact, 10 samples took more than 30 minutes to fit the model.
+Therefore, I further reduce the number of observations for each subject.
+I'll start with only the 1/10 of the time track (100 observations), take observations at bin midpoints 5, 15, 25, ...., 995.
+This is to preserve the non-linearity of latent function when reducing grid density
+
+$$g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
+
+```{r, class.source = "fold-show"}
+# subset 10 subjects from training dataset 
+df_train_sub <- train_df %>% filter(id %in% 1:10 & sind==bin)
+df_train_sub$id <- droplevels(df_train_sub$id)
+# levels(df_train_sub$id)
+# intersect(t, unique(df_train_sub$t))
+```
+
+- Below I construct basis functions for the fixed effect/random intercept.
+Let's use five basis functions to start with.
+I believe this is far less flexible than the FPCA model, but it still takes a really long time to run.
+It is actually not realistic to repeat 500 times.
+
+PS 50 subjects took 1.06 hours to fit with the same model structure.
+
+```{r, class.source="fold-show"}
+# construct design matrix for the fixed effect intercept
+t_sub <- unique(df_train_sub$t)
+fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
+dim(fix_xmat[[1]]$X)
+
+# basis functions
+fix_xmat <- data.frame(t=t_sub, 
+                       fix_xmat[[1]]$X)
+colnames(fix_xmat) <- c("t", paste0("B", 1:5))
+
+df_train_sub <- df_train_sub %>% left_join(fix_xmat, by = "t")
+```
+
+- Below I construct basis functions for the individual random effect (also a functinal intercept), with four basis functions to start with.
+
+```{r, class.source = "fold-show"}
+# construct design matrix for the random effect
+rand_xmat <- smoothCon(s(t, bs="cr", k=4), data = data.frame(t=t_sub))
+
+dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
+rand_xmat <- data.frame(t = t, rand_xmat[[1]]$X)
+
+df_train_sub <- df_train_sub %>%
+  left_join(rand_xmat, by = c("t"))
+```
+
+- Now I'll fit the model and record the time spent on model fitting
+
+The model fitting took about 6 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.
+
+```{r, eval=TRUE, class.source = "fold-show"}
+# model fit
+t1 <- Sys.time()
+fit_adglmm <- mixed_model(fixed = Y ~ 0+B1+B2+B3+B4+B5, 
+                          random = ~X1+X2+X3+X4 | id,
+                          data =  df_train_sub, family = binomial())
+t2 <- Sys.time()
+t2-t1
+
+# model
+summary(fit_adglmm)
+```
+
+- But anyways, let's try to use this method for prediction. Let's do that for 10 new test subjects.
+
+```{r}
+df_test_sub <- test_df %>% filter(id %in% 1:10 & sind==bin)
+df_test_sub$id <- droplevels(df_test_sub$id)
+# levels(df_test_sub$id)
+
+# basis functions for both fixed and random effects should be shared across sample
+df_test_sub <- df_test_sub %>% left_join(fix_xmat, by = "t") %>%
+  left_join(rand_xmat, by = "t")
+
+# given observation from 0.2, 0.4, 0.6, 0.8, predict the rest
+# this step in fact is very fast
+pred_adglmm1 <- predict(fit_adglmm, 
+                      newdata = df_test_sub %>% filter(t<=0.2),
+                      newdata2 =  df_test_sub %>% filter(t>0.2),
+                      type = "subject_specific", type_pred = "link",
+                      return_newdata = TRUE)
+
+pred_adglmm2 <- predict(fit_adglmm, 
+                      newdata = df_test_sub %>% filter(t<=0.4),
+                      newdata2 =  df_test_sub %>% filter(t>0.4),
+                      type = "subject_specific", type_pred = "link",
+                      return_newdata = TRUE)
+
+pred_adglmm3 <- predict(fit_adglmm, 
+                      newdata = df_test_sub %>% filter(t<=0.6),
+                      newdata2 =  df_test_sub %>% filter(t>0.6),
+                      type = "subject_specific", type_pred = "link",
+                      return_newdata = TRUE)
+
+pred_adglmm4 <- predict(fit_adglmm, 
+                      newdata = df_test_sub %>% filter(t<=0.8),
+                      newdata2 =  df_test_sub %>% filter(t>0.8),
+                      type = "subject_specific", type_pred = "link",
+                      return_newdata = TRUE)
+```
+
+```{r}
+pred_adglmm_all <- df_test_sub %>% 
+  left_join(pred_adglmm1$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>% 
+  left_join(pred_adglmm2$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>%
+  left_join(pred_adglmm3$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>%
+  left_join(pred_adglmm4$newdata2 %>% select(id, t, pred), by = c("id", "t"))
+
+colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:15], "pred1", "pred2", "pred3", "pred4")
+```
+
+```{r}
+
+pred_adglmm_all %>%
+  filter(id %in% 1:4) %>%
+  mutate_at(vars(eta_i, pred1, pred2, pred3, pred4), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred1, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred2, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred3, col = "0.5"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred4, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GLMMadaptive + spline basis prediction")
+
+```

### 2023-10-23 Docs/ProgressReportSim.Rmd
example code for one iteration of simulation

@@
 - Also, take number of knots and order from the FPCA model in step 3 (knots = 35, cubic spline basis). 
-- Do the same kind of projection to mean functions as well
+- We don't project the mean function here. Instead, we will use the re-evaluated mean function from the next part (bam model used for debias)
 
@@
 # project mean functions 
-lm_mod <- lm(fpca_mod$mu ~ B-1)
-mu_new <- Bnew %*% coef(lm_mod)
+# lm_mod <- lm(fpca_mod$mu ~ B-1)
+# mu_new <- Bnew %*% coef(lm_mod)
 ```
@@
 
-```{r}
-plot(mid_t, fpca_mod$mu, xlab="bin", ylab="Mean", pch=20, cex = 0.5)
-lines(t, mu_new, col="blue")
-```
-
-- The projection seems to work very well!
-
 ### Debias with eigenvalues
@@
 - The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
-- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
+- What needs re-evaluation are eigenvalues (variance estimates) and the mean function. They will be used in Laplace Approximation for out-of-sample prediction. 
 - We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction.
@@
 
+We will need the mean function from this model, derived below:
+
+This is pretty much a linear functions. Is that supposed to happen?
+
+```{r, class.source = "fold-show"}
+new_mu <- predict(debias_glmm, type = "terms")[, 1]
+new_mu <- unique(new_mu)
+plot(unique(train_df$t), new_mu, type = "l")
+```
+
 And now I try to extract the variance of scores (debiased eigenvalues), using the conclusion that connects smoothing parameters to variance
@@
 
-\begin{aligned}
+\[\begin{aligned}
 \phi_k(t) & = \sqrt{n_{bin}}\hat{\phi}_k(t)  \\
 \lambda_k & = \hat{\lambda}_k/n_{bin}
-
-\end{aligned}
+\end{aligned}\]
 
@@
 - Re-scaled eigenvalues
+
 ```{r}
-new_lambda/n_bin
+scaled_lambda <- new_lambda/n_bin
 ```
@@
 ```{r, fig.height=8, fig.width=8}
+efunctions_scaled <- efunctions_new*sqrt(n_bin)
+
 par(mfrow=c(2,2))
 
-plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid_t, sqrt(n_bin)*fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
+plot(t, efunctions_scaled[, 1], type="l", xlab="bin", ylab="PC1")
+plot(t, efunctions_scaled[, 2], type="l", xlab="bin", ylab="PC2")
+plot(t, efunctions_scaled[, 3], type="l", xlab="bin", ylab="PC3")
+plot(t, efunctions_scaled[, 4], type="l", xlab="bin", ylab="PC4")
 
@@
 
-```{r, results='hide'}
+```{r pred_debias_unscale, results='hide'}
 # approximation process
@@
     MyData <- list(K=K, PGF=PGF, 
-                   X=efunctions_new[1: max_rid, ], 
+                   X=efunctions_new[1:max_rid, ], 
                    mon.names=mon.names,
@@
                    y=df_it$Y, 
-                   tao=diag(new_lambda), f0=mu_new[1:max_rid, ])
+                   tao=diag(new_lambda), f0=new_mu[1:max_rid])
 
@@
     # prediction
-    eta_pred_out <- mu_new+efunctions_new%*%score
+    eta_pred_out <- new_mu+efunctions_new%*%score
     df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
@@
 
-<!-- ### Compared to biased eigenvalues -->
-
-```{r, results='hide'}
-## With eigenvalues from FPCA
-# container
-pred_list2 <- list()
-
-# per subject
-for(i in seq_along(test_id)){
-  df_i <- test_df %>% filter(id==test_id[i])
-
-  # per max obs time
-  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
-    df_it <- df_i %>% filter(t <= tmax)
-    max_rid <- nrow(df_it)
-
-    # into a list
-    MyData <- list(K=K, PGF=PGF, 
-                   X=efunctions_new[1: max_rid, ], 
-                   mon.names=mon.names,
-                   parm.names=parm.names, 
-                   pos.xi=pos.xi, 
-                   y=df_it$Y, 
-                   tao=diag(fpca_mod$evalues[1:K]), f0=mu_new[1:max_rid, ])
-
-    # fit laplace approximation
-    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
-    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
-    score <- Fit$Summary1[, "Mode"]
-
-    # prediction
-    eta_pred_out <- mu_new+efunctions_new%*%score
-    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
-  }
-
-  # df_i$pred0.2[df_i$t<=0.2] <- NA
-  # df_i$pred0.4[df_i$t<=0.4] <- NA
-  # df_i$pred0.6[df_i$t<=0.6] <- NA
-  # df_i$pred0.8[df_i$t<=0.8] <- NA
-
-  pred_list2[[i]] <- df_i
-}
-```
-
-```{r}
-df_pred2 <- bind_rows(pred_list2)
-
-df_pred2$pred0.2[df_pred2$t<=0.2] <- NA
-df_pred2$pred0.4[df_pred2$t<=0.4] <- NA
-df_pred2$pred0.6[df_pred2$t<=0.6] <- NA
-df_pred2$pred0.8[df_pred2$t<=0.8] <- NA
-# overview of predicted track
-p2<-df_pred2 %>%
-  filter(id %in% rand_test_id) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
-  labs(title = "Unscaled, biased eigenvalues")
-
-```
-
 <!-- ### Scaled eigenfunctions and eigenvalues -->
 
-```{r, results='hide'}
+```{r pred_debias_scale, results='hide'}
 ## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
@@
 pred_list3 <- list()
-efunctions_scaled <- efunctions_new*sqrt(n_bin)
-evalues_scaled <- new_lambda/n_bin
 
@@
                    y=df_it$Y, 
-                   tao=diag(evalues_scaled), f0=mu_new[1:max_rid, ])
+                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])
 
@@
     Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
-    # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
     score <- Fit$Summary1[, "Mode"]
@@
     # prediction
-    eta_pred_out <- mu_new+efunctions_scaled%*%score
+    eta_pred_out <- new_mu+efunctions_scaled%*%score
     df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
@@
 
-```{r, fig.height=5, fig.width=15}
-ggarrange(p1, p2, p3, nrow = 1, common.legend = T)
+```{r, fig.height=5, fig.width=10}
+ggarrange(p1, p3, nrow = 1, common.legend = T)
 ```
@@
 tb1 <- lapply(err1, function(x){
-  x %>%  group_by(id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>%
-    summarise_at(vars(err1, err2, err3, err4), mean)
-}) %>% bind_rows(.id = "Window")
-```
-
-```{r}
-# calcualte ISE
-err2 <- df_pred2 %>%
-  mutate(err1 = (pred0.2-eta_i)^2,
-         err2 = (pred0.4-eta_i)^2,
-         err3 = (pred0.6-eta_i)^2,
-         err4 = (pred0.8-eta_i)^2) %>%
-  select(id, t, starts_with("err"))
-
-err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
-# table(err1$window)
-err2 <- split(err2, f = err2$window)
-
-tb2 <- lapply(err2, function(x){
   x %>%  group_by(id) %>% 
@@
 
-ise <- full_join(tb1, tb2, by = "Window") %>% full_join(tb3, by = "Window")
-colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 3))
+ise <- full_join(tb1, tb3, by = "Window")
+colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
 ise %>%
@@
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
-  add_header_above(c(" "=1, "Maximum observation time" = 12))
+  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
 ```
@@
 # calcualte AUC
-auc_df2 <- df_pred2 %>%
-  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
-  select(Y, starts_with("pred"), window)
-
-auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))
-
-for(j in seq_along(obs_track)){
-  for(i in (j+1):5){
-    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
-    this_pred <- auc_df2[auc_df2$window==pred_window[i],
-                         paste0("pred", obs_track[j])]
-    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
-    auc_mat2[i ,j] <- this_perf@y.values[[1]]
-  }
-}
-
-```
-
-```{r}
-# calcualte AUC
 auc_df3 <- df_pred3 %>%
@@
 ```{r}
-# calcualte ISE
+# calcualte AUC
 options(knitr.kable.NA = "")
 
-auc <- cbind(auc_mat, auc_mat2, auc_mat3)
+auc <- cbind(auc_mat, auc_mat3)
 rownames(auc) <- unique(auc_df1$window)
-colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 3))
+colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 2))
 auc %>%
@@
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "= 1, "Unscaled, debiased" = 4, "Unscaled, biased" = 4, "Scaled, debiased" = 4)) %>%
-  add_header_above(c(" "=1, "Maximum observation time" = 12))
+  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+```
+
+## Investigate a bit more on the difference in prediction
+
+- The scaled and unscaled parameters seem to make a difference in prediction, which they shouldn't
+
+```{r, fig.height=8, fig.width=8}
+par(mfrow=c(2, 2))
+plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5, 
+     xlab="unscaled", ylab="scaled", main = "Up to 0.2")
+plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
+     xlab="unscaled", ylab="scaled", main = "Up to 0.4")
+plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
+     xlab="unscaled", ylab="scaled", main = "Up to 0.6")
+plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
+     xlab="unscaled", ylab="scaled", main = "Up to 0.8")
 ```
@@
 ## Flexible GLMM
-
-### BAM
-
-- The bam model below couldn't do out-of-sample prediction
-- It also take a long time to fit
-
-```{r,class.source='show', eval=FALSE}
-# try to fit a flexible GLMM
-t1 <- Sys.time()
-fit_glmm_sp <- bam(Y ~ s(t, bs="cr", k=10)+ti(id, t, bs=c("re", "cr"), k = c(5, 5), mc = c(TRUE, FALSE)), 
-                   family = binomial, data=train_df, 
-                   method = "fREML",
-                   discrete = TRUE)
-t2 <- Sys.time()
-```
-
-- I should construct basis functions and fit model with GLMMadaptvie::mixed_model instead
-
-### GLMMadaptive + basis functions
 
@@
 The model fitting took about 6 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.
+
+Also tried increasing number of basis functions, which ended up with "computationally singular". In fact, this method does not seem very numerically stable.
 

### 2023-10-23 Docs/SimExample.Rmd
example code for one iteration of simulation

---
title: "Example with One simulated data"
author: "Ying Jin"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: no
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(516)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(knitr)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```

```{r}
# define a function to fit local GLMM
# df must have: outcome Y, subject id
pred_latent <- function(df, n_node){
    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ=n_node)
    eta_hat <- predict(this_glm, type = "link")
    df$eta_hat <- eta_hat
    return(df)
}
```

# Generate one dataset

A dataset with 500 subjects and 1000 measurements on a regular grid is generated as follows:

\[\begin{aligned}
Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}\]

where:

- $t$ is equal-spaced on $[0, 1]$
- $f_0(t)=0$
- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively.

```{r gen_data}
N <- 500 # sample size
J <- 1000 # number of observation points

t = seq(0,1,len=J) # observations points

# mean function
f_0 <- function(s) 0

#eigenfunctions 
K <- 4 # number of eigenfunctions
phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
                     sin(4*pi*t),cos(4*pi*t))

# eigenvalues
lambda = 0.5^(0:(K-1))

## score
xi <- matrix(rnorm(N*K),N,K)
xi <- xi %*% diag(sqrt(lambda))

# subject-specific random effect
b_i <- xi %*% t(phi); # of size N by J

# latent gaussian function
eta_i <- t(vapply(1:N, function(x){
    f_0(t) + b_i[x,]
  }, numeric(J)))

# outcome binary function
Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
                N, J, byrow=FALSE)

# format into dataframe
# id = subject identifier (factor variable)
# sind = numeric value corresponding to the observed functional domain
# Y = functional response (binary)
# sind_inx = numeric value associated with order of "sind"
#            this is not necessary, but may be of convenience when you implement the method
train_df <- data.frame(id = factor(rep(1:N, each=J)),
                 t = rep(t, N), 
                 Y = as.vector(t(Y_i)),
                 eta_i = as.vector(t(eta_i)),
                 sind = rep(1:J, N))
```

# Fit fGFPCA model

## Step 1: bin data

- Bin every 10 consecutive observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))
```

## Step 2: fit local GLMMs

```{r}
# fit local GLMMs model on the binned training set
train_bin_lst <- split(train_df, f = train_df$bin)
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent) 
```

## Step 3: fit FPCA model on estiamted latent function

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mid_t <- df_est_latent$t[df_est_latent$bin==df_est_latent$sind]
mid_t <- unique(mid_t) # the time points corresponding to bin midpoints

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=N, 
                         ncol=n_bin, byrow = F) # row index subject, column binned time
fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T)
```

## Step 4: projection and re-evaluation

According to the fGFPCA paper, in this step we do two things:

1. Project the eigenfunctions with B-spline basis back to the original grid
2. Re-evaluate the eigenvalues and mean function

### Projection

```{r}
# order of b splines
p <- 3

# knots
knots <- 35
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)

# evaluate B-splines on binned grid
B <- spline.des(knots = knots_values, x = mid_t, ord = p + 1,
                            outer.ok = TRUE)$design
# evaluate B-splines on original grid
Bnew <- spline.des(knots = knots_values, x = unique(train_df$t), ord = p + 1,
                  outer.ok = TRUE)$design

# project binned eigenfunctions onto the original grid
efunctions_new <- matrix(NA, J, K)
for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    efunctions_new[,k] <- Bnew %*% coef(lm_mod)
}
```

### Re-evaluation

```{r}
df_phi <- data.frame(t = t, efunctions_new)
colnames(df_phi) <- c("t", paste0("phi", 1:4))

train_df <- train_df %>% left_join(df_phi, by = "t")
train_df$id <- as.factor(train_df$id)

# bam model
debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
```

```{r}
# re-evaluated mean
# linear functions? 
new_mu <- predict(debias_glmm, type = "terms")[, 1]
new_mu <- unique(new_mu)
```

```{r}
# re-evaluate eigenvalues
new_lambda <- 1/debias_glmm$sp[2:5]
```

### Re-scale mean and eigenfunctions

We scale the re-evaluated eigenvalues and eigenfunctions using the number of bins, so that they are on the same scale with the true values from data generation mechanism.

```{r}
# rescale eigenvalues
scaled_lambda <- new_lambda/n_bin

# rescale eigenfunctions
efunctions_scaled <- efunctions_new*sqrt(n_bin)
```

# Out-of-sample prediction with Laplace approximation

## Generate test data

200 subjects with 1000 measurements on a regular grid

```{r}
# generated testing data 
N_test <- 200
b_test <- matrix(rnorm(N_test*K), N_test, K) %*% diag(sqrt(lambda)) %*% t(phi) # of size N by J

# latent gaussian function
eta_test <- t(vapply(1:N_test, function(x){
    f_0(x) + b_i[x,]
  }, numeric(J)))

# outcome binary function
Y_test <- matrix(rbinom(N_test*J, size=1, prob=plogis(eta_test)), 
                N_test, J, byrow=FALSE)

# format into dataframe
test_df <- data.frame(id = factor(rep(1:N_test, each=J)),
                 t = rep(t, N_test), 
                 Y = as.vector(t(Y_test)),
                 eta_i = as.vector(t(eta_test)),
                 sind = rep(1:J, N_test))
# bin
test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
test_df$bin <- as.numeric(as.character(test_df$bin))
test_df$id <- as.factor(test_df$id)
```

## Laplace approximation

```{r}
# the model
Model <- function(parm, Data){
  xi <- parm[Data$pos.xi]

  # log-prior
  xi.prior <- dmvnorm(xi, mean = rep(0, Data$K), sigma=Data$tao, log = TRUE)

  # log-posterior likelihood
  eta <- Data$f0+Data$X %*% xi
  p <- exp(eta)/(1+exp(eta))
  LL <- sum(dbern(x=Data$y, prob=p, log = TRUE)) # log likelihood of Y|xi
  LP <- LL+sum(xi.prior) # joint log likelihood of (Y, xi)

  # output
  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LL,
                   yhat=Data$y, parm=parm)
  return(Modelout)
}
```

```{r}
# parameter names
name_lst <- as.list(rep(0, K))
names(name_lst) <- paste("xi", 1:K, sep = "")
parm.names <- as.parm.names(name_lst)
pos.xi <- grep("xi", parm.names)
# PGF <- function(Data) {
#   xi <- rnorm(Data$K)
#   return(xi)
# }
mon.names <- "LP"
```

- Maximum observation track: 0.2, 0.4, 0.6, 0.8
- prediction window: 0.2-0.4, 0.4-0.6, 0.6-0.8, 0.8-1

## Prediction with unscaled eigenfunctions and eigenvalues

```{r pred_debias_unscale, results='hide'}
# approximation process
# container
pred_list <- list()
test_id <- unique(test_df$id)

# prediction for all subjects
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

  # per max obs time
  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K, 
                   X=efunctions_new[1:max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(new_lambda), 
                   f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), 
                                Data=MyData, Method = "BFGS")
    # individual score
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- new_mu+efunctions_new%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  pred_list[[i]] <- df_i
}
```

```{r}
df_pred <- bind_rows(pred_list)

df_pred$pred0.2[df_pred$t<=0.2] <- NA
df_pred$pred0.4[df_pred$t<=0.4] <- NA
df_pred$pred0.6[df_pred$t<=0.6] <- NA
df_pred$pred0.8[df_pred$t<=0.8] <- NA

rand_test_id <- sample(test_id, 4)

p1<-df_pred %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Unscaled", col = "Maximum observation time")

```
## Prediction with scaled eigenfunctions and eigenvalues

```{r pred_debias_scale, results='hide'}
# container
pred_list2 <- list()

# per subject
for(i in seq_along(test_id)){
  df_i <- test_df %>% filter(id==test_id[i])

  # per max obs time
  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
    df_it <- df_i %>% filter(t <= tmax)
    max_rid <- nrow(df_it)

    # into a list
    MyData <- list(K=K,
                   X=efunctions_scaled[1: max_rid, ], 
                   mon.names=mon.names,
                   parm.names=parm.names, 
                   pos.xi=pos.xi, 
                   y=df_it$Y, 
                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])

    # fit laplace approximation
    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
    score <- Fit$Summary1[, "Mode"]

    # prediction
    eta_pred_out <- new_mu+efunctions_scaled%*%score
    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
  }

  pred_list2[[i]] <- df_i
}
```

```{r}
df_pred2 <- bind_rows(pred_list2)

df_pred2$pred0.2[df_pred2$t<=0.2] <- NA
df_pred2$pred0.4[df_pred2$t<=0.4] <- NA
df_pred2$pred0.6[df_pred2$t<=0.6] <- NA
df_pred2$pred0.8[df_pred2$t<=0.8] <- NA

# overview of predicted track
p2 <- df_pred2 %>%
  filter(id %in% rand_test_id) %>%
  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.2)+
  geom_line(aes(x=t, y=eta_i, col = "True"))+
  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
  facet_wrap(~id)+
  labs(title = "Scaled", col="Maximum observation time")

```

## Compare performance

- Individual prediction track (use four randomly-selected subjects as example)

```{r, fig.height=5, fig.width=10}
ggarrange(p1, p2, nrow = 1, common.legend = T)
```

- Compare integrated squared error

```{r}
# calcualte ISE
err1 <- df_pred %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err1$window = cut(err1$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err1 <- split(err1, f = err1$window)

tb1 <- lapply(err1, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

```{r}
# calcualte ISE
err2 <- df_pred2 %>%
  mutate(err1 = (pred0.2-eta_i)^2,
         err2 = (pred0.4-eta_i)^2,
         err3 = (pred0.6-eta_i)^2,
         err4 = (pred0.8-eta_i)^2) %>%
  select(id, t, starts_with("err"))

err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
# table(err1$window)
err2 <- split(err2, f = err2$window)

tb2 <- lapply(err2, function(x){
  x %>%  group_by(id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>%
    summarise_at(vars(err1, err2, err3, err4), mean)
}) %>% bind_rows(.id = "Window")
```

```{r}
# calcualte ISE
options(knitr.kable.NA = "")

ise <- full_join(tb1, tb2, by = "Window")
colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
ise %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 8))
```

- Compare AUC

```{r}
# calcualte AUC
auc_df1 <- df_pred %>%
  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
  select(Y, starts_with("pred"), window)

obs_track <- seq(0.2, 0.8, by = 0.2)
pred_window <- unique(auc_df1$window)
auc_mat <- matrix(NA, length(pred_window), length(obs_track))

for(j in seq_along(obs_track)){
  for(i in (j+1):5){
    this_Y <- auc_df1$Y[auc_df1$window==pred_window[i]]
    this_pred <- auc_df1[auc_df1$window==pred_window[i],
                         paste0("pred", obs_track[j])]
    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
    auc_mat[i ,j] <- this_perf@y.values[[1]]
  }
}

```

```{r}
# calcualte AUC
auc_df2 <- df_pred2 %>%
  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
  select(Y, starts_with("pred"), window)

auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))

for(j in seq_along(obs_track)){
  for(i in (j+1):5){
    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
    this_pred <- auc_df2[auc_df2$window==pred_window[i],
                         paste0("pred", obs_track[j])]
    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
    auc_mat2[i ,j] <- this_perf@y.values[[1]]
  }
}

```

```{r}
# calcualte AUC
options(knitr.kable.NA = "")

auc <- cbind(auc_mat, auc_mat2)
rownames(auc) <- unique(auc_df1$window)
colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 2))
auc %>%
  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
  add_header_above(c(" "=1, "Maximum observation time" = 8))
```

- It looks as if scaled eigenfunctions and eigenvalues led to worse prediction, especially when prediction is conditioning on observation up to t=0.4

### 2023-10-25 Docs/ProgressReportSim.Rmd
update mean function in the debias step

@@
   html_document:
-    self_contained: no
+    self_contained: yes
     number_sections: true
@@
 t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
+debias_glmm <- bam(Y ~ s(t, bs="cc", k=10)+
                      s(id, by=phi1, bs="re")+
@@
 
-This is pretty much a linear functions. Is that supposed to happen?
+The shape of the function depends very much on how the basis functions are constructed.
+For example, cubic spline would lead to a linear mean and cyclic cubic spline would lead to the function below.
+What's a good way to decided?
 
 ```{r, class.source = "fold-show"}
-new_mu <- predict(debias_glmm, type = "terms")[, 1]
-new_mu <- unique(new_mu)
-plot(unique(train_df$t), new_mu, type = "l")
+new_mu <- predict(debias_glmm, type = "terms")
+length(unique(new_mu[,1]))
+
+new_mu <- unique(new_mu[,1])
+plot(t, new_mu, type = "l")
 ```
@@
 
+```{r}
+data.frame(
+  df_pred$pred0.4[df_pred$t>0.4],
+  df_pred3$pred0.4[df_pred3$t>0.4]
+)
+
+```
+
 # Reference method

### 2023-11-09 Docs/ProgressReportSim.Rmd
pred difference from scaled or unscaled parameters

@@
 author: "Ying Jin"
-date: "2023-09-15"
+date: "`r Sys.Date()`"
 output: 
@@
 N_test <- 200
-b_test <- matrix(rnorm(N_test*K), N_test, K) %*% diag(sqrt(lambda)) %*% t(phi) # of size N by J
+# xi_test <- matrix(rnorm(N_test*K), N_test, K)
+xi_test <- rmvnorm(N_test, mean = rep(0, K), sigma = diag(lambda))
+b_test <- xi_test %*% t(phi) # of size N by J
 # dim(b_test)
@@
 eta_test <- t(vapply(1:N_test, function(x){
-    f_0(x) + b_i[x,]
+    f_0(x) + b_test[x,]
   }, numeric(J)))
@@
 # bin
-test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
-test_df$bin <- as.numeric(as.character(test_df$bin))
+# test_df$bin <- cut(test_df$sind, breaks = brks, include.lowest = T, labels = mid)
+# test_df$bin <- as.numeric(as.character(test_df$bin))
 test_df$id <- as.factor(test_df$id)
@@
 
-```{r}
+```{r model}
 # the model
@@
 pos.xi <- grep("xi", parm.names)
-PGF <- function(Data) {
-  xi <- rnorm(Data$K)
-  return(xi)
-}
+# PGF <- function(Data) {
+#   xi <- rnorm(Data$K)
+#   return(xi)
+# }
 mon.names <- "LP"
@@
 pred_list <- list()
+converge_vec <- matrix(NA, N_test, 4)
 # score <- array(NA, dim=c(length(test_id), K, 4))
@@
 # per subject
+# seq_along(test_id)
 for(i in seq_along(test_id)){
@@
     # into a list
-    MyData <- list(K=K, PGF=PGF, 
+    MyData <- list(K=K, 
                    X=efunctions_new[1:max_rid, ], 
@@
     # fit laplace approximation
-    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS", Iterations = 1000,
+                                CovEst = "Identity")
+    converge_vec[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
     # Fit <- LaplacesDemon(Model, Data=MyData, Initial.Values = parm = rep(0, K),  Method = "BFGS")
@@
 
-<!-- ### Scaled eigenfunctions and eigenvalues -->
-
 ```{r pred_debias_scale, results='hide'}
@@
 pred_list3 <- list()
-
+converge_vec2 <- matrix(NA, N_test, 4)
 # per subject
@@
     # into a list
-    MyData <- list(K=K, PGF=PGF, 
+    MyData <- list(K=K,
                    X=efunctions_scaled[1: max_rid, ], 
@@
     # fit laplace approximation
-    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS", Iterations = 1000,
+                                CovEst = "Identity")
+    converge_vec2[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
     score <- Fit$Summary1[, "Mode"]
@@
 }
+
+mean(converge_vec2)
 ```
@@
 
-- The scaled and unscaled parameters seem to make a difference in prediction, which they shouldn't
+When using the unscaled estimates of eigenvalues and eigenfunctions, the Laplace Approximation does not converge all the time (convergence rate = `r round(mean(converge_vec), 2)`).
+And when using scaled estimates, the convergence rate is even lower (`r round(mean(converge_vec2), 2)`).
 
-```{r, fig.height=8, fig.width=8}
-par(mfrow=c(2, 2))
-plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5, 
-     xlab="unscaled", ylab="scaled", main = "Up to 0.2")
-plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
-     xlab="unscaled", ylab="scaled", main = "Up to 0.4")
-plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
-     xlab="unscaled", ylab="scaled", main = "Up to 0.6")
-plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
-     xlab="unscaled", ylab="scaled", main = "Up to 0.8")
+When I increasing the maximum iteration to 1000, the predictive performance from two sets of estiamtes got very similar.
+However, the convergence rate is still very different.
+Below I try to show a example of the discrepancy in prediction.
+As we see, the most problematic situation is when observations is up to 0.4.
+
+```{r}
+# subjects with higher deiviation 
+# subjects with greatest difference
+id_diff <- df_pred %>% select(id, t, pred0.4) %>%
+  full_join(df_pred3 %>% select(id, t, pred0.4), by = c("id", "t")) %>% 
+  group_by(id) %>%
+  summarise(diff = mean((pred0.4.x-pred0.4.y)^2, na.rm = T)) %>% 
+  arrange(desc(diff))
+
+# subject 43, 79, 57, 148 have the most different prediction
+df_pred %>% filter(id == id_diff$id[1]) %>% 
+  select(t, starts_with("pred")) %>%
+  pivot_longer(starts_with("pred")) %>%
+  left_join(df_pred3 %>% filter(id == id_diff$id[1]) %>%
+              select(t, starts_with("pred")) %>%
+              pivot_longer(starts_with("pred")),
+            by = c("t", "name")) %>%
+  ggplot()+
+  geom_point(aes(x=value.x, y=value.y), na.rm = T)+
+  facet_wrap(~name)+
+  labs(title = paste0("Subject ", id_diff$id[1]), x="Unscaled", y="Scaled")
+```
+
+Think about the estimates.
+The scaled estimates of eigenvalues are much smaller than the unscaled once, meaning the prior distribution of score using scaled estimates would be much more centered thant the unscaled ones.
+This would also drive the posterior distribution to be in similar shapes.
+Since the distribution from scaled eigenvalues has such high density at certain values, could it be that the sampling quickly got trapped in a small interval that is far from the true optimal values?
+
+Take the same subject with observation up to 0.4 as an example.
+Below is the iterative history until convergence.
+In this case, unscaled estiamtes converged at the 67 iteration and the scaled one did not converge.
+
+```{r, results='hide'}
+this_id <- id_diff$id[1]
+this_t <- 0.4
+
+# data
+df_it <- test_df %>% filter(id==this_id & t <= this_t)
+max_rid <- nrow(df_it)
+
+#### use Unscaled params ####
+MyData_unscaled <- list(K=K, 
+               X=efunctions_new[1:max_rid, ], 
+               mon.names=mon.names,
+               parm.names=parm.names, 
+               pos.xi=pos.xi, 
+               y=df_it$Y, 
+               tao=diag(new_lambda), f0=new_mu[1:max_rid])
+Fit_unscaled <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData,
+                                     Method = "BFGS", Iterations = 1000,
+                            CovEst = "Identity")
+
+#### Scaled params ####
+MyData_scaled <- list(K=K, 
+               X=efunctions_scaled[1:max_rid, ], 
+               mon.names=mon.names,
+               parm.names=parm.names, 
+               pos.xi=pos.xi, 
+               y=df_it$Y, 
+               tao=diag(scaled_lambda), f0=new_mu[1:max_rid])
+Fit_scaled <- LaplaceApproximation(Model, parm = rep(0, K), 
+                                   Data=MyData_scaled, Method = "BFGS", 
+                                   Iterations = 1000,
+                            CovEst = "Identity")
+
+## iterative distribution, compared to true value
+true_xi_it <- data.frame(name = colnames(Fit_unscaled$History),
+                         true_xi = xi_test[id_diff$id[1], ])
 ```
@@
 ```{r}
-data.frame(
-  df_pred$pred0.4[df_pred$t>0.4],
-  df_pred3$pred0.4[df_pred3$t>0.4]
-)
+ggarrange(
+Fit_unscaled$History %>% as.data.frame() %>%
+  mutate(iter = 1:nrow(Fit$History)) %>%
+  pivot_longer(starts_with("xi")) %>%
+  left_join(true_xi_it, by = "name") %>%
+  mutate(value = value/sqrt(n_bin)) %>%
+  mutate_at(vars(value, true_xi), abs) %>%
+  ggplot()+
+  geom_point(aes(x=iter, y=value))+
+  geom_line(aes(x=iter, y = true_xi, col = "red"))+
+  facet_wrap(~name)+
+  labs(title = "Unscaled parameter"),
 
+Fit_scaled$History %>% as.data.frame() %>%
+  mutate(iter = 1:nrow(Fit_scaled$History)) %>%
+  pivot_longer(starts_with("xi")) %>%
+  left_join(true_xi_it, by = "name") %>%
+  mutate_at(vars(value, true_xi), abs) %>%
+  ggplot()+
+  geom_point(aes(x=iter, y=value))+
+  geom_line(aes(x=iter, y = true_xi, col = "red"))+
+  facet_wrap(~name)+
+  labs(title = "Scaled parameter"), nrow = 1, common.legend = T)
+```
+
+Let's zoom in on the later part of the scaled parameter. It looks like sampling bounced symmetrically around a certain line.
+
+```{r}
+Fit_scaled$History %>% as.data.frame() %>%
+  mutate(iter = 1:nrow(Fit_scaled$History)) %>%
+  filter(iter >= 100) %>%
+  pivot_longer(starts_with("xi")) %>%
+  left_join(true_xi_it, by = "name") %>%
+  mutate_at(vars(value, true_xi), abs) %>%
+  ggplot()+
+  geom_point(aes(x=iter, y=value), size = 0.2)+
+  # geom_line(aes(x=iter, y = true_xi, col = "red"))+
+  facet_wrap(~name, scales = "free")+
+  labs(title = "Scaled parameter, after 100th iteration")
 ```
@@
 
-```{r, class.source='fold-show'}
+```{r}
 # model fit
@@
 t2 <- Sys.time()
-# time 
-t2-t1
-# model
-summary(fit_adglmm)
+# t2-t1
 ```
@@
 
-```{r, class.source = "fold-show"}
+```{r}
 # subset 10 subjects from training dataset 
@@
 # intersect(t, unique(df_train_sub$t))
+# View(df_train_sub)
 ```
@@
 
-```{r, class.source="fold-show"}
+```{r}
 # construct design matrix for the fixed effect intercept
 t_sub <- unique(df_train_sub$t)
-fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
-dim(fix_xmat[[1]]$X)
+# fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
+fix_xmat <- ns(t_sub, df=5)
+# dim(fix_xmat[[1]]$X)
 
@@
 fix_xmat <- data.frame(t=t_sub, 
-                       fix_xmat[[1]]$X)
+                       fix_xmat)
 colnames(fix_xmat) <- c("t", paste0("B", 1:5))
@@
 
-```{r, class.source = "fold-show"}
+```{r}
 # construct design matrix for the random effect
 rand_xmat <- smoothCon(s(t, bs="cr", k=4), data = data.frame(t=t_sub))
+# rand_xmat <- ns(t_sub, df = 4)
 
-dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
+# dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
 rand_xmat <- data.frame(t = t, rand_xmat[[1]]$X)
@@
 
-The model fitting took about 6 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.
+The model fitting took about 12 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.
 
@@
 t2 <- Sys.time()
-t2-t1
+# t2-t1
 
-# model
-summary(fit_adglmm)
 ```
@@
 ```{r}
-df_test_sub <- test_df %>% filter(id %in% 1:10 & sind==bin)
+df_test_sub <- test_df %>% filter(id %in% 1:10 & sind %in% mid)
 df_test_sub$id <- droplevels(df_test_sub$id)
@@
 
-colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:15], "pred1", "pred2", "pred3", "pred4")
+colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:14], "pred1", "pred2", "pred3", "pred4")
 ```
@@
 ```
+
+# Notes
+
+Perhaps I should try to verify my thoughts about laplace approximation by different generation scheme. For example, what if the true eigenvalues are on larger scale?

### 2023-11-13 Docs/ProgressReportSim.Rmd
update figure

@@
 
-```{r, cache=TRUE}
+```{r}
 # fit model on the training set
@@
 
-# project mean functions 
-# lm_mod <- lm(fpca_mod$mu ~ B-1)
-# mu_new <- Bnew %*% coef(lm_mod)
 ```
@@
   ggplot()+
-  geom_point(aes(x=value.x, y=value.y), na.rm = T)+
+  geom_point(aes(x=value.x, y=value.y), na.rm = T, size = 0.2)+
   facet_wrap(~name)+
@@
 
-#### use Unscaled params ####
+# use unscaled params
 MyData_unscaled <- list(K=K, 
@@
                tao=diag(new_lambda), f0=new_mu[1:max_rid])
-Fit_unscaled <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData,
+Fit_unscaled <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData_unscaled,
                                      Method = "BFGS", Iterations = 1000,
@@
 
-#### Scaled params ####
+# use scaled params
 MyData_scaled <- list(K=K, 
@@
 Fit_unscaled$History %>% as.data.frame() %>%
-  mutate(iter = 1:nrow(Fit$History)) %>%
+  mutate(iter = 1:nrow(Fit_unscaled$History)) %>%
   pivot_longer(starts_with("xi")) %>%
@@
   labs(title = "Scaled parameter, after 100th iteration")
+```
+
+```{r converge_strata, eval=FALSE}
+# Look at prediction results stratified by convergence status
+converge_vec
+converge_vec2
+```
+
+```{r, eval=FALSE}
+# calcualte ISE
+err1_conv <- df_pred %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err1_conv$window = cut(err1_conv$t,
+                       breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err1_conv <- split(err1_conv, f = err1_conv$window)
+err1_conv[[1]]
+
+tb1 <- lapply(err1, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) 
+}) %>% bind_rows(.id = "Window")
 ```
@@
 Perhaps I should try to verify my thoughts about laplace approximation by different generation scheme. For example, what if the true eigenvalues are on larger scale?
+
+- Well, after increasing the scale of the true eigenvalues, the convergence rate did get closer between scaled and unscaled parameter. Also reversed as I expected.
+- But on second though, A more effecient way is to calcualted ISE and AUC stratified by convergence status

### 2023-11-13 Scripts/LaplacePredScale.R

@@ -1,9 +1,270 @@
 
-# I want to look at the posterior distribution from scaled and unscaled estiamte
-# also figure out why scaled parameters would lead to such low convergence rate
-# with scaled parameter, the posterior distribution of scores would be much, much more centered
-# does it mean that sampling got stuck in a neighborhood of false values and was never able to get close to the "true" values? 
+set.seed(516)
 
+#### Set up ####
+
+library(here)
+library(tidyverse)
+library(ggpubr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(kableExtra)
+library(knitr)
+library(mvtnorm)
+library(mgcv)
+library(splines)
+
+source(here("Code/GLMM-FPCA.R")) 
+source(here("Code/OutsampBayes.R"))
+
+#### generated training data ####
+
+N <- 500 # sample size
+J <- 1000 # number of observation points
+
+t = seq(0,1,len=J) # observations points
+
+# mean function
+f_0 <- function(s) 0 
+
+#eigenfunctions 
+K <- 4 # number of eigenfunctions
+phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
+                     sin(4*pi*t),cos(4*pi*t))
+
+# eigenvalues
+lambda = rep(10, 4)
+
+# score
+xi <- matrix(rnorm(N*K),N,K)
+xi <- xi %*% diag(sqrt(lambda))
+
+# subject-specific random effect
+b_i <- xi %*% t(phi) # of size N by J
+
+# latent gaussian function
+eta_i <- t(vapply(1:N, function(x){
+  f_0(t) + b_i[x,]
+}, numeric(J)))
+
+# outcome binary function
+Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
+              N, J, byrow=FALSE)
+
+train_df <- data.frame(id = factor(rep(1:N, each=J)),
+                       t = rep(t, N), 
+                       Y = as.vector(t(Y_i)),
+                       eta_i = as.vector(t(eta_i)),
+                       sind = rep(1:J, N))
+
+# visualization
+rand_id <- sample(N, size = 4)
+train_df %>% filter(id %in% rand_id) %>% 
+  ggplot()+
[…]

### 2023-11-14 Docs/ProgressReportSim.Rmd
try different optimizers

@@
     # fit laplace approximation
-    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS", Iterations = 1000,
+    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "NM", Iterations = 1000,
                                 CovEst = "Identity")
@@
 t_pred <- t2-t1 # About 3.5 minutes
+# mean(converge_vec)
 ```
@@
 # container
+t1 <- Sys.time()
 pred_list3 <- list()
@@
 }
-
-mean(converge_vec2)
+t2 <- Sys.time()
+# t2-t1
+# mean(converge_vec2)
 ```
@@
   add_header_above(c(" "=1, "Maximum observation time" = 8))
-```
-
-- Looks like the debiased eigenvalues helped.
-- Scaling, on the other hand, doesn't seem to help in general. But shouldn't scaling improve performance? 
-- Also, it seems like the longer the observed track, the less different the performance is.
+
+```
 
@@
 
-## Investigate a bit more on the difference in prediction
-
-When using the unscaled estimates of eigenvalues and eigenfunctions, the Laplace Approximation does not converge all the time (convergence rate = `r round(mean(converge_vec), 2)`).
-And when using scaled estimates, the convergence rate is even lower (`r round(mean(converge_vec2), 2)`).
-
-When I increasing the maximum iteration to 1000, the predictive performance from two sets of estiamtes got very similar.
-However, the convergence rate is still very different.
-Below I try to show a example of the discrepancy in prediction.
-As we see, the most problematic situation is when observations is up to 0.4.
-
-```{r}
-# subjects with higher deiviation 
-# subjects with greatest difference
-id_diff <- df_pred %>% select(id, t, pred0.4) %>%
-  full_join(df_pred3 %>% select(id, t, pred0.4), by = c("id", "t")) %>% 
-  group_by(id) %>%
-  summarise(diff = mean((pred0.4.x-pred0.4.y)^2, na.rm = T)) %>% 
-  arrange(desc(diff))
-
-# subject 43, 79, 57, 148 have the most different prediction
-df_pred %>% filter(id == id_diff$id[1]) %>% 
-  select(t, starts_with("pred")) %>%
-  pivot_longer(starts_with("pred")) %>%
-  left_join(df_pred3 %>% filter(id == id_diff$id[1]) %>%
-              select(t, starts_with("pred")) %>%
-              pivot_longer(starts_with("pred")),
-            by = c("t", "name")) %>%
-  ggplot()+
-  geom_point(aes(x=value.x, y=value.y), na.rm = T, size = 0.2)+
-  facet_wrap(~name)+
-  labs(title = paste0("Subject ", id_diff$id[1]), x="Unscaled", y="Scaled")
+## Numeric stability
+
+Following up on the discrepancy in predictive performance using scaled and unscaled estimates of eigenfunctions and eigenvalues, it is likely to be a issue of numeric stability.
+I would like to compare a few optimizers to see if they make any difference on the results.
+I'll record computation time, convergence rate using scaled and unscaled estimates, a brief summary of performance.
+
+1. BFGS 
+- computation time: 3.43min
+- convergance rate: 94% (unscaled), 75% (scaled)
+- performance is not too different
+
+2. Nelder-Mead
+- computation time: 3.48 min (unscaled), 3.03 min (scaled)
+- convergance rate: 100% (unscaled), 100% (scaled)
+- performance very similar. Scaled parameters can sometimes outperform unscaled ones.
+
+3. Newton-Ralphson
+- ran into numeric problems when using scaled parameter
+
+Over all, I think Nelder-Mead is probably our bdst option.
+It converged under all situations and is also fast.
+When looking at ISE, scaled and unscaled estiamtes had similar performance.
+In fact they outperform each other five times each.
+
+Below I compared the predictions using scaled and unscaled parameters:
+
+```{r, fig.height=8, fig.width=8}
+par(mfrow = c(2, 2))
+plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5,
+     main = "Up to 0.2", xlab = "Unscaled", ylab = "Scaled")
+plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
+     main = "Up to 0.4", xlab = "Unscaled", ylab = "Scaled")
+plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
+     main = "Up to 0.6", xlab = "Unscaled", ylab = "Scaled")
+plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
+     main = "Up to 0.8", xlab = "Unscaled", ylab = "Scaled")
 ```
@@
 Think about the estimates.
-The scaled estimates of eigenvalues are much smaller than the unscaled once, meaning the prior distribution of score using scaled estimates would be much more centered thant the unscaled ones.
+The scaled estimates of eigenvalues are much smaller than the unscaled ones, meaning the prior distribution of score using scaled estimates would be much more centered thant the unscaled ones.
 This would also drive the posterior distribution to be in similar shapes.
 Since the distribution from scaled eigenvalues has such high density at certain values, could it be that the sampling quickly got trapped in a small interval that is far from the true optimal values?
-
-Take the same subject with observation up to 0.4 as an example.
-Below is the iterative history until convergence.
-In this case, unscaled estiamtes converged at the 67 iteration and the scaled one did not converge.
+It did not happen when using Nelder-Mead optimizer and set maxiter=1000.
+
+Below I look at the convergence path for one arbitrary subject
 
 ```{r, results='hide'}
-this_id <- id_diff$id[1]
+this_id <- sample(test_id, 1)
 this_t <- 0.4
@@
 true_xi_it <- data.frame(name = colnames(Fit_unscaled$History),
-                         true_xi = xi_test[id_diff$id[1], ])
+                         true_xi = xi_test[this_id, ])
 ```
@@
   ggplot()+
-  geom_point(aes(x=iter, y=value))+
+  geom_point(aes(x=iter, y=value), size = 0.2)+
   geom_line(aes(x=iter, y = true_xi, col = "red"))+
@@
 
-Let's zoom in on the later part of the scaled parameter. It looks like sampling bounced symmetrically around a certain line.
-
-```{r}
-Fit_scaled$History %>% as.data.frame() %>%
-  mutate(iter = 1:nrow(Fit_scaled$History)) %>%
-  filter(iter >= 100) %>%
-  pivot_longer(starts_with("xi")) %>%
-  left_join(true_xi_it, by = "name") %>%
-  mutate_at(vars(value, true_xi), abs) %>%
-  ggplot()+
-  geom_point(aes(x=iter, y=value), size = 0.2)+
-  # geom_line(aes(x=iter, y = true_xi, col = "red"))+
-  facet_wrap(~name, scales = "free")+
-  labs(title = "Scaled parameter, after 100th iteration")
-```
-
-```{r converge_strata, eval=FALSE}
-# Look at prediction results stratified by convergence status
-converge_vec
-converge_vec2
-```
-
-```{r, eval=FALSE}
-# calcualte ISE
-err1_conv <- df_pred %>%
-  mutate(err1 = (pred0.2-eta_i)^2,
-         err2 = (pred0.4-eta_i)^2,
-         err3 = (pred0.6-eta_i)^2,
-         err4 = (pred0.8-eta_i)^2) %>%
-  select(id, t, starts_with("err"))
-
-err1_conv$window = cut(err1_conv$t,
-                       breaks = seq(0, 1, by = 0.2), include.lowest = T)
-# table(err1$window)
-err1_conv <- split(err1_conv, f = err1_conv$window)
-err1_conv[[1]]
-
-tb1 <- lapply(err1, function(x){
-  x %>%  group_by(id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) 
-}) %>% bind_rows(.id = "Window")
-```
+In this case the sampling bounced up and down, which is NM is supposed to do I think? But it reaches a stable status quickly with a few dozens of iterations.
+
+Potential drawback of Nelder-Mead: large dimensional problems.
 

### 2023-11-14 Code/GenData.R
start full simulation

@@ -1,66 +1,61 @@
 
-# This sciprt generateds 500 data sets
+# This sciprt generateds 500 data sets for the simulation study
 
-set.seed(314)
-N <- 500 # sample size
+set.seed(1114)
+
+library(mvtnorm)
+
+N_train <- 500 # training sample size
+N_test <- 100 # testing sample size
+N <- N_train+N_test
+
 J <- 1000 # number of observation points
+t = seq(0,1,len=J) # observation grid 
+
+# fixed effect: mean function
+f_0 <- function(s){return(0)}
+
+# random effects
+K <- 4 # number of eigenfunctions
+
+phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
+                     sin(4*pi*t),cos(4*pi*t)) # orthogonal random functions
+
+lambda = 0.5^(0:(K-1)) # eigenvalues/variance of scores
 
 sind = seq(0,1,len=J) # observations points
 
-
-f_0 <- function(s) 0 # mean functions
-f_0(5)
-
-
-#### eigenfunctions #####
-K <- 4 # number of eigenfunctions
-phi <- sqrt(2)*cbind(sin(2*pi*sind),cos(2*pi*sind),
-                     sin(4*pi*sind),cos(4*pi*sind))
-
+# visualize eigenfunctions
 par(mfrow=c(2, 2))
-plot(sind, phi[,1])
-plot(sind, phi[,2])
-plot(sind, phi[,3])
-plot(sind, phi[,4])
+plot(t, phi[,1])
+plot(t, phi[,2])
+plot(t, phi[,3])
+plot(t, phi[,4])
+
 
 
 #### generated data ####
 
-sim_data <- list()
+sim_data <- list() # container
 M <- 500 # number of datasets to generate
-lambda = 0.5^(0:(K-1)) # eigenvalues
 
 for(m in 1:M){
   
   # generate score
-  xi <- matrix(rnorm(N*K),N,K)
-  xi <- xi %*% diag(sqrt(lambda))
-  
-  # subject-specific random effect
-  b_i <- xi %*% t(phi); # of size N by J
-  
+  xi <- rmvnorm(N, mean = rep(0, K), sigma = diag(lambda))
+  bi <- xi %*% t(phi) # random effect functions
+
   # latent gaussian function
-  eta_i <- t(vapply(1:N, function(x){
-    f_0(sind) + b_i[x,]
-  }, numeric(J)))
+  eta_i <- f_0(t)+bi
   
-  # outcome binary function
[…]

### 2023-11-14 Code/Sim_fGFPCA.R

@@ -1,59 +1,72 @@
 # This script implements the full simulation 
 # for dynamic prediction of fGFPCA
+# on the simulated dataset generated from Code/GenData.R
 
 #### set up ####
 
-set.seed(516)
+set.seed(1114)
 
 library(here)
 library(tidyverse)
-theme_set(theme_minimal())
-
+library(ggpubr)
 library(refund)
 library(lme4)
 library(ROCR)
 library(GLMMadaptive)
+library(kableExtra)
+library(knitr)
+library(mvtnorm)
 library(mgcv)
 library(splines)
-
+library(LaplacesDemon)
+theme_set(theme_minimal())
 
 
 #### load simulated data ####
 
 load(here("Data/sim_data.RData"))
 
-# data structure
-M <- length(sim_data) # number of simulation
-N <- length(unique(sim_data[[1]]$id)) # sample size 
-J <- length(unique(sim_data[[1]]$sind)) # number of measurements for each subject
+N <- length(unique(sim_data[[1]]$id))
+J <- length(unique(sim_data[[1]]$t))
+t <- unique(sim_data[[1]]$t)
+M <- length(sim_data)
 
-# data preview
-rand_id <- sample(as.numeric(unique(sim_data[[1]]$id)), size=4)
-
-sim_data[[1]] %>% 
-  filter(id %in% rand_id) %>%
+# overview of a few samples
+rand_id <- sample(N, size = 4)
+sim_data[[357]] %>% filter(id %in% rand_id) %>% 
   ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  geom_line(aes(x=sind, y=exp(eta_i)/(1+exp(eta_i))))+
+  geom_point(aes(x=t, y=Y))+
+  geom_line(aes(x=t, y=plogis(eta_i)), col = "red")+
+  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
   facet_wrap(~id)
 
+#### fGFPCA Model estimation ####
 
-##### fGFPCA Model estimation #####
-
+# functions
 source(here("Code/GLMM-FPCA.R")) 
 source(here("Code/OutSampBayes.R"))
-# use pred_latent function to estimate latent function 
 
-# bin data
+# split data into train and test set
+N_train <- 500
+N_test <- 100
+
+# binning 
 bin_w <- 10 # bin width
 n_bin <- J/bin_w # number of bins
[…]

### 2023-11-16 Docs/SimExample.Rmd
add numeric stability to example file

@@
 test_id <- unique(test_df$id)
+converge_vec <- matrix(NA, N_test, 4) # convergence status
 
@@
                                 Data=MyData, Method = "BFGS")
+    converge_vec[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
     # individual score
@@
 ```
+
 ## Prediction with scaled eigenfunctions and eigenvalues
@@
 pred_list2 <- list()
+converge_vec2 <- matrix(NA, N_test, 4)
 
@@
     Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "BFGS")
+    converge_vec2[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
     score <- Fit$Summary1[, "Mode"]
@@
 - It looks as if scaled eigenfunctions and eigenvalues led to worse prediction, especially when prediction is conditioning on observation up to t=0.4
+
+## Numeric stability
+
+I start by looking at the convergence rate of scaled and un-scaled estimates:
+
+- unscaled: `r round(mean(converge_vec), 2)`
+- scaled: `r round(mean(converge_vec2), 2)`
+
+Then, I would like to see some individuals who performed the most different under the two set of estimates. I used the observations up to 0.4 as an example, since it is the most problematic scenario.
+
+Below I identify the subjects with most dependency in prediction
+
+```{r}
+# subjects with the most difference in prediction
+diff_subjs <- full_join(df_pred %>% select(id, t, pred0.4),
+          df_pred2 %>% select(id, t, pred0.4),
+          by = c("id", "t")) %>% 
+  mutate(diff = (pred0.4.x-pred0.4.y)^2) %>%
+  group_by(id) %>%
+  summarise(diff = sum(diff, na.rm = T)) %>% 
+  arrange(desc(diff)) %>% 
+  head(10)
+```
+
+Now I look at there convergence status
+
+- unscaled:
+
+```{r}
+converge_vec[diff_subjs$id,  2]
+```
+
+- scaled:
+
+```{r}
+converge_vec2[diff_subjs$id,  2]
+```
+
+And there prediction track (take the top 4):
+
+```{r}
+full_join(df_pred %>% select(id, t, Y, eta_i, pred0.4),
+          df_pred2 %>% select(id, t, pred0.4),
+          by = c("id", "t")) %>% 
+  filter(id %in% diff_subjs$id[1:4]) %>%
+    mutate_at(vars(eta_i, pred0.4.x, pred0.4.y), function(x){exp(x)/(1+exp(x))}) %>%
+    ggplot()+
+    geom_point(aes(x=t, y=Y), size = 0.2)+
+    geom_line(aes(x=t, y=eta_i, col = "True"))+
+    geom_line(aes(x=t, y=pred0.4.x, col = "unscaled"), linetype="dashed", na.rm=T)+
+    geom_line(aes(x=t, y=pred0.4.y, col = "scaled"), linetype="dashed", na.rm=T)+
+    facet_wrap(~id)+
+    labs(title = "Maximum observation time = 0.4")
+```

### 2023-11-16 Code/GLMM-FPCA.R

@@ -12,12 +12,12 @@ pred_latent <- function(df, n_node){
   
 # Additionally, get the average estimation across subject 
 # by extacting the random intercepts from each bin
-mean_latent <- function(df, n_node){
-    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ=n_node)
-    beta_hat <- coef(summary(this_glm))[1]
-    return(beta_hat)
-}
-  
+# mean_latent <- function(df, n_node){
+#     this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ=n_node)
+#     beta_hat <- coef(summary(this_glm))[1]
+#     return(beta_hat)
+# }
+#   
 
 
 

### 2023-11-16 Code/OutSampBayes.R

@@ -30,75 +30,75 @@ parm.names <- as.parm.names(name_lst)
 pos.xi <- grep("xi", parm.names)
 mon.names <- "LP"
 
+# 
+# ## a Bernoulli distribution at each time point
+# Model <- function(parm, Data){
+#   xi <- parm[Data$pos.xi]
+#   
+#   # log-prior
+#   xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)
+#   
+#   # log-posterior likelihood
+#   eta <- Data$f0+Data$X %*% xi
+#   p <- exp(eta)/(1+exp(eta))
+#   LL <- sum(dbern(x=Data$y, prob=p, log = TRUE)) # log likelihood of Y|xi
+#   LP <- LL+sum(xi.prior) # joint log likelihood of (Y, xi)
+#   
+#   # output
+#   Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LL,
+#                    yhat=Data$y, parm=parm)
+#   return(Modelout)
+# }
 
-## a Bernoulli distribution at each time point
-Model <- function(parm, Data){
-  xi <- parm[Data$pos.xi]
-  
-  # log-prior
-  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)
-  
-  # log-posterior likelihood
-  eta <- Data$f0+Data$X %*% xi
-  p <- exp(eta)/(1+exp(eta))
-  LL <- sum(dbern(x=Data$y, prob=p, log = TRUE)) # log likelihood of Y|xi
-  LP <- LL+sum(xi.prior) # joint log likelihood of (Y, xi)
-  
-  # output
-  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LL,
-                   yhat=Data$y, parm=parm)
-  return(Modelout)
-}
 
-
-# data: 
-out_pred_laplace <- function(mu, evalues, phi_mat, df_new, kpc, max_t){
-  
-  # put data into correct format
-  # ns <- as.vector(table(df_new$bin)) # number of observations
-  # hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
-  # nf <- ns-hs # number of failure
-  # max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
-  # df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)  
-  
-  # truncate data
[…]

### 2023-11-16 Code/Sim_fGFPCA.R

@@ -65,16 +65,18 @@ knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
 
 
 # result container
+M <- 10
+
 pred_list_all <- list()
 converge_state_list <- list()
 fit_time <- pred_time <- rep(NA, M)
 
 K <- 4 # number of eigenfunctions to use
 
+
 pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 
 # The whole process
-M <- 2
 for(m in 1:M){
   
   # for every simulated dataset
@@ -203,20 +205,25 @@ mean(fit_time) # average time for model fitting: 26 secs
 mean(pred_time) # average time for prediction: 133 secs 
 
 pred_list_all %>% lapply(dim)
+length(pred_list_all)
 
-rand_id <- sample(pred_list_all[[158]]$id, 4)
+# convergence status
+lapply(converge_state_list, mean)
+
+# visualize test samples
+rand_id <- sample(test_id, 4)
 
 # prediction results
-pred_list_all[[158]] %>% 
-  filter(id %in% sample(pred_list_all[[158]]$id, 4)) %>% 
-  mutate_at(vars(eta_i, pred1, pred2, pred3, pred4), function(x){exp(x)/(1+exp(x))}) %>%
+pred_list_all[[5]] %>% 
+  filter(id %in% rand_id) %>% 
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
   geom_point(aes(x=sind, y=Y), size = 0.2)+
   geom_line(aes(x=sind, y=eta_i, col = "True"))+
-  geom_line(aes(x=sind, y=pred1, col = "200"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=sind, y=pred2, col = "400"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=sind, y=pred3, col = "600"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=sind, y=pred4, col = "800"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
[…]

### 2023-11-20 Code/manuscript.Rmd
simulation fGFPCA output

--- 2023-05-17 Docs/manuscript.Rmd
@@
 ---
-title: "Dynamic Prediction with Fast Generalized Functional Principal Component analysis"
+title: "Simulation and Data Application Output"
 author: "Ying Jin"
-date: "2023-03-07"
-output: 
-  pdf_document:
-    citation_package: natbib
-    keep_text: true
-    fig_caption: true
-    latex_engine: pdflatex
-    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-ms.tex
-
+date: "2023-11-20"
+output: html_document
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 ```

### 2023-11-20 Docs/ProgressReportSim.Rmd
start full simulation
simulation fGFPCA output

@@
 
-# df <- read_rds(here("Data/nhanes_bi.rds"))
-# load(here("Data/ApplOutput_fGFPCA.RData"))
-
-# code
-source(here("Code/GLMM-FPCA.R")) 
-# source(here("Code/OutSampMLE.R"))
-source(here("Code/OutsampBayes.R"))
-# source(here("Code/OutsampBayes.R"))
-
+# load simulation results
+load(here("Data/SimOutput_fGFPCA.RData"))
 ```
@@
 
-### Porjection with B-spline basis
+### Projection with B-spline basis
 

### 2023-11-20 Docs/manuscript.Rmd
simulation fGFPCA output

@@
 ---
-title: "Dynamic Prediction with Fast Generalized Functional Principal Component analysis"
+title: "Tables and Figures to put in manuscript"
 author: "Ying Jin"
-date: "2023-03-07"
+date: "`r Sys.Date()`"
 output: 
-  pdf_document:
-    citation_package: natbib
-    keep_text: true
-    fig_caption: true
-    latex_engine: pdflatex
-    template: ~/Dropbox/miscelanea/svm-r-markdown-templates/svm-latex-ms.tex
-
+  html_document:
+    self_contained: yes
+    number_sections: true
+    code_folding: hide
+    toc: true
+    toc_depth: 3
+    toc_float: true
+    font: 12pt
 ---
@@
 knitr::opts_chunk$set(echo = FALSE)
+options(knitr.kable.NA = "")
+
+set.seed(1120)
+
+library(here)
+library(tidyverse)
+library(ggpubr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(kableExtra)
+library(knitr)
+library(mvtnorm)
+library(mgcv)
+library(splines)
+theme_set(theme_minimal())
+
+load(here("Data/SimOutput_fGFPCA.RData"))
+
+## prediction window 
+window <- seq(0, 1, by = 0.2)
+M <- length(pred_list_all)
 ```
 
-## R Markdown
+# Simulation
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+```{r fig_pred}
+rand_id <- sample(501:600, size = 4)
+# range(pred_list_all[[1]]$pred0.2, na.rm = T)
+pred_list_all[[1]] %>%
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "fGFPCA prediction")
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
 ```
 
-## Including Plots
+```{r tb_ise}
+## ISE container 
+ise_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+# dims: prediction window, max obs time, simulation iter
 
-You can also embed plots, for example:
+## calculation
+for(m in 1:M){
+  this_df <- pred_list_all[[m]]
+  ise_tb <- pred_list_all[[m]] %>%
+    mutate(err1 = (pred0.2-eta_i)^2,
+           err2 = (pred0.4-eta_i)^2,
+           err3 = (pred0.6-eta_i)^2,
+           err4 = (pred0.8-eta_i)^2) %>%
+    select(id, t, starts_with("err")) %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("err"))
+  ise_mat[, ,m] <- as.matrix(ise_tb)
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+}
+
+mean_ise <- apply(ise_mat, c(1, 2), mean)
+mean_ise <- data.frame(mean_ise) %>% 
+  mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+         .before = 1)
+colnames(mean_ise) <- c("Window", "0.2", "0.4", "0.6", "0.8")
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+```{r}
+mean_ise %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+```
+
+```{r auc_func}
+## a function to calculate AUC
+get_auc <- function(y, pred){
+  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
+    auc <- NA
+  }
+  else{
+    this_perf <- performance(prediction(pred, y), measure = "auc")
+    auc <- this_perf@y.values[[1]]
+  }
+  return(auc)
+}
+
+```
+
+```{r tb_auc}
+## auc container 
+auc_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+
+for(m in 1:M){
+  this_df <- pred_list_all[[m]]
+  auc_tb <- this_df %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred0.2),
+              auc2 = get_auc(Y, pred0.4),
+              auc3 = get_auc(Y, pred0.6),
+              auc4 = get_auc(Y, pred0.8)) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("auc"))
+  auc_mat[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc <- apply(auc_mat, c(1, 2), mean)
+mean_auc <- data.frame(mean_auc) %>% 
+  mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+         .before = 1)
+colnames(mean_auc) <- c("Window", "0.2", "0.4", "0.6", "0.8")
+```
+
+```{r}
+mean_auc %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area under the ROC curve") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+```
+
+```{r time}
+data.frame("Fit" = mean(fit_time)/60, "Prediction"= mean(pred_time)) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Computation time (minutes)") %>%
+  kable_styling(full_width = F)
+
+```

### 2023-11-20 Code/Sim_fGFPCA.R

@@ -30,6 +30,7 @@ N <- length(unique(sim_data[[1]]$id))
 J <- length(unique(sim_data[[1]]$t))
 t <- unique(sim_data[[1]]$t)
 M <- length(sim_data)
+K <- 4 # number of eigenfunctions to use
 
 # overview of a few samples
 rand_id <- sample(N, size = 4)
@@ -65,13 +66,12 @@ knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
 
 
 # result container
-M <- 10
-
+# M <- 10
+M
 pred_list_all <- list()
 converge_state_list <- list()
 fit_time <- pred_time <- rep(NA, M)
 
-K <- 4 # number of eigenfunctions to use
 
 
 pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
@@ -208,13 +208,13 @@ pred_list_all %>% lapply(dim)
 length(pred_list_all)
 
 # convergence status
-lapply(converge_state_list, mean)
+mean(sapply(converge_state_list, mean)) # all dataset converged
 
 # visualize test samples
 rand_id <- sample(test_id, 4)
 
 # prediction results
-pred_list_all[[5]] %>% 
+pred_list_all[[476]] %>% 
   filter(id %in% rand_id) %>% 
   mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
@@ -231,11 +231,98 @@ pred_list_all[[5]] %>%
 
 #### Save results ####
 
-save(pred_list_all, fit_time, pred_time, file = here("Data/SimOutput_fGFPCA.RData"))
+save(pred_list_all, fit_time, pred_time, converge_state_list,
+     file = here("Data/SimOutput_fGFPCA.RData"))
+
+#### Calculate ISE ####
+
+# load simulation results
+load(here("Data/SimOutput_fGFPCA.RData"))
+
+dim(pred_list_all[[1]])
+
+## prediction window 
+window <- seq(0, 1, by = 0.2)
+M <- length(pred_list_all)
+
+## ISE container 
+ise_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+# dims: prediction window, max obs time, simulation iter
+
+## calculation
+for(m in 1:M){
[…]

### 2023-11-27 Docs/ProgressReportSim.Rmd
simulation result summary

@@
 library(splines)
+library(LaplacesDemon)
 theme_set(theme_minimal())
@@
 # load simulation results
-load(here("Data/SimOutput_fGFPCA.RData"))
+# load(here("Data/SimOutput_fGFPCA.RData"))
 ```
@@
   facet_wrap(~id)
+```
+
+```{r}
+# functions
+source(here("Code/GLMM-FPCA.R")) 
+source(here("Code/OutSampBayes.R"))
 ```
@@
 
-```{r pred_debias_unscale, results='hide'}
+```{r pred_debias_scale, results='hide'}
 # approximation process
@@
     MyData <- list(K=K, 
-                   X=efunctions_new[1:max_rid, ], 
+                   X=efunctions_scaled[1:max_rid, ], 
                    mon.names=mon.names,
@@
                    y=df_it$Y, 
-                   tao=diag(new_lambda), f0=new_mu[1:max_rid])
+                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])
 
@@
   facet_wrap(~id)+
-  labs(title = "Unscaled, debiased eigenvalues")
-
-```
-
-```{r pred_debias_scale, results='hide'}
-## With debiased eigenvalues, also scale eigenvaleus and eigenfuncitons
-# container
-t1 <- Sys.time()
-pred_list3 <- list()
-converge_vec2 <- matrix(NA, N_test, 4)
-# per subject
-for(i in seq_along(test_id)){
-  df_i <- test_df %>% filter(id==test_id[i])
-
-  # per max obs time
-  for(tmax in c(0.2, 0.4, 0.6, 0.8)){
-    df_it <- df_i %>% filter(t <= tmax)
-    max_rid <- nrow(df_it)
-
-    # into a list
-    MyData <- list(K=K,
-                   X=efunctions_scaled[1: max_rid, ], 
-                   mon.names=mon.names,
-                   parm.names=parm.names, 
-                   pos.xi=pos.xi, 
-                   y=df_it$Y, 
-                   tao=diag(scaled_lambda), f0=new_mu[1:max_rid])
-
-    # fit laplace approximation
-    Fit <- LaplaceApproximation(Model, parm = rep(0, K), Data=MyData, Method = "NM", Iterations = 1000,
-                                CovEst = "Identity")
-    converge_vec2[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
-    score <- Fit$Summary1[, "Mode"]
-
-    # prediction
-    eta_pred_out <- new_mu+efunctions_scaled%*%score
-    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
-  }
-
-  # df_i$pred0.2[df_i$t<=0.2] <- NA
-  # df_i$pred0.4[df_i$t<=0.4] <- NA
-  # df_i$pred0.6[df_i$t<=0.6] <- NA
-  # df_i$pred0.8[df_i$t<=0.8] <- NA
-
-  pred_list3[[i]] <- df_i
-}
-t2 <- Sys.time()
-# t2-t1
-# mean(converge_vec2)
-```
-
-```{r}
-df_pred3 <- bind_rows(pred_list3)
-
-df_pred3$pred0.2[df_pred3$t<=0.2] <- NA
-df_pred3$pred0.4[df_pred3$t<=0.4] <- NA
-df_pred3$pred0.6[df_pred3$t<=0.6] <- NA
-df_pred3$pred0.8[df_pred3$t<=0.8] <- NA
-# overview of predicted track
-p3 <- df_pred3 %>%
-  filter(id %in% rand_test_id) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
-  labs(title = "Scaled, unbiased eigenvalues")
-
-```
-
-```{r}
-t_1sim <- (t_local_glmm+t_fpca+t_debias+t_pred)
-```
-
-I would like to see if the following operations can improve predictive performance:
-
-1. Debiasing eigenvalues from FPCA
-2. Scale the eigenfunctions and debiased eigenvalues (from mgcv::bam) so that they are close to the true values for data generation
-
-```{r, fig.height=5, fig.width=10}
-ggarrange(p1, p3, nrow = 1, common.legend = T)
-```
-
-- Time spent on one iteration: `r round(t_1sim, 2)` seconds
-- Debiasing eigenvalues helped with prediction. Scaling eigenfunctiosn and eigenvalues, on the other hand, ended up being more complicated.
-
-## ISE
+  labs(title = "fGFPCA")
+
+```
 
@@
 ```
-
-```{r}
-# calcualte ISE
-err3 <- df_pred3 %>%
-  mutate(err1 = (pred0.2-eta_i)^2,
-         err2 = (pred0.4-eta_i)^2,
-         err3 = (pred0.6-eta_i)^2,
-         err4 = (pred0.8-eta_i)^2) %>%
-  select(id, t, starts_with("err"))
-
-err3$window = cut(err3$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
-# table(err1$window)
-err3 <- split(err3, f = err3$window)
-
-tb3 <- lapply(err3, function(x){
-  x %>%  group_by(id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>%
-    summarise_at(vars(err1, err2, err3, err4), mean)
-}) %>% bind_rows(.id = "Window")
-```
-
-```{r}
-# calcualte ISE
-options(knitr.kable.NA = "")
-
-ise <- full_join(tb1, tb3, by = "Window")
-colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
-ise %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
-  add_header_above(c(" "=1, "Maximum observation time" = 8))
-
-```
-
-## AUC
 
@@
 
-```{r}
-# calcualte AUC
-auc_df3 <- df_pred3 %>%
-  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
-  select(Y, starts_with("pred"), window)
-
-auc_mat3 <- matrix(NA, length(pred_window), length(obs_track))
-
-for(j in seq_along(obs_track)){
-  for(i in (j+1):5){
-    this_Y <- auc_df3$Y[auc_df3$window==pred_window[i]]
-    this_pred <- auc_df3[auc_df3$window==pred_window[i],
-                         paste0("pred", obs_track[j])]
-    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
-    auc_mat3[i ,j] <- this_perf@y.values[[1]]
-  }
-}
-
-```
-
-```{r}
-# calcualte AUC
-options(knitr.kable.NA = "")
-
-auc <- cbind(auc_mat, auc_mat3)
-rownames(auc) <- unique(auc_df1$window)
-colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 2))
-auc %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" "= 1, "Unscaled" = 4, "Scaled" = 4)) %>%
-  add_header_above(c(" "=1, "Maximum observation time" = 8))
-```
-
 ## Numeric stability
@@
 In fact they outperform each other five times each.
-
-Below I compared the predictions using scaled and unscaled parameters:
-
-```{r, fig.height=8, fig.width=8}
-par(mfrow = c(2, 2))
-plot(df_pred$pred0.2, df_pred3$pred0.2, pch=20, cex = 0.5,
-     main = "Up to 0.2", xlab = "Unscaled", ylab = "Scaled")
-plot(df_pred$pred0.4, df_pred3$pred0.4, pch=20, cex = 0.5,
-     main = "Up to 0.4", xlab = "Unscaled", ylab = "Scaled")
-plot(df_pred$pred0.6, df_pred3$pred0.6, pch=20, cex = 0.5,
-     main = "Up to 0.6", xlab = "Unscaled", ylab = "Scaled")
-plot(df_pred$pred0.8, df_pred3$pred0.8, pch=20, cex = 0.5,
-     main = "Up to 0.8", xlab = "Unscaled", ylab = "Scaled")
-```
 
@@
 
-## GLMMadaptive
-
-- Here we can fit a model with random intercept and slope for time
+## GLMMadaptive linear time effect
+
+- Here we can fit a model with random intercept and slope for time. It is doable on 500 datasets, but obviously too simple for the data generation scheme
 
 $$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
-- But this model is obviously too simple for the data generation scheme
 
@@
 t1 <- Sys.time()
-fit_adglmm <- mixed_model(Y ~ t, random = ~ t | id, data =  train_df, family = binomial())
+fit_adglmm <- mixed_model(Y ~ t, random = ~ t | id, 
+                          data =  train_df, family = binomial())
 t2 <- Sys.time()
@@
 
+```{r}
+# out-of-sample prediction
+pred_lin_glmm <- list(
+  predict(fit_adglmm, 
+          newdata = test_df %>% filter(t<=0.2),
+          newdata2 =  test_df %>% filter(t>0.2),
+          type = "subject_specific", type_pred = "link",
+          return_newdata = TRUE)$newdata2,
+
+  predict(fit_adglmm, 
+          newdata = test_df %>% filter(t<=0.4),
+          newdata2 =  test_df %>% filter(t>0.4),
+          type = "subject_specific", type_pred = "link",
+          return_newdata = TRUE)$newdata2,
+
+  predict(fit_adglmm, 
+          newdata = test_df %>% filter(t<=0.6),
+          newdata2 =  test_df %>% filter(t>0.6),
+          type = "subject_specific", type_pred = "link",
+          return_newdata = TRUE)$newdata2,
+
+  predict(fit_adglmm, 
+          newdata = test_df %>% filter(t<=0.8),
+          newdata2 =  test_df %>% filter(t>0.8),
+          type = "subject_specific", type_pred = "link",
+          return_newdata = TRUE)$newdata2
+  )
+```
+
+```{r}
+pred_lin_glmm <- lapply(pred_lin_glmm, 
+                        function(df){df %>% select(id, t, pred)})
+
+pred_lin_glmm_df <- test_df %>%
+  left_join(pred_lin_glmm[[1]], by = c("id", "t")) %>%
+  rename(pred0.2 = pred) %>%
+  left_join(pred_lin_glmm[[2]], by = c("id", "t")) %>%
+  rename(pred0.4 = pred) %>%
+  left_join(pred_lin_glmm[[3]], by = c("id", "t")) %>%
+  rename(pred0.6 = pred) %>%
+  left_join(pred_lin_glmm[[4]], by = c("id", "t")) %>%
+  rename(pred0.8 = pred) 
+```
+
+```{r}
+p2 <- pred_lin_glmm_df %>%
+  filter(id %in% rand_test_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GLMMadaptive")
+```
+
+```{r, fig.height=5, fig.width=10}
+ggarrange(p1, p2, nrow = 1, common.legend = T)
+```
+
+```{r}
+# calcualte ISE
+err2 <- pred_lin_glmm_df %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err2$window = cut(err2$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err2 <- split(err2, f = err2$window)
+
+tb2 <- lapply(err2, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+options(knitr.kable.NA = "")
+
+ise <- full_join(tb1, tb2, by = "Window")
+colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
+ise %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"",
+        caption = "ISE") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+
+```
+
+```{r}
+# calcualte AUC
+auc_df2 <- pred_lin_glmm_df %>%
+  mutate(window = cut(t, breaks = seq(0, 1, by = 0.2), include.lowest = T)) %>%
+  select(Y, starts_with("pred"), window)
+
+auc_mat2 <- matrix(NA, length(pred_window), length(obs_track))
+
+for(j in seq_along(obs_track)){
+  for(i in (j+1):5){
+    this_Y <- auc_df2$Y[auc_df2$window==pred_window[i]]
+    this_pred <- auc_df2[auc_df2$window==pred_window[i],
+                         paste0("pred", obs_track[j])]
+    this_perf <- performance(prediction(this_pred, this_Y), measure = "auc")
+    auc_mat2[i ,j] <- this_perf@y.values[[1]]
+  }
+}
+
+```
+
+```{r}
+# calcualte AUC
+options(knitr.kable.NA = "")
+
+auc <- cbind(auc_mat, auc_mat2)
+rownames(auc) <- unique(auc_df1$window)
+colnames(auc) <-c(rep(seq(0.2, 0.8, by = 0.2), 2))
+auc %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"",
+        caption = "AUC") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+```
+
 ## Flexible GLMM
 
-- Fit on all 500 subjects take too long. I reduced the sample size to 10 
-- In fact, 10 samples took more than 30 minutes to fit the model.
-Therefore, I further reduce the number of observations for each subject.
-I'll start with only the 1/10 of the time track (100 observations), take observations at bin midpoints 5, 15, 25, ...., 995.
-This is to preserve the non-linearity of latent function when reducing grid density
-
-$$g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
+Due to computation time, I took only 10 subjects for training set. In this case, the model took 30 minutes to fit. Perhaps we could do a few more iterations?
+
+I use five basis functions for both fixed and random effects.
+
+$$g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^5\xi_{il}\phi_l(t)$$
 
@@
 # subset 10 subjects from training dataset 
-df_train_sub <- train_df %>% filter(id %in% 1:10 & sind==bin)
+df_train_sub <- train_df %>% filter(id %in% 1:10)
 df_train_sub$id <- droplevels(df_train_sub$id)
-# levels(df_train_sub$id)
-# intersect(t, unique(df_train_sub$t))
-# View(df_train_sub)
-```
-
-- Below I construct basis functions for the fixed effect/random intercept.
-Let's use five basis functions to start with.
-I believe this is far less flexible than the FPCA model, but it still takes a really long time to run.
-It is actually not realistic to repeat 500 times.
-
-PS 50 subjects took 1.06 hours to fit with the same model structure.
-
-```{r}
-# construct design matrix for the fixed effect intercept
-t_sub <- unique(df_train_sub$t)
-# fix_xmat <- smoothCon(s(t, bs="cr", k=5), data = data.frame(t=t_sub))
-fix_xmat <- ns(t_sub, df=5)
-# dim(fix_xmat[[1]]$X)
-
-# basis functions
-fix_xmat <- data.frame(t=t_sub, 
-                       fix_xmat)
-colnames(fix_xmat) <- c("t", paste0("B", 1:5))
-
-df_train_sub <- df_train_sub %>% left_join(fix_xmat, by = "t")
-```
-
-- Below I construct basis functions for the individual random effect (also a functinal intercept), with four basis functions to start with.
-
-```{r}
-# construct design matrix for the random effect
-rand_xmat <- smoothCon(s(t, bs="cr", k=4), data = data.frame(t=t_sub))
-# rand_xmat <- ns(t_sub, df = 4)
-
-# dim(rand_xmat[[1]]$X) # 5 basis functions for each of the 50 subjects
-rand_xmat <- data.frame(t = t, rand_xmat[[1]]$X)
-
-df_train_sub <- df_train_sub %>%
-  left_join(rand_xmat, by = c("t"))
-```
-
-- Now I'll fit the model and record the time spent on model fitting
-
-The model fitting took about 12 minutes at this data scale, which obviously is not realistic for 500 simulated datasets.
-
-Also tried increasing number of basis functions, which ended up with "computationally singular". In fact, this method does not seem very numerically stable.
-
-```{r, eval=TRUE, class.source = "fold-show"}
-# model fit
+
+df_test_sub <- test_df %>% filter(id %in% 1:10)
+df_test_sub$id <- droplevels(df_test_sub$id)
+# levels(df_test_sub$id)
+```
+
+```{r flex_adglmm, cache=T}
 t1 <- Sys.time()
-fit_adglmm <- mixed_model(fixed = Y ~ 0+B1+B2+B3+B4+B5, 
-                          random = ~X1+X2+X3+X4 | id,
+fit_adglmm_dense <- mixed_model(fixed = Y ~ 0+ns(t, df = 5), 
+                          random = ~ 0+ns(t, df = 5) | id, 
                           data =  df_train_sub, family = binomial())
@@
 # t2-t1
-
-```
-
-- But anyways, let's try to use this method for prediction. Let's do that for 10 new test subjects.
-
-```{r}
-df_test_sub <- test_df %>% filter(id %in% 1:10 & sind %in% mid)
-df_test_sub$id <- droplevels(df_test_sub$id)
-# levels(df_test_sub$id)
-
-# basis functions for both fixed and random effects should be shared across sample
-df_test_sub <- df_test_sub %>% left_join(fix_xmat, by = "t") %>%
-  left_join(rand_xmat, by = "t")
-
+```
+
+As we have talked about from last meeting, I wanna use ns() wrapped in the GLMMadaptive::mixed_model() to see if this could speed things up. The model fit took 
+`r round(t2-t1, 2)` mins to fit.
+
+```{r}
 # given observation from 0.2, 0.4, 0.6, 0.8, predict the rest
-# this step in fact is very fast
-pred_adglmm1 <- predict(fit_adglmm, 
+pred_nonlin_adglmm <- list(
+  predict(fit_adglmm_dense, 
           newdata = df_test_sub %>% filter(t<=0.2),
@@
           type = "subject_specific", type_pred = "link",
-                      return_newdata = TRUE)
-
-pred_adglmm2 <- predict(fit_adglmm, 
+          return_newdata = TRUE)$newdata2,
+  predict(fit_adglmm_dense, 
           newdata = df_test_sub %>% filter(t<=0.4),
@@
           type = "subject_specific", type_pred = "link",
-                      return_newdata = TRUE)
-
-pred_adglmm3 <- predict(fit_adglmm, 
+          return_newdata = TRUE)$newdata2,
+  predict(fit_adglmm_dense, 
           newdata = df_test_sub %>% filter(t<=0.6),
@@
           type = "subject_specific", type_pred = "link",
-                      return_newdata = TRUE)
-
-pred_adglmm4 <- predict(fit_adglmm, 
+          return_newdata = TRUE)$newdata2,
+  predict(fit_adglmm_dense, 
           newdata = df_test_sub %>% filter(t<=0.8),
@@
           type = "subject_specific", type_pred = "link",
-                      return_newdata = TRUE)
-```
-
-```{r}
-pred_adglmm_all <- df_test_sub %>% 
-  left_join(pred_adglmm1$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>% 
-  left_join(pred_adglmm2$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>%
-  left_join(pred_adglmm3$newdata2 %>% select(id, t, pred), by = c("id", "t")) %>%
-  left_join(pred_adglmm4$newdata2 %>% select(id, t, pred), by = c("id", "t"))
-
-colnames(pred_adglmm_all) <- c(colnames(pred_adglmm_all)[1:14], "pred1", "pred2", "pred3", "pred4")
-```
-
-```{r}
-
-pred_adglmm_all %>%
+          return_newdata = TRUE)$newdata2
+  )
+```
+
+```{r}
+pred_nonlin_adglmm <- lapply(pred_nonlin_adglmm, 
+                        function(df){df %>% select(id, t, pred)})
+
+pred_nonlin_adglmm_df <- df_test_sub %>%
+  left_join(pred_nonlin_adglmm[[1]], by = c("id", "t")) %>%
+  rename(pred0.2 = pred) %>%
+  left_join(pred_nonlin_adglmm[[2]], by = c("id", "t")) %>%
+  rename(pred0.4 = pred) %>%
+  left_join(pred_nonlin_adglmm[[3]], by = c("id", "t")) %>%
+  rename(pred0.6 = pred) %>%
+  left_join(pred_nonlin_adglmm[[4]], by = c("id", "t")) %>%
+  rename(pred0.8 = pred)
+
+# pred_nonlin_adglmm_df %>% filter(t>0.4)
+```
+
+```{r}
+p1_nonlin <- pred_nonlin_adglmm_df %>%
   filter(id %in% 1:4) %>%
-  mutate_at(vars(eta_i, pred1, pred2, pred3, pred4), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8),
+            function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
@@
   geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred1, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred2, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred3, col = "0.5"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred4, col = "0.8"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.5"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
   facet_wrap(~id)+
@@
 
-# Notes
-
-Perhaps I should try to verify my thoughts about laplace approximation by different generation scheme. For example, what if the true eigenvalues are on larger scale?
-
-- Well, after increasing the scale of the true eigenvalues, the convergence rate did get closer between scaled and unscaled parameter. Also reversed as I expected.
-- But on second though, A more effecient way is to calcualted ISE and AUC stratified by convergence status
+```{r}
+# calcualte ISE
+err3 <- pred_nonlin_adglmm_df %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err3$window = cut(err3$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err3 <- split(err3, f = err3$window)
+
+tb3 <- lapply(err3, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+For the comparing fGFPCA, if we decrease the measurement density to 100 points, a lot of numeric problems popped up in the either local models or Laplace Approximation process.
+It looks like when observation track is short, GLMMadaptive has a better prediction.
+
+```{r fGFPCA}
+# bin data
+# J_sub <- length(unique(df_train_sub$t))
+# df_train_sub$sind <- rep(1:J_sub, 10)
+# t_sub <- unique(df_train_sub$t)
+# bin_w <- 4
+
+# n_bin_sub <- J_sub/bin_w # number of bins
+# brks <- seq(0, J_sub, by = bin_w) # cutoff points on time index
+# mid <- (brks+bin_w/2)[1:n_bin_sub] # interval mid point time index
+# mid_t <- t_sub[mid] 
+# 
+# df_train_sub$bin <- cut(df_train_sub$sind,
+#                         breaks = brks, include.lowest = T, labels = mid)
+# df_train_sub$bin <- as.numeric(as.character(df_train_sub$bin))
+```
+
+```{r}
+# fit fGFPCA model on the training set
+t1 <- Sys.time()
+
+## Step 2: local GLMM and estimate latent function
+## on training set
+bin_lst_sub <- split(df_train_sub, f = df_train_sub$bin)
+df_est_latent_sub <- lapply(bin_lst_sub, function(x){pred_latent(x, n_node = 0)}) 
+# boundary hitting in this process
+df_est_latent_sub <- bind_rows(df_est_latent_sub) 
+# range(df_est_latent_sub$eta_hat)
+```
+
+```{r}
+## Step 3: FPCA
+uni_eta_hat_sub <- df_est_latent_sub %>% filter(bin==sind)
+mat_est_unique <- matrix(uni_eta_hat_sub$eta_hat,
+                         nrow=10, 
+                         ncol=n_bin, byrow = F) 
+fpca_mod_sub <- fpca.face(mat_est_unique, argvals = mid_t, var=T)
+```
+
+```{r}
+## Step 4: project and debias
+## Projection
+# values used for projection
+p <- 3 # order of b splines 
+knots <- 35 # number of knots (same from FPCA model)
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
+B <- spline.des(knots = knots_values, x = mid_t, ord = p + 1,
+                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
+Bnew <- spline.des(knots = knots_values, x = t, ord = p + 1,
+                   outer.ok = TRUE)$design  # evaluate B-splines on original grid
+df_phi <- matrix(NA, J, K) 
+for(k in 1:K){
+  lm_mod <- lm(fpca_mod_sub$efunctions[,k] ~ B-1)
+  df_phi[,k] <- Bnew %*% coef(lm_mod)
+}# project binned eigenfunctions onto the original grid
+```
+
+```{r}
+## debias
+df_phi <- data.frame(t = t, df_phi)
+colnames(df_phi) <- c("t", paste0("phi", 1:4))
+df_train_sub <- df_train_sub %>% 
+  select(-starts_with("phi")) %>%
+  left_join(df_phi, by = "t")
+train_df$id <- as.factor(train_df$id)
+debias_glmm <- bam(Y ~ s(t, bs="cc", k=10)+
+                     s(id, by=phi1, bs="re")+
+                     s(id, by=phi2, bs="re")+
+                     s(id, by=phi3, bs="re")+
+                     s(id, by=phi4, bs="re"), 
+                   family = binomial, data=df_train_sub, 
+                   method = "fREML",
+                   discrete = TRUE)
+new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean
+new_lambda <- 1/debias_glmm$sp[2:5] # extract re-evaluated lambda
+# rescale
+new_phi <- df_phi %>% select(starts_with("phi"))*sqrt(n_bin)
+new_phi <- as.matrix(new_phi)
+new_lambda <- new_lambda/n_bin
+t2 <- Sys.time()
+
+```
+
+```{r, results='hide'}
+df_pred_sub<- df_test_sub
+converge_st <- matrix(NA, 10, 4)
+
+for(i in 1:10){
+    df_i <- df_test_sub %>% filter(id==i)
+
+    # per max obs time
+    for(tmax in c(0.2, 0.4, 0.6, 0.8)){
+      df_it <- df_i %>% filter(t <= tmax)
+      max_rid <- nrow(df_it)
+
+      # into a list
+      MyData <- list(K=K, 
+                     X=new_phi[1:max_rid, ], 
+                     mon.names=mon.names,
+                     parm.names=parm.names, 
+                     pos.xi=pos.xi, 
+                     y=df_it$Y, 
+                     tao=diag(new_lambda), f0=new_mu[1:max_rid])
+
+      # fit laplace approximation
+      Fit <- LaplaceApproximation(Model, parm = rep(0, K), 
+                                  Data=MyData, Method = "NM", Iterations = 1000,
+                                  CovEst = "Identity")
+      converge_st[i, which(c(0.2, 0.4, 0.6, 0.8)==tmax)] <- Fit$Converged
+      score <- Fit$Summary1[, "Mode"]
+
+      # prediction
+      eta_pred_out <- new_mu+new_phi%*%score
+      df_pred_sub[df_pred_sub$id==i, paste0("pred", tmax)] <- eta_pred_out[,1]
+    }
+}
+
+```
+
+```{r}
+df_pred_sub$pred0.2[df_pred_sub$t<=0.2] <- NA
+df_pred_sub$pred0.4[df_pred_sub$t<=0.4] <- NA
+df_pred_sub$pred0.6[df_pred_sub$t<=0.6] <- NA
+df_pred_sub$pred0.8[df_pred_sub$t<=0.8] <- NA
+
+p2_fgfpca <- df_pred_sub %>%
+  filter(id %in% 1:4) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8),
+            function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.5"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "fGFPCA")
+
+```
+
+```{r}
+# calcualte ISE
+err4 <- df_pred_sub %>%
+  mutate(err1 = (pred0.2-eta_i)^2,
+         err2 = (pred0.4-eta_i)^2,
+         err3 = (pred0.6-eta_i)^2,
+         err4 = (pred0.8-eta_i)^2) %>%
+  select(id, t, starts_with("err"))
+
+err4$window = cut(err4$t, breaks = seq(0, 1, by = 0.2), include.lowest = T)
+# table(err1$window)
+err4 <- split(err4, f = err4$window)
+
+tb4 <- lapply(err4, function(x){
+  x %>%  group_by(id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>%
+    summarise_at(vars(err1, err2, err3, err4), mean)
+}) %>% bind_rows(.id = "Window")
+```
+
+```{r}
+# calcualte ISE
+options(knitr.kable.NA = "")
+
+ise <- full_join(tb4, tb3, by = "Window")
+colnames(ise) <-c("Window",  rep(seq(0.2, 0.8, by =0.2), 2))
+ise %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"",
+        caption = "ISE") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "= 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "=1, "Maximum observation time" = 8))
+```
+
+```{r, fig.height=5, fig.width=10}
+ggarrange(p1_nonlin, p2_fgfpca, nrow = 1, common.legend = T)
+```
+
+## Random thouds

### 2023-11-27 Docs/manuscript.Rmd
simulation result summary

--- 2023-05-17 Docs/manuscript.Rmd
@@
 knitr::opts_chunk$set(echo = FALSE)
+options(knitr.kable.NA = "")
+
+set.seed(1120)
+
+library(here)
+library(tidyverse)
+library(ggpubr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(kableExtra)
+library(knitr)
+library(mvtnorm)
+library(mgcv)
+library(splines)
+theme_set(theme_minimal())
+
+load(here("Data/SimOutput_fGFPCA.RData"))
+load(here("Data/SimOutput_GLMMadaptive.RData"))
+
+## prediction window 
+window <- seq(0, 1, by = 0.2)
+M <- length(pred_list_all)
 ```
 
-## R Markdown
+# Larger-scale simulation
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+## Figure
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+```{r fig_pred}
+rand_id <- sample(501:600, size = 4)
+# range(pred_list_all[[1]]$pred0.2, na.rm = T)
+fig_fgfpca <- pred_list_all[[1]] %>%
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "fGFPCA", col = "Maximum observation time")
 
-```{r cars}
-summary(cars)
+fig_adglmm <- pred_list_ref[[1]] %>%
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GLMMadaptive", col = "Maximum observation time")
+
 ```
 
-## Including Plots
+```{r}
+ggarrange(fig_fgfpca, fig_adglmm,
+          nrow = 1, common.legend = T)
+```
+## ISE
 
-You can also embed plots, for example:
+```{r}
+## ISE container 
+ise_mat <- ise_mat_ref <- array(NA, 
+                                dim = c(length(window)-2, length(window)-2, M))
+# dims: prediction window, max obs time, simulation iter
 
-```{r pressure, echo=FALSE}
-plot(pressure)
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+```{r ise_fGFPCA}
+## calculation
+for(m in 1:M){
+  this_df <- pred_list_all[[m]]
+  ise_tb <- pred_list_all[[m]] %>%
+    mutate(err1 = (pred0.2-eta_i)^2,
+           err2 = (pred0.4-eta_i)^2,
+           err3 = (pred0.6-eta_i)^2,
+           err4 = (pred0.8-eta_i)^2) %>%
+    select(id, t, starts_with("err")) %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("err"))
+  ise_mat[, ,m] <- as.matrix(ise_tb)
+
+}
+
+mean_ise <- apply(ise_mat, c(1, 2), mean)
+
+# mean_ise <- data.frame(mean_ise) %>%
+#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+#          .before = 1)
+colnames(mean_ise) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_GLMMadaptive}
+## calculation
+for(m in 1:M){
+  this_df <- pred_list_ref[[m]]
+  ise_tb <- pred_list_ref[[m]] %>%
+    mutate(err1 = (pred0.2-eta_i)^2,
+           err2 = (pred0.4-eta_i)^2,
+           err3 = (pred0.6-eta_i)^2,
+           err4 = (pred0.8-eta_i)^2) %>%
+    select(id, t, starts_with("err")) %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("err"))
+  ise_mat_ref[, ,m] <- as.matrix(ise_tb)
+
+}
+
+mean_ise_ref <- apply(ise_mat_ref, c(1, 2), mean)
+# mean_ise_ref <- data.frame(mean_ise_ref) %>% 
+#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+#          .before = 1)
+colnames(mean_ise_ref) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r}
+data.frame(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+           mean_ise, mean_ise_ref, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 8))
+```
+
+## AUC
+
+```{r auc_func}
+## a function to calculate AUC
+get_auc <- function(y, pred){
+  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
+    auc <- NA
+  }
+  else{
+    this_perf <- performance(prediction(pred, y), measure = "auc")
+    auc <- this_perf@y.values[[1]]
+  }
+  return(auc)
+}
+
+```
+
+```{r}
+## auc container 
+auc_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+auc_mat_ref <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+```
+
+```{r auc_fGFPCA}
+for(m in 1:M){
+  this_df <- pred_list_all[[m]]
+  auc_tb <- this_df %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred0.2),
+              auc2 = get_auc(Y, pred0.4),
+              auc3 = get_auc(Y, pred0.6),
+              auc4 = get_auc(Y, pred0.8)) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("auc"))
+  auc_mat[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc <- apply(auc_mat, c(1, 2), mean)
+# mean_auc <- data.frame(mean_auc) %>% 
+#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+#          .before = 1)
+colnames(mean_auc) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r auc_GLMMadaptive}
+for(m in 1:M){
+  this_df <- pred_list_ref[[m]]
+  auc_tb <- this_df %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred0.2),
+              auc2 = get_auc(Y, pred0.4),
+              auc3 = get_auc(Y, pred0.6),
+              auc4 = get_auc(Y, pred0.8)) %>%
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("auc"))
+  auc_mat_ref[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc_ref <- apply(auc_mat_ref, c(1, 2), mean)
+# mean_auc <- data.frame(mean_auc) %>% 
+#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+#          .before = 1)
+colnames(mean_auc_ref) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r}
+data.frame(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
+           mean_auc, mean_auc_ref, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area under the ROC curve") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 8))
+```
+
+```{r time}
+data.frame(
+  "Method" = c("fGFPCA", "GLMMadaptive"),
+  "Fit" = c(mean(fit_time)/60, mean(fit_time_ref)),
+           "Prediction"= c(mean(pred_time), mean(pred_time_ref)/60)) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Computation time (minutes)") %>%
+  kable_styling(full_width = F) 
+```
+
+I think we could say that while the total time spend on model fitting + prediction are similar between two methods, fGFPCA achieved much better flexibility and much better predictive performance of prediction under every scenario.
+
+# Small-scale simulation
+
+# NHANES data application

### 2023-11-27 Code/Sim_GLMMadaptive.R

@@ -1,75 +1,211 @@
+# This script implements the full simulation 
+# for dynamic prediction using GLMMadaptive
+# as a completing method 
+# on the simulated dataset generated from Code/GenData.R
 
-#### GLMMadaptive ####
+#### set up ####
 
-pred_lst_ref <- list()
-runtime_ref <- rep(NA, M)
+set.seed(1114)
 
+library(here)
+library(tidyverse)
+library(ggpubr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(kableExtra)
+library(knitr)
+library(mvtnorm)
+library(mgcv)
+library(splines)
+library(LaplacesDemon)
+theme_set(theme_minimal())
+
+# load simulated data set. From Code/GenData.R
+load(here("Data/sim_data.RData"))
+
+#### Data overview ####
+
+N <- length(unique(sim_data[[1]]$id))
+J <- length(unique(sim_data[[1]]$t))
+t <- unique(sim_data[[1]]$t)
+M <- length(sim_data)
+K <- 4 # number of eigenfunctions to use
+
+# overview of a few samples
+rand_id <- sample(N, size = 4)
+sim_data[[357]] %>% filter(id %in% rand_id) %>% 
+  ggplot()+
+  geom_point(aes(x=t, y=Y))+
+  geom_line(aes(x=t, y=plogis(eta_i)), col = "red")+
+  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
+  facet_wrap(~id)
+
+
+#### GLMMadaptive model estimation ####
+
+# split data into train and test set
+N_train <- 500
+N_test <- 100
+
+# containers
+pred_list_ref <- list()
+fit_time_ref <- pred_time_ref <- rep(NA, M)
 
 
 pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 
 for(m in 1:M){
+  
+  # data plit
   df <- sim_data[[m]]
+  train_df <- df %>% filter(id %in% 1:N_train)
+  test_df <- df %>% filter(!id %in% 1:N_train)
   
   t1 <- Sys.time()
-  # GLMM adaptive model
-  adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df, family = binomial())
-  
-  # container
-  df_est_latent <- df
[…]

### 2023-11-28 Docs/ProgressReportSim.Rmd
Add method to progress report

@@
 # subset 10 subjects from training dataset 
-df_train_sub <- train_df %>% filter(id %in% 1:10)
+df_train_sub <- train_df %>% filter(id %in% 1:10) %>%
+  filter(t >= 0.375 & t <= 0.708)
 df_train_sub$id <- droplevels(df_train_sub$id)
 
-df_test_sub <- test_df %>% filter(id %in% 1:10)
+df_test_sub <- test_df %>% filter(id %in% 1:10) %>% 
+  filter(t >= 0.375 & t <= 0.708)
 df_test_sub$id <- droplevels(df_test_sub$id)
@@
 t2 <- Sys.time()
-# t2-t1
+t2-t1
 ```
@@
 
+I wonder if we should decrease the obervation track instead of decrease density. Let's say we only think about 9am-5pm (t from 0.375 to 0.208).
+
 ```{r fGFPCA}

### 2023-11-28 Docs/manuscript.Rmd
Add method to progress report

@@
 ---
-title: "Tables and Figures to put in manuscript"
+title: "Manuscript progress report"
 author: "Ying Jin"
@@
 
+# Method
+
+## Assumptions
+
+- For each subject i in the population, a generalized outcome $Y_i(t)$ is generated along a variable t (for example, time), where $t \in (0, T)$. 
+- The outcome, at any specific t, follows an exponential family distribution characterized by a (latent) continuous function $\eta_i(t)$:
+
+$$g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)$$
+
+$$p(Y_i(t)) = h(Y_i(t))\frac{exp(\eta_i(t)T[Y_i(t)])}{A(\eta_i(t))}$$
+
+- The continuous latent function consists of a population-level fixed process and an individual-level random process
+
+$$\eta_i(t) = \beta_0(t)+b_i(t)$$
+
+## Observed data
+
+In practice we would observe the discrete realization of $\{Y_i(t), t\}$ along a dense grid.
+For simplicity, we assume the observation grid is regular (same across sample).
+When we have J observations points in $(0, T]$, then for the jth observation point, we denote the corresponding value of t as $t_j$, and the corresponding outcome at this point $Y_i(t_j)$.
+
+## fGFPCA Algorithm
+
+### Bin data:
+
+Choose a proper bin width $w$ considering model complexity and identifiability. For now let's say the bins are equal-length and non-overlapping.
+
+- Bin index $s = 1...S$
+- Index of bin midpoints $m_s$
+- Value of t corresponding to bin midpoints $t_{m_s}$
+- Bin endpoints: $(t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]$
+
+```{r, eval=FALSE}
+grid_exp <- data.frame(t = seq(0, 1, by = 0.01), y = 0)
+ggplot(grid_exp, aes(x=t, y=y))+
+  geom_point(size = 0.5)
+```
+
+### Local GLMMs
+
+At the every bin, we fit a local intercept-only model:
+
+$$g[E(Y_i(t_j))] =\eta_i(t_{m_s})= \beta_0(t_{m_s})+b_i(t_{m_s})$$
+where $t_j \in (t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]$.
+
+Here we are basically saying that the value of latent function is constant within the same bin, which clearly is a misspecification of the true latent process.
+
+From the model above. we will be able to estimate a $\hat{\eta_i}(t_{m_s})$ on the binned grid for every individual in the training sample.
+
+### FPCA
+
+Here, we fit a FPCA model on the $\hat{\eta_i}(t_{m_s})$ obtained from step 2:
+
+$$\hat{\eta}_i(t_{m_s}) = f_0(t_{m_s})+\sum_{k=1}^K\xi_{ik}\phi_{k}(t_{m_s})+\epsilon_i(t_{m_s})$$
+
+where $\xi_{ik}$ independently follows normal distribution $N(0, \lambda_k)$, and $\epsilon_i(t_{m_s})$ at each point follows $N(0, \sigma_2)$.
+
+From this model, we will be able to obtain the following estimates which are shared across population:
+
+- Population mean $\hat{f_0}(t_{m_s})$
+- Basis functions $\hat{\mathbf{\Phi}} = \{\hat{\phi}_1(t_{m_s}), ...,\hat{\phi}_K(t_{m_s}))\}$
+- Estimates of variance of scores $\hat{\lambda}_1...\hat{\lambda}_K$
+
+### Projection and Debias
+
+The mean and basis functions are evaluated on the binned grid.
+To extend it to the original measurement grid data was collected on, we project the estimated eigenfunctions $\hat{\mathbf{\Phi}}$ back use spline basis.
+Now we have extend the $\hat{\phi}_k(t_{m_s})$ to the original grid $\hat{\phi}_k(t_j)$
+
+Because of the misspecification of local GLMMs, the estimated eigenfunctions and eigenvalues are also biased by a constant multiplicative effect.
+Therefore, we use a GLMM to re-evaluate the mean function, eigenfunctions and eigenvalues.
+
+## Out-of-sample prediction
+
+Now, let's assume we have a new subject $u$ with $J_u$ observations ($J_u < J$). Then the log-likelihood of this new subject would be:
+
+$$l_u=\sum_{t_j<t_{J_u}}log(h(Y_u(t_j)))+\hat{\eta}_u(t_j)T(Y_u(t_j))-log(A[\hat{\eta}_u(t_j)])$$
+
+where $\hat{\eta}_u(t_j) = \hat{f}_0(t_j)+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_j)$.
+
+With estimates for the population-level parameters from fGFPCA algorithms above, we can estimate $\xi_{uk}$ by maximization of $l_u$.
+Direct maximization some times does not have closed form solution.
+Numeric maximization methods seem not very stable as well.
+So I have decided to used a Bayes approach (Laplace Approximation):
+
+- Prior distribution: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$
+- Posterior distribution: the likelihood of $l_u =l(Y_u(t_j)|\mathbf{\xi}_u)$
+
+Laplace Approximation would get the posterior mode of $\xi_{uk}$ through quadratic approximation.
+
 # Larger-scale simulation
+
+## Simulation set up
+
+Here we simulate binary data from cyclic latent process:
+
+\[\begin{aligned}
+Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
+\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
+\end{aligned}\]
+
+where:
+
+- $t$ is 1000 equal-spaced observations points on $[0, 1]$ (J = 1000).
+- $f_0(t)=0$
+- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively. 
+- Sample size $N = 500$
+- In the binning step, we bin every 10 observations
+- 500 simulations were implemented
 
@@
 ```
+
 ## ISE
@@
 
-# Small-scale simulation
+<!-- # Small-scale simulation  -->
 
-# NHANES data application
+<!-- # NHANES data application -->

### 2023-11-30 Docs/manuscript.Rmd
small-scale simulation

@@
 
-$$p(Y_i(t)) = h(Y_i(t))\frac{exp(\eta_i(t)T[Y_i(t)])}{A(\eta_i(t))}$$
+$$p(Y_i(t)) = h(Y_i(t))exp\{\eta_i(t)T[Y_i(t)]-A(\eta_i(t))\}$$
 
@@
 - 500 simulations were implemented
+
+## Reference method
+
+- GLMMadaptive
+- Here we can fit a model with random intercept and slope for time. It is doable on 500 datasets, but obviously too simple for the data generation scheme. We would expect it to perform terribly.
+
+$$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
 
@@
 
-<!-- # Small-scale simulation  -->
+# Small-scale simulation: one iteration
+
+Here we would like to fit fGFPCA and GLMMadaptive on a dataset with smaller sample size and/or smaller measurement density.
+For the GLMMadaptive model, we would set it up with spline basis functions so that its flexibility is comparable with fGFPCA model, such as:
+
+$$g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^5\xi_{il}\phi_l(t)$$
+
+I did the following things for data-reduction
+
+- Reduce sample size to 10 (for both training and testing)
+- Reduce measurement track by half (0.25-0.75)
+- Another way is the simple reduce the measurement density.
+For example, instead of every minute, we can use every 10 minutes.
+In this case, the binning step in fGFPCA would bin every 100 minutes in to one bin.
+- - When we have smaller density of measurement grid, both methods could run into numeric issues in both methods.
+For example, singularity happened the the local GLMM step of fGFPCA procedure.
+This might be able to get resoved by changing the optimizer (e.g.
+Nealder_Mead).
+Do we want to do that?
+Or perhaps we simply want to use the full track?
+
+- We may also need to reduce number of simulations (GLMMadaptive took 20-ish minutes for one simulaiton).
+
+Below are some of the outcomes:
+
+```{r load_subset_sim_output}
+load(here("Data/SubSimOutput_GLMMadaptive.RData"))
+load(here("Data/SubSimOutput_fGFPCA.RData"))
+```
+
+## Figure
+
+```{r fig_pred_subset}
+fig_fgfpca_subset <- df_pred_m_fGFPCA %>% filter(id %in% 501:504) %>%
+  mutate_at(vars(eta_i, pred0.35, pred0.45, pred0.55, pred0.65), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.35, col = "0.35"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.45, col = "0.45"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.55, col = "0.55"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.65, col = "0.65"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "fGFPCA", col = "Maximum observation time")
+
+fig_adglmm_subset <- df_pred_m_adglmm %>% filter(id %in% 501:504) %>%
+  mutate_at(vars(eta_i, pred0.35, pred0.45, pred0.55, pred0.65), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.35, col = "0.35"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.45, col = "0.45"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.55, col = "0.55"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.65, col = "0.65"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GLMMadaptive", col = "Maximum observation time")
+
+```
+
+```{r}
+ggarrange(fig_fgfpca_subset, fig_adglmm_subset,
+          nrow = 1, common.legend = T)
+```
+
+## ISE
+
+```{r ise_fGFPCA_subset}
+window2 <- seq(0.25, 0.75, by = 0.1)
+ise_mat_subset <- df_pred_m_fGFPCA %>%
+    mutate(err1 = (pred0.35-eta_i)^2,
+           err2 = (pred0.45-eta_i)^2,
+           err3 = (pred0.55-eta_i)^2,
+           err4 = (pred0.65-eta_i)^2) %>%
+    select(id, t, starts_with("err")) %>% 
+    mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != "[0.25,0.35]") %>% 
+    select(starts_with("err"))
+colnames(ise_mat_subset) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+```
+
+```{r ise_GLMMadaptive_subset}
+ise_mat_subset2 <- df_pred_m_adglmm %>%
+    mutate(err1 = (pred0.35-eta_i)^2,
+           err2 = (pred0.45-eta_i)^2,
+           err3 = (pred0.55-eta_i)^2,
+           err4 = (pred0.65-eta_i)^2) %>%
+    select(id, t, starts_with("err")) %>% 
+    mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != "[0.25,0.35]") %>% 
+    select(starts_with("err"))
+colnames(ise_mat_subset2) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+```
+
+```{r}
+data.frame(Window = c("(0.35,0.45]", "(0.45,0.55]", "(0.55,0.65]", "(0.65, 0.75]"),
+           ise_mat_subset, ise_mat_subset2, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "= 1, "Observation track" = 8))
+```
+
+## AUC
+
+```{r auc_fGFPCA_subset}
+auc_mat_subset <- df_pred_m_fGFPCA %>%
+  mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred0.35),
+            auc2 = get_auc(Y, pred0.45),
+            auc3 = get_auc(Y, pred0.55),
+            auc4 = get_auc(Y, pred0.65)) %>%
+  filter(window != "[0.25,0.35]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_subset) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+```
+
+```{r auc_GLMMadaptive_subset}
+auc_mat_subset2 <- df_pred_m_adglmm %>%
+  mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred0.35),
+            auc2 = get_auc(Y, pred0.45),
+            auc3 = get_auc(Y, pred0.55),
+            auc4 = get_auc(Y, pred0.65)) %>%
+  filter(window != "[0.25,0.35]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_subset2) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+```
+
+```{r}
+data.frame(Window = c("(0.35,0.45]", "(0.45,0.55]", "(0.55,0.65]", "(0.65, 0.75]"),
+           auc_mat_subset, auc_mat_subset2, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" "= 1, "Observation track" = 8))
+```
+
+There is huge difference in compuation time. GLMMadaptive took 23 minutes, while fGFPCA took less than 3 seconds.
 

### 2023-11-30 Code/Sim_Subset.R

@@ -1,86 +1,0 @@
-set.seed(314)
-
-library(here)
-library(tidyverse)
-# library(ggplot2)
-# theme_set(theme_minimal())
-# library(fcr)
-library(refund)
-library(lme4)
-library(ROCR)
-library(GLMMadaptive)
-library(mgcv)
-# library(ggpubr)
-# library(gridExtra)
-# library(kableExtra)
-
-
-
-
-#### load simulated data ####
-
-load(here("Data/sim_data.RData"))
-
-
-#### simulation set up ####
-
-J <- 1000 # number of observations points
-N <- 500 # sample size
-M <- 50 
-
-
-#### GLMM adaptive ####
-
-# cubic spline basis functions
-df_basis <- smoothCon(s(sind, bs = "cr", k = 20), data=sim_data[[1]])[[1]]$X
-colnames(df_basis) <- paste("B", 1:20, sep="")
-this_df <- data.frame(sim_data[[1]] %>% select(id, Y), df_basis)
-fix_f <- paste("Y ~", 
-                paste("B", 1:20, sep = "", collapse = " + "))
-fix_f <- as.formula(fix_f)
-rand_f <- paste
-
-adglmm <- mixed_model(this_f, random = ~ 1 | id, data =  this_df, family = binomial())
-
-
-for(m in 1:M){
-  df <- sim_data[[m]]
-  
-  t1 <- Sys.time()
-  # GLMM adaptive model
-  adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df, family = binomial())
-  
-  # container
-  df_est_latent <- df
-  df_est_latent[, 'pred_t200'] <- df_est_latent[, 'pred_t400'] <- df_est_latent[, 'pred_t600'] <- df_est_latent[, 'pred_t800'] <- NA 
-  
-  # time up to t
-  pred_t200 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 200),
-                       newdata2 =  df %>% filter(sind_inx > 200), type = "subject_specific", type_pred = "link",
-                       se.fit = TRUE, return_newdata = TRUE)
-  df_est_latent$pred_t200[df_est_latent$sind_inx>200] <- pred_t200$newdata2$pred
-  
-  pred_t400 <- predict(adglmm, newdata = df %>% filter(sind_inx <= 400),
-                       newdata2 =  df %>% filter(sind_inx > 400), type = "subject_specific", type_pred = "link",
[…]

### 2023-11-30 Code/Sim_Subset_fGFPCA.R

@@ -1,0 +1,202 @@
+set.seed(1114)
+
+library(here)
+library(tidyverse)
+library(ggplot2)
+theme_set(theme_minimal())
+library(fcr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+library(ggpubr)
+library(gridExtra)
+library(kableExtra)
+library(LaplacesDemon)
+
+
+
+
+#### load simulated data ####
+
+load(here("Data/sim_data.RData"))
+
+# reduce data
+## training data
+sub_sim_data <- sim_data[1:100] 
+M <- 100 # number of simulations
+
+sub_sim_data <- lapply(sub_sim_data, function(x){x %>% filter(id %in% c(1:10,501:510))})
+Ntr <- Nte <- 10 # trainig and testing sample size
+
+sub_sim_data <- lapply(sub_sim_data, function(x){x %>% filter(t > 0.25 & t <= 0.75)})
+J <- length(unique(sub_sim_data[[1]]$t))
+
+lapply(sub_sim_data, dim)
+head(sub_sim_data[[1]])
+
+#### One iteration example ####
+
+df_train_m <- sub_sim_data[[1]] %>% filter(id %in% 1:10)
+df_test_m <- sub_sim_data[[1]] %>% filter(!id %in% 1:10)
+
+
+
+#### fGFPCA ####
+
+# functions
+K <- 4
+source(here("Code/GLMM-FPCA.R")) 
+source(here("Code/OutSampBayes.R"))
+
+
+# binning 
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(min(df_train_m$sind)-1, max(df_train_m$sind), by = bin_w) # cutoff points on time index
+mid <- (brks+bin_w/2)[1:n_bin] # interval mid point time index
+t <- unique(df_train_m$t)
+mid_t <- unique(df_train_m$t[df_train_m$sind %in% mid]) # actually time corresponding to bin mid point
+
+# values used for projection
+p <- 3 # order of b splines 
+knots <- 35 # number of knots (same from FPCA model)
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
+
+
+# result container
+# M <- 10
+M <- 1
+# pred_list_all <- list()
+# converge_state_list <- list()
[…]

### 2023-11-30 Docs/manuscript.Rmd
meeting notes

@@
   html_document:
-    self_contained: yes
+    self_contained: no
     number_sections: true
@@
     font: 12pt
+    keep_tex: true
 ---
@@
 
+```{r fig_sim_data}
+rand_id <- sample(501:600, size = 4)
+pred_list_all[[1]] %>% filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  facet_wrap(~id)+
+  labs(title = "Simulated data")
+```
+
 ## Reference method
@@
 ```{r fig_pred}
-rand_id <- sample(501:600, size = 4)
 # range(pred_list_all[[1]]$pred0.2, na.rm = T)

### 2023-11-30 Code/Sim_Subset_fGFPCA.R

@@ -133,6 +133,7 @@ new_phi <- as.matrix(new_phi)
 
 new_lambda <- new_lambda/n_bin
 t2 <- Sys.time()
+t2-t1 # less than 1 second
 
 
 #### Prediction ####

### 2023-12-01 Docs/ProgressReportAppl.Rmd
data application

@@
 
-set.seed(516)
+set.seed(121)
 

### 2023-12-01 Docs/manuscript.Rmd
data application

@@
 
-<!-- # NHANES data application -->
+# NHANES data application
+
+We take 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction.
+
+In the GLMMadaptive model, I could only fit a model with random intercept. It took more than 20 minutes for model fitting.
+
+```{r load_nhanes}
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+load(here("Data/ApplOutput.RData"))
+```
+
+```{r nhance example}
+df_nhanes %>% select(SEQN, Z, sind) %>% 
+  mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
+  mutate(SEQN = factor(SEQN, labels = 1:8763)) %>%
+  ggplot()+
+  geom_tile(aes(x=sind, y = SEQN, fill = Z))+
+  labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
+  theme(axis.text.y = element_blank())
+
+```
+
+## AUC
+
+```{r auc_fGFPCA_nhanes}
+window3 <- seq(0, 1440, by = 240)
+
+auc_mat_nhanes <- nhanes_pred_fgfpca %>%
+  mutate(window = cut(sind, breaks = window3, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred240),
+            auc2 = get_auc(Y, pred480),
+            auc3 = get_auc(Y, pred720),
+            auc4 = get_auc(Y, pred960),
+            auc5 = get_auc(Y, pred1200)) %>%
+  filter(window != "[0,240]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_nhanes) <- c("4am", "8am", "12pm", "4pm", "8pm")
+```
+
+```{r auc_GLMMadaptive_nhanes}
+auc_mat_nhanes2 <- nhanes_pred_adglmm %>%
+  mutate(sind=t*1440) %>%
+  mutate(window = cut(sind, breaks = window3, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred240),
+            auc2 = get_auc(Y, pred480),
+            auc3 = get_auc(Y, pred720),
+            auc4 = get_auc(Y, pred960),
+            auc5 = get_auc(Y, pred1200)) %>%
+  filter(window != "[0,240]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_nhanes2) <- c("4am", "8am", "12pm", "4pm", "8pm")
+```
+
+```{r}
+data.frame(Window = c("4am-8am", "8am-12pm", "12pm-4pm", "4pm-8pm",
+                      "8pm-12am"),
+           auc_mat_nhanes, auc_mat_nhanes2, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 5, "GLMMadaptive" = 5)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 10))
+```

### 2023-12-01 Code/Data_Appl_NHANES.R

@@ -1,0 +1,291 @@
+# data application of NHANES
+
+set.seed(121)
+
+library(here)
+library(tidyverse)
+library(ggplot2)
+theme_set(theme_minimal())
+library(fcr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+library(ggpubr)
+library(gridExtra)
+library(kableExtra)
+library(LaplacesDemon)
+
+#### load data ####
+df <- read_rds(here("Data/nhanes_bi.rds"))
+df <- df %>% rename(id=SEQN, Y=Z)
+
+N <- length(unique(df$id)) # sample size 8763
+J <- max(df$sind) # 1440 measures for each subject
+t <- unique(df$sind)
+K <- 4 # number of eigenfunctions to use
+
+
+#### visulization ####
+rand_id <- sample(unique(df$id), size = 4)
+
+df %>% 
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  facet_wrap(~id)+
+  labs(x="Time", y = "Activity")+
+  theme(plot.margin = margin(0, 1, 0, 0))
+
+
+#### Data split ####
+# 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction
+train_id <- sample(unique(df$id), size = N*0.8)
+test_id <- setdiff(unique(df$id), train_id)
+
+train_df <- df %>% filter(id %in% train_id)
+test_df <- df %>% filter(id %in% test_id)
+
+
+##### fGFPCA #####
+
+# functions
+source(here("Code/GLMM-FPCA.R")) 
+source(here("Code/OutSampBayes.R"))
+
+# bin data
+bin_w <- 10 # bin width
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
+mid_t <- t[mid] # time value corresponding to mid points
+
+train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
+train_df$bin <- as.numeric(as.character(train_df$bin))
+
+# local GLMM
+t1 <- Sys.time()
+bin_lst <- split(train_df, f = train_df$bin)
+df_est_latent <- lapply(bin_lst, function(x){pred_latent(x, n_node = 0)}) 
[…]

### 2023-12-02 Docs/manuscript.Rmd
data application outcome, with problems

@@
 ```
+
+## Figure
+
+```{r fig_pred_anhanes}
+rand_id2 <- sample(unique(nhanes_pred_fgfpca$id), 4)
+
+fig_fgfpca_nhanes <- nhanes_pred_fgfpca %>% 
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
+  rename(t=sind) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=pred240, col = "4am"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred480, col = "8am"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred720, col = "12pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred960, col = "4pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred1200, col = "8pm"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "fGFPCA", col = "Maximum observation time")+
+  scale_x_continuous(breaks = window3, 
+                     labels = c("Midnight", "4am", "8am", "12pm", "4pm", "8pm", "Midnight"))
+```
+
+```{r}
+fig_adglmm_nhanes <- nhanes_pred_adglmm %>% 
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(t = t*1440) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=pred240, col = "4am"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred480, col = "8am"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred720, col = "12pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred960, col = "4pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred1200, col = "8pm"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GLMMadaptive", col = "Maximum observation time")+
+  scale_x_continuous(breaks = window3, 
+                     labels = c("Midnight", "4am", "8am", "12pm", "4pm", "8pm", "Midnight"))
+```
+
+```{r}
+ggarrange(fig_fgfpca_nhanes, fig_adglmm_nhanes, nrow = 1, common.legend = T)
+```
+
+## Issues
+
+In step 4 of fGFPCA, some eigenvalues, after re-evaluation, was very, very small. It also lost its monotone decreasing nature.
+
+The debias step of fGFPCA also took a very, very long time.

### 2023-12-02 Code/Data_Appl_NHANES.R

@@ -1,6 +1,6 @@
 # data application of NHANES
 
-set.seed(121)
+set.seed(1114)
 
 library(here)
 library(tidyverse)
@@ -16,6 +16,7 @@ library(ggpubr)
 library(gridExtra)
 library(kableExtra)
 library(LaplacesDemon)
+library(splines)
 
 #### load data ####
 df <- read_rds(here("Data/nhanes_bi.rds"))
@@ -25,18 +26,6 @@ N <- length(unique(df$id)) # sample size 8763
 J <- max(df$sind) # 1440 measures for each subject
 t <- unique(df$sind)
 K <- 4 # number of eigenfunctions to use
-
-
-#### visulization ####
-rand_id <- sample(unique(df$id), size = 4)
-
-df %>% 
-  filter(id %in% rand_id) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  facet_wrap(~id)+
-  labs(x="Time", y = "Activity")+
-  theme(plot.margin = margin(0, 1, 0, 0))
 
 
 #### Data split ####
@@ -65,7 +54,6 @@ train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = m
 train_df$bin <- as.numeric(as.character(train_df$bin))
 
 # local GLMM
-t1 <- Sys.time()
 bin_lst <- split(train_df, f = train_df$bin)
 df_est_latent <- lapply(bin_lst, function(x){pred_latent(x, n_node = 0)}) 
 df_est_latent <- bind_rows(df_est_latent) 
@@ -77,9 +65,9 @@ head(uni_eta_hat)
 mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                          nrow=length(train_id), 
                          ncol=n_bin, byrow = F) 
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T)
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T, npc=4)
 ncol(fpca_mod$efunctions)
-fpca_mod$evalues[1:4]
+fpca_mod$evalues
 
 # Re-evaluation
 ## grid extension
@@ -93,6 +81,7 @@ B <- spline.des(knots = knots_values, x = mid_t, ord = p + 1,
 Bnew <- spline.des(knots = knots_values, x = t, ord = p + 1,
[…]

### 2023-12-04 Docs/ProgressReportAppl.Rmd
small sample data application

@@
 # use pred_latent function to estimate latent function 
-source(here("Code/OutSampMLE.R"))
-# source(here("Code/OutsampBayes.R"))
+source(here("Code/OutsampBayes.R"))
 ```
@@
 
-- 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction
+- Original sample size took too much time in the debias step of fGFPCA method. 
+- In fact, the debais step does not work very well in this case
+- 500 subjects for training, 500 for out-of-sample prediction
 
@@
 
-train_id <- sample(unique(df$id), size = N*0.8)
-test_id <- setdiff(unique(df$id), train_id)
+Ntr <- Nte <- 500
+
+train_id <- unique(df$id)[1:500]
+test_id <- unique(df$id)[501:1000]
 
@@
 test_df <- df %>% filter(id %in% test_id)
+# 1440*500
 ```
@@
 mat_est_unique <- matrix(uni_eta_hat$eta_hat,
-                         nrow=length(train_id), 
+                         nrow=Ntr, 
                          ncol=n_bin, byrow = F) 
@@
 
-```{r}
-plot(mid, fpca_mod$mu, type = "l", xlab = "bin", ylab = "mean")
-```
-
-```{r, fig.height=8, fig.width=8}
-par(mfrow=c(2,2))
-
-plot(mid, fpca_mod$efunctions[, 1], type="l", xlab="bin", ylab="PC1")
-plot(mid, fpca_mod$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
-plot(mid, fpca_mod$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
-plot(mid, fpca_mod$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
-```
-
-```{r}
+```{r, eval=FALSE}
 # plot correlation matrix
@@
 
-Becuase I would like to focus this report on subjects with failed Laplace approximation, I will not do these two things for now.
+```{r projection}
+# Re-evaluation
+## grid extension
+p <- 3 # order of b splines 
+knots <- 35 # number of knots (same from FPCA model)
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
+
+B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
+                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
+Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
+                   outer.ok = TRUE)$design  # evaluate B-splines on original grid
+
+df_phi <- matrix(NA, J, K) 
+for(k in 1:K){
+  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+  df_phi[,k] <- Bnew %*% coef(lm_mod)
+}# project binned eigenfunctions onto the original grid
+```
+
+```{r, fig.height=8, fig.width=8}
+par(mfrow=c(2,2))
+
+plot(mid, fpca_mod$efunctions[, 1], pch=16, xlab="bin", ylab="PC1", col = "blue",
+     cex = 0.5)
+lines(1:J, df_phi[, 1])
+
+plot(mid, fpca_mod$efunctions[, 2], pch=16, xlab="bin", ylab="PC2", col = "blue",
+     cex = 0.5)
+lines(1:J, df_phi[, 2])
+
+plot(mid, fpca_mod$efunctions[, 3], pch=16, xlab="bin", ylab="PC3", col = "blue",
+     cex = 0.5)
+lines(1:J, df_phi[, 3])
+
+plot(mid, fpca_mod$efunctions[, 4], pch=16, xlab="bin", ylab="PC4", col = "blue",
+     cex = 0.5)
+lines(1:J, df_phi[, 4])
+```
+
+```{r revaluation}
+df_phi <- data.frame(sind = 1:J, df_phi)
+colnames(df_phi) <- c("sind", paste0("phi", 1:K))
+train_df <- train_df %>% 
+  left_join(df_phi, by = "sind")
+train_df$id <- as.factor(train_df$id)
+# class(train_df$id)
+# train_df$id2 <- factor(train_df$id, levels = train_id, labels = 1:length(train_id))
+# train_df$id2 <- as.numeric(train_df$id2)
+# range(unique(train_df$id2))
+
+t1 <- Sys.time()
+debias_glmm <- bam(Y ~ s(sind, bs="cc")+
+                     s(id, by=phi1, bs="re")+
+                     s(id, by=phi2, bs="re")+
+                     s(id, by=phi3, bs="re")+
+                     s(id, by=phi4, bs="re"), 
+                   family = binomial, 
+                   data=train_df, 
+                   method = "fREML",
+                   discrete = TRUE)
+t2 <- Sys.time()
+t2-t1
+```
+
+```{r}
+# mean
+new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean
+
+plot(mid, fpca_mod$mu, pch=16, xlab="bin", ylab="PC4", col = "blue",
+     cex = 0.5)
+lines(1:J, new_mu)
+
+# eigenvalues
+fpca_mod$evalues[1:K] # original
+lambda_scaled <- 1/debias_glmm$sp[2:5] # extract re-evaluated lambda
+fpca_mod$evalues[1:K]/lambda_scaled # ratio
+```
+
+```{r rescale}
+## rescale
+new_phi <- df_phi %>% select(starts_with("phi"))*sqrt(n_bin)
+new_phi <- as.matrix(new_phi)
+
+lambda_scaled <- lambda_scaled/n_bin
+```
 
 # Out-of-sample prediction
-
-follow up on the last report, I'll add some details here.
-
-Let's set:
-
-- $s$: bin index
-- $t_{m_s}$: midpoint time of bin s
-- $\mathscr{T}_s$: all observation points in bin s
-- $Y_i^s$ all observations in bin s from subject i. $Y_i^S := \{Y_i(t_j), t_j \in \mathscr{T}_s\}$
-
-Let's assume we have a new subject $u$ with maximum observation time $T_u$. Then the log-likelihood of this new subject would be:
 
@@
 
-## One subject example
-
-Now I am going to use one subject as an example to go over the approximation procedure:
-```{r, class.source = "fold-show"}
-## model
-Model <- function(parm, Data){
-  xi <- parm[Data$pos.xi]
-
-  # log-prior
-  xi.prior <- dmvnorm(xi, mean = rep(0, Data$J), sigma=Data$tao, log = TRUE)
-
-  # log-posterior likelihood
-  eta <- Data$f0+Data$X %*% xi
-  p <- exp(eta)/(1+exp(eta))
-  LL <- sum(dbinom(x=Data$y, size = Data$n, prob=p, log = TRUE)) # log likelihood of Y|xi
-  LP <- LL+sum(xi.prior) # unnormalized joint log likelihood of (Y, xi)
-
-  # output
-  Modelout <- list(LP=LP, Dev=-2*LL, Monitor=LP,
-                   yhat=Data$y, parm=parm)
-  return(Modelout)
-}
-```
-
 - Then we feed in data for one observation
 
-```{r, class.source = "fold-show"}
-rand_id <- test_id[15]
-df_new <- df %>% filter(id == rand_id & sind <= 540) # observation until 9am
+```{r}
+window <- seq(0, J, by = 8*60) # midnight,  8am, 4pm, midnight
 
-# put data into correct format
-ns <- as.vector(table(df_new$bin)) # number of observations
-hs <- df_new %>% group_by(bin) %>% summarize_at("Y", sum) %>% select(Y) %>% unlist()# number of success
-nf <- ns-hs # number of failure
-max_bin <- length(unique(df_new$bin)) # assume new skipped bins  
-df_new2 <- data.frame(bin = unique(df_new$bin), ns, hs, nf)
+test_df <- rename(test_df, t = sind)
 
-# extract estimates from FPCA model to be used in prior distribution
-tao <- diag(fpca_mod$evalues[1:K]) # variance of xi
+converge_state_m <- matrix(NA, nrow = length(test_id), 2)
+pred_list_m <- list()
 
-# extract estimates from FPCA model to be used in posterior distribution
-f0 <- fpca_mod$mu[1:max_bin]
-N <- nrow(df_new2) # "sample size" which is in fact number of observed bins in this case for a new subject
-n <- df_new2$ns # number of experiements at each bin
-y <- df_new2$hs # outcome, number of 1
-X <- fpca_mod$efunctions[1:max_bin, 1:K] # eigen/PC functions
-J <- K # number of parameters/scores
-mon.names <- "LP"
+t1 <- Sys.time()
+for(i in seq_along(test_id)){
+  df_i <- test_df %>% filter(id==test_id[i])
+  # per max obs time
+  for(tmax in window[2:3]){
+    df_it <- df_i %>% filter(t <= tmax)
+    max_rid <- nrow(df_it)
 
-# parameter names
-name_lst <- as.list(rep(0, J))
-names(name_lst) <- paste("xi", 1:J, sep = "")
-parm.names <- as.parm.names(name_lst) # names of parameters to estimate
-pos.xi <- grep("xi", parm.names)
+    # into a list
+    MyData <- list(K=K, 
+                   X=new_phi[1:max_rid, ], 
+                   mon.names=mon.names,
+                   parm.names=parm.names, 
+                   pos.xi=pos.xi, 
+                   y=df_it$Y, 
+                   tao=diag(lambda_scaled), f0=new_mu[1:max_rid])
 
-# data
-MyData <- list(J=J, X=X, mon.names=mon.names,
-                 parm.names=parm.names, pos.xi=pos.xi, y=y, n=n, tao=tao, f0=f0)
-```
+    # fit laplace approximation
+    Fit <- LaplaceApproximation(Model, 
+                                parm = rep(0, K), 
+                                Data=MyData, 
+                                Method = "NM", 
+                                Iterations = 1000,
+                                CovEst = "Identity")
+    converge_state_m[i, which(window[2:3]==tmax)] <- Fit$Converged
+    score <- Fit$Summary1[, "Mode"]
 
-```{r, class.source = "fold-show"}
-# fit laplace approximation
-# initial values generated by GIV function in the same package
-Fit <- LaplaceApproximation(Model, Data=MyData)
-```
+    # prediction
+    eta_pred_out <- new_mu+new_phi%*%score
+    df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
+  }
 
-- Summarizing output
+  pred_list_m[[i]] <- df_i
+}
+t2 <- Sys.time()
+t_pred <- t2-t1 # About 3.5 minutes
 
-```{r, class.source = "fold-show", fig.height=8, fig.width=8}
-# summarizing output
-print(Fit)
-plot(Fit, MyData)
+# mean(converge_state_m) # convergence rate 100%
 ```
@@
 ```{r}
-# prediction
-eta_pred_out <- fpca_mod$mu+fpca_mod$efunctions[, 1:K]%*%Fit$Summary1[, "Mode"]
-
-data.frame(bin = mid, eta_pred = eta_pred_out) %>% 
-  right_join(df %>% filter(id==rand_id) %>%  select(Y, sind, bin), by = "bin") %>%
-  ggplot()+
-  geom_point(aes(x=sind, y = Y))+
-  geom_line(aes(x=sind, y = exp(eta_pred)/(1+exp(eta_pred))))+
-  labs(title = rand_id)
-
+# check results
+df_pred <- bind_rows(pred_list_m)
+df_pred$pred480[df_pred$t<480] <- NA
+df_pred$pred960[df_pred$t<960] <- NA
 ```
 
-## Full data application output
-
-### fGFPCA
-
-```{r}
-# results generated from DataApplNHANES.R
-load(here("Data/ApplOutput_fGFPCA.RData"))
-```
-
-```{r message=FALSE, warning=FALSE}
-df_test_full <- df %>% 
-  filter(id %in% test_id) %>%
-  # filter(!id %in% skip_id) %>%
-  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin"))
-
-rand_id2 <- sample(unique(df_test$id), 4)
-# without interpolation, assume constant latent function value in each bin
-df_test_full%>%
-  filter(id %in% rand_id2) %>%
-  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>%
-  ggplot()+
-    geom_line(aes(x=bin, y = pred_t540, col = "9am"))+
-    geom_line(aes(x=bin, y = pred_t780, col = "1pm"))+
-    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"))+
-    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)",
-         title = "Constant latent function within bin")
-```
-
-<!-- Calculate AUC -->
-
-```{r}
-# break by prediction window
-lst_auc <- df_test_full %>% 
-  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
-                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-lst_auc <- split(lst_auc, f=lst_auc$window)
-```
-
-```{r}
-# no interpolation
-# up to 9am
-auc_t540 <- lst_auc[2:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
-auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 1pm
-auc_t780 <- lst_auc[3:4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t780, x$Y), measure = "auc"))
-auc_t780 <- lapply(auc_t780, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 5pm
-auc_t1020 <- lst_auc[4] %>% 
-  lapply(function(x)performance(prediction(x$pred_t1020, x$Y), measure = "auc"))
-auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
-```
-
-## GLMMadaptvie
+# GLMMadaptvie
 
@@
 ```{r}
-load(here("Data/ApplOutput_GLMMadaptive.RData"))
-```
-
-- `r round(t_est_adglmm, 2)` minutes on model fitting and `r round(t_pred_adglmm, 2)` minutes on out-of-sample prediction.
-
-- Below is the model fit on the training set
-
-```{r}
+t1 <- Sys.time()
+adglmm_mod <- mixed_model(Y ~ t, random = ~ t | id, 
+                          data = train_df %>% mutate(t=sind/J),
+                      family = binomial())
+t2 <- Sys.time()
+t_est_adglmm <- t2-t1 # model fitting took 20.82 mins
 summary(adglmm_mod)
@@
 ```{r}
-df_test_full2 <- df_test_full %>%
-  filter(id %in% rand_id2)%>%
-  select(id, sind, Y)
+test_df2 <- test_df %>% mutate(t=t/J)
+pred_list_m2 <- list(
+  predict(adglmm_mod, newdata = test_df2 %>% filter(t <= 480/J),
+          newdata2 =  test_df2 %>% filter(t>480/J), 
+          type = "subject_specific", type_pred = "link",
+          se.fit = FALSE, return_newdata = TRUE)$newdata2,
 
-df_test_full2$pred_t540 <- df_test_full2$pred_t780 <- df_test_full2$pred_t1020 <- NA 
-for(i in rand_id2){
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>540, "pred_t540"]<- adglmm_pred_t540$newdata2 %>% filter(id ==i) %>% select(pred)
-
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>780, "pred_t780"]<- adglmm_pred_t780$newdata2 %>% filter(id ==i) %>% select(pred)
-
-  df_test_full2[df_test_full2$id==i & df_test_full2$sind>1020, "pred_t1020"]<- adglmm_pred_t1020$newdata2 %>% filter(id ==i) %>% select(pred)
-}
-
-# colSums(is.na(df_test_full2))
+  predict(adglmm_mod, newdata = test_df2 %>% filter(t <= 960/J),
+          newdata2 =  test_df2 %>% filter(t>960/J), 
+          type = "subject_specific", type_pred = "link",
+          se.fit = FALSE, return_newdata = TRUE)$newdata2
+)
 ```
 
-- Below is an example of predicted probablity on four random subjects
-
-```{r, warning=FALSE}
-df_test_full2 %>% 
-  mutate_at(vars(pred_t540, pred_t780, pred_t1020), 
-            function(x)exp(x)/(1+exp(x))) %>%
-  ggplot()+
-    geom_line(aes(x=sind, y = pred_t540, col = "9am"))+
-    geom_line(aes(x=sind, y = pred_t780, col = "1pm"))+
-    geom_line(aes(x=sind, y = pred_t1020, col = "5pm"))+
-    geom_point(aes(x=sind, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "Estimated latent function (probablity scale)")
+```{r}
+# check results
+pred_list_m2 <- lapply(pred_list_m2, function(x){x %>% select(id, t, pred)})
+df_pred2 <- test_df2 %>% 
+  left_join(pred_list_m2[[1]], by = c("id", "t")) %>% 
+  rename(pred480 = pred) %>% 
+  left_join(pred_list_m2[[2]], by = c("id", "t")) %>% 
+  rename(pred960 = pred) %>%
+  mutate(t=t*J)
 ```
 
-<!-- Calculate AUC  -->
+# Results
 
 ```{r}
-# up to 540
-J <- 1440
-auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
-  # filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
-                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
-auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
+rand_id_test <- sample(unique(test_id), 4)
+ggarrange(
+  df_pred %>% 
+    filter(id %in% rand_id_test) %>%
+    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
+    ggplot()+
+    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
+    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
+    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "fGFPCA"),
 
-# up to 780
-auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
-  # filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
-                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
-auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
-
-# up to 1020
-auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
-  # filter(!id %in% skip_id) %>%
-  mutate(sind=sind*J) %>%
-  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
-                      labels = c("5pm-9pm","9pm-12pm"), 
-                      include.lowest = T))
-auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
-auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
-  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
-auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()
+  df_pred2 %>% 
+    filter(id %in% rand_id_test) %>%
+    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
+    ggplot()+
+    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
+    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
+    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "GLMMadaptive"),
+  nrow = 1, common.legend = T)
 
@@
 
-### AUC
+```{r auc_func}
+## a function to calculate AUC
+get_auc <- function(y, pred){
+  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
+    auc <- NA
+  }
+  else{
+    this_perf <- performance(prediction(pred, y), measure = "auc")
+    auc <- this_perf@y.values[[1]]
+  }
+  return(auc)
+}
+```
 
-- To use the same sample for comparison, I'd exclude the subjects with failed Laplace approximation from the test set for both methods
+```{r auc_fGFPCA_nhanes}
+auc_mat1 <- df_pred %>%
+  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred480),
+            auc2 = get_auc(Y, pred960)) %>%
+  filter(window != "[0,480]") %>% 
+  select(starts_with("auc"))
 
-```{r}
-options(knitr.kable.NA = '')
-tb_auc <- data.frame(auc_t540, 
-           c(NA, auc_t780), 
-           c(NA, NA, auc_t1020),
+colnames(auc_mat1) <- c("0-8am", "0am-4pm")
+```
 
-           auc_t540_adglmm,
-           c(NA, auc_t780_adglmm), 
-           c(NA, NA, auc_t1020_adglmm))
+```{r auc_GLMMadaptive_nhanes}
+auc_mat2 <- df_pred2 %>%
+  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred480),
+            auc2 = get_auc(Y, pred960)) %>%
+  filter(window != "[0,480]") %>% 
+  select(starts_with("auc"))
 
-colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)
+colnames(auc_mat2) <- c("0-8am", "0am-4pm")
 ```
@@
 ```{r}
-# time
-t_fit <- c(t_local_glmm+t_fpca, t_est_adglmm)
-units(t_fit) <- "mins"
-
-t_test <- c(t_pred, t_pred_adglmm)
-# units(t_pred)
-
-t_total <- t_fit+t_test
-
-idx1 <- idx2 <- idx3 <- c(1, 3, 3)
-names(idx1) <- c("Total time", as.character(round(t_total, 2)))
-names(idx2) <- c("Time on prediction", as.character(round(t_test, 2)))
-names(idx3) <- c("Time on model fitting", as.character(round(t_fit, 2)))
-
-t_fit <- as.character(round(t_fit, 2))
-t_test <- as.character(round(t_test, 2))
-t_total <- as.character(round(t_total, 2))
-
+data.frame(Window = c("8am-4pm", "4am-12pm"),
+           auc_mat1, auc_mat2, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 2, "GLMMadaptive" = 2)) %>%
+  add_header_above(c(" "= 1, "Observed track" = 4))
 ```
@@
 ```{r}
-tb_auc %>%
-  kable(digit = 4, table.attr = "style = \"color: black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(header=idx1) %>%
-  add_header_above(header=idx2) %>%
-  add_header_above(header=idx3) %>%
-  add_header_above(c(" "=1, "fGFPCA"=3, "GLMMadaptive"=3))
+nhanes_pred_fgfpca <- df_pred
+nhanes_pred_adglmm <- df_pred2
+save(nhanes_pred_fgfpca, nhanes_pred_adglmm,
+     file = here("Data/ApplOutput.RData"))
 ```
-
-- fGFPCA does not always have higher AUC, but definitely most of the time. 
-- In fact there are two scenarios where GLMMadaptive outperforms fGFPCA slightly: 1) given 9am to predict 1-5pm; and 2) given 1pm and predict 5-9pm.
-Both of them have a 4-hour interval apart from the maximum observation time and and prediction window.
-Could it have anything to do the cyclic trend in the data?
-Is 4h the length of a cycle?
-I tried to look at it but really couldn't see anything...there are too many points lumped together!
-
-```{r}
-df %>% filter(id %in% rand_id2) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  geom_vline(xintercept = c(540, 780, 1020, 1260))+
-  scale_x_continuous(breaks = c(540, 780, 1020, 1260),
-                     labels = c("9am", "1pm", "5pm", "9pm"))+
-  facet_wrap(~id)
-```
-
-## Failed Laplace approximation
-
-- This problem seems to be solved after changing the optimizer? (Method = "BFGS")
-- The new optimizer also took less time
-- The old one was SPG (Spectral projected gradient), which is a non-monotone line search. 
-- The new one is BFGS (quasi-newton, the same in glmmTMB)
-
-## Discussion
-
-- What would be the reference method to compare our performance to?
-Original GLMMadaptvie perhaps is feasible on the entire dataset.
-But what should we use if we wanna do sub-sample comparison?
-Should we even do that for this data application (or simulation alone)?
-<!-- - There are still three participants with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.  -->
-<!-- - Is interpolation really necessary?
-It does not improve perfor- Should I calculate subject-wise AUC and average them?
-But some subjects are all zero within the prediction window, making AUC for this single subjects impossible to calculate...

### 2023-12-04 Docs/manuscript.Rmd
small sample data application

--- 2023-11-30 Docs/manuscript.Rmd
@@
 
-<!-- # NHANES data application -->
+# NHANES data application
+
+We take 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction.
+
+In the GLMMadaptive model, I could only fit a model with random intercept. It took more than 20 minutes for model fitting.
+
+```{r load_nhanes}
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+load(here("Data/ApplOutput.RData"))
+```
+
+```{r nhance example}
+df_nhanes %>% select(SEQN, Z, sind) %>% 
+  mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
+  mutate(SEQN = factor(SEQN, labels = 1:8763)) %>%
+  ggplot()+
+  geom_tile(aes(x=sind, y = SEQN, fill = Z))+
+  labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
+  theme(axis.text.y = element_blank())
+
+```
+
+## AUC
+
+```{r auc_fGFPCA_nhanes}
+window3 <- c(0, 480, 960, 1440)
+
+auc_mat_nhanes <- nhanes_pred_fgfpca %>%
+  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred480),
+            auc2 = get_auc(Y, pred960)) %>%
+  filter(window != "[0,480]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_nhanes) <- c("0-8am", "0am-4pm")
+```
+
+```{r auc_GLMMadaptive_nhanes}
+auc_mat_nhanes2 <- nhanes_pred_adglmm %>%
+  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred480),
+            auc2 = get_auc(Y, pred960)) %>%
+  filter(window != "[0,480]") %>% 
+  select(starts_with("auc"))
+
+colnames(auc_mat_nhanes2) <- c("0-8am", "0am-4pm")
+```
+
+```{r}
+data.frame(Window = c("8am-4pm", "4am-12pm"),
+           auc_mat_nhanes, auc_mat_nhanes2, check.names = F) %>%
+  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "fGFPCA" = 2, "GLMMadaptive" = 2)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+```
+
+## Figure
+
+```{r}
+rand_id2 <- sample(unique(nhanes_pred_fgfpca$id), 4)
+
+ggarrange(
+  nhanes_pred_fgfpca %>% 
+    filter(id %in% rand_id2) %>%
+    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
+    ggplot()+
+    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
+    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
+    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "fGFPCA"),
+
+  nhanes_pred_adglmm %>% 
+    filter(id %in% rand_id2) %>%
+    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
+    ggplot()+
+    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
+    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
+    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
+    facet_wrap(~id)+
+    labs(x = "Time", y = "GLMMadaptive"),
+  nrow = 1, common.legend = T)
+```

### 2023-12-04 Docs/ProgressReportAppl.Rmd
WIT presentation figures

@@
 # use pred_latent function to estimate latent function 
+K <- 4
 source(here("Code/OutsampBayes.R"))
@@
 
+```{r PC}
+data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
+  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
+  pivot_longer(2:5, names_to = "PC") %>%
+  ggplot()+
+  geom_line(aes(x=t, y=value, col = PC))+
+  labs(x="Time", y="", title = "Eigenfunctions on the binned grid")
+```
+
 ```{r, eval=FALSE}
@@
 
-```{r, fig.height=8, fig.width=8}
-par(mfrow=c(2,2))
+```{r}
+df_pc1 <- data.frame(t=1:J,  df_phi) %>%
+  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
+  pivot_longer(2:5, names_to = "PC")
 
-plot(mid, fpca_mod$efunctions[, 1], pch=16, xlab="bin", ylab="PC1", col = "blue",
-     cex = 0.5)
-lines(1:J, df_phi[, 1])
+df_pc2 <- data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
+  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
+  pivot_longer(2:5, names_to = "PC")
 
-plot(mid, fpca_mod$efunctions[, 2], pch=16, xlab="bin", ylab="PC2", col = "blue",
-     cex = 0.5)
-lines(1:J, df_phi[, 2])
+left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
+  ggplot()+
+  geom_line(aes(x=t, y=value.x, col = PC), linewidth = 0.5)+
+  geom_point(aes(x=t, y=value.y, col = PC), na.rm = T, alpha = 0.5, size = 0.5)+
+  labs(x="Time", y="", title = "Re-evaluate eigenfunctions on the original grid")
 
-plot(mid, fpca_mod$efunctions[, 3], pch=16, xlab="bin", ylab="PC3", col = "blue",
-     cex = 0.5)
-lines(1:J, df_phi[, 3])
-
-plot(mid, fpca_mod$efunctions[, 4], pch=16, xlab="bin", ylab="PC4", col = "blue",
-     cex = 0.5)
-lines(1:J, df_phi[, 4])
 ```

### 2023-12-04 Docs/manuscript.Rmd
WIT presentation figures

@@
 output:
+  pdf_document:
+    toc: yes
+    toc_depth: '3'
+    keep_tex: yes
   html_document:
     self_contained: no
-    number_sections: true
+    number_sections: yes
     code_folding: hide
-    toc: true
+    toc: yes
     toc_depth: 3
-    toc_float: true
+    toc_float: yes
     font: 12pt
-    keep_tex: true
+    keep_tex: yes
 ---
@@
   geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=eta_i, col = "True"), show.legend = F)+
   facet_wrap(~id)+
-  labs(title = "Simulated data")
+  labs(title = "Simulated data")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
 ```
@@
 
-```{r fig_pred}
+```{r}
 # range(pred_list_all[[1]]$pred0.2, na.rm = T)
@@
   facet_wrap(~id)+
-  labs(title = "fGFPCA", col = "Maximum observation time")
+  labs(title = "fGFPCA", col = "Maximum observation time")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5))
 
@@
   facet_wrap(~id)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time")
+  labs(title = "GLMMadaptive", col = "Maximum observation time")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5))
 
@@
 
-```{r}
+```{r fig_sim, fig.height=4, fig.width=8}
 ggarrange(fig_fgfpca, fig_adglmm,
@@
            mean_ise, mean_ise_ref, check.names = F) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable(digits = 3, caption = "Integrated squared error", booktabs=T) %>%
   kable_styling(full_width = F) %>% 
@@
            mean_auc, mean_auc_ref, check.names = F) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area under the ROC curve") %>%
+  kable(digits = 3, caption = "Area under the ROC curve", booktabs = T) %>%
   kable_styling(full_width = F) %>% 
@@
            "Prediction"= c(mean(pred_time), mean(pred_time_ref)/60)) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Computation time (minutes)") %>%
+  kable(digits = 3, caption = "Computation time (minutes)", booktabs = T) %>%
   kable_styling(full_width = F) 
@@
            ise_mat_subset, ise_mat_subset2, check.names = F) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable(digits = 3, caption = "Integrated squared error") %>%
   kable_styling(full_width = F) %>% 
@@
            auc_mat_subset, auc_mat_subset2, check.names = F) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Integrated squared error") %>%
+  kable(digits = 3, caption = "Integrated squared error") %>%
   kable_styling(full_width = F) %>% 
@@
 
-```{r nhance example}
+```{r nhance_example}
 df_nhanes %>% select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% unique(nhanes_pred_adglmm$id) | SEQN %in% unique(nhanes_pred_fgfpca$id)) %>%
   mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
-  mutate(SEQN = factor(SEQN, labels = 1:8763)) %>%
+  mutate(SEQN = factor(SEQN)) %>%
   ggplot()+
@@
   labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
-  theme(axis.text.y = element_blank())
-
+  theme(axis.text.y = element_blank())+
+  scale_x_continuous(breaks = seq(0, 1440, by = 480),
+                     labels = c("Midnight", "8am", "4pm", "Midnight"))
 ```
@@
 
-colnames(auc_mat_nhanes) <- c("0-8am", "0am-4pm")
+colnames(auc_mat_nhanes) <- c("8am", "4pm")
 ```
@@
 
-colnames(auc_mat_nhanes2) <- c("0-8am", "0am-4pm")
+colnames(auc_mat_nhanes2) <- c("8am", "4pm")
 ```
@@
 ```{r}
-data.frame(Window = c("8am-4pm", "4am-12pm"),
+data.frame(Window = c("8am-4pm", "4am-midnight"),
            auc_mat_nhanes, auc_mat_nhanes2, check.names = F) %>%
-  kable(digits = 3, table.attr = "style = \"color: black;\"", caption = "Area Under the ROC curve") %>%
+  kable(digits = 3, caption = "Area Under the ROC curve", booktabs = T) %>%
   kable_styling(full_width = F) %>% 
@@
 
-```{r}
+```{r pred_nhanes, fig.height=4, fig.width=10}
 rand_id2 <- sample(unique(nhanes_pred_fgfpca$id), 4)
@@
     facet_wrap(~id)+
-    labs(x = "Time", y = "fGFPCA"),
+    labs(x = "Time", y = "fGFPCA")+
+  scale_x_continuous(breaks = seq(0, 1440, by = 480), 
+                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
+  theme(axis.text = element_text(size = 5)),
 
@@
     facet_wrap(~id)+
-    labs(x = "Time", y = "GLMMadaptive"),
+    labs(x = "Time", y = "GLMMadaptive")+
+  scale_x_continuous(breaks = seq(0, 1440, by = 480), 
+                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
+  theme(axis.text = element_text(size = 5)),
   nrow = 1, common.legend = T)
 ```
+
+```{r small_exp}
+# a smaller example of binary functions
+df_nhanes %>% select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% rand_id2 & Z == 1) %>% 
+  mutate(SEQN = as.factor(SEQN)) %>% 
+  ggplot()+
+  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
+  labs(x="Time", y="Subject", title = "NHANES binary activity indicator")
+```
+
+# Additional figures
+
+## fGFPCA algorithm demonstration
+
+```{r binning}
+df_nhanes %>% select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% rand_id2 & sind >= 650 & sind <= 750) %>% 
+  mutate(SEQN = as.factor(SEQN)) %>% 
+  ggplot()+
+  geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+
+  geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+
+  scale_y_continuous(breaks = c(0, 1))+
+  labs(x="Time", y="", title = "NHANES binary activity indicator")
+```
+
+```{r}
+df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% rand_id2 & sind >= 650 & sind <= 750) %>% 
+  mutate(bin = cut(sind, breaks = seq(650, 750, by = 10), 
+                   labels = seq(655, 745, by = 10),
+                   include.lowest = T)) 
+bin_mid <- seq(655, 745, by = 10)
+
+for(i in bin_mid){
+  df_bin_i <- df_fig %>% filter(bin==i) 
+  glm_i <- glmer(Z ~ 1 + (1|SEQN), data = df_bin_i, family = binomial, nAGQ=0)
+  eta_i <- predict(glm_i, type = "link")
+
+  df_fig$eta_hat[df_fig$bin==i] <- eta_i
+}
+
+df_fig %>% 
+  mutate(SEQN = as.factor(SEQN),
+         eta_hat = exp(eta_hat)/(1+exp(eta_hat)),
+         bin=as.numeric(as.character(bin))) %>%
+
+  ggplot()+
+  geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+
+  geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+
+  geom_point(aes(x=bin, y=eta_hat, col = SEQN))+
+  geom_line(aes(x=bin, y=eta_hat, col = SEQN))+
+  scale_y_continuous(breaks = c(0, 1))+
+  labs(x="Time", y="", title = "NHANES binary activity indicator")
+
+```

### 2023-12-04 Docs/manuscript.tex
WIT presentation figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Manuscript progress report},
  pdfauthor={Ying Jin},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Manuscript progress report}
\author{Ying Jin}
\date{2023-12-04}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{method}{%
\section{Method}\label{method}}

\hypertarget{assumptions}{%
\subsection{Assumptions}\label{assumptions}}

\begin{itemize}
\tightlist
\item
  For each subject i in the population, a generalized outcome \(Y_i(t)\)
  is generated along a variable t (for example, time), where
  \(t \in (0, T)\).
\item
  The outcome, at any specific t, follows an exponential family
  distribution characterized by a (latent) continuous function
  \(\eta_i(t)\):
\end{itemize}

\[g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)\]

\[p(Y_i(t)) = h(Y_i(t))exp\{\eta_i(t)T[Y_i(t)]-A(\eta_i(t))\}\]

\begin{itemize}
\tightlist
\item
  The continuous latent function consists of a population-level fixed
  process and an individual-level random process
\end{itemize}

\[\eta_i(t) = \beta_0(t)+b_i(t)\]

\hypertarget{observed-data}{%
\subsection{Observed data}\label{observed-data}}

In practice we would observe the discrete realization of
\(\{Y_i(t), t\}\) along a dense grid. For simplicity, we assume the
observation grid is regular (same across sample). When we have J
observations points in \((0, T]\), then for the jth observation point,
we denote the corresponding value of t as \(t_j\), and the corresponding
outcome at this point \(Y_i(t_j)\).

\hypertarget{fgfpca-algorithm}{%
\subsection{fGFPCA Algorithm}\label{fgfpca-algorithm}}

\hypertarget{bin-data}{%
\subsubsection{Bin data:}\label{bin-data}}

Choose a proper bin width \(w\) considering model complexity and
identifiability. For now let's say the bins are equal-length and
non-overlapping.

\begin{itemize}
\tightlist
\item
  Bin index \(s = 1...S\)
\item
  Index of bin midpoints \(m_s\)
\item
  Value of t corresponding to bin midpoints \(t_{m_s}\)
\item
  Bin endpoints: \((t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]\)
\end{itemize}

\hypertarget{local-glmms}{%
\subsubsection{Local GLMMs}\label{local-glmms}}

At the every bin, we fit a local intercept-only model:

\[g[E(Y_i(t_j))] =\eta_i(t_{m_s})= \beta_0(t_{m_s})+b_i(t_{m_s})\] where
\(t_j \in (t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]\).

Here we are basically saying that the value of latent function is
constant within the same bin, which clearly is a misspecification of the
true latent process.

From the model above. we will be able to estimate a
\(\hat{\eta_i}(t_{m_s})\) on the binned grid for every individual in the
training sample.

\hypertarget{fpca}{%
\subsubsection{FPCA}\label{fpca}}

Here, we fit a FPCA model on the \(\hat{\eta_i}(t_{m_s})\) obtained from
step 2:

\[\hat{\eta}_i(t_{m_s}) = f_0(t_{m_s})+\sum_{k=1}^K\xi_{ik}\phi_{k}(t_{m_s})+\epsilon_i(t_{m_s})\]

where \(\xi_{ik}\) independently follows normal distribution
\(N(0, \lambda_k)\), and \(\epsilon_i(t_{m_s})\) at each point follows
\(N(0, \sigma_2)\).

From this model, we will be able to obtain the following estimates which
are shared across population:

\begin{itemize}
\tightlist
\item
  Population mean \(\hat{f_0}(t_{m_s})\)
\item
  Basis functions
  \(\hat{\mathbf{\Phi}} = \{\hat{\phi}_1(t_{m_s}), ...,\hat{\phi}_K(t_{m_s}))\}\)
\item
  Estimates of variance of scores \(\hat{\lambda}_1...\hat{\lambda}_K\)
\end{itemize}

\hypertarget{projection-and-debias}{%
\subsubsection{Projection and Debias}\label{projection-and-debias}}

The mean and basis functions are evaluated on the binned grid. To extend
it to the original measurement grid data was collected on, we project
the estimated eigenfunctions \(\hat{\mathbf{\Phi}}\) back use spline
basis. Now we have extend the \(\hat{\phi}_k(t_{m_s})\) to the original
grid \(\hat{\phi}_k(t_j)\)

Because of the misspecification of local GLMMs, the estimated
eigenfunctions and eigenvalues are also biased by a constant
multiplicative effect. Therefore, we use a GLMM to re-evaluate the mean
function, eigenfunctions and eigenvalues.

\hypertarget{out-of-sample-prediction}{%
\subsection{Out-of-sample prediction}\label{out-of-sample-prediction}}

Now, let's assume we have a new subject \(u\) with \(J_u\) observations
(\(J_u < J\)). Then the log-likelihood of this new subject would be:

\[l_u=\sum_{t_j<t_{J_u}}log(h(Y_u(t_j)))+\hat{\eta}_u(t_j)T(Y_u(t_j))-log(A[\hat{\eta}_u(t_j)])\]

where
\(\hat{\eta}_u(t_j) = \hat{f}_0(t_j)+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_j)\).

With estimates for the population-level parameters from fGFPCA
algorithms above, we can estimate \(\xi_{uk}\) by maximization of
\(l_u\). Direct maximization some times does not have closed form
solution. Numeric maximization methods seem not very stable as well. So
I have decided to used a Bayes approach (Laplace Approximation):

\begin{itemize}
\tightlist
\item
  Prior distribution: \(\xi_{uk} \sim N(0, \hat{\lambda}_k)\)
\item
  Posterior distribution: the likelihood of
  \(l_u =l(Y_u(t_j)|\mathbf{\xi}_u)\)
\end{itemize}

Laplace Approximation would get the posterior mode of \(\xi_{uk}\)
through quadratic approximation.

\hypertarget{larger-scale-simulation}{%
\section{Larger-scale simulation}\label{larger-scale-simulation}}

\hypertarget{simulation-set-up}{%
\subsection{Simulation set up}\label{simulation-set-up}}

Here we simulate binary data from cyclic latent process:

\[\begin{aligned}
Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}\]

where:

\begin{itemize}
\tightlist
\item
  \(t\) is 1000 equal-spaced observations points on \([0, 1]\) (J =
  1000).
\item
  \(f_0(t)=0\)
\item
  \(\xi_k \sim N(0, \lambda_k)\), and
  \(\lambda_k = 1, 0.5, 0.25, 0.125\) for k = 1, 2, 3, 4 respectively.
\item
  Sample size \(N = 500\)
\item
  In the binning step, we bin every 10 observations
\item
  500 simulations were implemented
\end{itemize}

\includegraphics{manuscript_files/figure-latex/fig_sim_data-1.pdf}

\hypertarget{reference-method}{%
\subsection{Reference method}\label{reference-method}}

\begin{itemize}
\tightlist
\item
  GLMMadaptive
\item
  Here we can fit a model with random intercept and slope for time. It
  is doable on 500 datasets, but obviously too simple for the data
  generation scheme. We would expect it to perform terribly.
\end{itemize}

\[g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\]

\hypertarget{figure}{%
\subsection{Figure}\label{figure}}

\includegraphics{manuscript_files/figure-latex/fig_sim-1.pdf}

\hypertarget{ise}{%
\subsection{ISE}\label{ise}}

\begin{table}

\caption{\label{tab:unnamed-chunk-4}Integrated squared error}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-9}
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
\midrule
(0.2, 0.4] & 146.407 &  &  &  & 387.708 &  &  & \\
(0.4, 0.6] & 183.967 & 74.977 &  &  & 291.579 & 269.799 &  & \\
(0.6, 0.8] & 218.265 & 49.275 & 15.776 &  & 315.778 & 282.736 & 278.242 & \\
(0.8, 1.0] & 108.918 & 77.981 & 17.747 & 12.005 & 563.011 & 477.485 & 597.746 & 600.34\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{auc}{%
\subsection{AUC}\label{auc}}

\begin{table}

\caption{\label{tab:unnamed-chunk-6}Area under the ROC curve}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-9}
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
\midrule
(0.2, 0.4] & 0.748 &  &  &  & 0.591 &  &  & \\
(0.4, 0.6] & 0.664 & 0.734 &  &  & 0.524 & 0.596 &  & \\
(0.6, 0.8] & 0.715 & 0.790 & 0.803 &  & 0.669 & 0.694 & 0.687 & \\
(0.8, 1.0] & 0.740 & 0.755 & 0.781 & 0.784 & 0.514 & 0.556 & 0.526 & 0.564\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:time}Computation time (minutes)}
\centering
\begin{tabular}[t]{lrr}
\toprule
Method & Fit & Prediction\\
\midrule
fGFPCA & 0.725 & 1.592\\
GLMMadaptive & 2.287 & 0.017\\
\bottomrule
\end{tabular}
\end{table}

I think we could say that while the total time spend on model fitting +
prediction are similar between two methods, fGFPCA achieved much better
flexibility and much better predictive performance of prediction under
every scenario.

\hypertarget{small-scale-simulation-one-iteration}{%
\section{Small-scale simulation: one
iteration}\label{small-scale-simulation-one-iteration}}

Here we would like to fit fGFPCA and GLMMadaptive on a dataset with
smaller sample size and/or smaller measurement density. For the
GLMMadaptive model, we would set it up with spline basis functions so
that its flexibility is comparable with fGFPCA model, such as:

\[g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^5\xi_{il}\phi_l(t)\]

I did the following things for data-reduction

\begin{itemize}
\item
  Reduce sample size to 10 (for both training and testing)
\item
  Reduce measurement track by half (0.25-0.75)
\item
  Another way is the simple reduce the measurement density. For example,
  instead of every minute, we can use every 10 minutes. In this case,
  the binning step in fGFPCA would bin every 100 minutes in to one bin.
\item
  \begin{itemize}
  \tightlist
  \item
    When we have smaller density of measurement grid, both methods could
    run into numeric issues in both methods. For example, singularity
    happened the the local GLMM step of fGFPCA procedure. This might be
    able to get resoved by changing the optimizer (e.g.~Nealder\_Mead).
    Do we want to do that? Or perhaps we simply want to use the full
    track?
  \end{itemize}
\item
  We may also need to reduce number of simulations (GLMMadaptive took
  20-ish minutes for one simulaiton).
\end{itemize}

Below are some of the outcomes:

\hypertarget{figure-1}{%
\subsection{Figure}\label{figure-1}}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}

\hypertarget{ise-1}{%
\subsection{ISE}\label{ise-1}}

\begin{table}

\caption{\label{tab:unnamed-chunk-8}Integrated squared error}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r|r|r}
\hline
\multicolumn{1}{c|}{ } & \multicolumn{8}{c}{Observation track} \\
\cline{2-9}
\multicolumn{1}{c|}{ } & \multicolumn{4}{c|}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
\cline{2-5} \cline{6-9}
Window & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65] & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65]\\
\hline
(0.35,0.45] & 35.724 &  &  &  & 48.779 &  &  & \\
\hline
(0.45,0.55] & 77.379 & 22.690 &  &  & 74.537 & 34.232 &  & \\
\hline
(0.55,0.65] & 32.581 & 20.370 & 20.043 &  & 48.683 & 35.666 & 18.303 & \\
\hline
(0.65, 0.75] & 158.191 & 135.968 & 130.360 & 66.648 & 174.158 & 141.285 & 143.370 & 103.92\\
\hline
\end{tabular}
\end{table}

\hypertarget{auc-1}{%
\subsection{AUC}\label{auc-1}}

\begin{table}

\caption{\label{tab:unnamed-chunk-9}Integrated squared error}
\centering
\begin{tabular}[t]{l|r|r|r|r|r|r|r|r}
\hline
\multicolumn{1}{c|}{ } & \multicolumn{8}{c}{Observation track} \\
\cline{2-9}
\multicolumn{1}{c|}{ } & \multicolumn{4}{c|}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
\cline{2-5} \cline{6-9}
Window & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65] & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65]\\
\hline
(0.35,0.45] & 0.657 &  &  &  & 0.643 &  &  & \\
\hline
(0.45,0.55] & 0.611 & 0.690 &  &  & 0.624 & 0.702 &  & \\
\hline
(0.55,0.65] & 0.687 & 0.710 & 0.714 &  & 0.666 & 0.686 & 0.702 & \\
\hline
(0.65, 0.75] & 0.604 & 0.639 & 0.636 & 0.653 & 0.598 & 0.593 & 0.587 & 0.62\\
\hline
\end{tabular}
\end{table}

There is huge difference in compuation time. GLMMadaptive took 23
minutes, while fGFPCA took less than 3 seconds.

\hypertarget{nhanes-data-application}{%
\section{NHANES data application}\label{nhanes-data-application}}

We take 80\% (7010) subjects for training, 20\% (1753) for out-of-sample
prediction.

In the GLMMadaptive model, I could only fit a model with random
intercept. It took more than 20 minutes for model fitting.

\includegraphics{manuscript_files/figure-latex/nhance_example-1.pdf}

\hypertarget{auc-2}{%
\subsection{AUC}\label{auc-2}}

\begin{table}

\caption{\label{tab:unnamed-chunk-10}Area Under the ROC curve}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-5}
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{fGFPCA} & \multicolumn{2}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Window & 8am & 4pm & 8am & 4pm\\
\midrule
8am-4pm & 0.587 &  & 0.628 & \\
4am-midnight & 0.680 & 0.766 & 0.448 & 0.613\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{figure-2}{%
\subsection{Figure}\label{figure-2}}

\includegraphics{manuscript_files/figure-latex/pred_nhanes-1.pdf}

\includegraphics{manuscript_files/figure-latex/small_exp-1.pdf}

\hypertarget{additional-figures}{%
\section{Additional figures}\label{additional-figures}}

\hypertarget{fgfpca-algorithm-demonstration}{%
\subsection{fGFPCA algorithm
demonstration}\label{fgfpca-algorithm-demonstration}}

\includegraphics{manuscript_files/figure-latex/binning-1.pdf}

\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1.pdf}

\end{document}

### 2023-12-05 Docs/manuscript.Rmd
small scale simulation 100 iterations

@@
 
-# Small-scale simulation: one iteration
+# Small-scale simulation
 
@@
 
-$$g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^5\xi_{il}\phi_l(t)$$
+$$g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
 
-I did the following things for data-reduction
-
-- Reduce sample size to 10 (for both training and testing)
-- Reduce measurement track by half (0.25-0.75)
-- Another way is the simple reduce the measurement density.
-For example, instead of every minute, we can use every 10 minutes.
-In this case, the binning step in fGFPCA would bin every 100 minutes in to one bin.
-- - When we have smaller density of measurement grid, both methods could run into numeric issues in both methods.
-For example, singularity happened the the local GLMM step of fGFPCA procedure.
-This might be able to get resoved by changing the optimizer (e.g.
-Nealder_Mead).
-Do we want to do that?
-Or perhaps we simply want to use the full track?
-
-- We may also need to reduce number of simulations (GLMMadaptive took 20-ish minutes for one simulaiton).
-
-Below are some of the outcomes:
+I have used 100 subjects for training and testing, and repeated 100 times.
+When fitting GLMMadpative, I reduce the number of measurements to 1/10 by taking one every 10 observations.
+The prediction is on the original grid.
 
@@
 ```{r fig_pred_subset}
-fig_fgfpca_subset <- df_pred_m_fGFPCA %>% filter(id %in% 501:504) %>%
-  mutate_at(vars(eta_i, pred0.35, pred0.45, pred0.55, pred0.65), function(x){exp(x)/(1+exp(x))}) %>%
+rand_id2 <- sample(unique(pred_subset_fGFPCA[[1]]$id), 4)
+
+fig_fgfpca_subset <- pred_subset_fGFPCA[[1]] %>%
+  filter(id %in% rand_id2) %>%
+  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
@@
   geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.35, col = "0.35"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.45, col = "0.45"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.55, col = "0.55"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.65, col = "0.65"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
   facet_wrap(~id)+
-  labs(title = "fGFPCA", col = "Maximum observation time")
+  labs(title = "fGFPCA", col = "Maximum observation time")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
 
-fig_adglmm_subset <- df_pred_m_adglmm %>% filter(id %in% 501:504) %>%
-  mutate_at(vars(eta_i, pred0.35, pred0.45, pred0.55, pred0.65), function(x){exp(x)/(1+exp(x))}) %>%
+fig_adglmm_subset <- df_subset_adglmm %>% filter(id %in% rand_id2) %>%
+  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
@@
   geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.35, col = "0.35"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.45, col = "0.45"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.55, col = "0.55"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.65, col = "0.65"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
   facet_wrap(~id)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time")
+  labs(title = "GLMMadaptive", col = "Maximum observation time")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
 
@@
 
+```{r}
+M2 <- length(pred_subset_fGFPCA)
+## ISE container 
+ise_mat_subset <- ise_mat_subset_ref <- 
+  array(NA, dim = c(length(window)-2, length(window)-2, M2))
+```
+
 ```{r ise_fGFPCA_subset}
-window2 <- seq(0.25, 0.75, by = 0.1)
-ise_mat_subset <- df_pred_m_fGFPCA %>%
-    mutate(err1 = (pred0.35-eta_i)^2,
-           err2 = (pred0.45-eta_i)^2,
-           err3 = (pred0.55-eta_i)^2,
-           err4 = (pred0.65-eta_i)^2) %>%
+for(m in 1:M2){
+  this_df <- pred_subset_fGFPCA[[m]]
+  ise_tb_m <- this_df %>%
+    mutate(err1 = (pred0.2-eta_i)^2,
+           err2 = (pred0.4-eta_i)^2,
+           err3 = (pred0.6-eta_i)^2,
+           err4 = (pred0.8-eta_i)^2) %>%
     select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
     group_by(window, id) %>% 
@@
     summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0.25,0.35]") %>% 
-    select(starts_with("err"))
-colnames(ise_mat_subset) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+    filter(window != "[0,0.2]") %>% 
+    select(starts_with("err")) %>% as.matrix()
+  ise_mat_subset[,,m] <- ise_tb_m
+}
+
+ise_mat_subset <- apply(ise_mat_subset, c(1, 2), mean)
+
+colnames(ise_mat_subset) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r ise_GLMMadaptive_subset}
-ise_mat_subset2 <- df_pred_m_adglmm %>%
-    mutate(err1 = (pred0.35-eta_i)^2,
-           err2 = (pred0.45-eta_i)^2,
-           err3 = (pred0.55-eta_i)^2,
-           err4 = (pred0.65-eta_i)^2) %>%
+ise_mat_subset2 <- df_subset_adglmm %>%
+    mutate(err1 = (pred0.2-eta_i)^2,
+           err2 = (pred0.4-eta_i)^2,
+           err3 = (pred0.6-eta_i)^2,
+           err4 = (pred0.8-eta_i)^2) %>%
     select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
     group_by(window, id) %>% 
@@
     summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0.25,0.35]") %>% 
+    filter(window != "[0,0.2]") %>% 
     select(starts_with("err"))
-colnames(ise_mat_subset2) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+colnames(ise_mat_subset2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.35,0.45]", "(0.45,0.55]", "(0.55,0.65]", "(0.65, 0.75]"),
+data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
            ise_mat_subset, ise_mat_subset2, check.names = F) %>%
@@
 
+```{r}
+## auc container 
+auc_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, M2))
+auc_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, M2))
+```
+
 ```{r auc_fGFPCA_subset}
-auc_mat_subset <- df_pred_m_fGFPCA %>%
-  mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+for(m in 1:M2){
+  this_df <- pred_subset_fGFPCA[[m]]
+
+  auc_tb_m <- this_df %>%
+  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
   select(Y, starts_with("pred"), window) %>%
   group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.35),
-            auc2 = get_auc(Y, pred0.45),
-            auc3 = get_auc(Y, pred0.55),
-            auc4 = get_auc(Y, pred0.65)) %>%
-  filter(window != "[0.25,0.35]") %>% 
-  select(starts_with("auc"))
+  summarise(auc1 = get_auc(Y, pred0.2),
+            auc2 = get_auc(Y, pred0.4),
+            auc3 = get_auc(Y, pred0.6),
+            auc4 = get_auc(Y, pred0.8)) %>%
+  filter(window != "[0,0.2]") %>% 
+  select(starts_with("auc")) %>% as.matrix()
 
-colnames(auc_mat_subset) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+  auc_mat_subset[, , m] <- auc_tb_m
+
+}
+
+auc_mat_subset <- apply(auc_mat_subset, c(1,2), mean)
+colnames(auc_mat_subset) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r auc_GLMMadaptive_subset}
-auc_mat_subset2 <- df_pred_m_adglmm %>%
-  mutate(window = cut(t, breaks = window2, include.lowest = T)) %>% 
+auc_mat_subset2 <- df_subset_adglmm %>%
+  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
   select(Y, starts_with("pred"), window) %>%
   group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.35),
-            auc2 = get_auc(Y, pred0.45),
-            auc3 = get_auc(Y, pred0.55),
-            auc4 = get_auc(Y, pred0.65)) %>%
-  filter(window != "[0.25,0.35]") %>% 
+  summarise(auc1 = get_auc(Y, pred0.2),
+            auc2 = get_auc(Y, pred0.4),
+            auc3 = get_auc(Y, pred0.6),
+            auc4 = get_auc(Y, pred0.8)) %>%
+  filter(window != "[0,0.2]") %>% 
   select(starts_with("auc"))
 
-colnames(auc_mat_subset2) <- c("(0.25,0.35]", "(0.25,0.45]", "(0.25,0.55]", "(0.25,0.65]")
+colnames(auc_mat_subset2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.35,0.45]", "(0.45,0.55]", "(0.55,0.65]", "(0.65, 0.75]"),
+data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
            auc_mat_subset, auc_mat_subset2, check.names = F) %>%
@@
 ```{r pred_nhanes, fig.height=4, fig.width=10}
-rand_id2 <- sample(unique(nhanes_pred_fgfpca$id), 4)
+rand_id3 <- sample(unique(nhanes_pred_fgfpca$id), 4)
 
@@
   nhanes_pred_fgfpca %>% 
-    filter(id %in% rand_id2) %>%
+    filter(id %in% rand_id3) %>%
     mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
@@
   nhanes_pred_adglmm %>% 
-    filter(id %in% rand_id2) %>%
+    filter(id %in% rand_id3) %>%
     mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
@@
 df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id2 & Z == 1) %>% 
+  filter(SEQN %in% rand_id3 & Z == 1) %>% 
   mutate(SEQN = as.factor(SEQN)) %>% 
@@
 df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id2 & sind >= 650 & sind <= 750) %>% 
+  filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>% 
   mutate(SEQN = as.factor(SEQN)) %>% 
@@
 df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id2 & sind >= 650 & sind <= 750) %>% 
+  filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>% 
   mutate(bin = cut(sind, breaks = seq(650, 750, by = 10), 

### 2023-12-05 Docs/manuscript.tex
small scale simulation 100 iterations

@@
 \author{Ying Jin}
-\date{2023-12-04}
+\date{2023-12-05}
 
@@
 
-\[g(E(Y_i(t))) = \sum_{k=1}^5\zeta_{k}B_k(t)+\sum_{l=1}^5\xi_{il}\phi_l(t)\]
+\[g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)\]
 
-I did the following things for data-reduction
-
-\begin{itemize}
-\item
-  Reduce sample size to 10 (for both training and testing)
-\item
-  Reduce measurement track by half (0.25-0.75)
-\item
-  Another way is the simple reduce the measurement density. For example,
-  instead of every minute, we can use every 10 minutes. In this case,
-  the binning step in fGFPCA would bin every 100 minutes in to one bin.
-\item
-  \begin{itemize}
-  \tightlist
-  \item
-    When we have smaller density of measurement grid, both methods could
-    run into numeric issues in both methods. For example, singularity
-    happened the the local GLMM step of fGFPCA procedure. This might be
-    able to get resoved by changing the optimizer (e.g.~Nealder\_Mead).
-    Do we want to do that? Or perhaps we simply want to use the full
-    track?
-  \end{itemize}
-\item
-  We may also need to reduce number of simulations (GLMMadaptive took
-  20-ish minutes for one simulaiton).
-\end{itemize}
-
-Below are some of the outcomes:
+I have used 100 subjects for training and testing. When fitting
+GLMMadpative, I reduce the number of measurements to 1/10 by taking one
+every 10 observations. The prediction is on the original grid.
 
@@
 \cline{2-5} \cline{6-9}
-Window & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65] & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65]\\
+Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
 \hline
-(0.35,0.45] & 35.724 &  &  &  & 48.779 &  &  & \\
+(0.2,0.4] & 154.178 &  &  &  & 387.140 &  &  & \\
 \hline
-(0.45,0.55] & 77.379 & 22.690 &  &  & 74.537 & 34.232 &  & \\
+(0.4,0.6] & 168.881 & 70.248 &  &  & 248.395 & 415.355 &  & \\
 \hline
-(0.55,0.65] & 32.581 & 20.370 & 20.043 &  & 48.683 & 35.666 & 18.303 & \\
+(0.6,0.8] & 235.562 & 47.770 & 16.813 &  & 330.585 & 321.804 & 265.975 & \\
 \hline
-(0.65, 0.75] & 158.191 & 135.968 & 130.360 & 66.648 & 174.158 & 141.285 & 143.370 & 103.92\\
+(0.8, 1.0] & 97.413 & 77.682 & 19.947 & 12.015 & 186.022 & 525.561 & 349.314 & 116.319\\
 \hline
@@
 \cline{2-5} \cline{6-9}
-Window & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65] & (0.25,0.35] & (0.25,0.45] & (0.25,0.55] & (0.25,0.65]\\
+Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
 \hline
-(0.35,0.45] & 0.657 &  &  &  & 0.643 &  &  & \\
+(0.2,0.4] & 0.721 &  &  &  & 0.646 &  &  & \\
 \hline
-(0.45,0.55] & 0.611 & 0.690 &  &  & 0.624 & 0.702 &  & \\
+(0.4,0.6] & 0.658 & 0.725 &  &  & 0.620 & 0.665 &  & \\
 \hline
-(0.55,0.65] & 0.687 & 0.710 & 0.714 &  & 0.666 & 0.686 & 0.702 & \\
+(0.6,0.8] & 0.710 & 0.793 & 0.805 &  & 0.682 & 0.700 & 0.733 & \\
 \hline
-(0.65, 0.75] & 0.604 & 0.639 & 0.636 & 0.653 & 0.598 & 0.593 & 0.587 & 0.62\\
+(0.8, 1.0] & 0.712 & 0.725 & 0.761 & 0.764 & 0.647 & 0.582 & 0.657 & 0.725\\
 \hline

### 2023-12-06 Docs/manuscript.Rmd
small scale simulation

@@
 
-fig_adglmm_subset <- df_subset_adglmm %>% filter(id %in% rand_id2) %>%
+fig_adglmm_subset <- pred_subset_adglmm[[1]] %>% 
+  filter(id %in% rand_id2) %>%
   mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
@@
 
-```{r}
+```{r fit_sim_small, fig.height=4, fig.width=8}
 ggarrange(fig_fgfpca_subset, fig_adglmm_subset,
@@
 ```{r ise_GLMMadaptive_subset}
-ise_mat_subset2 <- df_subset_adglmm %>%
+for(m in 1:M2){
+  this_df <- pred_subset_adglmm[[m]]
+  ise_tb_m <- this_df %>%
     mutate(err1 = (pred0.2-eta_i)^2,
@@
     filter(window != "[0,0.2]") %>% 
-    select(starts_with("err"))
-colnames(ise_mat_subset2) <- c("0.2", "0.4", "0.6", "0.8")
+    select(starts_with("err")) %>% as.matrix()
+  ise_mat_subset_ref[,,m] <- ise_tb_m
+}
+
+ise_mat_subset_ref <- apply(ise_mat_subset_ref, c(1, 2), mean)
+
+colnames(ise_mat_subset_ref) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
-           ise_mat_subset, ise_mat_subset2, check.names = F) %>%
-  kable(digits = 3, caption = "Integrated squared error") %>%
+           ise_mat_subset, ise_mat_subset_ref, check.names = F) %>%
+  kable(digits = 3, caption = "Integrated squared error", booktabs = T) %>%
   kable_styling(full_width = F) %>% 
@@
 ```{r auc_GLMMadaptive_subset}
-auc_mat_subset2 <- df_subset_adglmm %>%
+for(m in 1:M2){
+  this_df <- pred_subset_adglmm[[m]]
+
+  auc_tb_m <- this_df %>%
   mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
@@
   filter(window != "[0,0.2]") %>% 
-  select(starts_with("auc"))
+  select(starts_with("auc")) %>% as.matrix()
 
-colnames(auc_mat_subset2) <- c("0.2", "0.4", "0.6", "0.8")
+  auc_mat_subset_ref[, , m] <- auc_tb_m
+
+}
+
+auc_mat_subset_ref <- apply(auc_mat_subset_ref, c(1,2), mean)
+colnames(auc_mat_subset_ref) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
-           auc_mat_subset, auc_mat_subset2, check.names = F) %>%
-  kable(digits = 3, caption = "Integrated squared error") %>%
+           auc_mat_subset, auc_mat_subset_ref, check.names = F) %>%
+  kable(digits = 3, caption = "Area under the ROC curve", booktabs = T) %>%
   kable_styling(full_width = F) %>% 
@@
 There is huge difference in compuation time. GLMMadaptive took 23 minutes, while fGFPCA took less than 3 seconds.
+
+```{r}
+data.frame(
+  "Method" = c("fGFPCA", "GLMMadaptive"),
+  "Fit" = c(mean(fit_time_subset_fGFPCA)/60, mean(fit_time_subset_adglmm)),
+           "Prediction"= c(mean(pred_time_subset_fGFPCA), mean(pred_time_subset_adglmm)/60)) %>%
+  kable(digits = 3, caption = "Computation time (minutes)", booktabs = T) %>%
+  kable_styling(full_width = F) 
+```
 

### 2023-12-06 Docs/manuscript.tex
small scale simulation

@@
 \author{Ying Jin}
-\date{2023-12-05}
+\date{2023-12-06}
 
@@
 
-\hypertarget{small-scale-simulation-one-iteration}{%
-\section{Small-scale simulation: one
-iteration}\label{small-scale-simulation-one-iteration}}
+\hypertarget{small-scale-simulation}{%
+\section{Small-scale simulation}\label{small-scale-simulation}}
 
@@
 
-I have used 100 subjects for training and testing. When fitting
-GLMMadpative, I reduce the number of measurements to 1/10 by taking one
-every 10 observations. The prediction is on the original grid.
+I have used 100 subjects for training and testing, and repeated 100
+times. When fitting GLMMadpative, I reduce the number of measurements to
+1/10 by taking one every 10 observations. The prediction is on the
+original grid.
 
@@
 
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}
+\includegraphics{manuscript_files/figure-latex/fit_sim_small-1.pdf}
 
@@
 \centering
-\begin{tabular}[t]{l|r|r|r|r|r|r|r|r}
-\hline
-\multicolumn{1}{c|}{ } & \multicolumn{8}{c}{Observation track} \\
-\cline{2-9}
-\multicolumn{1}{c|}{ } & \multicolumn{4}{c|}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cline{2-5} \cline{6-9}
+\begin{tabular}[t]{lrrrrrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Observation track} \\
+\cmidrule(l{3pt}r{3pt}){2-9}
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
 Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
-\hline
-(0.2,0.4] & 154.178 &  &  &  & 387.140 &  &  & \\
-\hline
-(0.4,0.6] & 168.881 & 70.248 &  &  & 248.395 & 415.355 &  & \\
-\hline
-(0.6,0.8] & 235.562 & 47.770 & 16.813 &  & 330.585 & 321.804 & 265.975 & \\
-\hline
-(0.8, 1.0] & 97.413 & 77.682 & 19.947 & 12.015 & 186.022 & 525.561 & 349.314 & 116.319\\
-\hline
+\midrule
+(0.2,0.4] & 151.709 &  &  &  & 374.368 &  &  & \\
+(0.4,0.6] & 189.147 & 77.904 &  &  & 268.805 & 464.796 &  & \\
+(0.6,0.8] & 226.314 & 51.773 & 17.050 &  & 321.530 & 286.806 & 230.564 & \\
+(0.8, 1.0] & 113.007 & 81.718 & 19.576 & 13.244 & 227.965 & 472.162 & 341.146 & 136.831\\
+\bottomrule
 \end{tabular}
@@
 
-\caption{\label{tab:unnamed-chunk-9}Integrated squared error}
+\caption{\label{tab:unnamed-chunk-10}Area under the ROC curve}
 \centering
-\begin{tabular}[t]{l|r|r|r|r|r|r|r|r}
-\hline
-\multicolumn{1}{c|}{ } & \multicolumn{8}{c}{Observation track} \\
-\cline{2-9}
-\multicolumn{1}{c|}{ } & \multicolumn{4}{c|}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cline{2-5} \cline{6-9}
+\begin{tabular}[t]{lrrrrrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Observation track} \\
+\cmidrule(l{3pt}r{3pt}){2-9}
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
 Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
-\hline
-(0.2,0.4] & 0.721 &  &  &  & 0.646 &  &  & \\
-\hline
-(0.4,0.6] & 0.658 & 0.725 &  &  & 0.620 & 0.665 &  & \\
-\hline
-(0.6,0.8] & 0.710 & 0.793 & 0.805 &  & 0.682 & 0.700 & 0.733 & \\
-\hline
-(0.8, 1.0] & 0.712 & 0.725 & 0.761 & 0.764 & 0.647 & 0.582 & 0.657 & 0.725\\
-\hline
+\midrule
+(0.2,0.4] & 0.745 &  &  &  & 0.674 &  &  & \\
+(0.4,0.6] & 0.663 & 0.733 &  &  & 0.623 & 0.671 &  & \\
+(0.6,0.8] & 0.712 & 0.789 & 0.802 &  & 0.691 & 0.700 & 0.732 & \\
+(0.8, 1.0] & 0.741 & 0.755 & 0.782 & 0.785 & 0.680 & 0.627 & 0.679 & 0.743\\
+\bottomrule
 \end{tabular}
@@
 minutes, while fGFPCA took less than 3 seconds.
+
+\begin{table}
+
+\caption{\label{tab:unnamed-chunk-11}Computation time (minutes)}
+\centering
+\begin{tabular}[t]{lrr}
+\toprule
+Method & Fit & Prediction\\
+\midrule
+fGFPCA & 0.043 & 1.405\\
+GLMMadaptive & 5.842 & 0.023\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-\caption{\label{tab:unnamed-chunk-10}Area Under the ROC curve}
+\caption{\label{tab:unnamed-chunk-12}Area Under the ROC curve}
 \centering
@@
 
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1.pdf}
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-13-1.pdf}
 

### 2023-12-18 Docs/NhanesApplExample.Rmd
example of issue with debias step in NHNES

---
title: "Problem of Step 4 (debias) of fGFPCA on NHANES data"
author: "Ying Jin"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(121)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```

```{r}
# load data
df <- read_rds(here("Data/nhanes_bi.rds"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 4
source(here("Code/OutsampBayes.R"))
```

Fit fGFPCA model

Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 500 participants as the training set

```{r data_subset}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)

Ntr <- 1000

train_id <- unique(df$id)[1:Ntr]
train_df <- df %>% filter(id %in% train_id)

# additional parameters
J <- length(unique(df$sind))
```

# Step 1: Bin data

- Bin every 10 observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))
```

# Step 2:  Local GLMM

```{r}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
# t2-t1 
```

```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
rand_id <- sample(train_id, 4)
df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

# Step 3: FPCA

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=Ntr, 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 4 PCs
t2 <- Sys.time()
# t2-t1 
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
  pivot_longer(2:5, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

- Eigenvalues: `r fpca_mod$evalues`
- Proportion of variance explained with four PCs: `r fpca_mod$pve`

# Step 4: Re-evaluation

## Extend the eigenfunctions back to the original grid with B-spline basis

```{r interplate}
# Re-evaluation
## grid extension
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design  # evaluate B-splines on original grid

df_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  df_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid
```

Below presents the result of re-evaluation. Points are values of PC functions on the binned grid estimated from FPCA, and lines are values of re-evaluated PC functions on the original measurement grid.

```{r}
df_pc1 <- data.frame(t=1:J,  df_phi) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC")

df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC")

left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
  ggplot()+
  geom_line(aes(x=t, y=value.x, col = PC), linewidth = 0.2)+
  geom_point(aes(x=t, y=value.y, col = PC), na.rm = T, alpha = 0.5, size = 0.5)+
  labs(x="Time", y="", title = "Re-evaluate eigenfunctions on the original grid")

```

## Debias the eigenvalues with GLMM

```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1
```

I ran into a few problems at this step. Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues. They also lost the decreasing nature. Specifically:

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`

The FPCA eigenvalues, as expected, decreases from PC1 to PC4. However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3. Also, the proportion of these two set of estimates is not constant, as below:

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:5])
```

### 2023-12-18 Docs/manuscript.Rmd
example of issue with debias step in NHNES

@@
 output:
-  pdf_document:
-    toc: yes
-    toc_depth: '3'
-    keep_tex: yes
   html_document:
@@
     font: 12pt
+    keep_tex: yes
+  pdf_document:
+    toc: yes
+    toc_depth: '3'
     keep_tex: yes
@@
   labs(title = "fGFPCA", col = "Maximum observation time")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5))
 
@@
   labs(title = "GLMMadaptive", col = "Maximum observation time")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
-
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5))
 ```
@@
 
-We take 80% (7010) subjects for training, 20% (1753) for out-of-sample prediction.
+We take 500 subjects for training, 500 for out-of-sample prediction.
+When using the full sample The debiase step (mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only works for intercept-only model.
+In the debias step, re-evaluated eigenvalues also lost their decreasing nature.
 
-In the GLMMadaptive model, I could only fit a model with random intercept. It took more than 20 minutes for model fitting.
+In addition to reduce sample size, maybe we could just drop the debias step and only interpolate the eigenfunctions?
 
@@
 
-```{r small_exp}
-# a smaller example of binary functions
-df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id3 & Z == 1) %>% 
-  mutate(SEQN = as.factor(SEQN)) %>% 
-  ggplot()+
-  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
-  labs(x="Time", y="Subject", title = "NHANES binary activity indicator")
-```
-
 # Additional figures
@@
 ```
+
+## Example of continuous and binary NHANES
+
+```{r}
+df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds"))
+p_cont <- df_nhanes_cont %>% 
+  filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>% 
+  select(starts_with("min"), SEQN) %>%
+  pivot_longer(1:1440, names_to = "Minute") %>% 
+  mutate(Minute = gsub("MIN", "", Minute)) %>%
+  mutate(Minute = as.numeric(Minute),
+         SEQN=as.factor(SEQN)) %>%
+  ggplot(aes(x=Minute, y=value, col=SEQN))+
+  geom_point(size = 0.5, alpha=0.5)+
+  geom_smooth(se=F, size = 0.5)+
+  scale_x_continuous(breaks = seq(0, 1440, by = 480),
+                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
+  theme(axis.text = element_text(size = 5))+
+  labs(y="", title = "NHANES Activitiy count (2003-2004)")
+```
+
+```{r}
+# a example of continuous and binary NHANES data
+p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% rand_id3 & Z == 1) %>% 
+  mutate(SEQN = as.factor(SEQN)) %>% 
+  ggplot()+
+  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
+  scale_x_continuous(breaks = seq(0, 1440, by = 480),
+                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
+  theme(axis.text = element_text(size = 5))+
+  labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)")
+```
+
+```{r data_exp, fig.height=4, fig.width=10}
+ggarrange(p_cont, p_bi, nrow = 1)
+```
+
+## Dynamic prediction example
+
+```{r dynpred_exp,  fig.height=4, fig.width=10}
+ggarrange(
+  pred_list_all[[1]] %>% 
+    filter(id==501) %>%
+    mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>%
+    ggplot()+
+    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
+    geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T, 
+              linetype = "dashed")+
+    geom_vline(xintercept = 0.2, linetype = "dashed")+
+    labs(x="Time", y="")+
+    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+    theme(axis.text = element_text(size = 5))+
+    ylim(-1.2, 1.7),
+
+ pred_list_all[[1]] %>% 
+    filter(id==501) %>%
+    mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>%
+    ggplot()+
+    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
+    geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T, 
+              linetype = "dashed")+
+    geom_vline(xintercept = 0.4, linetype = "dashed")+
+    labs(x="Time", y="")+
+    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+    theme(axis.text = element_text(size = 5))+
+   ylim(-1.2, 1.7),
+ nrow = 1, common.legend = T
+  )
+
+```

### 2023-12-18 Code/Sim_Subset_fGFPCA.R

@@ -15,6 +15,7 @@ library(gridExtra)
 library(kableExtra)
 library(LaplacesDemon)
 library(mvtnorm)
+library(splines)
 
 
 
@@ -23,7 +24,7 @@ library(mvtnorm)
 
 load(here("Data/sim_data.RData"))
 
-M <- 100 # number of simulations
+M <- 500 # number of simulations
 
 # reduce data
 sim_data <- lapply(sim_data[1:M], function(x){x %>% filter(id %in% c(1:200))})

### 2023-12-19 Docs/NhanesApplExample.Rmd
slight modification
try with three PCs

@@
 # use pred_latent function to estimate latent function 
-K <- 4
+K <- 3
 source(here("Code/OutsampBayes.R"))
@@
 
-Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 500 participants as the training set
+Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 1000 participants as the training set
 
@@
 
+Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.
+
 ```{r}
@@
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-# keep 4 PCs
+# keep 3 PCs
 t2 <- Sys.time()
 # t2-t1
+
+# fpca_mod$evalues
 ```
@@
 ```{r PC}
-data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
-  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
-  pivot_longer(2:5, names_to = "PC") %>%
+data.frame(t = mid, fpca_mod$efunctions) %>%
+  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
+  pivot_longer(2:4, names_to = "PC") %>%
   ggplot()+
@@
 df_pc1 <- data.frame(t=1:J,  df_phi) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
-  pivot_longer(2:5, names_to = "PC")
+  rename(PC1=X1, PC2=X2, PC3=X3) %>%
+  pivot_longer(2:4, names_to = "PC")
 
 df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
-  pivot_longer(2:5, names_to = "PC")
+  rename(PC1=X1, PC2=X2, PC3=X3) %>%
+  pivot_longer(2:4, names_to = "PC")
 
@@
 ## Debias the eigenvalues with GLMM
+
+Again, here we use the three PC functions in the random effect part.
 
@@
                      s(id, by=phi2, bs="re")+
-                     s(id, by=phi3, bs="re")+
-                     s(id, by=phi4, bs="re"), 
+                     s(id, by=phi3, bs="re"),
+                     # s(id, by=phi4, bs="re"), 
                    family = binomial, 
@@
 t2 <- Sys.time()
-# t2-t1
+# t2-t1 # 3.5 minutes
 ```
 
-I ran into a few problems at this step.
-Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues.
-They also lost the decreasing nature.
-Specifically:
+It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically
 
-- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
-- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
-- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`
+- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:4]`
+- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:4]`
+- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:3]`
 
-The FPCA eigenvalues, as expected, decreases from PC1 to PC4.
-However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3.
 Also, the proportion of these two set of estimates is not constant, as below:
@@
 # proportion of biased eigenvalues to re-evaluated eigenvalues
-fpca_mod$evalues/(1/debias_glmm$sp[2:5])
+fpca_mod$evalues/(1/debias_glmm$sp[2:4])
 ```

### 2023-12-19 Docs/ProgressReportAppl.Rmd
try with three PCs

@@
 
+- 60% subjects for training, 40% for out-of-sample prediction
 - Original sample size took too much time in the debias step of fGFPCA method. 
 - In fact, the debais step does not work very well in this case
-- 500 subjects for training, 500 for out-of-sample prediction
 
@@
 
-Ntr <- Nte <- 500
+Ntr <- round(N*0.6)
+Nte <- N - Ntr
 
-train_id <- unique(df$id)[1:500]
-test_id <- unique(df$id)[501:1000]
+train_id <- sample(unique(df$id), Ntr)
+train_df <- df %>% filter(id %in% train_id)
 
-train_df <- df %>% filter(id %in% train_id)
-test_df <- df %>% filter(id %in% test_id)
+test_df <- df %>% filter(!id %in% train_id)
+test_id <- unique(test_df$id)
+
 # 1440*500
@@
 t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
-
-# dim(fpca_mod$efunctions) # 27 eigenfunctions
-K <- 4
 ```
@@
 t2 <- Sys.time()
-t2-t1
+t2-t1 # 20 hours
+
+save(debias_glmm, train_df, test_df,
+      file = here("Data/Appl_debias_model.RData"))
+
 ```

### 2023-12-22 Docs/manuscript.Rmd
small scale simulation 500 iterations

@@
 ```{r}
-M2 <- length(pred_subset_fGFPCA)
+# M2 <- length(pred_subset_fGFPCA)
 ## ISE container 
-ise_mat_subset <- ise_mat_subset_ref <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M2))
+ise_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, length(pred_subset_fGFPCA)))
+
+pred_subset_adglmm <- pred_subset_adglmm[num_probs==0]
+ise_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, length(pred_subset_adglmm)))
+
 ```
@@
 ```{r ise_fGFPCA_subset}
-for(m in 1:M2){
+for(m in 1:length(pred_subset_fGFPCA)){
   this_df <- pred_subset_fGFPCA[[m]]
@@
 ```{r ise_GLMMadaptive_subset}
-for(m in 1:M2){
+for(m in 1:length(pred_subset_adglmm)){
   this_df <- pred_subset_adglmm[[m]]
@@
 ## auc container 
-auc_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, M2))
-auc_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, M2))
+auc_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, 
+                                    length(pred_subset_fGFPCA)))
+auc_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, 
+                                        length(pred_subset_adglmm)))
 ```
@@
 ```{r auc_fGFPCA_subset}
-for(m in 1:M2){
+for(m in 1:length(pred_subset_fGFPCA)){
   this_df <- pred_subset_fGFPCA[[m]]
@@
 ```{r auc_GLMMadaptive_subset}
-for(m in 1:M2){
+for(m in 1:length(pred_subset_adglmm)){
   this_df <- pred_subset_adglmm[[m]]
@@
 
-There is huge difference in compuation time. GLMMadaptive took 23 minutes, while fGFPCA took less than 3 seconds.
+Among 500 hundred iterations, `r sum(num_probs)` did not converge for the GLMMadaptive model. No numeric issue for fGFPCA.
 
@@
   "Method" = c("fGFPCA", "GLMMadaptive"),
-  "Fit" = c(mean(fit_time_subset_fGFPCA)/60, mean(fit_time_subset_adglmm)),
-           "Prediction"= c(mean(pred_time_subset_fGFPCA), mean(pred_time_subset_adglmm)/60)) %>%
+  "Fit" = c(mean(fit_time_subset_fGFPCA)/60, mean(fit_time_subset_adglmm, na.rm=T)),
+           "Prediction"= c(mean(pred_time_subset_fGFPCA), mean(pred_time_subset_adglmm, na.rm=T)/60)) %>%
   kable(digits = 3, caption = "Computation time (minutes)", booktabs = T) %>%

### 2023-12-22 Code/Sim_Subset_GLMMadaptive.R

@@ -24,7 +24,7 @@ load(here("Data/sim_data.RData"))
 
 # reduce data
 ## training data
-M <- 100 # number of simulations
+M <- 500 # number of simulations
 
 sim_data <- lapply(sim_data[1:M], function(x){x %>% filter(id %in% c(1:200))})
 lapply(sim_data, dim)
@@ -52,74 +52,80 @@ t_bin <- seq(0, J, by = 10)
 pred_list_all <- list()
 fit_time <- pred_time <- rep(NA, M)
 
+# incorporate a convergen check since some iteration would run into numeric issues
+# such as singularity
+# we skip such datasets, saving results only of the numeric-free datsets
+num_probs <- rep(NA, M)
+
+
 #### GLMMadaptive ####
 
 # The simulation process run into a numeric problem at the 95th iteration
 # (singularity)
 
-pb = txtProgressBar(min = 95, max = M, initial = 0, style = 3) 
-for(m in 95:M){
+pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
+for(m in 1:M){
   
   # data split and training data reduction
-  try_fit <- NA
+  df_train_m <- sim_data[[m]] %>% filter(id %in% 1:100 & sind %in% t_bin) 
   df_test_m <- sim_data[[m]] %>% filter(!id %in% 1:100)
   
-  # model fit
-  while(is.na(try_fit)){
-    df_train_m <- sim_data[[m]] %>% filter(id %in% 1:100 & sind %in% t_bin) 
-    try_fit <- tryCatch(expr={t1 <- Sys.time()
-                              fit_adglmm <- mixed_model(fixed = Y ~ 0+ns(t, df = 4), 
-                                                        random = ~ 0+ns(t, df = 4) | id, 
-                                                        data =  df_train_m, family = binomial())
-                              t2 <- Sys.time()},
-                        error = function(e){NA}
-                        )
-    t_bin <- t_bin+1
+  # model fit with a numeric check
+  t1 <- Sys.time()
+  fit_adglmm <- tryCatch(expr={mixed_model(fixed = Y ~ 0+ns(t, df = 4), 
+                                           random = ~ 0+ns(t, df = 4) | id, 
+                                           data =  df_train_m, family = binomial())},
[…]

### 2024-01-02 Docs/manuscript.Rmd
edit manuscript progress report

@@
 
-```{r binning}
+```{r binning, eval=FALSE}
 df_nhanes %>% select(SEQN, Z, sind) %>% 
@@
 
-```{r}
+```{r,eval=FALSE}
 df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>% 
@@
 
-```{r}
+```{r, eval=FALSE}
 df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds"))
@@
 
-```{r}
+```{r, eval=FALSE}
 # a example of continuous and binary NHANES data
@@
 
-```{r data_exp, fig.height=4, fig.width=10}
+```{r data_exp, fig.height=4, fig.width=10, eval=FALSE}
 ggarrange(p_cont, p_bi, nrow = 1)
@@
 
-```{r dynpred_exp,  fig.height=4, fig.width=10}
+```{r dynpred_exp,  fig.height=4, fig.width=10, eval=FALSE}
 ggarrange(

### 2024-01-02 Code/Sim_Subset_GLMMadaptive.R

@@ -132,6 +132,7 @@ close(pb)
 
 mean(fit_time)
 mean(pred_time)
+sum(num_probs)
 
 pred_list_all[[96]] %>% filter(t>0.2) %>% View()
 pred_list_all[[99]] %>% filter(t>0.2) %>% View()
@@ -143,6 +144,7 @@ pred_time_subset_adglmm <- pred_time
 pred_subset_adglmm <- pred_list_all
 
 save(fit_time_subset_adglmm, pred_time_subset_adglmm, pred_subset_adglmm, 
+     num_probs,
      file = here("Data/SubSimOutput_GLMMadaptive.RData"))
 
 

### 2024-01-03 Docs/manuscript.Rmd
try to fit generalized FOF model

--- 2023-12-22 Docs/manuscript.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = TRUE)
 options(knitr.kable.NA = "")
@@
 library(splines)
+library(boot)
 theme_set(theme_minimal())
@@
 
-# Additional figures
+# The second reference method:
 
-## fGFPCA algorithm demonstration
+We would like to use a second reference method for predictive performance comparision.
+This would be a generalized function-on-function regression where the history records are used to predict a equal length of outcome.
+Formulation as follows:
 
-```{r binning}
-df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>% 
-  mutate(SEQN = as.factor(SEQN)) %>% 
-  ggplot()+
-  geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+
-  geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+
-  scale_y_continuous(breaks = c(0, 1))+
-  labs(x="Time", y="", title = "NHANES binary activity indicator")
+$$g(E[Y_i(t)]) = \beta_0(t) + \beta_1 (t)Y_i(s),\ s<t$$
+
+In this case, the outcome must be evaluated at the same time points as the covariate.
+That is, s and t must have the same number of measurement points.
+For example, if t is (0, 0.2], then s can be either (0.2, 0.4), (0.4, 0.6), (0.6, 0.8) or (0.8, 1).
+
+What to do when the history t is (0, 0.4)?
+How share we prediction on 0.2 windows including (0.4, 0.6), (0.6, 0.8) or (0.8, 1)?
+Does this mean we do not utilize the full observed track but instead only take the most recent subtrack (0.2, 0.4)?
+Such scenarios remains unclear.
+
+As for the model fitting, I am not sure how to construct a function-on-function regression using bam or gam.
+While most functions I found in R does not account for the "generalized" part.
+At first glance, the model framework is the same as a regular functional regression model:
+
+$$g(E[Y_i(t)]) = \sum_{k}\xi_{k}\phi_k(t) + Y_i(t)\sum_{l}\zeta_{l}\phi_l(s),\ s<t$$
+
+so I fit a gam model where history is [0, 0.2) and future is [0.2, 0.4).
+
+```{r data_format, message=FALSE}
+load(here("Data/sim_data.RData"))
+
+# Let's start with a one-simulation example
+df_m <- sim_data[[1]]
+# class(df_m$Y)
+# df_m$Y <- as.factor(df_m$Y)
+
+# observation/prediction window 
+windows <- seq(0, 1, by = 0.2)
+df_m$window <- cut(df_m$t, breaks=windows, labels = 1:5,
+                       include.lowest = T)
+
+# split data
+N_train <- 500
+N_test <- 100
+train_df <- df_m %>% filter(id %in% 1:N_train)
+test_df <- df_m %>% filter(!id %in% 1:N_train)
+
+# put covariates and outcome together
+dfm1_tr <- train_df %>% filter(window==2) %>% 
+  rename(Yf=Y, tf=t) %>%
+  left_join(train_df %>% filter(window==1) %>% select(id, sind, t, Y) %>% 
+              mutate(sind=sind+200) %>%
+              rename(Yh=Y, th = t))
+
+dfm1_te <- test_df %>% filter(window==2) %>% 
+  rename(Yf=Y) %>%
+  left_join(test_df %>% filter(window==1) %>% select(id, sind, Y) %>% 
+              mutate(sind=sind+200) %>%
+              rename(Yh=Y))
+
+# class(dfm1_tr$Yh)
 ```
 
+```{r class.source='fold-show'}
+fitm1 <- bam(Yf ~ s(tf, bs="cc", k=20) + s(tf, bs="cc", k=20, by = Yh) , 
+                   family = binomial, data=dfm1_tr, 
+                   method = "fREML",
+                   discrete = TRUE)
+summary(fitm1)
+```
+
+However, the prediction result of this model is very wired.
+Below is the in-sample estimation of latent process.
+I am not sure if this is because of a wrong design matrix or ignoring the individual random effect.
+
 ```{r}
-df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>% 
-  mutate(bin = cut(sind, breaks = seq(650, 750, by = 10), 
-                   labels = seq(655, 745, by = 10),
-                   include.lowest = T)) 
-bin_mid <- seq(655, 745, by = 10)
+dfm1_tr$eta_est <- predict(fitm1, newdata = dfm1_tr, type = "link")
+dfm1_tr %>% filter(id %in% 1:4) %>% 
+  mutate_at(vars(eta_i, eta_est), inv.logit) %>%
+  ggplot()+
+  geom_line(aes(x=tf, y=eta_i, col = "true"))+
+  geom_line(aes(x=tf, y=eta_est, col = "pred"))+
+  facet_wrap(~id)
+```
 
-for(i in bin_mid){
-  df_bin_i <- df_fig %>% filter(bin==i) 
-  glm_i <- glmer(Z ~ 1 + (1|SEQN), data = df_bin_i, family = binomial, nAGQ=0)
-  eta_i <- predict(glm_i, type = "link")
+Thinking about the two problems above, perhaps I would want to use the following framework instead.
+This framework doesn't require same density of s and t.
+But I am not sure how to construct the design matrix for this.
 
-  df_fig$eta_hat[df_fig$bin==i] <- eta_i
-}
+$$g(E[Y_i(t)]) = \beta_0(t) + \int_{s}\beta_1(t, s)Y_i(s), \ s<t$$
 
-df_fig %>% 
-  mutate(SEQN = as.factor(SEQN),
-         eta_hat = exp(eta_hat)/(1+exp(eta_hat)),
-         bin=as.numeric(as.character(bin))) %>%
+or
 
-  ggplot()+
-  geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+
-  geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+
-  geom_point(aes(x=bin, y=eta_hat, col = SEQN))+
-  geom_line(aes(x=bin, y=eta_hat, col = SEQN))+
-  scale_y_continuous(breaks = c(0, 1))+
-  labs(x="Time", y="", title = "NHANES binary activity indicator")
+$$g(E[Y_i(t)]) = \beta_0(t) + F(Y_i(s), s, t), \ s<t$$
+
+Perhaps I can try using [refund::ff](https://rdrr.io/cran/refund/man/ff.html)
+
+```{r class.source='fold-show', error=TRUE}
+library(refund)
+
+th <- unique(dfm1_tr$th)
+tf <- unique(dfm1_tr$tf)
+
+Yhmat <- matrix(dfm1_tr$Yh, nrow=N_train, ncol = length(th), byrow = T)
+
+Yhmat2 <- Yhmat[rep(1:N_train, each = length(th)), ]
+Yf <- dfm1_tr$Yf
+# length(Yf)
+# dim(Yhmat2)
+
+fitm1 <- fgam(Yf ~ s(tf, bs="cc", k=10) + af(Yhmat2, argvals=th) , 
+                   family = binomial())
+# I couldn't get this function to work
+# I think for each observation in the future, all of the history is used as predictor
+# therefore the whole historical function is repeatedly used for each subject?
 
@@
 
-## Example of continuous and binary NHANES
+# Additional figures
 
-```{r}
-df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds"))
-p_cont <- df_nhanes_cont %>% 
-  filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>% 
-  select(starts_with("min"), SEQN) %>%
-  pivot_longer(1:1440, names_to = "Minute") %>% 
-  mutate(Minute = gsub("MIN", "", Minute)) %>%
-  mutate(Minute = as.numeric(Minute),
-         SEQN=as.factor(SEQN)) %>%
-  ggplot(aes(x=Minute, y=value, col=SEQN))+
-  geom_point(size = 0.5, alpha=0.5)+
-  geom_smooth(se=F, size = 0.5)+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480),
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
-  theme(axis.text = element_text(size = 5))+
-  labs(y="", title = "NHANES Activitiy count (2003-2004)")
-```
+<!-- ## fGFPCA algorithm demonstration -->
 
-```{r}
-# a example of continuous and binary NHANES data
-p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% rand_id3 & Z == 1) %>% 
-  mutate(SEQN = as.factor(SEQN)) %>% 
-  ggplot()+
-  geom_point(aes(x=sind, y = SEQN), size = 0.5)+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480),
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
-  theme(axis.text = element_text(size = 5))+
-  labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)")
-```
+<!-- ```{r binning, eval=FALSE} -->
+<!-- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
+<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
+<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
+<!--   ggplot()+ -->
+<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
+<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
+<!--   scale_y_continuous(breaks = c(0, 1))+ -->
+<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
+<!-- ``` -->
 
-```{r data_exp, fig.height=4, fig.width=10}
-ggarrange(p_cont, p_bi, nrow = 1)
-```
+<!-- ```{r,eval=FALSE} -->
+<!-- df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
+<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
+<!--   mutate(bin = cut(sind, breaks = seq(650, 750, by = 10),  -->
+<!--                    labels = seq(655, 745, by = 10), -->
+<!--                    include.lowest = T))  -->
+<!-- bin_mid <- seq(655, 745, by = 10) -->
 
-## Dynamic prediction example
+<!-- for(i in bin_mid){ -->
+<!--   df_bin_i <- df_fig %>% filter(bin==i)  -->
+<!--   glm_i <- glmer(Z ~ 1 + (1|SEQN), data = df_bin_i, family = binomial, nAGQ=0) -->
+<!--   eta_i <- predict(glm_i, type = "link") -->
 
-```{r dynpred_exp,  fig.height=4, fig.width=10}
-ggarrange(
-  pred_list_all[[1]] %>% 
-    filter(id==501) %>%
-    mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>%
-    ggplot()+
-    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
-    geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T, 
-              linetype = "dashed")+
-    geom_vline(xintercept = 0.2, linetype = "dashed")+
-    labs(x="Time", y="")+
-    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-    theme(axis.text = element_text(size = 5))+
-    ylim(-1.2, 1.7),
+<!--   df_fig$eta_hat[df_fig$bin==i] <- eta_i -->
+<!-- } -->
 
- pred_list_all[[1]] %>% 
-    filter(id==501) %>%
-    mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>%
-    ggplot()+
-    geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+
-    geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T, 
-              linetype = "dashed")+
-    geom_vline(xintercept = 0.4, linetype = "dashed")+
-    labs(x="Time", y="")+
-    scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-    theme(axis.text = element_text(size = 5))+
-   ylim(-1.2, 1.7),
- nrow = 1, common.legend = T
-  )
+<!-- df_fig %>%  -->
+<!--   mutate(SEQN = as.factor(SEQN), -->
+<!--          eta_hat = exp(eta_hat)/(1+exp(eta_hat)), -->
+<!--          bin=as.numeric(as.character(bin))) %>%  -->
 
-```
+<!--   ggplot()+ -->
+<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
+<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
+<!--   geom_point(aes(x=bin, y=eta_hat, col = SEQN))+ -->
+<!--   geom_line(aes(x=bin, y=eta_hat, col = SEQN))+ -->
+<!--   scale_y_continuous(breaks = c(0, 1))+ -->
+<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
+
+<!-- ``` -->
+
+<!-- ## Example of continuous and binary NHANES -->
+
+<!-- ```{r, eval=FALSE} -->
+<!-- df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds")) -->
+<!-- p_cont <- df_nhanes_cont %>%  -->
+<!--   filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>%  -->
+<!--   select(starts_with("min"), SEQN) %>% -->
+<!--   pivot_longer(1:1440, names_to = "Minute") %>%  -->
+<!--   mutate(Minute = gsub("MIN", "", Minute)) %>% -->
+<!--   mutate(Minute = as.numeric(Minute), -->
+<!--          SEQN=as.factor(SEQN)) %>% -->
+<!--   ggplot(aes(x=Minute, y=value, col=SEQN))+ -->
+<!--   geom_point(size = 0.5, alpha=0.5)+ -->
+<!--   geom_smooth(se=F, size = 0.5)+ -->
+<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
+<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
+<!--   theme(axis.text = element_text(size = 5))+ -->
+<!--   labs(y="", title = "NHANES Activitiy count (2003-2004)") -->
+<!-- ``` -->
+
+<!-- ```{r, eval=FALSE} -->
+<!-- # a example of continuous and binary NHANES data -->
+<!-- p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
+<!--   filter(SEQN %in% rand_id3 & Z == 1) %>%  -->
+<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
+<!--   ggplot()+ -->
+<!--   geom_point(aes(x=sind, y = SEQN), size = 0.5)+ -->
+<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
+<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
+<!--   theme(axis.text = element_text(size = 5))+ -->
+<!--   labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)") -->
+<!-- ``` -->
+
+<!-- ```{r data_exp, fig.height=4, fig.width=10, eval=FALSE} -->
+<!-- ggarrange(p_cont, p_bi, nrow = 1) -->
+<!-- ``` -->
+
+<!-- ## Dynamic prediction example -->
+
+<!-- ```{r dynpred_exp,  fig.height=4, fig.width=10, eval=FALSE} -->
+<!-- ggarrange( -->
+<!--   pred_list_all[[1]] %>%  -->
+<!--     filter(id==501) %>% -->
+<!--     mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>% -->
+<!--     ggplot()+ -->
+<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
+<!--     geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T,  -->
+<!--               linetype = "dashed")+ -->
+<!--     geom_vline(xintercept = 0.2, linetype = "dashed")+ -->
+<!--     labs(x="Time", y="")+ -->
+<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
+<!--     theme(axis.text = element_text(size = 5))+ -->
+<!--     ylim(-1.2, 1.7), -->
+
+<!--  pred_list_all[[1]] %>%  -->
+<!--     filter(id==501) %>% -->
+<!--     mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>% -->
+<!--     ggplot()+ -->
+<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
+<!--     geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T,  -->
+<!--               linetype = "dashed")+ -->
+<!--     geom_vline(xintercept = 0.4, linetype = "dashed")+ -->
+<!--     labs(x="Time", y="")+ -->
+<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
+<!--     theme(axis.text = element_text(size = 5))+ -->
+<!--    ylim(-1.2, 1.7), -->
+<!--  nrow = 1, common.legend = T -->
+<!--   ) -->
+
+<!-- ``` -->

### 2024-01-03 Code/Sim_GenFOF.R

@@ -66,26 +66,48 @@ test_df <- df_m %>% filter(!id %in% 1:N_train)
 # observed 0-0.2, predict 0.2-0.4, 0.4-0.6, 0.6-0.8, 0.8-1.0
 # remaining scenarios not clear. Let's wait for now. 
 
+
+# let me try a FOF regression first
 ## time window
 windows <- seq(0, 1, by = 0.2)
 train_df$window <- cut(train_df$t, breaks=windows, labels = 1:5,
                        include.lowest = T)
-train_df$window <- as.numeric(train_df$window)
-table(train_df$window, useNA = "always")
+class(train_df$window)
+table(train_df$window, useNA = "always") # each window has equal number of observations
 
-train_df %>%
-  mutate(sind = sind-200*(window-1)) %>% 
-  select(sind) %>% table(useNA = "always")
-
+# I would like to try a gam model for generalized FOF
+# first some data formatting
 ## observed 0-0.2, predict 0.2-0.4
+dfm1 <- train_df %>% filter(window==2) %>% 
+  rename(Yf=Y) %>%
+  left_join(train_df %>% filter(window==1) %>% select(id, sind, Y) %>% 
+              mutate(sind=sind+200) %>%
+              rename(Yh=Y))
 
-df_mt <- train_df %>%
-  filter(window == 1) %>%
-  select(id, Y) %>% rename(Y_w1 = Y) %>% 
-  mutate(Y)
-  
-  train_df %>% filter(window ==1 | window == 2) %>% 
-  pivot_wider(id_cols = id,
-              values_from = c(t, eta_i, sind, Y),
-              names_from = window)
-?gam
+## gam model
+fitm1 <- bam(Yf ~ s(t, bs="cc", k=10) + s(t, bs="cc", k=10, by = Yh) , 
+                   family = binomial, data=dfm1, 
+                   method = "fREML",
+                   discrete = TRUE)
+
+## prediction?
+test_df$window <- cut(test_df$t, breaks=windows, labels = 1:5,
+                       include.lowest = T)
+table(test_df$window, useNA = "always")
+dfm1_te <- test_df %>% filter(window==2) %>% 
+  rename(Yf=Y) %>%
+  left_join(test_df %>% filter(window==1) %>% select(id, sind, Y) %>% 
+              mutate(sind=sind+200) %>%
[…]

### 2024-01-16 Docs/NhanesApplExample.Rmd
time-varying variation of each PC in data application

--- 2023-12-18 Docs/NhanesApplExample.Rmd
@@
 
-set.seed(121)
+set.seed(1114)
 
@@
 
-train_id <- unique(df$id)[1:Ntr]
+train_id <- unique(df$id)[1:Ntr] # training sample size = 1000
 train_df <- df %>% filter(id %in% train_id)
@@
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-# keep 4 PCs
+# keep 3 PCs
 t2 <- Sys.time()
 # t2-t1
+
+# fpca_mod$evalues/sum(fpca_mod$evalues)
 ```
@@
 ```{r PC}
-data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
-  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
+data.frame(t = mid, fpca_mod$efunctions) %>%
+  rename(PC1 = X1, PC2=X2, PC3=X3, PC4=X4) %>%
   pivot_longer(2:5, names_to = "PC") %>%
@@
 
+Again, here we use the three PC functions in the random effect part.
+
 ```{r revaluation}
@@
 t2 <- Sys.time()
-# t2-t1
+# t2-t1 # 7.5 minutes
 ```
 
-I ran into a few problems at this step.
-Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues.
-They also lost the decreasing nature.
-Specifically:
+It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically
 
@@
 
-The FPCA eigenvalues, as expected, decreases from PC1 to PC4.
-However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3.
 Also, the proportion of these two set of estimates is not constant, as below:
@@
 ```
+
+### Percentage of variance explained by each PC
+
+We now that the variance of individual-specific random effects, in theory, is:
+
+\[\begin{aligned}
+Var(b_i(t)) & = Var(\sum_{k=1}^K \xi_{ik}\phi_k(t)) \\
+& = \sum_{k=1}
+^K\phi_k(t)^2\lambda_k
+\end{aligned}\]
+
+Here I'd like to compare the variantion for each PC calculated using estimated from the **biased** FPCA estimates and the supposedly **debiased** from the dibased GLMM model.
+
+```{r var_bi_fpca}
+df_pc_fpca <- data.frame(t = mid, fpca_mod$efunctions) %>%
+  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>% 
+  mutate(eval1=fpca_mod$evalues[1], 
+         eval2=fpca_mod$evalues[2], 
+         eval3=fpca_mod$evalues[3], 
+         eval4=fpca_mod$evalues[4]) %>%
+  mutate(var1 = eval1*PC1^2, var2 = eval2*PC2^2, var3 = eval3*PC3^2, var4 = eval4*PC4^2)
+
+```
+
+```{r var_bi_glmm}
+# re-evaluated eigenvalues
+evals_new <- 1/debias_glmm$sp[2:5]
+
+df_pc_glmm <- df_phi %>%
+  mutate(eval1=evals_new[1], 
+         eval2=evals_new[2], 
+         eval3=evals_new[3], 
+         eval4=evals_new[4]) %>%
+  mutate(var1 = eval1*phi1^2, var2 = eval2*phi2^2, var3 = eval3*phi3^2, var4 = eval4*phi4^2)
+```
+
+```{r pc_var}
+df_pc_fpca$est <- "FPCA"
+df_pc_glmm$est <- "Debias"
+
+bind_rows(
+  df_pc_fpca %>% select(t, starts_with("var"), est),
+  df_pc_glmm %>% select(sind, starts_with("var"), est) %>% rename(t=sind)
+) %>% 
+  rename(PC1=var1, PC2=var2, PC3=var3, PC4=var4) %>%
+  pivot_longer(2:5, names_to = "PC", values_to = "variance") %>%
+  ggplot()+
+  geom_line(aes(x=t, y=variance, col=est, group=est))+
+  facet_wrap(~PC)+
+  labs(title = "Time-varying variation of each PC")
+
+```
+
+Things I see from the plots above:
+
+1. PC1 basically has no effect in the earlier track. It explains variation later during the day
+2. PC2 seems to work primary at night and a bit earlier time interval from PC1
+3. PC2 and PC3 seems to have similar shape? In fact, I think PC4 has a similar shape as well
+
+I am also curious about how the individual scores change after debias
+
+```{r score_df}
+df_score_fpca <- data.frame(id = 1:Ntr, fpca_mod$scores) %>%
+  pivot_longer(2:5, names_to = "PC", values_to = "score") %>%
+  mutate(est = "FPCA")
+
+df_score_glmm <- data.frame(
+  id = 1:Ntr,
+  X1 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi1.")) %>% t(),
+  X2 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi2.")) %>% t(),
+  X3 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi3.")) %>% t(),
+  X4 = data.frame(t(debias_glmm$coefficients)) %>% select(starts_with("s.id..phi4.")) %>% t()
+  ) %>%
+  pivot_longer(2:5, names_to = "PC", values_to = "score") %>%
+  mutate(est = "Debias")
+
+```
+
+```{r}
+bind_rows(df_score_fpca, df_score_glmm) %>%
+  ggplot(aes(x=est, y=score, col=est))+
+  geom_boxplot()+
+  geom_jitter(size = 0.2, alpha=0.5)+
+  facet_wrap(~PC, scales = "free_y")+
+  labs(title = "Score of each PC")
+
+```

### 2024-01-16 Docs/manuscript.Rmd
time-varying variation of each PC in data application

@@
 
+```{r}
+load(here("Data/sim_data.RData"))
+df_m <- sim_data[[1]]
+df_m %>% filter(t>0.2 & t<=0.4)
+```
+
 We would like to use a second reference method for predictive performance comparision.
-This would be a generalized function-on-function regression where the history records are used to predict a equal length of outcome.
-Formulation as follows:
+In this madel, predictions on a specific interval are all made using the last few observations in the observed window.
+For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2($t_j: j=200, ...
+200-(L-1)$) as time-fixed covariate to predict any future time.
 
-$$g(E[Y_i(t)]) = \beta_0(t) + \beta_1 (t)Y_i(s),\ s<t$$
+Let's write out the model expression (with questionable notation). If the observations if up to $t_m$, then:
 
-In this case, the outcome must be evaluated at the same time points as the covariate.
-That is, s and t must have the same number of measurement points.
-For example, if t is (0, 0.2], then s can be either (0.2, 0.4), (0.4, 0.6), (0.6, 0.8) or (0.8, 1).
+\[\begin{aligned}
+g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
+l & = m,...m-(j-1)\\
+t &> t_m 
+\end{aligned}\]
 
-What to do when the history t is (0, 0.4)?
-How share we prediction on 0.2 windows including (0.4, 0.6), (0.6, 0.8) or (0.8, 1)?
-Does this mean we do not utilize the full observed track but instead only take the most recent subtrack (0.2, 0.4)?
-Such scenarios remains unclear.
+This is a simple function-on-scalar model with no random effect, meaning all subject with the same last observed outcome would have the same estimated/predicted latent track.
+This is clearly anti-intuitive.
+But adding random effects would make out-of-sample prediction impossible.
 
-As for the model fitting, I am not sure how to construct a function-on-function regression using bam or gam.
-While most functions I found in R does not account for the "generalized" part.
-At first glance, the model framework is the same as a regular functional regression model:
+Under this framework, for each dataset we need to fit four models:
 
-$$g(E[Y_i(t)]) = \sum_{k}\xi_{k}\phi_k(t) + Y_i(t)\sum_{l}\zeta_{l}\phi_l(s),\ s<t$$
+- Given 0-0.2, predict 0.2-1
+- Given 0-0.4, predict 0.4-1
+- Given 0-0.6, predict 0.6-1
+- Given 0-0.8, predict 0.8-1
 
-so I fit a gam model where history is [0, 0.2) and future is [0.2, 0.4).
+Below are some single dataset examples.
 
-```{r data_format, message=FALSE}
-load(here("Data/sim_data.RData"))
+## L = 1
 
-# Let's start with a one-simulation example
-df_m <- sim_data[[1]]
-# class(df_m$Y)
-# df_m$Y <- as.factor(df_m$Y)
+First let try to make prediction on an interval with only one historical observation (L = 1):
 
-# observation/prediction window 
+\[\begin{aligned}
+g(E[Y_i(t)]) &= \beta_0(t) +\beta_1(t) Y_i(t_{m})\\
+t &> t_m 
+\end{aligned}\]
+
+```{r data_format}
+L <- 1 # number of observations to use as predictor
+
+# prediction window 
 windows <- seq(0, 1, by = 0.2)
@@
 test_df <- df_m %>% filter(!id %in% 1:N_train)
-
-# put covariates and outcome together
-dfm1_tr <- train_df %>% filter(window==2) %>% 
-  rename(Yf=Y, tf=t) %>%
-  left_join(train_df %>% filter(window==1) %>% select(id, sind, t, Y) %>% 
-              mutate(sind=sind+200) %>%
-              rename(Yh=Y, th = t))
-
-dfm1_te <- test_df %>% filter(window==2) %>% 
-  rename(Yf=Y) %>%
-  left_join(test_df %>% filter(window==1) %>% select(id, sind, Y) %>% 
-              mutate(sind=sind+200) %>%
-              rename(Yh=Y))
-
-# class(dfm1_tr$Yh)
 ```
 
-```{r class.source='fold-show'}
-fitm1 <- bam(Yf ~ s(tf, bs="cc", k=20) + s(tf, bs="cc", k=20, by = Yh) , 
-                   family = binomial, data=dfm1_tr, 
+```{r fit_model, class.source='fold-show'}
+# format training data
+## given [0, 0.2], predictor would be
+y_obs_max <- train_df %>% filter(window==1) %>% arrange(desc(t)) %>%
+  group_by(id) %>%
+  slice(L) %>% select(id, Y) %>% rename(yl = Y)
+## to predict (0.2, 0.4), outcome:
+df_pred_tr <- train_df %>% filter(window!=1) %>%
+  left_join(y_obs_max, by = "id") %>% 
+  mutate_at(vars(Y, yl), as.factor)
+
+# model
+t1 <- Sys.time()
+fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
+                        s(t, bs="cc", k=20, by = yl),
+                   family = binomial, data=df_pred_tr, 
                    method = "fREML",
                    discrete = TRUE)
-summary(fitm1)
+t2 <- Sys.time()
+t2-t1
+
+summary(fit_gen_fosr)
 ```
 
-However, the prediction result of this model is very wired.
-Below is the in-sample estimation of latent process.
-I am not sure if this is because of a wrong design matrix or ignoring the individual random effect.
+```{r est_train}
+# estimation on the training set
+df_pred_tr$eta_est <- predict(fit_gen_fosr, type = "link")
 
-```{r}
-dfm1_tr$eta_est <- predict(fitm1, newdata = dfm1_tr, type = "link")
-dfm1_tr %>% filter(id %in% 1:4) %>% 
-  mutate_at(vars(eta_i, eta_est), inv.logit) %>%
+df_pred_tr %>%
+  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
+  filter(id %in% 1:4) %>%
   ggplot()+
-  geom_line(aes(x=tf, y=eta_i, col = "true"))+
-  geom_line(aes(x=tf, y=eta_est, col = "pred"))+
-  facet_wrap(~id)
+  geom_point(aes(x=t, y=Y), size = 0.5)+
+  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
+  geom_line(aes(x=t, y=eta_i, col = "ture"))+
+  facet_wrap(~id)+
+  labs(title = "In-sample estimation")
 ```
 
-Thinking about the two problems above, perhaps I would want to use the following framework instead.
-This framework doesn't require same density of s and t.
-But I am not sure how to construct the design matrix for this.
+```{r pred_test}
+# format testing data
+df_pred_te <- test_df %>% filter(window!=1) %>%
+  left_join(test_df %>% 
+              filter(window==1) %>% arrange(desc(t)) %>% group_by(id) %>%
+              slice(L) %>% select(id, Y) %>% rename(yl = Y),
+            by = "id") %>% 
+  mutate_at(vars(Y, yl), as.factor)
 
-$$g(E[Y_i(t)]) = \beta_0(t) + \int_{s}\beta_1(t, s)Y_i(s), \ s<t$$
+# predict using the FOSR model fit above
+df_pred_te$eta_est <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
 
-or
-
-$$g(E[Y_i(t)]) = \beta_0(t) + F(Y_i(s), s, t), \ s<t$$
-
-Perhaps I can try using [refund::ff](https://rdrr.io/cran/refund/man/ff.html)
-
-```{r class.source='fold-show', error=TRUE}
-library(refund)
-
-th <- unique(dfm1_tr$th)
-tf <- unique(dfm1_tr$tf)
-
-Yhmat <- matrix(dfm1_tr$Yh, nrow=N_train, ncol = length(th), byrow = T)
-
-Yhmat2 <- Yhmat[rep(1:N_train, each = length(th)), ]
-Yf <- dfm1_tr$Yf
-# length(Yf)
-# dim(Yhmat2)
-
-fitm1 <- fgam(Yf ~ s(tf, bs="cc", k=10) + af(Yhmat2, argvals=th) , 
-                   family = binomial())
-# I couldn't get this function to work
-# I think for each observation in the future, all of the history is used as predictor
-# therefore the whole historical function is repeatedly used for each subject?
+df_pred_te %>%
+  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.5)+
+  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
+  geom_line(aes(x=t, y=eta_i, col = "ture"))+
+  facet_wrap(~id)+
+  labs(title = "Out-of-sample prediction")
 

### 2024-01-16 Code/Data_Appl_NHANES.R

@@ -54,15 +54,15 @@ train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = m
 train_df$bin <- as.numeric(as.character(train_df$bin))
 
 # Step 2:  Local GLMM
-## fit model on the training set
 train_bin_lst <- split(train_df, f = train_df$bin)
+
 t1=Sys.time()
 df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
 t2= Sys.time()
 t2-t1 
 
-df_est_latent <- bind_rows(df_est_latent) 
 ## example estimated latent function
+df_est_latent <- bind_rows(df_est_latent) 
 rand_id <- sample(train_id, 4)
 df_est_latent %>% 
   filter(id %in% rand_id) %>%
@@ -76,58 +76,78 @@ df_est_latent %>%
 
 
 # FPCA
+
 uni_eta_hat <- df_est_latent %>% filter(bin==sind)
-head(uni_eta_hat)
 mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                          nrow=length(train_id), 
                          ncol=n_bin, byrow = F) 
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid_t, var=T, npc=4)
-ncol(fpca_mod$efunctions)
-fpca_mod$evalues
+# row index subject, column binned time
+
+t1 <- Sys.time()
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
+t2 <- Sys.time()
+t2-t1 
+
+data.frame(t = mid, fpca_mod$efunctions) %>%
+  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
+  pivot_longer(2:4, names_to = "PC") %>%
+  ggplot()+
+  geom_line(aes(x=t, y=value, col = PC))+
+  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
 
 # Re-evaluation
 ## grid extension
 p <- 3 # order of b splines 
 knots <- 35 # number of knots (same from FPCA model)
 knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
-knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
+knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
 
[…]

### 2024-01-16 Docs/manuscript.Rmd
the second reference method (FOSR)

@@
 df_m <- sim_data[[1]]
-df_m %>% filter(t>0.2 & t<=0.4)
 ```
@@
 We would like to use a second reference method for predictive performance comparision.
-In this madel, predictions on a specific interval are all made using the last few observations in the observed window.
-For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2($t_j: j=200, ...
+In this model, predictions on a specific interval are all made using the last few observations in the observed window.
+For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2 ($t_j: j=200, ...
 200-(L-1)$) as time-fixed covariate to predict any future time.
@@
 
-Under this framework, for each dataset we need to fit four models:
+Under this framework, for each dataset we need to fit four models (similar to the GLMMadaptive method):
 
@@
 
-## L = 1
+## Example of L = 1, given [0, 0.2] to predict (0.2, 1)
 
@@
   slice(L) %>% select(id, Y) %>% rename(yl = Y)
-## to predict (0.2, 0.4), outcome:
+## to predict (0.2, 1), outcome:
 df_pred_tr <- train_df %>% filter(window!=1) %>%
@@
   mutate_at(vars(Y, yl), as.factor)
+
+# range(df_pred_tr$sind)
 
@@
 
+```
+
+## Example of prediction on one dataset (L = 1)
+
+```{r}
+# table(df_m$window)
+train_df$window <- as.numeric(as.character(train_df$window))
+test_df$window <- as.numeric(as.character(test_df$window))
+test_df2 <- test_df
+
+t1 <- Sys.time()
+for(w in 1:4){
+  # model fit
+  # predictor (training set)
+  y_obs_max <- train_df %>% filter(window==w) %>% arrange(desc(t)) %>%
+    group_by(id) %>%
+    slice(L) %>% select(id, Y) %>% rename(yl = Y)
+  # outcome (training set)
+  df_pred_tr <- train_df %>% filter(window > w) %>%
+    left_join(y_obs_max, by = "id") %>% 
+    mutate_at(vars(Y, yl), as.factor)
+  # model 
+  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
+                        s(t, bs="cc", k=20, by = yl),
+                   family = binomial, data=df_pred_tr, 
+                   method = "fREML",
+                   discrete = TRUE)
+
+  # prediction
+  # predictor (testing set)
+  y_obs_max_te <- test_df %>% 
+                filter(window==w) %>% arrange(desc(t)) %>% group_by(id) %>%
+                slice(L) %>% select(id, Y) %>% rename(yl = Y)
+  # outcome (testing set)
+  df_pred_te <- test_df %>% filter(window > w) %>%
+    left_join(y_obs_max_te, by = "id") %>% 
+    mutate_at(vars(Y, yl), as.factor)
+
+  # predict using the FOSR model fit above
+  pred_name <- paste0("pred_w", w)
+  test_df[, pred_name] <- NA
+  test_df[test_df$window>w, pred_name] <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
+}
+t2 <- Sys.time()
+# t2-t1
+
+```
+
+```{r}
+test_df %>%
+  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.5)+
+  geom_line(aes(x=t, y=pred_w1, col = "Up to 0.1"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "Up to 0.2"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "Up to 0.3"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "Up to 0.4"), na.rm=T)+
+  geom_line(aes(x=t, y=eta_i, col = "ture"))+
+  facet_wrap(~id)+
+  labs(title = "Out-of-sample prediction")
+```
+
+## Example of prediction on one dataset (L = 5)
+
+```{r}
+t1 <- Sys.time()
+for(w in 1:4){
+  # model fit
+  # predictor (training set)
+  y_obs_max <- train_df %>% filter(window==w) %>% group_by(id) %>%
+    slice_max(t, n=5) %>% select(id, Y, sind) %>%
+    mutate(name = sind-min(sind)+1) %>%
+    pivot_wider(id_cols = id, values_from = Y, names_from = "name", names_prefix = "yl")
+  # outcome (training set)
+  df_pred_tr <- train_df %>% filter(window > w) %>%
+    left_join(y_obs_max, by = "id") %>% 
+    mutate_at(vars(Y, starts_with("yl")), as.factor) 
+  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
+                        s(t, bs="cc", k=20, by = yl5)+
+                        s(t, bs="cc", k=20, by = yl4)+
+                        s(t, bs="cc", k=20, by = yl3)+
+                        s(t, bs="cc", k=20, by = yl2)+
+                        s(t, bs="cc", k=20, by = yl1),
+                   family = binomial, data=df_pred_tr, 
+                   method = "fREML",
+                   discrete = TRUE)
+
+  # prediction
+  # predictor (testing set)
+  y_obs_max_te <- test_df2 %>% filter(window==w) %>% group_by(id) %>%
+    slice_max(t, n=5) %>% select(id, Y, sind) %>% 
+    mutate(name = sind-min(sind)+1) %>%
+    pivot_wider(id_cols = id, values_from = Y, names_from = name, names_prefix = "yl")
+  # outcome (testing set)
+  df_pred_te <- test_df2 %>% filter(window > w) %>%
+    left_join(y_obs_max_te, by = "id") %>% 
+    mutate_at(vars(Y, starts_with("yl")), as.factor)
+
+  # predict using the FOSR model fit above
+  pred_name <- paste0("pred_w", w)
+  test_df2[, pred_name] <- NA
+  test_df2[test_df2$window>w, pred_name] <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
+}
+t2 <- Sys.time()
+# t2-t1
+
+```
+
+```{r}
+test_df2 %>%
+  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
+  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
+  filter(id %in% rand_id) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.5)+
+  geom_line(aes(x=t, y=pred_w1, col = "Up to 0.1"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "Up to 0.2"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "Up to 0.3"), na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "Up to 0.4"), na.rm=T)+
+  geom_line(aes(x=t, y=eta_i, col = "ture"))+
+  facet_wrap(~id)+
+  labs(title = "Out-of-sample prediction")
+```
+
+This method is very, very fast, takes 10 seconds or so, much faster than both fGFPCA and GLMMadaptive. The performance lies in between them for most cases, sometimes worse than GLMMadaptive.
+
+Do we really want to use the lag-1 model?
+
+```{r ise_fosr}
+tb_ies_l5 <- test_df2 %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, window, starts_with("err")) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>%
+  select(starts_with("err"))
+
+tb_ies_l1 <- test_df %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, window, starts_with("err")) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>%
+  select(starts_with("err"))
+
+colnames(tb_ies_l1) <- colnames(tb_ies_l5) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r}
+data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
+           tb_ies_l1, tb_ies_l5, check.names = F) %>%
+  kable(digits = 3, caption = "Integrated squared error", booktabs = T) %>%
+  kable_styling(full_width = F) %>% 
+  add_header_above(c(" "=1, "L = 1" = 4, "L = 5" = 4)) %>%
+  add_header_above(c(" "= 1, "Observation track" = 8))
 ```

### 2024-01-19 Docs/NhanesApplExample.Rmd
summary stats for PC score estimates

@@
 library(LaplacesDemon)
+library(arsenal)
 theme_set(theme_minimal())
@@
 
-Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.
-
 ```{r}
@@
 
-- Eigenvalues: `r fpca_mod$evalues`
-- Proportion of variance explained with four PCs: `r fpca_mod$pve`
+- Proportion of variance explained with four PCs: `r round(fpca_mod$pve, 3)`
 
@@
 
-Again, here we use the three PC functions in the random effect part.
-
 ```{r revaluation}
@@
 train_df$id <- as.factor(train_df$id)
+```
 
+```{r, class.source='fold-show'}
 t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(sind, bs="cc")+
+debias_glmm <- bam(Y ~ s(sind, bs="cr")+
                      s(id, by=phi1, bs="re")+
@@
 t2 <- Sys.time()
-# t2-t1 # 7.5 minutes
+# t2-t1 # less than 7.5 minutes
 ```
 
-It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically
+It seems that the re-evaluated eigenvalues lost their decreasing nature. The eigenvalue corresponding to PC2 is smaller than PC3. Specifically
 
-- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
-- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
-- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`
+- The smoothing parameters from the mgcv::bam model: `r round(debias_glmm$sp[2:5], 4)`
 
-Also, the proportion of these two set of estimates is not constant, as below:
+- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r round(1/debias_glmm$sp[2:5], 3)`
+- The **biased** eigenvalues from FPCA (step 3): `r round(fpca_mod$evalues[1:4], 3)`
 
-```{r, class.source = 'fold-show'}
-# proportion of biased eigenvalues to re-evaluated eigenvalues
-fpca_mod$evalues/(1/debias_glmm$sp[2:5])
-```
+Below I try to investigate the cause of this problem.
 
-### Percentage of variance explained by each PC
+### Variance explained by each PC
 
-We now that the variance of individual-specific random effects, in theory, is:
+The variance of individual-specific random effects, in theory, is:
 
@@
 
-Here I'd like to compare the variantion for each PC calculated using estimated from the **biased** FPCA estimates and the supposedly **debiased** from the dibased GLMM model.
+Here I'd like to compare the variation from each PC calculated using estimated from the **biased** FPCA estimates (binned grid) and the supposedly **debiased** from the dibased GLMM model (original grid).
 
@@
   facet_wrap(~PC)+
-  labs(title = "Time-varying variation of each PC")
+  labs(title = "Time-varying variation of each PC", x="Time", y = "Variation",
+       col = "")
 
@@
 
-I am also curious about how the individual scores change after debias
+### Estimates of score
+
+I am also curious about how the individual score estimates change after debias
 
@@
 
+df_score <- bind_rows(df_score_fpca, df_score_glmm) %>%
+  mutate(PC = gsub("X", "PC", PC)) 
 ```
 
-```{r}
-bind_rows(df_score_fpca, df_score_glmm) %>%
+```{r score_box}
+df_score %>%
   ggplot(aes(x=est, y=score, col=est))+
@@
   facet_wrap(~PC, scales = "free_y")+
-  labs(title = "Score of each PC")
+  labs(title = "Score estimates of each PC", x = "", y="Score estimates",
+       col = "")
 
 ```
+
+```{r score_hist}
+df_score %>% 
+  ggplot()+
+  geom_histogram(aes(x=score), bins = 100)+
+  facet_grid(rows = vars(PC), cols = vars(est), scales = "free")+
+  labs(title = "Distribution of score estimates", x = "Score estimates",
+       y= "")
+
+```
+
+From figures above, the score estimates on the binned grid (FPCA) of the 3rd PC is quite skewed negatively (and the 2nd PC slightly positively?)
+
+Below are summary statistics
+
+```{r score_sum_stat}
+df_score %>%
+  pivot_wider(names_from = PC, values_from = score) %>%
+  tableby(est ~ PC1 + PC2 + PC3 + PC4, data = .,
+          control = tableby.control(total = F, 
+                                    test = F)) %>% 
+  summary() %>% 
+  kable(table.attr = "style = \"color: black;\"", escape =F) %>% 
+  kable_styling(full_width = F)
+
+```

### 2024-01-19 Docs/manuscript.Rmd
change spline basis for GFOSR model

@@
 t1 <- Sys.time()
-fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
-                        s(t, bs="cc", k=20, by = yl),
+fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) + 
+                        s(t, bs="cr", k=20, by = yl),
                    family = binomial, data=df_pred_tr, 
@@
   # model 
-  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
-                        s(t, bs="cc", k=20, by = yl),
+  fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) + 
+                        s(t, bs="cr", k=20, by = yl),
                    family = binomial, data=df_pred_tr, 
@@
     mutate_at(vars(Y, starts_with("yl")), as.factor) 
-  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
-                        s(t, bs="cc", k=20, by = yl5)+
-                        s(t, bs="cc", k=20, by = yl4)+
-                        s(t, bs="cc", k=20, by = yl3)+
-                        s(t, bs="cc", k=20, by = yl2)+
-                        s(t, bs="cc", k=20, by = yl1),
+  fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) + 
+                        s(t, bs="cr", k=20, by = yl5)+
+                        s(t, bs="cr", k=20, by = yl4)+
+                        s(t, bs="cr", k=20, by = yl3)+
+                        s(t, bs="cr", k=20, by = yl2)+
+                        s(t, bs="cr", k=20, by = yl1),
                    family = binomial, data=df_pred_tr, 
@@
            tb_ies_l1, tb_ies_l5, check.names = F) %>%
-  kable(digits = 3, caption = "Integrated squared error", booktabs = T) %>%
+  kable(digits = 3, caption = "Integrated squared error", booktabs = T,
+        table.attr = "style = \"color: black;\"") %>%
   kable_styling(full_width = F) %>% 

### 2024-01-23 Docs/manuscript.Rmd
GFOSR simulation output

--- 2024-01-16 Docs/manuscript.Rmd
@@
 
-load(here("Data/SimOutput_fGFPCA.RData"))
-load(here("Data/SimOutput_GLMMadaptive.RData"))
-
 ## prediction window 
 window <- seq(0, 1, by = 0.2)
-M <- length(pred_list_all)
+M <- 500
 ```
@@
 
-```{r, eval=FALSE}
-grid_exp <- data.frame(t = seq(0, 1, by = 0.01), y = 0)
-ggplot(grid_exp, aes(x=t, y=y))+
-  geom_point(size = 0.5)
-```
-
 ### Local GLMMs
@@
 Laplace Approximation would get the posterior mode of $\xi_{uk}$ through quadratic approximation.
+
+# Reference method
+
+## GLMMadaptive
+
+- For large datasets, we can fit a model with random intercept and slope for time.
+It is doable on 500 datasets, but obviously too simple for the data generation scheme.
+We would expect it to perform terribly.
+
+$$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
+
+- For small datasets, we would like to fit fGFPCA and GLMMadaptive on a dataset with smaller sample size and/or smaller measurement density.
+For the GLMMadaptive model, we would set it up with spline basis functions so that its flexibility is comparable with fGFPCA model, such as:
+
+$$g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
+
+## Generalized functional-on-scalar regression
+
+We would like to use a second reference method for predictive performance comparision.
+In this model, predictions on a specific interval are all made using the last few observations in the observed window.
+For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2 ($t_j: j=200, ...
+200-(L-1)$) as time-fixed covariate to predict any future time.
+
+Let's write out the model expression (with questionable notation). If the observations if up to $t_m$, then:
+
+\[\begin{aligned}
+g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
+l & = m,...m-(j-1)\\
+t &> t_m 
+\end{aligned}\]
+
+This is a simple function-on-scalar model with no random effect, meaning all subject with the same last observed outcome would have the same estimated/predicted latent track.
+This is clearly anti-intuitive.
+But adding random effects would make out-of-sample prediction impossible.
+
+Under both frameworks, for each dataset we need to fit four models (similar to the GLMMadaptive method):
+
+- Given 0-0.2, predict 0.2-1
+- Given 0-0.4, predict 0.4-1
+- Given 0-0.6, predict 0.6-1
+- Given 0-0.8, predict 0.8-1
+
+But prediction performance will be reported by equal-length time window.
 
@@
 
-```{r fig_sim_data}
-rand_id <- sample(501:600, size = 4)
-pred_list_all[[1]] %>% filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"), show.legend = F)+
-  facet_wrap(~id)+
-  labs(title = "Simulated data")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
-```
-
-## Reference method
-
-- GLMMadaptive
-- Here we can fit a model with random intercept and slope for time. It is doable on 500 datasets, but obviously too simple for the data generation scheme. We would expect it to perform terribly.
-
-$$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
-
 ## Figure
+
+```{r load_sim1}
+load(here("Data/SimOutput_fGFPCA.RData"))
+load(here("Data/SimOutput_GLMMadaptive.RData"))
+load(here("Data/SimOutput_GFOSR_L1.RData"))
+pred_list_gfosr_l1 <- pred_list_gfofr
+t_vec_gfosr_l1 <- t_vec_gfofr
+rm(pred_list_gfofr, t_vec_gfofr)
+load(here("Data/SimOutput_GFOSR_L5.RData"))
+pred_list_gfosr_l5 <- pred_list_gfofr
+t_vec_gfosr_l5 <- t_vec_gfofr
+rm(pred_list_gfofr, t_vec_gfofr)
+
+rand_id <- sample(501:600, size = 2)
+```
 
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
-  labs(title = "fGFPCA", col = "Maximum observation time")+
+  facet_wrap(~id, nrow = 1)+
+  labs(title = "fGFPCA", col = "Maximum observation time", x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5))
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
 
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time")+
+  facet_wrap(~id, nrow = 1)+
+  labs(title = "GLMMadaptive", col = "Maximum observation time", 
+       x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5))
-
-```
-
-```{r fig_sim, fig.height=4, fig.width=8}
-ggarrange(fig_fgfpca, fig_adglmm,
-          nrow = 1, common.legend = T)
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+fig_gfosr_l1 <- pred_list_gfosr_l1[[1]] %>%
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id, nrow = 1)+
+  labs(title = "GFOSR (L=1)", col = "Maximum observation time",
+       x="Time", y="")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+fig_gfosr_l5 <- pred_list_gfosr_l5[[1]] %>%
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id, nrow = 1)+
+  labs(title = "GFOSR (L=5)", col = "Maximum observation time",
+       x="Time", y="")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+```
+
+```{r fig_sim, fig.height=8, fig.width=5}
+ggarrange(fig_fgfpca, fig_adglmm, fig_gfosr_l1, fig_gfosr_l5,
+          ncol = 1, common.legend = T)
 ```
@@
 ## ISE container 
-ise_mat <- ise_mat_ref <- array(NA, 
-                                dim = c(length(window)-2, length(window)-2, M))
+ise_fgfpca <- ise_adglmm <- ise_gfosr_l1 <- ise_gfosr_l5 <- 
+  array(NA, dim = c(length(window)-2, length(window)-2, M))
 # dims: prediction window, max obs time, simulation iter
@@
 for(m in 1:M){
-  this_df <- pred_list_all[[m]]
+  # this_df <- pred_list_all[[m]]
   ise_tb <- pred_list_all[[m]] %>%
@@
     select(starts_with("err"))
-  ise_mat[, ,m] <- as.matrix(ise_tb)
-
-}
-
-mean_ise <- apply(ise_mat, c(1, 2), mean)
+  ise_fgfpca[, ,m] <- as.matrix(ise_tb)
+
+}
+
+mean_ise_fgfpca <- apply(ise_fgfpca, c(1, 2), mean)
 
@@
 #          .before = 1)
-colnames(mean_ise) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_GLMMadaptive}
+colnames(mean_ise_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_adglmm}
 ## calculation
 for(m in 1:M){
-  this_df <- pred_list_ref[[m]]
+  # this_df <- pred_list_gfosr_l1[[m]]
   ise_tb <- pred_list_ref[[m]] %>%
@@
     select(starts_with("err"))
-  ise_mat_ref[, ,m] <- as.matrix(ise_tb)
-
-}
-
-mean_ise_ref <- apply(ise_mat_ref, c(1, 2), mean)
-# mean_ise_ref <- data.frame(mean_ise_ref) %>% 
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_ise_ref) <- c("0.2", "0.4", "0.6", "0.8")
+  ise_adglmm[, ,m] <- as.matrix(ise_tb)
+}
+
+mean_ise_adglmm <- apply(ise_adglmm, c(1, 2), mean)
+colnames(mean_ise_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_gfosr_l1}
+## calculation
+for(m in 1:M){
+  # this_df <- pred_list_ref[[m]]
+  ise_tb <- pred_list_gfosr_l1[[m]] %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, starts_with("err"), window) %>% 
+    # mutate(window = factor(window, levels = 1:5, 
+    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>% 
+    select(starts_with("err"))
+  ise_gfosr_l1[, ,m] <- as.matrix(ise_tb)
+}
+
+mean_ise_gfosr_l1 <- apply(ise_gfosr_l1, c(1, 2), mean)
+colnames(mean_ise_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_gfosr_l5}
+## calculation
+for(m in 1:M){
+  # this_df <- pred_list_ref[[m]]
+  ise_tb <- pred_list_gfosr_l5[[m]] %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, starts_with("err"), window) %>% 
+    # mutate(window = factor(window, levels = 1:5, 
+    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>% 
+    select(starts_with("err"))
+  ise_gfosr_l5[, ,m] <- as.matrix(ise_tb)
+}
+
+mean_ise_gfosr_l5 <- apply(ise_gfosr_l5, c(1, 2), mean)
+colnames(mean_ise_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-           mean_ise, mean_ise_ref, check.names = F) %>%
-  kable(digits = 3, caption = "Integrated squared error", booktabs=T) %>%
+rbind(mean_ise_fgfpca, mean_ise_adglmm, 
+           mean_ise_gfosr_l1, mean_ise_gfosr_l5) %>%
+  as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
+                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
+         .before=1) %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 8))
+  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
+                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
 ```
@@
 ## auc container 
-auc_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
-auc_mat_ref <- array(NA, dim = c(length(window)-2, length(window)-2, M))
+auc_fgfpca <- auc_adglmm <- auc_gfosr_l1 <- auc_gfosr_l5 <- 
+  array(NA, dim = c(length(window)-2, length(window)-2, M))
 ```
@@
     select(starts_with("auc"))
-  auc_mat[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc <- apply(auc_mat, c(1, 2), mean)
+  auc_fgfpca[, ,m] <- as.matrix(auc_tb)
+}
+
+mean_auc_fgfpca <- apply(auc_fgfpca, c(1, 2), mean)
 # mean_auc <- data.frame(mean_auc) %>% 
@@
 #          .before = 1)
-colnames(mean_auc) <- c("0.2", "0.4", "0.6", "0.8")
+colnames(mean_auc_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
     select(starts_with("auc"))
-  auc_mat_ref[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_ref <- apply(auc_mat_ref, c(1, 2), mean)
-# mean_auc <- data.frame(mean_auc) %>% 
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_auc_ref) <- c("0.2", "0.4", "0.6", "0.8")
+  auc_adglmm[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc_adglmm <- apply(auc_adglmm, c(1, 2), mean)
+colnames(mean_auc_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
+```
+```{r auc_gfosr_l1}
+for(m in 1:M){
+  this_df <- pred_list_gfosr_l1[[m]]
+  auc_tb <- this_df %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred_w1),
+              auc2 = get_auc(Y, pred_w2),
+              auc3 = get_auc(Y, pred_w3),
+              auc4 = get_auc(Y, pred_w4)) %>%
+    filter(window != 1) %>% 
+    select(starts_with("auc"))
+  auc_gfosr_l1[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc_gfosr_l1 <- apply(auc_gfosr_l1, c(1, 2), mean)
+colnames(mean_auc_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r auc_gfosr_l5}
+for(m in 1:M){
+  this_df <- pred_list_gfosr_l5[[m]]
+  auc_tb <- this_df %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred_w1),
+              auc2 = get_auc(Y, pred_w2),
+              auc3 = get_auc(Y, pred_w3),
+              auc4 = get_auc(Y, pred_w4)) %>%
+    filter(window != 1) %>% 
+    select(starts_with("auc"))
+  auc_gfosr_l5[, ,m] <- as.matrix(auc_tb)
+
+}
+
+mean_auc_gfosr_l5 <- apply(auc_gfosr_l5, c(1, 2), mean)
+colnames(mean_auc_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-           mean_auc, mean_auc_ref, check.names = F) %>%
-  kable(digits = 3, caption = "Area under the ROC curve", booktabs = T) %>%
+rbind(mean_auc_fgfpca, mean_auc_adglmm, 
+           mean_auc_gfosr_l1, mean_auc_gfosr_l5) %>%
+  as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
+                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
+         .before=1) %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 8))
-```
-
-```{r time}
+  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
+                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+```
+
+## Time (minutes)
+
+I chose not to report fitting and prediction time separately because it is too difficult to tell them apart for the GFOSR models.
+
+```{r}
 data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive"),
-  "Fit" = c(mean(fit_time)/60, mean(fit_time_ref)),
-           "Prediction"= c(mean(pred_time), mean(pred_time_ref)/60)) %>%
-  kable(digits = 3, caption = "Computation time (minutes)", booktabs = T) %>%
+  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
+  "Time" = c(mean(fit_time/60+pred_time),
+             mean(fit_time_ref+pred_time_ref/60),
+             mean(t_vec_gfosr_l1/60), 
+             mean(t_vec_gfosr_l5/60))) %>%
+  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
   kable_styling(full_width = F) 
@@
 
-I think we could say that while the total time spend on model fitting + prediction are similar between two methods, fGFPCA achieved much better flexibility and much better predictive performance of prediction under every scenario.
-
 # Small-scale simulation
 
-Here we would like to fit fGFPCA and GLMMadaptive on a dataset with smaller sample size and/or smaller measurement density.
-For the GLMMadaptive model, we would set it up with spline basis functions so that its flexibility is comparable with fGFPCA model, such as:
-
-$$g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
-
-I have used 100 subjects for training and testing, and repeated 100 times.
-When fitting GLMMadpative, I reduce the number of measurements to 1/10 by taking one every 10 observations.
+I have used 100 subjects for training and testing, and repeated 500 times.
+When fitting GLMMadpative, I reduce the number of measurements in the training dataset to 1/10 by taking one every 10 observations.
 The prediction is on the original grid.
 
-```{r load_subset_sim_output}
+```{r load_sim2}
 load(here("Data/SubSimOutput_GLMMadaptive.RData"))
+
 load(here("Data/SubSimOutput_fGFPCA.RData"))
+pred_subset_adglmm <- pred_subset_adglmm[!num_probs]
+fit_time_subset_adglmm <- fit_time_subset_adglmm[!num_probs]
+pred_time_subset_adglmm <- pred_time_subset_adglmm[!num_probs]
+
+load(here("Data/SubSimOutput_GFOSR_L1.RData"))
+pred_subset_gfosr_l1 <- pred_list_gfofr_subset
+t_subset_gfosr_l1 <- t_vec_gfofr_subset
+rm(pred_list_gfofr_subset, t_vec_gfofr_subset)
+
+load(here("Data/SubSimOutput_GFOSR_L5.RData"))
+pred_subset_gfosr_l5 <- pred_list_gfofr_subset
+t_subset_gfosr_l5 <- t_vec_gfofr_subset
+rm(pred_list_gfofr_subset, t_vec_gfofr_subset)
+
+rand_id <- sample(101:200, size = 2)
 ```
@@
 ```{r fig_pred_subset}
-rand_id2 <- sample(unique(pred_subset_fGFPCA[[1]]$id), 4)
-
-fig_fgfpca_subset <- pred_subset_fGFPCA[[1]] %>%
-  filter(id %in% rand_id2) %>%
+fig_fgfpca2 <- pred_subset_fGFPCA[[1]] %>%
+  filter(id %in% rand_id) %>%
   mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
@@
   facet_wrap(~id)+
-  labs(title = "fGFPCA", col = "Maximum observation time")+
+  labs(title = "fGFPCA", col = "Maximum observation time",
+       x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5))
-
-fig_adglmm_subset <- pred_subset_adglmm[[1]] %>% 
-  filter(id %in% rand_id2) %>%
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+fig_adglmm2 <- pred_subset_adglmm[[1]] %>% 
+  filter(id %in% rand_id) %>%
   mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
@@
   facet_wrap(~id)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time")+
+  labs(title = "GLMMadaptive", col = "Maximum observation time",
+       x = "Time", y = "")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5))
-```
-
-```{r fit_sim_small, fig.height=4, fig.width=8}
-ggarrange(fig_fgfpca_subset, fig_adglmm_subset,
-          nrow = 1, common.legend = T)
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+fig_gfosr_l1_2 <- pred_subset_gfosr_l1[[1]] %>% 
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GFOSR (L=1)", col = "Maximum observation time",
+       x = "Time", y = "")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+
+fig_gfosr_l5_2 <- pred_subset_gfosr_l5[[1]] %>% 
+  filter(id %in% rand_id) %>%
+  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
+  facet_wrap(~id)+
+  labs(title = "GFOSR (L=5)", col = "Maximum observation time",
+       x = "Time", y = "")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(axis.text = element_text(size = 5), 
+        title = element_text(size = 8))
+```
+
+```{r fig_sim2, fig.height=8, fig.width=5}
+ggarrange(fig_fgfpca2, fig_adglmm2, fig_gfosr_l1_2, fig_gfosr_l5_2,
+          ncol = 1, common.legend = T)
 ```
@@
 ```{r}
-# M2 <- length(pred_subset_fGFPCA)
 ## ISE container 
-ise_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, length(pred_subset_fGFPCA)))
-
-pred_subset_adglmm <- pred_subset_adglmm[num_probs==0]
-ise_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, length(pred_subset_adglmm)))
+ise_fgfpca2 <- ise_adglmm2 <- ise_gfosr_l1_2 <- ise_gfosr_l5_2 <- 
+  array(NA, dim = c(length(window)-2, length(window)-2, M))
+# dims: prediction window, max obs time, simulation iter
 
@@
     select(starts_with("err")) %>% as.matrix()
-  ise_mat_subset[,,m] <- ise_tb_m
-}
-
-ise_mat_subset <- apply(ise_mat_subset, c(1, 2), mean)
-
-colnames(ise_mat_subset) <- c("0.2", "0.4", "0.6", "0.8")
+  ise_fgfpca2[,,m] <- ise_tb_m
+}
+
+mean_ise_fgfpca2 <- apply(ise_fgfpca2, c(1, 2), mean)
+
+colnames(mean_ise_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
     select(starts_with("err")) %>% as.matrix()
-  ise_mat_subset_ref[,,m] <- ise_tb_m
-}
-
-ise_mat_subset_ref <- apply(ise_mat_subset_ref, c(1, 2), mean)
-
-colnames(ise_mat_subset_ref) <- c("0.2", "0.4", "0.6", "0.8")
+  ise_adglmm2[,,m] <- ise_tb_m
+}
+
+mean_ise_adglmm2<- apply(ise_adglmm2, c(1, 2), mean, na.rm = T)
+
+colnames(mean_ise_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_gfosr_l1_subset}
+for(m in 1:length(pred_subset_gfosr_l1)){
+  this_df <- pred_subset_gfosr_l1[[m]]
+  ise_tb_m <- this_df %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, starts_with("err"), window) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>% 
+    select(starts_with("err")) %>% as.matrix()
+  ise_gfosr_l1_2[,,m] <- ise_tb_m
+}
+
+mean_ise_gfosr_l1_2<- apply(ise_gfosr_l1_2, c(1, 2), mean, na.rm = T)
+
+colnames(mean_ise_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r ise_gfosr_l5_subset}
+for(m in 1:length(pred_subset_gfosr_l5)){
+  this_df <- pred_subset_gfosr_l5[[m]]
+  ise_tb_m <- this_df %>%
+    mutate(err1 = (pred_w1-eta_i)^2,
+           err2 = (pred_w2-eta_i)^2,
+           err3 = (pred_w3-eta_i)^2,
+           err4 = (pred_w4-eta_i)^2) %>%
+    select(id, t, starts_with("err"), window) %>% 
+    group_by(window, id) %>% 
+    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
+    group_by(window) %>% 
+    summarize_at(vars(err1, err2, err3, err4), mean) %>%
+    filter(window != 1) %>% 
+    select(starts_with("err")) %>% as.matrix()
+  ise_gfosr_l5_2[,,m] <- ise_tb_m
+}
+
+mean_ise_gfosr_l5_2 <- apply(ise_gfosr_l5_2, c(1, 2), mean, na.rm = T)
+
+colnames(mean_ise_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
-           ise_mat_subset, ise_mat_subset_ref, check.names = F) %>%
-  kable(digits = 3, caption = "Integrated squared error", booktabs = T) %>%
+rbind(mean_ise_fgfpca2, mean_ise_adglmm2, 
+           mean_ise_gfosr_l1_2, mean_ise_gfosr_l5_2) %>%
+  as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
+                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
+         .before=1) %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" "= 1, "Observation track" = 8))
+  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
+                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
 ```
@@
 ## auc container 
-auc_mat_subset <- array(NA, dim = c(length(window)-2, length(window)-2, 
-                                    length(pred_subset_fGFPCA)))
-auc_mat_subset_ref <- array(NA, dim = c(length(window)-2, length(window)-2, 
-                                        length(pred_subset_adglmm)))
+auc_fgfpca2 <- auc_adglmm2 <- auc_gfosr_l1_2 <- auc_gfosr_l5_2 <- 
+  array(NA, dim = c(length(window)-2, length(window)-2, M))
 ```
@@
   this_df <- pred_subset_fGFPCA[[m]]
-
   auc_tb_m <- this_df %>%
@@
 
-  auc_mat_subset[, , m] <- auc_tb_m
-
-}
-
-auc_mat_subset <- apply(auc_mat_subset, c(1,2), mean)
-colnames(auc_mat_subset) <- c("0.2", "0.4", "0.6", "0.8")
+  auc_fgfpca2[, , m] <- auc_tb_m
+
+}
+
+mean_auc_fgfpca2 <- apply(auc_fgfpca2, c(1,2), mean)
+colnames(mean_auc_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
   this_df <- pred_subset_adglmm[[m]]
-
   auc_tb_m <- this_df %>%
@@
 
-  auc_mat_subset_ref[, , m] <- auc_tb_m
-
-}
-
-auc_mat_subset_ref <- apply(auc_mat_subset_ref, c(1,2), mean)
-colnames(auc_mat_subset_ref) <- c("0.2", "0.4", "0.6", "0.8")
+  auc_adglmm2[, , m] <- auc_tb_m
+
+}
+
+mean_auc_adglmm2 <- apply(auc_adglmm2, c(1,2), mean, na.rm = T)
+colnames(mean_auc_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r auc_GFOSR_subset_l1}
+for(m in 1:length(pred_subset_gfosr_l1)){
+  this_df <- pred_subset_gfosr_l1[[m]]
+  auc_tb_m <- this_df %>%
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred_w1),
+            auc2 = get_auc(Y, pred_w2),
+            auc3 = get_auc(Y, pred_w3),
+            auc4 = get_auc(Y, pred_w4)) %>%
+  filter(window != 1) %>% 
+  select(starts_with("auc")) %>% as.matrix()
+
+  auc_gfosr_l1_2[, , m] <- auc_tb_m
+
+}
+
+mean_auc_gfosr_l1_2 <- apply(auc_gfosr_l1_2, c(1,2), mean)
+colnames(mean_auc_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
+```
+
+```{r auc_GFOSR_subset_l5}
+for(m in 1:length(pred_subset_gfosr_l5)){
+  this_df <- pred_subset_gfosr_l5[[m]]
+  auc_tb_m <- this_df %>%
+  select(Y, starts_with("pred"), window) %>%
+  group_by(window) %>%
+  summarise(auc1 = get_auc(Y, pred_w1),
+            auc2 = get_auc(Y, pred_w2),
+            auc3 = get_auc(Y, pred_w3),
+            auc4 = get_auc(Y, pred_w4)) %>%
+  filter(window != 1) %>% 
+  select(starts_with("auc")) %>% as.matrix()
+
+  auc_gfosr_l5_2[, , m] <- auc_tb_m
+
+}
+
+mean_auc_gfosr_l5_2 <- apply(auc_gfosr_l5_2, c(1,2), mean)
+colnames(mean_auc_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
 ```
@@
 ```{r}
-data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
-           auc_mat_subset, auc_mat_subset_ref, check.names = F) %>%
-  kable(digits = 3, caption = "Area under the ROC curve", booktabs = T) %>%
+rbind(mean_auc_fgfpca2, mean_auc_adglmm2, 
+           mean_auc_gfosr_l1_2, mean_auc_gfosr_l5_2) %>%
+  as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
+                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
+         .before=1) %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "=1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" "= 1, "Observation track" = 8))
+  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
+                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
+  add_header_above(c(" "= 1, "Maximum observation time" = 4))
 ```
@@
 Among 500 hundred iterations, `r sum(num_probs)` did not converge for the GLMMadaptive model. No numeric issue for fGFPCA.
+
+# Time
 
@@
 data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive"),
-  "Fit" = c(mean(fit_time_subset_fGFPCA)/60, mean(fit_time_subset_adglmm, na.rm=T)),
-           "Prediction"= c(mean(pred_time_subset_fGFPCA), mean(pred_time_subset_adglmm, na.rm=T)/60)) %>%
-  kable(digits = 3, caption = "Computation time (minutes)", booktabs = T) %>%
+  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
+  "Time" = c(mean(fit_time_subset_fGFPCA/60+pred_time_subset_fGFPCA),
+             mean(fit_time_subset_adglmm+pred_time_subset_adglmm/60),
+             mean(t_subset_gfosr_l1/60), 
+             mean(t_subset_gfosr_l5/60))) %>%
+  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
   kable_styling(full_width = F) 
@@
   nrow = 1, common.legend = T)
-```
-
-# The second reference method:
-
-```{r}
-load(here("Data/sim_data.RData"))
-df_m <- sim_data[[1]]
-df_m %>% filter(t>0.2 & t<=0.4)
-```
-
-We would like to use a second reference method for predictive performance comparision.
-In this madel, predictions on a specific interval are all made using the last few observations in the observed window.
-For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2($t_j: j=200, ...
-200-(L-1)$) as time-fixed covariate to predict any future time.
-
-Let's write out the model expression (with questionable notation). If the observations if up to $t_m$, then:
-
-\[\begin{aligned}
-g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
-l & = m,...m-(j-1)\\
-t &> t_m 
-\end{aligned}\]
-
-This is a simple function-on-scalar model with no random effect, meaning all subject with the same last observed outcome would have the same estimated/predicted latent track.
-This is clearly anti-intuitive.
-But adding random effects would make out-of-sample prediction impossible.
-
-Under this framework, for each dataset we need to fit four models:
-
-- Given 0-0.2, predict 0.2-1
-- Given 0-0.4, predict 0.4-1
-- Given 0-0.6, predict 0.6-1
-- Given 0-0.8, predict 0.8-1
-
-Below are some single dataset examples.
-
-## L = 1
-
-First let try to make prediction on an interval with only one historical observation (L = 1):
-
-\[\begin{aligned}
-g(E[Y_i(t)]) &= \beta_0(t) +\beta_1(t) Y_i(t_{m})\\
-t &> t_m 
-\end{aligned}\]
-
-```{r data_format}
-L <- 1 # number of observations to use as predictor
-
-# prediction window 
-windows <- seq(0, 1, by = 0.2)
-df_m$window <- cut(df_m$t, breaks=windows, labels = 1:5,
-                       include.lowest = T)
-
-# split data
-N_train <- 500
-N_test <- 100
-train_df <- df_m %>% filter(id %in% 1:N_train)
-test_df <- df_m %>% filter(!id %in% 1:N_train)
-```
-
-```{r fit_model, class.source='fold-show'}
-# format training data
-## given [0, 0.2], predictor would be
-y_obs_max <- train_df %>% filter(window==1) %>% arrange(desc(t)) %>%
-  group_by(id) %>%
-  slice(L) %>% select(id, Y) %>% rename(yl = Y)
-## to predict (0.2, 0.4), outcome:
-df_pred_tr <- train_df %>% filter(window!=1) %>%
-  left_join(y_obs_max, by = "id") %>% 
-  mutate_at(vars(Y, yl), as.factor)
-
-# model
-t1 <- Sys.time()
-fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
-                        s(t, bs="cc", k=20, by = yl),
-                   family = binomial, data=df_pred_tr, 
-                   method = "fREML",
-                   discrete = TRUE)
-t2 <- Sys.time()
-t2-t1
-
-summary(fit_gen_fosr)
-```
-
-```{r est_train}
-# estimation on the training set
-df_pred_tr$eta_est <- predict(fit_gen_fosr, type = "link")
-
-df_pred_tr %>%
-  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
-  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
-  filter(id %in% 1:4) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.5)+
-  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
-  geom_line(aes(x=t, y=eta_i, col = "ture"))+
-  facet_wrap(~id)+
-  labs(title = "In-sample estimation")
-```
-
-```{r pred_test}
-# format testing data
-df_pred_te <- test_df %>% filter(window!=1) %>%
-  left_join(test_df %>% 
-              filter(window==1) %>% arrange(desc(t)) %>% group_by(id) %>%
-              slice(L) %>% select(id, Y) %>% rename(yl = Y),
-            by = "id") %>% 
-  mutate_at(vars(Y, yl), as.factor)
-
-# predict using the FOSR model fit above
-df_pred_te$eta_est <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
-
-df_pred_te %>%
-  mutate_at(vars(eta_est, eta_i), function(x){exp(x)/(1+exp(x))}) %>%
-  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
-  filter(id %in% rand_id) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.5)+
-  geom_line(aes(x=t, y=eta_est, col = "estimate"))+
-  geom_line(aes(x=t, y=eta_i, col = "ture"))+
-  facet_wrap(~id)+
-  labs(title = "Out-of-sample prediction")
-
 ```

### 2024-01-24 Docs/manuscript.Rmd
again, fix the figures

@@
 
-rand_id <- sample(501:600, size = 2)
+rand_id <- sample(501:600, size = 4)
 ```
@@
 
-```{r fig_sim, fig.height=8, fig.width=5}
+```{r fig_sim, fig.height=8, fig.width=8}
 ggarrange(fig_fgfpca, fig_adglmm, fig_gfosr_l1, fig_gfosr_l5,
@@
 ```
+
 ```{r auc_gfosr_l1}
@@
 
-rand_id <- sample(101:200, size = 2)
+rand_id <- sample(101:200, size = 4)
 ```
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
+  facet_wrap(~id, nrow = 1)+
   labs(title = "fGFPCA", col = "Maximum observation time",
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
+  facet_wrap(~id, nrow = 1)+
   labs(title = "GLMMadaptive", col = "Maximum observation time",
@@
   geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
+  facet_wrap(~id, nrow=1)+
   labs(title = "GFOSR (L=1)", col = "Maximum observation time",
@@
   geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id)+
+  facet_wrap(~id, nrow=1)+
   labs(title = "GFOSR (L=5)", col = "Maximum observation time",
@@
 
-```{r fig_sim2, fig.height=8, fig.width=5}
+```{r fig_sim2, fig.height=8, fig.width=8}
 ggarrange(fig_fgfpca2, fig_adglmm2, fig_gfosr_l1_2, fig_gfosr_l5_2,

### 2024-01-31 Docs/NhanesApplExample.Rmd
reversed eigenvalue problems

--- 2023-12-19 Docs/NhanesApplExample.Rmd
@@
 
-set.seed(121)
+set.seed(130)
 
@@
 # use pred_latent function to estimate latent function 
-K <- 3
+K <- 4
 source(here("Code/OutsampBayes.R"))
@@
 
-Fit fGFPCA model
-
-Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 1000 participants as the training set
+# Fit fGFPCA model
 
@@
 df <- df %>% rename(id=SEQN, Y=Z)
-
-Ntr <- 1000
-
-train_id <- unique(df$id)[1:Ntr]
-train_df <- df %>% filter(id %in% train_id)
-
-# additional parameters
+N <- length(unique(df$id))
 J <- length(unique(df$sind))
@@
 
-train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
-train_df$bin <- as.numeric(as.character(train_df$bin))
+df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
+df$bin <- as.numeric(as.character(df$bin))
 ```
 
-# Step 2:  Local GLMM
+# Step 2-4:  Local GLMM, fPCA, splines interpolation and GLMM reevaluation
 
-```{r}
-# fit model on the training set
-train_bin_lst <- split(train_df, f = train_df$bin)
+I'd like to investigate with different sample sizes.
+
+```{r, cache=TRUE, results='hide'}
+# sample size 
+N_sub <- rep(500, 4)
+fpca_model_list <- list()
+interp_pc_est <- list()
+debias_model_list <- list()
+time_vec <- rep(NA, length(N_sub))
+
+pb <- txtProgressBar(min=0, max = length(N_sub), style = 3)
+for(i in seq_along(N_sub)){
+
+  # training sample of a certain size
+  Ntr <- N_sub[i]
+  train_id <- sample(unique(df$id), size = Ntr) 
+  train_df <- df %>% filter(id %in% train_id)
 
   t1=Sys.time()
+  # local GLMM
+  train_bin_lst <- split(train_df, f = train_df$bin)
   df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
-t2= Sys.time()
-# t2-t1 
-```
+  df_est_latent <- bind_rows(df_est_latent)
 
-```{r}
-df_est_latent <- bind_rows(df_est_latent) 
-# head(df_est_latent)
+  # FPCA
+  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+  mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
+  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
+  fpca_model_list[[i]] <- fpca_mod
 
-# example estimated latent function
-rand_id <- sample(train_id, 4)
-df_est_latent %>% 
-  filter(id %in% rand_id) %>%
-  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
-  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
-  facet_wrap(~id, scales = "free")+
-  labs(x = "Time", y = "Estimated latent function (probablity scale)")
-```
-
-# Step 3: FPCA
-
-Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.
-
-```{r}
-uni_eta_hat <- df_est_latent %>% filter(bin==sind)
-mat_est_unique <- matrix(uni_eta_hat$eta_hat,
-                         nrow=Ntr, 
-                         ncol=n_bin, byrow = F) 
-# row index subject, column binned time
-
-t1 <- Sys.time()
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-# keep 3 PCs
-t2 <- Sys.time()
-# t2-t1
-
-# fpca_mod$evalues
-```
-
-```{r PC}
-data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC") %>%
-  ggplot()+
-  geom_line(aes(x=t, y=value, col = PC))+
-  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
-```
-
-- Eigenvalues: `r fpca_mod$evalues`
-- Proportion of variance explained with four PCs: `r fpca_mod$pve`
-
-# Step 4: Re-evaluation
-
-## Extend the eigenfunctions back to the original grid with B-spline basis
-
-```{r interplate}
-# Re-evaluation
-## grid extension
+  # interpolation
   p <- 3 # order of b splines 
@@
   }# project binned eigenfunctions onto the original grid
-```
+  interp_pc_est[[i]] <- df_phi
 
-Below presents the result of re-evaluation.
-Points are values of PC functions on the binned grid estimated from FPCA, and lines are values of re-evaluated PC functions on the original measurement grid.
-
-```{r}
-df_pc1 <- data.frame(t=1:J,  df_phi) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC")
-
-df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC")
-
-left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
-  ggplot()+
-  geom_line(aes(x=t, y=value.x, col = PC), linewidth = 0.2)+
-  geom_point(aes(x=t, y=value.y, col = PC), na.rm = T, alpha = 0.5, size = 0.5)+
-  labs(x="Time", y="", title = "Re-evaluate eigenfunctions on the original grid")
-
-```
-
-## Debias the eigenvalues with GLMM
-
-Again, here we use the three PC functions in the random effect part.
-
-```{r revaluation}
+  # debias model
   df_phi <- data.frame(sind = 1:J, df_phi)
@@
   train_df$id <- as.factor(train_df$id)
-
-t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(sind, bs="cc")+
+  debias_glmm <- bam(Y ~ s(sind, bs="cr")+
                      s(id, by=phi1, bs="re")+
                      s(id, by=phi2, bs="re")+
-                     s(id, by=phi3, bs="re"),
-                     # s(id, by=phi4, bs="re"), 
+                     s(id, by=phi3, bs="re")+
+                     s(id, by=phi4, bs="re"), 
                    family = binomial, 
@@
                    discrete = TRUE)
+  debias_model_list[[i]] <- debias_glmm
   t2 <- Sys.time()
-# t2-t1 # 3.5 minutes
+
+  time_vec[i] <- as.numeric(t2-t1, unit = "secs")
+
+  setTxtProgressBar(pb, i)
+}
+
+close(pb)
 ```
 
-It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically
+<!-- - Time spent on model fitting -->
 
-- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:4]`
-- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:4]`
-- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:3]`
+```{r, eval=FALSE}
+time_vec
+```
 
-Also, the proportion of these two set of estimates is not constant, as below:
+<!-- - FPCA eigenvalues -->
 
-```{r, class.source = 'fold-show'}
-# proportion of biased eigenvalues to re-evaluated eigenvalues
-fpca_mod$evalues/(1/debias_glmm$sp[2:4])
+```{r, eval=FALSE}
+lapply(fpca_model_list, function(x){x$evalues})
+lapply(fpca_model_list, function(x){x$pve})
 ```
+
+- Eigenvalues from debias models
+
+```{r}
+lapply(debias_model_list, function(x){1/x$sp[2:5]})
+```
+
+- PC functions
+
+```{r}
+pc_plot_list <- list()
+
+for(i in  1:length(N_sub)){
+
+  df_phi1 <- data.frame(t=mid, fpca_model_list[[i]]$efunctions)
+  df_phi2 <- data.frame(t = 1:J, interp_pc_est[[i]])
+
+  this_plot <- left_join(df_phi2, df_phi1, by = "t") %>%
+    ggplot()+
+    geom_point(aes(x=t, y=X1.y, col = "PC1"), na.rm = T, size = 0.5)+
+    geom_point(aes(x=t, y=X2.y, col = "PC2"), na.rm = T, size = 0.5)+
+    geom_point(aes(x=t, y=X3.y, col = "PC3"), na.rm = T, size = 0.5)+
+    geom_point(aes(x=t, y=X4.y, col = "PC4"), na.rm = T, size = 0.5)+
+    geom_line(aes(x=t, y=X1.x, col = "PC1"))+
+    geom_line(aes(x=t, y=X2.x, col = "PC2"))+
+    geom_line(aes(x=t, y=X3.x, col = "PC3"))+
+    geom_line(aes(x=t, y=X4.x, col = "PC4"))+
+    labs(x="time", y = "", title = paste0("Iter", i))
+
+  pc_plot_list[[i]] <- this_plot
+
+}
+
+ggarrange(plotlist = pc_plot_list, common.legend = T)
+```
+
+The "reverse eigenvalue problem" does not always happen.
+When training sample size is small (like maybe 200), the problem happened less.
+So this is definately a numeric problem.
+But a very small sample also would induce sigularity issues.
+
+Some other thoughts:
+
+1. What about estimation method of smooth parameter? It seems GCV doesn't treat model as mixed effects. REML may be too slow without the "discrete=T" option on the full dataset. 
+2. Is is possible to impose monotone constraint on smoothing parameter estimation?

### 2024-02-02 Docs/manuscript.Rmd
rotated figures and tables

@@
 library(boot)
+
+# plot settings
 theme_set(theme_minimal())
-
+cols <- c("0.2"= "#FDCC8A", "0.4"="#FC8D59", "0.6"="#E34A33", "0.8"="#B30000", "True"="#000000")
+# brewer.pal(5, "OrRd")
+
+# define a color palette 
 ## prediction window 
@@
 M <- 500
+
 ```
@@
 
-## Figure
+## Figures
 
@@
 
-```{r}
-# range(pred_list_all[[1]]$pred0.2, na.rm = T)
-fig_fgfpca <- pred_list_all[[1]] %>%
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
+```{r fig_large_sim, fig.height=8, fig.width=8}
+bind_rows(
+  pred_list_all[[1]] %>% filter(id %in% rand_id) %>% 
+    mutate(method="fGFPCA"),
+  pred_list_ref[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GLMMadaptive"),
+  pred_list_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=1)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_list_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=5)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4)
+) %>% mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
   ggplot()+
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "fGFPCA", col = "Maximum observation time", x = "Time", y="")+
+  facet_grid(rows = vars(id), cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_adglmm <- pred_list_ref[[1]] %>%
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time", 
-       x = "Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_gfosr_l1 <- pred_list_gfosr_l1[[1]] %>%
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "GFOSR (L=1)", col = "Maximum observation time",
-       x="Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_gfosr_l5 <- pred_list_gfosr_l5[[1]] %>%
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "GFOSR (L=5)", col = "Maximum observation time",
-       x="Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-```
-
-```{r fig_sim, fig.height=8, fig.width=8}
-ggarrange(fig_fgfpca, fig_adglmm, fig_gfosr_l1, fig_gfosr_l5,
-          ncol = 1, common.legend = T)
-```
-
-## ISE
-
-```{r}
+  theme(legend.position = "bottom")+
+  scale_color_manual(values = cols)
+```
+
+## ISE and AUC
+
+```{r ise}
 ## ISE container 
@@
 
-```{r}
-rbind(mean_ise_fgfpca, mean_ise_adglmm, 
-           mean_ise_gfosr_l1, mean_ise_gfosr_l5) %>%
-  as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
-                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
-         .before=1) %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
-                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 4))
-```
-
-## AUC
-
 ```{r auc_func}
@@
 
-```{r}
-rbind(mean_auc_fgfpca, mean_auc_adglmm, 
-           mean_auc_gfosr_l1, mean_auc_gfosr_l5) %>%
-  as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
-                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
-         .before=1) %>%
+```{r table_sim1}
+fgfpca1 <- rbind(mean_ise_fgfpca ,mean_auc_fgfpca) %>% as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
+gfosr_l5_1 <- rbind(mean_ise_gfosr_l5 ,mean_auc_gfosr_l5) %>% as_tibble()
+gfosr_l1_1 <- rbind(mean_ise_gfosr_l1 ,mean_auc_gfosr_l1) %>% as_tibble()
+adglmm1 <- rbind(mean_ise_adglmm ,mean_auc_adglmm) %>% as_tibble()
+
+bind_cols(fgfpca1, gfosr_l5_1, gfosr_l1_1, adglmm1, .name_repair = "minimal") %>%
   kable(digits = 3, booktabs=T,
@@
   kable_styling(full_width = F) %>% 
-  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
-                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+  add_header_above(c(" " = 1, "fGFPCA" = 4, "GFOSR (L=5)" = 4, 
+                     "GFOSR (L=1)" = 4, "GLMMadaptive" = 4)) %>%
+  add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
+  group_rows(index = c("ISE"= 4, "AUC" = 4))
+
 ```
@@
 
-```{r fig_pred_subset}
-fig_fgfpca2 <- pred_subset_fGFPCA[[1]] %>%
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
+```{r fig_small_sim, fig.height=8, fig.width=8}
+bind_rows(
+  pred_subset_fGFPCA[[1]] %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
+  pred_subset_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)") %>% 
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_subset_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_subset_adglmm[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
   ggplot()+
@@
   geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "fGFPCA", col = "Maximum observation time",
-       x = "Time", y="")+
+  facet_grid(rows = vars(id), cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_adglmm2 <- pred_subset_adglmm[[1]] %>% 
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow = 1)+
-  labs(title = "GLMMadaptive", col = "Maximum observation time",
-       x = "Time", y = "")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_gfosr_l1_2 <- pred_subset_gfosr_l1[[1]] %>% 
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow=1)+
-  labs(title = "GFOSR (L=1)", col = "Maximum observation time",
-       x = "Time", y = "")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-
-fig_gfosr_l5_2 <- pred_subset_gfosr_l5[[1]] %>% 
-  filter(id %in% rand_id) %>%
-  mutate_at(vars(eta_i, starts_with("pred")), function(x){exp(x)/(1+exp(x))}) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred_w1, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w2, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w3, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred_w4, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_wrap(~id, nrow=1)+
-  labs(title = "GFOSR (L=5)", col = "Maximum observation time",
-       x = "Time", y = "")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(axis.text = element_text(size = 5), 
-        title = element_text(size = 8))
-```
-
-```{r fig_sim2, fig.height=8, fig.width=8}
-ggarrange(fig_fgfpca2, fig_adglmm2, fig_gfosr_l1_2, fig_gfosr_l5_2,
-          ncol = 1, common.legend = T)
-```
-
-## ISE
-
-```{r}
+  theme(legend.position = "bottom")+
+  scale_color_manual(values = cols)
+```
+
+## ISE and AUC
+
+```{r ise_subset}
 ## ISE container 
@@
 
-```{r}
-rbind(mean_ise_fgfpca2, mean_ise_adglmm2, 
-           mean_ise_gfosr_l1_2, mean_ise_gfosr_l5_2) %>%
-  as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
-                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
-         .before=1) %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  group_rows(index = c("fGFPCA" = 4, "GLMMadaptive" = 4,
-                     "GFOSR (L=1)" = 4, "GFOSR (L=5)" = 4)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 4))
-```
-
-## AUC
-
-```{r}
+```{r auc2}
 ## auc container 
@@
 
-```{r}
-rbind(mean_auc_fgfpca2, mean_auc_adglmm2, 
-           mean_auc_gfosr_l1_2, mean_auc_gfosr_l5_2) %>%
-  as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", 
-                        "(0.6, 0.8]", "(0.8, 1.0]"), 4),
-         .before=1) %>%
+```{r table_sim2}
+fgfpca2 <- rbind(mean_ise_fgfpca2 ,mean_auc_fgfpca2) %>% as_tibble() %>%
+  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
+gfosr_l5_2 <- rbind(mean_ise_gfosr_l5_2 ,mean_auc_gfosr_l5_2) %>% as_tibble()
+gfosr_l1_2 <- rbind(mean_ise_gfosr_l1_2 ,mean_auc_gfosr_l1_2) %>% as_tibble()
+adglmm2 <- rbind(mean_ise_adglmm2 ,mean_auc_adglmm2) %>% as_tibble()
+
+bind_cols(fgfpca2, gfosr_l5_2, gfosr_l1_2, adglmm2, .name_repair = "minimal") %>%
   kable(digits = 3, booktabs=T,
@@
 
-```{r load_nhanes}
+```{r load_nhanes, eval=FALSE}
 df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
@@
 
-```{r nhance_example}
+```{r nhance_example, eval=FALSE}
 df_nhanes %>% select(SEQN, Z, sind) %>% 
@@
 
-```{r auc_fGFPCA_nhanes}
+```{r auc_fGFPCA_nhanes, eval=FALSE}
 window3 <- c(0, 480, 960, 1440)
@@
 
-```{r auc_GLMMadaptive_nhanes}
+```{r auc_GLMMadaptive_nhanes, eval=FALSE}
 auc_mat_nhanes2 <- nhanes_pred_adglmm %>%
@@
 
-```{r}
+```{r, eval=FALSE}
 data.frame(Window = c("8am-4pm", "4am-midnight"),
@@
 
-```{r pred_nhanes, fig.height=4, fig.width=10}
+```{r pred_nhanes, fig.height=4, fig.width=10, eval=FALSE}
 rand_id3 <- sample(unique(nhanes_pred_fgfpca$id), 4)

### 2024-02-06 Docs/manuscript.Rmd
figures and tables

@@
 output:
+  pdf_document:
+    toc: yes
+    toc_depth: '3'
+    keep_tex: yes
   html_document:
@@
     font: 12pt
-    keep_tex: yes
-  pdf_document:
-    toc: yes
-    toc_depth: '3'
     keep_tex: yes
@@
   add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
-  group_rows(index = c("ISE"= 4, "AUC" = 4))
+  group_rows(index = c("ISE"= 4, "AUC" = 4)) %>%
+  landscape()
 
@@
 
-We take 500 subjects for training, 500 for out-of-sample prediction.
+We take 60% subjects (N = 5257) for training, 40% (n = 3506) for out-of-sample prediction.
 When using the full sample The debiase step (mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only works for intercept-only model.
@@
 
-In addition to reduce sample size, maybe we could just drop the debias step and only interpolate the eigenfunctions?
-
-```{r load_nhanes, eval=FALSE}
-df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
-load(here("Data/ApplOutput.RData"))
+```{r load_nhanes}
+# df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+load(here("Data/ApplOutput_fGFPCA.RData"))
+load(here("Data/ApplOutput_GLMMadaptive.RData"))
+
+load(here("Data/Appl_GFOSR_L5.RData"))
+load(here("Data/Appl_GFOSR_L1.RData"))
+
+pred_appl_gfosr_l1 <- pred_appl_gfosr_l1 %>%
+  rename(pred360=pred_w1, pred720=pred_w2, pred1080=pred_w3)
+pred_appl_gfosr_l5 <- pred_appl_gfosr_l5 %>%
+  rename(pred360=pred_w1, pred720=pred_w2, pred1080=pred_w3)
+
 ```
@@
 
-```{r auc_fGFPCA_nhanes, eval=FALSE}
-window3 <- c(0, 480, 960, 1440)
-
-auc_mat_nhanes <- nhanes_pred_fgfpca %>%
-  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred480),
-            auc2 = get_auc(Y, pred960)) %>%
-  filter(window != "[0,480]") %>% 
+```{r}
+# window
+window <- seq(0, 1440, by = 360)
+```
+
+```{r auc_appl}
+auc_appl_fgfpca <- pred_nhanes_fgfpca %>% 
+    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred360),
+              auc2 = get_auc(Y, pred720),
+              auc3 = get_auc(Y, pred1080)) %>%
+    filter(window != "[0,360]") %>% 
     select(starts_with("auc"))
 
-colnames(auc_mat_nhanes) <- c("8am", "4pm")
-```
-
-```{r auc_GLMMadaptive_nhanes, eval=FALSE}
-auc_mat_nhanes2 <- nhanes_pred_adglmm %>%
-  mutate(window = cut(t, breaks = window3, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred480),
-            auc2 = get_auc(Y, pred960)) %>%
-  filter(window != "[0,480]") %>% 
+auc_appl_gfosr_l5 <- pred_appl_gfosr_l5 %>% 
+  select(-window) %>%
+    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred360),
+              auc2 = get_auc(Y, pred720),
+              auc3 = get_auc(Y, pred1080)) %>%
+    filter(window != "[0,360]") %>% 
     select(starts_with("auc"))
 
-colnames(auc_mat_nhanes2) <- c("8am", "4pm")
-```
-
-```{r, eval=FALSE}
-data.frame(Window = c("8am-4pm", "4am-midnight"),
-           auc_mat_nhanes, auc_mat_nhanes2, check.names = F) %>%
-  kable(digits = 3, caption = "Area Under the ROC curve", booktabs = T) %>%
+auc_appl_gfosr_l1 <- pred_appl_gfosr_l1 %>% 
+  select(-window) %>%
+    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred360),
+              auc2 = get_auc(Y, pred720),
+              auc3 = get_auc(Y, pred1080)) %>%
+    filter(window != "[0,360]") %>% 
+    select(starts_with("auc"))
+
+auc_appl_adglmm <- pred_nhanes_adglmm %>%
+    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
+    select(Y, starts_with("pred"), window) %>%
+    group_by(window) %>%
+    summarise(auc1 = get_auc(Y, pred360),
+              auc2 = get_auc(Y, pred720),
+              auc3 = get_auc(Y, pred1080)) %>%
+    filter(window != "[0,360]") %>% 
+    select(starts_with("auc"))
+
+colnames(auc_appl_fgfpca) <- colnames(auc_appl_gfosr_l5) <- colnames(auc_appl_gfosr_l1) <- colnames(auc_appl_adglmm) <- c("6am", "12pm", "6pm")
+```
+
+```{r table_appl}
+bind_cols(Window = c("6am-12pm", "12pm-6pm", "6pm-12am"),
+          auc_appl_fgfpca, auc_appl_gfosr_l5, auc_appl_gfosr_l1, auc_appl_adglmm,
+          .name_repair = "minimal") %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
   kable_styling(full_width = F) %>% 
-  add_header_above(c(" "=1, "fGFPCA" = 2, "GLMMadaptive" = 2)) %>%
-  add_header_above(c(" "= 1, "Maximum observation time" = 4))
+  add_header_above(c(" " = 1, "fGFPCA" = 3, "GFOSR (L=5)" = 3, 
+                     "GFOSR (L=1)" = 3, "GLMMadaptive" = 3)) %>%
+  add_header_above(c(" " = 1, "Maximum observation time" = 12)) %>%
+  landscape()
 ```
@@
 
-```{r pred_nhanes, fig.height=4, fig.width=10, eval=FALSE}
-rand_id3 <- sample(unique(nhanes_pred_fgfpca$id), 4)
-
-ggarrange(
-  nhanes_pred_fgfpca %>% 
-    filter(id %in% rand_id3) %>%
-    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
+```{r fig_appl, fig.height=8, fig.width=8}
+rand_id <- sample(unique(pred_nhanes_adglmm$id), 4)
+cols <- cols[1:3]
+names(cols) <- c("6am", "12pm", "6pm")
+bind_rows(
+  pred_nhanes_fgfpca %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
+  pred_appl_gfosr_l5 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)"),
+  pred_appl_gfosr_l1 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)"),
+  pred_nhanes_adglmm %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
+  mutate_at(vars(pred360, pred720, pred1080), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
   ggplot()+
-    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
-    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
-    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "fGFPCA")+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480), 
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
-  theme(axis.text = element_text(size = 5)),
-
-  nhanes_pred_adglmm %>% 
-    filter(id %in% rand_id3) %>%
-    mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
-    ggplot()+
-    geom_line(aes(x=t, y = pred480, col = "8am"), linetype = "dashed", na.rm = T)+
-    geom_line(aes(x=t, y = pred960, col = "4pm"),linetype = "dashed", na.rm = T)+
-    geom_point(aes(x=t, y = Y, col = "Outcome"), size = 0.2)+
-    facet_wrap(~id)+
-    labs(x = "Time", y = "GLMMadaptive")+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480), 
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))+
-  theme(axis.text = element_text(size = 5)),
-  nrow = 1, common.legend = T)
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  geom_line(aes(x=sind, y=pred360, col = "6am"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred720, col = "12pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linetype="dashed", na.rm=T)+
+  facet_grid(rows = vars(id), cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="")+
+  scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  theme(legend.position = "bottom", axis.text.x = element_text(size = 7))+
+  scale_color_manual(values = cols)
 ```

### 2024-02-06 Docs/manuscript.tex
figures and tables

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 \author{Ying Jin}
-\date{2023-12-06}
+\date{2024-02-06}
 
@@
 
+\hypertarget{reference-method}{%
+\section{Reference method}\label{reference-method}}
+
+\hypertarget{glmmadaptive}{%
+\subsection{GLMMadaptive}\label{glmmadaptive}}
+
+\begin{itemize}
+\tightlist
+\item
+  For large datasets, we can fit a model with random intercept and slope
+  for time. It is doable on 500 datasets, but obviously too simple for
+  the data generation scheme. We would expect it to perform terribly.
+\end{itemize}
+
+\[g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\]
+
+\begin{itemize}
+\tightlist
+\item
+  For small datasets, we would like to fit fGFPCA and GLMMadaptive on a
+  dataset with smaller sample size and/or smaller measurement density.
+  For the GLMMadaptive model, we would set it up with spline basis
+  functions so that its flexibility is comparable with fGFPCA model,
+  such as:
+\end{itemize}
+
+\[g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)\]
+
+\hypertarget{generalized-functional-on-scalar-regression}{%
+\subsection{Generalized functional-on-scalar
+regression}\label{generalized-functional-on-scalar-regression}}
+
+We would like to use a second reference method for predictive
+performance comparision. In this model, predictions on a specific
+interval are all made using the last few observations in the observed
+window. For example, if we are given observations on {[}0, 0.2{]} (j =
+1\ldots200), we may use L observations taken right before t=0.2
+(\(t_j: j=200, ... 200-(L-1)\)) as time-fixed covariate to predict any
+future time.
+
+Let's write out the model expression (with questionable notation). If
+the observations if up to \(t_m\), then:
+
+\[\begin{aligned}
+g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
+l & = m,...m-(j-1)\\
+t &> t_m 
+\end{aligned}\]
+
+This is a simple function-on-scalar model with no random effect, meaning
+all subject with the same last observed outcome would have the same
+estimated/predicted latent track. This is clearly anti-intuitive. But
+adding random effects would make out-of-sample prediction impossible.
+
+Under both frameworks, for each dataset we need to fit four models
+(similar to the GLMMadaptive method):
+
+\begin{itemize}
+\tightlist
+\item
+  Given 0-0.2, predict 0.2-1
+\item
+  Given 0-0.4, predict 0.4-1
+\item
+  Given 0-0.6, predict 0.6-1
+\item
+  Given 0-0.8, predict 0.8-1
+\end{itemize}
+
+But prediction performance will be reported by equal-length time window.
+
 \hypertarget{larger-scale-simulation}{%
@@
 
-\includegraphics{manuscript_files/figure-latex/fig_sim_data-1.pdf}
+\hypertarget{figures}{%
+\subsection{Figures}\label{figures}}
 
-\hypertarget{reference-method}{%
-\subsection{Reference method}\label{reference-method}}
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SimOutput\_fGFPCA.RData"}\NormalTok{))}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SimOutput\_GLMMadaptive.RData"}\NormalTok{))}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SimOutput\_GFOSR\_L1.RData"}\NormalTok{))}
+\NormalTok{pred\_list\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfofr}
+\NormalTok{t\_vec\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ t\_vec\_gfofr}
+\FunctionTok{rm}\NormalTok{(pred\_list\_gfofr, t\_vec\_gfofr)}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SimOutput\_GFOSR\_L5.RData"}\NormalTok{))}
+\NormalTok{pred\_list\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfofr}
+\NormalTok{t\_vec\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ t\_vec\_gfofr}
+\FunctionTok{rm}\NormalTok{(pred\_list\_gfofr, t\_vec\_gfofr)}
 
-\begin{itemize}
-\tightlist
-\item
-  GLMMadaptive
-\item
-  Here we can fit a model with random intercept and slope for time. It
-  is doable on 500 datasets, but obviously too simple for the data
-  generation scheme. We would expect it to perform terribly.
-\end{itemize}
+\NormalTok{rand\_id }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\DecValTok{501}\SpecialCharTok{:}\DecValTok{600}\NormalTok{, }\AttributeTok{size =} \DecValTok{4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
-\[g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\]
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_list\_all[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_list\_ref[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{),}
+\NormalTok{  pred\_list\_gfosr\_l1[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_list\_gfosr\_l5[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4)}
+\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(eta\_i, pred0}\FloatTok{.2}\NormalTok{, pred0}\FloatTok{.4}\NormalTok{, pred0}\FloatTok{.6}\NormalTok{, pred0}\FloatTok{.8}\NormalTok{), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"Time"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.2}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{manuscript_files/figure-latex/fig_large_sim-1.pdf}
+
+\hypertarget{ise-and-auc}{%
+\subsection{ISE and AUC}\label{ise-and-auc}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# ISE container }
+\NormalTok{ise\_fgfpca }\OtherTok{\textless{}{-}}\NormalTok{ ise\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ ise\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ ise\_gfosr\_l5 }\OtherTok{\textless{}{-}}
+  \FunctionTok{array}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{dim =} \FunctionTok{c}\NormalTok{(}\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, M))}
+\CommentTok{\# dims: prediction window, max obs time, simulation iter}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# calculation}
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+  \CommentTok{\# this\_df \textless{}{-} pred\_list\_all[[m]]}
+\NormalTok{  ise\_tb }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_all[[m]] }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred0}\FloatTok{.2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred0}\FloatTok{.4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred0}\FloatTok{.6}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred0}\FloatTok{.8}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{))}
+\NormalTok{  ise\_fgfpca[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(ise\_tb)}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_fgfpca }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_fgfpca, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+
+\CommentTok{\# mean\_ise \textless{}{-} data.frame(mean\_ise) \%\textgreater{}\%}
+\CommentTok{\#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),}
+\CommentTok{\#          .before = 1)}
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_fgfpca) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# calculation}
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+  \CommentTok{\# this\_df \textless{}{-} pred\_list\_gfosr\_l1[[m]]}
+\NormalTok{  ise\_tb }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_ref[[m]] }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred0}\FloatTok{.2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred0}\FloatTok{.4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred0}\FloatTok{.6}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred0}\FloatTok{.8}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{))}
+\NormalTok{  ise\_adglmm[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(ise\_tb)}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_adglmm }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_adglmm, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_adglmm) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# calculation}
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+  \CommentTok{\# this\_df \textless{}{-} pred\_list\_ref[[m]]}
+\NormalTok{  ise\_tb }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfosr\_l1[[m]] }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred\_w1}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred\_w2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred\_w3}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred\_w4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \CommentTok{\# mutate(window = factor(window, levels = 1:5, }
+    \CommentTok{\#                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) \%\textgreater{}\% }
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{))}
+\NormalTok{  ise\_gfosr\_l1[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(ise\_tb)}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_gfosr\_l1 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_gfosr\_l1, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_gfosr\_l1) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# calculation}
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+  \CommentTok{\# this\_df \textless{}{-} pred\_list\_ref[[m]]}
+\NormalTok{  ise\_tb }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfosr\_l5[[m]] }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred\_w1}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred\_w2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred\_w3}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred\_w4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \CommentTok{\# mutate(window = factor(window, levels = 1:5, }
+    \CommentTok{\#                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) \%\textgreater{}\% }
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{))}
+\NormalTok{  ise\_gfosr\_l5[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(ise\_tb)}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_gfosr\_l5 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_gfosr\_l5, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_gfosr\_l5) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# a function to calculate AUC}
+\NormalTok{get\_auc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(y, pred)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{sum}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(y))}\SpecialCharTok{\textgreater{}}\DecValTok{0} \SpecialCharTok{|} \FunctionTok{sum}\NormalTok{(}\FunctionTok{is.na}\NormalTok{(pred))}\SpecialCharTok{\textgreater{}}\DecValTok{0}\NormalTok{)\{}
+\NormalTok{    auc }\OtherTok{\textless{}{-}} \ConstantTok{NA}
+\NormalTok{  \}}
+  \ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{    this\_perf }\OtherTok{\textless{}{-}} \FunctionTok{performance}\NormalTok{(}\FunctionTok{prediction}\NormalTok{(pred, y), }\AttributeTok{measure =} \StringTok{"auc"}\NormalTok{)}
+\NormalTok{    auc }\OtherTok{\textless{}{-}}\NormalTok{ this\_perf}\SpecialCharTok{@}\NormalTok{y.values[[}\DecValTok{1}\NormalTok{]]}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(auc)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# auc container }
+\NormalTok{auc\_fgfpca }\OtherTok{\textless{}{-}}\NormalTok{ auc\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ auc\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ auc\_gfosr\_l5 }\OtherTok{\textless{}{-}}
+  \FunctionTok{array}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{dim =} \FunctionTok{c}\NormalTok{(}\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, M))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_all[[m]]}
+\NormalTok{  auc\_tb }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.2}\NormalTok{),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.4}\NormalTok{),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.6}\NormalTok{),}
+              \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.8}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+\NormalTok{  auc\_fgfpca[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(auc\_tb)}
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_fgfpca }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_fgfpca, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\CommentTok{\# mean\_auc \textless{}{-} data.frame(mean\_auc) \%\textgreater{}\% }
+\CommentTok{\#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),}
+\CommentTok{\#          .before = 1)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_fgfpca) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_ref[[m]]}
+\NormalTok{  auc\_tb }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.2}\NormalTok{),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.4}\NormalTok{),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.6}\NormalTok{),}
+              \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.8}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+\NormalTok{  auc\_adglmm[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(auc\_tb)}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_adglmm }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_adglmm, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_adglmm) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfosr\_l1[[m]]}
+\NormalTok{  auc\_tb }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w1),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w2),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w3),}
+              \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w4)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+\NormalTok{  auc\_gfosr\_l1[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(auc\_tb)}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_gfosr\_l1 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_gfosr\_l1, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_gfosr\_l1) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{M)\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfosr\_l5[[m]]}
+\NormalTok{  auc\_tb }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w1),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w2),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w3),}
+              \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w4)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+\NormalTok{  auc\_gfosr\_l5[, ,m] }\OtherTok{\textless{}{-}} \FunctionTok{as.matrix}\NormalTok{(auc\_tb)}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_gfosr\_l5 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_gfosr\_l5, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_gfosr\_l5) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fgfpca1 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_fgfpca ,mean\_auc\_fgfpca) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Window =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"(0.2, 0.4]"}\NormalTok{, }\StringTok{"(0.4, 0.6]"}\NormalTok{, }\StringTok{"(0.6, 0.8]"}\NormalTok{, }\StringTok{"(0.8, 1.0]"}\NormalTok{), }\DecValTok{2}\NormalTok{), }\AttributeTok{.before=}\DecValTok{1}\NormalTok{)}
+\NormalTok{gfosr\_l5\_1 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_gfosr\_l5 ,mean\_auc\_gfosr\_l5) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{()}
+\NormalTok{gfosr\_l1\_1 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_gfosr\_l1 ,mean\_auc\_gfosr\_l1) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{()}
+\NormalTok{adglmm1 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_adglmm ,mean\_auc\_adglmm) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{() }
+
+\FunctionTok{bind\_cols}\NormalTok{(fgfpca1, gfosr\_l5\_1, gfosr\_l1\_1, adglmm1, }\AttributeTok{.name\_repair =} \StringTok{"minimal"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs=}\NormalTok{T,}
+        \AttributeTok{table.attr=}\StringTok{"style=}\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"fGFPCA"} \OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"GFOSR (L=5)"} \OtherTok{=} \DecValTok{4}\NormalTok{, }
+                     \StringTok{"GFOSR (L=1)"} \OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"GLMMadaptive"} \OtherTok{=} \DecValTok{4}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"Maximum observation time"} \OtherTok{=} \DecValTok{16}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_rows}\NormalTok{(}\AttributeTok{index =} \FunctionTok{c}\NormalTok{(}\StringTok{"ISE"}\OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"AUC"} \OtherTok{=} \DecValTok{4}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{landscape}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{landscape}\begin{table}
+\centering
+\begin{tabular}{lrrrrrrrrrrrrrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{16}{c}{Maximum observation time} \\
+\cmidrule(l{3pt}r{3pt}){2-17}
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GFOSR (L=5)} & \multicolumn{4}{c}{GFOSR (L=1)} & \multicolumn{4}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13} \cmidrule(l{3pt}r{3pt}){14-17}
+Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{17}{l}{\textbf{ISE}}\\
+\hspace{1em}(0.2, 0.4] & 146.407 &  &  &  & 274.947 &  &  &  & 362.476 &  &  &  & 387.708 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 183.967 & 74.977 &  &  & 277.438 & 220.209 &  &  & 286.614 & 262.552 &  &  & 291.579 & 269.799 &  & \\
+\hspace{1em}(0.6, 0.8] & 218.265 & 49.275 & 15.776 &  & 322.338 & 373.108 & 325.017 &  & 385.701 & 410.508 & 389.305 &  & 315.778 & 282.736 & 278.242 & \\
+\hspace{1em}(0.8, 1.0] & 108.918 & 77.981 & 17.747 & 12.005 & 290.982 & 318.079 & 350.850 & 333.580 & 328.482 & 341.274 & 354.211 & 347.067 & 563.011 & 477.485 & 597.746 & 600.340\\
+\addlinespace[0.3em]
+\multicolumn{17}{l}{\textbf{AUC}}\\
+\hspace{1em}(0.2, 0.4] & 0.748 &  &  &  & 0.686 &  &  &  & 0.624 &  &  &  & 0.591 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 0.664 & 0.734 &  &  & 0.563 & 0.630 &  &  & 0.543 & 0.590 &  &  & 0.524 & 0.596 &  & \\
+\hspace{1em}(0.6, 0.8] & 0.715 & 0.790 & 0.803 &  & 0.669 & 0.628 & 0.676 &  & 0.604 & 0.577 & 0.615 &  & 0.669 & 0.694 & 0.687 & \\
+\hspace{1em}(0.8, 1.0] & 0.740 & 0.755 & 0.781 & 0.784 & 0.626 & 0.606 & 0.552 & 0.584 & 0.588 & 0.564 & 0.537 & 0.551 & 0.514 & 0.556 & 0.526 & 0.564\\
+\bottomrule
+\end{tabular}
+\end{table}
+\end{landscape}
+
+\hypertarget{time-minutes}{%
+\subsection{Time (minutes)}\label{time-minutes}}
+
+I chose not to report fitting and prediction time separately because it
+is too difficult to tell them apart for the GFOSR models.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{data.frame}\NormalTok{(}
+  \StringTok{"Method"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GLMMadaptive"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{),}
+  \StringTok{"Time"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time),}
+             \FunctionTok{mean}\NormalTok{(fit\_time\_ref}\SpecialCharTok{+}\NormalTok{pred\_time\_ref}\SpecialCharTok{/}\DecValTok{60}\NormalTok{),}
+             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l1}\SpecialCharTok{/}\DecValTok{60}\NormalTok{), }
+             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs =}\NormalTok{ T, }\AttributeTok{table.attr =} \StringTok{"style = }\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+\centering
+\begin{tabular}{lr}
+\toprule
+Method & Time\\
+\midrule
+fGFPCA & 2.317\\
+GLMMadaptive & 2.304\\
+GFOSR (L=1) & 0.025\\
+GFOSR (L=5) & 0.144\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\hypertarget{small-scale-simulation}{%
+\section{Small-scale simulation}\label{small-scale-simulation}}
+
+I have used 100 subjects for training and testing, and repeated 500
+times. When fitting GLMMadpative, I reduce the number of measurements in
+the training dataset to 1/10 by taking one every 10 observations. The
+prediction is on the original grid.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SubSimOutput\_GLMMadaptive.RData"}\NormalTok{))}
+
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SubSimOutput\_fGFPCA.RData"}\NormalTok{))}
+\NormalTok{pred\_subset\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_adglmm[}\SpecialCharTok{!}\NormalTok{num\_probs]}
+\NormalTok{fit\_time\_subset\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ fit\_time\_subset\_adglmm[}\SpecialCharTok{!}\NormalTok{num\_probs]}
+\NormalTok{pred\_time\_subset\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ pred\_time\_subset\_adglmm[}\SpecialCharTok{!}\NormalTok{num\_probs]}
+
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SubSimOutput\_GFOSR\_L1.RData"}\NormalTok{))}
+\NormalTok{pred\_subset\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfofr\_subset}
+\NormalTok{t\_subset\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ t\_vec\_gfofr\_subset}
+\FunctionTok{rm}\NormalTok{(pred\_list\_gfofr\_subset, t\_vec\_gfofr\_subset)}
+
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/SubSimOutput\_GFOSR\_L5.RData"}\NormalTok{))}
+\NormalTok{pred\_subset\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_list\_gfofr\_subset}
+\NormalTok{t\_subset\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ t\_vec\_gfofr\_subset}
+\FunctionTok{rm}\NormalTok{(pred\_list\_gfofr\_subset, t\_vec\_gfofr\_subset)}
+
+\NormalTok{rand\_id }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\DecValTok{101}\SpecialCharTok{:}\DecValTok{200}\NormalTok{, }\AttributeTok{size =} \DecValTok{4}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/fig_sim-1.pdf}
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_subset\_fGFPCA[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_subset\_gfosr\_l5[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_subset\_gfosr\_l1[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_subset\_adglmm[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(eta\_i, pred0}\FloatTok{.2}\NormalTok{, pred0}\FloatTok{.4}\NormalTok{, pred0}\FloatTok{.6}\NormalTok{, pred0}\FloatTok{.8}\NormalTok{), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"Time"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.2}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols)}
+\end{Highlighting}
+\end{Shaded}
 
-\hypertarget{ise}{%
-\subsection{ISE}\label{ise}}
+\includegraphics{manuscript_files/figure-latex/fig_small_sim-1.pdf}
 
-\begin{table}
+\hypertarget{ise-and-auc-1}{%
+\subsection{ISE and AUC}\label{ise-and-auc-1}}
 
-\caption{\label{tab:unnamed-chunk-4}Integrated squared error}
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# ISE container }
+\NormalTok{ise\_fgfpca2 }\OtherTok{\textless{}{-}}\NormalTok{ ise\_adglmm2 }\OtherTok{\textless{}{-}}\NormalTok{ ise\_gfosr\_l1\_2 }\OtherTok{\textless{}{-}}\NormalTok{ ise\_gfosr\_l5\_2 }\OtherTok{\textless{}{-}}
+  \FunctionTok{array}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{dim =} \FunctionTok{c}\NormalTok{(}\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, M))}
+\CommentTok{\# dims: prediction window, max obs time, simulation iter}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_fGFPCA))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_fGFPCA[[m]]}
+\NormalTok{  ise\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred0}\FloatTok{.2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred0}\FloatTok{.4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred0}\FloatTok{.6}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred0}\FloatTok{.8}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+\NormalTok{  ise\_fgfpca2[,,m] }\OtherTok{\textless{}{-}}\NormalTok{ ise\_tb\_m}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_fgfpca2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_fgfpca2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean)}
+
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_fgfpca2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_adglmm))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_adglmm[[m]]}
+\NormalTok{  ise\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred0}\FloatTok{.2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred0}\FloatTok{.4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred0}\FloatTok{.6}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred0}\FloatTok{.8}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+\NormalTok{  ise\_adglmm2[,,m] }\OtherTok{\textless{}{-}}\NormalTok{ ise\_tb\_m}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_adglmm2}\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_adglmm2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean, }\AttributeTok{na.rm =}\NormalTok{ T)}
+
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_adglmm2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_gfosr\_l1))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_gfosr\_l1[[m]]}
+\NormalTok{  ise\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred\_w1}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred\_w2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred\_w3}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred\_w4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+\NormalTok{  ise\_gfosr\_l1\_2[,,m] }\OtherTok{\textless{}{-}}\NormalTok{ ise\_tb\_m}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_gfosr\_l1\_2}\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_gfosr\_l1\_2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean, }\AttributeTok{na.rm =}\NormalTok{ T)}
+
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_gfosr\_l1\_2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_gfosr\_l5))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_gfosr\_l5[[m]]}
+\NormalTok{  ise\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{err1 =}\NormalTok{ (pred\_w1}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err2 =}\NormalTok{ (pred\_w2}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err3 =}\NormalTok{ (pred\_w3}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,}
+           \AttributeTok{err4 =}\NormalTok{ (pred\_w4}\SpecialCharTok{{-}}\NormalTok{eta\_i)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(id, t, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window, id) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarise\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), sum) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{summarize\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(err1, err2, err3, err4), mean) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"err"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+\NormalTok{  ise\_gfosr\_l5\_2[,,m] }\OtherTok{\textless{}{-}}\NormalTok{ ise\_tb\_m}
+\NormalTok{\}}
+
+\NormalTok{mean\_ise\_gfosr\_l5\_2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(ise\_gfosr\_l5\_2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), mean, }\AttributeTok{na.rm =}\NormalTok{ T)}
+
+\FunctionTok{colnames}\NormalTok{(mean\_ise\_gfosr\_l5\_2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# auc container }
+\NormalTok{auc\_fgfpca2 }\OtherTok{\textless{}{-}}\NormalTok{ auc\_adglmm2 }\OtherTok{\textless{}{-}}\NormalTok{ auc\_gfosr\_l1\_2 }\OtherTok{\textless{}{-}}\NormalTok{ auc\_gfosr\_l5\_2 }\OtherTok{\textless{}{-}}
+  \FunctionTok{array}\NormalTok{(}\ConstantTok{NA}\NormalTok{, }\AttributeTok{dim =} \FunctionTok{c}\NormalTok{(}\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\FunctionTok{length}\NormalTok{(window)}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, M))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_fGFPCA))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_fGFPCA[[m]]}
+\NormalTok{  auc\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.2}\NormalTok{),}
+            \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.4}\NormalTok{),}
+            \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.6}\NormalTok{),}
+            \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.8}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+
+\NormalTok{  auc\_fgfpca2[, , m] }\OtherTok{\textless{}{-}}\NormalTok{ auc\_tb\_m}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_fgfpca2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_fgfpca2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_fgfpca2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_adglmm))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_adglmm[[m]]}
+\NormalTok{  auc\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(t, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.2}\NormalTok{),}
+            \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.4}\NormalTok{),}
+            \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.6}\NormalTok{),}
+            \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred0}\FloatTok{.8}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,0.2]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+
+\NormalTok{  auc\_adglmm2[, , m] }\OtherTok{\textless{}{-}}\NormalTok{ auc\_tb\_m}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_adglmm2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_adglmm2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), mean, }\AttributeTok{na.rm =}\NormalTok{ T)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_adglmm2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_gfosr\_l1))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_gfosr\_l1[[m]]}
+\NormalTok{  auc\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w1),}
+            \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w2),}
+            \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w3),}
+            \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w4)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+
+\NormalTok{  auc\_gfosr\_l1\_2[, , m] }\OtherTok{\textless{}{-}}\NormalTok{ auc\_tb\_m}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_gfosr\_l1\_2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_gfosr\_l1\_2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_gfosr\_l1\_2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\ControlFlowTok{for}\NormalTok{(m }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(pred\_subset\_gfosr\_l5))\{}
+\NormalTok{  this\_df }\OtherTok{\textless{}{-}}\NormalTok{ pred\_subset\_gfosr\_l5[[m]]}
+\NormalTok{  auc\_tb\_m }\OtherTok{\textless{}{-}}\NormalTok{ this\_df }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w1),}
+            \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w2),}
+            \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w3),}
+            \AttributeTok{auc4 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred\_w4)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as.matrix}\NormalTok{()}
+
+\NormalTok{  auc\_gfosr\_l5\_2[, , m] }\OtherTok{\textless{}{-}}\NormalTok{ auc\_tb\_m}
+
+\NormalTok{\}}
+
+\NormalTok{mean\_auc\_gfosr\_l5\_2 }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(auc\_gfosr\_l5\_2, }\FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{), mean)}
+\FunctionTok{colnames}\NormalTok{(mean\_auc\_gfosr\_l5\_2) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"0.2"}\NormalTok{, }\StringTok{"0.4"}\NormalTok{, }\StringTok{"0.6"}\NormalTok{, }\StringTok{"0.8"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fgfpca2 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_fgfpca2 ,mean\_auc\_fgfpca2) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Window =} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"(0.2, 0.4]"}\NormalTok{, }\StringTok{"(0.4, 0.6]"}\NormalTok{, }\StringTok{"(0.6, 0.8]"}\NormalTok{, }\StringTok{"(0.8, 1.0]"}\NormalTok{), }\DecValTok{2}\NormalTok{), }\AttributeTok{.before=}\DecValTok{1}\NormalTok{)}
+\NormalTok{gfosr\_l5\_2 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_gfosr\_l5\_2 ,mean\_auc\_gfosr\_l5\_2) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{()}
+\NormalTok{gfosr\_l1\_2 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_gfosr\_l1\_2 ,mean\_auc\_gfosr\_l1\_2) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{()}
+\NormalTok{adglmm2 }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(mean\_ise\_adglmm2 ,mean\_auc\_adglmm2) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{as\_tibble}\NormalTok{() }
+
+\FunctionTok{bind\_cols}\NormalTok{(fgfpca2, gfosr\_l5\_2, gfosr\_l1\_2, adglmm2, }\AttributeTok{.name\_repair =} \StringTok{"minimal"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs=}\NormalTok{T,}
+        \AttributeTok{table.attr=}\StringTok{"style=}\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"fGFPCA"} \OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"GFOSR (L=5)"} \OtherTok{=} \DecValTok{4}\NormalTok{, }
+                     \StringTok{"GFOSR (L=1)"} \OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"GLMMadaptive"} \OtherTok{=} \DecValTok{4}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"Maximum observation time"} \OtherTok{=} \DecValTok{16}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{group\_rows}\NormalTok{(}\AttributeTok{index =} \FunctionTok{c}\NormalTok{(}\StringTok{"ISE"}\OtherTok{=} \DecValTok{4}\NormalTok{, }\StringTok{"AUC"} \OtherTok{=} \DecValTok{4}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{landscape}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{landscape}\begin{table}
 \centering
-\begin{tabular}[t]{lrrrrrrrr}
+\begin{tabular}{lrrrrrrrrrrrrrrrr}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observation time} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
-Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
+\multicolumn{1}{c}{ } & \multicolumn{16}{c}{Maximum observation time} \\
+\cmidrule(l{3pt}r{3pt}){2-17}
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GFOSR (L=5)} & \multicolumn{4}{c}{GFOSR (L=1)} & \multicolumn{4}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13} \cmidrule(l{3pt}r{3pt}){14-17}
+Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
 \midrule
-(0.2, 0.4] & 146.407 &  &  &  & 387.708 &  &  & \\
-(0.4, 0.6] & 183.967 & 74.977 &  &  & 291.579 & 269.799 &  & \\
-(0.6, 0.8] & 218.265 & 49.275 & 15.776 &  & 315.778 & 282.736 & 278.242 & \\
-(0.8, 1.0] & 108.918 & 77.981 & 17.747 & 12.005 & 563.011 & 477.485 & 597.746 & 600.34\\
+\addlinespace[0.3em]
+\multicolumn{17}{l}{\textbf{ISE}}\\
+\hspace{1em}(0.2, 0.4] & 150.641 &  &  &  & 286.853 &  &  &  & 367.282 &  &  &  & 374.836 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 188.065 & 77.380 &  &  & 286.628 & 227.225 &  &  & 290.781 & 265.809 &  &  & 268.390 & 463.337 &  & \\
+\hspace{1em}(0.6, 0.8] & 224.001 & 51.533 & 16.879 &  & 328.612 & 374.572 & 326.522 &  & 386.705 & 408.427 & 388.427 &  & 319.731 & 287.001 & 233.510 & \\
+\hspace{1em}(0.8, 1.0] & 112.163 & 81.368 & 19.592 & 13.399 & 301.165 & 324.817 & 353.888 & 335.139 & 331.488 & 342.935 & 354.800 & 347.627 & 228.335 & 469.769 & 331.509 & 131.993\\
+\addlinespace[0.3em]
+\multicolumn{17}{l}{\textbf{AUC}}\\
+\hspace{1em}(0.2, 0.4] & 0.746 &  &  &  & 0.679 &  &  &  & 0.620 &  &  &  & 0.672 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 0.662 & 0.734 &  &  & 0.549 & 0.627 &  &  & 0.536 & 0.590 &  &  & 0.622 & 0.671 &  & \\
+\hspace{1em}(0.6, 0.8] & 0.710 & 0.787 & 0.801 &  & 0.656 & 0.617 & 0.668 &  & 0.598 & 0.574 & 0.613 &  & 0.689 & 0.698 & 0.731 & \\
+\hspace{1em}(0.8, 1.0] & 0.740 & 0.754 & 0.781 & 0.784 & 0.611 & 0.588 & 0.551 & 0.578 & 0.580 & 0.556 & 0.536 & 0.551 & 0.681 & 0.629 & 0.679 & 0.743\\
 \bottomrule
@@
 \end{table}
+\end{landscape}
 
-\hypertarget{auc}{%
-\subsection{AUC}\label{auc}}
+Among 500 hundred iterations, 6 did not converge for the GLMMadaptive
+model. No numeric issue for fGFPCA.
+
+\hypertarget{time}{%
+\section{Time}\label{time}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{data.frame}\NormalTok{(}
+  \StringTok{"Method"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GLMMadaptive"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{),}
+  \StringTok{"Time"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time\_subset\_fGFPCA}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time\_subset\_fGFPCA),}
+             \FunctionTok{mean}\NormalTok{(fit\_time\_subset\_adglmm}\SpecialCharTok{+}\NormalTok{pred\_time\_subset\_adglmm}\SpecialCharTok{/}\DecValTok{60}\NormalTok{),}
+             \FunctionTok{mean}\NormalTok{(t\_subset\_gfosr\_l1}\SpecialCharTok{/}\DecValTok{60}\NormalTok{), }
+             \FunctionTok{mean}\NormalTok{(t\_subset\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs =}\NormalTok{ T, }\AttributeTok{table.attr =} \StringTok{"style = }\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }
+\end{Highlighting}
+\end{Shaded}
 
 \begin{table}
-
-\caption{\label{tab:unnamed-chunk-6}Area under the ROC curve}
 \centering
-\begin{tabular}[t]{lrrrrrrrr}
+\begin{tabular}{lr}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observation time} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
-Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
+Method & Time\\
 \midrule
-(0.2, 0.4] & 0.748 &  &  &  & 0.591 &  &  & \\
-(0.4, 0.6] & 0.664 & 0.734 &  &  & 0.524 & 0.596 &  & \\
-(0.6, 0.8] & 0.715 & 0.790 & 0.803 &  & 0.669 & 0.694 & 0.687 & \\
-(0.8, 1.0] & 0.740 & 0.755 & 0.781 & 0.784 & 0.514 & 0.556 & 0.526 & 0.564\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:time}Computation time (minutes)}
-\centering
-\begin{tabular}[t]{lrr}
-\toprule
-Method & Fit & Prediction\\
-\midrule
-fGFPCA & 0.725 & 1.592\\
-GLMMadaptive & 2.287 & 0.017\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-I think we could say that while the total time spend on model fitting +
-prediction are similar between two methods, fGFPCA achieved much better
-flexibility and much better predictive performance of prediction under
-every scenario.
-
-\hypertarget{small-scale-simulation}{%
-\section{Small-scale simulation}\label{small-scale-simulation}}
-
-Here we would like to fit fGFPCA and GLMMadaptive on a dataset with
-smaller sample size and/or smaller measurement density. For the
-GLMMadaptive model, we would set it up with spline basis functions so
-that its flexibility is comparable with fGFPCA model, such as:
-
-\[g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)\]
-
-I have used 100 subjects for training and testing, and repeated 100
-times. When fitting GLMMadpative, I reduce the number of measurements to
-1/10 by taking one every 10 observations. The prediction is on the
-original grid.
-
-\hypertarget{figure-1}{%
-\subsection{Figure}\label{figure-1}}
-
-\includegraphics{manuscript_files/figure-latex/fit_sim_small-1.pdf}
-
-\hypertarget{ise-1}{%
-\subsection{ISE}\label{ise-1}}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-8}Integrated squared error}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Observation track} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
-Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
-\midrule
-(0.2,0.4] & 151.709 &  &  &  & 374.368 &  &  & \\
-(0.4,0.6] & 189.147 & 77.904 &  &  & 268.805 & 464.796 &  & \\
-(0.6,0.8] & 226.314 & 51.773 & 17.050 &  & 321.530 & 286.806 & 230.564 & \\
-(0.8, 1.0] & 113.007 & 81.718 & 19.576 & 13.244 & 227.965 & 472.162 & 341.146 & 136.831\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{auc-1}{%
-\subsection{AUC}\label{auc-1}}
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-10}Area under the ROC curve}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Observation track} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
-\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
-Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
-\midrule
-(0.2,0.4] & 0.745 &  &  &  & 0.674 &  &  & \\
-(0.4,0.6] & 0.663 & 0.733 &  &  & 0.623 & 0.671 &  & \\
-(0.6,0.8] & 0.712 & 0.789 & 0.802 &  & 0.691 & 0.700 & 0.732 & \\
-(0.8, 1.0] & 0.741 & 0.755 & 0.782 & 0.785 & 0.680 & 0.627 & 0.679 & 0.743\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-There is huge difference in compuation time. GLMMadaptive took 23
-minutes, while fGFPCA took less than 3 seconds.
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-11}Computation time (minutes)}
-\centering
-\begin{tabular}[t]{lrr}
-\toprule
-Method & Fit & Prediction\\
-\midrule
-fGFPCA & 0.043 & 1.405\\
-GLMMadaptive & 5.842 & 0.023\\
+fGFPCA & 1.629\\
+GLMMadaptive & 5.479\\
+GFOSR (L=1) & 0.010\\
+GFOSR (L=5) & 0.054\\
 \bottomrule
@@
 
-We take 80\% (7010) subjects for training, 20\% (1753) for out-of-sample
-prediction.
+We take 60\% subjects (N = 5257) for training, 40\% (n = 3506) for
+out-of-sample prediction. When using the full sample The debiase step
+(mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only
+works for intercept-only model. In the debias step, re-evaluated
+eigenvalues also lost their decreasing nature.
 
-In the GLMMadaptive model, I could only fit a model with random
-intercept. It took more than 20 minutes for model fitting.
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# df\_nhanes \textless{}{-} read\_rds(here("Data/nhanes\_bi.rds"))}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/ApplOutput\_fGFPCA.RData"}\NormalTok{))}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/ApplOutput\_GLMMadaptive.RData"}\NormalTok{))}
 
-\includegraphics{manuscript_files/figure-latex/nhance_example-1.pdf}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/Appl\_GFOSR\_L5.RData"}\NormalTok{))}
+\FunctionTok{load}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/Appl\_GFOSR\_L1.RData"}\NormalTok{))}
 
-\hypertarget{auc-2}{%
-\subsection{AUC}\label{auc-2}}
+\NormalTok{pred\_appl\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred360=}\NormalTok{pred\_w1, }\AttributeTok{pred720=}\NormalTok{pred\_w2, }\AttributeTok{pred1080=}\NormalTok{pred\_w3)}
+\NormalTok{pred\_appl\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred360=}\NormalTok{pred\_w1, }\AttributeTok{pred720=}\NormalTok{pred\_w2, }\AttributeTok{pred1080=}\NormalTok{pred\_w3)}
+\end{Highlighting}
+\end{Shaded}
 
-\begin{table}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{df\_nhanes }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(SEQN, Z, sind) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{filter}\NormalTok{(SEQN }\SpecialCharTok{\%in\%} \FunctionTok{unique}\NormalTok{(nhanes\_pred\_adglmm}\SpecialCharTok{$}\NormalTok{id) }\SpecialCharTok{|}\NormalTok{ SEQN }\SpecialCharTok{\%in\%} \FunctionTok{unique}\NormalTok{(nhanes\_pred\_fgfpca}\SpecialCharTok{$}\NormalTok{id)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{Z =} \FunctionTok{factor}\NormalTok{(Z, }\AttributeTok{levels =} \DecValTok{0}\SpecialCharTok{:}\DecValTok{1}\NormalTok{, }\AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"Inactive"}\NormalTok{, }\StringTok{"Active"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{SEQN =} \FunctionTok{factor}\NormalTok{(SEQN)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_tile}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y =}\NormalTok{ SEQN, }\AttributeTok{fill =}\NormalTok{ Z))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x=}\StringTok{"Time"}\NormalTok{, }\AttributeTok{y=}\StringTok{"Subject"}\NormalTok{, }\AttributeTok{title =} \StringTok{"Overview of NHANES binary activity indicator"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.y =} \FunctionTok{element\_blank}\NormalTok{())}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1440}\NormalTok{, }\AttributeTok{by =} \DecValTok{480}\NormalTok{),}
+                     \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"Midnight"}\NormalTok{, }\StringTok{"8am"}\NormalTok{, }\StringTok{"4pm"}\NormalTok{, }\StringTok{"Midnight"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-\caption{\label{tab:unnamed-chunk-12}Area Under the ROC curve}
+\hypertarget{auc}{%
+\subsection{AUC}\label{auc}}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# window}
+\NormalTok{window }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1440}\NormalTok{, }\AttributeTok{by =} \DecValTok{360}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{auc\_appl\_fgfpca }\OtherTok{\textless{}{-}}\NormalTok{ pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(sind, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred360),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred720),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred1080)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,360]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+
+\NormalTok{auc\_appl\_gfosr\_l5 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(sind, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred360),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred720),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred1080)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,360]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+
+\NormalTok{auc\_appl\_gfosr\_l1 }\OtherTok{\textless{}{-}}\NormalTok{ pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(sind, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred360),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred720),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred1080)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,360]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+
+\NormalTok{auc\_appl\_adglmm }\OtherTok{\textless{}{-}}\NormalTok{ pred\_nhanes\_adglmm }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{window =} \FunctionTok{cut}\NormalTok{(sind, }\AttributeTok{breaks =}\NormalTok{ window, }\AttributeTok{include.lowest =}\NormalTok{ T)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{select}\NormalTok{(Y, }\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"pred"}\NormalTok{), window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{group\_by}\NormalTok{(window) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{auc1 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred360),}
+              \AttributeTok{auc2 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred720),}
+              \AttributeTok{auc3 =} \FunctionTok{get\_auc}\NormalTok{(Y, pred1080)) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{filter}\NormalTok{(window }\SpecialCharTok{!=} \StringTok{"[0,360]"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+    \FunctionTok{select}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"auc"}\NormalTok{))}
+
+\FunctionTok{colnames}\NormalTok{(auc\_appl\_fgfpca) }\OtherTok{\textless{}{-}} \FunctionTok{colnames}\NormalTok{(auc\_appl\_gfosr\_l5) }\OtherTok{\textless{}{-}} \FunctionTok{colnames}\NormalTok{(auc\_appl\_gfosr\_l1) }\OtherTok{\textless{}{-}} \FunctionTok{colnames}\NormalTok{(auc\_appl\_adglmm) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"6am"}\NormalTok{, }\StringTok{"12pm"}\NormalTok{, }\StringTok{"6pm"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{bind\_cols}\NormalTok{(}\AttributeTok{Window =} \FunctionTok{c}\NormalTok{(}\StringTok{"6am{-}12pm"}\NormalTok{, }\StringTok{"12pm{-}6pm"}\NormalTok{, }\StringTok{"6pm{-}12am"}\NormalTok{),}
+\NormalTok{          auc\_appl\_fgfpca, auc\_appl\_gfosr\_l5, auc\_appl\_gfosr\_l1, auc\_appl\_adglmm,}
+          \AttributeTok{.name\_repair =} \StringTok{"minimal"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs=}\NormalTok{T,}
+        \AttributeTok{table.attr=}\StringTok{"style=}\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"fGFPCA"} \OtherTok{=} \DecValTok{3}\NormalTok{, }\StringTok{"GFOSR (L=5)"} \OtherTok{=} \DecValTok{3}\NormalTok{, }
+                     \StringTok{"GFOSR (L=1)"} \OtherTok{=} \DecValTok{3}\NormalTok{, }\StringTok{"GLMMadaptive"} \OtherTok{=} \DecValTok{3}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{add\_header\_above}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{" "} \OtherTok{=} \DecValTok{1}\NormalTok{, }\StringTok{"Maximum observation time"} \OtherTok{=} \DecValTok{12}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{landscape}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{landscape}\begin{table}
 \centering
-\begin{tabular}[t]{lrrrr}
+\begin{tabular}{lrrrrrrrrrrrr}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Maximum observation time} \\
-\cmidrule(l{3pt}r{3pt}){2-5}
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{fGFPCA} & \multicolumn{2}{c}{GLMMadaptive} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Window & 8am & 4pm & 8am & 4pm\\
+\multicolumn{1}{c}{ } & \multicolumn{12}{c}{Maximum observation time} \\
+\cmidrule(l{3pt}r{3pt}){2-13}
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{fGFPCA} & \multicolumn{3}{c}{GFOSR (L=5)} & \multicolumn{3}{c}{GFOSR (L=1)} & \multicolumn{3}{c}{GLMMadaptive} \\
+\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
+Window & 6am & 12pm & 6pm & 6am & 12pm & 6pm & 6am & 12pm & 6pm & 6am & 12pm & 6pm\\
 \midrule
-8am-4pm & 0.587 &  & 0.628 & \\
-4am-midnight & 0.680 & 0.766 & 0.448 & 0.613\\
+6am-12pm & 0.703 &  &  & 0.689 &  &  & 0.684 &  &  & 0.581 &  & \\
+12pm-6pm & 0.536 & 0.709 &  & 0.519 & 0.699 &  & 0.520 & 0.654 &  & 0.532 & 0.701 & \\
+6pm-12am & 0.716 & 0.679 & 0.773 & 0.677 & 0.678 & 0.723 & 0.677 & 0.676 & 0.708 & 0.514 & 0.565 & 0.626\\
 \bottomrule
@@
 \end{table}
+\end{landscape}
 
-\hypertarget{figure-2}{%
-\subsection{Figure}\label{figure-2}}
+\hypertarget{figure-1}{%
+\subsection{Figure}\label{figure-1}}
 
-\includegraphics{manuscript_files/figure-latex/pred_nhanes-1.pdf}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{rand\_id }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\FunctionTok{unique}\NormalTok{(pred\_nhanes\_adglmm}\SpecialCharTok{$}\NormalTok{id), }\DecValTok{4}\NormalTok{)}
+\NormalTok{cols }\OtherTok{\textless{}{-}}\NormalTok{ cols[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{3}\NormalTok{]}
+\FunctionTok{names}\NormalTok{(cols) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"6am"}\NormalTok{, }\StringTok{"12pm"}\NormalTok{, }\StringTok{"6pm"}\NormalTok{)}
+\FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{),}
+\NormalTok{  pred\_nhanes\_adglmm }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(pred360, pred720, pred1080), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred360, }\AttributeTok{col =} \StringTok{"6am"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred720, }\AttributeTok{col =} \StringTok{"12pm"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred1080, }\AttributeTok{col =} \StringTok{"6pm"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"Time"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1440}\NormalTok{, }\AttributeTok{by =} \DecValTok{360}\NormalTok{),}
+                     \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"0am"}\NormalTok{, }\StringTok{"6am"}\NormalTok{, }\StringTok{"12pm"}\NormalTok{, }\StringTok{"6pm"}\NormalTok{, }\StringTok{"12am"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{, }\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{size =} \DecValTok{7}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols)}
+\end{Highlighting}
+\end{Shaded}
 
-\includegraphics{manuscript_files/figure-latex/small_exp-1.pdf}
+\includegraphics{manuscript_files/figure-latex/fig_appl-1.pdf}
 
@@
 
-\hypertarget{fgfpca-algorithm-demonstration}{%
-\subsection{fGFPCA algorithm
-demonstration}\label{fgfpca-algorithm-demonstration}}
-
-\includegraphics{manuscript_files/figure-latex/binning-1.pdf}
-
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-13-1.pdf}
-
 \end{document}

### 2024-02-06 Code/DataAppl/Data_Appl_GLMMadaptive.R

@@ -1,0 +1,102 @@
+# Reference method: GLMMadaptive 
+set.seed(1114)
+
+library(here)
+library(tidyverse)
+library(ggplot2)
+theme_set(theme_minimal())
+library(fcr)
+library(refund)
+library(lme4)
+library(ROCR)
+library(GLMMadaptive)
+library(mgcv)
+library(ggpubr)
+library(gridExtra)
+library(kableExtra)
+library(LaplacesDemon)
+library(splines)
+
+#### load data ####
+df <- read_rds(here("Data/nhanes_bi.rds"))
+df <- df %>% rename(id=SEQN, Y=Z)
+
+N <- length(unique(df$id)) # sample size 8763
+J <- max(df$sind) # 1440 measures for each subject
+t <- unique(df$sind)
+K <- 4 # number of eigenfunctions to use
+
+
+#### Data split ####
+
+# 60% (5257) subjects for training, 40% (3506) for out-of-sample prediction
+train_id <- sample(unique(df$id), size = N*0.6)
+test_id <- setdiff(unique(df$id), train_id)
+
+train_df <- df %>% filter(id %in% train_id)
+test_df <- df %>% filter(id %in% test_id)
+
+
+#### fit GLMMadaptvie ####
+# on training set
+# only able to fit a random intecept
+head(train_df)
+
+t1 <- Sys.time()
+adglmm_mod <- mixed_model(Y ~ t, random = ~ 1 | id, 
+                          data = train_df %>% mutate(t=sind/J),
+                          family = binomial())
+t2 <- Sys.time()
+t_est_adglmm <- t2-t1 # model fitting took 20.82 mins
+summary(adglmm_mod)
+
+# prediction
+head(test_df)
+test_df <- test_df %>% mutate(t=sind/J)
+
+
+df_pred1 <- predict(adglmm_mod, 
+                    newdata = test_df %>% filter(sind <= 360),
+                    newdata2 =  test_df %>% filter(sind > 360), 
+                    type = "subject_specific", type_pred = "link",
+                    se.fit = FALSE, return_newdata = TRUE)$newdata2
+head(df_pred1)
+df_pred1 <- df_pred1 %>% rename(pred360=pred)
+
+df_pred2 <- predict(adglmm_mod, 
+                    newdata = test_df %>% filter(sind <= 720),
+                    newdata2 =  test_df %>% filter(sind > 720), 
[…]

### 2024-02-06 Code/DataAppl/Data_Appl_NHANES.R

@@ -89,8 +89,8 @@ t2 <- Sys.time()
 t2-t1 
 
 data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC") %>%
+  rename(PC1 = X1, PC2=X2, PC3=X3, PC4=X4) %>%
+  pivot_longer(2:5, names_to = "PC") %>%
   ggplot()+
   geom_line(aes(x=t, y=value, col = PC))+
   labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
@@ -115,12 +115,12 @@ for(k in 1:K){
 
 # plot
 df_pc1 <- data.frame(t=1:J,  df_phi) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC") 
+  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
+  pivot_longer(2:5, names_to = "PC") 
 
 df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1=X1, PC2=X2, PC3=X3) %>%
-  pivot_longer(2:4, names_to = "PC") 
+  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
+  pivot_longer(2:5, names_to = "PC") 
 
 left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
   ggplot()+
@@ -135,52 +135,56 @@ train_df <- train_df %>%
   left_join(df_phi, by = "sind")
 train_df$id <- as.factor(train_df$id)
 
-t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(sind, bs="cc")+
-                     s(id, by=phi1, bs="re")+
-                     s(id, by=phi2, bs="re")+
-                     s(id, by=phi3, bs="re")+
-                     s(id, by=phi4, bs="re"), 
-                   family = binomial, 
-                   data=train_df, 
-                   method = "fREML",
-                   discrete = TRUE)
-t2 <- Sys.time()
-t2-t1 # 21 hours
-save(debias_glmm, file = here("Data/Appl_debias_model.RData"))
+# don't run this part unless it's absolutely necessary!
+# It takes 20 hours!
+# t1 <- Sys.time()
[…]

### 2024-02-07 Docs/NhanesApplExample.Rmd
try to simulate using estimated from debias model

@@
 
-I'd like to investigate with different sample sizes.
-
 ```{r, cache=TRUE, results='hide'}
 # sample size 
-N_sub <- rep(500, 4)
-fpca_model_list <- list()
-interp_pc_est <- list()
-debias_model_list <- list()
-time_vec <- rep(NA, length(N_sub))
-
-pb <- txtProgressBar(min=0, max = length(N_sub), style = 3)
-for(i in seq_along(N_sub)){
+Ntr <- 500
 
 # training sample of a certain size
-  Ntr <- N_sub[i]
 train_id <- sample(unique(df$id), size = Ntr) 
@@
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-  fpca_model_list[[i]] <- fpca_mod
 
@@
 }# project binned eigenfunctions onto the original grid
-  interp_pc_est[[i]] <- df_phi
 
@@
 
-  setTxtProgressBar(pb, i)
-}
-
-close(pb)
 ```
@@
 2. Is is possible to impose monotone constraint on smoothing parameter estimation?
+
+```{r, class.source='fold-show'}
+load(here("Data/Appl_debias_model.RData"))
+summary(debias_glmm)
+
+# re-evaluated mean
+new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean
+plot(1:J, new_mu)
+
+# re-evaluated eigenvalues
+new_lambda <- 1/debias_glmm$sp[2:5]
+new_lambda <- new_lambda/n_bin
+
+# re-evaluated PC functions
+df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>%
+  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})
+df_phi %>% pivot_longer(2:5) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=value, col = name))
+
+```
+
+```{r gen_data}
+# generate data with the estimates from thedebias step
+
+# generate score
+xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(new_lambda))
+bi <- xi %*% t(as.matrix(df_phi[, 2:5])) # random effect functions
+dim(bi)
+
+# latent gaussian function
+new_mu <- matrix(new_mu, nrow = 1)
+eta_i <- new_mu+bi
+dim(new_mu)
+
+  this_df <- data.frame(id = factor(rep(1:N, each=J)),
+                        t = rep(t, N), 
+                        eta_i = as.vector(t(eta_i)),
+                        sind = rep(1:J, N))
+
+  this_df$Y <- rbinom(N*J, size=1, prob=plogis(this_df$eta_i))
+
+  sim_data[[m]] <- this_df
+]
+```

### 2024-02-07 Docs/manuscript.Rmd
try to simulate using estimated from debias model

@@
   rename(pred360=pred_w1, pred720=pred_w2, pred1080=pred_w3)
-
 ```
@@
 ```{r fig_appl, fig.height=8, fig.width=8}
-rand_id <- sample(unique(pred_nhanes_adglmm$id), 4)
+rand_id 
 cols <- cols[1:3]
@@
 
+```{r eval=FALSE}
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+df_nhanes %>% 
+  select(SEQN, Z, sind) %>%
+  filter(SEQN %in% rand_id) %>%
+  filter(sind %in% 356:360) %>%
+  pivot_wider(names_from = sind, values_from = Z) %>% 
+  select(-SEQN) %>% distinct(.)
+
+pred_appl_gfosr_l5 %>% 
+  filter(id %in% rand_id) %>%
+  select(id, sind, pred360) %>%
+  pivot_wider(names_from = sind, values_from = pred360) %>% 
+  select(-id) %>% distinct(.)
+```
+
 <!-- ## fGFPCA algorithm demonstration -->

### 2024-02-13 Docs/NhanesApplExample.Rmd
reversed eigenvalue problem: simulation

--- 2024-01-31 Docs/NhanesApplExample.Rmd
@@
 
-# Fit fGFPCA model
-
-```{r data_subset}
+```{r bin_data}
 # head(df)
@@
 N <- length(unique(df$id))
+Ntr <- 500
 J <- length(unique(df$sind))
-```
 
-# Step 1: Bin data
+load(here("Data/Appl_debias_model.RData"))
 
-- Bin every 10 observations
-
-```{r, message=FALSE}
-# bin data
+# parameters used in binning
 bin_w <- 10 # bin width
@@
 mid <- (brks+bin_w/2)[1:n_bin] # mid points
-
-df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
-df$bin <- as.numeric(as.character(df$bin))
 ```
 
-# Step 2-4:  Local GLMM, fPCA, splines interpolation and GLMM reevaluation
+# Estimates
 
-I'd like to investigate with different sample sizes.
+- Eigenvalues
 
-```{r, cache=TRUE, results='hide'}
-# sample size 
-N_sub <- rep(500, 4)
-fpca_model_list <- list()
-interp_pc_est <- list()
-debias_model_list <- list()
-time_vec <- rep(NA, length(N_sub))
+```{r evals}
+evals <- 1/debias_glmm$sp[2:5]/n_bin
+```
 
-pb <- txtProgressBar(min=0, max = length(N_sub), style = 3)
-for(i in seq_along(N_sub)){
+- PC functions
 
-  # training sample of a certain size
-  Ntr <- N_sub[i]
-  train_id <- sample(unique(df$id), size = Ntr) 
-  train_df <- df %>% filter(id %in% train_id)
+```{r}
+df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>%
+  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})
 
-  t1=Sys.time()
+df_phi %>% pivot_longer(2:5) %>%
+  ggplot(aes(x=sind, y=value))+
+  geom_line()+
+  facet_wrap(~name)
+```
+
+- Mean function
+
+```{r}
+mu <- predict(debias_glmm, type = "terms")[1:J, 1]
+plot(1:J, mu, type = "l")
+```
+
+# Simulation using re-evaluated estimates
+
+```{r gen_data}
+# generate score
+xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
+xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
+  ggplot()+
+  geom_boxplot(aes(x=PC, y=score))+
+  geom_jitter(aes(x=PC, y=score), size = 0.5)
+
+# individual random effects
+bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
+
+# add in mean 
+eta <- apply(bi, 2, function(x){x+mu})
+gen_df <- data.frame(id = factor(rep(1:Ntr, each=J)),
+                     sind = rep(1:J, Ntr), 
+                     eta_i = as.vector(eta))
+
+# binary outcome
+gen_df$Y <- rbinom(Ntr*J, size=1, prob=plogis(gen_df$eta_i))
+
+gen_df %>% filter(id %in% 1:4) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
+  facet_wrap(~id)+
+  labs(title = "Generated data")
+```
+
+# fGFPCA on generated data
+
+```{r fGFPCA1}
+# bin 
+gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
+gen_df$bin <- as.numeric(as.character(gen_df$bin))
+
 # local GLMM
-  train_bin_lst <- split(train_df, f = train_df$bin)
+train_bin_lst <- split(gen_df, f = gen_df$bin)
 df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
@@
 
+df_est_latent %>% filter(id %in% 1:4) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=plogis(eta_i), col = "True"))+
+  geom_line(aes(x=sind, y=plogis(eta_hat), col = "Estimate"))+
+  facet_wrap(~id)+
+  labs(title = "Generated data")
+```
+
+```{r fGFPCA2}
 # FPCA
@@
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-  fpca_model_list[[i]] <- fpca_mod
+```
 
+```{r re_eval_PC}
 # interpolation
@@
 
-  df_phi <- matrix(NA, J, K) 
+est_phi <- matrix(NA, J, K) 
 for(k in 1:K){
   lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
-    df_phi[,k] <- Bnew %*% coef(lm_mod)
-  }# project binned eigenfunctions onto the original grid
-  interp_pc_est[[i]] <- df_phi
+  est_phi[,k] <- Bnew %*% coef(lm_mod)
+} 
+colnames(est_phi) <- paste0("phi", 1:K)
+```
 
+I would like to look at estimated PC functions:
+
+```{r}
+# PC functions after re-evaluation
+bind_rows(
+  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "re-evaluated"),
+  data.frame(df_phi, type = "True")
+) %>%
+  pivot_longer(2:5) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=value, col=type))+
+  facet_wrap(~name)
+```
+
+It looks like something went wrong in the estimation of PC3. The estimated line is either identical or off by a sign compared to the true PC functions. That said, PC2 seems a bit off too.
+
+What about the estimated PC functions from FPCA without re-evaluation?
+
+```{r}
+fpca_phi <- fpca_mod$efunctions*sqrt(n_bin)
+colnames(fpca_phi) <- paste0("phi", 1:K)
+bind_rows(
+  data.frame(df_phi, type = "True"),
+  data.frame(sind = mid, fpca_phi, type = "FPCA")
+) %>%
+  pivot_longer(2:5) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=value, col=type),size = 0.5)+
+  facet_wrap(~name)
+```
+
+Looks like FPCA did not capture the shape of PC2 and PC3.
+
+Next I'd like to re-evaluate the eigenvalues
+
+```{r}
 # debias model
-  df_phi <- data.frame(sind = 1:J, df_phi)
-  colnames(df_phi) <- c("sind", paste0("phi", 1:K))
-  train_df <- train_df %>% 
+gen_df <- gen_df %>% 
   left_join(df_phi, by = "sind")
-  train_df$id <- as.factor(train_df$id)
-  debias_glmm <- bam(Y ~ s(sind, bs="cr")+
+gen_df$id <- as.factor(gen_df$id)
+debias_glmm2 <- bam(Y ~ s(sind, bs="cr")+
                    s(id, by=phi1, bs="re")+
@@
                  family = binomial,
-                   data=train_df, 
+                 data=gen_df,
                  method = "fREML",
                  discrete = TRUE)
-  debias_model_list[[i]] <- debias_glmm
-  t2 <- Sys.time()
-
-  time_vec[i] <- as.numeric(t2-t1, unit = "secs")
-
-  setTxtProgressBar(pb, i)
-}
-
-close(pb)
 ```
 
-<!-- - Time spent on model fitting -->
+Eigenvalues before and after re-evaluation
 
-```{r, eval=FALSE}
-time_vec
+```{r, class.source='fold-show'}
+# fpca
+fpca_mod$evalues/n_bin
+
+# re-evaluated
+1/debias_glmm2$sp[2:5]
 ```
 
-<!-- - FPCA eigenvalues -->
+It looks like the flip happened again.
 
-```{r, eval=FALSE}
-lapply(fpca_model_list, function(x){x$evalues})
-lapply(fpca_model_list, function(x){x$pve})
-```
-
-- Eigenvalues from debias models
-
-```{r}
-lapply(debias_model_list, function(x){1/x$sp[2:5]})
-```
-
-- PC functions
-
-```{r}
-pc_plot_list <- list()
-
-for(i in  1:length(N_sub)){
-
-  df_phi1 <- data.frame(t=mid, fpca_model_list[[i]]$efunctions)
-  df_phi2 <- data.frame(t = 1:J, interp_pc_est[[i]])
-
-  this_plot <- left_join(df_phi2, df_phi1, by = "t") %>%
-    ggplot()+
-    geom_point(aes(x=t, y=X1.y, col = "PC1"), na.rm = T, size = 0.5)+
-    geom_point(aes(x=t, y=X2.y, col = "PC2"), na.rm = T, size = 0.5)+
-    geom_point(aes(x=t, y=X3.y, col = "PC3"), na.rm = T, size = 0.5)+
-    geom_point(aes(x=t, y=X4.y, col = "PC4"), na.rm = T, size = 0.5)+
-    geom_line(aes(x=t, y=X1.x, col = "PC1"))+
-    geom_line(aes(x=t, y=X2.x, col = "PC2"))+
-    geom_line(aes(x=t, y=X3.x, col = "PC3"))+
-    geom_line(aes(x=t, y=X4.x, col = "PC4"))+
-    labs(x="time", y = "", title = paste0("Iter", i))
-
-  pc_plot_list[[i]] <- this_plot
-
-}
-
-ggarrange(plotlist = pc_plot_list, common.legend = T)
-```
-
-The "reverse eigenvalue problem" does not always happen.
-When training sample size is small (like maybe 200), the problem happened less.
-So this is definately a numeric problem.
-But a very small sample also would induce sigularity issues.
-
-Some other thoughts:
-
-1. What about estimation method of smooth parameter? It seems GCV doesn't treat model as mixed effects. REML may be too slow without the "discrete=T" option on the full dataset. 
-2. Is is possible to impose monotone constraint on smoothing parameter estimation?
+At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self.
+For some reason, FPCA was not able to estimated PC2 and PC3 very well, and the GLMM model later on impose more penalty on the second random term.

### 2024-02-16 Docs/manuscript.Rmd
update figures

@@
 library(boot)
+library(RColorBrewer)
 
@@
 theme_set(theme_minimal())
-cols <- c("0.2"= "#FDCC8A", "0.4"="#FC8D59", "0.6"="#E34A33", "0.8"="#B30000", "True"="#000000")
-# brewer.pal(5, "OrRd")
-
-# define a color palette 
+cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
+names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")
+# display.brewer.all()
+# display.brewer.pal(4, "Set2")
+# brewer.pal(5, "Set2")
+
 ## prediction window 
@@
   geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm=T, linewidth = 1.0)+
   facet_grid(rows = vars(id), cols = vars(method))+
@@
   geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linewidth=1, na.rm=T)+
   facet_grid(rows = vars(id), cols = vars(method))+
@@
 ```{r fig_appl, fig.height=8, fig.width=8}
-rand_id 
+# redefine color scale
 cols <- cols[1:3] 
 names(cols) <- c("6am", "12pm", "6pm")
+
+# selct four subjects
+plot_id <- sort(c(68959, 68624, 64198, 80517))
+
+# figure
 bind_rows(
-  pred_nhanes_fgfpca %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
-  pred_appl_gfosr_l5 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)"),
-  pred_appl_gfosr_l1 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)"),
-  pred_nhanes_adglmm %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
+  pred_nhanes_fgfpca %>% filter(id %in% plot_id) %>% mutate(method="fGFPCA"),
+  pred_appl_gfosr_l5 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=5)"),
+  pred_appl_gfosr_l1 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=1)"),
+  pred_nhanes_adglmm %>% filter(id %in% plot_id) %>% mutate(method = "GLMMadaptive")) %>%
   mutate_at(vars(pred360, pred720, pred1080), 
@@
   geom_point(aes(x=sind, y=Y), size = 0.2)+
-  geom_line(aes(x=sind, y=pred360, col = "6am"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=sind, y=pred720, col = "12pm"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linetype="dashed", na.rm=T)+
+  geom_line(aes(x=sind, y=pred360, col = "6am"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=sind, y=pred720, col = "12pm"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linewidth=1, na.rm=T)+
   facet_grid(rows = vars(id), cols = vars(method))+
@@
   scale_color_manual(values = cols)
+```
+
+The fGFPCA prediction conditional on observation up to 6am looks identical across subjects.
+This is not due to coding issue.
+In fact, if we look at individual tracks, out of 3506 subjects there are 1473 unique predicted tracks.
+There is a high change the selected four subjects all have the same predicted tracks.
+
+PS number of unique predicted tracks is:
+- 3456 when conditional on 12 pm
+- 3488 when conditional on 6pm
+
+I think this is closedly determined by the variety of observed outcome across individuals.
+When conditioning on 6am, there are 1475 unique observed tracks; and 3465 and 3488 respectively when obesrved up to 12pm and 6pm.
+As we see, these numbers are very close to the number of unique prediction tracks in fGFPCA.
+
+I'd note that the four randomly selected subjects from the last report happen to have exactly the same observations from 0-6am.
+That is why they also have identical prediction after 6am.
+Therefore, in this report I arbitrarily assigned subjects to be displayed: 62161, .
+
+In addition: 
+- For GFOSR (L5), number of unique predicted tracks are 31, 32, 32
+- For GFOSR (L1), number of unique predicted tracks are 2, 2, 2
+- For GLMMadaptive, number of unique predicted tracks are 294, 712, 1533.
+
+```{r, eval=FALSE}
+length(unique(pred_nhanes_adglmm$id))
+pred_nhanes_fgfpca %>%
+  # filter(id %in% rand_id) %>%
+  select(id, sind, Y) %>%
+  filter(sind <=  360) %>%
+  pivot_wider(names_from = sind, values_from = Y, names_prefix = "sind") %>%
+  distinct(., pick(starts_with("sind")), .keep_all = T) %>%
+  select(id) %>%
+  t() %>%
+  sample(., size = 4)
+
+  # ggplot(aes(x=sind, y=pred360, col = as.factor(id), group = as.factor(id)))+
+  # geom_line()
+```
+
+```{r, eval=FALSE}
+pred_nhanes_fgfpca %>%
+  select(id, sind, pred360) %>%
+  filter(sind>360) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=pred360, col = as.factor(id)), show.legend = F)
+
+pred_nhanes_fgfpca %>%
+  select(id, sind, pred720) %>%
+  filter(sind>720) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=pred720, col = as.factor(id)), show.legend = F)
+
 ```

### 2024-02-16 Docs/manuscript.tex
update figures

@@
 \author{Ying Jin}
-\date{2024-02-06}
+\date{2024-02-16}
 
@@
   \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{), }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
   \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
@@
   \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{), }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
   \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
@@
 \begin{Highlighting}[]
-\NormalTok{rand\_id }\OtherTok{\textless{}{-}} \FunctionTok{sample}\NormalTok{(}\FunctionTok{unique}\NormalTok{(pred\_nhanes\_adglmm}\SpecialCharTok{$}\NormalTok{id), }\DecValTok{4}\NormalTok{)}
-\NormalTok{cols }\OtherTok{\textless{}{-}}\NormalTok{ cols[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{3}\NormalTok{]}
+\CommentTok{\# redefine color scale}
+\NormalTok{cols }\OtherTok{\textless{}{-}}\NormalTok{ cols[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{3}\NormalTok{] }
 \FunctionTok{names}\NormalTok{(cols) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"6am"}\NormalTok{, }\StringTok{"12pm"}\NormalTok{, }\StringTok{"6pm"}\NormalTok{)}
+
+\CommentTok{\# selct four subjects}
+\NormalTok{plot\_id }\OtherTok{\textless{}{-}} \FunctionTok{sort}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{68959}\NormalTok{, }\DecValTok{68624}\NormalTok{, }\DecValTok{64198}\NormalTok{, }\DecValTok{80517}\NormalTok{))}
+
+\CommentTok{\# figure}
 \FunctionTok{bind\_rows}\NormalTok{(}
-\NormalTok{  pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
-\NormalTok{  pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{),}
-\NormalTok{  pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{),}
-\NormalTok{  pred\_nhanes\_adglmm }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+\NormalTok{  pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{),}
+\NormalTok{  pred\_nhanes\_adglmm }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
   \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(pred360, pred720, pred1080), }
@@
   \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred360, }\AttributeTok{col =} \StringTok{"6am"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred720, }\AttributeTok{col =} \StringTok{"12pm"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
-  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred1080, }\AttributeTok{col =} \StringTok{"6pm"}\NormalTok{), }\AttributeTok{linetype=}\StringTok{"dashed"}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred360, }\AttributeTok{col =} \StringTok{"6am"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred720, }\AttributeTok{col =} \StringTok{"12pm"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred1080, }\AttributeTok{col =} \StringTok{"6pm"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
   \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
@@
 
+The fGFPCA prediction conditional on observation up to 6am looks
+identical across subjects. This is not due to coding issue. In fact, if
+we look at individual tracks, out of 3506 subjects there are 1473 unique
+predicted tracks. There is a high change the selected four subjects all
+have the same predicted tracks.
+
+PS number of unique predicted tracks is: - 3456 when conditional on 12
+pm - 3488 when conditional on 6pm
+
+I think this is closedly determined by the variety of observed outcome
+across individuals. When conditioning on 6am, there are 1475 unique
+observed tracks; and 3465 and 3488 respectively when obesrved up to 12pm
+and 6pm. As we see, these numbers are very close to the number of unique
+prediction tracks in fGFPCA.
+
+I'd note that the four randomly selected subjects from the last report
+happen to have exactly the same observations from 0-6am. That is why
+they also have identical prediction after 6am. Therefore, in this report
+I arbitrarily assigned subjects to be displayed: 62161, .
+
+In addition: - For GFOSR (L5), number of unique predicted tracks are 31,
+32, 32 - For GFOSR (L1), number of unique predicted tracks are 2, 2, 2 -
+For GLMMadaptive, number of unique predicted tracks are 294, 712, 1533.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{length}\NormalTok{(}\FunctionTok{unique}\NormalTok{(pred\_nhanes\_adglmm}\SpecialCharTok{$}\NormalTok{id))}
+\NormalTok{pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%}
+  \CommentTok{\# filter(id \%in\% rand\_id) \%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(id, sind, Y) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(sind }\SpecialCharTok{\textless{}=}  \DecValTok{360}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ sind, }\AttributeTok{values\_from =}\NormalTok{ Y, }\AttributeTok{names\_prefix =} \StringTok{"sind"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{distinct}\NormalTok{(., }\FunctionTok{pick}\NormalTok{(}\FunctionTok{starts\_with}\NormalTok{(}\StringTok{"sind"}\NormalTok{)), }\AttributeTok{.keep\_all =}\NormalTok{ T) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(id) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{t}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{sample}\NormalTok{(., }\AttributeTok{size =} \DecValTok{4}\NormalTok{)}
+
+  \CommentTok{\# ggplot(aes(x=sind, y=pred360, col = as.factor(id), group = as.factor(id)))+}
+  \CommentTok{\# geom\_line()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(id, sind, pred360) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(sind}\SpecialCharTok{\textgreater{}}\DecValTok{360}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred360, }\AttributeTok{col =} \FunctionTok{as.factor}\NormalTok{(id)), }\AttributeTok{show.legend =}\NormalTok{ F)}
+
+\NormalTok{pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(id, sind, pred720) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(sind}\SpecialCharTok{\textgreater{}}\DecValTok{720}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred720, }\AttributeTok{col =} \FunctionTok{as.factor}\NormalTok{(id)), }\AttributeTok{show.legend =}\NormalTok{ F)}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{additional-figures}{%
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{df\_nhanes }\OtherTok{\textless{}{-}} \FunctionTok{read\_rds}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/nhanes\_bi.rds"}\NormalTok{))}
+\NormalTok{df\_nhanes }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(SEQN, Z, sind) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(SEQN }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{filter}\NormalTok{(sind }\SpecialCharTok{\%in\%} \DecValTok{356}\SpecialCharTok{:}\DecValTok{360}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ sind, }\AttributeTok{values\_from =}\NormalTok{ Z) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{SEQN) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{distinct}\NormalTok{(.)}
+
+\NormalTok{pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{select}\NormalTok{(id, sind, pred360) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ sind, }\AttributeTok{values\_from =}\NormalTok{ pred360) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{distinct}\NormalTok{(.)}
+\end{Highlighting}
+\end{Shaded}
+
 \end{document}

### 2024-02-20 Docs/NhanesApplExample.Rmd
appendix figure

@@
 ```{r}
+theme_set(theme_minimal())
+cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
+names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")
+
 # PC functions after re-evaluation
 bind_rows(
-  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "re-evaluated"),
-  data.frame(df_phi, type = "True")
+  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "Simulation"),
+  data.frame(df_phi, type = "Generation")
 ) %>%
+  rename("PC1" = phi1, "PC2" = phi2, "PC3" = phi3, "PC4" = phi4) %>%
   pivot_longer(2:5) %>%
   ggplot()+
-  geom_line(aes(x=sind, y=value, col=type))+
-  facet_wrap(~name)
+  geom_line(aes(x=sind, y=value, col=type), linewidth=1)+
+  facet_wrap(~name)+
+  scale_color_brewer(palette = "Set2")+
+   scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  labs(x = "Time", y="", color = "")
 ```

### 2024-02-27 Docs/NhanesApplExample.Rmd
more exploration on reversed eigenvalue problem

@@
 
-set.seed(130)
+set.seed(227)
 
@@
 library(arsenal)
+library(RColorBrewer)
 theme_set(theme_minimal())
@@
 mu <- predict(debias_glmm, type = "terms")[1:J, 1]
-plot(1:J, mu, type = "l")
+# plot(1:J, mu, type = "l", main = "Population mean")
+plot(1:J, plogis(mu), type = "l", main = "Population mean on the probablity scale")
 ```
+
+- Deviation from each PC from the population mean
+
+Below are figures for $\hat{f}_0(t)+2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale
+
+```{r}
+df_pc_dev <- df_phi %>%
+  mutate(mu = mu) %>%
+  mutate(phi1_dev = mu + 2*sqrt(evals[1])*phi1,
+         phi2_dev = mu + 2*sqrt(evals[2])*phi2,
+         phi3_dev = mu + 2*sqrt(evals[3])*phi3,
+         phi4_dev = mu + 2*sqrt(evals[4])*phi4)
+
+df_pc_dev %>%
+  pivot_longer(7:10) %>%
+  mutate(value = plogis(value), mu = plogis(mu)) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=value))+
+  geom_line(aes(x=sind, y=mu), col = "red")+
+  facet_wrap(~name) 
+```
+
+It looks like subjects with high loadings on PC2 would be more active than population average during afternoon/night, and less active in the morning.
+PC3, on the other hand, seems to indicate more active at night and early morning and less active in midday.
+They both reflect more active at night time.
+
+- Individual score
+
+```{r}
+data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
+  pivot_longer(1:4) %>%
+  ggplot(aes(x=name, y=value))+
+  geom_boxplot()+
+  geom_jitter(size = 0.2, alpha = 0.2)
+```
+
+Compared to the other three PCs, the distribution of $\hat{\xi}_{i3}$ seems to be shifted lower (positively skewed?).
 
@@
 xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
-xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
-  ggplot()+
-  geom_boxplot(aes(x=PC, y=score))+
-  geom_jitter(aes(x=PC, y=score), size = 0.5)
+# xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
+#   ggplot()+
+#   geom_boxplot(aes(x=PC, y=score))+
+#   geom_jitter(aes(x=PC, y=score), size = 0.5)
 
@@
 
-Looks like FPCA did not capture the shape of PC2 and PC3.
+Next I'd like to re-evaluate the eigenvalues.
 
-Next I'd like to re-evaluate the eigenvalues
+PS from the mgcv [user manual](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf): *Note that discrete=TRUEmay result in re-ordering of variables in tensor product smooths for improved efficiency, and sp must be
+supplied in re-ordered order*.
+Does this also apply to random effects?
+Does it have anything to do with the reversed eigenvalue functions?
+(I don't believe this is the case though, based on the experiments.
+Basciall I fit the model on a subset of 100 subjects setting discrete = F, and the smoothing parameters still flipped).
 
@@
 
-Eigenvalues before and after re-evaluation
+- Eigenvalues before and after re-evaluation
 
@@
 
+```{r, eval=FALSE}
+new_df <- gen_df %>% filter(id == 1)
+est_phi <- predict(debias_glmm2, newdata = new_df, type = "terms")[, 2:5] %>% 
+  data.frame()
+est_phi %>%
+  mutate(sind=1:J) %>%
+  pivot_longer(1:4) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y= value))+
+  facet_wrap(~name)
+```
+
+```{r eval=FALSE}
+coef1 <- coef(debias_glmm2)[c("s(id):phi1.1", "s(id):phi2.1", "s(id):phi3.1", "s(id):phi4.1")]
+coef1 <- matrix(coef1, nrow = J, ncol = 4, byrow = T)
+(as.matrix(df_phi[, 2:5]) * coef1) %>% data.frame() %>%
+  mutate(sind=1:J) %>%
+  pivot_longer(1:4) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y= value))+
+  facet_wrap(~name)
+
+```
+
 At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self.
-For some reason, FPCA was not able to estimated PC2 and PC3 very well, and the GLMM model later on impose more penalty on the second random term.
+It looks like FPCA captures the shape of eigenvalues very well, but for some reason, mgcv::bam thinks it is necessary to penalize the second eigenfunctions more than the third function.
+Why is that?
+Could it be that PC2 is more wiggly?
+
+If we compare the eigenvalues used for data generation to the ones estimated by bam, we can see bam actually does a good job.
+But the order of eigenfunctions flipped, and that is because FPCA re-ordered the eigenfunctions based on eigenvalues.
+
+```{r, class.source='fold-show'}
+# generation
+evals
+# estimation
+1/debias_glmm2$sp[2:5]
+```

### 2024-02-28 Docs/NhanesApplExample.Rmd
more explorations on the reversed eigenvalue problem

@@
                  method = "fREML",
-                 discrete = TRUE)
+                 discrete = TRUE,
+                 use.chol = TRUE)
 ```
@@
 It looks like the flip happened again.
-
-```{r, eval=FALSE}
-new_df <- gen_df %>% filter(id == 1)
-est_phi <- predict(debias_glmm2, newdata = new_df, type = "terms")[, 2:5] %>% 
-  data.frame()
-est_phi %>%
-  mutate(sind=1:J) %>%
-  pivot_longer(1:4) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y= value))+
-  facet_wrap(~name)
-```
-
-```{r eval=FALSE}
-coef1 <- coef(debias_glmm2)[c("s(id):phi1.1", "s(id):phi2.1", "s(id):phi3.1", "s(id):phi4.1")]
-coef1 <- matrix(coef1, nrow = J, ncol = 4, byrow = T)
-(as.matrix(df_phi[, 2:5]) * coef1) %>% data.frame() %>%
-  mutate(sind=1:J) %>%
-  pivot_longer(1:4) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y= value))+
-  facet_wrap(~name)
-
-```
 
@@
 ```
+
+# Other random explorations
+
+- what if I fit the bam model in a subset of data (e.g. bin midpoints) ?
+
+```{r}
+try_glmm <- bam(Y ~ s(sind, bs="cr")+
+                   s(id, by=phi1, bs="re")+
+                   s(id, by=phi2, bs="re")+
+                   s(id, by=phi3, bs="re")+
+                   s(id, by=phi4, bs="re"),
+                 family = binomial,
+                 data=gen_df %>% filter(sind==bin),
+                 method = "fREML",
+                 discrete = TRUE,
+                 use.chol = TRUE)
+
+1/try_glmm$sp[2:5]
+
+```
+
+- What about sub-intervals of time (e.g. every three hours? )
+
+```{r}
+evals_by_hour <- gen_df %>% 
+  mutate(hour = cut(sind, breaks = seq(0, J, by = 180), labels = 1:8)) %>%
+  group_by(hour) %>%
+  # filter(hour %in% 1:3) %>%
+  group_map(~{
+    try_glmm <- bam(Y ~ s(sind, bs="cr")+
+                   s(id, by=phi1, bs="re")+
+                   s(id, by=phi2, bs="re")+
+                   s(id, by=phi3, bs="re")+
+                   s(id, by=phi4, bs="re"),
+                 family = binomial,
+                 data=.x,
+                 method = "fREML",
+                 discrete = TRUE)
+    1/try_glmm$sp[2:5]
+  })
+
+bind_rows(evals_by_hour) %>%
+  mutate(hour = seq(3, 24, by = 3)) %>% 
+  pivot_longer(1:4) %>% 
+  ggplot(aes(x=hour, y=value, group=name, col=name))+
+  geom_point()+
+  labs(x="Hour", y= "Evals")
+```

### 2024-02-29 Docs/NhanesApplExample.Rmd
update meeting notes

@@
 
-Below are figures for $\hat{f}_0(t)+2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale
+Below are figures for $\hat{f}_0(t) \pm 2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale
 
 ```{r}
-df_pc_dev <- df_phi %>%
-  mutate(mu = mu) %>%
-  mutate(phi1_dev = mu + 2*sqrt(evals[1])*phi1,
-         phi2_dev = mu + 2*sqrt(evals[2])*phi2,
-         phi3_dev = mu + 2*sqrt(evals[3])*phi3,
-         phi4_dev = mu + 2*sqrt(evals[4])*phi4)
+df_pc_dev <- bind_rows(
+  data.frame(sind = 1:J, mu=mu, pc = "PC1",
+             high = mu + 2*sqrt(evals[1])*df_phi$phi1,
+             low = mu - 2*sqrt(evals[1])*df_phi$phi1),
+  data.frame(sind = 1:J, mu=mu, pc = "PC2",
+             high = mu + 2*sqrt(evals[2])*df_phi$phi2,
+             low = mu - 2*sqrt(evals[2])*df_phi$phi2),
+  data.frame(sind = 1:J, mu=mu, pc = "PC3",
+             high = mu + 2*sqrt(evals[3])*df_phi$phi3,
+             low = mu - 2*sqrt(evals[3])*df_phi$phi3),
+  data.frame(sind = 1:J, mu=mu, pc = "PC4",
+             high = mu + 2*sqrt(evals[4])*df_phi$phi4,
+             low = mu - 2*sqrt(evals[4])*df_phi$phi4)
+)
 
+```
+
+```{r}
 df_pc_dev %>%
-  pivot_longer(7:10) %>%
-  mutate(value = plogis(value), mu = plogis(mu)) %>%
   ggplot()+
-  geom_line(aes(x=sind, y=value))+
-  geom_line(aes(x=sind, y=mu), col = "red")+
-  facet_wrap(~name) 
+  geom_ribbon(aes(x=sind, y=mu, ymin=low, ymax=high, alpha = 0.2))+
+  geom_line(aes(x=sind, y=mu))+
+  facet_wrap(~pc) 
 ```
@@
 They both reflect more active at night time.
+
+There are a lot of flips between upper and lower bound in PC 2-4. This is certainly possible since the value of eigenfunctions can be either positive or negative. But is this an abnormal observation?
 

### 2024-03-08 Docs/NhanesApplExample.Rmd
weekly update

--- 2024-02-27 Docs/NhanesApplExample.Rmd
@@
 ---
-title: "Problem of Step 4 (debias) of fGFPCA on NHANES data"
+title: "Reverse eigenvalue problem"
 author: "Ying Jin"
@@
 
-set.seed(227)
+set.seed(1114)
 
@@
 
-```{r}
+```{r load_data}
 # load data
 df <- read_rds(here("Data/nhanes_bi.rds"))
+df <- df %>% rename(id=SEQN, Y=Z)
 
-# code
+N <- length(unique(df$id))
+J <- length(unique(df$sind))
+K <- 4 # number of eigenfunctions to use
+```
+
+```{r load_code}
 source(here("Code/GLMM-FPCA.R")) 
-# use pred_latent function to estimate latent function 
-K <- 4
 source(here("Code/OutsampBayes.R"))
@@
 
-```{r bin_data}
-# head(df)
-df <- df %>% rename(id=SEQN, Y=Z)
-N <- length(unique(df$id))
-Ntr <- 500
-J <- length(unique(df$sind))
+It seems that the reversed eigenvalue problem has to do with 1) the choice of spline basis and 2) sample size.
 
-load(here("Data/Appl_debias_model.RData"))
+# Spline basis
 
-# parameters used in binning
+I'll take a subset of 500 subject and try different spline basis.
+
+```{r data_split}
+# or for sanity checks, use a subset of 1000 subjects
+Ntry <- 500
+train_id <- sample(unique(df$id), size = Ntry)
+train_df <- df %>% filter(id %in% train_id)
+```
+
+```{r set_up}
+# basis to try
+bs_vec <- c("cr", "cs", "tp", "bs", "ps")
+evals <- matrix(NA, length(bs_vec), K)
+```
+
+```{r fGFPCA}
+# Step 1: Bin every 10 observations
 bin_w <- 10 # bin width
@@
 mid <- (brks+bin_w/2)[1:n_bin] # mid points
-```
 
-# Estimates
+train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
+train_df$bin <- as.numeric(as.character(train_df$bin))
 
-- Eigenvalues
-
-```{r evals}
-evals <- 1/debias_glmm$sp[2:5]/n_bin
-```
-
-- PC functions
-
-```{r}
-df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>%
-  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})
-
-df_phi %>% pivot_longer(2:5) %>%
-  ggplot(aes(x=sind, y=value))+
-  geom_line()+
-  facet_wrap(~name)
-```
-
-- Mean function
-
-```{r}
-mu <- predict(debias_glmm, type = "terms")[1:J, 1]
-# plot(1:J, mu, type = "l", main = "Population mean")
-plot(1:J, plogis(mu), type = "l", main = "Population mean on the probablity scale")
-```
-
-- Deviation from each PC from the population mean
-
-Below are figures for $\hat{f}_0(t)+2\sqrt{\hat{\lambda}}_k\phi_l(k)$, but on the probablity scale
-
-```{r}
-df_pc_dev <- df_phi %>%
-  mutate(mu = mu) %>%
-  mutate(phi1_dev = mu + 2*sqrt(evals[1])*phi1,
-         phi2_dev = mu + 2*sqrt(evals[2])*phi2,
-         phi3_dev = mu + 2*sqrt(evals[3])*phi3,
-         phi4_dev = mu + 2*sqrt(evals[4])*phi4)
-
-df_pc_dev %>%
-  pivot_longer(7:10) %>%
-  mutate(value = plogis(value), mu = plogis(mu)) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=value))+
-  geom_line(aes(x=sind, y=mu), col = "red")+
-  facet_wrap(~name) 
-```
-
-It looks like subjects with high loadings on PC2 would be more active than population average during afternoon/night, and less active in the morning.
-PC3, on the other hand, seems to indicate more active at night and early morning and less active in midday.
-They both reflect more active at night time.
-
-- Individual score
-
-```{r}
-data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
-  pivot_longer(1:4) %>%
-  ggplot(aes(x=name, y=value))+
-  geom_boxplot()+
-  geom_jitter(size = 0.2, alpha = 0.2)
-```
-
-Compared to the other three PCs, the distribution of $\hat{\xi}_{i3}$ seems to be shifted lower (positively skewed?).
-
-# Simulation using re-evaluated estimates
-
-```{r gen_data}
-# generate score
-xi <- rmvnorm(Ntr, mean = rep(0, K), sigma = diag(evals))
-# xi %>% data.frame() %>% pivot_longer(1:4, names_to = "PC", values_to = "score") %>%
-#   ggplot()+
-#   geom_boxplot(aes(x=PC, y=score))+
-#   geom_jitter(aes(x=PC, y=score), size = 0.5)
-
-# individual random effects
-bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
-
-# add in mean 
-eta <- apply(bi, 2, function(x){x+mu})
-gen_df <- data.frame(id = factor(rep(1:Ntr, each=J)),
-                     sind = rep(1:J, Ntr), 
-                     eta_i = as.vector(eta))
-
-# binary outcome
-gen_df$Y <- rbinom(Ntr*J, size=1, prob=plogis(gen_df$eta_i))
-
-gen_df %>% filter(id %in% 1:4) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
-  facet_wrap(~id)+
-  labs(title = "Generated data")
-```
-
-# fGFPCA on generated data
-
-```{r fGFPCA1}
-# bin 
-gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
-gen_df$bin <- as.numeric(as.character(gen_df$bin))
-
-# local GLMM
-train_bin_lst <- split(gen_df, f = gen_df$bin)
+# Step 2:  Local GLMM
+train_bin_lst <- split(train_df, f = train_df$bin)
 df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
@@
 
-df_est_latent %>% filter(id %in% 1:4) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  geom_line(aes(x=sind, y=plogis(eta_i), col = "True"))+
-  geom_line(aes(x=sind, y=plogis(eta_hat), col = "Estimate"))+
-  facet_wrap(~id)+
-  labs(title = "Generated data")
-```
-
-```{r fGFPCA2}
 # FPCA
 uni_eta_hat <- df_est_latent %>% filter(bin==sind)
-mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
-fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
-```
+mat_est_unique <- matrix(uni_eta_hat$eta_hat,
+                         nrow=length(train_id), 
+                         ncol=n_bin, byrow = F) 
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
 
-```{r re_eval_PC}
-# interpolation
+# Re-evaluation
+## grid extension
 p <- 3 # order of b splines 
@@
 
+df_phi <- matrix(NA, J, K) 
+for(k in 1:K){
+  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+  df_phi[,k] <- Bnew %*% coef(lm_mod)
+}# project binned eigenfunctions onto the original grid
+
+## debias
+## let's try scale the eigenfunctions before re-evaluation
+df_phi <- data.frame(sind = 1:J, df_phi)
+colnames(df_phi) <- c("sind", paste0("phi", 1:K))
+# df_phi <- df_phi %>% mutate(
+#   phi1 = sqrt(n_bin)*phi1,
+#   phi2 = sqrt(n_bin)*phi2,
+#   phi3 = sqrt(n_bin)*phi3,
+#   phi4 = sqrt(n_bin)*phi4
+# )
+train_df <- train_df %>% 
+  left_join(df_phi, by = "sind")
+train_df$id <- as.factor(train_df$id)
+
+```
+
+```{r try_splines, results='hide'}
+pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)
+
+t1 <- Sys.time()
+for(i in seq_along(bs_vec)){
+
+this_glmm <- bam(Y ~ s(sind, bs = bs_vec[i])+
+                     s(id, by=phi1, bs="re")+
+                     s(id, by=phi2, bs="re")+
+                     s(id, by=phi3, bs="re")+
+                     s(id, by=phi4, bs="re"),
+                   family = binomial,
+                   data=train_df,
+                   method = "fREML",
+                   discrete = TRUE)
+evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
+setTxtProgressBar(pb, i)
+
+}
+
+close(pb)
+t2 <- Sys.time()
+# t2-t1 # about 5 minutes
+```
+
+```{r}
+data.frame(bs_vec, evals)
+```
+
+It seems the flip happened for both cubic spline basis sets, but not the other ones. Also, when sample size gets larger, the flip happens almost certainly for all spline basis.
+
+When I increase the number of PC from 4 to 5, the second and third eigenvalues still flipped almost all the time.
+
+```{r, eval=FALSE}
+bind_rows(
+data.frame(sind=1:J, mu = predict(this_glmm, type="terms")[1:J, 1]) %>% mutate(type = "BAM"),
+data.frame(sind=mid, mu = fpca_mod$mu) %>% mutate(type = "FPCA")
+) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=mu, col=type))+
+  labs(title="Population mean estimate")
+
+```
+
+# Simulation using estimates from NHANES data
+
+## Estimates use for simualtion
+
+For now I will use estimates from the model fit on the whole NHANES dataset, even though the spline basis is not proper (bs = "cc").
+
+```{r load_est}
+load(here("Data/Appl_debias_model.RData"))
+
+# eigenvalues
+true_evals <- 1/debias_glmm$sp[2:5]/n_bin
+
+# mean 
+true_mu <- predict(debias_glmm, type = "terms")[, 1]
+true_mu <- matrix(true_mu, nrow = J)[, 1]
+# plot(1:J, true_mu)
+
+# eigenfunctions
+df_phi <- debias_glmm$model %>% select(sind, starts_with("phi")) %>% distinct(.) %>% 
+  mutate_at(vars(starts_with("phi")), function(x){x*sqrt(n_bin)})
+```
+
+Below are some visualizations of these estimates
+
+```{r plot_pc_func}
+df_phi %>%
+  pivot_longer(starts_with("phi")) %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=value))+
+  facet_wrap(~name)+
+  labs(title = "Eigenfunctions")
+```
+
+```{r plot_pc_dev_from_mean}
+df_pc_dev <- bind_rows(
+  data.frame(sind = 1:J, mu=true_mu, pc = "PC1",
+             high = true_mu + 2*sqrt(true_evals[1])*df_phi$phi1,
+             low = true_mu - 2*sqrt(true_evals[1])*df_phi$phi1),
+  data.frame(sind = 1:J, mu=true_mu, pc = "PC2",
+             high = true_mu + 2*sqrt(true_evals[2])*df_phi$phi2,
+             low = true_mu - 2*sqrt(true_evals[2])*df_phi$phi2),
+  data.frame(sind = 1:J, mu=true_mu, pc = "PC3",
+             high = true_mu + 2*sqrt(true_evals[3])*df_phi$phi3,
+             low = true_mu - 2*sqrt(true_evals[3])*df_phi$phi3),
+  data.frame(sind = 1:J, mu=true_mu, pc = "PC4",
+             high = true_mu + 2*sqrt(true_evals[4])*df_phi$phi4,
+             low = true_mu - 2*sqrt(true_evals[4])*df_phi$phi4)
+)
+
+# plot
+df_pc_dev %>%
+  ggplot()+
+  geom_ribbon(aes(x=sind, y=mu, ymin=low, ymax=high, alpha = 0.2))+
+  geom_line(aes(x=sind, y=mu))+
+  facet_wrap(~pc)+
+  labs(title = "Deviation of each PC from population mean")
+```
+
+```{r plot_score}
+data.frame(matrix(debias_glmm$coefficients[-c(1:9)], ncol = 4)) %>%
+  pivot_longer(1:4) %>%
+  ggplot(aes(x=name, y=value))+
+  geom_boxplot()+
+  geom_jitter(size = 0.2, alpha = 0.2)+
+  stat_summary(fun=mean, geom="point", color="red")+
+  labs(title = "Individual scores")
+```
+
+## Data generation
+
+```{r gen_data_unbound}
+# interpolate FPCA mean
+# interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
+# true_mu <- interp_mu_fpca$y
+
+# increase sample size?
+Ntry <- 500
+# generate score
+xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))
+
+# individual random effects
+bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
+
+# add in unbounded mean
+eta <- apply(bi, 2, function(x){x+true_mu})
+gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
+                     sind = rep(1:J, Ntry),
+                     eta_i = as.vector(eta))
+
+# binary outcome
+gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
+```
+
+```{r plot_data}
+# comparing generated data to the actual data
+bind_rows(
+    gen_df %>% 
+      group_by(sind) %>% 
+      summarize(active = mean(Y)) %>%
+      mutate(type="Simulation"),
+
+    df %>% 
+      group_by(sind) %>% 
+      summarize(active = mean(Y)) %>%
+      mutate(type="NHANES")
+) %>%
+  ggplot()+
+      geom_line(aes(x=sind, y=active, col = type))+
+      labs(title="Proportion of active subjects by minute")
+
+```
+
+This is the first discrepancy I have found - It looks like the simulated data is more active and more variable than the real data.
+Doesn't that mean the algorithm overestimates the population activity and variability?
+Since the mean estimates will surely change in Step 4 (debias), would the overestimation be caused by that?
+
+## fGFPCA
+
+```{r exp_gen_data}
+# bin data
+# using the same bin length 
+gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
+gen_df$bin <- as.numeric(as.character(gen_df$bin))
+
+# Step 2:  Local GLMM
+gen_bin_lst <- split(gen_df, f = gen_df$bin)
+df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+df_est_latent <- bind_rows(df_est_latent)
+## Figure
+# df_est_latent %>% filter(id %in% 1:4) %>%
+#   ggplot()+
+#   geom_point(aes(x=sind, y=Y), size = 0.5)+
+#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
+#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
+#   facet_wrap(~id)
+
+# FPCA
+uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+mat_est_unique <- matrix(uni_eta_hat$eta_hat,
+                        nrow=Ntry, 
+                        ncol=n_bin, byrow = F) 
+# dim(mat_est_unique)
+fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
+```
+
+```{r}
+# Re-evaluation
 est_phi <- matrix(NA, J, K) 
+p <- 3 # order of b splines 
+knots <- 35 # number of knots (same from FPCA model)
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
+
+B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
+                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
+Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
+                   outer.ok = TRUE)$design
 for(k in 1:K){
@@
   est_phi[,k] <- Bnew %*% coef(lm_mod)
-} 
-colnames(est_phi) <- paste0("phi", 1:K)
-```
+}# project binned eigenfunctions onto the original grid
 
-I would like to look at estimated PC functions:
-
-```{r}
-theme_set(theme_minimal())
-cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
-names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")
-
-# PC functions after re-evaluation
-bind_rows(
-  data.frame(sind=1:J, est_phi*sqrt(n_bin), type = "Simulation"),
-  data.frame(df_phi, type = "Generation")
-) %>%
-  rename("PC1" = phi1, "PC2" = phi2, "PC3" = phi3, "PC4" = phi4) %>%
-  pivot_longer(2:5) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=value, col=type), linewidth=1)+
-  facet_wrap(~name)+
-  scale_color_brewer(palette = "Set2")+
-   scale_x_continuous(breaks = seq(0, 1440, by = 360),
-                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
-  labs(x = "Time", y="", color = "")
-```
-
-It looks like something went wrong in the estimation of PC3. The estimated line is either identical or off by a sign compared to the true PC functions. That said, PC2 seems a bit off too.
-
-What about the estimated PC functions from FPCA without re-evaluation?
-
-```{r}
-fpca_phi <- fpca_mod$efunctions*sqrt(n_bin)
-colnames(fpca_phi) <- paste0("phi", 1:K)
-bind_rows(
-  data.frame(df_phi, type = "True"),
-  data.frame(sind = mid, fpca_phi, type = "FPCA")
-) %>%
-  pivot_longer(2:5) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=value, col=type),size = 0.5)+
-  facet_wrap(~name)
-```
-
-Next I'd like to re-evaluate the eigenvalues.
-
-PS from the mgcv [user manual](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf): *Note that discrete=TRUEmay result in re-ordering of variables in tensor product smooths for improved efficiency, and sp must be
-supplied in re-ordered order*.
-Does this also apply to random effects?
-Does it have anything to do with the reversed eigenvalue functions?
-(I don't believe this is the case though, based on the experiments.
-Basciall I fit the model on a subset of 100 subjects setting discrete = F, and the smoothing parameters still flipped).
-
-```{r}
-# debias model
-gen_df <- gen_df %>% 
-  left_join(df_phi, by = "sind")
+## debias
+est_phi <- data.frame(sind = 1:J, est_phi)
+colnames(est_phi) <- c("sind", paste0("phi", 1:K))
+gen_df <- gen_df %>% left_join(est_phi, by = "sind")
 gen_df$id <- as.factor(gen_df$id)
-debias_glmm2 <- bam(Y ~ s(sind, bs="cr")+
+### model fit with B-splines basis
+t1 <- Sys.time()
+this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                    s(id, by=phi1, bs="re")+
@@
                  discrete = TRUE)
+t2 <- Sys.time()
 ```
 
-- Eigenvalues before and after re-evaluation
+Below I compared the estimated eigenvalues from simulated data to the true eigenvalues. As we see, the reverse did not happen on the simulated data.
 
-```{r, class.source='fold-show'}
-# fpca
-fpca_mod$evalues/n_bin
+```{r new_evals, class.source='fold-show'}
+# estimated eigenvalues
+1/this_glmm$sp[2:5]/n_bin
 
-# re-evaluated
-1/debias_glmm2$sp[2:5]
+# true eigenvalues
+true_evals
 ```
+Below is the comparision between true and esimtated eigenfunctions.
 
-It looks like the flip happened again.
-
-```{r, eval=FALSE}
-new_df <- gen_df %>% filter(id == 1)
-est_phi <- predict(debias_glmm2, newdata = new_df, type = "terms")[, 2:5] %>% 
-  data.frame()
-est_phi %>%
-  mutate(sind=1:J) %>%
-  pivot_longer(1:4) %>%
+```{r new_PC}
+bind_rows(df_phi %>% mutate(type = "True"),
+          est_phi %>% mutate(type = "Estimate",
+                             phi1 = phi1*sqrt(n_bin),
+                             phi2 = phi2*sqrt(n_bin),
+                             phi3 = phi3*sqrt(n_bin),
+                             phi4 = phi4*sqrt(n_bin))) %>%
+  pivot_longer(2:5) %>%
   ggplot()+
-  geom_line(aes(x=sind, y= value))+
+  geom_line(aes(x=sind, y=value, col = type))+
   facet_wrap(~name)
@@
 
-```{r eval=FALSE}
-coef1 <- coef(debias_glmm2)[c("s(id):phi1.1", "s(id):phi2.1", "s(id):phi3.1", "s(id):phi4.1")]
-coef1 <- matrix(coef1, nrow = J, ncol = 4, byrow = T)
-(as.matrix(df_phi[, 2:5]) * coef1) %>% data.frame() %>%
-  mutate(sind=1:J) %>%
-  pivot_longer(1:4) %>%
+And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4.
+The re-evaluation seem to drive the mean function closer to the truth.
+But in that case, why would generated data much more active compared to the NHANES data?
+
+```{r compare_mean}
+bind_rows(
+  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
+  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]) %>% mutate(type = "BAM"),
+  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
+) %>%
   ggplot()+
-  geom_line(aes(x=sind, y= value))+
-  facet_wrap(~name)
+      geom_line(aes(x=sind, y=mu, col = type))+
+      labs(title="Mean estimates")
 
@@
 
-At this point, I think we could say the reversed eigenvalue problem is an artifact of the dataset it self.
-It looks like FPCA captures the shape of eigenvalues very well, but for some reason, mgcv::bam thinks it is necessary to penalize the second eigenfunctions more than the third function.
-Why is that?
-Could it be that PC2 is more wiggly?
+# Simulation with the FPCA mean
 
-If we compare the eigenvalues used for data generation to the ones estimated by bam, we can see bam actually does a good job.
-But the order of eigenfunctions flipped, and that is because FPCA re-ordered the eigenfunctions based on eigenvalues.
+I would like to experiments if I could simulated data whose activity level and variation is closer to NHANES, using the interpolated FPCA mean and un-evaluated eigenvalues.
 
-```{r, class.source='fold-show'}
-# generation
-evals
-# estimation
-1/debias_glmm2$sp[2:5]
+```{r load_fpca_mu}
+# load FPCA model from NHANES
+load(here("Data/Appl_fpca_model.RData"))
+
+# compare means on NHANES data from FPCA and debias model
+bind_rows(
+  data.frame(sind=1:J, mu = true_mu, type = "BAM"),
+  data.frame(sind=mid, mu = fpca_mod$mu, type = "FPCA")
+) %>% 
+  ggplot()+
+  geom_line(aes(x=sind, y=mu, col = type))+
+  labs(title = "Mean estimates on the NHANES dataset")
+
 ```
+
+## Data generation
+
+```{r gen_data_fpca_mu}
+# interpolate FPCA mean
+interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
+true_mu <- interp_mu_fpca$y
+
+# FPCA eigenvalues
+true_evals <- fpca_mod$evalues/n_bin
+
+# maybe increase sample size?
+# Ntry <- 1000
+
+# generate score
+xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))
+
+# individual random effects
+bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
+
+# add in un-evaluated mean
+eta <- apply(bi, 2, function(x){x+true_mu})
+gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
+                     sind = rep(1:J, Ntry),
+                     eta_i = as.vector(eta))
+
+# binary outcome
+gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
+```
+
+```{r plot_data_fpca_mu}
+# comparing generated data to the actual data
+bind_rows(
+    gen_df %>% 
+      group_by(sind) %>% 
+      summarize(active = mean(Y)) %>%
+      mutate(type="Simulation"),
+
+    df %>% 
+      group_by(sind) %>% 
+      summarize(active = mean(Y)) %>%
+      mutate(type="NHANES")
+) %>%
+  ggplot()+
+      geom_line(aes(x=sind, y=active, col = type))+
+      labs(title="Proportion of active subjects by minute")
+
+```
+
+Looks like by using the FPCA mean, I was able to lower the activity level of generated data down to the level of NHANES data.
+
+## fGFPCA
+
+```{r fgfpca_fpca_mean}
+# bin data
+# using the same bin length 
+gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
+gen_df$bin <- as.numeric(as.character(gen_df$bin))
+
+# Step 2:  Local GLMM
+gen_bin_lst <- split(gen_df, f = gen_df$bin)
+df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+df_est_latent <- bind_rows(df_est_latent)
+## Figure
+# df_est_latent %>% filter(id %in% 1:4) %>%
+#   ggplot()+
+#   geom_point(aes(x=sind, y=Y), size = 0.5)+
+#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
+#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
+#   facet_wrap(~id)
+
+# FPCA
+uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+mat_est_unique <- matrix(uni_eta_hat$eta_hat,
+                        nrow=Ntry, 
+                        ncol=n_bin, byrow = F) 
+# dim(mat_est_unique)
+fpca_mod_sim <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
+```
+
+```{r}
+# Re-evaluation
+est_phi <- matrix(NA, J, K) 
+p <- 3 # order of b splines 
+knots <- 35 # number of knots (same from FPCA model)
+knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
+
+B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
+                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
+
+Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
+                   outer.ok = TRUE)$design
+for(k in 1:K){
+  lm_mod <- lm(fpca_mod_sim$efunctions[,k] ~ B-1)
+  est_phi[,k] <- Bnew %*% coef(lm_mod)
+}# project binned eigenfunctions onto the original grid
+
+## debias
+est_phi <- data.frame(sind = 1:J, est_phi)
+colnames(est_phi) <- c("sind", paste0("phi", 1:K))
+gen_df <- gen_df %>% left_join(est_phi, by = "sind")
+gen_df$id <- as.factor(gen_df$id)
+### model fit with B-splines basis
+t1 <- Sys.time()
+this_glmm <- bam(Y ~ s(sind, bs = "cr")+
+                   s(id, by=phi1, bs="re")+
+                   s(id, by=phi2, bs="re")+
+                   s(id, by=phi3, bs="re")+
+                   s(id, by=phi4, bs="re"),
+                 family = binomial,
+                 data=gen_df,
+                 method = "fREML",
+                 discrete = TRUE)
+t2 <- Sys.time()
+```
+
+Estimated eigenvalues
+
+```{r}
+1/this_glmm$sp[2:5]/sqrt(n_bin)
+```
+
+Using the FPCA estimates, I was able to lower the activity but not variation.
+I was also not able to recreate the reversed eigenvalue problem.
+Also, singularity issue happened in the binnin step when using the FPCA estimates to generate data.

### 2024-03-08 Docs/NhanesApplExample.Rmd
try NHANES data but truncated to only daytime

@@
 
-set.seed(1114)
+set.seed(308)
 
@@
 
-I'll take a subset of 500 subject and try different spline basis.
+I'll take a subset of 500 subject and try different spline basis. In addition, I truncated the dataset to keep records during the daytime (8am-6pm).
 
 ```{r data_split}
-# or for sanity checks, use a subset of 1000 subjects
+# for sanity checks, use a subset of 500 subjects
 Ntry <- 500
@@
 train_df <- df %>% filter(id %in% train_id)
+
+# only the day time? 
+# 8am - 6pm (sind: 480 - 1080)
+train_df <- train_df %>% filter(sind > 480 & sind <= 1080)
+train_df$sind <- train_df$sind-480
+J <- max(train_df$sind)
 ```
@@
 
-It seems the flip happened for both cubic spline basis sets, but not the other ones. Also, when sample size gets larger, the flip happens almost certainly for all spline basis.
+It seems for the entire day, the flip happened for both cubic spline basis sets, but not the other ones.
+Also, when sample size gets larger, the flip happens almost certainly for all spline basis.
+Bin width did not make a difference.
+Nor did increasing number of PCs from 4 to 5.
 
-When I increase the number of PC from 4 to 5, the second and third eigenvalues still flipped almost all the time.
+But if I restrict the time period down to 8am-6pm, the eigenvalues are not reversed, and it says that way with larger sample size!
 
@@
 
+```{r}
+df_est_latent %>% 
+  filter(id %in% sample(train_id, 4)) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=plogis(eta_hat), col = "Local GLMM"))+
+  facet_wrap(~id)
+```
+
 # Simulation using estimates from NHANES data
@@
 load(here("Data/Appl_debias_model.RData"))
+J <- length(unique(df$sind))
+n_bin <- J/bin_w # number of bins
+brks <- seq(0, J, by = bin_w) # cutoff points
+mid <- (brks+bin_w/2)[1:n_bin] # mid points
 

### 2024-03-08 Code/DataAppl/Data_Appl_NHANES.R

@@ -57,6 +57,11 @@ mid <- (brks+bin_w/2)[1:n_bin] # mid points
 train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
 train_df$bin <- as.numeric(as.character(train_df$bin))
 
+## checks
+head(train_df)
+table(train_df$bin)
+
+
 # Step 2:  Local GLMM
 train_bin_lst <- split(train_df, f = train_df$bin)
 
@@ -90,16 +95,19 @@ t1 <- Sys.time()
 fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
 t2 <- Sys.time()
 t2-t1 
-save(fpca_mod, file = here("Data/Appl_fpca_model.RData"))
 
-
-data.frame(t = mid, fpca_mod$efunctions) %>%
-  rename(PC1 = X1, PC2=X2, PC3=X3, PC4=X4) %>%
-  pivot_longer(2:5, names_to = "PC") %>%
+## checks
+fpca_mod$evalues
+data.frame(sind = mid, 
+           fpca_mod$efunctions) %>%
+  pivot_longer(2:5) %>%
   ggplot()+
-  geom_line(aes(x=t, y=value, col = PC))+
-  facet_wrap(~PC)+
-  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
+  geom_line(aes(x=sind, y=value))+
+  facet_wrap(~name)
+
+plot(mid, fpca_mod$mu, type = "l")
+
+save(fpca_mod, file = here("Data/Appl_fpca_model.RData"))
 
 # Re-evaluation
 ## grid extension
@@ -146,17 +154,19 @@ df_phi <- df_phi %>% mutate(
   phi4 = sqrt(n_bin)*phi4
 )
 train_df <- train_df %>% 
-  select(!starts_with("phi")) %>%
+  # select(!starts_with("phi")) %>%
   left_join(df_phi, by = "sind")
 train_df$id <- as.factor(train_df$id)
 
 
-
+head(df_phi)
 
 # don't run this part unless it's absolutely necessary!
 # It takes 20 hours!
+head(train_df)
+
 t1 <- Sys.time()
-debias_glmm <- bam(Y ~ s(sind)+
+debias_glmm <- bam(Y ~ s(sind, bs = "bs")+
                      s(id, by=phi1, bs="re")+
[…]

### 2024-03-19 Docs/NhanesApplExample.Rmd
add intercept

@@
 
-# Spline basis
+# Truncated track
 
@@
 J <- max(train_df$sind)
-```
-
-```{r set_up}
-# basis to try
-bs_vec <- c("cr", "cs", "tp", "bs", "ps")
-evals <- matrix(NA, length(bs_vec), K)
 ```
@@
 ```{r try_splines, results='hide'}
-pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)
+# pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)
+# 
+# t1 <- Sys.time()
+# for(i in seq_along(bs_vec)){
 
-t1 <- Sys.time()
-for(i in seq_along(bs_vec)){
-
-this_glmm <- bam(Y ~ s(sind, bs = bs_vec[i])+
+this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                      s(id, by=phi1, bs="re")+
@@
                    discrete = TRUE)
-evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
-setTxtProgressBar(pb, i)
-
-}
-
-close(pb)
-t2 <- Sys.time()
+# evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
+# setTxtProgressBar(pb, i)
+# 
+# }
+# 
+# close(pb)
+# t2 <- Sys.time()
 # t2-t1 # about 5 minutes
@@
 
-```{r}
-data.frame(bs_vec, evals)
+```{r, class.source='fold-show'}
+# re-evaluated eigenvalues
+1/this_glmm$sp[2:5]/n_bin
 ```
@@
 # mean 
-true_mu <- predict(debias_glmm, type = "terms")[, 1]
+# because of the parametrization of mgcv
+# we need to add the constant intercept back
+true_mu <- predict(debias_glmm, type = "terms")[, 1]+coef(debias_glmm)[1]
 true_mu <- matrix(true_mu, nrow = J)[, 1]
@@
 
-This is the first discrepancy I have found - It looks like the simulated data is more active and more variable than the real data.
-Doesn't that mean the algorithm overestimates the population activity and variability?
-Since the mean estimates will surely change in Step 4 (debias), would the overestimation be caused by that?
+After adding the intercept back, the generated data is still slightly more active and variable than the true data. Is it enough to make a difference?
 
@@
 ## Figure
-# df_est_latent %>% filter(id %in% 1:4) %>%
-#   ggplot()+
-#   geom_point(aes(x=sind, y=Y), size = 0.5)+
-#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
-#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
-#   facet_wrap(~id)
+df_est_latent %>% filter(id %in% 1:4) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.5)+
+  geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
+  geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
+  facet_wrap(~id)
 
@@
 gen_df$id <- as.factor(gen_df$id)
+
 ### model fit with B-splines basis
@@
   data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
-  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]) %>% mutate(type = "BAM"),
+  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"),
   data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
@@
       labs(title="Mean estimates")
-
 ```
-
-# Simulation with the FPCA mean
-
-I would like to experiments if I could simulated data whose activity level and variation is closer to NHANES, using the interpolated FPCA mean and un-evaluated eigenvalues.
-
-```{r load_fpca_mu}
-# load FPCA model from NHANES
-load(here("Data/Appl_fpca_model.RData"))
-
-# compare means on NHANES data from FPCA and debias model
-bind_rows(
-  data.frame(sind=1:J, mu = true_mu, type = "BAM"),
-  data.frame(sind=mid, mu = fpca_mod$mu, type = "FPCA")
-) %>% 
-  ggplot()+
-  geom_line(aes(x=sind, y=mu, col = type))+
-  labs(title = "Mean estimates on the NHANES dataset")
-
-```
-
-## Data generation
-
-```{r gen_data_fpca_mu}
-# interpolate FPCA mean
-interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
-true_mu <- interp_mu_fpca$y
-
-# FPCA eigenvalues
-true_evals <- fpca_mod$evalues/n_bin
-
-# maybe increase sample size?
-# Ntry <- 1000
-
-# generate score
-xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))
-
-# individual random effects
-bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
-
-# add in un-evaluated mean
-eta <- apply(bi, 2, function(x){x+true_mu})
-gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
-                     sind = rep(1:J, Ntry),
-                     eta_i = as.vector(eta))
-
-# binary outcome
-gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
-```
-
-```{r plot_data_fpca_mu}
-# comparing generated data to the actual data
-bind_rows(
-    gen_df %>% 
-      group_by(sind) %>% 
-      summarize(active = mean(Y)) %>%
-      mutate(type="Simulation"),
-
-    df %>% 
-      group_by(sind) %>% 
-      summarize(active = mean(Y)) %>%
-      mutate(type="NHANES")
-) %>%
-  ggplot()+
-      geom_line(aes(x=sind, y=active, col = type))+
-      labs(title="Proportion of active subjects by minute")
-
-```
-
-Looks like by using the FPCA mean, I was able to lower the activity level of generated data down to the level of NHANES data.
-
-## fGFPCA
-
-```{r fgfpca_fpca_mean}
-# bin data
-# using the same bin length 
-gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
-gen_df$bin <- as.numeric(as.character(gen_df$bin))
-
-# Step 2:  Local GLMM
-gen_bin_lst <- split(gen_df, f = gen_df$bin)
-df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
-df_est_latent <- bind_rows(df_est_latent)
-## Figure
-# df_est_latent %>% filter(id %in% 1:4) %>%
-#   ggplot()+
-#   geom_point(aes(x=sind, y=Y), size = 0.5)+
-#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
-#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
-#   facet_wrap(~id)
-
-# FPCA
-uni_eta_hat <- df_est_latent %>% filter(bin==sind)
-mat_est_unique <- matrix(uni_eta_hat$eta_hat,
-                        nrow=Ntry, 
-                        ncol=n_bin, byrow = F) 
-# dim(mat_est_unique)
-fpca_mod_sim <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
-```
-
-```{r}
-# Re-evaluation
-est_phi <- matrix(NA, J, K) 
-p <- 3 # order of b splines 
-knots <- 35 # number of knots (same from FPCA model)
-knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
-knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
-
-B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
-                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
-
-Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
-                   outer.ok = TRUE)$design
-for(k in 1:K){
-  lm_mod <- lm(fpca_mod_sim$efunctions[,k] ~ B-1)
-  est_phi[,k] <- Bnew %*% coef(lm_mod)
-}# project binned eigenfunctions onto the original grid
-
-## debias
-est_phi <- data.frame(sind = 1:J, est_phi)
-colnames(est_phi) <- c("sind", paste0("phi", 1:K))
-gen_df <- gen_df %>% left_join(est_phi, by = "sind")
-gen_df$id <- as.factor(gen_df$id)
-### model fit with B-splines basis
-t1 <- Sys.time()
-this_glmm <- bam(Y ~ s(sind, bs = "cr")+
-                   s(id, by=phi1, bs="re")+
-                   s(id, by=phi2, bs="re")+
-                   s(id, by=phi3, bs="re")+
-                   s(id, by=phi4, bs="re"),
-                 family = binomial,
-                 data=gen_df,
-                 method = "fREML",
-                 discrete = TRUE)
-t2 <- Sys.time()
-```
-
-Estimated eigenvalues
-
-```{r}
-1/this_glmm$sp[2:5]/sqrt(n_bin)
-```
-
-Using the FPCA estimates, I was able to lower the activity but not variation.
-I was also not able to recreate the reversed eigenvalue problem.
-Also, singularity issue happened in the binnin step when using the FPCA estimates to generate data.

### 2024-03-19 Code/DataAppl/Data_Appl_NHANES.R

@@ -41,9 +41,15 @@ train_df <- df %>% filter(id %in% train_id)
 test_df <- df %>% filter(id %in% test_id)
 
 # or for sanity checks, use a subset of 1000 subjects
-Ntry <- 1000
+Ntry <- 1500
 train_id <- sample(unique(df$id), size = Ntry)
 train_df <- df %>% filter(id %in% train_id)
+
+## truncate to 8am-6pm (480-1080)?
+train_df <- train_df %>% filter(sind > 480 & sind <= 1080) %>%
+  mutate(sind=sind-480)
+head(train_df)
+J <- max(train_df$sind)
 
 ##### fGFPCA #####
 
@@ -56,6 +62,8 @@ mid <- (brks+bin_w/2)[1:n_bin] # mid points
 
 train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
 train_df$bin <- as.numeric(as.character(train_df$bin))
+
+
 
 ## checks
 head(train_df)

### 2024-03-19 Code/Simulation_N100/Sim_Subset_fGFPCA.R

@@ -70,6 +70,8 @@ pred_list_all <- list()
 converge_state_list <- list()
 fit_time <- pred_time <- rep(NA, M)
 
+M <- 10
+
 
 #### fGFPCA
 
@@ -117,7 +119,7 @@ for(m in 1:M){
                      family = binomial, data=df_train_m, 
                      method = "fREML",
                      discrete = TRUE)
-  new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean
+  new_mu <- predict(debias_glmm, type = "terms")[1:J, 1]+coef(debias_glmm)[1] # extract re-evaluated mean
   new_lambda <- 1/debias_glmm$sp[2:5] # extract re-evaluated lambda
   # rescale
   new_phi <- df_phi %>% select(starts_with("phi"))*sqrt(n_bin)
@@ -203,63 +205,29 @@ save(fit_time_subset_fGFPCA, pred_time_subset_fGFPCA, pred_subset_fGFPCA,
 
 
 #### ISE ####
-window <- seq(0, 1, by = 0.2)
-
-# ise_mat <- array(NA, dim = c(length(window)-2, length(window)-2, M))
-ise_mat <- df_pred %>%
-  mutate(err1 = (pred0.2-eta_i)^2,
-         err2 = (pred0.4-eta_i)^2,
-         err3 = (pred0.6-eta_i)^2,
-         err4 = (pred0.8-eta_i)^2) %>% 
-  select(id, t, starts_with("err")) %>% 
-  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-  group_by(window, id) %>% 
-  summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-  group_by(window) %>% 
-  summarize_at(vars(err1, err2, err3, err4), mean) %>%
-  filter(window != "[0,0.2]") %>% 
-  select(starts_with("err"))
-# ise_mat[, ,m] <- as.matrix(ise_tb)
+ise_fgfpca2 <- array(NA, dim = c(length(window)-2, length(window)-2, 10))
 
 
-# dims: prediction window, max obs time, simulation iter
-mean_ise <- apply(ise_mat, c(1, 2), mean)
-mean_ise <- data.frame(mean_ise) %>% 
-  mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
[…]

### 2024-03-19 Code/Simulation_N500/Sim_fGFPCA.R

@@ -127,7 +127,7 @@ for(m in 1:M){
                      family = binomial, data=train_df, 
                      method = "fREML",
                      discrete = TRUE)
-  new_mu <- predict(debias_glmm, type = "terms")[1:J, 1] # extract re-evaluated mean
+  new_mu <- predict(debias_glmm, type = "terms")[1:J, 1]+coef(debias_glmm)[1] # extract re-evaluated mean
   new_lambda <- 1/debias_glmm$sp[2:5] # extract re-evaluated lambda
   # rescale
   new_phi <- df_phi %>% select(starts_with("phi"))*sqrt(n_bin)
@@ -214,7 +214,7 @@ mean(sapply(converge_state_list, mean)) # all dataset converged
 rand_id <- sample(test_id, 4)
 
 # prediction results
-pred_list_all[[476]] %>% 
+pred_list_all[[4]] %>% 
   filter(id %in% rand_id) %>% 
   mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
@@ -271,7 +271,7 @@ for(m in 1:M){
 }
 
 mean_ise <- apply(ise_mat, c(1, 2), mean)
-mean_ise <- data.frame(mean_ies) %>% 
+mean_ise <- data.frame(mean_ise) %>% 
   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
          .before = 1)
 colnames(mean_ise) <- c("Window", "0.2", "0.4", "0.6", "0.8")

### 2024-03-26 Docs/NhanesApplExample.Rmd
repeated simulation use NHANES estimates

@@
 
-```{r fGFPCA}
+```{r}
 # Step 1: Bin every 10 observations
@@
 # we need to add the constant intercept back
-true_mu <- predict(debias_glmm, type = "terms")[, 1]+coef(debias_glmm)[1]
-true_mu <- matrix(true_mu, nrow = J)[, 1]
+true_mu <- predict(debias_glmm, type = "terms")[1:J, 1]+coef(debias_glmm)[1]
 # plot(1:J, true_mu)
@@
 
-## Data generation
+## Simulation
 
-```{r gen_data_unbound}
-# interpolate FPCA mean
-# interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
-# true_mu <- interp_mu_fpca$y
+Let's repeat the simulation a few times:
 
-# increase sample size?
-Ntry <- 500
+- With different sample sizes
+- With a lower-bounded mean
+
+### Sample size
+
+```{r gen_data_N}
+# increase sample size
+Ntry_vec <- c(500, 1000, 1500, 2000)
+data_list <- list()
+
+for(n in seq_along(Ntry_vec)){
+
+  Ntry <- Ntry_vec[n]
+
   # generate score
@@
   gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
+
+  data_list[[n]] <- gen_df
+
+}
 ```
 
-```{r plot_data}
-# comparing generated data to the actual data
-bind_rows(
-    gen_df %>% 
-      group_by(sind) %>% 
+```{r plot_data, message=FALSE}
+data_list %>% bind_rows(.id = "size") %>%
+  mutate(size = factor(size, levels = 1:length(Ntry_vec), labels = Ntry_vec)) %>% 
+  group_by(size, sind) %>% 
   summarize(active = mean(Y)) %>%
-      mutate(type="Simulation"),
-
-    df %>% 
-      group_by(sind) %>% 
-      summarize(active = mean(Y)) %>%
-      mutate(type="NHANES")
-) %>%
+  left_join(df %>% group_by(sind) %>% summarize(true_active = mean(Y)),
+            by = "sind") %>%
   ggplot()+
-      geom_line(aes(x=sind, y=active, col = type))+
+  geom_line(aes(x=sind, y=active, col = size), alpha = 0.5)+
+  geom_smooth(aes(x=sind, y=active, col = size), linewidth = 0.5)+
+  geom_line(aes(x=sind, y=true_active), col = "black", 
+            linewidth=0.5)+
   labs(title="Proportion of active subjects by minute")
@@
 
-## fGFPCA
+```{r}
+# containers
+time_vec <- rep(NA, length(Ntry_vec))
+fpca_evals <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
+bam_evals <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
+```
 
-```{r exp_gen_data}
+```{r fGFPCA}
+for(n in seq_along(Ntry_vec)){
+
+  Ntry <- Ntry_vec[n]
+  gen_df <- data_list[[n]]
+
   # bin data
@@
   df_est_latent <- bind_rows(df_est_latent)
-## Figure
-df_est_latent %>% filter(id %in% 1:4) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.5)+
-  geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
-  geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
-  facet_wrap(~id)
 
@@
                           ncol=n_bin, byrow = F) 
-# dim(mat_est_unique)
   fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
-```
 
-```{r}
   # Re-evaluation
@@
   t2 <- Sys.time()
+
+  time_vec[n] <- t2-t1
+  fpca_evals[n, ] <- fpca_mod$evalues
+  bam_evals[n, ] <- 1/this_glmm$sp[2:5]/n_bin
+
+}
 ```
 
-Below I compared the estimated eigenvalues from simulated data to the true eigenvalues. As we see, the reverse did not happen on the simulated data.
+- eigenvalues from FPCA
 
-```{r new_evals, class.source='fold-show'}
-# estimated eigenvalues
-1/this_glmm$sp[2:5]/n_bin
+```{r}
+# time_vec
+data.frame(size = Ntry_vec, fpca_evals)
+```
 
+- eigenvalues from BAM
+
+```{r}
+data.frame(size = Ntry_vec, bam_evals)
+```
+
+- true eigenvalues used to generate data
+
+```{r}
 # true eigenvalues
@@
 ```
-Below is the comparision between true and esimtated eigenfunctions.
 
-```{r new_PC}
-bind_rows(df_phi %>% mutate(type = "True"),
-          est_phi %>% mutate(type = "Estimate",
-                             phi1 = phi1*sqrt(n_bin),
-                             phi2 = phi2*sqrt(n_bin),
-                             phi3 = phi3*sqrt(n_bin),
-                             phi4 = phi4*sqrt(n_bin))) %>%
-  pivot_longer(2:5) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=value, col = type))+
-  facet_wrap(~name)
+<!-- Below is the comparision between true and esimtated eigenfunctions.  -->
+
+<!-- ```{r new_PC} -->
+<!-- bind_rows(df_phi %>% mutate(type = "True"), -->
+<!--           est_phi %>% mutate(type = "Estimate", -->
+<!--                              phi1 = phi1*sqrt(n_bin), -->
+<!--                              phi2 = phi2*sqrt(n_bin), -->
+<!--                              phi3 = phi3*sqrt(n_bin), -->
+<!--                              phi4 = phi4*sqrt(n_bin))) %>% -->
+<!--   pivot_longer(2:5) %>% -->
+<!--   ggplot()+ -->
+<!--   geom_line(aes(x=sind, y=value, col = type))+ -->
+<!--   facet_wrap(~name) -->
+<!-- ``` -->
+
+<!-- And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4.
+The re-evaluation seem to drive the mean function closer to the truth.
+But in that case, why would generated data much more active compared to the NHANES data?
+-->
+
+<!-- ```{r compare_mean} -->
+<!-- bind_rows( -->
+<!--   data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"), -->
+<!--   data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"), -->
+<!--   data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA") -->
+<!-- ) %>% -->
+<!--   ggplot()+ -->
+<!--       geom_line(aes(x=sind, y=mu, col = type))+ -->
+<!--       labs(title="Mean estimates") -->
+<!-- ``` -->
+
+## Bounded mean
+
+- Here, we put a lower bound to the mean function, so that the probablity of active status is greater than 0.2
+
+```{r}
+bd_mean <- ifelse(plogis(true_mu)>=0.2, true_mu, logit(0.2))
+plot(1:J, plogis(true_mu), type = "l", col = "black", 
+     main = "Population mean on the probablity scale",
+     xlab = "sind", ylab = "")
+lines(1:J, plogis(bd_mean), col = "red")
+
+```
+```{r gen_data_bd_mean}
+# increase sample size
+data_list_bdmean <- list()
+
+for(n in seq_along(Ntry_vec)){
+
+  Ntry <- Ntry_vec[n]
+
+  # generate score
+  xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))
+
+  # individual random effects
+  bi <- as.matrix(df_phi[, 2:5]) %*% t(xi)
+
+  # add in unbounded mean
+  eta <- apply(bi, 2, function(x){x+bd_mean})
+  gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
+                       sind = rep(1:J, Ntry),
+                       eta_i = as.vector(eta))
+
+  # binary outcome
+  gen_df$Y <- rbinom(Ntry*J, size=1, prob=plogis(gen_df$eta_i))
+
+  data_list_bdmean[[n]] <- gen_df
+
+}
 ```
 
-And below is the comparision between true mean, mean estimated from FPCA, and re-evaluated mean from the BAM model in Step 4.
-The re-evaluation seem to drive the mean function closer to the truth.
-But in that case, why would generated data much more active compared to the NHANES data?
+```{r plot_data_bdmean, message=FALSE}
+data_list_bdmean %>% bind_rows(.id = "size") %>%
+  mutate(size = factor(size, levels = 1:length(Ntry_vec), labels = Ntry_vec)) %>% 
+  group_by(size, sind) %>% 
+  summarize(active = mean(Y)) %>%
+  left_join(df %>% group_by(sind) %>% summarize(true_active = mean(Y)),
+            by = "sind") %>%
+  ggplot()+
+  geom_line(aes(x=sind, y=active, col = size), alpha = 0.5)+
+  geom_smooth(aes(x=sind, y=active, col = size), linewidth = 0.5)+
+  geom_line(aes(x=sind, y=true_active), col = "black", 
+            linewidth=0.5)+
+  labs(title="Proportion of active subjects by minute")
 
-```{r compare_mean}
-bind_rows(
-  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
-  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"),
-  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
-) %>%
-  ggplot()+
-      geom_line(aes(x=sind, y=mu, col = type))+
-      labs(title="Mean estimates")
 ```
+
+```{r}
+# containers
+fpca_evals_bdmean <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
+bam_evals_bdmean <- matrix(NA, nrow = length(Ntry_vec), ncol = K)
+```
+
+```{r fGFPCA_bdmean}
+for(n in seq_along(Ntry_vec)){
+
+  Ntry <- Ntry_vec[n]
+  gen_df <- data_list_bdmean[[n]]
+
+  # bin data
+  # using the same bin length 
+  gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
+  gen_df$bin <- as.numeric(as.character(gen_df$bin))
+
+  # Step 2:  Local GLMM
+  gen_bin_lst <- split(gen_df, f = gen_df$bin)
+  df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
+  df_est_latent <- bind_rows(df_est_latent)
+
+  # FPCA
+  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
+  mat_est_unique <- matrix(uni_eta_hat$eta_hat,
+                          nrow=Ntry, 
+                          ncol=n_bin, byrow = F) 
+  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
+
+  # Re-evaluation
+  est_phi <- matrix(NA, J, K) 
+  p <- 3 # order of b splines 
+  knots <- 35 # number of knots (same from FPCA model)
+  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
+  knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)
+
+  B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
+                  outer.ok = TRUE)$design  # evaluate B-splines on binned grid
+  Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
+                     outer.ok = TRUE)$design
+  for(k in 1:K){
+    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
+    est_phi[,k] <- Bnew %*% coef(lm_mod)
+  }# project binned eigenfunctions onto the original grid
+
+  ## debias
+  est_phi <- data.frame(sind = 1:J, est_phi)
+  colnames(est_phi) <- c("sind", paste0("phi", 1:K))
+  gen_df <- gen_df %>% left_join(est_phi, by = "sind")
+  gen_df$id <- as.factor(gen_df$id)
+
+  ### model fit with B-splines basis
+  t1 <- Sys.time()
+  this_glmm <- bam(Y ~ s(sind, bs = "cr")+
+                   s(id, by=phi1, bs="re")+
+                   s(id, by=phi2, bs="re")+
+                   s(id, by=phi3, bs="re")+
+                   s(id, by=phi4, bs="re"),
+                 family = binomial,
+                 data=gen_df,
+                 method = "fREML",
+                 discrete = TRUE)
+  t2 <- Sys.time()
+
+  # time_vec[n] <- t2-t1
+  fpca_evals_bdmean[n, ] <- fpca_mod$evalues
+  bam_evals_bdmean[n, ] <- 1/this_glmm$sp[2:5]/n_bin
+
+}
+```
+
+- eigenvalues from FPCA
+
+```{r}
+# time_vec
+data.frame(size = Ntry_vec, fpca_evals_bdmean)
+```
+
+- eigenvalues from BAM
+
+```{r}
+data.frame(size = Ntry_vec, bam_evals_bdmean)
+```

### 2024-05-24 Docs/Presentation.Rmd
try simulation with corrected mean (need rerun on all 500 sims)

--- 2024-02-16 Docs/manuscript.Rmd
@@
 ---
-title: "Manuscript progress report"
+title: "Presentation report"
 author: "Ying Jin"
@@
 ```
+
+This documents mostly produces and saves some figures and tables used for presentation.
 

### 2024-05-24 Code/Simulation_N100/Sim_Subset_fGFPCA.R

@@ -70,8 +70,8 @@ pred_list_all <- list()
 converge_state_list <- list()
 fit_time <- pred_time <- rep(NA, M)
 
-M <- 10
-
+# M <- 10
+M
 
 #### fGFPCA
 
@@ -190,8 +190,9 @@ close(pb)
 
 
 # results
-mean(fit_time)/60 # in minutes
+mean(fit_time, na.rm = T)/60 # in minutes
 class(fit_time)
+mean(pred_time, na.rm=T)
 sum(lapply(converge_state_list, mean)!=1) # all datasets converged
 pred_list_all[[1]] %>%
   filter(t > 0.4) %>%
@@ -205,6 +206,7 @@ save(fit_time_subset_fGFPCA, pred_time_subset_fGFPCA, pred_subset_fGFPCA,
 
 
 #### ISE ####
+window <- seq(0, 1, by = 0.2)
 ise_fgfpca2 <- array(NA, dim = c(length(window)-2, length(window)-2, 10))
 
 
@@ -214,9 +216,9 @@ for(m in 1:length(pred_subset_fGFPCA)){
     mutate(err1 = (pred0.2-eta_i)^2,
            err2 = (pred0.4-eta_i)^2,
            err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
+           err4 = (pred0.8-eta_i)^2) %>% 
     select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
+    mutate(window = cut(t, breaks = seq(0, 1, by=0.2), include.lowest = T)) %>% 
     group_by(window, id) %>% 
     summarise_at(vars(err1, err2, err3, err4), sum) %>% 
     group_by(window) %>% 

### 2024-05-24 Code/Simulation_N500/Sim_fGFPCA.R

@@ -66,7 +66,7 @@ knots_values <- knots_values * (max(mid_t) - min(mid_t)) + min(mid_t)
 
 
 # result container
-# M <- 10
+M <- 10
 M
 pred_list_all <- list()
 converge_state_list <- list()
@@ -211,11 +211,11 @@ length(pred_list_all)
 mean(sapply(converge_state_list, mean)) # all dataset converged
 
 # visualize test samples
-rand_id <- sample(test_id, 4)
+rand_id <- sample(test_id, 2)
 
 # prediction results
 pred_list_all[[4]] %>% 
-  filter(id %in% rand_id) %>% 
+  filter(id %in% sample(test_id, 2)) %>% 
   mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), function(x){exp(x)/(1+exp(x))}) %>%
   ggplot()+
   geom_point(aes(x=sind, y=Y), size = 0.2)+

### 2024-05-28 Docs/Presentation.Rmd
presentation papers

--- 2024-02-07 Docs/manuscript.Rmd
@@
 theme_set(theme_minimal())
-cols <- c("0.2"= "#FDCC8A", "0.4"="#FC8D59", "0.6"="#E34A33", "0.8"="#B30000", "True"="#000000")
-# brewer.pal(5, "OrRd")
+cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
+names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")
+# display.brewer.all()
+# display.brewer.pal(4, "Set2")
+# brewer.pal(5, "Set2")
 
-# define a color palette 
 ## prediction window 
@@
 
-# Method
-
-## Assumptions
-
-- For each subject i in the population, a generalized outcome $Y_i(t)$ is generated along a variable t (for example, time), where $t \in (0, T)$. 
-- The outcome, at any specific t, follows an exponential family distribution characterized by a (latent) continuous function $\eta_i(t)$:
-
-$$g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)$$
-
-$$p(Y_i(t)) = h(Y_i(t))exp\{\eta_i(t)T[Y_i(t)]-A(\eta_i(t))\}$$
-
-- The continuous latent function consists of a population-level fixed process and an individual-level random process
-
-$$\eta_i(t) = \beta_0(t)+b_i(t)$$
-
-## Observed data
-
-In practice we would observe the discrete realization of $\{Y_i(t), t\}$ along a dense grid.
-For simplicity, we assume the observation grid is regular (same across sample).
-When we have J observations points in $(0, T]$, then for the jth observation point, we denote the corresponding value of t as $t_j$, and the corresponding outcome at this point $Y_i(t_j)$.
-
-## fGFPCA Algorithm
-
-### Bin data:
-
-Choose a proper bin width $w$ considering model complexity and identifiability. For now let's say the bins are equal-length and non-overlapping.
-
-- Bin index $s = 1...S$
-- Index of bin midpoints $m_s$
-- Value of t corresponding to bin midpoints $t_{m_s}$
-- Bin endpoints: $(t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]$
-
-### Local GLMMs
-
-At the every bin, we fit a local intercept-only model:
-
-$$g[E(Y_i(t_j))] =\eta_i(t_{m_s})= \beta_0(t_{m_s})+b_i(t_{m_s})$$
-where $t_j \in (t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]$.
-
-Here we are basically saying that the value of latent function is constant within the same bin, which clearly is a misspecification of the true latent process.
-
-From the model above. we will be able to estimate a $\hat{\eta_i}(t_{m_s})$ on the binned grid for every individual in the training sample.
-
-### FPCA
-
-Here, we fit a FPCA model on the $\hat{\eta_i}(t_{m_s})$ obtained from step 2:
-
-$$\hat{\eta}_i(t_{m_s}) = f_0(t_{m_s})+\sum_{k=1}^K\xi_{ik}\phi_{k}(t_{m_s})+\epsilon_i(t_{m_s})$$
-
-where $\xi_{ik}$ independently follows normal distribution $N(0, \lambda_k)$, and $\epsilon_i(t_{m_s})$ at each point follows $N(0, \sigma_2)$.
-
-From this model, we will be able to obtain the following estimates which are shared across population:
-
-- Population mean $\hat{f_0}(t_{m_s})$
-- Basis functions $\hat{\mathbf{\Phi}} = \{\hat{\phi}_1(t_{m_s}), ...,\hat{\phi}_K(t_{m_s}))\}$
-- Estimates of variance of scores $\hat{\lambda}_1...\hat{\lambda}_K$
-
-### Projection and Debias
-
-The mean and basis functions are evaluated on the binned grid.
-To extend it to the original measurement grid data was collected on, we project the estimated eigenfunctions $\hat{\mathbf{\Phi}}$ back use spline basis.
-Now we have extend the $\hat{\phi}_k(t_{m_s})$ to the original grid $\hat{\phi}_k(t_j)$
-
-Because of the misspecification of local GLMMs, the estimated eigenfunctions and eigenvalues are also biased by a constant multiplicative effect.
-Therefore, we use a GLMM to re-evaluate the mean function, eigenfunctions and eigenvalues.
-
-## Out-of-sample prediction
-
-Now, let's assume we have a new subject $u$ with $J_u$ observations ($J_u < J$). Then the log-likelihood of this new subject would be:
-
-$$l_u=\sum_{t_j<t_{J_u}}log(h(Y_u(t_j)))+\hat{\eta}_u(t_j)T(Y_u(t_j))-log(A[\hat{\eta}_u(t_j)])$$
-
-where $\hat{\eta}_u(t_j) = \hat{f}_0(t_j)+\sum_{k=1}^K \xi_{uk}\hat{\phi}(t_j)$.
-
-With estimates for the population-level parameters from fGFPCA algorithms above, we can estimate $\xi_{uk}$ by maximization of $l_u$.
-Direct maximization some times does not have closed form solution.
-Numeric maximization methods seem not very stable as well.
-So I have decided to used a Bayes approach (Laplace Approximation):
-
-- Prior distribution: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$
-- Posterior distribution: the likelihood of $l_u =l(Y_u(t_j)|\mathbf{\xi}_u)$
-
-Laplace Approximation would get the posterior mode of $\xi_{uk}$ through quadratic approximation.
-
-# Reference method
-
-## GLMMadaptive
-
-- For large datasets, we can fit a model with random intercept and slope for time.
-It is doable on 500 datasets, but obviously too simple for the data generation scheme.
-We would expect it to perform terribly.
-
-$$g(E(Y_i(t))) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$$
-
-- For small datasets, we would like to fit fGFPCA and GLMMadaptive on a dataset with smaller sample size and/or smaller measurement density.
-For the GLMMadaptive model, we would set it up with spline basis functions so that its flexibility is comparable with fGFPCA model, such as:
-
-$$g(E(Y_i(t))) = \sum_{k=1}^4\zeta_{k}B_k(t)+\sum_{l=1}^4\xi_{il}\phi_l(t)$$
-
-## Generalized functional-on-scalar regression
-
-We would like to use a second reference method for predictive performance comparision.
-In this model, predictions on a specific interval are all made using the last few observations in the observed window.
-For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2 ($t_j: j=200, ...
-200-(L-1)$) as time-fixed covariate to predict any future time.
-
-Let's write out the model expression (with questionable notation). If the observations if up to $t_m$, then:
-
-\[\begin{aligned}
-g(E[Y_i(t)]) &= \beta_0(t) +\sum_{l=1}^L \beta_l (t)Y_i(t_l)\\
-l & = m,...m-(j-1)\\
-t &> t_m 
-\end{aligned}\]
-
-This is a simple function-on-scalar model with no random effect, meaning all subject with the same last observed outcome would have the same estimated/predicted latent track.
-This is clearly anti-intuitive.
-But adding random effects would make out-of-sample prediction impossible.
-
-Under both frameworks, for each dataset we need to fit four models (similar to the GLMMadaptive method):
-
-- Given 0-0.2, predict 0.2-1
-- Given 0-0.4, predict 0.4-1
-- Given 0-0.6, predict 0.6-1
-- Given 0-0.8, predict 0.8-1
-
-But prediction performance will be reported by equal-length time window.
-
-# Larger-scale simulation
-
-## Simulation set up
-
-Here we simulate binary data from cyclic latent process:
-
-\[\begin{aligned}
-Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
-\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
-\end{aligned}\]
-
-where:
-
-- $t$ is 1000 equal-spaced observations points on $[0, 1]$ (J = 1000).
-- $f_0(t)=0$
-- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively. 
-- Sample size $N = 500$
-- In the binning step, we bin every 10 observations
-- 500 simulations were implemented
-
-## Figures
+# Simulation
 
@@
 
-rand_id <- sample(501:600, size = 4)
 ```
-
-```{r fig_large_sim, fig.height=8, fig.width=8}
-bind_rows(
-  pred_list_all[[1]] %>% filter(id %in% rand_id) %>% 
-    mutate(method="fGFPCA"),
-  pred_list_ref[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GLMMadaptive"),
-  pred_list_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=1)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_list_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=5)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4)
-) %>% mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
-                .funs = function(x){exp(x)/(1+exp(x))}) %>%
-  mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_grid(rows = vars(id), cols = vars(method))+
-  labs(col = "Maximum observation time", x = "Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(legend.position = "bottom")+
-  scale_color_manual(values = cols)
-```
-
-## ISE and AUC
-
-```{r ise}
-## ISE container 
-ise_fgfpca <- ise_adglmm <- ise_gfosr_l1 <- ise_gfosr_l5 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-# dims: prediction window, max obs time, simulation iter
-
-```
-
-```{r ise_fGFPCA}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_all[[m]]
-  ise_tb <- pred_list_all[[m]] %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err"))
-  ise_fgfpca[, ,m] <- as.matrix(ise_tb)
-
-}
-
-mean_ise_fgfpca <- apply(ise_fgfpca, c(1, 2), mean)
-
-# mean_ise <- data.frame(mean_ise) %>%
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_ise_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_adglmm}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_gfosr_l1[[m]]
-  ise_tb <- pred_list_ref[[m]] %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err"))
-  ise_adglmm[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_adglmm <- apply(ise_adglmm, c(1, 2), mean)
-colnames(mean_ise_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l1}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_ref[[m]]
-  ise_tb <- pred_list_gfosr_l1[[m]] %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    # mutate(window = factor(window, levels = 1:5, 
-    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err"))
-  ise_gfosr_l1[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_gfosr_l1 <- apply(ise_gfosr_l1, c(1, 2), mean)
-colnames(mean_ise_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l5}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_ref[[m]]
-  ise_tb <- pred_list_gfosr_l5[[m]] %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    # mutate(window = factor(window, levels = 1:5, 
-    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err"))
-  ise_gfosr_l5[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_gfosr_l5 <- apply(ise_gfosr_l5, c(1, 2), mean)
-colnames(mean_ise_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_func}
-## a function to calculate AUC
-get_auc <- function(y, pred){
-  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
-    auc <- NA
-  }
-  else{
-    this_perf <- performance(prediction(pred, y), measure = "auc")
-    auc <- this_perf@y.values[[1]]
-  }
-  return(auc)
-}
-
-```
-
-```{r}
-## auc container 
-auc_fgfpca <- auc_adglmm <- auc_gfosr_l1 <- auc_gfosr_l5 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-```
-
-```{r auc_fGFPCA}
-for(m in 1:M){
-  this_df <- pred_list_all[[m]]
-  auc_tb <- this_df %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred0.2),
-              auc2 = get_auc(Y, pred0.4),
-              auc3 = get_auc(Y, pred0.6),
-              auc4 = get_auc(Y, pred0.8)) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("auc"))
-  auc_fgfpca[, ,m] <- as.matrix(auc_tb)
-}
-
-mean_auc_fgfpca <- apply(auc_fgfpca, c(1, 2), mean)
-# mean_auc <- data.frame(mean_auc) %>% 
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_auc_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GLMMadaptive}
-for(m in 1:M){
-  this_df <- pred_list_ref[[m]]
-  auc_tb <- this_df %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred0.2),
-              auc2 = get_auc(Y, pred0.4),
-              auc3 = get_auc(Y, pred0.6),
-              auc4 = get_auc(Y, pred0.8)) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("auc"))
-  auc_adglmm[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_adglmm <- apply(auc_adglmm, c(1, 2), mean)
-colnames(mean_auc_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_gfosr_l1}
-for(m in 1:M){
-  this_df <- pred_list_gfosr_l1[[m]]
-  auc_tb <- this_df %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred_w1),
-              auc2 = get_auc(Y, pred_w2),
-              auc3 = get_auc(Y, pred_w3),
-              auc4 = get_auc(Y, pred_w4)) %>%
-    filter(window != 1) %>% 
-    select(starts_with("auc"))
-  auc_gfosr_l1[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_gfosr_l1 <- apply(auc_gfosr_l1, c(1, 2), mean)
-colnames(mean_auc_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_gfosr_l5}
-for(m in 1:M){
-  this_df <- pred_list_gfosr_l5[[m]]
-  auc_tb <- this_df %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred_w1),
-              auc2 = get_auc(Y, pred_w2),
-              auc3 = get_auc(Y, pred_w3),
-              auc4 = get_auc(Y, pred_w4)) %>%
-    filter(window != 1) %>% 
-    select(starts_with("auc"))
-  auc_gfosr_l5[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_gfosr_l5 <- apply(auc_gfosr_l5, c(1, 2), mean)
-colnames(mean_auc_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r table_sim1}
-fgfpca1 <- rbind(mean_ise_fgfpca ,mean_auc_fgfpca) %>% as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
-gfosr_l5_1 <- rbind(mean_ise_gfosr_l5 ,mean_auc_gfosr_l5) %>% as_tibble()
-gfosr_l1_1 <- rbind(mean_ise_gfosr_l1 ,mean_auc_gfosr_l1) %>% as_tibble()
-adglmm1 <- rbind(mean_ise_adglmm ,mean_auc_adglmm) %>% as_tibble()
-
-bind_cols(fgfpca1, gfosr_l5_1, gfosr_l1_1, adglmm1, .name_repair = "minimal") %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" " = 1, "fGFPCA" = 4, "GFOSR (L=5)" = 4, 
-                     "GFOSR (L=1)" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
-  group_rows(index = c("ISE"= 4, "AUC" = 4)) %>%
-  landscape()
-
-```
-
-## Time (minutes)
-
-I chose not to report fitting and prediction time separately because it is too difficult to tell them apart for the GFOSR models.
-
-```{r}
-data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
-  "Time" = c(mean(fit_time/60+pred_time),
-             mean(fit_time_ref+pred_time_ref/60),
-             mean(t_vec_gfosr_l1/60), 
-             mean(t_vec_gfosr_l5/60))) %>%
-  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
-  kable_styling(full_width = F) 
-```
-
-# Small-scale simulation
-
-I have used 100 subjects for training and testing, and repeated 500 times.
-When fitting GLMMadpative, I reduce the number of measurements in the training dataset to 1/10 by taking one every 10 observations.
-The prediction is on the original grid.
 
@@
 rm(pred_list_gfofr_subset, t_vec_gfofr_subset)
-
-rand_id <- sample(101:200, size = 4)
 ```
 
-## Figure
+```{r sim_example, fig.height=2, fig.width=8}
+pred_list_all[[1]] %>% filter(id %in% sample(501: 600, 4)) %>% 
+  mutate(eta_i = plogis(eta_i)) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "eta_i"), show.legend = F)+
+  facet_grid(cols=vars(id))+
+  scale_color_manual(values = "#FC8D62")+
+  scale_x_continuous(breaks = seq(0, 1000, by = 200))+
+  labs(x="t")
+```
 
-```{r fig_small_sim, fig.height=8, fig.width=8}
-bind_rows(
-  pred_subset_fGFPCA[[1]] %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
-  pred_subset_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)") %>% 
+```{r fig_large_sim, fig.height=8, fig.width=8}
+sim1 <- bind_rows(
+  pred_list_all[[1]] %>% filter(id == 580) %>% 
+    mutate(method="fGFPCA"),
+  pred_list_ref[[1]] %>% filter(id == 580) %>%
+    mutate(method = "GLMMadaptive"),
+  pred_list_gfosr_l1[[1]] %>% filter(id == 580) %>%
+    mutate(method = "GFOSR (L=1)") %>%
     rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)") %>%
+  pred_list_gfosr_l5[[1]] %>% filter(id == 580) %>%
+    mutate(method = "GFOSR (L=5)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4)
+) %>% mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive")))
+
+sim2 <- bind_rows(
+  pred_subset_fGFPCA[[1]] %>% filter(id == 189) %>% mutate(method="fGFPCA"),
+  pred_subset_gfosr_l5[[1]] %>% filter(id == 189) %>% mutate(method = "GFOSR (L=5)") %>% 
     rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_adglmm[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
+  pred_subset_gfosr_l1[[1]] %>% filter(id == 189) %>% mutate(method = "GFOSR (L=1)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_subset_adglmm[[1]] %>% filter(id == 189) %>% mutate(method = "GLMMadaptive")) %>%
   mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
@@
   mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive")))
+```
+
+```{r sim, fig.height=4, fig.width=10}
+bind_rows(sim1 %>% mutate(sim = "Simulation 1"), sim2 %>% mutate(sim = "Simulation 2")) %>%
   ggplot()+
   geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"))+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linetype="dashed", na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linetype="dashed", na.rm=T)+
-  facet_grid(rows = vars(id), cols = vars(method))+
+  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm=T, linewidth = 1.0)+
+  facet_grid(rows = vars(sim), cols = vars(method))+
   labs(col = "Maximum observation time", x = "Time", y="")+
   scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(legend.position = "bottom")+
+  theme(legend.position = "right")+
   scale_color_manual(values = cols)
@@
 
-## ISE and AUC
+# Data application
 
-```{r ise_subset}
-## ISE container 
-ise_fgfpca2 <- ise_adglmm2 <- ise_gfosr_l1_2 <- ise_gfosr_l5_2 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-# dims: prediction window, max obs time, simulation iter
+```{r nhanes_exp, fig.width=8, fig.height=5}
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+df_nhanes %>% 
+  select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% sample(unique(df_nhanes$SEQN), size = 100)) %>%
+  filter(Z == 1) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=SEQN), size = 0.2)+
+  labs(x = "")+
+  labs(x = "Time", y="Subject")+
+  scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  theme(axis.text.y = element_blank())
 
 ```
-
-```{r ise_fGFPCA_subset}
-for(m in 1:length(pred_subset_fGFPCA)){
-  this_df <- pred_subset_fGFPCA[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_fgfpca2[,,m] <- ise_tb_m
-}
-
-mean_ise_fgfpca2 <- apply(ise_fgfpca2, c(1, 2), mean)
-
-colnames(mean_ise_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_GLMMadaptive_subset}
-for(m in 1:length(pred_subset_adglmm)){
-  this_df <- pred_subset_adglmm[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_adglmm2[,,m] <- ise_tb_m
-}
-
-mean_ise_adglmm2<- apply(ise_adglmm2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l1_subset}
-for(m in 1:length(pred_subset_gfosr_l1)){
-  this_df <- pred_subset_gfosr_l1[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_gfosr_l1_2[,,m] <- ise_tb_m
-}
-
-mean_ise_gfosr_l1_2<- apply(ise_gfosr_l1_2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l5_subset}
-for(m in 1:length(pred_subset_gfosr_l5)){
-  this_df <- pred_subset_gfosr_l5[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_gfosr_l5_2[,,m] <- ise_tb_m
-}
-
-mean_ise_gfosr_l5_2 <- apply(ise_gfosr_l5_2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc2}
-## auc container 
-auc_fgfpca2 <- auc_adglmm2 <- auc_gfosr_l1_2 <- auc_gfosr_l5_2 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-```
-
-```{r auc_fGFPCA_subset}
-for(m in 1:length(pred_subset_fGFPCA)){
-  this_df <- pred_subset_fGFPCA[[m]]
-  auc_tb_m <- this_df %>%
-  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.2),
-            auc2 = get_auc(Y, pred0.4),
-            auc3 = get_auc(Y, pred0.6),
-            auc4 = get_auc(Y, pred0.8)) %>%
-  filter(window != "[0,0.2]") %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_fgfpca2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_fgfpca2 <- apply(auc_fgfpca2, c(1,2), mean)
-colnames(mean_auc_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GLMMadaptive_subset}
-for(m in 1:length(pred_subset_adglmm)){
-  this_df <- pred_subset_adglmm[[m]]
-  auc_tb_m <- this_df %>%
-  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.2),
-            auc2 = get_auc(Y, pred0.4),
-            auc3 = get_auc(Y, pred0.6),
-            auc4 = get_auc(Y, pred0.8)) %>%
-  filter(window != "[0,0.2]") %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_adglmm2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_adglmm2 <- apply(auc_adglmm2, c(1,2), mean, na.rm = T)
-colnames(mean_auc_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GFOSR_subset_l1}
-for(m in 1:length(pred_subset_gfosr_l1)){
-  this_df <- pred_subset_gfosr_l1[[m]]
-  auc_tb_m <- this_df %>%
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred_w1),
-            auc2 = get_auc(Y, pred_w2),
-            auc3 = get_auc(Y, pred_w3),
-            auc4 = get_auc(Y, pred_w4)) %>%
-  filter(window != 1) %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_gfosr_l1_2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_gfosr_l1_2 <- apply(auc_gfosr_l1_2, c(1,2), mean)
-colnames(mean_auc_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GFOSR_subset_l5}
-for(m in 1:length(pred_subset_gfosr_l5)){
-  this_df <- pred_subset_gfosr_l5[[m]]
-  auc_tb_m <- this_df %>%
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred_w1),
-            auc2 = get_auc(Y, pred_w2),
-            auc3 = get_auc(Y, pred_w3),
-            auc4 = get_auc(Y, pred_w4)) %>%
-  filter(window != 1) %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_gfosr_l5_2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_gfosr_l5_2 <- apply(auc_gfosr_l5_2, c(1,2), mean)
-colnames(mean_auc_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r table_sim2}
-fgfpca2 <- rbind(mean_ise_fgfpca2 ,mean_auc_fgfpca2) %>% as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
-gfosr_l5_2 <- rbind(mean_ise_gfosr_l5_2 ,mean_auc_gfosr_l5_2) %>% as_tibble()
-gfosr_l1_2 <- rbind(mean_ise_gfosr_l1_2 ,mean_auc_gfosr_l1_2) %>% as_tibble()
-adglmm2 <- rbind(mean_ise_adglmm2 ,mean_auc_adglmm2) %>% as_tibble()
-
-bind_cols(fgfpca2, gfosr_l5_2, gfosr_l1_2, adglmm2, .name_repair = "minimal") %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" " = 1, "fGFPCA" = 4, "GFOSR (L=5)" = 4, 
-                     "GFOSR (L=1)" = 4, "GLMMadaptive" = 4)) %>%
-  add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
-  group_rows(index = c("ISE"= 4, "AUC" = 4)) %>%
-  landscape()
-
-```
-
-Among 500 hundred iterations, `r sum(num_probs)` did not converge for the GLMMadaptive model. No numeric issue for fGFPCA.
-
-# Time
-
-```{r}
-data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
-  "Time" = c(mean(fit_time_subset_fGFPCA/60+pred_time_subset_fGFPCA),
-             mean(fit_time_subset_adglmm+pred_time_subset_adglmm/60),
-             mean(t_subset_gfosr_l1/60), 
-             mean(t_subset_gfosr_l5/60))) %>%
-  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
-  kable_styling(full_width = F) 
-```
-
-# NHANES data application
-
-We take 60% subjects (N = 5257) for training, 40% (n = 3506) for out-of-sample prediction.
-When using the full sample The debiase step (mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only works for intercept-only model.
-In the debias step, re-evaluated eigenvalues also lost their decreasing nature.
 
@@
 
-```{r nhance_example, eval=FALSE}
-df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% unique(nhanes_pred_adglmm$id) | SEQN %in% unique(nhanes_pred_fgfpca$id)) %>%
-  mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
-  mutate(SEQN = factor(SEQN)) %>%
-  ggplot()+
-  geom_tile(aes(x=sind, y = SEQN, fill = Z))+
-  labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
-  theme(axis.text.y = element_blank())+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480),
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))
-```
-
-## AUC
-
-```{r}
-# window
-window <- seq(0, 1440, by = 360)
-```
-
-```{r auc_appl}
-auc_appl_fgfpca <- pred_nhanes_fgfpca %>% 
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_gfosr_l5 <- pred_appl_gfosr_l5 %>% 
-  select(-window) %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_gfosr_l1 <- pred_appl_gfosr_l1 %>% 
-  select(-window) %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_adglmm <- pred_nhanes_adglmm %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-colnames(auc_appl_fgfpca) <- colnames(auc_appl_gfosr_l5) <- colnames(auc_appl_gfosr_l1) <- colnames(auc_appl_adglmm) <- c("6am", "12pm", "6pm")
-```
-
-```{r table_appl}
-bind_cols(Window = c("6am-12pm", "12pm-6pm", "6pm-12am"),
-          auc_appl_fgfpca, auc_appl_gfosr_l5, auc_appl_gfosr_l1, auc_appl_adglmm,
-          .name_repair = "minimal") %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" " = 1, "fGFPCA" = 3, "GFOSR (L=5)" = 3, 
-                     "GFOSR (L=1)" = 3, "GLMMadaptive" = 3)) %>%
-  add_header_above(c(" " = 1, "Maximum observation time" = 12)) %>%
-  landscape()
-```
-
-## Figure
-
-```{r fig_appl, fig.height=8, fig.width=8}
-rand_id 
+```{r fig_appl_paper, fig.height=4, fig.width=10}
+# redefine color scale
 cols <- cols[1:3] 
 names(cols) <- c("6am", "12pm", "6pm")
+
+# selct four subjects
+plot_id <- sort(c(64198, 80517))
+
+# figure
 bind_rows(
-  pred_nhanes_fgfpca %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
-  pred_appl_gfosr_l5 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)"),
-  pred_appl_gfosr_l1 %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)"),
-  pred_nhanes_adglmm %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
+  pred_nhanes_fgfpca %>% filter(id %in% plot_id) %>% mutate(method="fGFPCA"),
+  pred_appl_gfosr_l5 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=5)"),
+  pred_appl_gfosr_l1 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=1)"),
+  pred_nhanes_adglmm %>% filter(id %in% plot_id) %>% mutate(method = "GLMMadaptive")) %>%
   mutate_at(vars(pred360, pred720, pred1080), 
@@
                      labels = c("0am", "6am", "12pm", "6pm", "12am"))+
-  theme(legend.position = "bottom", axis.text.x = element_text(size = 7))+
+  theme(legend.position = "right", axis.text.x = element_text(size = 7))+
   scale_color_manual(values = cols)
 ```
-
-# Additional figures
-
-```{r eval=FALSE}
-df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
-df_nhanes %>% 
-  select(SEQN, Z, sind) %>%
-  filter(SEQN %in% rand_id) %>%
-  filter(sind %in% 356:360) %>%
-  pivot_wider(names_from = sind, values_from = Z) %>% 
-  select(-SEQN) %>% distinct(.)
-
-pred_appl_gfosr_l5 %>% 
-  filter(id %in% rand_id) %>%
-  select(id, sind, pred360) %>%
-  pivot_wider(names_from = sind, values_from = pred360) %>% 
-  select(-id) %>% distinct(.)
-```
-
-<!-- ## fGFPCA algorithm demonstration -->
-
-<!-- ```{r binning, eval=FALSE} -->
-<!-- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
-<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
-<!--   scale_y_continuous(breaks = c(0, 1))+ -->
-<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
-<!-- ``` -->
-
-<!-- ```{r,eval=FALSE} -->
-<!-- df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
-<!--   mutate(bin = cut(sind, breaks = seq(650, 750, by = 10),  -->
-<!--                    labels = seq(655, 745, by = 10), -->
-<!--                    include.lowest = T))  -->
-<!-- bin_mid <- seq(655, 745, by = 10) -->
-
-<!-- for(i in bin_mid){ -->
-<!--   df_bin_i <- df_fig %>% filter(bin==i)  -->
-<!--   glm_i <- glmer(Z ~ 1 + (1|SEQN), data = df_bin_i, family = binomial, nAGQ=0) -->
-<!--   eta_i <- predict(glm_i, type = "link") -->
-
-<!--   df_fig$eta_hat[df_fig$bin==i] <- eta_i -->
-<!-- } -->
-
-<!-- df_fig %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN), -->
-<!--          eta_hat = exp(eta_hat)/(1+exp(eta_hat)), -->
-<!--          bin=as.numeric(as.character(bin))) %>%  -->
-
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
-<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
-<!--   geom_point(aes(x=bin, y=eta_hat, col = SEQN))+ -->
-<!--   geom_line(aes(x=bin, y=eta_hat, col = SEQN))+ -->
-<!--   scale_y_continuous(breaks = c(0, 1))+ -->
-<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
-
-<!-- ``` -->
-
-<!-- ## Example of continuous and binary NHANES -->
-
-<!-- ```{r, eval=FALSE} -->
-<!-- df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds")) -->
-<!-- p_cont <- df_nhanes_cont %>%  -->
-<!--   filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>%  -->
-<!--   select(starts_with("min"), SEQN) %>% -->
-<!--   pivot_longer(1:1440, names_to = "Minute") %>%  -->
-<!--   mutate(Minute = gsub("MIN", "", Minute)) %>% -->
-<!--   mutate(Minute = as.numeric(Minute), -->
-<!--          SEQN=as.factor(SEQN)) %>% -->
-<!--   ggplot(aes(x=Minute, y=value, col=SEQN))+ -->
-<!--   geom_point(size = 0.5, alpha=0.5)+ -->
-<!--   geom_smooth(se=F, size = 0.5)+ -->
-<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
-<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
-<!--   theme(axis.text = element_text(size = 5))+ -->
-<!--   labs(y="", title = "NHANES Activitiy count (2003-2004)") -->
-<!-- ``` -->
-
-<!-- ```{r, eval=FALSE} -->
-<!-- # a example of continuous and binary NHANES data -->
-<!-- p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & Z == 1) %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = SEQN), size = 0.5)+ -->
-<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
-<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
-<!--   theme(axis.text = element_text(size = 5))+ -->
-<!--   labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)") -->
-<!-- ``` -->
-
-<!-- ```{r data_exp, fig.height=4, fig.width=10, eval=FALSE} -->
-<!-- ggarrange(p_cont, p_bi, nrow = 1) -->
-<!-- ``` -->
-
-<!-- ## Dynamic prediction example -->
-
-<!-- ```{r dynpred_exp,  fig.height=4, fig.width=10, eval=FALSE} -->
-<!-- ggarrange( -->
-<!--   pred_list_all[[1]] %>%  -->
-<!--     filter(id==501) %>% -->
-<!--     mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>% -->
-<!--     ggplot()+ -->
-<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
-<!--     geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T,  -->
-<!--               linetype = "dashed")+ -->
-<!--     geom_vline(xintercept = 0.2, linetype = "dashed")+ -->
-<!--     labs(x="Time", y="")+ -->
-<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
-<!--     theme(axis.text = element_text(size = 5))+ -->
-<!--     ylim(-1.2, 1.7), -->
-
-<!--  pred_list_all[[1]] %>%  -->
-<!--     filter(id==501) %>% -->
-<!--     mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>% -->
-<!--     ggplot()+ -->
-<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
-<!--     geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T,  -->
-<!--               linetype = "dashed")+ -->
-<!--     geom_vline(xintercept = 0.4, linetype = "dashed")+ -->
-<!--     labs(x="Time", y="")+ -->
-<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
-<!--     theme(axis.text = element_text(size = 5))+ -->
-<!--    ylim(-1.2, 1.7), -->
-<!--  nrow = 1, common.legend = T -->
-<!--   ) -->
-
-<!-- ``` -->

### 2024-05-28 Docs/manuscript.Rmd
presentation papers

@@
 
-```{r fig_large_sim, fig.height=8, fig.width=8}
+```{r fig_large_sim_paper, fig.height=8, fig.width=8}
 bind_rows(
@@
 
-## Time (minutes)
-
-I chose not to report fitting and prediction time separately because it is too difficult to tell them apart for the GFOSR models.
-
-```{r}
-data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
-  "Time" = c(mean(fit_time/60+pred_time),
-             mean(fit_time_ref+pred_time_ref/60),
-             mean(t_vec_gfosr_l1/60), 
-             mean(t_vec_gfosr_l5/60))) %>%
-  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
-  kable_styling(full_width = F) 
-```
-
 # Small-scale simulation
@@
 
-```{r fig_small_sim, fig.height=8, fig.width=8}
+```{r fig_small_sim_paper, fig.height=8, fig.width=8}
 bind_rows(
@@
 
+```{r}
+sep_plot <- bind_rows(
+  pred_subset_fGFPCA[[1]] %>% filter(id==189) %>% mutate(method="fGFPCA"),
+  pred_subset_gfosr_l5[[1]] %>% filter(id==189) %>% mutate(method = "GFOSR (L=5)") %>% 
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_subset_gfosr_l1[[1]] %>% filter(id==189) %>% mutate(method = "GFOSR (L=1)") %>%
+    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
+  pred_subset_adglmm[[1]] %>% filter(id==189) %>% mutate(method = "GLMMadaptive")) %>%
+  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linewidth=1, na.rm=T)+
+  facet_grid(cols = vars(method))+
+  labs(col = "Maximum observation time", x = "t", y="")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
+```
+
+```{r, fig_small_sim_pre, fig.height=2, fig.width=8}
+sep_plot+scale_color_manual(values = cols, guide="none")+
+  #theme(legend.position = "bottom")+
+  scale_color_manual(values = cols, guide = "none")
+```
+
+```{r, legend, fig.height=0.2, fig.width=8}
+as_ggplot(ggpubr::get_legend(sep_plot+theme(legend.position = "bottom")+
+                               scale_color_manual(values = cols)))
+```
+
 ## ISE and AUC
@@
 
+Average time per iteration in minutes
+
 ```{r}
-data.frame(
-  "Method" = c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)"),
-  "Time" = c(mean(fit_time_subset_fGFPCA/60+pred_time_subset_fGFPCA),
+time_tb <- data.frame(sim1 = c(mean(fit_time/60+pred_time),
+             mean(fit_time_ref+pred_time_ref/60),
+             mean(t_vec_gfosr_l1/60), 
+             mean(t_vec_gfosr_l5/60)),
+           sim2 = c(mean(fit_time_subset_fGFPCA/60+pred_time_subset_fGFPCA),
              mean(fit_time_subset_adglmm+pred_time_subset_adglmm/60),
              mean(t_subset_gfosr_l1/60), 
-             mean(t_subset_gfosr_l5/60))) %>%
+             mean(t_subset_gfosr_l5/60))) %>% t()
+colnames(time_tb) <- c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)")
+rownames(time_tb) <- c("Simulation 1", "Simulation 2")
+```
+
+```{r time_tb}
+time_tb %>% 
   kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
@@
 
-```{r fig_appl, fig.height=8, fig.width=8}
+```{r fig_appl_paper, fig.height=8, fig.width=8}
 # redefine color scale
@@
 
+```{r fig_appl_pre, fig.height=4, fig.width=8}
+# selct four subjects
+plot_id <- sort(c(64198, 80517))
+
+# figure
+bind_rows(
+  pred_nhanes_fgfpca %>% filter(id %in% plot_id) %>% mutate(method="fGFPCA"),
+  pred_appl_gfosr_l5 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=5)"),
+  pred_appl_gfosr_l1 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=1)"),
+  pred_nhanes_adglmm %>% filter(id %in% plot_id) %>% mutate(method = "GLMMadaptive")) %>%
+  mutate_at(vars(pred360, pred720, pred1080), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+  ggplot()+
+  geom_point(aes(x=sind, y=Y), size = 0.2)+
+  geom_line(aes(x=sind, y=pred360, col = "6am"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=sind, y=pred720, col = "12pm"), linewidth=1, na.rm=T)+
+  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linewidth=1, na.rm=T)+
+  facet_grid(rows = vars(id), cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="")+
+  scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  theme(legend.position = "bottom", axis.text.x = element_text(size = 7))+
+  scale_color_manual(values = cols)
+```
+
 The fGFPCA prediction conditional on observation up to 6am looks identical across subjects.
@@
 ```
-
-# Additional figures
-
-```{r eval=FALSE}
-df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
-df_nhanes %>% 
-  select(SEQN, Z, sind) %>%
-  filter(SEQN %in% rand_id) %>%
-  filter(sind %in% 356:360) %>%
-  pivot_wider(names_from = sind, values_from = Z) %>% 
-  select(-SEQN) %>% distinct(.)
-
-pred_appl_gfosr_l5 %>% 
-  filter(id %in% rand_id) %>%
-  select(id, sind, pred360) %>%
-  pivot_wider(names_from = sind, values_from = pred360) %>% 
-  select(-id) %>% distinct(.)
-```
-
-<!-- ## fGFPCA algorithm demonstration -->
-
-<!-- ```{r binning, eval=FALSE} -->
-<!-- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
-<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
-<!--   scale_y_continuous(breaks = c(0, 1))+ -->
-<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
-<!-- ``` -->
-
-<!-- ```{r,eval=FALSE} -->
-<!-- df_fig <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & sind >= 650 & sind <= 750) %>%  -->
-<!--   mutate(bin = cut(sind, breaks = seq(650, 750, by = 10),  -->
-<!--                    labels = seq(655, 745, by = 10), -->
-<!--                    include.lowest = T))  -->
-<!-- bin_mid <- seq(655, 745, by = 10) -->
-
-<!-- for(i in bin_mid){ -->
-<!--   df_bin_i <- df_fig %>% filter(bin==i)  -->
-<!--   glm_i <- glmer(Z ~ 1 + (1|SEQN), data = df_bin_i, family = binomial, nAGQ=0) -->
-<!--   eta_i <- predict(glm_i, type = "link") -->
-
-<!--   df_fig$eta_hat[df_fig$bin==i] <- eta_i -->
-<!-- } -->
-
-<!-- df_fig %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN), -->
-<!--          eta_hat = exp(eta_hat)/(1+exp(eta_hat)), -->
-<!--          bin=as.numeric(as.character(bin))) %>%  -->
-
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = Z, col = SEQN), size = 0.5)+ -->
-<!--   geom_vline(xintercept = seq(650, 750, by = 10), linetype="dashed")+ -->
-<!--   geom_point(aes(x=bin, y=eta_hat, col = SEQN))+ -->
-<!--   geom_line(aes(x=bin, y=eta_hat, col = SEQN))+ -->
-<!--   scale_y_continuous(breaks = c(0, 1))+ -->
-<!--   labs(x="Time", y="", title = "NHANES binary activity indicator") -->
-
-<!-- ``` -->
-
-<!-- ## Example of continuous and binary NHANES -->
-
-<!-- ```{r, eval=FALSE} -->
-<!-- df_nhanes_cont <- readRDS(here("Data/NHANES_AC_processed.rds")) -->
-<!-- p_cont <- df_nhanes_cont %>%  -->
-<!--   filter(SEQN %in% sample(unique(df_nhanes_cont$SEQN), 4) & WEEKDAY==6) %>%  -->
-<!--   select(starts_with("min"), SEQN) %>% -->
-<!--   pivot_longer(1:1440, names_to = "Minute") %>%  -->
-<!--   mutate(Minute = gsub("MIN", "", Minute)) %>% -->
-<!--   mutate(Minute = as.numeric(Minute), -->
-<!--          SEQN=as.factor(SEQN)) %>% -->
-<!--   ggplot(aes(x=Minute, y=value, col=SEQN))+ -->
-<!--   geom_point(size = 0.5, alpha=0.5)+ -->
-<!--   geom_smooth(se=F, size = 0.5)+ -->
-<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
-<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
-<!--   theme(axis.text = element_text(size = 5))+ -->
-<!--   labs(y="", title = "NHANES Activitiy count (2003-2004)") -->
-<!-- ``` -->
-
-<!-- ```{r, eval=FALSE} -->
-<!-- # a example of continuous and binary NHANES data -->
-<!-- p_bi <- df_nhanes %>% select(SEQN, Z, sind) %>%  -->
-<!--   filter(SEQN %in% rand_id3 & Z == 1) %>%  -->
-<!--   mutate(SEQN = as.factor(SEQN)) %>%  -->
-<!--   ggplot()+ -->
-<!--   geom_point(aes(x=sind, y = SEQN), size = 0.5)+ -->
-<!--   scale_x_continuous(breaks = seq(0, 1440, by = 480), -->
-<!--                      labels = c("Midnight", "8am", "4pm", "Midnight"))+ -->
-<!--   theme(axis.text = element_text(size = 5))+ -->
-<!--   labs(x="Time", y="Subject", title = "NHANES binary activity indicator (2011-2014)") -->
-<!-- ``` -->
-
-<!-- ```{r data_exp, fig.height=4, fig.width=10, eval=FALSE} -->
-<!-- ggarrange(p_cont, p_bi, nrow = 1) -->
-<!-- ``` -->
-
-<!-- ## Dynamic prediction example -->
-
-<!-- ```{r dynpred_exp,  fig.height=4, fig.width=10, eval=FALSE} -->
-<!-- ggarrange( -->
-<!--   pred_list_all[[1]] %>%  -->
-<!--     filter(id==501) %>% -->
-<!--     mutate(eta_i = ifelse(t<=0.2, eta_i, NA)) %>% -->
-<!--     ggplot()+ -->
-<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
-<!--     geom_line(aes(x=t, y=pred0.2, col = "Predicted"), na.rm = T,  -->
-<!--               linetype = "dashed")+ -->
-<!--     geom_vline(xintercept = 0.2, linetype = "dashed")+ -->
-<!--     labs(x="Time", y="")+ -->
-<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
-<!--     theme(axis.text = element_text(size = 5))+ -->
-<!--     ylim(-1.2, 1.7), -->
-
-<!--  pred_list_all[[1]] %>%  -->
-<!--     filter(id==501) %>% -->
-<!--     mutate(eta_i = ifelse(t<=0.4, eta_i, NA)) %>% -->
-<!--     ggplot()+ -->
-<!--     geom_line(aes(x=t, y=eta_i, col = "Observed"), na.rm = T)+ -->
-<!--     geom_line(aes(x=t, y=pred0.4, col = "Predicted"), na.rm = T,  -->
-<!--               linetype = "dashed")+ -->
-<!--     geom_vline(xintercept = 0.4, linetype = "dashed")+ -->
-<!--     labs(x="Time", y="")+ -->
-<!--     scale_x_continuous(breaks = seq(0, 1, by = 0.2))+ -->
-<!--     theme(axis.text = element_text(size = 5))+ -->
-<!--    ylim(-1.2, 1.7), -->
-<!--  nrow = 1, common.legend = T -->
-<!--   ) -->
-
-<!-- ``` -->

### 2024-05-28 Docs/manuscript.tex
presentation papers

@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \author{Ying Jin}
-\date{2024-02-16}
+\date{2024-05-28}
 
@@
 }
-\hypertarget{method}{%
-\section{Method}\label{method}}
-
-\hypertarget{assumptions}{%
-\subsection{Assumptions}\label{assumptions}}
+\section{Method}\label{method}
+
+\subsection{Assumptions}\label{assumptions}
 
@@
 
-\hypertarget{observed-data}{%
-\subsection{Observed data}\label{observed-data}}
+\subsection{Observed data}\label{observed-data}
 
@@
 
-\hypertarget{fgfpca-algorithm}{%
-\subsection{fGFPCA Algorithm}\label{fgfpca-algorithm}}
-
-\hypertarget{bin-data}{%
-\subsubsection{Bin data:}\label{bin-data}}
+\subsection{fGFPCA Algorithm}\label{fgfpca-algorithm}
+
+\subsubsection{Bin data:}\label{bin-data}
 
@@
 
-\hypertarget{local-glmms}{%
-\subsubsection{Local GLMMs}\label{local-glmms}}
+\subsubsection{Local GLMMs}\label{local-glmms}
 
@@
 
-\hypertarget{fpca}{%
-\subsubsection{FPCA}\label{fpca}}
+\subsubsection{FPCA}\label{fpca}
 
@@
 
-\hypertarget{projection-and-debias}{%
-\subsubsection{Projection and Debias}\label{projection-and-debias}}
+\subsubsection{Projection and Debias}\label{projection-and-debias}
 
@@
 
-\hypertarget{out-of-sample-prediction}{%
-\subsection{Out-of-sample prediction}\label{out-of-sample-prediction}}
+\subsection{Out-of-sample prediction}\label{out-of-sample-prediction}
 
@@
 
-\hypertarget{reference-method}{%
-\section{Reference method}\label{reference-method}}
-
-\hypertarget{glmmadaptive}{%
-\subsection{GLMMadaptive}\label{glmmadaptive}}
+\section{Reference method}\label{reference-method}
+
+\subsection{GLMMadaptive}\label{glmmadaptive}
 
@@
 
-\hypertarget{generalized-functional-on-scalar-regression}{%
 \subsection{Generalized functional-on-scalar
-regression}\label{generalized-functional-on-scalar-regression}}
+regression}\label{generalized-functional-on-scalar-regression}
 
@@
 
-\hypertarget{larger-scale-simulation}{%
-\section{Larger-scale simulation}\label{larger-scale-simulation}}
-
-\hypertarget{simulation-set-up}{%
-\subsection{Simulation set up}\label{simulation-set-up}}
+\section{Larger-scale simulation}\label{larger-scale-simulation}
+
+\subsection{Simulation set up}\label{simulation-set-up}
 
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\subsection{Figures}\label{figures}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/fig_large_sim-1.pdf}
-
-\hypertarget{ise-and-auc}{%
-\subsection{ISE and AUC}\label{ise-and-auc}}
+\includegraphics{manuscript_files/figure-latex/fig_large_sim_paper-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_list\_all[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{580}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_list\_ref[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{580}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{),}
+\NormalTok{  pred\_list\_gfosr\_l1[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{580}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_list\_gfosr\_l5[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{580}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4)}
+\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(eta\_i, pred0}\FloatTok{.2}\NormalTok{, pred0}\FloatTok{.4}\NormalTok{, pred0}\FloatTok{.6}\NormalTok{, pred0}\FloatTok{.8}\NormalTok{), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{), }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{na.rm=}\NormalTok{T, }\AttributeTok{linewidth =} \FloatTok{1.0}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"t"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.2}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols, }\AttributeTok{guide=}\StringTok{"none"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{manuscript_files/figure-latex/fig_large_sim_pre-1.pdf}
+
+\subsection{ISE and AUC}\label{ise-and-auc}
 
@@
 
-\begin{landscape}\begin{table}
-\centering
-\begin{tabular}{lrrrrrrrrrrrrrrrr}
+\begin{landscape}
+\begin{longtable}[t]{lrrrrrrrrrrrrrrrr}
 \toprule
@@
 \multicolumn{17}{l}{\textbf{ISE}}\\
-\hspace{1em}(0.2, 0.4] & 146.407 &  &  &  & 274.947 &  &  &  & 362.476 &  &  &  & 387.708 &  &  & \\
-\hspace{1em}(0.4, 0.6] & 183.967 & 74.977 &  &  & 277.438 & 220.209 &  &  & 286.614 & 262.552 &  &  & 291.579 & 269.799 &  & \\
-\hspace{1em}(0.6, 0.8] & 218.265 & 49.275 & 15.776 &  & 322.338 & 373.108 & 325.017 &  & 385.701 & 410.508 & 389.305 &  & 315.778 & 282.736 & 278.242 & \\
-\hspace{1em}(0.8, 1.0] & 108.918 & 77.981 & 17.747 & 12.005 & 290.982 & 318.079 & 350.850 & 333.580 & 328.482 & 341.274 & 354.211 & 347.067 & 563.011 & 477.485 & 597.746 & 600.340\\
+\hspace{1em}(0.2, 0.4] & 146.405 &  &  &  & 274.947 &  &  &  & 362.476 &  &  &  & 387.708 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 183.965 & 74.979 &  &  & 277.438 & 220.209 &  &  & 286.614 & 262.552 &  &  & 291.579 & 269.799 &  & \\
+\hspace{1em}(0.6, 0.8] & 218.264 & 49.287 & 15.788 &  & 322.338 & 373.108 & 325.017 &  & 385.701 & 410.508 & 389.305 &  & 315.778 & 282.736 & 278.242 & \\
+\hspace{1em}(0.8, 1.0] & 108.914 & 77.985 & 17.759 & 12.018 & 290.982 & 318.079 & 350.850 & 333.580 & 328.482 & 341.274 & 354.211 & 347.067 & 563.011 & 477.485 & 597.746 & 600.340\\
 \addlinespace[0.3em]
@@
 \bottomrule
-\end{tabular}
-\end{table}
+\end{longtable}
 \end{landscape}
 
-\hypertarget{time-minutes}{%
-\subsection{Time (minutes)}\label{time-minutes}}
-
-I chose not to report fitting and prediction time separately because it
-is too difficult to tell them apart for the GFOSR models.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{data.frame}\NormalTok{(}
-  \StringTok{"Method"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GLMMadaptive"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{),}
-  \StringTok{"Time"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time),}
-             \FunctionTok{mean}\NormalTok{(fit\_time\_ref}\SpecialCharTok{+}\NormalTok{pred\_time\_ref}\SpecialCharTok{/}\DecValTok{60}\NormalTok{),}
-             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l1}\SpecialCharTok{/}\DecValTok{60}\NormalTok{), }
-             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs =}\NormalTok{ T, }\AttributeTok{table.attr =} \StringTok{"style = }\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{full\_width =}\NormalTok{ F) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-\centering
-\begin{tabular}{lr}
-\toprule
-Method & Time\\
-\midrule
-fGFPCA & 2.317\\
-GLMMadaptive & 2.304\\
-GFOSR (L=1) & 0.025\\
-GFOSR (L=5) & 0.144\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{small-scale-simulation}{%
-\section{Small-scale simulation}\label{small-scale-simulation}}
+\section{Small-scale simulation}\label{small-scale-simulation}
 
@@
 
-\hypertarget{figure}{%
-\subsection{Figure}\label{figure}}
+\subsection{Figure}\label{figure}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/fig_small_sim-1.pdf}
-
-\hypertarget{ise-and-auc-1}{%
-\subsection{ISE and AUC}\label{ise-and-auc-1}}
+\includegraphics{manuscript_files/figure-latex/fig_small_sim_paper-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{sep\_plot }\OtherTok{\textless{}{-}} \FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_subset\_fGFPCA[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{189}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_subset\_gfosr\_l5[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{189}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_subset\_gfosr\_l1[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{189}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+    \FunctionTok{rename}\NormalTok{(}\AttributeTok{pred0.2=}\NormalTok{pred\_w1, }\AttributeTok{pred0.4=}\NormalTok{pred\_w2, }\AttributeTok{pred0.6=}\NormalTok{pred\_w3, }\AttributeTok{pred0.8=}\NormalTok{pred\_w4),}
+\NormalTok{  pred\_subset\_adglmm[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id}\SpecialCharTok{==}\DecValTok{189}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(eta\_i, pred0}\FloatTok{.2}\NormalTok{, pred0}\FloatTok{.4}\NormalTok{, pred0}\FloatTok{.6}\NormalTok{, pred0}\FloatTok{.8}\NormalTok{), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{eta\_i, }\AttributeTok{col =} \StringTok{"True"}\NormalTok{), }\AttributeTok{linetype =} \StringTok{"dashed"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.2}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.2"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.4}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.4"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.6}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.6"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{t, }\AttributeTok{y=}\NormalTok{pred0}\FloatTok{.8}\NormalTok{, }\AttributeTok{col =} \StringTok{"0.8"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"t"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.2}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{sep\_plot}\SpecialCharTok{+}\FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols, }\AttributeTok{guide=}\StringTok{"none"}\NormalTok{)}\SpecialCharTok{+}
+  \CommentTok{\#theme(legend.position = "bottom")+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols, }\AttributeTok{guide =} \StringTok{"none"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Scale for colour is already present.
+## Adding another scale for colour, which will replace the existing scale.
+\end{verbatim}
+
+\includegraphics{manuscript_files/figure-latex/fig_small_sim_pre-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{as\_ggplot}\NormalTok{(ggpubr}\SpecialCharTok{::}\FunctionTok{get\_legend}\NormalTok{(sep\_plot}\SpecialCharTok{+}\FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}\SpecialCharTok{+}
+                               \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols)))}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{manuscript_files/figure-latex/legend-1.pdf}
+
+\subsection{ISE and AUC}\label{ise-and-auc-1}
 
@@
 \multicolumn{17}{l}{\textbf{ISE}}\\
-\hspace{1em}(0.2, 0.4] & 150.641 &  &  &  & 286.853 &  &  &  & 367.282 &  &  &  & 374.836 &  &  & \\
-\hspace{1em}(0.4, 0.6] & 188.065 & 77.380 &  &  & 286.628 & 227.225 &  &  & 290.781 & 265.809 &  &  & 268.390 & 463.337 &  & \\
-\hspace{1em}(0.6, 0.8] & 224.001 & 51.533 & 16.879 &  & 328.612 & 374.572 & 326.522 &  & 386.705 & 408.427 & 388.427 &  & 319.731 & 287.001 & 233.510 & \\
-\hspace{1em}(0.8, 1.0] & 112.163 & 81.368 & 19.592 & 13.399 & 301.165 & 324.817 & 353.888 & 335.139 & 331.488 & 342.935 & 354.800 & 347.627 & 228.335 & 469.769 & 331.509 & 131.993\\
+\hspace{1em}(0.2, 0.4] & 150.636 &  &  &  & 286.853 &  &  &  & 367.282 &  &  &  & 374.836 &  &  & \\
+\hspace{1em}(0.4, 0.6] & 188.095 & 77.401 &  &  & 286.628 & 227.225 &  &  & 290.781 & 265.809 &  &  & 268.390 & 463.337 &  & \\
+\hspace{1em}(0.6, 0.8] & 224.067 & 51.634 & 16.969 &  & 328.612 & 374.572 & 326.522 &  & 386.705 & 408.427 & 388.427 &  & 319.731 & 287.001 & 233.510 & \\
+\hspace{1em}(0.8, 1.0] & 112.190 & 81.407 & 19.701 & 13.485 & 301.165 & 324.817 & 353.888 & 335.139 & 331.488 & 342.935 & 354.800 & 347.627 & 228.335 & 469.769 & 331.509 & 131.993\\
 \addlinespace[0.3em]
@@
 \bottomrule
-\end{tabular}
-\end{table}
+\end{longtable}
 \end{landscape}
@@
 
-\hypertarget{time}{%
-\section{Time}\label{time}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{data.frame}\NormalTok{(}
-  \StringTok{"Method"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GLMMadaptive"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{),}
-  \StringTok{"Time"} \OtherTok{=} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time\_subset\_fGFPCA}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time\_subset\_fGFPCA),}
+\section{Time}\label{time}
+
+Average time per iteration in minutes
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{time\_tb }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{sim1 =} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time),}
+             \FunctionTok{mean}\NormalTok{(fit\_time\_ref}\SpecialCharTok{+}\NormalTok{pred\_time\_ref}\SpecialCharTok{/}\DecValTok{60}\NormalTok{),}
+             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l1}\SpecialCharTok{/}\DecValTok{60}\NormalTok{), }
+             \FunctionTok{mean}\NormalTok{(t\_vec\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{)),}
+           \AttributeTok{sim2 =} \FunctionTok{c}\NormalTok{(}\FunctionTok{mean}\NormalTok{(fit\_time\_subset\_fGFPCA}\SpecialCharTok{/}\DecValTok{60}\SpecialCharTok{+}\NormalTok{pred\_time\_subset\_fGFPCA),}
              \FunctionTok{mean}\NormalTok{(fit\_time\_subset\_adglmm}\SpecialCharTok{+}\NormalTok{pred\_time\_subset\_adglmm}\SpecialCharTok{/}\DecValTok{60}\NormalTok{),}
              \FunctionTok{mean}\NormalTok{(t\_subset\_gfosr\_l1}\SpecialCharTok{/}\DecValTok{60}\NormalTok{), }
-             \FunctionTok{mean}\NormalTok{(t\_subset\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+             \FunctionTok{mean}\NormalTok{(t\_subset\_gfosr\_l5}\SpecialCharTok{/}\DecValTok{60}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{t}\NormalTok{()}
+\FunctionTok{colnames}\NormalTok{(time\_tb) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GLMMadaptive"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{)}
+\FunctionTok{rownames}\NormalTok{(time\_tb) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Simulation 1"}\NormalTok{, }\StringTok{"Simulation 2"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{time\_tb }\SpecialCharTok{\%\textgreater{}\%} 
   \FunctionTok{kable}\NormalTok{(}\AttributeTok{digits =} \DecValTok{3}\NormalTok{, }\AttributeTok{booktabs =}\NormalTok{ T, }\AttributeTok{table.attr =} \StringTok{"style = }\SpecialCharTok{\textbackslash{}"}\StringTok{color:black;}\SpecialCharTok{\textbackslash{}"}\StringTok{"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
@@
 
-\begin{table}
-\centering
-\begin{tabular}{lr}
+\begin{longtable}[t]{lrrrr}
 \toprule
-Method & Time\\
+ & fGFPCA & GLMMadaptive & GFOSR (L=1) & GFOSR (L=5)\\
 \midrule
-fGFPCA & 1.629\\
-GLMMadaptive & 5.479\\
-GFOSR (L=1) & 0.010\\
-GFOSR (L=5) & 0.054\\
+Simulation 1 & 2.584 & 2.304 & 0.025 & 0.144\\
+Simulation 2 & 2.415 & 5.479 & 0.010 & 0.054\\
 \bottomrule
-\end{tabular}
-\end{table}
-
-\hypertarget{nhanes-data-application}{%
-\section{NHANES data application}\label{nhanes-data-application}}
+\end{longtable}
+
+\section{NHANES data application}\label{nhanes-data-application}
 
@@
 
-\hypertarget{auc}{%
-\subsection{AUC}\label{auc}}
+\subsection{AUC}\label{auc}
 
@@
 
-\begin{landscape}\begin{table}
-\centering
-\begin{tabular}{lrrrrrrrrrrrr}
+\begin{landscape}
+\begin{longtable}[t]{lrrrrrrrrrrrr}
 \toprule
@@
 \midrule
-6am-12pm & 0.703 &  &  & 0.689 &  &  & 0.684 &  &  & 0.581 &  & \\
-12pm-6pm & 0.536 & 0.709 &  & 0.519 & 0.699 &  & 0.520 & 0.654 &  & 0.532 & 0.701 & \\
-6pm-12am & 0.716 & 0.679 & 0.773 & 0.677 & 0.678 & 0.723 & 0.677 & 0.676 & 0.708 & 0.514 & 0.565 & 0.626\\
+6am-12pm & 0.701 &  &  & 0.689 &  &  & 0.684 &  &  & 0.581 &  & \\
+12pm-6pm & 0.530 & 0.708 &  & 0.519 & 0.699 &  & 0.520 & 0.654 &  & 0.532 & 0.701 & \\
+6pm-12am & 0.708 & 0.679 & 0.774 & 0.677 & 0.678 & 0.723 & 0.677 & 0.676 & 0.708 & 0.514 & 0.565 & 0.626\\
 \bottomrule
-\end{tabular}
-\end{table}
+\end{longtable}
 \end{landscape}
 
-\hypertarget{figure-1}{%
-\subsection{Figure}\label{figure-1}}
+\subsection{Figure}\label{figure-1}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/fig_appl-1.pdf}
+\includegraphics{manuscript_files/figure-latex/fig_appl_paper-1.pdf}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# selct four subjects}
+\NormalTok{plot\_id }\OtherTok{\textless{}{-}} \FunctionTok{sort}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\DecValTok{64198}\NormalTok{, }\DecValTok{80517}\NormalTok{))}
+
+\CommentTok{\# figure}
+\FunctionTok{bind\_rows}\NormalTok{(}
+\NormalTok{  pred\_nhanes\_fgfpca }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\StringTok{"fGFPCA"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=5)"}\NormalTok{),}
+\NormalTok{  pred\_appl\_gfosr\_l1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GFOSR (L=1)"}\NormalTok{),}
+\NormalTok{  pred\_nhanes\_adglmm }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ plot\_id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method =} \StringTok{"GLMMadaptive"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate\_at}\NormalTok{(}\FunctionTok{vars}\NormalTok{(pred360, pred720, pred1080), }
+                \AttributeTok{.funs =} \ControlFlowTok{function}\NormalTok{(x)\{}\FunctionTok{exp}\NormalTok{(x)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\FunctionTok{exp}\NormalTok{(x))\}) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{method=}\FunctionTok{factor}\NormalTok{(method, }
+         \AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"fGFPCA"}\NormalTok{, }\StringTok{"GFOSR (L=5)"}\NormalTok{, }\StringTok{"GFOSR (L=1)"}\NormalTok{,}\StringTok{"GLMMadaptive"}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{Y), }\AttributeTok{size =} \FloatTok{0.2}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred360, }\AttributeTok{col =} \StringTok{"6am"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred720, }\AttributeTok{col =} \StringTok{"12pm"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{geom\_line}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{sind, }\AttributeTok{y=}\NormalTok{pred1080, }\AttributeTok{col =} \StringTok{"6pm"}\NormalTok{), }\AttributeTok{linewidth=}\DecValTok{1}\NormalTok{, }\AttributeTok{na.rm=}\NormalTok{T)}\SpecialCharTok{+}
+  \FunctionTok{facet\_grid}\NormalTok{(}\AttributeTok{rows =} \FunctionTok{vars}\NormalTok{(id), }\AttributeTok{cols =} \FunctionTok{vars}\NormalTok{(method))}\SpecialCharTok{+}
+  \FunctionTok{labs}\NormalTok{(}\AttributeTok{col =} \StringTok{"Maximum observation time"}\NormalTok{, }\AttributeTok{x =} \StringTok{"Time"}\NormalTok{, }\AttributeTok{y=}\StringTok{""}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1440}\NormalTok{, }\AttributeTok{by =} \DecValTok{360}\NormalTok{),}
+                     \AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"0am"}\NormalTok{, }\StringTok{"6am"}\NormalTok{, }\StringTok{"12pm"}\NormalTok{, }\StringTok{"6pm"}\NormalTok{, }\StringTok{"12am"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{, }\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{size =} \DecValTok{7}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =}\NormalTok{ cols)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{manuscript_files/figure-latex/fig_appl_pre-1.pdf}
 
@@
 
-\hypertarget{additional-figures}{%
-\section{Additional figures}\label{additional-figures}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{df\_nhanes }\OtherTok{\textless{}{-}} \FunctionTok{read\_rds}\NormalTok{(}\FunctionTok{here}\NormalTok{(}\StringTok{"Data/nhanes\_bi.rds"}\NormalTok{))}
-\NormalTok{df\_nhanes }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{select}\NormalTok{(SEQN, Z, sind) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(SEQN }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{filter}\NormalTok{(sind }\SpecialCharTok{\%in\%} \DecValTok{356}\SpecialCharTok{:}\DecValTok{360}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ sind, }\AttributeTok{values\_from =}\NormalTok{ Z) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{SEQN) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{distinct}\NormalTok{(.)}
-
-\NormalTok{pred\_appl\_gfosr\_l5 }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{filter}\NormalTok{(id }\SpecialCharTok{\%in\%}\NormalTok{ rand\_id) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{select}\NormalTok{(id, sind, pred360) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ sind, }\AttributeTok{values\_from =}\NormalTok{ pred360) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{id) }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{distinct}\NormalTok{(.)}
-\end{Highlighting}
-\end{Shaded}
-
 \end{document}

### 2024-05-28 Code/Figures.R

@@ -1,0 +1,21 @@
+
+# This script saves the code for some additional figures
+# produced for presentation or reports
+
+#### NHANES example ####
+
+# 2011-2014 binary
+
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+df_nhanes %>% 
+  select(SEQN, Z, sind) %>% 
+  filter(SEQN %in% sample(unique(df_nhanes), size = 100)) %>% head()
+  pivot_wider(names_from = sind, values_from = Z) %>% 
+  select(-SEQN) %>% distinct(.)
+
+
+pred_appl_gfosr_l5 %>% 
+  filter(id %in% rand_id) %>%
+  select(id, sind, pred360) %>%
+  pivot_wider(names_from = sind, values_from = pred360) %>% 
+  select(-id) %>% distinct(.)

### 2024-05-28 Code/GaussianMoM.R

@@ -1,183 +1,0 @@
-library(fcr)
-library(ggplot2)
-library(tidyverse)
-
-# https://cran.r-project.org/web/packages/fcr/vignettes/dynamic-prediction.html
-
-
-#### Data ####
-df = content
-df_train = subset(content, include == 1)
-df_test = subset(content, include == 0)
-
-ggplot(df)+
-  geom_line(aes(x = argvals, y = Y, col = as.factor(subj)), show.legend = F)
-
-
-
-#### functional concurrent regression ####
-
-k = 12
-
-# functional domain
-tnew = unique(df$argvals)
-tnew = sort(tnew)
-
-# smoothing time-varying covariate: measured with error
-ggplot(df_train)+
-  geom_line(aes(x = argvals, y = waz.true, col = as.factor(subj)), show.legend = F)
-
-df_waz <- data.frame("y" = df_train$waz, "subj" = df_train$subj, "argvals" = df_train$argvals)
-fit_waz <- face.sparse(df_waz, newdata = df_waz, knots = k, argvals.new = tnew) # not 
-## mean function
-plot(tnew, fit_waz$mu.new)
-## covariance function
-heatmap(fit_waz$Cor.new)
-heatmap(fit_waz$Cor.raw.new)
-## eigenfunctions
-fit_waz$eigenvalues
-plot(tnew, fit_waz$eigenfunctions[, 1])
-plot(tnew, fit_waz$eigenfunctions[, 2])
-plot(tnew, fit_waz$eigenfunctions[, 3])
-plot(tnew, fit_waz$eigenfunctions[, 4])
-## smoothed value 
-df_train$sm_waz <- fit_waz$y.pred
-
-
-# fit FCR 
-K2 <- 15
-
-fit_fcr <- fcr(formula = Y ~ s(argvals, k=K2, bs="ps") + s(argvals, by=Male, k=K2, bs="ps") + s(argvals, by=sm_waz, k=K2, bs="ps"), 
-               argvals = "argvals", subj = "subj", data = df_train, use_bam = T, argvals.new = tnew,
-               face.args = list(knots = K2, pve = 0.99))
-## plot coefficients
-par(mfrow = c(1, 1))
-plot(fit_fcr,select=3)
-## plot covariance function
-plot(fit_fcr, plot.covariance=TRUE) # eigen functions different from tutorial
-
-#### Prediction #####
-# in-sample: for subject used in the model fitting
[…]

### 2024-06-25 Docs/Presentation.Rmd
meeting notes on the manuscript

@@
 
+```{r load_nhanes}
+# binary
+df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
+
+# continuous
+df_nhanes_cont <- read_rds(here("Data/NHANES_AC_processed.rds"))
+df_exp <- df_nhanes_cont %>%
+  filter(SEQN == 21910 | SEQN == 29599) %>%
+  filter(WEEKDAY==5) %>% 
+  group_by(Gender) %>%
+  sample_n(1) %>%
+  select(SEQN, Gender, starts_with("MIN")) %>%
+  pivot_longer(starts_with("MIN")) %>%
+  mutate(minute = as.numeric(gsub("MIN", "", name))) %>%
+  mutate(value_bi = value > 100)
+
+```
+
+```{r cont_func_exp, fig.height=4, fig.width=6}
+df_exp %>%
+  ggplot(aes(x=minute, y=value, col=Gender))+
+  geom_point(size = 0.5, alpha = 0.5)+
+  geom_smooth(method = "loess", formula = y ~ x, show.legend = F)+
+  # scale_y_continuous(transform = "log")+
+  scale_color_manual(values = c("#FC8D62", "#8DA0CB"))+
+  scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  labs(y = "Activity count", x="")
+```
+
+```{r gen_func_exp, fig.height=4, fig.width=6}
+df_exp %>%
+  filter(value_bi) %>% 
+  ggplot(aes(x=minute, y=Gender, col=Gender))+
+  geom_point(size = 0.5, alpha = 0.5, show.legend = F)+
+  # scale_y_continuous(transform = "log")+
+  scale_color_manual(values = c("#FC8D62", "#8DA0CB"))+
+  scale_x_continuous(breaks = seq(0, 1440, by = 360),
+                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
+  labs(y = "Activity indicator", x="")
+```
+
 ```{r nhanes_exp, fig.width=8, fig.height=5}
-df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
 df_nhanes %>% 
@@
 
-```{r load_nhanes}
+```{r load_nhanes_result}
 # df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))

### 2024-07-19 Docs/manuscript.Rmd
figures with prediction interval

@@
     keep_tex: yes
+  pdf_document:
+    toc: yes
+    toc_depth: '3'
+    keep_tex: yes
 ---
@@
 
-set.seed(1120)
+set.seed(1114)
 
@@
 
+# code 
+source(here("Code/PredEval.R"))
+
 ## prediction window 
 window <- seq(0, 1, by = 0.2)
-M <- 500
+M <- 5
 
@@
 ```{r load_sim1}
-load(here("Data/SimOutput_fGFPCA.RData"))
-load(here("Data/SimOutput_GLMMadaptive.RData"))
-load(here("Data/SimOutput_GFOSR_L1.RData"))
-pred_list_gfosr_l1 <- pred_list_gfofr
-t_vec_gfosr_l1 <- t_vec_gfofr
-rm(pred_list_gfofr, t_vec_gfofr)
-load(here("Data/SimOutput_GFOSR_L5.RData"))
-pred_list_gfosr_l5 <- pred_list_gfofr
-t_vec_gfosr_l5 <- t_vec_gfofr
-rm(pred_list_gfofr, t_vec_gfofr)
+load(here("Data/TrialRun/SimOutput_fGFPCA.RData"))
+load(here("Data/TrialRun/SimOutput_GLMMadaptive.RData"))
+load(here("Data/TrialRun/SimOutput_GFOSR_L1.RData"))
+load(here("Data/TrialRun/SimOutput_GFOSR_L5.RData"))
 
@@
 
-```{r fig_large_sim_paper, fig.height=8, fig.width=8}
+```{r DataFormat}
+pred_list_fGFPCA <- lapply(pred_list_fGFPCA, function(x){x %>% select(-bin)})
+name_vec <- colnames(pred_list_fGFPCA[[1]])
+pred_list_fGFPCA <- lapply(pred_list_fGFPCA, 
+                           function(x){
+                      x[x$t<=0.2, c("pred0.2", "pred0.2_lb", "pred0.2_ub")] <- NA
+                      x[x$t<=0.4, c("pred0.4", "pred0.4_lb", "pred0.4_ub")] <- NA
+                      x[x$t<=0.6, c("pred0.6", "pred0.6_lb", "pred0.6_ub")] <- NA
+                      x[x$t<=0.8, c("pred0.8", "pred0.8_lb", "pred0.8_ub")] <- NA
+                      return(x)    
+                       })
+
+pred_list_GLMMad <- lapply(pred_list_GLMMad, function(x){
+  colnames(x) <- name_vec
+  return(x)
+})
+
+pred_list_gfofr <- lapply(pred_list_gfofr, function(x){
+  x <- x %>% select(-window)
+  colnames(x) <- name_vec
+  return(x)
+})
+
+pred_list_gfofr_l5 <- lapply(pred_list_gfofr_l5, function(x){
+  x <- x %>% select(-window)
+  colnames(x) <- name_vec
+  return(x)
+})
+
+```
+
+```{r large_sim_plot, fig.height=8, fig.width=8}
 bind_rows(
-  pred_list_all[[1]] %>% filter(id %in% rand_id) %>% 
+  pred_list_fGFPCA[[1]] %>% filter(id %in% rand_id) %>% 
     mutate(method="fGFPCA"),
-  pred_list_ref[[1]] %>% filter(id %in% rand_id) %>%
+  pred_list_GLMMad[[1]] %>% filter(id %in% rand_id) %>%
     mutate(method = "GLMMadaptive"),
-  pred_list_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=1)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_list_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=5)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4)
-) %>% mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
+  pred_list_gfofr[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=1)"),
+  pred_list_gfofr_l5[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=5)") 
+) %>% 
+  mutate_at(vars(eta_i, starts_with("pred")), 
                 .funs = function(x){exp(x)/(1+exp(x))}) %>%
@@
   geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm=T, linewidth = 1.0)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm=T, linewidth = 1.0)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm=T, linewidth = 1.0)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm=T, linewidth = 1.0)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm = T)+
+  geom_ribbon(aes(x=t, ymin=pred0.2_lb, ymax = pred0.2_ub,
+                  col = "0.2", fill = "0.2", alpha = 0.1),
+              linetype="dashed")+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm = T)+
+  geom_ribbon(aes(x=t, ymin=pred0.4_lb, ymax = pred0.4_ub,
+                  col = "0.4", fill = "0.4", alpha = 0.1),
+              linetype="dashed")+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm = T)+
+  geom_ribbon(aes(x=t, ymin=pred0.6_lb, ymax = pred0.6_ub,
+                  col = "0.6", fill = "0.6", alpha = 0.1),
+              linetype="dashed")+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm = T)+
+  geom_ribbon(aes(x=t, ymin=pred0.8_lb, ymax = pred0.8_ub,
+                  col = "0.8", fill = "0.8", alpha = 0.1),
+              linetype="dashed")+
+  guides(alpha = "none", col="none")+
   facet_grid(rows = vars(id), cols = vars(method))+
@@
 
-## ISE and AUC
+## Preformance
 
-```{r ise}
-## ISE container 
-ise_fgfpca <- ise_adglmm <- ise_gfosr_l1 <- ise_gfosr_l5 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-# dims: prediction window, max obs time, simulation iter
+Please note that here, the GLMM has a **square** time for both fixed effect and random effect.
 
+### Time (minutes)
+
+```{r}
+data.frame(Method = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"), 
+           Time = c(mean(time_fGFPCA),
+                    mean(time_gfofr_l5),
+                    mean(time_gfofr),
+                    mean(time_GLMMad))) %>%
+  kable(digits = 3, booktabs=T,
+        table.attr="style=\"color:black;\"") %>%
+  kable_styling(full_width = F)
 ```
 
-```{r ise_fGFPCA}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_all[[m]]
-  ise_tb <- pred_list_all[[m]] %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err"))
-  ise_fgfpca[, ,m] <- as.matrix(ise_tb)
-
-}
-
-mean_ise_fgfpca <- apply(ise_fgfpca, c(1, 2), mean)
-
-# mean_ise <- data.frame(mean_ise) %>%
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_ise_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_adglmm}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_gfosr_l1[[m]]
-  ise_tb <- pred_list_ref[[m]] %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err"))
-  ise_adglmm[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_adglmm <- apply(ise_adglmm, c(1, 2), mean)
-colnames(mean_ise_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l1}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_ref[[m]]
-  ise_tb <- pred_list_gfosr_l1[[m]] %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    # mutate(window = factor(window, levels = 1:5, 
-    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err"))
-  ise_gfosr_l1[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_gfosr_l1 <- apply(ise_gfosr_l1, c(1, 2), mean)
-colnames(mean_ise_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l5}
-## calculation
-for(m in 1:M){
-  # this_df <- pred_list_ref[[m]]
-  ise_tb <- pred_list_gfosr_l5[[m]] %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    # mutate(window = factor(window, levels = 1:5, 
-    #                        labels = c("[0,0.2]", "(0.2,0.4]", "(0.4,0.6]",))) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err"))
-  ise_gfosr_l5[, ,m] <- as.matrix(ise_tb)
-}
-
-mean_ise_gfosr_l5 <- apply(ise_gfosr_l5, c(1, 2), mean)
-colnames(mean_ise_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_func}
-## a function to calculate AUC
-get_auc <- function(y, pred){
-  if(sum(is.na(y))>0 | sum(is.na(pred))>0){
-    auc <- NA
-  }
-  else{
-    this_perf <- performance(prediction(pred, y), measure = "auc")
-    auc <- this_perf@y.values[[1]]
-  }
-  return(auc)
-}
-
-```
+### ISE
 
 ```{r}
-## auc container 
-auc_fgfpca <- auc_adglmm <- auc_gfosr_l1 <- auc_gfosr_l5 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-```
-
-```{r auc_fGFPCA}
-for(m in 1:M){
-  this_df <- pred_list_all[[m]]
-  auc_tb <- this_df %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred0.2),
-              auc2 = get_auc(Y, pred0.4),
-              auc3 = get_auc(Y, pred0.6),
-              auc4 = get_auc(Y, pred0.8)) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("auc"))
-  auc_fgfpca[, ,m] <- as.matrix(auc_tb)
-}
-
-mean_auc_fgfpca <- apply(auc_fgfpca, c(1, 2), mean)
-# mean_auc <- data.frame(mean_auc) %>% 
-#   mutate(Window = c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"),
-#          .before = 1)
-colnames(mean_auc_fgfpca) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GLMMadaptive}
-for(m in 1:M){
-  this_df <- pred_list_ref[[m]]
-  auc_tb <- this_df %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred0.2),
-              auc2 = get_auc(Y, pred0.4),
-              auc3 = get_auc(Y, pred0.6),
-              auc4 = get_auc(Y, pred0.8)) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("auc"))
-  auc_adglmm[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_adglmm <- apply(auc_adglmm, c(1, 2), mean)
-colnames(mean_auc_adglmm) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_gfosr_l1}
-for(m in 1:M){
-  this_df <- pred_list_gfosr_l1[[m]]
-  auc_tb <- this_df %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred_w1),
-              auc2 = get_auc(Y, pred_w2),
-              auc3 = get_auc(Y, pred_w3),
-              auc4 = get_auc(Y, pred_w4)) %>%
-    filter(window != 1) %>% 
-    select(starts_with("auc"))
-  auc_gfosr_l1[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_gfosr_l1 <- apply(auc_gfosr_l1, c(1, 2), mean)
-colnames(mean_auc_gfosr_l1) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_gfosr_l5}
-for(m in 1:M){
-  this_df <- pred_list_gfosr_l5[[m]]
-  auc_tb <- this_df %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred_w1),
-              auc2 = get_auc(Y, pred_w2),
-              auc3 = get_auc(Y, pred_w3),
-              auc4 = get_auc(Y, pred_w4)) %>%
-    filter(window != 1) %>% 
-    select(starts_with("auc"))
-  auc_gfosr_l5[, ,m] <- as.matrix(auc_tb)
-
-}
-
-mean_auc_gfosr_l5 <- apply(auc_gfosr_l5, c(1, 2), mean)
-colnames(mean_auc_gfosr_l5) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r table_sim1}
-fgfpca1 <- rbind(mean_ise_fgfpca ,mean_auc_fgfpca) %>% as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
-gfosr_l5_1 <- rbind(mean_ise_gfosr_l5 ,mean_auc_gfosr_l5) %>% as_tibble()
-gfosr_l1_1 <- rbind(mean_ise_gfosr_l1 ,mean_auc_gfosr_l1) %>% as_tibble()
-adglmm1 <- rbind(mean_ise_adglmm ,mean_auc_adglmm) %>% as_tibble()
-
-bind_cols(fgfpca1, gfosr_l5_1, gfosr_l1_1, adglmm1, .name_repair = "minimal") %>%
+cbind(calc_ISE(pred_list_fGFPCA, window),
+      calc_ISE(pred_list_gfofr, window),
+      calc_ISE(pred_list_gfofr_l5, window),
+      calc_ISE(pred_list_GLMMad, window)) %>%
+  data.frame(window = c("[0, 0.2]", "(0,2, 0.4]", "(0.4, 0.6]", 
+                    "(0.6, 0.8]","(0.8, 1]"),
+             ., 
+             check.names = F) %>%
   kable(digits = 3, booktabs=T,
@@
   add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
-  group_rows(index = c("ISE"= 4, "AUC" = 4)) %>%
   landscape()
-
 ```
 
-# Small-scale simulation
+It looks like the predictive performance can be seriously worse if the prediction window is far from the observation window.
 
-I have used 100 subjects for training and testing, and repeated 500 times.
-When fitting GLMMadpative, I reduce the number of measurements in the training dataset to 1/10 by taking one every 10 observations.
-The prediction is on the original grid.
-
-```{r load_sim2}
-load(here("Data/SubSimOutput_GLMMadaptive.RData"))
-
-load(here("Data/SubSimOutput_fGFPCA.RData"))
-pred_subset_adglmm <- pred_subset_adglmm[!num_probs]
-fit_time_subset_adglmm <- fit_time_subset_adglmm[!num_probs]
-pred_time_subset_adglmm <- pred_time_subset_adglmm[!num_probs]
-
-load(here("Data/SubSimOutput_GFOSR_L1.RData"))
-pred_subset_gfosr_l1 <- pred_list_gfofr_subset
-t_subset_gfosr_l1 <- t_vec_gfofr_subset
-rm(pred_list_gfofr_subset, t_vec_gfofr_subset)
-
-load(here("Data/SubSimOutput_GFOSR_L5.RData"))
-pred_subset_gfosr_l5 <- pred_list_gfofr_subset
-t_subset_gfosr_l5 <- t_vec_gfofr_subset
-rm(pred_list_gfofr_subset, t_vec_gfofr_subset)
-
-rand_id <- sample(101:200, size = 4)
-```
-
-## Figure
-
-```{r fig_small_sim_paper, fig.height=8, fig.width=8}
-bind_rows(
-  pred_subset_fGFPCA[[1]] %>% filter(id %in% rand_id) %>% mutate(method="fGFPCA"),
-  pred_subset_gfosr_l5[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=5)") %>% 
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_gfosr_l1[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GFOSR (L=1)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_adglmm[[1]] %>% filter(id %in% rand_id) %>% mutate(method = "GLMMadaptive")) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
-                .funs = function(x){exp(x)/(1+exp(x))}) %>%
-  mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linewidth=1, na.rm=T)+
-  facet_grid(rows = vars(id), cols = vars(method))+
-  labs(col = "Maximum observation time", x = "Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
-  theme(legend.position = "bottom")+
-  scale_color_manual(values = cols)
-```
+### AUC
 
 ```{r}
-sep_plot <- bind_rows(
-  pred_subset_fGFPCA[[1]] %>% filter(id==189) %>% mutate(method="fGFPCA"),
-  pred_subset_gfosr_l5[[1]] %>% filter(id==189) %>% mutate(method = "GFOSR (L=5)") %>% 
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_gfosr_l1[[1]] %>% filter(id==189) %>% mutate(method = "GFOSR (L=1)") %>%
-    rename(pred0.2=pred_w1, pred0.4=pred_w2, pred0.6=pred_w3, pred0.8=pred_w4),
-  pred_subset_adglmm[[1]] %>% filter(id==189) %>% mutate(method = "GLMMadaptive")) %>%
-  mutate_at(vars(eta_i, pred0.2, pred0.4, pred0.6, pred0.8), 
-                .funs = function(x){exp(x)/(1+exp(x))}) %>%
-  mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
-  ggplot()+
-  geom_point(aes(x=t, y=Y), size = 0.2)+
-  geom_line(aes(x=t, y=eta_i, col = "True"), linetype = "dashed")+
-  geom_line(aes(x=t, y=pred0.2, col = "0.2"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), linewidth=1, na.rm=T)+
-  facet_grid(cols = vars(method))+
-  labs(col = "Maximum observation time", x = "t", y="")+
-  scale_x_continuous(breaks = seq(0, 1, by = 0.2))
-```
-
-```{r, fig_small_sim_pre, fig.height=2, fig.width=8}
-sep_plot+scale_color_manual(values = cols, guide="none")+
-  #theme(legend.position = "bottom")+
-  scale_color_manual(values = cols, guide = "none")
-```
-
-```{r, legend, fig.height=0.2, fig.width=8}
-as_ggplot(ggpubr::get_legend(sep_plot+theme(legend.position = "bottom")+
-                               scale_color_manual(values = cols)))
-```
-
-## ISE and AUC
-
-```{r ise_subset}
-## ISE container 
-ise_fgfpca2 <- ise_adglmm2 <- ise_gfosr_l1_2 <- ise_gfosr_l5_2 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-# dims: prediction window, max obs time, simulation iter
-
-```
-
-```{r ise_fGFPCA_subset}
-for(m in 1:length(pred_subset_fGFPCA)){
-  this_df <- pred_subset_fGFPCA[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_fgfpca2[,,m] <- ise_tb_m
-}
-
-mean_ise_fgfpca2 <- apply(ise_fgfpca2, c(1, 2), mean)
-
-colnames(mean_ise_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_GLMMadaptive_subset}
-for(m in 1:length(pred_subset_adglmm)){
-  this_df <- pred_subset_adglmm[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred0.2-eta_i)^2,
-           err2 = (pred0.4-eta_i)^2,
-           err3 = (pred0.6-eta_i)^2,
-           err4 = (pred0.8-eta_i)^2) %>%
-    select(id, t, starts_with("err")) %>% 
-    mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != "[0,0.2]") %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_adglmm2[,,m] <- ise_tb_m
-}
-
-mean_ise_adglmm2<- apply(ise_adglmm2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l1_subset}
-for(m in 1:length(pred_subset_gfosr_l1)){
-  this_df <- pred_subset_gfosr_l1[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_gfosr_l1_2[,,m] <- ise_tb_m
-}
-
-mean_ise_gfosr_l1_2<- apply(ise_gfosr_l1_2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r ise_gfosr_l5_subset}
-for(m in 1:length(pred_subset_gfosr_l5)){
-  this_df <- pred_subset_gfosr_l5[[m]]
-  ise_tb_m <- this_df %>%
-    mutate(err1 = (pred_w1-eta_i)^2,
-           err2 = (pred_w2-eta_i)^2,
-           err3 = (pred_w3-eta_i)^2,
-           err4 = (pred_w4-eta_i)^2) %>%
-    select(id, t, starts_with("err"), window) %>% 
-    group_by(window, id) %>% 
-    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
-    group_by(window) %>% 
-    summarize_at(vars(err1, err2, err3, err4), mean) %>%
-    filter(window != 1) %>% 
-    select(starts_with("err")) %>% as.matrix()
-  ise_gfosr_l5_2[,,m] <- ise_tb_m
-}
-
-mean_ise_gfosr_l5_2 <- apply(ise_gfosr_l5_2, c(1, 2), mean, na.rm = T)
-
-colnames(mean_ise_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc2}
-## auc container 
-auc_fgfpca2 <- auc_adglmm2 <- auc_gfosr_l1_2 <- auc_gfosr_l5_2 <- 
-  array(NA, dim = c(length(window)-2, length(window)-2, M))
-```
-
-```{r auc_fGFPCA_subset}
-for(m in 1:length(pred_subset_fGFPCA)){
-  this_df <- pred_subset_fGFPCA[[m]]
-  auc_tb_m <- this_df %>%
-  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.2),
-            auc2 = get_auc(Y, pred0.4),
-            auc3 = get_auc(Y, pred0.6),
-            auc4 = get_auc(Y, pred0.8)) %>%
-  filter(window != "[0,0.2]") %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_fgfpca2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_fgfpca2 <- apply(auc_fgfpca2, c(1,2), mean)
-colnames(mean_auc_fgfpca2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GLMMadaptive_subset}
-for(m in 1:length(pred_subset_adglmm)){
-  this_df <- pred_subset_adglmm[[m]]
-  auc_tb_m <- this_df %>%
-  mutate(window = cut(t, breaks = window, include.lowest = T)) %>% 
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred0.2),
-            auc2 = get_auc(Y, pred0.4),
-            auc3 = get_auc(Y, pred0.6),
-            auc4 = get_auc(Y, pred0.8)) %>%
-  filter(window != "[0,0.2]") %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_adglmm2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_adglmm2 <- apply(auc_adglmm2, c(1,2), mean, na.rm = T)
-colnames(mean_auc_adglmm2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GFOSR_subset_l1}
-for(m in 1:length(pred_subset_gfosr_l1)){
-  this_df <- pred_subset_gfosr_l1[[m]]
-  auc_tb_m <- this_df %>%
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred_w1),
-            auc2 = get_auc(Y, pred_w2),
-            auc3 = get_auc(Y, pred_w3),
-            auc4 = get_auc(Y, pred_w4)) %>%
-  filter(window != 1) %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_gfosr_l1_2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_gfosr_l1_2 <- apply(auc_gfosr_l1_2, c(1,2), mean)
-colnames(mean_auc_gfosr_l1_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r auc_GFOSR_subset_l5}
-for(m in 1:length(pred_subset_gfosr_l5)){
-  this_df <- pred_subset_gfosr_l5[[m]]
-  auc_tb_m <- this_df %>%
-  select(Y, starts_with("pred"), window) %>%
-  group_by(window) %>%
-  summarise(auc1 = get_auc(Y, pred_w1),
-            auc2 = get_auc(Y, pred_w2),
-            auc3 = get_auc(Y, pred_w3),
-            auc4 = get_auc(Y, pred_w4)) %>%
-  filter(window != 1) %>% 
-  select(starts_with("auc")) %>% as.matrix()
-
-  auc_gfosr_l5_2[, , m] <- auc_tb_m
-
-}
-
-mean_auc_gfosr_l5_2 <- apply(auc_gfosr_l5_2, c(1,2), mean)
-colnames(mean_auc_gfosr_l5_2) <- c("0.2", "0.4", "0.6", "0.8")
-```
-
-```{r table_sim2}
-fgfpca2 <- rbind(mean_ise_fgfpca2 ,mean_auc_fgfpca2) %>% as_tibble() %>%
-  mutate(Window = rep(c("(0.2, 0.4]", "(0.4, 0.6]", "(0.6, 0.8]", "(0.8, 1.0]"), 2), .before=1)
-gfosr_l5_2 <- rbind(mean_ise_gfosr_l5_2 ,mean_auc_gfosr_l5_2) %>% as_tibble()
-gfosr_l1_2 <- rbind(mean_ise_gfosr_l1_2 ,mean_auc_gfosr_l1_2) %>% as_tibble()
-adglmm2 <- rbind(mean_ise_adglmm2 ,mean_auc_adglmm2) %>% as_tibble()
-
-bind_cols(fgfpca2, gfosr_l5_2, gfosr_l1_2, adglmm2, .name_repair = "minimal") %>%
+cbind(calc_AUC(pred_list_fGFPCA, window),
+      calc_AUC(pred_list_gfofr, window),
+      calc_AUC(pred_list_gfofr_l5, window),
+      calc_AUC(pred_list_GLMMad, window)) %>%
+  data.frame(window = c("[0, 0.2]", "(0,2, 0.4]", "(0.4, 0.6]", 
+                    "(0.6, 0.8]","(0.8, 1]"),
+             ., 
+             check.names = F) %>%
   kable(digits = 3, booktabs=T,
@@
   add_header_above(c(" " = 1, "Maximum observation time" = 16)) %>%
-  group_rows(index = c("ISE"= 4, "AUC" = 4)) %>%
-  landscape()
-
-```
-
-Among 500 hundred iterations, `r sum(num_probs)` did not converge for the GLMMadaptive model. No numeric issue for fGFPCA.
-
-# Time
-
-Average time per iteration in minutes
-
-```{r}
-time_tb <- data.frame(sim1 = c(mean(fit_time/60+pred_time),
-             mean(fit_time_ref+pred_time_ref/60),
-             mean(t_vec_gfosr_l1/60), 
-             mean(t_vec_gfosr_l5/60)),
-           sim2 = c(mean(fit_time_subset_fGFPCA/60+pred_time_subset_fGFPCA),
-             mean(fit_time_subset_adglmm+pred_time_subset_adglmm/60),
-             mean(t_subset_gfosr_l1/60), 
-             mean(t_subset_gfosr_l5/60))) %>% t()
-colnames(time_tb) <- c("fGFPCA", "GLMMadaptive", "GFOSR (L=1)", "GFOSR (L=5)")
-rownames(time_tb) <- c("Simulation 1", "Simulation 2")
-```
-
-```{r time_tb}
-time_tb %>% 
-  kable(digits = 3, booktabs = T, table.attr = "style = \"color:black;\"") %>%
-  kable_styling(full_width = F) 
-```
-
-# NHANES data application
-
-We take 60% subjects (N = 5257) for training, 40% (n = 3506) for out-of-sample prediction.
-When using the full sample The debiase step (mgcv::bam) in fGFPCA took a really a long time, and GLMM adaptive only works for intercept-only model.
-In the debias step, re-evaluated eigenvalues also lost their decreasing nature.
-
-```{r load_nhanes}
-# df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
-load(here("Data/ApplOutput_fGFPCA.RData"))
-load(here("Data/ApplOutput_GLMMadaptive.RData"))
-
-load(here("Data/Appl_GFOSR_L5.RData"))
-load(here("Data/Appl_GFOSR_L1.RData"))
-
-pred_appl_gfosr_l1 <- pred_appl_gfosr_l1 %>%
-  rename(pred360=pred_w1, pred720=pred_w2, pred1080=pred_w3)
-pred_appl_gfosr_l5 <- pred_appl_gfosr_l5 %>%
-  rename(pred360=pred_w1, pred720=pred_w2, pred1080=pred_w3)
-```
-
-```{r nhance_example, eval=FALSE}
-df_nhanes %>% select(SEQN, Z, sind) %>% 
-  filter(SEQN %in% unique(nhanes_pred_adglmm$id) | SEQN %in% unique(nhanes_pred_fgfpca$id)) %>%
-  mutate(Z = factor(Z, levels = 0:1, labels = c("Inactive", "Active"))) %>%
-  mutate(SEQN = factor(SEQN)) %>%
-  ggplot()+
-  geom_tile(aes(x=sind, y = SEQN, fill = Z))+
-  labs(x="Time", y="Subject", title = "Overview of NHANES binary activity indicator")+
-  theme(axis.text.y = element_blank())+
-  scale_x_continuous(breaks = seq(0, 1440, by = 480),
-                     labels = c("Midnight", "8am", "4pm", "Midnight"))
-```
-
-## AUC
-
-```{r}
-# window
-window <- seq(0, 1440, by = 360)
-```
-
-```{r auc_appl}
-auc_appl_fgfpca <- pred_nhanes_fgfpca %>% 
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_gfosr_l5 <- pred_appl_gfosr_l5 %>% 
-  select(-window) %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_gfosr_l1 <- pred_appl_gfosr_l1 %>% 
-  select(-window) %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-auc_appl_adglmm <- pred_nhanes_adglmm %>%
-    mutate(window = cut(sind, breaks = window, include.lowest = T)) %>% 
-    select(Y, starts_with("pred"), window) %>%
-    group_by(window) %>%
-    summarise(auc1 = get_auc(Y, pred360),
-              auc2 = get_auc(Y, pred720),
-              auc3 = get_auc(Y, pred1080)) %>%
-    filter(window != "[0,360]") %>% 
-    select(starts_with("auc"))
-
-colnames(auc_appl_fgfpca) <- colnames(auc_appl_gfosr_l5) <- colnames(auc_appl_gfosr_l1) <- colnames(auc_appl_adglmm) <- c("6am", "12pm", "6pm")
-```
-
-```{r table_appl}
-bind_cols(Window = c("6am-12pm", "12pm-6pm", "6pm-12am"),
-          auc_appl_fgfpca, auc_appl_gfosr_l5, auc_appl_gfosr_l1, auc_appl_adglmm,
-          .name_repair = "minimal") %>%
-  kable(digits = 3, booktabs=T,
-        table.attr="style=\"color:black;\"") %>%
-  kable_styling(full_width = F) %>% 
-  add_header_above(c(" " = 1, "fGFPCA" = 3, "GFOSR (L=5)" = 3, 
-                     "GFOSR (L=1)" = 3, "GLMMadaptive" = 3)) %>%
-  add_header_above(c(" " = 1, "Maximum observation time" = 12)) %>%
   landscape()
@@
 
-## Figure
+It seems that the GFOSR models, while are the fastest, also performs the worst in AUC.
 
-```{r fig_appl_paper, fig.height=8, fig.width=8}
-# redefine color scale
-cols <- cols[1:3] 
-names(cols) <- c("6am", "12pm", "6pm")
+### Prediction interval
 
-# selct four subjects
-plot_id <- sort(c(68959, 68624, 64198, 80517))
+Not sure how to display. Maybe start with if the prediction interval covers the truth? That is, coverage rate at each time point, averaged over all simulation and subjects?
 
-# figure
-bind_rows(
-  pred_nhanes_fgfpca %>% filter(id %in% plot_id) %>% mutate(method="fGFPCA"),
-  pred_appl_gfosr_l5 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=5)"),
-  pred_appl_gfosr_l1 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=1)"),
-  pred_nhanes_adglmm %>% filter(id %in% plot_id) %>% mutate(method = "GLMMadaptive")) %>%
-  mutate_at(vars(pred360, pred720, pred1080), 
-                .funs = function(x){exp(x)/(1+exp(x))}) %>%
-  mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+```{r, fig.height=4, fig.width=16}
+bind_rows(calc_predint_cover(pred_list_fGFPCA) %>% mutate(method = "fGFPCA"),
+          calc_predint_cover(pred_list_gfofr_l5) %>% mutate(method = "GFOSR (L=5)"),
+          calc_predint_cover(pred_list_gfofr) %>% mutate(method = "GFOSR (L=1)"),
+          calc_predint_cover(pred_list_GLMMad) %>% mutate(method = "GLMMadaptive")) %>%
   ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  geom_line(aes(x=sind, y=pred360, col = "6am"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=sind, y=pred720, col = "12pm"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linewidth=1, na.rm=T)+
-  facet_grid(rows = vars(id), cols = vars(method))+
-  labs(col = "Maximum observation time", x = "Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1440, by = 360),
-                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
-  theme(legend.position = "bottom", axis.text.x = element_text(size = 7))+
+  geom_line(aes(x=t, y=cover0.2, col = "0.2"), na.rm = T)+
+  geom_line(aes(x=t, y=cover0.4, col = "0.4"), na.rm = T)+
+  geom_line(aes(x=t, y=cover0.6, col = "0.6"), na.rm = T)+
+  geom_line(aes(x=t, y=cover0.8, col = "0.8"), na.rm = T)+
+  facet_grid(cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="Coverage rate")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(legend.position = "bottom")+
   scale_color_manual(values = cols)
 ```
-
-```{r fig_appl_pre, fig.height=4, fig.width=8}
-# selct four subjects
-plot_id <- sort(c(64198, 80517))
-
-# figure
-bind_rows(
-  pred_nhanes_fgfpca %>% filter(id %in% plot_id) %>% mutate(method="fGFPCA"),
-  pred_appl_gfosr_l5 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=5)"),
-  pred_appl_gfosr_l1 %>% filter(id %in% plot_id) %>% mutate(method = "GFOSR (L=1)"),
-  pred_nhanes_adglmm %>% filter(id %in% plot_id) %>% mutate(method = "GLMMadaptive")) %>%
-  mutate_at(vars(pred360, pred720, pred1080), 
-                .funs = function(x){exp(x)/(1+exp(x))}) %>%
-  mutate(method=factor(method, 
-         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
-  ggplot()+
-  geom_point(aes(x=sind, y=Y), size = 0.2)+
-  geom_line(aes(x=sind, y=pred360, col = "6am"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=sind, y=pred720, col = "12pm"), linewidth=1, na.rm=T)+
-  geom_line(aes(x=sind, y=pred1080, col = "6pm"), linewidth=1, na.rm=T)+
-  facet_grid(rows = vars(id), cols = vars(method))+
-  labs(col = "Maximum observation time", x = "Time", y="")+
-  scale_x_continuous(breaks = seq(0, 1440, by = 360),
-                     labels = c("0am", "6am", "12pm", "6pm", "12am"))+
-  theme(legend.position = "bottom", axis.text.x = element_text(size = 7))+
-  scale_color_manual(values = cols)
-```
-
-The fGFPCA prediction conditional on observation up to 6am looks identical across subjects.
-This is not due to coding issue.
-In fact, if we look at individual tracks, out of 3506 subjects there are 1473 unique predicted tracks.
-There is a high change the selected four subjects all have the same predicted tracks.
-
-PS number of unique predicted tracks is:
-- 3456 when conditional on 12 pm
-- 3488 when conditional on 6pm
-
-I think this is closedly determined by the variety of observed outcome across individuals.
-When conditioning on 6am, there are 1475 unique observed tracks; and 3465 and 3488 respectively when obesrved up to 12pm and 6pm.
-As we see, these numbers are very close to the number of unique prediction tracks in fGFPCA.
-
-I'd note that the four randomly selected subjects from the last report happen to have exactly the same observations from 0-6am.
-That is why they also have identical prediction after 6am.
-Therefore, in this report I arbitrarily assigned subjects to be displayed: 62161, .
-
-In addition: 
-- For GFOSR (L5), number of unique predicted tracks are 31, 32, 32
-- For GFOSR (L1), number of unique predicted tracks are 2, 2, 2
-- For GLMMadaptive, number of unique predicted tracks are 294, 712, 1533.
-
-```{r, eval=FALSE}
-length(unique(pred_nhanes_adglmm$id))
-pred_nhanes_fgfpca %>%
-  # filter(id %in% rand_id) %>%
-  select(id, sind, Y) %>%
-  filter(sind <=  360) %>%
-  pivot_wider(names_from = sind, values_from = Y, names_prefix = "sind") %>%
-  distinct(., pick(starts_with("sind")), .keep_all = T) %>%
-  select(id) %>%
-  t() %>%
-  sample(., size = 4)
-
-  # ggplot(aes(x=sind, y=pred360, col = as.factor(id), group = as.factor(id)))+
-  # geom_line()
-```
-
-```{r, eval=FALSE}
-pred_nhanes_fgfpca %>%
-  select(id, sind, pred360) %>%
-  filter(sind>360) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=pred360, col = as.factor(id)), show.legend = F)
-
-pred_nhanes_fgfpca %>%
-  select(id, sind, pred720) %>%
-  filter(sind>720) %>%
-  ggplot()+
-  geom_line(aes(x=sind, y=pred720, col = as.factor(id)), show.legend = F)
-
-```

### 2024-07-19 Code/PredEval.R

@@ -1,100 +1,107 @@
-# This script codes the computaiton of evaluation of predictive performance
+# This script codes the functions of evaluation of predictive performance
 
-library(ROCR)
-library(GLMMadaptive)
+#### A function to calculate ISE ####
 
+## calculation
+# pred list: must contain columns: pred.., eta_i, id
 
-obs_t <- c(195, 395, 595)
-
-ise_fgfpca <- rep(NA, 3)
-ise_adglmm <- rep(NA, 3)
-
-auc_fgfpca <- rep(NA, 3)
-auc_adglmm <- rep(NA, 3)
-
-comp_t_fgfpca <- rep(NA, 3)
-comp_t_adglmm <- rep(NA, 3)
-
-adglmm <- mixed_model(Y ~ sind, random = ~ sind | id, data =  df,
-                      family = binomial())
-
-##### Midpoint ISE ##### 
-
-# true eta and true y
-# head(df_true_eta)
-# table(df_true_eta$id)
-# unique(df_true_eta$mid)
-
-eta_true_mat <- df_true_eta %>%
-  pivot_wider(names_from = id, values_from = true) %>%
-  column_to_rownames("mid")
-
-y_mat <- df %>% select(id, Y, sind_inx) %>% 
-  pivot_wider(names_from = id, values_from = Y) %>%
-  column_to_rownames("sind_inx")
-
-# observation up to 195
-
-for(j in 1: length(obs_t)){
+calc_ISE <- function(pred_list, window){
   
-  t <- obs_t[j]
-  pred_t <- which(mid > t) # unobserved bin index (binned grid)
-  pred_t_full <- (t+1):max(mid) # unobserved time index (original grid)
+  M <- length(pred_list)  # number of simulations
   
-  # fast GFPCA prediction container
-  eta_hat_mat <- matrix(NA, nrow=length(mid), ncol = N)
-  eta_hat_mat_full <- matrix(NA, nrow=J, ncol = N) # latent Gaussian function on original grid by interpolation
- 
-  # prediction
-  tstart <- proc.time()
-  for(i in 1:N){
-    eta_hat_mat[ , i] <- out_samp_dyn_pred(df_new = df %>% filter(id==i & sind_inx<=t) %>%
-                                             select(-eta_i),
-                                           fpca_fit = fpca_fit)$eta_pred
[…]

### 2024-07-31 Code/Simulation_N500/Sim_fGFPCA.R
sampling quantiles

@@ -24,7 +24,6 @@ library(splines)
 library(rstan)
 theme_set(theme_minimal())
 
-my_logis <- function(x){exp(x)/(1+exp(x))}
 
 #### load simulated data ####
 
@@ -41,7 +40,7 @@ rand_id <- sample(N, size = 4)
 sim_data[[357]] %>% filter(id %in% rand_id) %>% 
   ggplot()+
   geom_point(aes(x=t, y=Y))+
-  geom_line(aes(x=t, y=my_logis(eta_i)), col = "red")+
+  geom_line(aes(x=t, y=exp(eta_i)/(1+exp(eta_i))), col = "red")+
   #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
   facet_wrap(~id)
 
@@ -186,16 +185,22 @@ for(m in 1:M){
       )
       
       scores_tmax <- summary(fit)$summary[1:K, "mean"]
-      sd_scores_tmax <- summary(fit)$summary[1:K, "sd"]
+      # sd_scores_tmax <- summary(fit)$summary[1:K, "sd"]
+      qt_score_tmax <- summary(fit)$summary[1:K, c("2.5%", "97.5%")]
       
       # latent function predictions
       eta_pred_out <- new_mu+new_phi%*%scores_tmax
       df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
       
       # prediction interval
-      sd_eta <- sqrt((new_phi^2) %*% sd_scores_tmax^2)
-      df_i[ , paste0("pred", tmax, "_lb")] <- as.vector(eta_pred_out[, 1]-qnorm(0.975)*sd_eta)
-      df_i[ , paste0("pred", tmax, "_ub")] <- as.vector(eta_pred_out[, 1]+qnorm(0.975)*sd_eta)
+      # sd_eta <- sqrt((new_phi^2) %*% sd_scores_tmax^2)
+      # df_i[ , paste0("pred", tmax, "_lb")] <- as.vector(eta_pred_out[, 1]-qnorm(0.975)*sd_eta)
+      # df_i[ , paste0("pred", tmax, "_ub")] <- as.vector(eta_pred_out[, 1]+qnorm(0.975)*sd_eta)
+      # 
+      # quantile interval
[…]

### 2024-07-31 Code/TryStan.R

@@ -1,11 +1,26 @@
 library(rstan)
 
-test_id <- 501:504
+#### true values ####
+
+# fixed effect: mean function
+true_mu <- rep(0, J)
+
+#eigenfunctions
+true_phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
+                     sin(4*pi*t),cos(4*pi*t)) # orthogonal random functions
+
+true_lambda = 0.5^(0:(K-1)) # eigenvalues/variance of scores
+
+
+
+#### stan ####
+test_df <- sim_data[[1]] %>% filter(id %in% 501:600)
+test_id <- unique(test_df$id) 
 tmax_vec <- c(0.2, 0.4, 0.6, 0.8)
 df_pred <- list()
 
-for(m in test_id){
-
+for(id in seq_along(test_id)){
+  m <- test_id[[id]]
   df_i <- test_df %>% filter(id==m)
   
   scores <- matrix(NA, K, length(tmax_vec))
@@ -19,9 +34,9 @@ for(m in test_id){
     # into a list
     stanData <- list(
       J = J, Ju = nrow(df_it), Y = df_it$Y, K = K,
-      efuncs = new_phi, 
-      b0 = new_mu,
-      lambda = new_lambda
+      efuncs = true_phi, 
+      b0 = true_mu,
+      lambda = true_lambda
     )
     
     fit <- stan(
@@ -54,16 +69,16 @@ for(m in test_id){
     
     
     # latent function predictions
-    eta_pred_out <- new_mu+new_phi%*%scores[ ,i]
+    eta_pred_out <- true_mu+true_phi%*%scores[ ,i]
     df_i[ , paste0("pred", tmax)] <- eta_pred_out[,1]
     
     # prediction interval
-    sd_eta <- sqrt((new_phi^2) %*% sd_scores[ ,i]^2)
+    sd_eta <- sqrt((true_phi^2) %*% sd_scores[ ,i]^2)
     df_i[ , paste0("pred", tmax, "_lb")] <- as.vector(eta_pred_out[, 1]-qnorm(0.975)*sd_eta)
     df_i[ , paste0("pred", tmax, "_ub")] <- as.vector(eta_pred_out[, 1]+qnorm(0.975)*sd_eta)
   }
   
-  df_pred[[m-500]] <- df_i
+  df_pred[[id]] <- df_i
 
 }
 
[…]

### 2024-08-08 Docs/manuscript.Rmd
sampling quantiles
investigate the prediction interval

@@
 
+pred_list_GLMMad <- pred_list_GLMMad[1:5]
 pred_list_GLMMad <- lapply(pred_list_GLMMad, function(x){
@@
 ```
+
+For the fGFPCA method, which interval should we use? Sampling quantile would cross zero and the coverage rate seems to fluctuate a lot.
+
+## A closer look at confidence/credible interval
+
+Since using the true parameters did not change the results, I don't think ignoring prior variance made a huge difference.
+If we plot the variance of mean (from global bam) or the variance of estimated $\eta$, the trend would in fact be roughly cyclic.
+
+I am not sure how to take into account the prior variation. As below:
+
+$$Var(\hat{\eta}(t)) = Var(\hat{\mu}_0(t)+
+\sum_k \hat{\phi}_k(t)
+\hat{\xi}_k)$$
+
+$$
+\begin{aligned}
+Var(\hat{\eta}(t) | \hat{\mu}_0(t), \hat{\lambda}_k, \hat{\phi}_k(t)) & = Var(\hat{\mu}_0(t))+
+Var(\sum_k \hat{\phi}_k(t)
+\hat{\xi}_k)\\
+
+& = Var(\hat{\mu}_0(t))+ \sum_k \hat{\phi}_k^2(t)
+\hat{\lambda}_k
+\end{aligned}
+$$
+
+Not sure why we need to take into account the variation of mean function (from global BAM).
+
+If we want to get rid of the conditions (let's for now assume the terms are independent):
+
+$$Var(\hat{\eta}(t))  = Var(\hat{\mu}_0(t))+
+Var(\sum_k \hat{\phi}_k(t)
+\hat{\xi}_k)$$
+
+The second term would be difficult to calculate. Is the estimated mean and variance of basis functions available?
 

### 2024-08-08 Code/Simulation_N500/Sim_GLMMadaptive.R

@@ -50,7 +50,8 @@ N_train <- 500
 N_test <- 100
 
 # containers
-M <- length(sim_data)
+# M <- length(sim_data)
+M <- 5
 pred_list_GLMMad <- list()
 time_GLMMad <- pred_time_ref <- rep(NA, M)
 
@@ -65,7 +66,7 @@ for(m in 1:M){
   
   t1 <- Sys.time()
   # fit GLMM adaptive model: linear time fixed and random effect
-  adglmm <- mixed_model(Y ~ t + I(t^2), random = ~ t + I(t^2) | id,
+  adglmm <- mixed_model(Y ~ t, random = ~ t | id,
                         data =  train_df, family = binomial())
   t2 <- Sys.time()
   

### 2024-08-26 Docs/manuscript.Rmd
run small simulation for fGFPCA

@@
 ```{r load_sim1}
-load(here("Data/TrialRun/SimOutput_fGFPCA.RData"))
-load(here("Data/TrialRun/SimOutput_GLMMadaptive.RData"))
-load(here("Data/TrialRun/SimOutput_GFOSR_L1.RData"))
-load(here("Data/TrialRun/SimOutput_GFOSR_L5.RData"))
+load(here("Data/SimN500/SimOutput_fGFPCA.RData"))
+load(here("Data/SimN500/SimOutput_GLMMadaptive.RData"))
+load(here("Data/SimN500/SimOutput_GFOSR_L1.RData"))
+load(here("Data/SimN500/SimOutput_GFOSR_L5.RData"))
 
@@
 
-pred_list_GLMMad <- pred_list_GLMMad[1:5]
+# pred_list_GLMMad <- pred_list_GLMMad[1:5]
 pred_list_GLMMad <- lapply(pred_list_GLMMad, function(x){
@@
 
-pred_list_gfofr <- lapply(pred_list_gfofr, function(x){
+pred_list_gfofr_l1 <- lapply(pred_list_gfofr_l1, function(x){
   x <- x %>% select(-window)
@@
     mutate(method = "GLMMadaptive"),
-  pred_list_gfofr[[1]] %>% filter(id %in% rand_id) %>%
+  pred_list_gfofr_l1[[1]] %>% filter(id %in% rand_id) %>%
     mutate(method = "GFOSR (L=1)"),
@@
   geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm = T)+
+  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm = T)+
+  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm = T)+
+  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm = T)+
+  guides(alpha = "none", col="none")+
+  facet_grid(rows = vars(id), cols = vars(method))+
+  labs(col = "Maximum observation time", x = "Time", y="")+
+  scale_x_continuous(breaks = seq(0, 1, by = 0.2))+
+  theme(legend.position = "bottom")+
+  scale_color_manual(values = cols)
+```
+
+```{r large_sim_plot_ci, fig.height=8, fig.width=8}
+bind_rows(
+  pred_list_fGFPCA[[1]] %>% filter(id %in% rand_id) %>% 
+    mutate(method="fGFPCA"),
+  pred_list_GLMMad[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GLMMadaptive"),
+  pred_list_gfofr_l1[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=1)"),
+  pred_list_gfofr_l5[[1]] %>% filter(id %in% rand_id) %>%
+    mutate(method = "GFOSR (L=5)") 
+) %>% 
+  mutate_at(vars(eta_i, starts_with("pred")), 
+                .funs = function(x){exp(x)/(1+exp(x))}) %>%
+  mutate(method=factor(method, 
+         levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+  ggplot()+
+  geom_point(aes(x=t, y=Y), size = 0.2)+
+  geom_line(aes(x=t, y=eta_i, col = "True"))+
+  # geom_line(aes(x=t, y=pred0.2, col = "0.2"), na.rm = T)+
   geom_ribbon(aes(x=t, ymin=pred0.2_lb, ymax = pred0.2_ub,
@@
               linetype="dashed")+
-  geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm = T)+
+  # geom_line(aes(x=t, y=pred0.4, col = "0.4"), na.rm = T)+
   geom_ribbon(aes(x=t, ymin=pred0.4_lb, ymax = pred0.4_ub,
@@
               linetype="dashed")+
-  geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm = T)+
+  # geom_line(aes(x=t, y=pred0.6, col = "0.6"), na.rm = T)+
   geom_ribbon(aes(x=t, ymin=pred0.6_lb, ymax = pred0.6_ub,
@@
               linetype="dashed")+
-  geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm = T)+
+  # geom_line(aes(x=t, y=pred0.8, col = "0.8"), na.rm = T)+
   geom_ribbon(aes(x=t, ymin=pred0.8_lb, ymax = pred0.8_ub,
@@
   theme(legend.position = "bottom")+
-  scale_color_manual(values = cols)
+  scale_color_manual(values = cols)+
+  scale_fill_manual(values = cols)
 ```
@@
                     mean(time_gfofr_l5),
-                    mean(time_gfofr),
+                    mean(time_gfofr_l1),
                     mean(time_GLMMad))) %>%
@@
 cbind(calc_ISE(pred_list_fGFPCA, window),
-      calc_ISE(pred_list_gfofr, window),
+      calc_ISE(pred_list_gfofr_l1, window),
       calc_ISE(pred_list_gfofr_l5, window),
@@
 cbind(calc_AUC(pred_list_fGFPCA, window),
-      calc_AUC(pred_list_gfofr, window),
+      calc_AUC(pred_list_gfofr_l1, window),
       calc_AUC(pred_list_gfofr_l5, window),
@@
           calc_predint_cover(pred_list_gfofr_l5) %>% mutate(method = "GFOSR (L=5)"),
-          calc_predint_cover(pred_list_gfofr) %>% mutate(method = "GFOSR (L=1)"),
+          calc_predint_cover(pred_list_gfofr_l1) %>% mutate(method = "GFOSR (L=1)"),
           calc_predint_cover(pred_list_GLMMad) %>% mutate(method = "GLMMadaptive")) %>%

### 2024-08-26 Code/Simulation_N100/Sim_Subset_fGFPCA.R

@@ -17,7 +17,9 @@ library(kableExtra)
 library(mvtnorm)
 library(splines)
 library(rstan)
-
+library(RColorBrewer)
+cols <- c(brewer.pal(4, "Set2"), "#000000") # define a color palette 
+names(cols) <- c("0.2", "0.4", "0.6", "0.8", "True")
 
 
 #### load simulated data ####
@@ -28,10 +30,10 @@ M <- 500 # number of simulations
 
 # reduce data
 sim_data <- lapply(sim_data[1:M], function(x){x %>% filter(id %in% c(1:200))})
-lapply(sim_data, dim)
-head(sim_data[[1]])
-unique(sim_data[[1]]$id)
-unique(sim_data[[1]]$t)
+# lapply(sim_data, dim)
+# head(sim_data[[1]])
+# unique(sim_data[[1]]$id)
+# unique(sim_data[[1]]$t)
 
 Ntr <- Nte <- 100 # trainig and testing sample size
 J <- 1000
@@ -41,7 +43,7 @@ J <- 1000
 #### set up ####
 K <- 4 # number of eigenfunctions
 source(here("Code/GLMM-FPCA.R")) 
-source(here("Code/OutSampBayes.R"))
+# source(here("Code/OutSampBayes.R"))
 
 # values for binning 
 bin_w <- 10 # bin width
@@ -66,14 +68,15 @@ window <- seq(0.2, 0.8, by = 0.2)
 
 #### containers ####
 
+# M <- 5
 pred_list_all <- list()
-converge_state_list <- list()
-fit_time <- pred_time <- rep(NA, M)
+# converge_state_list <- list()
+time_vec <- rep(NA, M)
 
-# M <- 10
+
 M
 
-#### fGFPCA
+#### fGFPCA ####
 
 pb = txtProgressBar(min = 0, max = M, initial = 0, style = 3) 
 for(m in 1:M){
@@ -111,6 +114,7 @@ for(m in 1:M){
   colnames(df_phi) <- c("t", paste0("phi", 1:4))
   df_train_m <- df_train_m %>% left_join(df_phi, by = "t")
   df_train_m$id <- droplevels(df_train_m$id)
+  # t1 <- Sys.time()
   debias_glmm <- bam(Y ~ s(t, bs="cc", k=10)+
                        s(id, by=phi1, bs="re")+
                        s(id, by=phi2, bs="re")+
@@ -119,21 +123,22 @@ for(m in 1:M){
                      family = binomial, data=df_train_m, 
                      method = "fREML",
[…]

### 2024-08-26 Code/Simulation_N500/Sim_GenFOSR.R

@@ -45,12 +45,11 @@ sim_data[[357]] %>% filter(id %in% rand_id) %>%
 
 N_train <- 500 # training sample size
 N_test <- 100 # testing sample szie 
-# L <- 5 # number of last observations used as time-fixed predictor
-L <- 5
+L <- 1 # number of last observations used as time-fixed predictor
 windows <- seq(0, 1, by = 0.2) # prediction window
 
 #### Simulation ####
-M <-5
+# M <-5
 time_gfofr <- rep(NA, M)
 pred_list_gfofr <- list()
 
@@ -79,19 +78,19 @@ for(m in 1:M){
     # outcome (training set)
     df_pred_tr <- train_df %>% filter(window > w) %>%
       left_join(y_obs_max, by = "id") %>% 
-      mutate_at(vars(Y, starts_with("yl")), as.factor) 
-    # fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) + 
+      mutate_at(vars(Y, starts_with("yl")), as.factor)
+    # fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) +
     #                       s(t, bs="cr", k=20, by = yl5)+
     #                       s(t, bs="cr", k=20, by = yl4)+
     #                       s(t, bs="cr", k=20, by = yl3)+
     #                       s(t, bs="cr", k=20, by = yl2)+
     #                       s(t, bs="cr", k=20, by = yl1),
-    #                     family = binomial, data=df_pred_tr, 
+    #                     family = binomial, data=df_pred_tr,
     #                     method = "fREML",
     #                     discrete = TRUE)
-    fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) + 
+    fit_gen_fosr <- bam(Y ~ s(t, bs="cr", k=20) +
                           s(t, bs="cr", k=20, by = yl1),
-                        family = binomial, data=df_pred_tr, 
+                        family = binomial, data=df_pred_tr,
                         method = "fREML",
                         discrete = TRUE)
     
@@ -170,14 +169,14 @@ ggsave(here("Images/IntervalExp1_3.jpeg"), height=12, width = 5)
 
 #### Save results ####
 
-pred_list_gfofr_l5 <- pred_list_gfofr
-time_gfofr_l5 <- time_gfofr
[…]

### 2024-09-02 Docs/manuscript.Rmd
rerun small simulation

@@
   html_document:
-    self_contained: no
+    self_contained: yes
     number_sections: yes

### 2024-09-02 Code/Figures/FiguresSim.R

@@ -39,27 +39,37 @@ source(here("Code/PredEval.R"))
 
 
 
-#### Large scale simulation ####
+#### Simulation output ####
 # load data
+## N = 500
 load(here("Data/SimN500/SimOutput_fGFPCA.RData"))
 load(here("Data/SimN500/SimOutput_GLMMadaptive.RData"))
 load(here("Data/SimN500/SimOutput_GFOSR_L1.RData"))
 load(here("Data/SimN500/SimOutput_GFOSR_L5.RData"))
 
+## N = 100
+load(here("Data/SimN100/SubSimOutput_fGFPCA.RData"))
+load(here("Data/SimN100/SubSimOutput_GLMMadaptive.RData"))
+load(here("Data/SimN100/SubSimOutput_GFOSR_L1.RData"))
+load(here("Data/SimN100/SubSimOutput_GFOSR_L5.RData"))
+
 # choose 4 subjects to plot
 rand_id <- sample(501:600, size = 4)
+rand_id <- sample(101:200, size = 4)
 
-# format
+#### Format ####
+##### N = 500 #####
 pred_list_fGFPCA <- lapply(pred_list_fGFPCA, function(x){x %>% select(-bin)})
 name_vec <- colnames(pred_list_fGFPCA[[1]])
-pred_list_fGFPCA <- lapply(pred_list_fGFPCA, 
+pred_list_fGFPCA <- lapply(pred_list_fGFPCA,
                            function(x){
                              x[x$t<=0.2, c("pred0.2", "pred0.2_lb", "pred0.2_ub")] <- NA
                              x[x$t<=0.4, c("pred0.4", "pred0.4_lb", "pred0.4_ub")] <- NA
                              x[x$t<=0.6, c("pred0.6", "pred0.6_lb", "pred0.6_ub")] <- NA
                              x[x$t<=0.8, c("pred0.8", "pred0.8_lb", "pred0.8_ub")] <- NA
-                             return(x)    
+                             return(x)
                            })
+
 
 pred_list_GLMMad <- lapply(pred_list_GLMMad, function(x){
   colnames(x) <- name_vec
@@ -77,6 +87,44 @@ pred_list_gfofr_l5 <- lapply(pred_list_gfofr_l5, function(x){
   colnames(x) <- name_vec
   return(x)
 })
+
+
+##### N = 100 #####
+# N = 100
+name_vec <- colnames(pred_subset_fGFPCA[[1]])
[…]

### 2024-09-02 Code/Simulation_N100/Sim_Subset_GLMMadaptive.R

@@ -175,13 +175,12 @@ pred_list_all[[1]] %>%
 
 #### Save results ####
 
-fit_time_subset_adglmm <- fit_time
-pred_time_subset_adglmm <- pred_time
+fit_time_subset_adglmm <- time_vec
 pred_subset_adglmm <- pred_list_all
 
-save(fit_time_subset_adglmm, pred_time_subset_adglmm, pred_subset_adglmm, 
+save(fit_time_subset_adglmm, pred_subset_adglmm, 
      num_probs,
-     file = here("Data/SubSimOutput_GLMMadaptive.RData"))
+     file = here("Data/SimN100/SubSimOutput_GLMMadaptive.RData"))
 
 
 

### 2024-10-22 Docs/Presentation.Rmd
remake figure for presentation

@@
 ```{r sim, fig.height=4, fig.width=10}
-bind_rows(sim1 %>% mutate(sim = "Simulation 1"), sim2 %>% mutate(sim = "Simulation 2")) %>%
+bind_rows(sim1 %>% mutate(sim = "Original Sample"), sim2 %>% mutate(sim = "Reduced sample")) %>%
   ggplot()+
@@
 
-```{r load_nhanes_result}
+```{r load_nhanes_result, eval=F}
 # df_nhanes <- read_rds(here("Data/nhanes_bi.rds"))
@@
 
-```{r fig_appl_paper, fig.height=4, fig.width=10}
+```{r fig_appl_paper, fig.height=4, fig.width=10, eval=F}
 # redefine color scale

### 2024-12-17 Docs/fGFPCA_DynPred_arXiv/Manuscript.tex
final update

\documentclass[12pt]{article}

\usepackage[letterpaper,top=2cm,bottom=2cm,left=2cm,right=2cm,marginparwidth=1.75cm]{geometry}
% double space
\usepackage{setspace}
\doublespacing
% author
\usepackage{authblk} 
% put figures at the end
% \usepackage[nomarkers, nolists]{endfloat}

\usepackage{graphicx}
\usepackage{pdflscape}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{tocloft}
\usepackage{blindtext}
\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}
\usepackage{xcolor}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
% keywords 
\providecommand{\keywords}[1]{\textbf{\textit{Keywords }} #1}

\title{Dynamic Prediction of High-density Generalized Functional Data with Fast Generalized
Functional Principal Component Analysis}
\author[1,*]{Ying Jin}
\author[1]{Andrew Leroux}
\affil[1]{Department of Biostatistics and Informatics, University of Colorado Anschutz Medical Campus, Aurora, CO, USA}
\affil[*]{Corresponding author: Ying Jin, ying.jin@cuanschutz.edu}

\date{}

\begin{document}
\maketitle

\begin{abstract}
Dynamic prediction, which typically refers to the prediction of future outcomes using historical records, is often of interest in biomedical research. For datasets with large sample sizes, high measurement density, and complex correlation structures, traditional methods based on Generalized Linear Mixed Models are often infeasible because of the computational burden associated with \textit{both} data size and model complexity. Moreover, many implementations of mixed-effects models do not directly facilitate out-of-sample predictions for generalized outcomes. To address these issues, we develop a novel approach for dynamic predictions based on a recently developed method estimating complex patterns of variation for exponential family data: fast Generalized Functional Principal Components Analysis (fGFPCA). Our method is able to handle large-scale, high-density repeated measures much more efficiently than existing methods, with its implementation feasible even on personal computational resources (e.g., a standard desktop or laptop computer). The proposed method makes highly flexible and accurate predictions of future trajectories for data that exhibit high degrees of nonlinearity, and allows for out-of-sample predictions to be obtained without reestimating any parameters. A simulation study is designed and implemented to illustrate the advantages of this method. To demonstrate its practical utility, we also conducted a case study to predict diurnal active/inactive patterns using accelerometry data from the National Health and Nutrition Examination Survey (NHANES) 2011-2014. Both the simulation study and the data application demonstrate the better predictive performance and high computational efficiency of the proposed method compared to existing methods. The proposed method also obtains more personalized prediction that improves as more information becomes available, which is an essential goal of dynamic prediction that other methods fail to achieve. 
\end{abstract}

\keywords{Functional data, Dynamic prediction, Repeated measures, Computational efficiency}

\section{Introduction}
\label{sec:introduction}

% Dynamic prediction methods

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, where we hope to make predictions that are both 1) individualized; and 2) improves as new information is collected. Such problems are typically termed dynamic prediction. The two desirable properties of these methods are important to support timely clinical decision-making that is customized to each unique individual. Examples of such problems include predicting the change in certain biomarkers over time \citep{leroux2016}, predicting the future status of physical activity using historical data collected by wearable devices, predicting the development of lesion based on imaging data, et cetera.

Traditionally, this type of data has been referred to as ``repeated measures" and modeled either marginally, such as generalized estimating
equations, or conditionally on specific subjects, such as mixed-effect
models \citep{davidian2003, Laird1982, liang1986, lindstrom1990}. Subject-specific predictions are often made based on the correlation between repeated measures from the same subject. Examples include \citet{GLMMadaptive}, which makes predictions by fitting a generalized linear mixed model (GLMM) and approximates the posterior distribution of parameters of interest with a multivariate normal distribution and their posterior mean. It adopts the Monte Carlo scheme to efficiently approximate the posterior likelihood. Moreover, such repeated measures are often modeled together with a time-to-event outcome to predict the risk of an event, which is updated with the value of repeated measures \citep{Rizopoulos2017, suresh2017}. Popular model frameworks include joint models and landmarking. The former specifies a joint distribution of the repeated measure and the time-to-event outcome. Random effects are typically introduced to model repeated measures and/or introduce correlation between multiple series of repeated measures. The latter estimates models and makes predictions conditioning on a series of ``landmark" times, although these models typically use time-dependent fixed effects alone for prediction and do not include random effects. \citep{vanHouwelingen2007, Rizopoulos2011}

% dynamic prediction challenges 
% the reason for turning to FDA
However, with
the notable exception of \citet{leroux2016}, existing methods are either 
limited in their flexibility (both fixed and random effect structures,
e.g.~random slope and intercept), computationally unfeasible for large
datasets, or both. Moreover, when the repeated measure is non-Gaussian, out-of-sample prediction often requires reestimating part or all of the model parameters, which adds to the computational burden. These issues are particularly acute for densely measured data, as the computational burden increases substantially with increased data size and measurement density, motivating investigators to look for more efficient methods. A statistical framework well-suited for high-density data with complex association structures is known as Functional Data Analysis (FDA) \citep{fdaR2024, fda2005}, which conceptualizes high-density repeated measures as the discrete realization of a continuous latent process or a continuous function measured with error. To better explain and illustrate the methodological framework of the FDA, we use data from the National Health and Nutrition Examination Survey (NHANES) as a motivational example throughout this manuscript. The NHANES data from 2011-2014 includes minute-level physical activities collected with wearable activity monitors, which were then dichotomized into binary activity indicators. Figure \ref{fig:nhanes_data} shows the data of six participants. The black dots indicate the activity status of each minute, with 1 indicating active and 0 inactive. The measurement grid is very dense with 1440 repeated measures over one day. The red lines are the estimated probability curves derived from the observed outcomes, which are interpreted as the \textit{latent} process that specifies the distribution of outcomes at each minute. As the figure reveals, these diurnal trends (and hence, latent functions) are highly nonlinear. Moreover, there is substantial heterogeneity between individuals, which cannot be fully explained using participant characteristics modeled as fixed effects, with unexplained variation continuing to exhibit a high degree of nonlinearity.

% dynamic prediction of functional data, gaussian
Methodology for estimating functional regression models with subject-specific random effects has focused primarily on continuous/Gaussian outcomes, often modeling
subject-specific random effects with Functional Principal Component Analysis (FPCA) \citep{chiou2012, goldberg2014, shang2017, kraus2015}. Existing methods often use partial observations for prediction with an intercept-only model.
\citet{leroux2016} proposed
the Functional Concurrent Regression (FCR) framework, which can incorporate
the effect of subject-specific predictors, accompanied by an R package \emph{fcr} \citep{fcrpkg}. \citet{delaigo2016} achieved
similar goals using Markov Chains to model the transition between discrete realizations of a continuous function, though the model it proposed did not include random effects.

% functional methods on generalized outcome
Software and methods for non-Gaussian functional data are currently even more limited. Software is largely limited to the \emph{refund} \citep{refundpkg} and \emph{registr} \cite{Wrobel2018}
packages in the statistical software R, both developed based on the Generalized Functional Principal Component Analysis (GFPCA) framework. Functional
regression models of non-Gaussian outcomes have also been proposed. \citet{chen2013} proposed approaches to fit
marginal functional models that are compatible with multilevel generalized outcomes, although marginal approaches are not suitable for making subject-specific dynamic predictions. \citet{gertheiss2017} identified the distinction between marginal and conditional estimates and pointed out that directly applying a marginal method (e.g. single-level FPCA) to conditional estimation could cause bias. It also proposed to address this problem using a two-stage joint estimation strategy. \citet{goldsmith2015} proposed estimating parameters with Bayesian method in \emph{Stan}, but the model
framework only takes into account the fixed effect of time-invariant
covariates. \citet{linde2019} used an adapted Bayesian
variational algorithm for FPCA of binary and count data. However, most of these methods are not adaptable to dynamic prediction problems, and others can be time consuming in parameter estimation, making it unfeasible for large-scale, high-density datasets, such as the following motivation example of NHANES accelerometer data in Section \ref{sec:data}.

To address the challenges mentioned above, in this paper, we propose a novel method based on the principles of GLMM and GFPCA, which is fast and scalable for large datasets with high-density repeated measures and readily allows for dynamic predictions of new subjects without the need for parameter reestimation. Section \ref{sec:data} introduces the NHANES activity indicator as a motivation example. Section \ref{sec:method} explains the model framework of the proposed method, laying out the algorithm for implementation. In Section \ref{sec:simulation}, we illustrate the performance and efficiency of the proposed method in a simulation study, comparing our method to classic methods used for dynamic prediction of non-Gaussian repeated measures. In Section \ref{sec:case_study},
we demonstrate the practical use of this method in our motivation dataset: the minute-level active/inactive diurnal patterns from the 2011-2014 National Health and Nutrition Examination Survey (NHANES). We conclude with a discussion in Section \ref{sec:discussion}.

\section{NHANES minute-level activity indicator}
\label{sec:data}
% motivating example

The National Health and Nutrition Examination Survey (NHANES) is a large study conducted by United States Centers for Disease Control and Prevention (CDC). The study focuses on non-institutionalized US population and covers many aspects of public health, including demographic,
socioeconomic, nutrition and
dietary intake, as well as physical fitness and activity. In this paper, we will use the physical activity monitor (PAM) data from 2011-2014 as a motivation example of dynamically predicting generalized functional data.

The analysis sample includes 8763 participants aged 20 to 80 years old, with an average age of 49.53. 52.05\% (4561)  of the sample are female. The range of BMI is 13.6-82.90 with an average of 29.08. The raw sub-second accelerometry data over a week was summarized at the minute level using the Monitor Independent Movement Summary (MIMS) unit. As a result, a single subject will have 1440 observations over each day. The minute-level MIMS was then dichotomized into minute-level binary activity indicators, where 1 indicates active status and 0 sedentary. The cutoff value of dichotomization was determined to be 10.558, following the suggestions in \citet{preprocess} and \citet{Koster2016}. The threshold value was developed based on its association with other activity measures. Finally, active/inactive status at each minute is summarized across days using the median value, rounded up in the event of a non-integer median. Ultimately, the preprocessing results in 1440 binary observations per participant.  \citet{leroux2022} provides additional details on data preprocessing steps.

Our goal is to predict future (later in the day) probability that an individual is active given their historical (earlier in the day) data. Given the minute-level activity indicators during a certain time interval, we hope to make personalized predictions of the afterward activity pattern that are accurate, efficient and can update as the observed interval extends. Figure \ref{fig:nhanes_dynpred} is an illustration of this problem using the activity indicator tracks of two subjects. The black dots are observed binary indicators, taking on values of either 0 (inactive) or 1 (active), and the dashed vertical lines represent the end of observation. From left to right, the three panels represent the increasing length of the observed interval, respectively 0-6am, 0am-12pm, and 0am-6pm. With these observations earlier in the day, the minute-level probability of activity was predicted and visualized with the red curve, which changes as more observations are collected.

% motivation example figure
\begin{figure}
\centering

\begin{subfigure}{0.9\textwidth}
\includegraphics[width=\textwidth]{Figs/MotiveExp_a.pdf}
\caption{Observed binary activity indicators and predicted probability curves of six participants over one day in the NHANES dataset.}   
\label{fig:nhanes_data}
\end{subfigure}

\begin{subfigure}{0.9\textwidth}
\includegraphics[width=\textwidth]{Figs/MotiveExp_b.pdf}
\caption{Illustration of dynamic prediction of minute-level binary activity indicator using two participants from the NHANES study.}
\label{fig:nhanes_dynpred}   
\end{subfigure}

\caption{Visual illustration of the structure of the NHANES minute-level activity indicator as a motivation example for the dynamic prediction problem. The black dots indicate the observed activity status and the red lines represent the predicted probability of an active minute based on given observations.}
\label{fig:MotiveExp}
\end{figure}

\section{Method}
\label{sec:method}

\subsection{GFPCA}
\label{subsec:gfpca}

We first introduce assumptions and notation that will be used throughout this paper. The data we observe is a series of generalized outcomes collected densely across a functional domain (e.g., time/space), whose distribution can be specified by a continuous latent function. Specifically, let $i=1,\ldots,N$ denote subject. The outcome of interest is a realization of a non-Gaussian random variable ($Y_i(t)$) observed discretely along a domain $t \in T$, generated by a continuous latent function ($\eta_i(t)$). We assume that the outcome follows an exponential family distribution through a link function $g$: $g[E(Y_i(t))] = \eta_i(t)$. Though functional data like this can be observed either densely or sparsely. Here, like our motivating example in Section \ref{sec:data}, we assume that the data are collected densely along a common grid for all participants, ${t_j, j = 1..J}$, along the functional domain $T$. In the NHANES data, $t_j$ is each minute of the day, and $Y_i(t_j)$ is the binary active/inactive indicator associated with the $t_j$ minute.

The GFPCA framework assumes that for each individual $i$, their latent function consists of a population-level functional fixed effect and an individual-specific functional random effect: $g(Y_i(t)) = \eta_i(t) = f_0(t)+ b_i(t), \hspace{0.2cm} t  \in T$. $b_i(t)$, the subject-specific random effect, is a zero-mean Gaussian process characterized by an unknown covariance operator $\Sigma_b$: $\Sigma_b(u, t) = Var(b_i(u), b_i(t))$, $u,t \in T$.

If $b_i$ is highly nonlinear, it can be modeled using standard GLMM software by specifying a sufficiently flexible spline basis for the random effect, $b_i(t) = \sum_{k=1}^K \xi_{ik} \phi_k(t)$ with $\phi_k$ being the basis functions, $K$ indicating the total number of basis functions used for estimation, and $\boldsymbol{\xi}_{i} \sim N(0, \Sigma_\xi)$. However, for most spline bases, the specification of the covariance matrix for the subject-specific coefficients ($\Sigma_\xi$) is not generally known in advance, and an unstructured covariance matrix is most likely appropriate. For highly nonlinear $b_i$, $K$ may be relatively large, resulting in a total of $K \times (K-1)/2$ variance parameters to be estimated. As a result, model estimation is often infeasible using traditional methods for generalized linear mixed models due to the heavy computational burden associated with a large number of variance parameters, especially with a large sample size and high density of observed data (e.g., $N = 4561$ and $J = 1440$ in the NHANES data). Moreover, when the intent is to obtain out-of-sample predictions, the individual random effects of new subjects can not be directly predicted using most existing software without re-estimating part or all of the parameters, which imposes further computational burden.

As an alternative to general purpose spline bases (e.g., B-splines, cubic regresison lines, etc.), one may consider a more parsimonous model which is the basis for GFPCA methodology. By the Karhunen-Loeve theorem, $b_i$ can be represented as an infinite linear combination of orthogonal functions. The GFPCA model then becomes: $
    g(Y_i(t)) = \eta_i(t) = f_0(t)+\sum_{k=1}^{\infty}\xi_{ik}\phi_k(t), \hspace{0.2cm}
    t  \in T$, where the basis functions $\phi_k(t)$ are pairwise orthogonal and the coefficients $\xi_{k}$ are mutually independent random variables. If the function $\eta_i(t)$ is Guassian and the $\phi_k(t)$ are orthonormal, then the subject-specific coefficients $\xi_k$ are independent zero-mean Gaussian random variables:  $\xi_{ik} \sim N(0, \lambda_k)$ with decreasing variance such that $\lambda_1 > \lambda_2 >...$. This expansion is also known as the eigenvector transformation, with $\phi_k(t)$ as 
 eigenfunctions and $\lambda_k$ as eigenvalues. In practice this infinite sum is approximated using a relatively small number of basis functions as higher order bases often contribute relatively little to the percentage of variance explained. The model then becomes:
\begin{equation}
%\begin{aligned}
   \eta_i(t) = f_0(t)+\sum_{k=1}^{K}\xi_{ik}\phi_k(t), \hspace{0.3cm}
    t  \in T.
    \label{eq:gfpca}
%\end{aligned}
\end{equation}

The latent function in Model \ref{eq:gfpca} consists of a
\textit{population} mean function $f_0(t)$ and an \textit{individual-specific} process $\sum_{k=1}^{K}\xi_{ik}\phi_k(t)$, the latter is a linear combination of $K$ functional principal components (FPC) $\phi_k(t)$. The FPCs are shared across population, but the coefficients $\xi_{ik}$ is specific to each subject. If we make an analogy to GLMM, then $f_0(t)$ can be perceived as a functional ``fixed effect", and $\xi_{ik}$ are similar to the ``random slopes" in the random effect. Because the $\xi_{ik}$ are independent within subjects, the number of variance parameters to be estimated has now been reduced from $K \times (K-1)/2$ to $K$, a substantial reduction in computational complexity. However, the primary methodologic challenge is how to estimate the eigenfunctions.

Many methods of parameter estimation for Model \ref{eq:gfpca}  have been proposed. However, there is a lack of straightforward implementation in practice, especially for large datasets with complex structures. Many existing methods are limited due to computational efficiency and/or model flexibility. For example, \citet{hall2018} modelled the generalized outcomes using a latent Gaussian process model, with experimentation on small, sparse dataset. The method assumes variation of the latent process about its population mean is relatively small. 
\citet{gertheiss2017} proposed and compared a two-stage estimation approach and a Bayesian joint estimation approach, with a simulation on sparse repeated measures generated from a latent process composed of two eigenfunctions. The generation mechanism is limited in both scale and flexibility. \citet{Li2018} proposed an efficient algorithm for FPCA on exponential data, under the assumption that a low-rank Singular Value Decomposition (SVD) exists of the natural parameters. The density of observation is greater in this paper compared to the previous ones, with 50 repeated measures from each subject. However, it is still far below the NHANES activity indicator. Its performance and efficiency on large, dense datasets remains to be explored. In terms of implementation, the software or package development of these proposed methods has been slow. Examples include a variational EM algorithm proposed by \citet{wrobel2019} for binary functional data, accompanied by the \textit{registr} package on GitHub \citep{Wrobel2018}.

\subsection{fGFPCA}
\label{subset:fGFPCA}

Despite the lack of efficient estimation or straightforward implementation of fGFPCA on generalized functional data, for Gaussian functions there exists fast implementation of esimtation methods, such as the Fast Covariance Estimation (FACE) algorithm \citep{face}. \citet{fGFPCA} combines the computational efficiencies of FACE with data binning to create a fast implementation of GFPCA (fGFPCA). The method speeds up the computation significantly by estimating the latent function $\eta_i$ on a binned grid slightly sparser than the observation grid, followed by an application of fast methods for FPCA (e.g., FACE) to the binned data to smooth across bins. A final normalization step follows to obtain estimates on the non-binned domain.

The fGFPCA algorithm is laid out as the following steps:

\begin{enumerate}
    \item Bin observations

Here we bin the densely collected data into small, non-overlapping intervals with equal length $w$. We denote the index of bins as $s = 1,...,S$, and the index of the midpoint of the sth bin as  $m_s$. The sth bin includes observations collected over $(t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}]$.

The binning step assumes neighboring observations are generated from the same value of latent function $\eta_i(t_{m_s})$, thus allowing us to pool information from neighboring observations and estimate $\eta_i(t_{m_s})$ at the expense of minor misspecification on a slightly sparser grid. we Note that this paper set up bins as equally-spaced for simplicity. However, this method can be applied to unequally spaced bins as well.

    \item Fit local intercept-only generalized mixed models within each bin

To estimate the latent function $\eta_i(t_{m_s})$, we fit a local Generalized Linear Mixed Model (GLMM) within every bin. These models include only a fixed and a random intercept, that is, at the $s^{\text{th}}$ bin: 
\begin{equation}
    g[E(Y_i(t_j))] =\eta_i(t_{m_s})= \beta_0(t_{m_s})+b_i(t_{m_s}),
    \hspace{0.5cm}t_j \in (t_{m_s}-\frac{w}{2}, t_{m_s}+\frac{w}{2}].
    \label{eq:local_glmm}
\end{equation}

With Model \ref{eq:local_glmm} above, an individual-specific intercept $\hat{\eta}_i(t_{m_s})$ can be estimated within each bin. Connecting these values from the same subject across all bins, we will get the estimated latent functions  $\hat{\eta}_i$ for each individual on the binned grid $S$.

    \item Implement FPCA

Since the estimated latent function $\hat{\eta}_i(t_{m_s})$ is continuous, we could apply the FPCA algorithm and fit the following model: 
\begin{equation}
    \hat{\eta}_i(t_{m_s}) = f_0(t_{m_s})+\sum_{k=1}^K\xi_{ik}\phi_{k}(t_{m_s})+\epsilon_i(t_{m_s}).
    \label{eq:fpca}
\end{equation}

The FPCA model produces estimates of smooth FPCs $\hat{\phi}_k(t_{m_s})$. The
number of PC functions, $K$, is often determined by the cumulative percentage of variance explained. \citet{zhou2024} also proposed selection using sum of squared 2nd-order differences (SSSOD).

    \item Re-evaluate parameters using a global GLMM

As mentioned in step 1, the mis-specification of local models \ref{eq:local_glmm} will lead to biased estimates, including variance of coefficients $\hat{\lambda}_k$ and population mean $\hat{f}_0$. This step corrects this bias by fitting a global GLMM, re-estimating $f_0$ as the fixed-effect and $\lambda_k$ as the random-effect. We use the term ``global" as it is uses all the observations on the original observation grid, unlike the ``local" models in Step 2 which use only obervations within a bin.

Note that the PC functions $\hat{\phi}_k$ from Step 3 are estimated on the binned grid. To fit the global model on the observations grid, we need to extend the $\hat{\phi}_k(t_{m_s})$ back to the observation grid $\hat{\phi}_k(t_j)$. Many methods are applicable here since the estimated PC functions are not biased. In this paper, we project grid S back to grid J using regression on a rich set of spline basis functions: $
g[E(Y_i(t_j))|\hat{\phi}_k(t_j)] = f_0(t_j)+\sum_{k=1}^K \xi_{ik}\hat{\phi}_k(t_j)
$.

From this model, parameter estimates that will be used later for dynamic prediction are obtained, including the population mean $\hat{f}_0(t)$ and variance of coefficients $\hat{\lambda}_k$. This model could be efficiently fit using the \textit{bam} function in the \textit{mgcv} packages in R. 
\end{enumerate}

In addition to fGFPCA, alternative methods exist to obtain the parameter estimates to be used for dynamic prediction. One potential choice was recently proposed by \citet{weishampel2023}. In this paper, each binary function is ``smoothed" separately using a general additive model (GAM), and FPCA was then fit on the collection of smoothed tracks. This method can be efficient in some scenarios, but does not work well for our motivation example of the physical activity data. Since a large proportion of subjects are sedatary at night, the hours of consecutive zeros can cause identifiability issues when fitting the individual GAM models. This shortcoming is illustrated in \citet{leroux2022} through a detailed comparison between this method and fGFPCA on the NHANES data. Another option is the \textit{registr} package by \citet{Wrobel2018} as we mentioned in the Introduction. It can achieve similar efficiency as fGFPCA \cite{fGFPCA}, but its application is restricted to binary data and cannot be directly applied to other exponential family distributions.

\subsection{Dynamic prediction}
\label{subsec:dyn_pred}

In this section we show how to use the results obtained from fGFPCA to obtain dynamic predictions. We formulate the problem in terms of partitioning the functional data along the domain in two parts and using data from one element of the partition to predict a participant's latent function in the other element of the partition. This corresponds to using historical data to predict the future if the functional domain is time. Note that this is our framework, but methods are generally applicable to any combination of observed data partition where data are either observed or unobserved.

Assume we have a subject, $u$, which was not used for model fitting. For this subject, we only observe the outcome along a portion of the domain: $Y_u(t_j)$, $j=1,...,J_u$ and $J_u < J$. With the population level estimates obtained from  Step 4 described in Section \ref{subset:fGFPCA}, including population mean $\hat{f}_0$, FPCs $\hat{\phi}_k$ and variance of individual scores $\hat{\lambda}_k$, the prediction at any point $t_j$ beyond the observed track ($j > J_u$) gets down to the estimation of out-of-sample coefficient $\hat{\xi}_{uk}$, as follows:

\[
\begin{aligned}
    \hat{\eta}_u(t_j)|\hat{\mathbf{\Theta}} &= \hat{f}_0(t_j)+\sum_{k=1}^K \hat{\xi}_{uk}\hat{\phi}_k(t_j)
    \hspace{0.3cm}  J_u < j < J  \\
    \hat{\mathbf{\Theta}} & = \{\hat{f}_0(t), \hat{\phi}_k(t), \hat{\lambda}_k, k = 1...K \}.
\end{aligned}
\]

Since the outcome $Y_u(t_j)$ follows an exponential family distribution characterized by the latent process $\eta_u(t_j)$, it is straightforward to write out the posterior log-likelihood of the observed track:

\[
\begin{aligned}
    l_u = l(\mathbf{Y}_u | \hat{\boldsymbol{\xi}}_u, 
    \hat{\mathbf{\Theta}}) &=\sum_{j<J_u}log(h(Y_u(t_j)))+\eta_u(t_j)T(Y_u(t_j))-log(A[\eta_u(t_j)])\\
    \eta_u(t_j) & = \hat{f}_0(t_j)+\sum_{k=1}^K\xi_{uk}\hat{\phi}_k(t_j)\\
    \mathbf{Y}_u &= \{Y_u(t_j), j = 1,...,J_u \}, \hspace{0.3cm} \hat{\boldsymbol{\xi}}_u = \{\hat{\xi}_{uk}, l = 1,..., K \}\\
\end{aligned}
%\label{eq:pos_llh}
\]

The form of $h(.)$, $T(.)$, and $A(.)$ depends on the specific type of exponential family distribution. We also know the prior distribution of coefficients: $\xi_{uk} \sim N(0, \hat{\lambda}_k)$. Therefore, the maximum likelihood estimator of the subject-specific coefficients can be obtained as:

\[
    \hat{\boldsymbol{\xi}}_u = 
    \underset{\boldsymbol{\xi}_u}{\mathrm{argmin}} \ L(\boldsymbol{\xi}_u|\boldsymbol{Y}_u, 
    \hat{\mathbf{\Theta}}) = \frac{p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\mathbf{\Theta}})p(\boldsymbol{\xi}_u|\hat{\mathbf{\Theta}})}{\int p(\boldsymbol{Y}_u|\boldsymbol{\xi}_u, \hat{\mathbf{\Theta}})p(\boldsymbol{\xi}_u|\hat{\mathbf{\Theta}})d\boldsymbol{\xi}_u}
\]
With generalized outcomes, the likelihood above often does not have a closed-form solution. In such cases, we could either apply numeric maximization or use methods of approximation. For example, in R, the likelihood can be optimized using MCMC sampling and the L-BFGS (limited Broyden–Fletcher–Goldfarb–Shanno) algorithm from the \textit{rstan}. Alternative methods, such as Laplace Approximation, are also available though we found them to be numerically unstable in our data application and thus do not explore them further here.

The construction of prediction intervals is also straightforward according to likelihood theory. The covariance matrix of MLE is asymptotically the inverse of observed information matrix. We are able to get the pointwise posterior variance estimates of the latent function by taking the second derivative of the posterior likelihood conditioning on $\hat{f}_0(t_j)$, $\hat{\phi}_k(t_j)$ and $\hat{\lambda}_k$, as follows:

\[
Var(\hat{\eta}_u(t_j)|\hat{\mathbf{\Theta}})  = Var(\hat{f}_0(t_j)+\sum_{k=1}^K\hat{\phi}_k(t_j)\hat{\xi}_{uk})
=\sum_{k=1}^K \hat{\phi}_k^2(t_j)\hat{\lambda}_k,
\hspace{0.3cm}  J_u < j < J.
\]

Wald prediction interval could be constructed with this variance estimator at each point along the functional domain. In practice, when using Bayesian methods for parameter estimation, we could also construct a ``credible" prediction interval by taking the sampling quantiles. Given $\hat{\mathbf{\Theta}}$, the two intervals are very similar numerically in our simulations and data application, though this may not be true in general.

\section{Simulation study}
\label{sec:simulation}

\subsection{Data generation scheme}
\label{subser:gen_data}

In this section, we implement a simulation study to illustrate the predictive performance and
computational efficiency of the proposed method. The functional outcome is generated from a cyclic latent process as follows:

\[
\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}), \hspace{0.3cm} t  \in [0, 1] \\
\eta_i(t) &=\xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)\\
\end{aligned}
\]

The latent process has a zero mean for simplicity ($f_0(.) = 0$). The true FPCs $\phi_k(t) = \{\sqrt{2}sin(2\pi t), \sqrt{2}cos(2\pi t), \sqrt{2}sin(4\pi t), \sqrt{2}cos(4\pi t) \}$. Coefficients are generated independently from zero-mean Gaussian distribution with decreasing variance: $\xi_{ik} \sim N(0, 0.5^{k-1})$, $k=1, 2, 3, 4$. For each subject, we generate 1000 observations ($J = 1000$) along $t \in [0, 1]$ on a regular grid. Every simulated dataset is split into the ``training set" used for model estimation, and a ``test set" for evaluating out-of-sample model performance. That is, the model estimated using only the training set is used to predict future observations of subjects in the test set based on their partially observed tracks. The proposed method is compared to existing methods that are commonly used for similar purposes. However, the density of the simulated data is high and beyond the capacity of most these methods (see Section \ref{subsec:compete_methods} for details). Therefore we set up two scenarios with different training sample size: 1) a larger training set of 500 subjects ($N_{train} = 500$); and 2) a smaller training set of 100 subjects ($N_{train} = 100$). The test sample sizes of the two scenarios are the same ($N_{test} = 100$). In the second scenario, we also need to reduce the measurement density, taking only one every ten observations and discarding the rest.

\subsection{Competing Approaches}
\label{subsec:compete_methods}

In addition to fGFPCA, we consider two competing methods as a basis for comparison.

\subsubsection{GLMMadaptive}
\label{subsubsec:adglmm}

``GLMMadaptive" is the abbreviation of ``Generalized Linear Mixed Models using Adaptive Gaussian Quadrature". Developed by \citet{Rizopoulos2017}, this method fits a generalized mixed effect model and obtains estimates by approximating the integral over the random effects using the adaptive Gauss-Hermite quadrature rule, accompanied by an R package \textit{GLMMadaptive}. Compared to traditional methods in $\it{lme4}$ and $\it{nlme}$, the approximation has improved the efficiency substantially and offers functionality for making dynamic predictions. According to \citet{Stringer2023}, \textit{GLMMadaptive} can fit the same model 2-4 times faster than \textit{lme4} without compromising performance. However, with the measurement density of our simulated data, this method is still not ``fast" enough when using models with complex, non-linear subject-specific effects. That is, GLMMadaptive is only feasible (within reasonable time) at the cost of model flexibility. With this restriction, GLMMadaptive is only able to estimate a random intercept and slope model on the ``larger" training dataset ($N_{train}=500, J=1000$). Specifically, the model:
\begin{equation}
g(E(Y_i(t_j))) = \beta_0+\beta_1t_j+b_{i0}+b_{i1}t_j.   
    \label{eq:glmm_linear}
\end{equation}

This model essentially has a linear structure and is clearly not sufficiently flexible to adequately model the complex data generation mechanism. Therefore, we would expect low predictive accuracy compared to the proposed method.

In the second scenario, the smaller training size makes it possible to increase the flexibility of the GLMMadaptive model by incorporating a small set of spline basis functions:

\begin{equation}
    g(E(Y_i(t_j))) = \sum_{k=1}^4\zeta_{k}B_k(t_j)+\sum_{l=1}^4\xi_{il}\phi_l(t_j),
    \label{eq:glmm_sp}
\end{equation}

where the $\phi_l(.)$ are natural cubic splines.

However, to reduce the model fitting time within a reasonable range, we still need to reduce the observations density by taking only one out of ten neighboring observations. The means 90\% of the observations are discared the the informaiton within cannot be utilized.

\subsubsection{GFOSR}
\label{subsubsec:gfosr}

Generalized Function on Scalar Regression (GFOSR) estimates the association between a functional response and scalar covariates {\color{purple}\citep{Reiss2010, fda2005, Bauer2018}}. In our context, this framework allows for directly modeling the future trajectories (functional outcome) explicitly conditional on the scalar covariates (historical/observed data). Specifically, we predict future trajectories using several of the most recent observations from the observed track. The concept is similar to a distributed lag model approach \citep{Baek2022}, though the coefficients are functional and not necessarily decaying. For a new subject $u$, if the observed track stopped at $t_m$ ($t_m < T$, $m < J$), and we use the last $L$ observations as predictors, then the model used for predicting the future can be rigorously expressed as: 
\begin{equation}
g(E[Y_i(t_j)|Y_i(v), v\leq t_m]) = \beta_0(t_j) +\sum_{l} \beta_{m-l+1} (t_j)Y_i(t_l), \hspace{0.3cm}
l  = m,...,m-(L-1),  \hspace{0.3cm}
t_j > t_m 
\label{eq:gfosr}  
\end{equation}

The larger L is, the more flexible this model can be, though without penalization, multicollinearity can be an issue. Here we limit our investigation to two different values: $L = 1$ or $L = 5$.

In our case, since the predictor (historical data) is itself non-Gaussian functional data, we could also consider a Function-on-Function regression framework \citep{Morris2015, fdaR2024,Goldsmith2011}. The FOSR and FOFR frameworks are both quite flexible, but for the latter, we are not aware of established methods for prediction where both the predictor and outcome are generalized functions. Here, we choose to use the FOSR approach.

\subsection{Evaluation metrics}

Out-of-sample predictive performance is evaluated with two metrics: 1) Integrated Squared Error (ISE) and
Area Under the Receiver Operating Curve (AUC). ISE is the squared error of the predicted latent function over an interval of t: $ISE_i = \sum_{t_j}(\hat{\eta}_i(t_j)-\eta_i(t_j))^2$. It assesses the
prediction accuracy of a continuous latent process. 
The second metric, AUC, focuses on comparing the predicted latent function to the observed binary outcome. Both metrics are reported as average across subjects and simulated datasets.

The uncertainty of the predictions are evaluated by prediction intervals and presented in the Supplement. For the proposed method fGFPCA, we calculated the credible interval using the sampling quantiles of latent function at each measurement point. This interval directly expresses the probability that the true value of $\eta$ lies within the range. For the completing methods, GFOSR and GLMMadaptive, Wald intervals are calculated using the standard error estimates obtained directly from the corresponding packages.

\subsection{Results}

\subsubsection{Large training set}
\label{sss:sim_result_large}

Here we fit the fGFPCA model on a larger training set with 500 subjects and compare its predictive performance on an additional 100 subjects to the simple linear GLMMadaptive model \ref{eq:glmm_linear} and the GFOSR model \ref{eq:gfosr} with $L=1$ and $L=5$. In the test set, future predictions are made based on observations given up to certain cut-off points $t_{J_u} \in \{0.2, 0.4, 0.6, 0.8\}$. That is, given observations along 0-0.2; prediction is made on 0.2-1; given observations along 0-0.4, prediction is made on 0.4-1, and so forth. The prediction performance is presented by equal length intervals: (0.2, 0.4], (0.4, 0.6], (0.6. 0.8] and (0.8, 1], so that the predictive performance is comparable on an equal number of observations and across common grids.

Figure \ref{fig:sim1} presents an example of predictions for two randomly selected test subjects from one of the 500 simulated datasets. As the figure reveals, the four methods provide predictions with very different degrees of nonlinearity. fGFPCA results in the most flexible predicted tracks and is largely capable of capturing the individual and cyclic patterns of latent functions. The shape of GFOSR prediction depends on the number of observations used. With the last five observations, the predicted curves resemble quadratic or cubic functions without periodicity. When using only the last observation, the predictions hardly reveal any individual-specific deviation from the population. This can be explained by the model formulation of GFOSR in Equation \ref{eq:gfosr}. Since this model does not include random effects and there are only two possible outcomes for each observation (0 or 1), there would be only two unique predicted tracks across the test sample when $L=1$. Finally, the GLMMadaptive model results in monotone predictions due to the linear random effects structure on the log-odds scale. We reiterate that this is a computational limitation of the current implementation in our simulation scenario and not a limitation of the method in general.

ISE and AUC are summarized in Table \ref{tab:sim_large}, with lower values of ISE and higher values of AUC representing better predictive performance. fGFPCA provided the best predictive performance in all scenarios and won over the other three methods by a large margin. The improvement of ISE across all scenarios ranges from 13.1\% to 97.3\%, with an average improvement of 63.9\%. AUC improved from 4.5\% to 48.1\%, with an average of 24.7\%. Ideally for dynamic prediction, the accumulation of additional data would result in improved performance. As the maximum observation time extends from 0.2 to 0.8, the predicted tracks from the proposed method move closer to the true latent functions. The other methods, on the other hand, do not change much as observation tracks extend. The more flexible fGFPCA on the left side in Figure \ref{fig:sim1} clearly shows this behavior. Table \ref{tab:sim_large} shows this result holds across subjects and simulated datasets. When comparing the same prediction window across different maximum observation times (rows), ISE decreases significantly, and AUC increases for fGFPCA as the observation track gets longer. However, the other three methods did not show such improvement. GLMMadaptive actually resulted in worse performance with more observations.

Regarding computation time, GFOSR is the fastest among the four models, taking about 1.5 seconds for each simulated dataset when using the last observation and 9 seconds when using the last five observations. However, the low accuracy of the predictions limits the appeal of very fast computation in this context. In contrast, fGFPCA takes approximately 5.8 minutes for each simulated dataset, while the linear GLMMadaptive runs slightly faster at 3.35 minutes on average.

\subsubsection{Small training set}
\label{sssec:sim_result_small}

In this section, we present the results when the training sample size and measurement density are reduced to accommodate a more flexible GLMMadaptive estimated with a nonlinear subject-specific random effect using a low rank natural spline basis. The model Equation \ref{eq:glmm_sp} still takes a very long time to fit on the full gird even with smaller training set. Therefore, as mentioned in Section \ref{subser:gen_data}, we reduced the data density by taking one out of every 10 neighboring observations, causing 90\% of the observations discarded. The out-of-sample prediction, on the other hand, is on the original measurement grid. The other three methods (fGFPCA and GFOSR with the last one or five observations) did not need data reduction and utilized all information in the training samples.

We also note that the GLMMadaptive model is not as numerically stable as the other three methods since consecutive zeros or ones can cause extreme values of parameter estimation. As a result, the mixed model occasionally runs into convergence issues. The simulated datasets with numeric problems are discarded. In the summary of predictive performance in Table \ref{tab:sim_small}, the AUC and ISE of GLMMadaptive are averaged over 496 datasets without numeric problems. The other three methods did not suffer from such issues and thus averaged over all 500 datasets.

Figure \ref{fig:sim2} presents predictions for two participants in the same format as~\ref{fig:sim1}. As a result of increased model flexibility for GLMMadaptive, individual prediction tracks on the right column in Figure \ref{fig:sim2} are much more nonlinear when compared to the previous section, and also approach the true latent function as observed tracks are updated. However, predictions from this model are still considerably worse than fGFPCA, as seen in Table \ref{tab:sim_small}, with fGFPCA resulting in 11.1\% - 94.6\% (average 61.5\%) improvement in ISE and 0.9\% - 45.1\% (average 19.5\%) improvement in AUC across all scenarios. Computationally, since the size of the training data was reduced, both fGFPCA and GFOSR spent less time on model fitting. On average, fGFPCA took 5.23 minutes, while GFOSR took 0.6 and 4.4 seconds for $L=1$ and $L=5$  respectively. However, for GLMMadaptive, the increased flexibility offset computational gains associated with a decreased sample size, taking 6.35 minutes to fit on average.

% figure
\begin{figure}
    \centering

    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{Figs/simN500.pdf}
    \caption{Training sample size = 500}
    \label{fig:sim1}
    \end{subfigure}

    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{Figs/simN100.pdf}
    \caption{Training sample size = 100}
    \label{fig:sim2}
    \end{subfigure}

    \caption{Individual out-of-sample dynamic prediction of the latent process of two randomly selected subjects from the simulated data . The top panel shows prediction from model fit on 500 subjects, and the bottom panel 100 subjects. The black dots indicate observed outcome and black dashed line the true latent function. Solid colored lines are the predicted track, with the color of lines indicating different maximum observation time. }
\end{figure}

% tables
\begin{landscape}\begin{table}

\begin{subtable}{\textwidth}
\small
\centering
\begin{tabular}{lrrrrrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{16}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-17}
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GFOSR (L=5)} & \multicolumn{4}{c}{GFOSR (L=1)} & \multicolumn{4}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13} \cmidrule(l{3pt}r{3pt}){14-17}
Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
\midrule
\addlinespace[0.3em]
\multicolumn{17}{l}{\textbf{ISE}}\\
\hspace{1em}(0.2, 0.4] & 184.19 &  &  &  & 274.95 &  &  &  & 362.48 &  &  &  & 387.71 &  &  & \\
\hspace{1em}(0.4, 0.6] & 218.57 & 113.45 &  &  & 277.44 & 220.21 &  &  & 286.61 & 262.55 &  &  & 291.58 & 269.80 &  & \\
\hspace{1em}(0.6, 0.8] & 274.57 & 76.51 & 27.07 &  & 322.34 & 373.11 & 325.02 &  & 385.70 & 410.51 & 389.31 &  & 315.78 & 282.74 & 278.24 & \\
\hspace{1em}(0.8, 1.0] & 119.59 & 106.93 & 25.46 & 16.41 & 290.98 & 318.08 & 350.85 & 333.58 & 328.48 & 341.27 & 354.21 & 347.07 & 563.01 & 477.49 & 597.75 & 600.34\\
\addlinespace[0.3em]
\multicolumn{17}{l}{\textbf{AUC}}\\
\hspace{1em}(0.2, 0.4] & 0.737 &  &  &  & 0.686 &  &  &  & 0.624 &  &  &  & 0.591 &  &  & \\
\hspace{1em}(0.4, 0.6] & 0.645 & 0.711 &  &  & 0.563 & 0.630 &  &  & 0.543 & 0.590 &  &  & 0.524 & 0.596 &  & \\
\hspace{1em}(0.6, 0.8] & 0.699 & 0.779 & 0.799 &  & 0.669 & 0.628 & 0.676 &  & 0.604 & 0.577 & 0.615 &  & 0.669 & 0.694 & 0.687 & \\
\hspace{1em}(0.8, 1.0] & 0.736 & 0.741 & 0.779 & 0.783 & 0.626 & 0.606 & 0.552 & 0.584 & 0.588 & 0.564 & 0.537 & 0.551 & 0.514 & 0.556 & 0.526 & 0.564\\
\bottomrule
\end{tabular}
\caption{Training sample size = 500}
\label{tab:sim_large}
\end{subtable}

\hfill

\begin{subtable}{\textwidth}
\small
\centering
\begin{tabular}{lrrrrrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{16}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-17}
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GFOSR (L=5)} & \multicolumn{4}{c}{GFOSR (L=1)} & \multicolumn{4}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){10-13} \cmidrule(l{3pt}r{3pt}){14-17}
Window & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8 & 0.2 & 0.4 & 0.6 & 0.8\\
\midrule
\addlinespace[0.3em]
\multicolumn{17}{l}{\textbf{ISE}}\\
\hspace{1em}(0.2, 0.4] & 191.05 &  &  &  & 286.85 &  &  &  & 367.28 &  &  &  & 374.47 &  &  & \\
\hspace{1em}(0.4, 0.6] & 224.08 & 118.19 &  &  & 286.63 & 227.23 &  &  & 290.78 & 265.81 &  &  & 268.20 & 462.62 &  & \\
\hspace{1em}(0.6, 0.8] & 283.93 & 81.57 & 30.16 &  & 328.61 & 374.57 & 326.52 &  & 386.71 & 408.43 & 388.43 &  & 319.28 & 286.71 & 233.55 & \\
\hspace{1em}(0.8, 1.0] & 124.41 & 114.15 & 28.87 & 18.95 & 301.17 & 324.82 & 353.89 & 335.14 & 331.49 & 342.94 & 354.80 & 347.63 & 228.24 & 469.24 & 331.53 & 131.91\\
\addlinespace[0.3em]
\multicolumn{17}{l}{\textbf{AUC}}\\
\hspace{1em}(0.2, 0.4] & 0.734 &  &  &  & 0.679 &  &  &  & 0.620 &  &  &  & 0.672 &  &  & \\
\hspace{1em}(0.4, 0.6] & 0.644 & 0.710 &  &  & 0.549 & 0.627 &  &  & 0.536 & 0.590 &  &  & 0.622 & 0.671 &  & \\
\hspace{1em}(0.6, 0.8] & 0.695 & 0.775 & 0.800 &  & 0.656 & 0.617 & 0.668 &  & 0.598 & 0.574 & 0.613 &  & 0.689 & 0.698 & 0.730 & \\
\hspace{1em}(0.8, 1.0] & 0.735 & 0.739 & 0.778 & 0.782 & 0.611 & 0.588 & 0.551 & 0.578 & 0.580 & 0.556 & 0.536 & 0.551 & 0.680 & 0.629 & 0.678 & 0.743\\
\bottomrule
\end{tabular}

\caption{Training sample size = 100}
\label{tab:sim_small}
\end{subtable}

\caption{Predictive performance of models fit on simulated data. Both the Integrated Squared Error (ISE) and AUC are averaged across all simulated datasets without numeric issues.}
\label{tab:sim}
\end{table}
\end{landscape}

\section{Case study}
\label{sec:case_study}

This section explores the application of the proposed method to a real-world dataset. Specifically, we conducted a case study is on the NHANES 2011-2014 minute-level active/inactive indicators introduced in Section \ref{sec:data}. We aim to predict active/inactive states later in the day using active/inactive status observed earlier in the day. Similar to the simulation study, the sample was split into a training set ($N_{train}$ = 5257) and a test set ($N_{test}$ = 3506). The models are fit in the training set and then evaluated in the test set. Since we do not know the true latent functions in this case, predictive performance is evaluated using AUC alone. For these data, we were only able to fit the GLMMadaptive model as in \eqref{eq:glmm_linear}.

Figure \ref{fig:appl} presents the individual prediction tracks of four randomly selected participants in the test set across different methods. The figure follows a similar format as Figures \ref{fig:sim1} and \ref{fig:sim2}, with red, green, and blue lines respectively indicating prediction obtained given observation from 0-6am, 0am-12pm and 0am-6pm. As expected, fGFPCA, shown in the left column, resulted in more nonlinearity in the predicted tracks and increased variability across individuals, neither of which is observed in predictions obtained from the three other methods. As in the simulation study, GLMMadaptive, shown in the right column, results in monotone predictions as it only includes a linear time effect. GFOSR provides more non-linear predictions but with substantially less heterogeneity between individuals when compared to fGFPCA. Additionally, the predictions from fGFPCA updates substantially as observations are accumulated, visually seen as the different shapes of tracks of different colors for the same prediction time. This behavior is less apparent in the other methods considered here, with mean shifts largely characterizing the difference in prediction tracks.

Table \ref{tab:appl} summarizes the AUC across different methods. This table is structured similarly to Tables \ref{tab:sim_large} and \ref{tab:sim_small}, where AUC is calculated on equal-length time windows for comparability (6am-12pm, 12pm-6pm, 6pm-12am). As the table shows, fGFPCA has the highest AUC in any situation, with the largest improvement seen for prediction windows further away from the observation window. For example, given data from 0am-6am, the AUC of fGFPCA is 0.118 higher than GLMM on 6am-12pm, but 0.192 higher on 6pm-12am. The improvement of predictive accuracy of fGFPCA is also more rapid as more observations are collected. In fact, the GFOSR and GLMM models showed worse performance as the prediction window moves away from the observation window, while fGFPCA declined and then improved. The latter finding is intuitively consistent with the fact that physical activity exhibits cyclic diurnal patterns. From these results, we conclude that fGFPCA is better capable of capturing and predicting the heterogeneous, individualized and nonlinear latent processes that characterize active/inactive patterns across the day.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{Figs/DataAppl.pdf}
\caption{Individual out-of-sample prediction tracks of the latent process of the NHANES data. The black dots indicate observed outcomes. Solid colored lines are the predicted track, with the color of lines indicating different maximum observation time. }
\label{fig:appl}
\end{figure}

\begin{table}
\small
\centering
\begin{tabular}{lrrrrrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{12}{c}{Maximum observation time} \\
\cmidrule(l{3pt}r{3pt}){2-13}
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{fGFPCA} & \multicolumn{3}{c}{GFOSR (L=5)} & \multicolumn{3}{c}{GFOSR (L=1)} & \multicolumn{3}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7} \cmidrule(l{3pt}r{3pt}){8-10} \cmidrule(l{3pt}r{3pt}){11-13}
Window & 6am & 12pm & 6pm & 6am & 12pm & 6pm & 6am & 12pm & 6pm & 6am & 12pm & 6pm\\
\midrule
6am-12pm & 0.699 &  &  & 0.689 &  &  & 0.684 &  &  & 0.581 &  & \\
12pm-6pm & 0.534 & 0.711 &  & 0.519 & 0.699 &  & 0.520 & 0.654 &  & 0.532 & 0.701 & \\
6pm-12am & 0.706 & 0.689 & 0.774 & 0.677 & 0.678 & 0.723 & 0.677 & 0.676 & 0.708 & 0.514 & 0.565 & 0.626\\
\bottomrule
\end{tabular}
\caption{Predictive performance of models fit on the NHANES data}
\label{tab:appl}
\end{table}

\section{Discussion}
\label{sec:discussion}

In this paper, we proposed a fast, scalable method for dynamically predicting generalized functional data. Compared with existing methods, using generalized linear mixed model software, the proposed method offers substantial computational efficiencies and thus can be applied to larger samples with high density functional data. The prediction it makes is also much more dynamic, personalized and accurate, thus achieves the essential goals of dynamic prediction. The simulation study and data application presented in Sections \ref{sec:simulation} and \ref{sec:case_study}, respectively, illustrate the proposed approach's feasibility and improved performance in comparison to other methods when data exhibit highly non-linear latent trends.

Despite the clear improvements in predictive performance and the computational efficiency of fGFPCA, a few open questions remain. In both the simulation study and data application, the tuning parameter for the fGFPCA approach, bin width, was chosen arbitrarily. In reality, such choices can affect the predictive performance and sensitivity to the choice of bin should be assessed in practice. If the bins are very narrow local GLMMs may be nonidentifiable, resulting in bins with no point estimates. We expect this to happen when bins have a very large proportion of only $0$ or $1$ observed for binary data. Moreover, the current methodology has not been extended to outcomes measured on a sparse or irregular grid, though this is possible in principle. Other methods for GFPCA mentioned in Section \ref{sec:introduction} may be sufficient to deal with these kinds of small, sparse data. One could then use the dynamic prediction framework presented in the current work.

While fGFPCA showed an advantage in computational efficiency over existing methods based on GLMM, the GLMM framework has the benefit of being able to easily add covariates to the model (e.g., demographic information). An extension of the fast GFPCA methodology used here to covariate-adjusted models is under active development, and we expect our dynamic prediction framework to directly apply once the methodology has been developed for model estimation.

Despite the limitations of the current work, our proposed method works fast and well for the type of data it was designed for: large, high-density functional data. The current work also opens a number of exciting avenues for future work on different types of functional data. For example, if multiple functions are collected for each individual (e.g., multiple days of minute-by-minute physical activity record, where each day is a single function), then a multilevel GFPCA framework may be more appropriate. We are actively developing a similar prediction framework for multilevel generalized functional data using recent work extending fGFPCA to multilevel data. Another potential future direction is the dynamic prediction of multivariate functional data. Here, different functional outcomes are collected from the same subjects, which are often correlated with each other. For example, activity count and heart rate can be measured by wearable devices at the same time. The future measures of one function can be predicted not only by its own past record but also by the other functions. We leave these exciting extensions as a promising area of future research.

\bibliography{refs}

% Supplement

\newpage

\renewcommand{\thesection}{Supplement \Roman{section}:}
\renewcommand{\thefigure}{S.\arabic{figure}}
\setcounter{section}{0}
\setcounter{figure}{0}
\section{Prediction interval}

The Figure \ref{fig:sim_ci} below presents the coverage rate of the 95\% prediction interval along the unobserved track. The prediction interval of fGFPCA is derived from the pointwise sampling quantiles of the estiamted latent function at each measurement point. On the other hand, the prediction intervals of the other three methods other are Wald intervals calculated by the estimated standard deviation. Coverage is averaged pointwise across all simulated datasets, and the color of curves indicates different length of observed track. Figure \ref{fig:simN500_ci} on the top panel visualizes the output of the first simulation scenario, where training sample size is 500 and the GLMM model is linear. Figure \ref{fig:simN100_ci} on the bottom panel visualizes the output of the second simulation scenario, where training sample size is 100 and the GLMM model incorporates spline basis functions.

Comparatively, fGFPCA achieves the highest coverage rate across the predicted track among the four models, and the coverage rate improves as the observed track extends. However, it did not reach the nominal value of 95\% for a large proportion of the track. The other three methods seem to severly underestimate the uncertainty of prediction, causing the coverage rate to be close to zero. GLMM model has a better coverage when with greater model flexibility, but still not as well as the fGFPCA intervals.

Improvement of the coverage rate of prediction interval from fGFPCA is an interesting topic for future research. The current interval is a posterior estimate conditional on parameters estimated from fGFPCA algorithm, thus fails to take into account the variability from the population-level estimates from fGFPCA algorithm. This is likely the cause of the coverage rate being lower than nominal level. However, quantiying such variation is a non-trival problem, especially the variation of estimated functional principal components.

\begin{figure}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{Figs/simN500_ci.pdf}
    \caption{Training sample size = 500}
    \label{fig:simN500_ci}
    \end{subfigure}

    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{Figs/simN100_ci.pdf}
    \caption{Training sample size = 100}
    \label{fig:simN100_ci}
    \end{subfigure}

    \caption{Converage rate of prediction interval averaged across all simulated datasets without numeric issue.}
    \label{fig:sim_ci}

\end{figure}

\end{document}

### 2024-12-17 Code/Figures/FiguresAppl.R

@@ -85,12 +85,14 @@ bind_rows(
   pred_nhanes_adglmm %>% filter(id %in% rand_id) %>%
     mutate(method = "GLMMadaptive"),
   pred_appl_gfosr_l1 %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=1)"),
+    mutate(method = "DLM (L=1)"),
   pred_appl_gfosr_l5 %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=5)") 
+    mutate(method = "DLM (L=5)") 
 ) %>% 
   mutate_at(vars(starts_with("pred")), function(x)exp(x)/(1+exp(x))) %>% 
   mutate(id = paste("ID","=", id)) %>%
+  mutate(method=factor(method, 
+                       levels = c("fGFPCA", "DLM (L=5)", "DLM (L=1)","GLMMadaptive"))) %>%
   ggplot()+
   geom_point(aes(x=sind, y=Y), size = 0.1, alpha = 0.3)+
   geom_line(aes(x=sind, y = pred360, col = "6am"), linewidth=1.1)+

### 2024-12-17 Code/Figures/FiguresSim.R

@@ -46,15 +46,13 @@ load(here("Data/SimN500/SimOutput_fGFPCA.RData"))
 load(here("Data/SimN500/SimOutput_GLMMadaptive.RData"))
 load(here("Data/SimN500/SimOutput_GFOSR_L1.RData"))
 load(here("Data/SimN500/SimOutput_GFOSR_L5.RData"))
+rand_id <- sample(501:600, size = 2)
 
 ## N = 100
 load(here("Data/SimN100/SubSimOutput_fGFPCA.RData"))
 load(here("Data/SimN100/SubSimOutput_GLMMadaptive.RData"))
 load(here("Data/SimN100/SubSimOutput_GFOSR_L1.RData"))
 load(here("Data/SimN100/SubSimOutput_GFOSR_L5.RData"))
-
-# choose 4 subjects to plot
-rand_id <- sample(501:600, size = 2)
 rand_id <- sample(101:200, size = 2)
 
 #### Format ####
@@ -126,6 +124,8 @@ pred_list_gfofr_l5 <- lapply(pred_subset_gfofr_l5, function(x){
 })
 
 
+
+
 #### Figure ####
 # individual prediction tracks
 bind_rows(
@@ -134,14 +134,14 @@ bind_rows(
   pred_list_GLMMad[[1]] %>% filter(id %in% rand_id) %>%
     mutate(method = "GLMMadaptive"),
   pred_list_gfofr_l1[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=1)"),
+    mutate(method = "DLM (L=1)"),
   pred_list_gfofr_l5[[1]] %>% filter(id %in% rand_id) %>%
-    mutate(method = "GFOSR (L=5)") 
+    mutate(method = "DLM (L=5)") 
 ) %>% 
   mutate_at(vars(eta_i, starts_with("pred")), 
             .funs = function(x){exp(x)/(1+exp(x))}) %>%
   mutate(method=factor(method, 
-                       levels = c("fGFPCA", "GFOSR (L=5)", "GFOSR (L=1)","GLMMadaptive"))) %>%
+                       levels = c("fGFPCA", "DLM (L=5)", "DLM (L=1)","GLMMadaptive"))) %>%
   mutate(id = factor(id, levels = unique(id), labels = paste0("Subject ", 1:2))) %>%
   ggplot()+
   geom_point(aes(x=t, y=Y), size = 0.1, alpha = 0.3)+
[…]

### 2024-12-17 Code/SimData.R

@@ -1,7 +1,13 @@
+
+# This scripts include codes to simulated data
+# corresponding to Section 4 in the manuscript
+
 # clear workspace
 rm(list=ls())
+
 # set seed
 set.seed(1090)
+
 # load and, if necessary, install packages used in simulating data 
 # (some of these you'll need for model fitting)
 pckgs <- c("tidyverse","mgcv","refund","lme4")
@@ -12,14 +18,14 @@ invisible(
         })
 )
 
-### Simulate data ###
+#### Simulate data ####
 
 ## basic parameters
 # number of subjects
 N <- 500
 # number of observations per subject (observations per function)
 J <- 1000
-# functional domain, equally spaced grid on 0,1
+# functional domain, equally spaced grid on [0,1]
 sind <- seq(0,1,len=J)
 
 ## simulate outcomes g(E[Y_i(s)]) = \eta_i(s) = f_0(s) + f_1(s)X_i + b_i(s)

### 2024-12-17 Code/Simulation_N100/Sim_Subset_GLMMadaptive.R

@@ -1,3 +1,8 @@
+
+# This script includes code for simulation on the original sample
+# using the competing method GLMMadaptive
+# 
+
 set.seed(1114)
 
 library(here)
